### 2022-06-08 examples/example_2/descr.tex
artificial examples ready

\documentclass{article}
\usepackage[utf8]{inputenc}

\title{Feature effect with variale-size bins}
\author{Vasilis Gkolemis}
\date{June 2021}

\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{float}
\graphicspath{ {.} }

\begin{document}

\maketitle

\section{Toy example}

A simple toy example for illustrating the importance of variable-size feature effect. Let's say that the correct feature effect for the feature of interest is (see \ref{im:1}):

\begin{equation} \label{eq:feature-effect}
  f_{\mathtt{ALE}}(x_s) = c +
  \begin{cases}
    10x_s &,  0 \leq x_s < 5 \\
    -10x_s &, 5 \leq x_s < 10 \\
    0 &, 10 \leq x_s < 100 \\
  \end{cases}
\end{equation}
%
where c is a normalizing constant.

\begin{figure}[!h]\label{im:1}
  \centering
  \includegraphics[width=.7\linewidth]{Figure_1.png}
  \caption{Ground-truth feature effect}
\end{figure}

We generate N = 500 datapoints which are distributed in a special way;
\(\frac{500}{3}\) are in \([0,5]\), \(\frac{500}{3}\) in \([5,10]\)
and \(\frac{500}{3}\) in \([10,100]\). The points are more dense in
the first two piecewise linear parts and more narrow later. We also
make the hypothesis that there are correlations with other features,
the local effect computed on each data points gets an added Gaussian
noise:

\begin{equation} \label{eq:data-effect}
  \frac{\partial f}{ \partial x_s} (x_s) = \epsilon +
  \begin{cases}
    10 &,  0 \leq x_s < 5 \\
    -10 &, 5 \leq x_s < 10 \\
    0 &, 10 \leq x_s < 100 \\
  \end{cases}
\end{equation}

where \(\epsilon \sim \mathcal{N}(0, 2)\). See the local effect of
each data point in figure \ref{im:2}.

\begin{figure}[!h]\label{im:2}
\centering
\includegraphics[width=0.7\linewidth]{Figure_2.png}
\caption{Local effect per point}
\end{figure}

The example is created to highlight the importance of variable bins;
there are two very strong effects in areas \([0,5]\), \([5,10]\), and
then the effect is zero. Ideally, we would like to create only three
bins of different-size, i.e. \([0,5]\), \([5,10]\), \([10,100]\).

\section{Fixed-size}

With fixed-size bins, there are two solutions; If we create large bins
we miss the first two high-resolution effects at the begining (see
fig.\ref{im:fixed-10}). If we create small bins, we capture the
effects at the beginning, but we split effect of \([10,100]\) in many
bins leading to noisy artifacts (see fig \ref{im:fixed-100},
\ref{im:fixed-1000}).

\begin{figure}[!h]
\centering
\includegraphics[width=0.9\linewidth]{Figure_dale_fixed_10.png}
\caption{DALE fixed-size with 10 bins}
\label{im:fixed-10}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=0.9\linewidth]{Figure_dale_fixed_100.png}
\caption{DALE fixed-size with 100 bins}
\label{im:fixed-100}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=0.9\linewidth]{Figure_dale_fixed_1000.png}
\caption{DALE fixed-size with 1000 bins}
\label{im:fixed-1000}
\end{figure}

\section{Variable-size}

With variable-size we can find the optimal bins, given that there is
enough resolution i.e. the maximum number of bins is large enough to
enable creating small bins in the beginning. For example, with maximum
number of bins equal to 10 (see figure \ref{im:var-10}), it is
impossible to capture the effect in the beginning. But with maximum
number of bins \(\geq 20\) (see \ref{im:var-20} \ref{im:var-100}), the
bins are created (almost) perfectly. We also observe that the actual
number of the hyperparamer maximum number of bins is not important;
all values \(\geq 20\) give robust results. Finally, we confirm that
with create the 'correct' bins, we create perfect feature effect
(almost identical with the ground-truth).

\begin{figure}[!h]
\centering
\includegraphics[width=0.9\linewidth]{Figure_dale_var_10.png}
\caption{DALE variable-size with 10 bins}
\label{im:var-10}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=0.9\linewidth]{Figure_dale_var_20.png}
\caption{DALE variable-size with 20 bins}
\label{im:var-20}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=0.9\linewidth]{Figure_dale_var_100.png}
\caption{DALE variable-size with 100 bins}
\label{im:var-100}
\end{figure}

\section{Comparing fixed-size with Variable-size}

We evaluate fixed-size and variable-size with two metrics; Loss and
MSE. Loss is the sum of the standard error of each bin and MSE is the
Mean squared error between the obtained feature effect plot and the
ground-truth. Our goal is to choose the plot with minimum sum of
standard errors (loss) as the best feature effect
estimation. Therefore, we want to check whether minimising the loss is
a good criterion for getting the plot with the smallest MSE. We set a
threshold of mimimum 10 points per bin for ensuring robust standard
error estimations. Therefore we cannot evaluate the loss in cases of
large number of fixed-size bins.

For the loss, in figure \ref{im:loss} we obsere that:

\begin{itemize}
\item for fixed size, the standard error is not a good indicator for
  choosing the correct number of bins. Firstly, we cannot evaluate it
  for large number of bins, since there are not enough points inside
  all bins. Secondly, since the variance inside the bins is big,
  standard error is dominated by the number of samples.
\item for variable size, the sum of standard errors is a valid
  indicator indicator. When loss is minimized (K=20, 40, 60) the MSE
  is also minimized.
\end{itemize}

\begin{figure}[h]
\centering
\includegraphics[width=0.9\linewidth]{Figure_7.png}
\caption{Loss vs number of bins (means number of bins when fixed size, and maximum number of bins when variable size)}
\label{im:loss}
\end{figure}

For the MSE, in figure \ref{im:mse}, we obsere that:

\begin{itemize}
\item for fixed-size, the MSE is smaller for larger number of bins (makes sense)
\item for variable-size, the MSE is smaller for larger maximum number of bins (makes sense)
\item the variable-size creates in general more accurate feature effect plots
\item as the maximum number of bins grows larger we get more robust
  feature effect plots. For example, K=21 leads to large MSE, whereas K=20 is almost optimal. In contrast, K=71 is very close to K=70 which is also almost optimal. In general, for K>50 the result is close to perfect independently of the exact value of K.
\end{itemize}

\begin{figure}[h]
\centering
\includegraphics[width=0.9\linewidth]{Figure_9.png}
\caption{MSE vs number of bins (means number of bins when fixed size, and maximum number of bins when variable size)}
\label{im:mse}
\end{figure}

\section{Best solutions based on standard error}

In figure \ref{im:best-fixed} we see the best fixed-size feature
effect plot based on loss and in figure \ref{im:best-var} the best variable-size based on the same criterion.

\begin{figure}[h]
\centering
\includegraphics[width=0.9\linewidth]{Figure_10.png}
\caption{Best solution, based on standard error, for fixed-size bins is the creation of 2 bins}
\label{im:best-fixed}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=0.9\linewidth]{Figure_11.png}
\caption{Best solution based on standard error for variable-size bins; it creates the 3 correct bins, with maximum number of bins equal to 80.}
\label{im:best-var}
\end{figure}

\section{Conclusion}

It is easy to show that it is important to create plots of
variable-size bins. Sum of standard errors can be an indicator for
choosing the correct plot, but it is not very robust. When the
clusters are not very clear, it tends to choose bigger bins (sometimes
even allocating all points in a single bin). I am still searching for
alternative clustering criteria.

\end{document}

### 2022-06-08 paper/paper_in_bullets.tex
artificial examples ready

\documentclass{article}
\usepackage[utf8]{inputenc}

\title{Idea in bullets}
\author{Vasilis Gkolemis}
\date{June 2021}

\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{float}
\graphicspath{ {./../examples/} }

\begin{document}

\maketitle

\section{Intro}

ALE plots are the best feature effect technique, because
blablabla. However, they have three drawbacks; Firstly, the do not
provide a measure of uncertainty about the effect they calculate (in
other words ). Secondly, they sometimes calculate different depending
on the number of bins (and we do not have a measure to choose which
one is closer to reality). Finally, for making accurate ALE plots, we
sometimes need to use variable-size bins.

\section{Can we trust the ALE plot?}

In ALE, the user defines the bin-size, through one of the following
hyperparameters:

\begin{itemize}
\item \(K\), number of bins
\item \(dx\), bin-size
\item \(\texttt{min\_points\_per\_bin}\)
\end{itemize}
%
which have `1-1' relation:

\begin{equation*}
 dx \Leftrightarrow K \Leftrightarrow \texttt{min\_points\_per\_bin}
\end{equation*}
%
Therefore, we refer to them interchangeably.

\subsubsection*{Statement}

Altering the number of bins \( (K) \), leads to different ALE
plots. Depending on \(K\), we receive quite different ALE without any
indicator which one is the correct. In other words, we need a metric
to evaluate each plot.

\subsubsection*{Example}

We define the following model:

\begin{equation} \label{eq:bullet-1-model}
  f(x_1, x_2) = x_1x_2 +
  \begin{cases}
    -5  + 0.3 x_1 &,  0  \leq x_1 < 20 \\
    1   + 7  (x_1 - 20) &,  20 \leq x_1 < 40 \\
    141 - 1.5(x_1 - 40) &,  40 \leq x_1 < 60 \\
    111 + 0  (x_1 - 60) &,  60 \leq x_1 < 80 \\
    111 - 5  (x_1 - 80) &,  80 \leq x_1 < 100
  \end{cases}
\end{equation}
%
where \(x_1 \perp x_2\) and \(x_2 \sim \mathcal{N}(\mu=0, \sigma=4)\).
Therefore, the gradients wrt. \(x_1\) are:

\begin{equation} \label{eq:bullet-1-data-effect}
  \frac{\partial f}{ \partial x_1} (x_1) = x_2 +
  \begin{cases}
    0.3 &,  0  \leq x_1 < 20 \\
    7   &,  20 \leq x_1 < 40 \\
    1.5 &,  40 \leq x_1 < 60 \\
    0   &,  60 \leq x_1 < 80 \\
    -5  &,  80 \leq x_1 < 100
  \end{cases}
\end{equation}
%
where \(x_2 \sim \mathcal{N}(\mu=0, \sigma=4)\). The ground truth ALE is:

\begin{equation} \label{eq:bullet-1-gt-ale}
  f_{\mathtt{ALE}}(x_s) = c +
  \begin{cases}
    -5  + 0.3 x_s &,  0  \leq x_s < 20 \\
    1   + 7  (x_s - 20) &,  20 \leq x_s < 40 \\
    141 - 1.5(x_s - 40) &,  40 \leq x_s < 60 \\
    111 + 0  (x_s - 60) &,  60 \leq x_s < 80 \\
    111 - 5  (x_s - 80) &,  80 \leq x_s < 100
  \end{cases}
\end{equation}
%
where \(c\) is a normalizing constant.

We generate N = 100 data points uniformly in the region \([0,100]\),
i.e., \(\mathcal{D} \sim \mathcal{U}(0, 100)\). The produced feature
effect plot is shown in figure \ref{fig:bullet-1-im-1}.

We observe that we get different plots for \(K=\{3, 5, 20, 100\}\),
without information which one to trust. If we set a threshold on
\(\texttt{min\_points\_per\_bin}\), then we discard plots (c) and (d)
beacuse they violate the threshold. In this case, among (a) and (b),
we cannot know which one to trust (We cannot trust both, since they
provide different effects).

For \(K=3\), the available resolution is smaller than the one
required, leading to an erroneous estimation. For \(K=5\), the feature
effect matches the correct resolution. For \(K=20, 100\), the feature
effect provides the general trend, but with noisy artifacts due to
small number of points per bin (violating
\texttt{min\_points\_per\_bin}).

\begin{figure}[!h]
  \centering
  \includegraphics[width=.49\linewidth]{bullet_1/im_1.png}
  \includegraphics[width=.49\linewidth]{bullet_1/im_2.png}\\
  \includegraphics[width=.49\linewidth]{bullet_1/im_3.png}
  \includegraphics[width=.49\linewidth]{bullet_1/im_4.png}
  \caption{DALE effect for (a) \(K = 3\), (b) \(K = 5\), (c)
    \(K = 20\), (d) \(K = 100\)}
  \label{fig:bullet-1-im-1}
\end{figure}

\subsubsection*{Proposal}

We propose standard error as a metric for informing \textbf{to what extend
we should trust the feature effect plot}. \textbf{Standard error
  shows the expected error in the computation of the feature effect
  plot.} However, there are some constraints we must notice:

\begin{enumerate}
\item Our computations are based on the hypothesis that inside all
  bins \textbf{the gradient wrt. to the feature of interest doesn't
    depend on the value feature of interest}. In our example, in the
  interval \([0,20)\) the gradient is \(x_2 + 0.3\) (independent of
  \(x_2\)). But in the interval \([0,40)\) the gradient is 0.3 if
  \(x_1 < 20\) and 7 otherwise (not independent of
  \(x_2\)). Unfortunately, we cannot when this is the case. We just
  know that as the bins grow larger, it is more possible to violated
  this hypothesis, as in Figure~\ref{fig:bullet-1-im-2}(a). In this
  case both the ALE effect and the standard error are wrong.
\item The standard error should be trusted when it is estimated by a
  large of data points. For example, in plots (c) and (d), there are
  bins with less than 10 points. Therefore, in these cases, we cannot
  trust the plot or the standard error.
\item For being confident about the feature effect plot, we should
  check the region covered by 2 or 3 times the standard error. For
  example in~\ref{fig:bullet-1-im-2}, we observe that in plots (a) and
  (b), the green region showing the standard error is very big (covers
  the region from almost -100 to 100). Therefore, the effect cannot be
  trusted.
\end{enumerate}

\begin{figure}[!h]
  \centering
  \includegraphics[width=.49\linewidth]{bullet_1/im_5.png}
  \includegraphics[width=.49\linewidth]{bullet_1/im_6.png}\\
  \includegraphics[width=.49\linewidth]{bullet_1/im_7.png}
  \includegraphics[width=.49\linewidth]{bullet_1/im_8.png}
  \caption{DALE effect with standard error for (a) \(K = 3\), (b)
    \(K = 5\), (c) \(K = 20\), (d) \(K = 100\)}
  \label{fig:bullet-1-im-2}
\end{figure}

If the above criteria are met, the standard error is a useful
metric. For example, if we repeat the experiment with a bigger dataset
\( (N = 10000) \) points, we get the results of figure
\ref{fig:bullet-1-im-3}. We observe that standard error gives accurate
trustability regions, in all plots, i.e. (b), (c), (d), apart from
(a). In (a), (b) and (c), it correctly us to trust the feature effect
plots. Misleadingly, it also informs to trust plot (a), which is
wrong. This because in this case, the first constraint has been
violated.

\begin{figure}[!h]
  \centering
  \includegraphics[width=.49\linewidth]{bullet_1/im_9.png}
  \includegraphics[width=.49\linewidth]{bullet_1/im_10.png}\\
  \includegraphics[width=.49\linewidth]{bullet_1/im_11.png}
  \includegraphics[width=.49\linewidth]{bullet_1/im_12.png}
  \caption{DALE effect with standard error for (a) \(K = 3\), (b)
    \(K = 5\), (c) \(K = 20\), (d) \(K = 100\)}
  \label{fig:bullet-1-im-3}
\end{figure}

\subsubsection*{Conclusion}

If the user can create small bins, i.e. (\(dx\) small enough
\(\rightarrow K \) big enough, to respect constraint 1 and has enough
points inside each bin (constraint 2), then the standard error reveals
to what extend we can trust the feature effect plot.

\section{Choose the more accurate feature effect plot to trust}

\subsubsection*{Statement}

Apart from having a metric of uncertainty about the feature effect
plot (standard error), we must also have a metric to choose the most
accurate feature effect plot.

\subsubsection*{Proposal}

We propose the minimization of the accumulated standard deviation (or
accumulated variance). For ALE with K bins, let's notate as:

\begin{itemize}
\item \(dx^K\) the length of each bin
\item \(p_i^K \) the number of the training points inside the \(i-th\) bin
\item \(\sigma_i^K \) the std of the local effects of the training points inside the \(i-th\) bin
\end{itemize}
%
We will minimize:

\begin{align} \label{eq:min-criterion}
  K_{min} &= \text{argmin}_{K} \quad  [dx^K  \sum_i^K \sigma_i^K * (1 - d_i^K)]\\
  & \text{s.t. } p_i^K \geq \texttt{min\_points\_per\_bin } \forall i
\end{align}
%
where \(d_i^K = 0.2 * \frac{p_i^K}{N} \in [0,0.1] \) works as a `discount', favoring the
creation of bigger bins in cases of similar standard deviation.

\subsubsection*{Example}

Let's see how the proposed metric works in the example of Chapter
2. We set \texttt{min\_points\_per\_bin = 10}. We repeat the procedure
for different number of points \(N=50, 100, 1000, 10000\). The results
are shown in \ref{fig:bullet-2-im-1}. For \(N=50\) we can evaluate the
loss only up to 4 bins (3 is the best) and for \(N=100\) up to 7 (5 is
the best). In all other cases 5 bins is the best solution (sensible),
with all multiples of 5 are also good options. All results make sense.

\begin{figure}[!h]
  \centering
  \includegraphics[width=.49\linewidth]{bullet_2/im_1.png}
  \includegraphics[width=.49\linewidth]{bullet_2/im_2.png}\\
  \includegraphics[width=.49\linewidth]{bullet_2/im_3.png}
  \includegraphics[width=.49\linewidth]{bullet_2/im_4.png}
  \caption{Loss (=accumulated standard error) for datasets of size (a)
    \(N = 50\), (b) \(N = 100\), (c) \(N = 1000\), (d) \(N = 10000\)}
  \label{fig:bullet-2-im-1}
\end{figure}

\subsubsection*{Conclusion}

Minimizing the accumulated standard deviation is a good indicator for
choosing the most accurate feature effect plot.

\section{Variable-size bins are important in many cases}

\subsubsection*{Statement}

Spliting the space in equally sized-bins is not always a good option.

\subsubsection*{Example}

Let's see the example of Figure \ref{fig:bullet-3-im-1}, where the
ground truth effect is a piecewise linear function with 3 parts. There
are two problems with applying equally-sized bins:

\begin{itemize}
\item The first two parts have length 20, whereas the third part has
  length 60. For capturing the first two effects we need
  \(dx \leq 20\), but this resolution adds noise to the third bin that
  could be bigger.
\item the data points are not uniformly split along the axis. From
  x=40 until x=100, they are sparse. This makes it difficult to split
  the space, with many bins since we won't have enough points per bin.
\end{itemize}

Therefore as we can see in figure \ref{fig:bullet-3-im-2}, with a
threshold of \(min\_points\_per\_bin = 10\), we can evaluate until
\(K=4\), which has the best loss. As we see, \(K=4\) has lower
resolution than needed, badly estimating feature effect (Figure
\ref{fig:bullet-3-im-2}(b)).

\begin{figure}[!h]
  \centering
  \includegraphics[width=.49\linewidth]{bullet_3/im_1.png}
  \includegraphics[width=.49\linewidth]{bullet_3/im_2.png}\\
  \caption{Left: Ground truth feature effect plot, Right: local effect
    for each point}
  \label{fig:bullet-3-im-1}
\end{figure}

\begin{figure}[!h]
  \centering
  \includegraphics[width=.49\linewidth]{bullet_3/im_3.png}
  \includegraphics[width=.49\linewidth]{bullet_3/im_4.png}\\
  \caption{Left: Ground truth feature effect plot, Right: local effect for each point}
  \label{fig:bullet-3-im-2}
\end{figure}

\subsubsection*{Proposal}

We propose the creation of variable-size bins. The objective is the
minimisation of the loss as described in Section 3. In our example,
the algorithm created 5 bins (figure \ref{fig:bullet-3-im-3} (a)) that
lead to the feature effect of (figure \ref{fig:bullet-3-im-3} (b)),
which is almost perfect.

\begin{figure}[!h]
  \centering
  \includegraphics[width=.49\linewidth]{bullet_3/im_5.png}
  \includegraphics[width=.49\linewidth]{bullet_3/im_7.png}\\
  \caption{Left: Ground truth feature effect plot, Right: local effect for each point}
  \label{fig:bullet-3-im-3}
\end{figure}

\subsubsection*{Conclusion}

\section{Variable-size bins in case of non-linear feautre effects}

\end{document}

### 2022-06-08 examples/bullet_1.py

@@ -1,0 +1,106 @@
+import numpy as np
+import feature_effect as fe
+import examples.example_utils as utils
+
+
+def gen_model_params():
+    params = [{"b": 0, "from": 0., "to": 20.},
+              {"b": 5., "from": 20, "to": 40},
+              {"b": -5, "from": 40, "to": 60},
+              {"b": 1., "from": 60, "to": 80},
+              {"b": -1., "from": 80, "to": 100}]
+    x_start = -5
+    utils.find_a(params, x_start)
+    return params
+
+
+def generate_samples(N):
+    eps = 1e-05
+    x = np.random.uniform(0, 100-eps, size=int(N))
+    x = np.expand_dims(np.concatenate((np.array([0.0]), x, np.array([100. - eps]))), axis=-1)
+    return x
+
+
+example_dir = "./examples/bullet_1/"
+
+# parameters
+N = 80
+noise_level = 4.
+K_max_fixed = 200
+min_points_per_bin = 10
+
+# init functions
+seed = 4834545
+np.random.seed(seed)
+
+# define functions
+model_params = gen_model_params()
+model_jac = utils.create_noisy_jacobian(model_params, noise_level, seed)
+
+# generate data and data effect
+data = np.sort(generate_samples(N=N), axis=0)
+model = utils.create_model(model_params, data)
+y = model(data)
+data_effect = model_jac(data)
+
+# dale without standard error
+dale = fe.DALE(data, model, model_jac=model_jac)
+dale.fit(alg_params={"nof_bins": 3, "min_points_per_bin": min_points_per_bin})
+dale.plot(error=False, savefig=example_dir + "im_1.png")
+
+dale.fit(alg_params={"nof_bins": 5, "min_points_per_bin": min_points_per_bin})
+dale.plot(error=False, savefig=example_dir + "im_2.png")
+
+dale.fit(alg_params={"nof_bins": 20, "min_points_per_bin": min_points_per_bin})
+dale.plot(error=False, savefig=example_dir + "im_3.png")
+
+dale.fit(alg_params={"nof_bins": 100, "min_points_per_bin": min_points_per_bin})
+dale.plot(error=False, savefig=example_dir + "im_4.png")
+
+# dale with standard error
+dale = fe.DALE(data, model, model_jac=model_jac)
+dale.fit(alg_params={"nof_bins": 3, "min_points_per_bin": min_points_per_bin})
[…]

### 2022-06-08 examples/bullet_2.py

@@ -1,0 +1,63 @@
+import numpy as np
+import feature_effect as fe
+import examples.example_utils as utils
+import matplotlib.pyplot as plt
+
+def gen_model_params():
+    params = [{"b":0, "from": 0., "to": 20.},
+              {"b":5., "from": 20, "to": 40},
+              {"b":-5, "from": 40, "to": 60},
+              {"b":1., "from": 60, "to": 80},
+              {"b":-1., "from": 80, "to": 100}]
+    x_start = -5
+    utils.find_a(params, x_start)
+    return params
+
+
+def generate_samples(N):
+    eps = 1e-05
+    x = np.random.uniform(0, 100-eps, size=int(N))
+    x = np.expand_dims(np.concatenate((np.array([0.0]), x, np.array([100. - eps]))), axis=-1)
+    return x
+
+def compute_dale_for_N(N, seed):
+    np.random.seed(seed)
+    data = np.sort(generate_samples(N=N), axis=0)
+    model = utils.create_model(model_params, data)
+    dale_fixed = utils.fit_multiple_K(data, model, model_jac, K_max_fixed, min_points_per_bin, method="fixed-size")
+    return dale_fixed
+
+
+example_dir = "./examples/bullet_2/"
+noise_level = 4.
+K_max_fixed = 200
+min_points_per_bin = 10
+seed = 233
+model_params = gen_model_params()
+model_jac = utils.create_noisy_jacobian(model_params, noise_level, seed)
+
+
+dale_list = []
+# N = 1000
+# dale_list.append(compute_dale_for_N(N, seed=seed))
+# utils.plot_loss(dale_list[-1], savefig=example_dir + "im_3.png")
+
+# for 50 points
+N = 50
+dale_list.append(compute_dale_for_N(N, seed=seed))
+utils.plot_loss(dale_list[-1], savefig=example_dir + "im_1.png")
+
+# for 100 points
+N = 100
+dale_list.append(compute_dale_for_N(N, seed=seed))
+utils.plot_loss(dale_list[-1], savefig=example_dir + "im_2.png")
+
+# for 1000 points
+N = 1000
+dale_list.append(compute_dale_for_N(N, seed=seed))
+utils.plot_loss(dale_list[-1], savefig=example_dir + "im_3.png")
+
+# for 10000 points
+N = 10000
+dale_list.append(compute_dale_for_N(N, seed=seed))
+utils.plot_loss(dale_list[-1], savefig=example_dir + "im_4.png")

### 2022-06-08 examples/bullet_3.py

@@ -43,15 +43,18 @@ def generate_samples(N):
                                        x2,
                                        np.array([41.0, 44., 50., 53, 57]),
                                        x3,
+                                       np.array([87, 92]),
                                        np.array([100-eps]))), axis=-1)
     return x
+
+example_dir = "./examples/bullet_3/"
 
 # experiment parameters
 N = 500
 noise_level = 4.
 K_max_fixed = 40
 K_max_var = 40
-min_points_per_bin = 20
+min_points_per_bin = 10
 
 # set seed
 seed = 4854894
@@ -67,49 +70,60 @@ model = utils.create_model(model_params, data)
 y = model(data)
 data_effect = model_jac(data)
 
-
 # check bin creation
-bin_est = fe.bin_estimation.BinEstimatorDP(data, data_effect, feature=0, K=45)
+bin_est = fe.BinEstimatorDP(data, data_effect, feature=0, K=45)
 limits = bin_est.solve_dp(min_points=min_points_per_bin)
 clusters = np.squeeze(np.digitize(data, limits))
 
 plt.figure()
-plt.title("Dynamic programming")
+plt.title("Local effect")
 plt.plot(data, data_effect, "bo")
-plt.vlines(limits, ymin=np.min(data_effect), ymax=np.max(data_effect))
+# plt.vlines(limits, ymin=np.min(data_effect), ymax=np.max(data_effect))
+plt.savefig(example_dir + "im_2.png")
 plt.show(block=False)
 
 # plot data effects and gt effect
-utils.plot_gt_effect(data, y)
-
-
-dale0 = fe.DALE(data, model, model_jac)
-dale0.fit(method="fixed-size", alg_params={"nof_bins": 301, "min_points_per_bin": min_points_per_bin})
-dale0.plot(gt=model)
-
-
-dale1 = fe.DALE(data, model, model_jac)
-dale1.fit(method="variable-size", alg_params={"max_nof_bins": 40, "min_points_per_bin": min_points_per_bin})
-dale1.plot(gt=model)
+utils.plot_gt_effect(data, y, savefig=example_dir + "im_1.png")
 
 # compute loss and mse for many different K
 dale_fixed = utils.fit_multiple_K(data, model, model_jac, K_max_fixed, min_points_per_bin, method="fixed-size")
+utils.plot_loss(dale_fixed, savefig=example_dir + "im_3.png")
+
[…]

### 2022-06-08 examples/bullet_4.py

@@ -1,0 +1,105 @@
+import copy
+import matplotlib.pyplot as plt
+import numpy as np
+import feature_effect as fe
+
+
+def sinc(k):
+    return np.sin(k)/k
+
+
+def model(x):
+    tau = 2.65
+    w = np.pi
+
+    ind = x[:, 0] > tau
+    y = sinc(w * x[:, 0] ** 2) + x[:, 0] * x[:, 1] * 10
+    y[ind] = sinc(w * tau ** 2) + x[:, 0][ind] * x[:, 1][ind] * 10
+    return y
+
+
+def model_jac(x_tmp):
+    x = copy.deepcopy(x_tmp)
+    tau = 2.65
+    w = np.pi
+
+    ind = x[:, 0] > tau
+
+    y1 = (2 * np.cos(w * x[:, 0] ** 2) / x[:, 0] - np.sin(w * x[:, 0] ** 2) * 2 / w / x[:, 0]**3) + x[:, 1] * 10
+    y1[ind] = x[:, 1][ind] * 10
+
+    y2 = x[:, 0]
+
+    grad = np.stack((y1, y2), axis=-1)
+    return grad
+
+
+def gen_data(N, noise_level):
+    eps = 1e-03
+    stop = 5
+    x1 = np.concatenate((np.random.uniform(0. + eps, 2.65, size=int(N)),
+                         np.array([2.7]),
+                         np.random.normal(3.75, .3, size=int(N/12)),
+                         np.array([4.99])
+                         )
+                        )
+
+    x2 = np.random.normal(loc=0, scale=noise_level, size=(int(x1.shape[0])))
+    x = np.stack((x1, x2), axis=-1)
+    return x
+
+
+def create_gt_effect_norm(data):
+    def gt_effect_un(x_tmp):
+        x = copy.deepcopy(x_tmp)
+        tau = 2.65
+        w = np.pi
+
+        ind = x > tau
+
+        y = sinc(w * x ** 2)
+        y[ind] = sinc(w * tau ** 2)
+        return y
+
+    z = np.mean(gt_effect_un(data[:,0]))
+
+    def tmp(x):
+        return gt_effect_un(x) - z
+    return tmp
+
+example_dir = "./examples/bullet_4/"
+
+np.random.seed(25)
+# generate points
+N = 1000
+noise_level = .1
+data = gen_data(N, noise_level)
+
+gt_effect_norm = create_gt_effect_norm(data)
[…]

### 2022-06-08 examples/example_1.py

@@ -60,17 +60,11 @@ For Variable-size bins:
 
 """
 import numpy as np
+import feature_effect as fe
 import examples.example_utils as utils
+import matplotlib.pyplot as plt
 
-# define piecewise linear function
-def f_params():
-    def find_a(params, x_start):
-        params[0]["a"] = x_start
-        for i, param in enumerate(params):
-            if i < len(params) - 1:
-                a_next = param["a"] + (param["to"] - param["from"]) * param["b"]
-                params[i + 1]["a"] = a_next
-
+def gen_model_params():
     params = [{"b":0.3, "from": 0., "to": 10.},
               {"b":7., "from": 10, "to": 20},
               {"b":-1.5, "from": 20, "to": 30},
@@ -82,7 +76,7 @@ def f_params():
               {"b":0., "from": 80, "to": 90},
               {"b":-5., "from": 90, "to": 100}]
     x_start = -1
-    find_a(params, x_start)
+    utils.find_a(params, x_start)
     return params
 
 
@@ -95,45 +89,62 @@ def generate_samples(N):
 
 # parameters
 N = 10000
-noise_level = 3.
-K_max_fixed = 50
+noise_level = 4.
+K_max_fixed = 40
 K_max_var = 40
-min_points_per_bin = 10
+min_points_per_bin = 20
 
 # init functions
 seed = 4834545
 np.random.seed(seed)
 
-model = utils.create_model(f_params)
-model_jac = utils.create_noisy_jacobian(f_params, noise_level, seed)
-data = generate_samples(N=N)
+# define functions
+model_params = gen_model_params()
+model_jac = utils.create_noisy_jacobian(model_params, noise_level, seed)
+
+# generate data and data effect
+data = np.sort(generate_samples(N=N), axis=0)
+model = utils.create_model(model_params, data)
 y = model(data)
 data_effect = model_jac(data)
 
+# check bin creation
+bin_est = fe.bin_estimation.BinEstimatorDP(data, data_effect, feature=0, K=74)
+limits = bin_est.solve_dp(min_points=min_points_per_bin)
+clusters = np.squeeze(np.digitize(data, limits))
+
+plt.figure()
+plt.title("Dynamic programming")
+plt.plot(data, data_effect, "bo")
+plt.vlines(limits, ymin=np.min(data_effect), ymax=np.max(data_effect))
+plt.show(block=False)
+
 # plot data effects and gt effect
 utils.plot_gt_effect(data, y)
-utils.plot_data_effect(data, data_effect)
+
[…]

### 2022-06-08 examples/example_2.py

@@ -1,5 +1,9 @@
-"""Goal: show that if data non-uniformly distributed, you have to find the K that matches the piecewise linear regions.
-This is not feasible in all cases. Piecewise linear regions, under normal circumstances, are have not equal length.
+"""
+if feature effect is made from variable size piecewise-linear functions:
+- fixed-size approach:
+  - (a) if big-size bins, fails in high-resolution effects
+  - (b) if small-size bins, suffers from instabilities in low-resolution effects
+variable-size finds the bins correctly.
 """
 
 import numpy as np
@@ -16,9 +20,9 @@ def create_model_params():
                 a_next = param["a"] + (param["to"] - param["from"]) * param["b"]
                 params[i + 1]["a"] = a_next
 
-    params = [{"b":10, "from": 0., "to": 5.},
-              {"b":-10, "from": 5., "to": 10.},
-              {"b":0. , "from": 10., "to": 100}]
+    params = [{"b":10, "from": 0., "to": 20.},
+              {"b":-10, "from": 20., "to": 40.},
+              {"b":0. , "from": 40., "to": 100}]
 
     x_start = -25.
     find_a(params, x_start)
@@ -28,65 +32,76 @@ def create_model_params():
 # generate samples
 def generate_samples(N):
     eps = 1e-05
-    x1 = np.random.uniform(0., 5, size=int(N / 3))
-    x2 = np.random.uniform(5, 10, size=int(N / 3))
-    x3 = np.random.uniform(10, 99, size=int(N / 3))
-    x = np.expand_dims(np.concatenate((np.array([0.0]), x1, x2, x3, np.array([100-eps]))), axis=-1)
+    x1 = np.random.uniform(0., 100, size=int(N))
+    x = np.expand_dims(np.concatenate((np.array([0.0]),
+                                       x1,
+                                       np.array([100-eps]))), axis=-1)
     return x
 
 # experiment parameters
-N = 5000
-noise_level = 3.
-K_max_fixed = 30
-K_max_var = 20
-min_points_per_bin = 10
+N = 500
+noise_level = 4.
+K_max_fixed = 40
+K_max_var = 40
+min_points_per_bin = 20
 
 # set seed
-seed = 4834545
+seed = 4854894
 np.random.seed(seed)
 
 # define functions
 model_params = create_model_params()
[…]

### 2022-06-08 examples/example_utils.py

@@ -4,7 +4,16 @@ import matplotlib.pyplot as plt
 import feature_effect as fe
 
 
-def create_model(params):
+def find_a(params, x_start):
+    params[0]["a"] = x_start
+    for i, param in enumerate(params):
+        if i < len(params) - 1:
+            a_next = param["a"] + (param["to"] - param["from"]) * param["b"]
+            params[i + 1]["a"] = a_next
+
+
+
+def create_model(params, data):
     def create_f1(params):
         limits = [param["from"] for param in params]
         limits.append(params[-1]["to"])
@@ -23,7 +32,7 @@ def create_model(params):
         return f1
 
     f1 = create_f1(params)
-    z = np.mean(f1(np.linspace(0.0, 99.99999, 100000)))
+    z = np.mean(f1(data))
 
     def f1_center(x):
         return f1(x) - z
@@ -43,8 +52,8 @@ def create_noisy_jacobian(params, noise_level, seed):
         y = np.array([r['b'] for r in res1])
 
         # add noise
-        rv = norm()
-        noise = rv.rvs(size=y.shape[0], random_state=seed)*noise_level
+        rv = norm(loc=0, scale=noise_level)
+        noise = rv.rvs(size=y.shape[0], random_state=seed)
         return np.expand_dims(y+noise, -1)
     return compute_data_effect
 
@@ -57,11 +66,13 @@ def create_gt_bins(params):
     return gt_bins
 
 
-def compute_mse(dale, gt_func):
-    xx = np.linspace(0., 99.9999, 1000)
+def compute_mse(dale, gt_effect):
+    start = dale.dale_params["feature_0"]["limits"][0]
+    stop = dale.dale_params["feature_0"]["limits"][-1]
+    xx = np.linspace(start, stop, 10000)
     y_pred = dale.eval(xx, s=0)[0]
-    y_gt = gt_func(xx)
-    return np.sum(np.abs(y_pred - y_gt)) * .001
+    y_gt = gt_effect(xx)
+    return np.mean(np.square(y_pred - y_gt))
 
 
 def fit_multiple_K(data, model, model_jac, K_max, min_points_per_bin, method):
@@ -76,13 +87,16 @@ def fit_multiple_K(data, model, model_jac, K_max, min_points_per_bin, method):
     return dale_list
 
 
-def plot_gt_effect(points, y):
+def plot_gt_effect(points, y, savefig=False):
     plt.figure()
[…]

### 2022-06-08 feature_effect/__init__.py

@@ -3,3 +3,4 @@ from feature_effect.dale import DALE
 from feature_effect.ale import ALE
 from feature_effect.pdp import PDP
 from feature_effect.mplot import MPlot
+from feature_effect.bin_estimation import BinEstimatorDP

### 2022-06-08 feature_effect/ale.py

@@ -31,45 +31,8 @@ def compute_ale_parameters(data: np.ndarray, model: np.ndarray, feature: int, al
     # compute parameters
     ale_params = utils.compute_fe_parameters(data[:, feature], data_effect, limits, min_points_per_bin)
     ale_params["method"] = "fixed-size"
+    ale_params["min_points_per_bin"] = min_points_per_bin
     return ale_params
-
-
-# def ale(x: np.ndarray, data: np.ndarray, model: typing.Callable, feature: int, k: int = 100):
-#     """Compute ALE at points x.
-#
-#     Functional implementation of DALE at the s-th feature. Computation is
-#     made on-the-fly.
-#
-#     Parameters
-#     ----------
-#     x: ndarray, shape (N,)
-#       The points to evaluate DALE on
-#     data: ndarray, shape (N,D)
-#       The training set
-#     model: Callable
-#     feature: int
-#       Index of the feature
-#     k: int
-#       number of bins
-#
-#     Returns
-#     -------
-#
-#     """
-#     # compute
-#     parameters = compute_ale_parameters(data, model, feature, k)
-#     y = utils.compute_accumulated_effect(x,
-#                                          limits=parameters["limits"],
-#                                          bin_effect=parameters["bin_effect"],
-#                                          dx=parameters["dx"])
-#     y -= parameters["z"]
-#     var = utils.compute_accumulated_effect(x,
-#                                            limits=parameters["limits"],
-#                                            bin_effect=parameters["bin_estimator_variance"],
-#                                            dx=parameters["dx"],
-#                                            square=True)
-#     return y, var
-
 
 class ALE:
     def __init__(self, data: np.ndarray, model: typing.Callable):
@@ -78,7 +41,7 @@ class ALE:
 
         self.data_effect = None
         self.feature_effect = None
-        self.parameters = None
+        self.ale_params = None
 
     @staticmethod
     def _ale_func(params):
@@ -145,6 +108,15 @@ class ALE:
     def eval(self, x: np.ndarray, s: int):
         return self.feature_effect["feature_" + str(s)](x)
 
-    def plot(self, s: int, block=False, gt=None, gt_bins=None):
+    def plot(self, s: int = 0, std: bool = True, block=False, gt=None, gt_bins=None):
         title = "ALE: Effect of feature %d" % (s + 1)
[…]

### 2022-06-08 feature_effect/bin_estimation.py

@@ -3,16 +3,16 @@ import feature_effect.utils as utils
 
 
 class BinEstimatorDP:
-    def __init__(self, data, data_effect, model, feature, K):
+    def __init__(self, data, data_effect, feature, K):
         self.x_min = np.min(data[:, feature])
         self.x_max = np.max(data[:, feature])
         self.K = K
         self.dx = (self.x_max - self.x_min) / K
         self.big_M = 1.e+10
         self.data = data
+        self.nof_points = data.shape[0]
         self.data_effect = data_effect
         self.feature = feature
-        self.model = model
 
         self.limits = None
         self.dx_list = None
@@ -32,7 +32,9 @@ class BinEstimatorDP:
         if data_effect.size < min_points:
             cost = big_cost
         else:
-            cost = np.std(data_effect) * (stop-start) / np.sqrt(data_effect.size)
+            # cost = np.std(data_effect) * (stop-start) / np.sqrt(data_effect.size)
+            discount_for_more_points = (1 - .3*(data_effect.size / self.nof_points))
+            cost = np.var(data_effect) * (stop-start) * discount_for_more_points
         return cost
 
     def _index_to_position(self, index_start, index_stop):

### 2022-06-08 feature_effect/dale.py

@@ -32,7 +32,6 @@ def compute_dale_parameters(data: np.ndarray, data_effect: np.ndarray, feature: 
       - z: float, the normalizer
     """
     assert method in ["fixed-size", "variable-size"]
-
     if method == "fixed-size":
         # alg params
         K = alg_params["nof_bins"] if "nof_bins" in alg_params.keys() else 30
@@ -44,19 +43,28 @@ def compute_dale_parameters(data: np.ndarray, data_effect: np.ndarray, feature: 
         # compute dale params
         dale_params = utils.compute_fe_parameters(data[:, feature], data_effect[:, feature], limits, min_points_per_bin)
         dale_params["method"] = method
+        dale_params["min_points_per_bin"] = min_points_per_bin
+        dale_params["nof_bins"] = K
     elif method == "variable-size":
         # alg params
         K = alg_params["max_nof_bins"] if "max_nof_bins" in alg_params.keys() else 30
         min_points_per_bin = alg_params["min_points_per_bin"] if "min_points_per_bin" in alg_params.keys() else 10
-        bin_estimator = be.BinEstimatorDP(data, data_effect, None, feature, K)
 
         # estimate bins
-        limits = bin_estimator.solve_dp(min_points_per_bin)
+        if "limits" in alg_params.keys():
+            limits = alg_params["limits"]
+        else:
+            bin_estimator = be.BinEstimatorDP(data, data_effect, feature, K)
+            limits = bin_estimator.solve_dp(min_points_per_bin)
 
         # compute dale parameters
         dale_params = utils.compute_fe_parameters(data[:, feature], data_effect[:, feature], limits, min_points_per_bin)
-        dale_params["bin_estimator"] = bin_estimator
+        if "limits" not in alg_params.keys():
+            dale_params["bin_estimator"] = bin_estimator
+
         dale_params["method"] = method
+        dale_params["max_nof_bins"] = K
+        dale_params["min_points_per_bin"] = min_points_per_bin
     return dale_params
 
 
@@ -88,12 +96,18 @@ class DALE:
             y = utils.compute_accumulated_effect(x, limits=params["limits"], bin_effect=params["bin_effect"],
                                                  dx=params["dx"])
             y -= params["z"]
-            var = utils.compute_accumulated_effect(x,
+            estimator_var = utils.compute_accumulated_effect(x,
                                                    limits=params["limits"],
                                                    bin_effect=params["bin_estimator_variance"],
                                                    dx=params["dx"],
[…]

### 2022-06-08 feature_effect/utils.py

@@ -315,12 +315,14 @@ def compute_normalizer(limits: np.ndarray, bin_effect: np.ndarray, x):
 
 
 def compute_loss(points_per_bin, bin_variance_nans, dx, min_points_per_bin):
-    big_m = 1.e+3
     if np.sum(points_per_bin < min_points_per_bin) > 0:
-        error = big_m
+        error = np.NaN
     else:
-        error_per_bin = np.sqrt(bin_variance_nans) * dx / np.sqrt(points_per_bin)
-        error = np.sum(error_per_bin)
+        discount_for_more_points = .2*(points_per_bin / np.sum(points_per_bin))
+        error_per_bin = bin_variance_nans * (1. - discount_for_more_points) * dx
+        error = np.sqrt(np.sum(error_per_bin))
+        # error_per_bin = np.sqrt(bin_variance_nans) * dx / np.sqrt(points_per_bin)
+        # error = np.sum(error_per_bin)
     return error
 
 
@@ -332,7 +334,7 @@ def compute_fe_parameters(data, data_effect, limits, min_points_per_bin):
     bin_effect_nans, points_per_bin = compute_bin_effect_mean(data, data_effect, limits)
 
     # add empty bins
-    is_bin_empty = bin_effect_nans == np.NaN
+    is_bin_empty = np.isnan(bin_effect_nans)
 
     # compute effect variance in each bin
     bin_variance_nans, bin_estimator_variance_nans = compute_bin_effect_variance(data, data_effect, limits, bin_effect_nans)
@@ -347,7 +349,6 @@ def compute_fe_parameters(data, data_effect, limits, min_points_per_bin):
 
     loss = compute_loss(points_per_bin, bin_variance_nans, dx, min_points_per_bin)
 
-    # TODO fix compute Z
     z = compute_normalizer(limits, bin_effect, data)
 
     parameters = {"nof_bins": limits.shape[0] - 1,

### 2022-06-08 feature_effect/visualization.py

@@ -2,18 +2,23 @@ import matplotlib.pyplot as plt
 import numpy as np
 
 
-def feature_effect_plot(params, eval, feature, title=None, block=False, gt=False, gt_bins=None):
+def feature_effect_plot(params, eval, feature, error, min_points_per_bin, title=None, block=False, gt=False, gt_bins=None, savefig=False):
     assert all(name in params for name in ["first_empty_bin", "limits", "dx", "is_bin_empty", "bin_estimator_variance", "bin_effect"])
-
-    first_empty_bin = params["first_empty_bin"]
     limits = params["limits"]
     dx = params["dx"]
-    is_bin_empty = params["is_bin_empty"]
-    bin_estimator_variance = params["bin_estimator_variance"]
+    bin_variance = params["bin_variance"]
     bin_effect = params["bin_effect"]
 
+    is_bin_empty = params["points_per_bin"] < min_points_per_bin
+
+    bins_under_limit = np.argwhere(params["points_per_bin"] < min_points_per_bin)
+    if bins_under_limit.shape[0] > 0:
+        first_empty_bin = bins_under_limit[0][0]
+    else:
+        first_empty_bin = None
+
     x = np.linspace(params["limits"][0], params["limits"][-1], 10000)
-    y, var = eval(x, feature)
+    y, estimator_var, var = eval(x, feature)
 
     fig, (ax1, ax2) = plt.subplots(2, 1, sharex=True)
 
@@ -23,54 +28,96 @@ def feature_effect_plot(params, eval, feature, title=None, block=False, gt=False
         fig.suptitle(title)
 
     # first subplot
-    feature_effect(ax1, x, y, var, limits, first_empty_bin, gt=gt)
+    feature_effect(ax1, x, y, estimator_var, var, first_empty_bin, limits, min_points_per_bin, error=error, gt=gt)
 
     # second subplot
-    effects_per_bin(ax2, bin_effect, bin_estimator_variance, is_bin_empty, limits, dx, gt_bins)
+    effects_per_bin(ax2, bin_effect, bin_variance, error, is_bin_empty,
+                    limits, dx, gt_bins, min_points_per_bin)
+
+    if savefig:
+        plt.savefig(savefig)
+
     if block is False:
         plt.show(block=False)
     else:
         plt.show()
 
 
-def feature_effect(ax1, x, y, var, limits, first_empty_bin, gt=None):
+def feature_effect(ax1, x, y, estimator_var, var, first_empty, limits, point_limit, error=True, gt=None):
[…]

### 2022-06-08 tests/test_bin_estimation.py

@@ -1,0 +1,64 @@
+import pytest
+import numpy as np
+import examples.example_utils as example_utils
+import feature_effect.utils as utils
+import matplotlib.pyplot as plt
+import feature_effect as fe
+
+def gen_model_params():
+    params = [{"b":0.3, "from": 0., "to": 10.},
+              {"b":7., "from": 10, "to": 20},
+              {"b":-1.5, "from": 20, "to": 30},
+              {"b":0., "from": 30, "to": 40},
+              {"b":-5., "from": 40, "to": 50},
+              {"b":0.3, "from": 50, "to": 60},
+              {"b":7., "from": 60, "to": 70},
+              {"b":-1.5, "from": 70, "to": 80},
+              {"b":0., "from": 80, "to": 90},
+              {"b":-5., "from": 90, "to": 100}]
+    x_start = -1
+    example_utils.find_a(params, x_start)
+    return params
+
+
+def generate_samples(N):
+    eps = 1e-05
+    x = np.random.uniform(0, 100-eps, size=int(N))
+    x = np.expand_dims(np.concatenate((np.array([0.0]), x, np.array([100. - eps]))), axis=-1)
+    return x
+
+
+# parameters
+N = 10000
+noise_level = 4.
+K_max_fixed = 40
+K_max_var = 40
+min_points_per_bin = 10
+
+# init functions
+seed = 4834545
+np.random.seed(seed)
+
+# define functions
+model_params = gen_model_params()
+model_jac = example_utils.create_noisy_jacobian(model_params, noise_level, seed)
+
+# generate data and data effect
+data = np.sort(generate_samples(N=N), axis=0)
+model = example_utils.create_model(model_params, data)
+y = model(data)
+data_effect = model_jac(data)
+
+
+# example_utils.plot_data_effect(data, data_effect)
+
+# check bin creation
+bin_est = fe.bin_estimation.BinEstimatorDP(data, data_effect, feature=0, K=75)
+limits = bin_est.solve_dp(min_points=10)
+clusters = np.squeeze(np.digitize(data, limits))
+
+plt.figure()
+plt.title("Dynamic programming")
+plt.plot(data, data_effect, "bo")
+plt.vlines(limits, ymin=np.min(data_effect), ymax=np.max(data_effect))
+plt.show(block=False)

### 2022-06-08 tests/test_functional.py

@@ -217,12 +217,12 @@ class TestExample1:
 
         # feature 1
         x = np.linspace(0, 1, 1000)
-        pred, _ = dale.eval(x, s=0)
+        pred, _, _ = dale.eval(x, s=0)
         gt = self.ale(x)
         assert np.allclose(pred, gt, atol=1.e-2)
 
         # feature 2
-        pred, _ = dale.eval(x, s=1)
+        pred, _, _ = dale.eval(x, s=1)
         gt = self.ale(x)
         assert np.allclose(pred, gt, atol=1.e-2)
 
@@ -374,11 +374,11 @@ class TestExample2:
 
         # feature 1
         x = np.linspace(0, 1, 1000)
-        pred, _ = dale.eval(x, s=0)
+        pred, _, _ = dale.eval(x, s=0)
         gt = self.ale_1(x)
         assert np.allclose(pred, gt, atol=self.atol)
 
         # feature 2
-        pred, _ = dale.eval(x, s=1)
+        pred, _, _ = dale.eval(x, s=1)
         gt = self.ale_2(x)
         assert np.allclose(pred, gt, atol=self.atol)

### 2022-06-10 acml/acml22_paper.tex
started acml paper

%\documentclass[wcp,gray]{jmlr} % test grayscale version
\documentclass[wcp]{jmlr}

% givasile packages
\usepackage{bbm}
\usepackage{xfrac}

% givasile commands
\newcommand{\xc}{\mathbf{x}_c}
\newcommand{\Xc}{\mathbf{X}_c}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Ysloc}{Y_{s}^{\text{local}}}
\newcommand{\ysloc}{y_{s}^{\text{local}}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\xii}{\mathbf{x}^{(i)}}
\newcommand{\yi}{y^{(i)}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\var}{\mathrm{Var}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\1}{\mathbbm{1}}
\newcommand{\D}{\mathcal{D}}

% The following packages will be automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e

%\usepackage{rotating}% for sideways figures and tables
\usepackage{longtable}% for long tables

% The booktabs package is used by this sample document
% (it provides \toprule, \midrule and \bottomrule).
% Remove the next line if you don't require it.
\usepackage{booktabs}
% The siunitx package is used by this sample document
% to align numbers in a column by their decimal point.
% Remove the next line if you don't require it.
%\usepackage[load-configurations=version-1]{siunitx} % newer version
%\usepackage{siunitx}
%\usepackage{natbib}

% Do not comment the following commands:
\pagenumbering{gobble}
\newcommand{\cs}[1]{\texttt{\char`\\#1}}
\makeatletter
\let\Ginclude@graphics\@org@Ginclude@graphics 
\makeatother

\jmlrvolume{}
\jmlryear{2022}
\jmlrworkshop{ACML 2022}

\title[Short Title]{Full Title of Article}

 % Use \Name{Author Name} to specify the name.
 % If the surname contains spaces, enclose the surname
 % in braces, e.g. \Name{John {Smith Jones}} similarly
 % if the name has a "von" part, e.g \Name{Jane {de Winter}}.
 % If the first letter in the forenames is a diacritic
 % enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}

 % Two authors with the same address
 % \author{\Name{Author Name1} \Email{abc@sample.com}\and
 %  \Name{Author Name2} \Email{xyz@sample.com}\\
 %  \addr Address}

 % Three or more authors with the same address:
 % \author{\Name{Author Name1} \Email{an1@sample.com}\\
 %  \Name{Author Name2} \Email{an2@sample.com}\\
 %  \Name{Author Name3} \Email{an3@sample.com}\\
 %  \Name{Author Name4} \Email{an4@sample.com}\\
 %  \Name{Author Name5} \Email{an5@sample.com}\\
 %  \Name{Author Name6} \Email{an6@sample.com}\\
 %  \Name{Author Name7} \Email{an7@sample.com}\\
 %  \Name{Author Name8} \Email{an8@sample.com}\\
 %  \Name{Author Name9} \Email{an9@sample.com}\\
 %  \Name{Author Name10} \Email{an10@sample.com}\\
 %  \Name{Author Name11} \Email{an11@sample.com}\\
 %  \Name{Author Name12} \Email{an12@sample.com}\\
 %  \Name{Author Name13} \Email{an13@sample.com}\\
 %  \Name{Author Name14} \Email{an14@sample.com}\\
 %  \addr Address}

 % Authors with different addresses:
  \author{\Name{Author Name1} \Email{abc@sample.com}\\
  \addr Address 1
  \AND
  \Name{Author Name2} \Email{xyz@sample.com}\\
  \addr Address 2
 }

\editors{Emtiyaz Khan and Mehmet Gonen}

\begin{document}

\maketitle

\begin{abstract}
This is the abstract for this article.
\end{abstract}
\begin{keywords}
List of keywords separated by semicolon.
\end{keywords}

\section{Introduction}
Main contents here.

\section{Related Work}

\section{Probabilistic ALE plots}

\subsection{The ALE method}\label{sec:ALE}

In this section, we introduce the reader to the feature effect method
ALE. Given that \(f: \R^S \rightarrow \R \) is known, we can measure
the \textit{local} effect of the \(s\)-th feature at a specific point
\( \x = (\xc, x_s)\) of the input space \(\mathcal{X}\), as
\(f_s(\x) = \sfrac{\partial f(\x)}{\partial x_s}\). ALE models the
\textit{local} feature effect at \(x_s\) as an expectation over the
distribution of the unkown (latent) features
\(\E_{p(\Xc;x_s)}[f_s(x_s, \xc)]\). Afterwards, ALE measures the
\textit{global} effect at \(x_s\) as an accumulation of the expected
\textit{local} effects:
%
\begin{equation}
  \label{eq:ale-definition}
  f_{\mathtt{ALE}}(x_s) =
  \int_{x_{s, \text{min}}}^{x_s} \E_{p(\Xc;x_s=z)}[f_s(x_s, \xc)] \partial z
\end{equation}

In real cases, it is infeasible to compute
eq.~\eqref{eq:ale-definition} analytically. Therefore, we reside on
estimating the effect from the training set. Let's denote the
available dataset as \(\D = \{\xii, \yi \}_{i=1}^N\), where \( \xii \)
is the \(i\)-th feature vector of the training data and \(\yi\) is the
\(i\)-th label. Apley et. al proposed spliting the axis into \( K \)
equal-sized bins, find the set of points that lie in each bin, i.e.
\( \mathcal{S}_k = \{ \xii : x_s^{(i)} \in [z_{k-1}, z_k) \} \) and,
finally, find the local effect at each bin as the mean value of the
population, i.e.
\(\hat{\mu}_{s,k} = \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in
  \mathcal{S}_k} [f_s(\xii)]\). The global effect at \(x_s\) is then
estimated through:
%
\begin{equation}
  \label{eq:ale-approximation}
  \hat{f}_{\mathtt{ALE}}^s(x)
  = \Delta x \sum_{k=1}^{k_{x_s}} \hat{\mu}_{s,k}
  = \Delta x \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in \mathcal{S}_k} [f_s(\xii)]
\end{equation}
%
where \(\Delta x\) is the bin size and \(k_{x_s}\) the bin-index of the value \(x_s\).
%
%
\subsection{ALE plots with uncertainty quantification}

ALE directly defines the local effect as an expectation over the
unknown features and the global effect as integral over the
expectations. Therefore, both local and global effects are modeled as
simple variables. In constrast, for integrating ALE plots in a
probabilistic framework, we model the local effect as a random
variable, dependent on the random variable \(\Xc\) representing the
latent features:

\begin{equation}
    \Ysloc \sim p(\ysloc ; x_s)
    = \int_{\xc} \delta(\ysloc-f_s(\x)) p(\xc;x_s) \partial \xc
    \label{eq:local-effect-rv}
\end{equation}
%
Following the core idea of ALE, we define the global feature effect as
the random variable \(Y_s\) which is the integration of the local
effects:

\begin{equation}
  Y_s \sim p(y_s ; x_s) = \int_{x_{s,min}}^{x_s} p(\ysloc;x_s=z) \partial z
  \label{eq:global-effect-rv}
\end{equation}
%
Through eqs.~(\ref{eq:local-effect-rv}), (\ref{eq:global-effect-rv})
we formulate the ALE in a fully probabilistic manner. We notice, that
the definition of ALE as given by~\cite{Apley2020} is simply the
expected value of \(Y_s\), i.e.
\(f_{\mathtt{ALE}}(x_s) = \E[Y_s ; x_s]\). For simpler notation we
denote \(\E[Y_s ; x_s]\) as \(\mu_s (x_s)\).

For measuring the uncertainty, it is also important to capture the
variance of the global effect. Notating the variance of the feature
effect \(\var[Y_s ; x_s]\) as \(\sigma_s^2\), we can compute it
through:

\begin{equation}
  \sigma_s^2 (x_s)
  = \int_{x_{s,min}}^{x_s} \int_{\xc} p(\xc|x_s) (f_s(\x) - \mu_s(x_s))^2 \partial \xc (\partial z)^2
\end{equation}

As before, it is infeasible to compute the the variance
\(\sigma_s^2(x_s)\) in closed-form. Therefore, we reside on estimating
them from the available examples of the training set. The estimation
of the variance can also done through the available points:

\begin{equation}
  \label{eq:effect-var-approx}
  \hat{\sigma}_s^2(x_s)
  = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \hat{\sigma}_{s,k}^2
  = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in \mathcal{S}_k} (f_s(\xii) - \hat{\mu}_{s,k})^2 
\end{equation}

\subsection{Unsupervised metric for assessing the quality of ALE plots}

\subsection{ALE plots with variable-size bins}

\subsection{Variable-size bins as an optimization problem}

\section{Experiments}

\subsection{Synthetic Data sets}

A figure in Fig.~\ref{fig:spiral}. Please use high quality graphics
for your camera-ready submission -- if you can use a vector graphics
format such as \texttt{.eps} or \texttt{.pdf}.
\begin{figure}[htp]
\begin{center}
\includegraphics[width=0.5\textwidth]{spiral.eps}
\caption{A spiral.}\label{fig:spiral}
\end{center}
\end{figure}

An example of citation~\cite{DBLP:conf/acml/2009}.

\subsection{Real Data sets}

\section{Conclusion}

\subsection{Subsection Title}

%\acks{Acknowledgements should go at the end, before appendices and references. You can uncomment this for the camera-ready version on paper acceptance.}

%\bibliographystyle{plain}
\bibliography{acml22}

\appendix

\section{Derivations}\label{apd:first}

Derivations

\begin{multline}
    \Ysloc \sim p(\ysloc ; x_s)
    = \int{} p(\ysloc|\xc;x_s) p(\xc;x_s) \partial \xc \\
    = \int_{\xc} \delta(\ysloc-f_s(\x)) p(\xc;x_s) \partial \xc
    = \int_{\xc} \1(\ysloc=f_s(\x)) p(\xc;x_s) \partial \xc
\end{multline}

Some statistics for the local feature effect:

\begin{equation}
  \E[ \Ysloc ; x_s] = \int_{\xc} p(\xc|x_s) f_s(\x) \partial \xc
\end{equation}

\begin{equation}
  \var[ \Ysloc ; x_s] = \int_{\xc} p(\xc|x_s) (f_s(\x) - \E[ \Ysloc ; x_s])^2 \partial \xc
\end{equation}

Some statistics for the global feature effect:

\begin{equation}
  \E[Y_s ; x_s] = \int_{x_{s,min}}^{x_s} \E(\ysloc;z) \partial z
\end{equation}

\begin{equation}
  \var[Y_s ; x_s] = \int_{x_{s,min}}^{x_s} \var[ \Ysloc ; x_s] (\partial z)^2
\end{equation}

\section{Second Appendix}\label{apd:second}

This is the second appendix.

\end{document}

### 2022-06-10 paper/paper_in_bullets.tex
started acml paper

@@
 
-ALE plots are the best feature effect technique, because
-blablabla. However, they have three drawbacks; Firstly, the do not
-provide a measure of uncertainty about the effect they calculate (in
-other words ). Secondly, they sometimes calculate different depending
-on the number of bins (and we do not have a measure to choose which
-one is closer to reality). Finally, for making accurate ALE plots, we
-sometimes need to use variable-size bins.
+ALE plots are the best feature effect (FE) technique, because ...
+However, they have three drawbacks; Firstly, they do not provide
+uncertainty about the measured effect, i.e.\ how confident to be that
+the provided effect is the correct one. Secondly, in many cases, the
+provide different effect depending on the number of bins (and we do
+not have a measure to choose which one is closer to reality). Finally,
+there are cases where it is impossible to create accurate ALE plots,
+with fixed-size bins.
 
@@
 %
-which have `1-1' relation:
+There is `1-1' relation between these hyperparameters:
 
 \begin{equation*}
- dx \Leftrightarrow K \Leftrightarrow \texttt{min\_points\_per\_bin}
+ dx \leftrightarrow K \leftrightarrow \texttt{min\_points\_per\_bin}
 \end{equation*}
@@
 
-Altering the number of bins \( (K) \), leads to different ALE
-plots. Depending on \(K\), we receive quite different ALE without any
-indicator which one is the correct. In other words, we need a metric
-to evaluate each plot.
+Altering the number of bins \( (K) \) leads to different ALE plots and
+we do not have any indicator which plot is closest to the truth.
 
@@
 
+\subsubsection*{Statement}
+
+The combination of non-linear and linear parts makes variable-size
+feature effect even more important.
+
+\begin{itemize}
+\item If we have enough samples, we need small bins to approach
+  accurately the non-linear parts.
+\item We need wider bins for robust estimation of the linear parts
+\end{itemize}
+
+\subsubsection*{Example}
+
+We define the following model:
+
+\begin{equation} \label{eq:bullet-4-model}
+  f(x_1, x_2) = 10x_1x_2 +
+  \begin{cases}
+    sinc(\pi x_1^2)  &,  0  \leq x_1 < \tau \\
+    sinc(\pi \tau^2) &, \tau \leq x_1 \leq 5
+  \end{cases}
+\end{equation}
+%
+where \(x_1 \perp x_2 \Rightarrow p(x_1, x_2) = p(x_1)p(x_2)\) and \(\tau = 2.65\). The data generation distribution for \(x_1\) is:
+
+\begin{equation} \label{eq:bullet-4-generative-x-1}
+  p(x_1) =
+  \begin{cases}
+    \mathcal{U}(x_1; 0 , \tau)/2  &,  0  \leq x_1 < \tau \\
+    \mathcal{N}(x_1; \mu = (5 + \tau)/2 , \sigma=0.3)/2 &, \tau \leq x_1 \leq 5
+  \end{cases}
+\end{equation}
+
+and for \(x_2\):
+
+\begin{equation} \label{eq:bullet-4-generative-x-2}
+  p(x_2) = \mathcal{N}(x_2; \mu = 0 , \sigma=0.1)
+\end{equation}
+
+Therefore, the gradients wrt. \(x_1\) are:
+
+\begin{equation} \label{eq:bullet-4-data-effect}
+  \frac{\partial f}{ \partial x_1} (x_1) = 10x_2 +
+  \begin{cases}
+    2 \frac{\cos(\pi x_1^2)}{x_1} - 2 \frac{\sin(\pi x_1^2)}{\pi x_1^3}  &,  0  \leq x_1 < \tau \\
+    0   &, \tau \leq x_1 \leq 5
+  \end{cases}
+\end{equation}
+%
+where \(x_2 \sim \mathcal{N}(\mu=0, \sigma=4)\). The ground truth ALE is:
+
+\begin{equation} \label{eq:bullet-4-gt-ale}
+  f_{\mathtt{ALE}}(x_s) = c +
+  \begin{cases}
+    sinc(\pi x_1^2)  &,  0  \leq x_1 < \tau \\
+    sinc(\pi \tau^2)  &,  0  \leq x_1 < \tau \\
+  \end{cases}
+\end{equation}
+%
+where \(c\) is a normalizing constant. We generate N = 1000 data
+points. In figure~\ref{fig:bullet-4-im-1}, we see the data points with
+their respective gradients. In figure~\ref{fig:bullet-4-im-2}, we see
+the fixed-size ALE plots; if we set \(K=50\), we capture the
+non-linear parts in the first half, but we fail in linear second
+part. Furthermore, we have empty bins, so we cannot trust the plot and
+the standard error on the second half. The largest number of bins with
+more than ten points in each bin, is \(K=5\) which fails to capture
+the high-resolution effects. Variable size bins solves both problems
+and provides the correct feature-effect.
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=.49\linewidth]{bullet_4/im_1.png}
+  \caption{Data points with their gradients}
+  \label{fig:bullet-4-im-1}
+\end{figure}
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=.48\linewidth]{bullet_4/im_2.png}
+  \includegraphics[width=.48\linewidth]{bullet_4/im_3.png}
+  \caption{Fixed-size bins feature effect. Left: with K=50 and Right: with K=5.}
+  \label{fig:bullet-4-im-2}
+\end{figure}
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=.75\linewidth]{bullet_4/im_4.png}
+  \caption{Variable-size bins.}
+  \label{fig:bullet-4-im-3}
+\end{figure}
+
+\subsubsection*{Proposal}
+
+\subsubsection*{Conclusion}
+
 \end{document}

### 2022-06-10 examples/bullet_4.py

@@ -99,7 +99,7 @@ dale2.fit(method="variable-size", alg_params={"max_nof_bins": 50})
 dale2.plot(gt=gt_effect_norm, savefig=example_dir + "im_4.png")
 
 
-limits = np.concatenate((np.linspace(0, 2.65, 99), np.array([5.01])))
-dale3 = fe.DALE(data, model, model_jac)
-dale3.fit(features=[0], method="variable-size", alg_params={"limits": limits})
-dale3.plot(gt=gt_effect_norm)
+# limits = np.concatenate((np.linspace(0, 2.65, 99), np.array([5.01])))
+# dale3 = fe.DALE(data, model, model_jac)
+# dale3.fit(features=[0], method="variable-size", alg_params={"limits": limits})
+# dale3.plot(gt=gt_effect_norm)

### 2022-06-14 acml/acml22_paper.tex
continued part of the structure

@@
 \newcommand{\xii}{\mathbf{x}^{(i)}}
+\newcommand{\xsi}{x_s^{(i)}}
+\newcommand{\xci}{\mathbf{x}_c^{(i)}}
 \newcommand{\yi}{y^{(i)}}
@@
 
+The decision about the bin size has major influence in the feature
+effect plot. Therefore, it is very important (a) to inform the user to
+what extend they should trust the ALE plot and, consequently, decide
+the bin size with the optimal effect. In the end, we will see that it i
+
+\newpage
+~\newpage
+
 \section{Related Work}
 
-\section{Probabilistic ALE plots}
+\newpage
 
-\subsection{The ALE method}\label{sec:ALE}
+\section{Probabilistic formulation of ALE plots}
 
-In this section, we introduce the reader to the feature effect method
-ALE. Given that \(f: \R^S \rightarrow \R \) is known, we can measure
-the \textit{local} effect of the \(s\)-th feature at a specific point
-\( \x = (\xc, x_s)\) of the input space \(\mathcal{X}\), as
+\subsection{ALE background}\label{sec:ALE}
+
+This section introduces the reader to the ALE formulation of feature
+effect. Given that \(f: \R^S \rightarrow \R \) is known, we can
+measure the effect of the \(s\)-th feature at a specific point
+\( \x = (\xc, x_s)\) of the input space \(\mathcal{X}\) as
 \(f_s(\x) = \sfrac{\partial f(\x)}{\partial x_s}\). ALE models the
-\textit{local} feature effect at \(x_s\) as an expectation over the
-distribution of the unkown (latent) features
-\(\E_{p(\Xc;x_s)}[f_s(x_s, \xc)]\). Afterwards, ALE measures the
-\textit{global} effect at \(x_s\) as an accumulation of the expected
-\textit{local} effects:
+\textit{local} effect at \(x_s\) as the expected effect over the
+distribution of the unkown (latent) features \(\Xc\), i.e.
+\(\E_{p(\xc;x_s)}[f_s(x_s, \xc)]\). Afterwards, the \textit{global}
+effect at \(x_s\) is the accumulation of the \textit{local} effects:
 %
@@
   f_{\mathtt{ALE}}(x_s) =
-  \int_{x_{s, \text{min}}}^{x_s} \E_{p(\Xc;x_s=z)}[f_s(x_s, \xc)] \partial z
+  \int_{x_{s, \text{min}}}^{x_s} \E_{p(\xc;x_s=z)}[f_s(x_s, \xc)] \partial z
 \end{equation}
@@
 is the \(i\)-th feature vector of the training data and \(\yi\) is the
-\(i\)-th label. Apley et. al proposed spliting the axis into \( K \)
-equal-sized bins, find the set of points that lie in each bin, i.e.
-\( \mathcal{S}_k = \{ \xii : x_s^{(i)} \in [z_{k-1}, z_k) \} \) and,
-finally, find the local effect at each bin as the mean value of the
-population, i.e.
-\(\hat{\mu}_{s,k} = \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in
-  \mathcal{S}_k} [f_s(\xii)]\). The global effect at \(x_s\) is then
-estimated through:
+\(i\)-th label.~\cite{Apley2020} proposed spliting the axis into
+\( K \) equal-sized bins, find the set of points that lie in each bin,
+i.e.  \( \mathcal{S}_k = \{ \xii : x_s^{(i)} \in [z_{k-1}, z_k) \} \)
+and, finally, find the local effect at each bin as the mean value of
+the population \(\hat{\mu}_{s,k}\). The global effect is then
+estimated through the following formula:
 %
@@
   \label{eq:ale-approximation}
-  \hat{f}_{\mathtt{ALE}}^s(x)
+  \hat{f}_{\mathtt{ALE}}(x_s)
   = \Delta x \sum_{k=1}^{k_{x_s}} \hat{\mu}_{s,k}
-  = \Delta x \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in \mathcal{S}_k} [f_s(\xii)]
+  = \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xsi \in \mathcal{S}_k} [f(z_k, \xci) - f(z_{k-1}, \xci)]
 \end{equation}
 %
-where \(\Delta x\) is the bin size and \(k_{x_s}\) the bin-index of the value \(x_s\).
+where \(\Delta x\) is the bin size and \(k_{x_s}\) the index of the
+bin that includes \(x_s\).
 %
-%
-\subsection{ALE plots with uncertainty quantification}
 
-ALE directly defines the local effect as an expectation over the
-unknown features and the global effect as integral over the
-expectations. Therefore, both local and global effects are modeled as
-simple variables. In constrast, for integrating ALE plots in a
-probabilistic framework, we model the local effect as a random
-variable, dependent on the random variable \(\Xc\) representing the
-latent features:
+\subsection{Quantification of the uncertainty of the estimation}
+
+We propose an alternative estimation of the feature effect inside each
+bin. The alternative estimation is better because it (a) separates the
+decision about the bin limits from the local effect of each point, (b)
+secures from out-of-distribution sampling and (c) provides better
+variance estimation inside the bin. The alternative estimation is:
 
 \begin{equation}
-    \Ysloc \sim p(\ysloc ; x_s)
-    = \int_{\xc} \delta(\ysloc-f_s(\x)) p(\xc;x_s) \partial \xc
-    \label{eq:local-effect-rv}
+  \label{eq:1}
+  \tilde{f}_{\texttt{ALE}}(x_s) =
+  \Delta x \sum_{k=1}^{k_{x_s}} \tilde{\mu}_{s,k} =
+  \Delta x \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xsi \in \mathcal{S}_k} f_s(\xii)
 \end{equation}
 %
-Following the core idea of ALE, we define the global feature effect as
-the random variable \(Y_s\) which is the integration of the local
-effects:
-
-\begin{equation}
-  Y_s \sim p(y_s ; x_s) = \int_{x_{s,min}}^{x_s} p(\ysloc;x_s=z) \partial z
-  \label{eq:global-effect-rv}
-\end{equation}
-%
-Through eqs.~(\ref{eq:local-effect-rv}), (\ref{eq:global-effect-rv})
-we formulate the ALE in a fully probabilistic manner. We notice, that
-the definition of ALE as given by~\cite{Apley2020} is simply the
-expected value of \(Y_s\), i.e.
-\(f_{\mathtt{ALE}}(x_s) = \E[Y_s ; x_s]\). For simpler notation we
-denote \(\E[Y_s ; x_s]\) as \(\mu_s (x_s)\).
-
-For measuring the uncertainty, it is also important to capture the
-variance of the global effect. Notating the variance of the feature
-effect \(\var[Y_s ; x_s]\) as \(\sigma_s^2\), we can compute it
-through:
-
-\begin{equation}
-  \sigma_s^2 (x_s)
-  = \int_{x_{s,min}}^{x_s} \int_{\xc} p(\xc|x_s) (f_s(\x) - \mu_s(x_s))^2 \partial \xc (\partial z)^2
-\end{equation}
-
-As before, it is infeasible to compute the the variance
-\(\sigma_s^2(x_s)\) in closed-form. Therefore, we reside on estimating
-them from the available examples of the training set. The estimation
-of the variance can also done through the available points:
+Our goal is to inform the user to what extent they should trust the
+ALE plot. For this purpose, we introduce two metrics; the variance and
+the standard error of the estimation. The variance of the estimation
+is:
 
@@
   \label{eq:effect-var-approx}
-  \hat{\sigma}_s^2(x_s)
-  = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \hat{\sigma}_{s,k}^2
-  = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in \mathcal{S}_k} (f_s(\xii) - \hat{\mu}_{s,k})^2 
+  \tilde{\sigma}_s^2(x_s)
+  = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \tilde{\sigma}_{s,k}^2
+  = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in \mathcal{S}_k} (f_s(\xii) - \tilde{\mu}_{s,k})^2 
 \end{equation}
+%
+The standard error of the estimation is:
+
+\begin{equation}
+  \label{eq:effect-std-error}
+  \text{SE}(x_s) = \Delta x \sum_{k=1}^{k_{x_s}} (\frac{\hat{\sigma}_{s,k}^2}{|\mathcal{S}_k|})^\frac{1}{2}
+\end{equation}
+
+The two metrics, (\ref{eq:effect-var-approx}) and
+(\ref{eq:effect-std-error}), should be trusted only in case they
+respect some contstraints. Under these constraints, the user can be
+confident that the correct feature effect is inside two or three
+standard errors errors around the mean estimation.
+
+Our computations are based on the hypothesis that inside
+all bins \textbf{the gradient wrt. to the feature of interest doesn't
+  depend on the value feature of interest}. Unfortunately, we cannot
+when this is the case. We just know that as the bins grow larger, it
+is more possible to violated this hypothesis. In this case both the
+ALE effect and the standard error are wrong.
+
+The standard error should be trusted when it is estimated by a large
+of data points. For example, in plots (c) and (d), there are bins with
+less than 10 points. Therefore, in these cases, we cannot trust the
+plot or the standard error.
 
@@
 
+Based on the observations of the previous chapter, we want to create a
+single metric for choosing the best feature effect plot. Our goal is
+to minimize the variance, given that we have points inside eac bin
+
+We propose the minimization of the accumulated standard deviation (or
+accumulated variance). For ALE with K bins, let's notate as:
+
+\begin{itemize}
+\item \(dx^K\) the length of each bin
+\item \(p_i^K \) the number of the training points inside the \(i-th\) bin
+\item \(\sigma_i^K \) the std of the local effects of the training points inside the \(i-th\) bin
+\end{itemize}
+%
+We will minimize:
+
+\begin{align} \label{eq:min-criterion}
+  K_{min} &= \text{argmin}_{K} \quad  [dx^K  \sum_i^K \sigma_i^K * (1 - d_i^K)]\\
+  & \text{s.t. } p_i^K \geq \texttt{min\_points\_per\_bin } \forall i
+\end{align}
+%
+where \(d_i^K = 0.2 * \frac{p_i^K}{N} \in [0,0.1] \) works as a `discount', favoring the
+creation of bigger bins in cases of similar standard deviation.
+
 \subsection{ALE plots with variable-size bins}
 
-\subsection{Variable-size bins as an optimization problem}
+\newpage
+~\newpage
 
@@
 \subsection{Synthetic Data sets}
+
+\begin{enumerate}
+\item One example to show that (a) variance and (b) standard error are
+  good estimates to what extend to trust the ALE plot.
+\item One example to show the importance of variable size bins
+\end{enumerate}
+
+\newpage
+~\newpage
+
+\subsection{Real Data sets}
+
+\newpage
+
+\section{Conclusion}
 
@@
 
-\subsection{Real Data sets}
-
-\section{Conclusion}
-
-\subsection{Subsection Title}
+\newpage
 
@@
 \bibliography{acml22}
+\newpage
 
@@
 
-\section{Derivations}\label{apd:first}
+\section{Symbols}\label{apd:symbols}
+
+Spaces and points:
+
+\begin{itemize}
+\item \(\mathcal{X} \subseteq \R^S \): the input space
+\item \(\x = (x_s, \xc) \in \mathcal{X}\)
+\item \(x_s \in \R \)
+\item \(\xc \in \R^{S-1}\)
+\item
+\end{itemize}
+%
+Functions:
+\begin{itemize}
+\item \(f: \R^S \rightarrow \R\): black-box function
+\item \(f_s(\x) = \sfrac{\partial f(\x)}{\partial x_s}: \R^S \rightarrow \R\): effect at point \(\x\)
+\item 
+\end{itemize}
+%
+ALE:
+\begin{itemize}
+\item \(\sfrac{f(z_{k_x}) - f(z_{k_x-1})}{\Delta x}\): effect of \(x_s\) at \(\x\)
+\item \(\E_{p(\xc;x_s)}[f_s(x_s, \xc)]\): local effect of \(x_s\) at \(x_s\)
+\item \(f_{\texttt{ALE}}(x_s): \R \rightarrow \R\): global effect of \(x_s\) at \(x_s\)
+\item \(\hat{f}_{\texttt{ALE}}(x_s): \R \rightarrow \R\): \textit{estimator} of global effect of \(x_s\) at \(x_s\)
+\end{itemize}
+
+\section{Derivations}\label{apd:derivations}
 
@@
 \end{document}
+
+% \subsection{Probabilistic formulation of ALE plots}
+
+% In this section, we will introduce a probabilistic formulation for ALE
+% plots. As discussed before, ALE directly defines the local effect as
+% the expectated value over the unknown features,
+% i.e. \(\E_{p(\xc;x_s)}[f_s(x_s, \xc)]\). Therefore, the local and the
+% global effect, which is the integration of the local effects, are
+% modeled as deterministic variables.
+
+% For adjusting ALE plots into a probabilistic framework, we model the
+% local effect with a random variable conditioned on feature \(x_s\),
+% i.e., \(\Ysloc \sim p(\ysloc ; x_s)\):
+
+% \begin{equation}
+%     p(\ysloc ; x_s)
+%     = \int_{\xc} \delta(\ysloc-f_s(\x)) p(\xc;x_s) \partial \xc
+%     \label{eq:local-effect-rv}
+% \end{equation}
+% %
+% Following the core idea of ALE, we define the global feature effect as
+% the accumulation of the local effects. Therefore, the feature effect
+% of the \(s\)-th feature is \(Y_s \sim p(y_s ; x_s) \):
+% \begin{equation}
+%   p(y_s ; x_s) = \int_{x_{s,min}}^{x_s} p(\ysloc;x_s=z) \partial z
+%   \label{eq:global-effect-rv}
+% \end{equation}
+% %
+% In eqs.~(\ref{eq:local-effect-rv}), (\ref{eq:global-effect-rv}) we
+% reformulate ALE plots in a fully probabilistic manner. We are
+% particularly interested into two summary statistics of \(Y_s\), the
+% expected value \(\mu_s (x_s) = \E[Y_s ; x_s]\) and the variance
+% \(\sigma_s^2(x_s) = \var[Y_s ; x_s]\). In the appendix, we show that
+% the expected value coincides with the definition of ALE plots in
+% eq.~(\ref{eq:ale-definition}), i.e.
+% \(\mu_s (x_s) = f_{\texttt{ALE}(x)s}\). The variance can be computed
+% as:
+
+% \begin{equation}
+%   \sigma_s^2 (x_s)
+%   = \int_{x_{s,min}}^{x_s} \E_{p(\xc|x_s = z)}[(f_s(\x) - \mu_s(z))^2](\partial z)^2
+% \end{equation}
+
+% As before, it is infeasible to compute the the variance
+% \(\sigma_s^2(x_s)\) in analytically. Therefore, we reside on
+% estimating them from the available examples of the training set:
+
+% \begin{equation}
+%   \label{eq:effect-var-approx}
+%   \hat{\sigma}_s^2(x_s)
+%   = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \hat{\sigma}_{s,k}^2
+%   = (\Delta x)^2 \sum_{k=1}^{k_{x_s}} \frac{1}{|\mathcal{S}_k|} \sum_{i:\xii \in \mathcal{S}_k} (f_s(\xii) - \hat{\mu}_{s,k})^2 
+% \end{equation}
+
+% .. add a description for standard error
+
+% \begin{equation}
+%   \label{eq:effect-std-error}
+%   \text{SE}(x_s) = \Delta x \sum_{k=1}^{k_{x_s}} (\frac{\hat{\sigma}_{s,k}^2}{|\mathcal{S}_k|})^\frac{1}{2}
+% \end{equation}

### 2022-06-14 acml/acml22_paper.tex
fixed example 1

@@
 \usepackage{xfrac}
+
+\usepackage{graphicx}
+\graphicspath{{../examples/}}
 
@@
 
-\begin{enumerate}
-\item One example to show that (a) variance and (b) standard error are
-  good estimates to what extend to trust the ALE plot.
-\item One example to show the importance of variable size bins
-\end{enumerate}
+\subsubsection{Case 1}
+
+One example to show that (a) variance and (b) standard error are good
+estimates to what extend to trust the ALE plot. More specifically; (a)
+the variance shows how far away from the ALE plot the global effect
+can be and (b) the standard error shows how wrong the ALE plot can be.
+Test to see how it changes with different sample size.
+
+\begin{figure}[htp]
+\begin{center}
+  \includegraphics[width=0.3\textwidth]{example_1/im_1.png}
+  \includegraphics[width=0.3\textwidth]{example_1/im_3.png}
+  \includegraphics[width=0.3\textwidth]{example_1/im_5.png}\\
+  \includegraphics[width=0.3\textwidth]{example_1/im_2.png}
+  \includegraphics[width=0.3\textwidth]{example_1/im_4.png}
+  \includegraphics[width=0.3\textwidth]{example_1/im_6.png}
+\caption{A spiral.}\label{fig:edfd}
+\end{center}
+\end{figure}
 
 \newpage
-~\newpage
 
-\subsection{Real Data sets}
+\subsubsection{Case 2}
+
+One example to show that (a) variance minimization is a good criterion
+in case of linear feature and fixed-size feature effect and (b)
+variable-size bins estimates as well as the best fixed-size.
+
+\newpage
+
+\subsubsection{Case 3}
+
+One example to show that variance minimization is a good criterion in
+case of linear feature and variable-size feature effect. In this case,
+variable-size bins estimates better than any possible fixed size.
+
+\newpage
+
+\subsubsection{Case 4}
+
+One example to show that (a) variance minimization is a good criterion
+in case of non-linear feature effect. In this case, show that
+variable-size bins estimates better than any possible fixed size.
+
+\newpage
+
+\subsection{Real Dataset}
 
@@
 
-%\acks{Acknowledgements should go at the end, before appendices and references. You can uncomment this for the camera-ready version on paper acceptance.}
+% \acks{Acknowledgements should go at the end, before appendices and
+% references. You can uncomment this for the camera-ready version on
+% paper acceptance.}
 

### 2022-06-14 examples/example_1.py

@@ -1,81 +1,14 @@
-"""Goal: show that if uniformly distributed data, the biggest K wins in fix-length bins.
-
-The set-up
-==========
-* The data points are uniformly distributed in [0,100]
-* The feature-effect is made from 10 piece-wise linear parts of length 10;
-
-For different values of parameters:
-* N: nof points
-* noise_level: how much noise to add in the data_effect estimation
-
-Questions to be addressed
-=========================
-For fixed-size bins:
-* what is the best K in terms of accuracy?
-* Does standard mse_fixed_size or variance estimator chooses the best K?
-For variable-sized bins:
-* what is the best (a) number and (b) size of bins, in terms of mse_fixed_size?
-* does dynamic programming based on mse_fixed_size finds them and for which K?
-* does dynamic programming based on standard mse_fixed_size finds them and for which K?
-
-Final Goals
-===========
-* show that choosing the correct bin-size K is crucial
-* in cases of uniformly distributed data, choosing a big K is always a good option - or even the best option
- (independently of noise level and nof points)
-* show that variable-size bins, has equivalent accuracy with fixed-bin in case of fixed-size piecewise linear effect
- and our algorithm is able to find variable-sizes with equivalent accuracy in this case
-* show that standard mse_fixed_size is a good estimator of the best bin size
-
-Experiments
-===========
-
-(a) For noise_level = 0
-=======================
-For fixed-size bins:
-* The best K are the ones that permit changing bin every 10
-i.e. K = {10, 20, 30, ..., 100} or dx = {10, 10/2=5 ,10/3=3.33, ... , 10/10=1}
-* but in the range of big K, all choices have good accuracy
-* Standard mse_fixed_size perfectly chooses the best candidates
-For Variable-size bins:
-* Finds good bins based on accuracy
-* Finds good bins base on standard mse_fixed_size
-
-
-The conclusions above, hold independently of the number of points. In case of limited points, variable-size bins can be
-"wrong" but this is due to the absence of points, not a problem of the method.
-
-
-(b) For noise_level = 3.0
-=======================
[…]

### 2022-07-08 paper/summary.tex
added paper summary

\documentclass{article}
\usepackage[utf8]{inputenc}

\title{Paper summary}
\author{Vasilis Gkolemis}
\date{July 2021}

\usepackage{graphicx}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[margin=3cm]{geometry}
\usepackage{float}
\graphicspath{ {./../examples/} }

\begin{document}

\maketitle

\section{Introduction}

\paragraph{Description.}

XAI literature distinguishes between local and global interpretation methods~(\cite{Molnar2020}). Global interpretation methods aim at explaining the overall behavior of an ML model. Many of these global interpretation methods are confounded by feature interactions, meaning that they can produce misleading explanations when feature interactions are present. In these cases, they often average individual effects of local interpretation methods and thereby obfuscate heterogeneous effects induced by feature interactions~(\cite{Herbinger2022}). This so-called aggregation bias~(\cite{Mehrabi2019}) is responsible for producing global explanations that often conceal that individual instances may deviate from the global explanation. Therefore, global methods must quantify and inform about \textit{the uncertainty of the global explanation}, i.e., how certain we are that a global explanation is valid if applied to a local individual drawn at random. The uncertainty of the global explanation emerges from the natural characteristics of the experiment, i.e.,~the data generating distribution and the black-box function.

In real ML scenarios, we do not know the data generating distribution for computing the expectations and the uncertainty. The local effects are estimated from the training set's limited instances, and the global effect is computed by aggregating them. Many methods, such as ALE, require an appropriate grouping of samples (partitioning of the feature space) for aggregating local effects that are as homogeneous as possible. If this grouping is not done correctly, it may erroneously mix heterogeneous local effects, not due to the unavoidable heterogeneity of the data generating mechanism but due to improper grouping.

ALE is a SotA method for measuring the global feature effect, but so far, it has two limitations. First, it does not quantify the uncertainty of the global explanation. Second, like most global explanation methods, it requires grouping together samples before computing the global effect. So far, this grouping is done by blindly splitting the feature space in \(K\) fixed-size non-overlapping intervals, where \(K\) is a hyperparameter provided by the user.

In this paper, first, we extend the ALE definition for quantifying the uncertainty of the global explanation. Second, we readjust ALE to work for variable-size bins and formulate the partitioning in non-overlapping intervals as an unsupervised clustering problem. In the unsupervised clustering problem, we minimize an objective which has as lower-bound the (unavoidable) heterogeneity, i.e. the aggregated uncertainty of the global explanation. Therefore, we aim to minimize the added uncertainty induced by the wrong grouping of samples. In other words, we aim to find the optimal grouping that adds no uncertainty over the unavoidable heterogeneity. We finally solve the minimization problem by finding the global optimum using dynamic programming. Our method works out of the box without requiring any input by the user. We provide a theoretical and empirical evaluation of our method.
% In all cases, to estimate
% \textit{the uncertainty of the approximation}~(\cite{Molnar2022}),
% i.e., how certain we are that the limited-samples Monte-Carlo
% approximates well the expectation over the data. These two types of
% uncertainty have not been introduced yet for ALE~(\cite{Apley2020}), a
% SotA method for measuring the global feature effect.

\paragraph{Problem Statement - Contribution.} The contribution of our
paper in bullets:

\begin{itemize}
\item Reformulation of ALE method to quantify the uncertainty of the
  global explanation
\item Formal definition of the variable-size interval splitting as an
  unsupervised clustering problem
\item Method for finding a global optimum in the clustering
  problem
\item Theoretical evaluation of our method (e.g.~show that the
  objective's lower bound is the unavoidable heterogeneity due to the
  characteristics of the problem, highlight specific cases,
  e.g. specific generative distribution and specific black-box function)
\item Empirical evaluation of the method in artificial and real
  datasets
\end{itemize}

\section{Mathematical formulation}

\paragraph{Effect at point \(x_s\).} In the intro, we described the
\textit{uncertainty of the global explanation} as a metric of how
certain we are that the global explanation is valid if applied to a
randomly-drawn individual. ALE plots measure the \(s\)-th feature
effect at point \(x_s\) as the expected change in the output \(y\), if
we slightly change the value of the feature of interest \(x_s\):

\begin{equation}
  \label{eq:ALE_expectation}
  \mu(x_S) = \mathbb{E}_{\mathbf{x_c}|x_s}\left [\frac{\partial
      f}{\partial x_s} \right ]
\end{equation}

\noindent
We argue that it is also crucial to compute the variance of the change
as a measure of the uncertainty of the local change:

\begin{equation}
  \label{eq:2}
  \sigma^2(x_S) = \mathrm{Var}_{\mathbf{x_c}|x_s}\left [\frac{\partial
      f}{\partial x_s} \right ]
\end{equation}

\noindent
The variance in eq.~(\ref{eq:2}) informs us about the heterogeneous
effects hiding behind the explanation. The heterogeneous effects is a
consequence of the uncertainty about the values of \(\mathbf{x_c}\),
i.e.~the uncertainty introduced by the conditional distribution
\(p(\mathbf{x_c}|x_s)\), and the formula of the black-box function
\(f\).

\paragraph{Effect at interval \([z_{k-1}, z_k]\).} In real scenarios,
we approximate \(\mu(x_s)\), \(\sigma^2(x_s)\) using the finite number
of samples the are available in the training set. Since the
possibility to have a training instance in the interval
\([x_s - h, x_s + h]\) is zero in the limit where \(h \to 0\), we are
obliged to estimate the local effect in larger neighboors. The mean
and variance of the interval-effect defined in
\(z_{K-1} \leq x_s < z_k\) are:

\begin{equation}
  \label{eq:ALE_expectation_bin}
  \mu_k(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}}  \int_{z_{k-1}}^{z_k} \mathbb{E}_{\mathbf{x_c}|x_s}\left [\frac{\partial
      f}{\partial x_s} \right ] \partial x_s
\end{equation}

\noindent

\begin{equation}
  \label{eq:ALE_variance_bin}
  \sigma^2(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k} p(\mathbf{x_c}|x_s)\left (\frac{\partial
      f}{\partial x_s}(x_s, \mathbf{x_c}) - \mu_k(z_{k-1}, z_k) \right )^2 \partial x_s
\end{equation}

\noindent
Eq.~(\ref{eq:ALE_variance_bin}) a second term to the uncertainty of
the global explanation defined in Eq.(\ref{eq:2}); the uncertainty
about the exact position of \(x_S\). Since we quantify the local
effect up to the resolution of the interval, we mix together all the
effects in the interval \(z_{K-1} \leq x_s < z_k\).
\noindent

\paragraph{Approximation of effect at interval \([z_{k-1}, z_k]\).}
Eqs.~(\ref{eq:ALE_expectation_bin}),~(\ref{eq:ALE_variance_bin}) are
approximated by the instances of the training set that lie inside the
\(k\)-th interval, i.e.
\( \mathcal{S}_k = \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_{k} \} \):

\begin{equation}
  \label{eq:mean_estimation}
  \hat{\mu}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ \frac{\partial f}{\partial x_s}(\mathbf{x}^i) \right ]
\end{equation}

\begin{equation}
  \label{eq:variance_estimation}
  \hat{\sigma}_k(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
  \mathcal{S}_k} \left [ \frac{\partial f}{\partial x_s}(\mathbf{x}^i) - \hat{\mu}_k(z_{k-1}, z_k) \right ]^2
\end{equation}

\paragraph{Uncertainty of the global effect.}

Eq.~(\ref{eq:variance_estimation}) gives an approximation of the
uncertainty of the bin effect. The uncertainty of the global effect is
simply the sum of the uncertainties in the bin effects. The
approximation is unbiased only if the points are uniformly distributed
in \([z_{k-1}, z_k]\). (TODOs: Check what happens otherwise).

\paragraph{Minimizing the uncertainty}

Solving the problem of finding (a) the optimal number of bins \(K\) and (b) the optimal bin limits for each bin \([z_{k-1}, z_k] \forall k\) to minimize:

\begin{equation}
  \label{eq:1}
  \mathcal{L} = \sum_{k=0}^K \hat{\sigma}_k(z_{k-1}, z_k)
\end{equation}
%
The constraints are that all bins must include more than \(\tau\)
points, i.e., \(|\mathcal{S}_k| \geq \tau\).

\noindent
TODOS. Show theoretically that \(\mathcal{L} \geq \int_{x_{s, min}}^{x_{s, max}}\sigma^2(x_s) \partial x_s\)

\paragraph{Uncertainty of the approximation.}

In all experiments, it also important to measure the uncertainty of
the approximation. The uncertainty of the approximation can be
quantified with two approaches:

\begin{itemize}
\item Splitting the dataset in many folds and (re)estimating
  \(\hat{\mu}(z_{k-1}, z_k), \hat{\sigma}_k(z_{k-1}, z_k)\)
\item Using the central limit theorem, we can (under assumptions) say
  that the standard error of the approximation in
  eq.~(\ref{eq:mean_estimation}) is
  \(\mathrm{std\_error} =
  \frac{\hat{\sigma}_k}{\sqrt{|\mathcal{S}_k|}}\).
\end{itemize}

\section{Evaluation}

\paragraph{Experiments.}

% Design two artificial examples to convince
% that:

% \begin{itemize}
% \item It is important to quantify the uncertainty of the explanation
%   \(\Rightarrow\) Design an example where we can obtain the
%   ground-truth ALE effect and show the change in the uncertainty as we
%   subsample the training set. Show qualitatively that the variance
%   (and the standard error) is growing as we subsample the training
%   set. Show that variance and standard error are valid compared to the
%   ground truth.
% \item It is important to split the feature axis in appropriate
%   non-overlaping bins \(\Rightarrow\) Design an example where by
%   changing the number of bins we observe that: with large bins, we
%   compute a coarse-scale ALE estimation missing fine-grain details
%   and, if ALE estimation instead of DALE, suffers from OOD
%   sampling. With small bins, the estimation of \(\mu\), \(\sigma^2\)
%   (computed only in non-empty bins and interpolated in the others) is
%   noisy. Show that with variable-size bins optimized to minimize the
%   variance we obtain the best estimation.

% \end{itemize}

\paragraph{Metrics.}

% Normalized Mean Squared Error (NMSE) between
% ground truth and approximation.

% \begin{table}[!h]
%   \centering
%   \begin{tabular}{|c|c|c|}
%     - & \texttt{ALE} vs \texttt{GT\_ALE} & ALE \\
%     \hline
%     \multirow{4}{*}{\texttt{Metric}} & \texttt{KL-Divergence} & \\
%     & \texttt{JS-Divergence} & \\
%     & \(r^2\) (\(\mu^2,\hat{\mu}^2\)) & \\
%     & \texttt{NMSE} (\(\mu^2,\hat{\mu}^2\)) & \\
%   \end{tabular}
%   \caption{Metrics that can be used}
%   \label{tab:metrics}
% \end{table}

\bibliographystyle{plain}
\bibliography{summary_bibliography.bib}

\end{document}

### 2022-07-25 paper/summary.tex
fixed some tests

@@
 \begin{equation}
-  \label{eq:ALE_expectation}
+  \label{eq:ALE_mean}
   \mu(x_S) = \mathbb{E}_{\mathbf{x_c}|x_s}\left [\frac{\partial
@@
 \begin{equation}
-  \label{eq:2}
+  \label{eq:ALE_var}
   \sigma^2(x_S) = \mathrm{Var}_{\mathbf{x_c}|x_s}\left [\frac{\partial
@@
 
+\section{Toy Example}
+
+\paragraph{Example 1.} We use the following example:
+
+\begin{equation}
+  \label{eq:model_1}
+  \begin{gathered}
+    f(x_1, x_2) = b_0 + b_1 x_1 + b_2 x_2 + b_3 x_1 x_2\\
+    x_1 \sim \mathcal{U}(0,1) \\
+    x_2 = x_1 + \epsilon, \epsilon \sim \mathcal{N}(\mu=0, \sigma_2^2)
+  \end{gathered}
+\end{equation}
+%
+Therefore, according to Eq.~(\ref{eq:ALE_mean})
+\(\mu_(x_1) = b_1 + b_3 x_1\) and according to Eq.~(\ref{eq:ALE_var})
+\(\sigma^2(x_1) = b_3^2 \sigma_2^2 \Rightarrow \sigma(x_1) = b_3
+\sigma_2 \).
+
+\paragraph{ALE with uncertainty.}
+
+In figure~\ref{fig:bullet-1-im-3} we observe the ALE effect with
+uncertainty for 3 different values of \(\sigma_2 = \{0.01, 0.1,
+1\}\). In all case the mean effect (ALE) is the same, but the
+uncertainty of the global explanation quantifies the impact of the
+heterogeneous effects hiding behing the global explanation.
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=.32\linewidth]{./example_gromping_paper/ALE_gt_s2_0_01.png}
+  \includegraphics[width=.32\linewidth]{./example_gromping_paper/ALE_gt_s2_0_1.png}
+  \includegraphics[width=.32\linewidth]{./example_gromping_paper/ALE_gt_s2_1_0.png}
+  \caption{ALE (a) \(\sigma_2^2 = 0.01\), (b) \(\sigma_2^2 = 0.1\), (c) \(\sigma_2^2 = 1.\)}
+  \label{fig:bullet-1-im-3}
+\end{figure}
+
+\paragraph{Bin Splitting.}
+
+In figure~\ref{fig:im2}, set up for
+\(b_0=0, b_1=1, b_2=1, b_3=2, \sigma_2^2 = 0.1\). Therefore, the
+unavoidable uncertainty is \(\sigma^2(x_1) = b_3^2 \sigma_2^2 =
+0.4\). We observe that as the bins become more dense the added
+uncertainty due to binning becomes less. But as the bins become more
+dense, less points lie inside them and the estimation is poor. The
+vertical lines show the maximum number of bins for dataset sizes, if I
+want at least 25 points per bin.
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=.7\linewidth]{./example_gromping_paper/bin_varinace.png}
+  \caption{}
+  \label{fig:im2}
+\end{figure}
+
 \section{Evaluation}

### 2022-07-25 examples/example_gromping_paper.py

@@ -1,18 +1,10 @@
 import matplotlib.pyplot as plt
 import numpy as np
 import feature_effect as fe
-import scipy.stats as sps
-import scipy.integrate as integrate
-import example_models.models as models
 import example_models.distributions as dist
 
 
-# define model
-b0 = 0
-b1 = 1
-b2 = 1
-b3 = 10
-model = models.LinearWithInteraction(b0, b1, b2, b3)
+path = "/home/givasile/papers/mdale/examples/example_gromping_paper/"
 
 # define distribution
 D = 2
@@ -20,34 +12,81 @@ x1_min = 0
 x1_max = 1
 x2_sigma = 2.
 gen_dist = dist.Correlated1(D, x1_min, x1_max, x2_sigma)
+axis_limits = gen_dist.axis_limits
 
 
-# generate points
-X = gen_dist.generate(N=100000)
+# define model
+def model_dale(b0, b1, b2, b3, x2_sigma):
+
+    # dale
+    dale_mean = lambda x: b1 + b3*x
+    dale_mean_int = lambda x: b1*x + b3/2*x**2
+    dale_var = lambda x: b3**2 * x2_sigma
+    dale_var_int = lambda x: b3**2 * x2_sigma
+
+    dale_gt = fe.DALEGroundTruth(dale_mean, dale_mean_int, dale_var, dale_var_int, axis_limits)
+    dale_gt.fit(features=0)
+    return dale_gt
+
+
+# plot
+def plot1(x2_sigma):
+    plt.figure()
+    plt.title("ALE s2=" + str(x2_sigma))
+    xs = np.linspace(axis_limits[0,0], axis_limits[1,0], 100)
+    dale_gt = model_dale(0, 1, 1, 2, x2_sigma)
+    y, var, _ = dale_gt.eval(xs, s=0, uncertainty=True)
+    plt.plot(xs, y, "b--", label="mean ALE")
+    plt.fill_between(xs, y-np.sqrt(var), y+np.sqrt(var), alpha=0.2, label="std ALE")
+    plt.xlabel("x1")
+    # plt.ylim(-4, 4)
+    plt.ylabel("f_ALE")
+    plt.legend()
+    plt.savefig(path + "ALE_gt_s2_" + str(x2_sigma).replace(".", "_") + ".png")
+    plt.show(block=False)
+
+
+plot1(x2_sigma=0.01)
+plot1(x2_sigma=0.1)
[…]

### 2022-07-25 feature_effect/bin_estimation.py

@@ -90,8 +90,8 @@ class GreedyBase:
                     close_bin = True
                 else:
                     # compare the added loss from closing or keep open in a greedy manner
-                    loss_1 = self.bin_loss(left_lim, limits[i+1])
-                    loss_2 = self.bin_loss(left_lim, limits[i+2])
+                    loss_1 = self.bin_loss(left_lim, limits[i+1])[0]
+                    loss_2 = self.bin_loss(left_lim, limits[i+2])[0]
 
                     if loss_1 == self.big_M or loss_1 == 0:
                         close_bin = False
@@ -124,7 +124,7 @@ class Greedy(GreedyBase):
         dy_dxs = self.data_effect[:, self.feature]
         _, effect_1 = utils.filter_points_belong_to_bin(xs, dy_dxs, np.array([start, stop]))
         loss = np.var(effect_1) if effect_1.size >= self.min_points else self.big_M
-        return loss
+        return loss, loss
 
     def bin_valid(self, start, stop):
         xs = self.data[:, self.feature]
@@ -159,7 +159,7 @@ class GreedyGroundTruth(GreedyBase):
         var1 = var1 / (stop-start)
         var2 = integrate.normalization_constant_1D(mean_var, start, stop)
         var2 = var2 / (stop-start)
-        return var1 + var2
+        return var1 + var2, var1 + var2
 
     def bin_valid(self, start, stop):
         return True
@@ -210,7 +210,7 @@ class DPBase:
             cost = 0
         else:
             start, stop = self._index_to_position(index_before, index_next, K)
-            cost = self._cost_of_bin(start, stop, min_points)
+            cost = self._cost_of_bin(start, stop, min_points)[0]
         return cost
 
     def _argmatrix_to_limits(self, K):
@@ -311,7 +311,7 @@ class DP(DPBase):
             # cost = np.std(data_effect) * (stop-start) / np.sqrt(data_effect.size)
             discount_for_more_points = (1 - .3*(data_effect.size / self.nof_points))
             cost = np.var(data_effect) * (stop-start) * discount_for_more_points
-        return cost
+        return cost, np.var(data_effect)
 
 
 class DPGroundTruth(DPBase):
@@ -339,7 +339,9 @@ class DPGroundTruth(DPBase):
         bin_length_pcg = (stop - start) / (self.xs_max - self.xs_min)
         discount_for_more_points = (1 - .3*bin_length_pcg)
[…]

### 2022-07-25 feature_effect/dale.py

@@ -22,7 +22,7 @@ class DALEGroundTruth(FeatureEffectBase):
         if not uncertainty:
             return self.mean_int(x)
         else:
-            return self.mean_int(x), self.var_int(x)
+            return self.mean_int(x), self.var_int(x), None
 
     def plot(self, s: int, normalized: bool = True, nof_points: int = 30) -> None:
         """Plot the s-th feature
@@ -47,20 +47,17 @@ class DALEBinsGT(FeatureEffectBase):
 
         # bin estimation
         if alg_params["bin_method"] == "fixed":
-            bin_est = be.FixedSizeGT(self.axis_limits, feature=s)
-            bin_est.solve(min_points = alg_params["min_points_per_bin"],
-                          K = alg_params["nof_bins"])
+            bin_est = be.FixedSizeGT(self.mean, self.var, self.axis_limits, feature=s)
+            bin_est.solve(min_points=alg_params["min_points_per_bin"], K=alg_params["nof_bins"])
         elif alg_params["bin_method"] == "greedy":
             bin_est = be.GreedyGroundTruth(self.mean, self.var, self.axis_limits, feature=s)
         elif alg_params["bin_method"] == "dp":
             bin_est = be.DPGroundTruth(self.mean, self.var, self.axis_limits, feature=s)
-        self.bin_est = bin_est
 
         # stats per bin
         dale_params = utils.compute_bin_statistics_gt(self.mean, self.var, bin_est.limits)
         dale_params["limits"] = bin_est.limits
         return dale_params
-
 
     def eval_unnorm(self, x: np.ndarray, s: int, uncertainty: bool = False):
         params = self.feature_effect["feature_" + str(s)]
@@ -87,7 +84,6 @@ class DALEBinsGT(FeatureEffectBase):
         vis.plot_1D(x, y, title="ALE GT Bins for feature %d" % (s+1))
 
 
-
 class DALE(FeatureEffectBase):
     def __init__(self, data, model, model_jac, axis_limits=None):
         # assertions
@@ -102,7 +98,6 @@ class DALE(FeatureEffectBase):
         super(DALE, self).__init__(axis_limits)
 
         self.data_effect = None
-
 
     def compile(self):
         if self.model_jac is not None:
@@ -141,7 +136,6 @@ class DALE(FeatureEffectBase):
         dale_params["limits"] = bin_est.limits
         dale_params["alg_params"] = alg_params
         return dale_params
-
 
[…]

### 2022-07-25 feature_effect/pdp.py

@@ -9,6 +9,7 @@ import feature_effect.utils_integrate as utils_integrate
 
 class FeatureEffectBase:
     empty_symbol = 1e8
+
     def __init__(self, axis_limits: np.ndarray) -> None:
         """
         :param dim: int

### 2022-07-25 examples/example_2.py
fixed examples

@@ -1,156 +1,0 @@
-import matplotlib.pyplot as plt
-import numpy as np
-import feature_effect as fe
-import scipy.stats as sps
-import scipy.integrate as integrate
-import pytest
-from functools import partial
-
-np.random.seed(21)
-
-class OpaqueModel:
-
-    def __init__(self, params):
-        self.params = params
-        self.s2 = 0.1
-
-    def _linear_part(self, x, a, b, x0):
-        return a + b*(x[:, 0]-x0) + x[:, 0]*x[:, 1]
-
-    def _create_cond(self, x, i, s):
-        par = self.params
-        if x.ndim >= 2:
-            return np.logical_and(x[:, s] >= par[i]["from"], x[:, s] <= par[i]["to"])
-        elif x.ndim == 1:
-            return np.logical_and(x >= par[i]["from"], x <= par[i]["to"])
-
-    def _create_func(self, i, func):
-        par = self.params
-        return partial(func, a=par[i]["a"], b=par[i]["b"], x0=par[i]["from"])
-
-    def predict(self, x):
-        """f(x1, x2) = a + b*x1 + x1x2
-        """
-        par = self.params
-        condlist = [self._create_cond(x, i, s=0) for i in range(4)]
-        funclist = [self._create_func(i, self._linear_part) for i in range(4)]
-
-        y = np.zeros(x.shape[0])
-        for i, cond in enumerate(condlist):
-            y[cond] = funclist[i](x[cond, :])
-        return y
-
-    def jacobian(self, x):
-        par = self.params
-        condlist = [self._create_cond(x, i, s=0) for i in range(4)]
-
-        def df_dx1(x, a, b, x0):
-            return b + x[:, 1]
-
-        def df_dx2(x, a, b, x0):
-            return x[:, 0]
-
-        funclist1 = [self._create_func(i, df_dx1) for i in range(4)]
-        funclist2 = [self._create_func(i, df_dx2) for i in range(4)]
-        y1 = np.zeros(x.shape[0])
-        y2 = np.zeros(x.shape[0])
-        for i, cond in enumerate(condlist):
-            y1[cond] = funclist1[i](x[cond, :])
-            y2[cond] = funclist2[i](x[cond, :])
-
-        return np.stack([y1,y2], axis=-1)
-
-
[…]

### 2022-07-25 examples/example_gromping_paper.py

@@ -3,7 +3,7 @@ import numpy as np
 import feature_effect as fe
 import example_models.distributions as dist
 
-
+savefig = False
 path = "/home/givasile/papers/mdale/examples/example_gromping_paper/"
 
 # define distribution
@@ -42,7 +42,8 @@ def plot1(x2_sigma):
     # plt.ylim(-4, 4)
     plt.ylabel("f_ALE")
     plt.legend()
-    plt.savefig(path + "ALE_gt_s2_" + str(x2_sigma).replace(".", "_") + ".png")
+    if savefig:
+        plt.savefig(path + "ALE_gt_s2_" + str(x2_sigma).replace(".", "_") + ".png")
     plt.show(block=False)
 
 
@@ -85,7 +86,8 @@ plt.hlines(b3**2*x2_sigma, 0, 40, label="unavoidable variance")
 plt.xlabel("K")
 plt.ylabel("mean bin variance")
 plt.legend()
-plt.savefig(path + "bin_varinace.png")
+if savefig:
+    plt.savefig(path + "bin_varinace.png")
 plt.show(block=False)
 
 # dale = fe.DALE(X, model.predict, model.jacobian)

### 2022-07-27 paper/summary.tex
added proof in paper summary

@@
 
+\newcommand{\dfdx}{\frac{\partial f}{\partial x_s}}
+\newcommand{\xc}{\mathbf{x_c}}
+\newcommand{\DY}{\mathbf{\Delta Y}}
 \begin{document}
@@
 \section{Introduction}
+\label{sec:introduction}
 
@@
 
-XAI literature distinguishes between local and global interpretation methods~(\cite{Molnar2020}).
+XAI literature distinguishes between local and global interpretation methods~\cite{Molnar2020}.
 Global interpretation methods aim at explaining the overall behavior of an ML model.
-Many of these global interpretation methods are confounded by feature interactions, meaning that they can produce misleading explanations when feature interactions are present.
-In these cases, they often average individual effects of local interpretation methods and thereby obfuscate heterogeneous effects induced by feature interactions~(\cite{Herbinger2022}).
-This so-called aggregation bias~(\cite{Mehrabi2019}) is responsible for producing global explanations that often conceal that individual instances may deviate from the global explanation.
+Many of these global interpretation methods are confounded by feature interactions,
+meaning that they can produce misleading explanations when feature interactions are present.
+In these cases, they often average individual effects of local interpretation methods and thereby obfuscate heterogeneous effects induced by feature interactions~\cite{Herbinger2022}.
+
+This so-called aggregation bias~\cite{Mehrabi2019} is responsible for producing global explanations that often conceal that individual instances may deviate from the global explanation.
 Therefore, global methods must quantify and inform about \textit{the uncertainty of the global explanation}, i.e., how certain we are that a global explanation is valid if applied to a local individual drawn at random.
@@
 Second, we readjust ALE to work for variable-size bins and formulate the partitioning in non-overlapping intervals as an unsupervised clustering problem.
-In the unsupervised clustering problem, we minimize an objective which has as lower-bound the (unavoidable) heterogeneity, i.e.
-the aggregated uncertainty of the global explanation.
+In the unsupervised clustering case, we minimize an objective which has as lower-bound the (unavoidable) heterogeneity, i.e.\ the aggregated uncertainty of the global explanation.
 Therefore, we aim to minimize the added uncertainty induced by the wrong grouping of samples.
@@
   characteristics of the problem, highlight specific cases,
-  e.g. specific generative distribution and specific black-box function)
+  e.g.\ specific generative distribution and specific black-box function)
 \item Empirical evaluation of the method in artificial and real
@@
 \section{Mathematical formulation}
+\label{sec:mathematical-formulation}
 
@@
 certain we are that the global explanation is valid if applied to a
-randomly-drawn individual. ALE plots measure the \(s\)-th feature
-effect at point \(x_s\) as the expected change in the output \(y\), if
+randomly-drawn individual.
+ALE plots measure the \(s\)-th feature
+effect at point \(x_s\) as the \textbf{expected} change in the output \(y\), if
 we slightly change the value of the feature of interest \(x_s\):
@@
   \label{eq:ALE_mean}
-  \mu(x_S) = \mathbb{E}_{\mathbf{x_c}|x_s}\left [\frac{\partial
-      f}{\partial x_s} \right ]
+  \mu(x_s) = \mathbb{E}_{\xc|x_s}\left [\dfdx (x_s, \xc)\right ]
 \end{equation}
@@
 \noindent
-We argue that it is also crucial to compute the variance of the change
-as a measure of the uncertainty of the local change:
+We model the s-th feature effect at point \(x_s\) (change in the output \(y\) wrt a slight change in the feature of interest \(x_s\)) as the random variable $\DY ; x_s $.
+For notation convenience, we will refer to the s-th feature effect as \(\DY\), ommiting the \(;xs\) part.
+The randomness has its origins in the ignorance of the values of the rest of the features, denoted with \(\mathbf{X}_c\).
+Therefore, the s-th feature effect is defined as:
+
+\begin{equation}
+  \label{eq:fe_rv}
+  \DY = g(x_s) = \dfdx (x_s, \mathbf{X}_c)
+\end{equation}
+%
+As shown in Eq.~\eqref{eq:ALE_mean}, ALE is only interested in the expected value of \(\DY\).
+Instead, we are also interested in the variance of \(\DY\) for measuring the uncertainty of the local change:
 
@@
   \label{eq:ALE_var}
-  \sigma^2(x_S) = \mathrm{Var}_{\mathbf{x_c}|x_s}\left [\frac{\partial
-      f}{\partial x_s} \right ]
+  \sigma^2(x_s) = \mathrm{Var}_{\xc|x_s}\left [\dfdx (x_s, \xc) \right ]
 \end{equation}
@@
 \noindent
-The variance in eq.~(\ref{eq:2}) informs us about the heterogeneous
-effects hiding behind the explanation. The heterogeneous effects is a
-consequence of the uncertainty about the values of \(\mathbf{x_c}\),
-i.e.~the uncertainty introduced by the conditional distribution
-\(p(\mathbf{x_c}|x_s)\), and the formula of the black-box function
-\(f\).
+The variance in Eq.~\eqref{eq:ALE_var} informs us about the heterogeneous
+effects hiding behind the explanation.
 
-\paragraph{Effect at interval \([z_{k-1}, z_k]\).} In real scenarios,
-we approximate \(\mu(x_s)\), \(\sigma^2(x_s)\) using the finite number
-of samples the are available in the training set. Since the
-possibility to have a training instance in the interval
-\([x_s - h, x_s + h]\) is zero in the limit where \(h \to 0\), we are
-obliged to estimate the local effect in larger neighboors. The mean
-and variance of the interval-effect defined in
-\(z_{K-1} \leq x_s < z_k\) are:
+\paragraph{Effect at interval \([z_{k-1}, z_k]\).}
+
+In real scenarios, we have ignorance about the generative distribution \(\p(x_s, \mathbf{x}_c)\),
+residing to Monte-Carlo approximations of \(\mu(x_s)\), \(\sigma^2(x_s)\) using the samples of the training set.
+Therefore, it is impossible to estimate Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
+point \(x_s\)\, since the  possibility to observe a sample in the interval
+\([x_s - h, x_s + h]\) is zero, in the limit where \(h \to 0\).
+Therefore, we are obliged to estimate the local effect in larger intervals (\(h>0\)).
+We refer to the expected value and the variance of the feature effect at the interval \([z_{k-1}, z_k)\), as:
 
 \begin{equation}
-  \label{eq:ALE_expectation_bin}
-  \mu_k(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}}  \int_{z_{k-1}}^{z_k} \mathbb{E}_{\mathbf{x_c}|x_s}\left [\frac{\partial
-      f}{\partial x_s} \right ] \partial x_s
+  \label{eq:mu_bin}
+  \mu_k = \mu(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k}
+  \mathbb{E}_{\xc|x_s=z}\left [\frac{\partial f}{\partial x_s} \right ] \partial z
 \end{equation}
@@
 \begin{equation}
-  \label{eq:ALE_variance_bin}
-  \sigma^2(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k} p(\mathbf{x_c}|x_s)\left (\frac{\partial
-      f}{\partial x_s}(x_s, \mathbf{x_c}) - \mu_k(z_{k-1}, z_k) \right )^2 \partial x_s
+  \label{eq:var_bin}
+  \sigma^2_k = \sigma^2(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k}
+  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
+      f}{\partial x_s}(x_s, \xc) - \mu_k )^2 \right] \partial z
 \end{equation}
@@
 \noindent
-Eq.~(\ref{eq:ALE_variance_bin}) a second term to the uncertainty of
-the global explanation defined in Eq.(\ref{eq:2}); the uncertainty
-about the exact position of \(x_S\). Since we quantify the local
-effect up to the resolution of the interval, we mix together all the
-effects in the interval \(z_{K-1} \leq x_s < z_k\).
+We prove that the interval-variance \(\sigma^2_k \) is:
+
+\begin{equation}
+ \sigma^2_k = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k} \sigma^2(z) + \rho^2(z) \partial z
+\end{equation}
+%
+where \(\rho(x_s) = \mu(x_s) - \mu_k\).
+The proof is at Section \ref{sec:proofs}.
+We observe that the interval-variance is the mean point-variance of the points
+inside the interval \([z_{k-1}, z_k]\), plus the mean of the residual term \(\rho\).
+The mean point-variance is the unavoidable variance, due to the uncertainty of the global explanation.
+The mean of the residual term is an extra variance (uncertainty) due to limiting the granularity of the effect at
+the bin level.
+
 \noindent
@@
 \paragraph{Approximation of effect at interval \([z_{k-1}, z_k]\).}
-Eqs.~(\ref{eq:ALE_expectation_bin}),~(\ref{eq:ALE_variance_bin}) are
+Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} are
 approximated by the instances of the training set that lie inside the
@@
   \hat{\mu}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
-    \mathcal{S}_k} \left [ \frac{\partial f}{\partial x_s}(\mathbf{x}^i) \right ]
+    \mathcal{S}_k} \left [ \dfdx(\mathbf{x}^i) \right ]
 \end{equation}
@@
   \hat{\sigma}_k(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
-  \mathcal{S}_k} \left [ \frac{\partial f}{\partial x_s}(\mathbf{x}^i) - \hat{\mu}_k(z_{k-1}, z_k) \right ]^2
+  \mathcal{S}_k} \left [ \dfdx(\mathbf{x}^i) - \hat{\mu}_k(z_{k-1}, z_k) \right ]^2
 \end{equation}
@@
 
-Eq.~(\ref{eq:variance_estimation}) gives an approximation of the
-uncertainty of the bin effect. The uncertainty of the global effect is
-simply the sum of the uncertainties in the bin effects. The
+Eq.~\eqref{eq:variance_estimation} gives an approximation of the
+uncertainty of the bin effect.
+The uncertainty of the global effect is
+simply the sum of the uncertainties in the bin effects.
+The
 approximation is unbiased only if the points are uniformly distributed
@@
 \noindent
-TODOS. Show theoretically that \(\mathcal{L} \geq \int_{x_{s, min}}^{x_{s, max}}\sigma^2(x_s) \partial x_s\)
+TODOS. Show theoretically that \(\mathcal{L} \geq \int_{x_{s, \min}}^{x_{s, \max}}\sigma^2(x_s) \partial x_s\)
 
@@
 In all experiments, it also important to measure the uncertainty of
-the approximation. The uncertainty of the approximation can be
+the approximation.
+The uncertainty of the approximation can be
 quantified with two approaches:
@@
   that the standard error of the approximation in
-  eq.~(\ref{eq:mean_estimation}) is
+  eq.~\eqref{eq:mean_estimation} is
   \(\mathrm{std\_error} =
@@
 \section{Toy Example}
+\label{sec:toy-example}
 
@@
 %
-Therefore, according to Eq.~(\ref{eq:ALE_mean})
-\(\mu_(x_1) = b_1 + b_3 x_1\) and according to Eq.~(\ref{eq:ALE_var})
+Therefore, according to Eq.~\eqref{eq:ALE_mean}
+\(\mu_(x_1) = b_1 + b_3 x_1\) and according to Eq.~\eqref{eq:ALE_var}
 \(\sigma^2(x_1) = b_3^2 \sigma_2^2 \Rightarrow \sigma(x_1) = b_3
@@
 uncertainty for 3 different values of \(\sigma_2 = \{0.01, 0.1,
-1\}\). In all case the mean effect (ALE) is the same, but the
+1\}\).
+In all case the mean effect (ALE) is the same, but the
 uncertainty of the global explanation quantifies the impact of the
-heterogeneous effects hiding behing the global explanation.
+heterogeneous effects hiding behind the global explanation.
 
@@
   \centering
-  \includegraphics[width=.32\linewidth]{./example_gromping_paper/ALE_gt_s2_0_01.png}
-  \includegraphics[width=.32\linewidth]{./example_gromping_paper/ALE_gt_s2_0_1.png}
-  \includegraphics[width=.32\linewidth]{./example_gromping_paper/ALE_gt_s2_1_0.png}
+  \includegraphics[width=.32\linewidth]{./example_gromping_paper/ALE_gt_s2_0_01}
+  \includegraphics[width=.32\linewidth]{./example_gromping_paper/ALE_gt_s2_0_1}
+  \includegraphics[width=.32\linewidth]{./example_gromping_paper/ALE_gt_s2_1_0}
   \caption{ALE (a) \(\sigma_2^2 = 0.01\), (b) \(\sigma_2^2 = 0.1\), (c) \(\sigma_2^2 = 1.\)}
@@
 In figure~\ref{fig:im2}, set up for
-\(b_0=0, b_1=1, b_2=1, b_3=2, \sigma_2^2 = 0.1\). Therefore, the
+\(b_0=0, b_1=1, b_2=1, b_3=2, \sigma_2^2 = 0.1\).
+Therefore, the
 unavoidable uncertainty is \(\sigma^2(x_1) = b_3^2 \sigma_2^2 =
-0.4\). We observe that as the bins become more dense the added
-uncertainty due to binning becomes less. But as the bins become more
-dense, less points lie inside them and the estimation is poor. The
+0.4\).
+We observe that as the bins become denser the added
+uncertainty due to binning becomes less.
+But as the bins become denser, fewer points lie inside them and the estimation is poor.
+The
 vertical lines show the maximum number of bins for dataset sizes, if I
@@
   \centering
-  \includegraphics[width=.7\linewidth]{./example_gromping_paper/bin_varinace.png}
+  \includegraphics[width=.7\linewidth]{./example_gromping_paper/bin_varinace}
   \caption{}
@@
 \section{Evaluation}
+\label{sec:evaluation}
 
@@
 
+\section{Proofs}
+\label{sec:proofs}
+
+Given that:
+
+\[
+  \sigma^2_k = \sigma^2(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k}
+  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
+      f}{\partial x_s}(x_s, \xc) - \mu_k )^2 \right] \partial z
+\]
+
+\[
+\sigma^2(x_s) = \mathrm{Var}_{\xc|x_s}\left [\dfdx (x_s, \xc) \right ]
+\]
+
+\[
+\(\rho(x_s) = \mu(x_s) - \mu_k\)
+\]
+
+We want to prove that:
+
+\[
+ \sigma^2_k = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k} \sigma^2(z) + \rho^2(z) \partial z
+\]
+
+Proof:
+
+\begin{gather}
+  \sigma^2_k = \sigma^2(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k}
+  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial f}{\partial x_s}(z, \xc) - \mu_k )^2 \right] \partial z \\
+  = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k}
+  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial f}{\partial x_s} - \mu(z) + \rho(z) )^2 \right] \partial z \\
+  = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k} \left(
+  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial f}{\partial x_s} - \mu(z) )^2 \right ]  +
+  \mathbb{E}_{\xc|x_s=z} \left [ \rho(z)^2 \right] +
+  \mathbb{E}_{\xc|x_s=z} \left [ 2(\frac{\partial f}{\partial x_s} - \mu(z) )\rho(z) \right ] \right )  \partial z \\
+  = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k} \left (
+  \sigma^2(x_s)  + \rho^2(z) + 2 ( \mu(z) -  \mu(z) )\rho(z) \right )  \partial z \\
+  = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k} \sigma^2(x_s)  + \rho^2(z) \partial z \\
+\end{gather}
+
 \bibliographystyle{plain}
-\bibliography{summary_bibliography.bib}
+\bibliography{summary_bibliography}
 

### 2022-07-27 example_models/distributions.py

@@ -1,9 +1,6 @@
-import matplotlib.pyplot as plt
 import numpy as np
-import feature_effect as fe
 import scipy.stats as sps
-import scipy.integrate as integrate
-import pytest
+
 
 class GenerativeBase:
     def __init__(self):

### 2022-07-27 example_models/models.py

@@ -1,12 +1,7 @@
 import matplotlib.pyplot as plt
 import typing
 import numpy as np
-import feature_effect as fe
-import scipy.stats as sps
-import scipy.integrate as integrate
-import pytest
 from functools import partial
-
 
 
 class ModelBase:
@@ -22,7 +17,6 @@ class ModelBase:
         """Callable (x: np.ndarray (N,D)) -> np.ndarray (N, D)
         """
         raise NotImplementedError
-
 
     def plot(self, axis_limits: np.ndarray, nof_points: int,
              X: typing.Union[None, np.ndarray] = None) -> None:
@@ -80,11 +74,11 @@ class SquareWithInteraction(ModelBase):
 
 
 class PiecewiseLinear(ModelBase):
-
     def __init__(self, params):
         self.params = params
 
-    def _linear_part(self, x, a, b, x0):
+    @staticmethod
+    def _linear_part(x, a, b, x0):
         return a + b*(x[:, 0]-x0) + x[:, 0]*x[:, 1]
 
     def _create_cond(self, x, i, s):
@@ -101,7 +95,6 @@ class PiecewiseLinear(ModelBase):
     def predict(self, x):
         """f(x1, x2) = a + b*x1 + x1x2
         """
-        par = self.params
         condlist = [self._create_cond(x, i, s=0) for i in range(4)]
         funclist = [self._create_func(i, self._linear_part) for i in range(4)]
 
@@ -111,7 +104,6 @@ class PiecewiseLinear(ModelBase):
         return y
 
     def jacobian(self, x):
-        par = self.params
         condlist = [self._create_cond(x, i, s=0) for i in range(4)]
 
         def df_dx1(x, a, b, x0):
@@ -129,7 +121,6 @@ class PiecewiseLinear(ModelBase):
             y2[cond] = funclist2[i](x[cond, :])
 
         return np.stack([y1,y2], axis=-1)
-
 
 
 class LinearWithInteraction3D(ModelBase):

### 2022-07-27 feature_effect/pdp.py

@@ -31,14 +31,11 @@ class FeatureEffectBase:
         """
         raise NotImplementedError
 
-
     def fit_feature(self, s: int, alg_params: typing.Dict = None) -> typing.Dict:
         raise NotImplementedError
 
-
     def plot(self, s: int, normalized: bool = True, nof_points: int = 30) -> None:
         raise NotImplementedError
-
 
     def compute_z(self, s: int) -> float:
         func = partial(self.eval_unnorm, s=s, uncertainty=False)
@@ -47,16 +44,20 @@ class FeatureEffectBase:
         z = utils_integrate.normalization_constant_1D(func, start, stop)
         return z
 
-
     def fit(self,
             features: typing.Union[str, list] = "all",
             alg_params: typing.Union[None, dict] = {},
             compute_z: bool = True) -> None:
         """Compute normalization constants for asked features
 
-        :param features: list of features to compute the normalization constant
+        :param
         :returns: None
 
+        Parameters
+        ----------
+        features: list of features to compute the normalization constant
+        alg_params
+        compute_z
         """
         features = helpers.prep_features(features, self.dim)
         for s in features:

### 2022-09-05 AISTATS2023/sample_paper.tex
initialized AISTATS2023 submission

\documentclass[twoside]{article}

\usepackage{aistats2023}
% If your paper is accepted, change the options for the package
% aistats2023 as follows:
%
%\usepackage[accepted]{aistats2023}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

% If you set papersize explicitly, activate the following three lines:
%\special{papersize = 8.5in, 11in}
%\setlength{\pdfpageheight}{11in}
%\setlength{\pdfpagewidth}{8.5in}

% If you use natbib package, activate the following three lines:
\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}
\bibliographystyle{plainnat}

% If you use BibTeX in apalike style, activate the following line:
%\bibliographystyle{apalike}

\begin{document}

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Instructions for Paper Submissions to AISTATS 2023}

\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }

\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]

\begin{abstract}
  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
  both left and right-hand margins. Use 10~point type, with a vertical
  spacing of 11~points. The \textbf{Abstract} heading must be centered,
  bold, and in point size 12. Two line spaces precede the
  Abstract. The Abstract must be limited to one paragraph.
\end{abstract}

\section{INTRODUCTION}

XAI literature distinguishes between local and global interpretation methods~\citep{Molnar2020}. Local methods explain a specific prediction, whereas global methods explain the entire model behavior. Global methods provide a universal explanation, summarizing the numerous local explanations into a single interpretable outcome (number or plot). For example, if a user wants to know which features are significant (feature importance) or whether a particular feature has a positive or negative effect on the output (feature effect), they should opt for a global explainability technique. Aggregating the individual explanations for producing a global one comes at a cost. In cases where feature interactions are strong, the global explanation may obfuscate heterogeneous effects~\citep{Herbinger2022} that exist under the hood, a phenomenon called aggregation bias~\citep{Mehrabi2019}.

Feature effect forms a fundamental category of global explainability methods. The goal of the feature effect is to isolate the average impact of a single feature on the output. Feature effect methods suffer from aggregation bias because the rationale behind the average effect might be unclear. For example, a feature with zero average effect may indicate that the feature has no effect on the output or has a highly positive effect in some cases and a highly negative one in others.

There are two widely-used feature effect methods; Partial Dependence Plots (PDPlots) (Friedman, 2001) and Aggregated Local Effects (ALE)\citep{Apley2020}. PDPlots have been criticized for producing erroneous feature effect plots when the input features are correlated due to marginalizing over out-of-distribution synthetic instances. Therefore, ALE has been established as the state-of-the-art feature effect method since it can isolate feature effects in situations where input features are highly correlated.

However, ALE faces two crucial drawbacks. First, it lacks a way to inform the user about potential heterogeneous effects that hide behind the average effect. In contrast, in Partial Dependence Plots (PDP), the heterogeneous effects can be spotted by exploring the Individual Conditional Expectations (ICE). Second, ALE requires an additional step, where the axis of the feature of interest is split in \(K\) fixed-size non-overlapping intervals, where \(K\) is a hyperparameter provided by the user. This splitting is done blindly, which can lead to inconsistent explanations.

In this paper, we extend ALE with a probabilistic component for measuring the uncertainty of the global explanation. The uncertainty of the global explanation expresses how certain we are that the global (expected) explanation is valid if applied to an instance drawn at random, or, in other words, the level of heterogeneous effects hidden behind the expected explanation. The uncertainty of the global explanation emerges from the natural characteristics of the experiment, i.e.,~the data generating distribution and the black-box function.

Furthermore, we automate the step of splitting the axis into non-overlapping intervals. The need for non-overlapping bins emerges from the ignorance of the data-generating distribution, enforcing all estimations to be based on the limited instances of the training set. Therefore, there is an implicit trade-off behind the formation of bins. Each bin must include enough instances for a robust estimation of the bin feature effect (expected value), and the uncertainty of the explanation (variance). On the other hand, each bin should include points with similar local effects. Therefore, we transform the bin splitting step into an unsupervised clustering problem, encoding the trade-off mentioned above in the objective function. We formally show that the objective of the clustering problem has lower-bound the (unavoidable) heterogeneity, i.e., the aggregated uncertainty of the global explanation. Therefore, we aim to find the optimal grouping of samples that adds the slightest uncertainty over the unavoidable heterogeneity. We finally solve the minimization problem by finding the global optimum using dynamic programming. Our method works out of the box without requiring any input by the user. We provide a theoretical and empirical evaluation of our method.

\section{BACKGROUND AND RELATED WORK}

\section{THE U-DALE METHOD}

\subsection{Uncertainty Quantification}

\subsubsection{Methodology}

\subsection{Bin Spliting as a Clustering Problem}

\subsubsection{Methodology}

\subsubsection{Algorithms}

\section{SYNTHETIC EXAMPLES}

\section{REAL-WORLD EXAMPLES}

\section{GENERAL FORMATTING INSTRUCTIONS}

The camera-ready versions of the accepted papers are 8 pages,
plus any additional pages needed for references.

Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
Each column is 3.25~inches wide (19.5~picas).  The space
between the columns is .25~inches wide (1.5~picas).  The left margin is 0.88~inches (5.28~picas).
Use 10~point type with a vertical spacing of
11~points. Please use US Letter size paper instead of A4.

Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
Top rule is 4~points thick and bottom rule is 1~point thick.
Allow 1/4~inch space above and below title to rules.

Author descriptions are center-justified, initial caps.  The lead
author is to be listed first (left-most), and the Co-authors are set
to follow.  If up to three authors, use a single row of author
descriptions, each one center-justified, and all set side by side;
with more authors or unusually long names or institutions, use more
rows.

Use one-half line space between paragraphs, with no indent.

\section{FIRST LEVEL HEADINGS}

First level headings are all caps, flush left, bold, and in point size
12. Use one line space before the first level heading and one-half line space
after the first level heading.

\subsection{Second Level Heading}

Second level headings are initial caps, flush left, bold, and in point
size 10. Use one line space before the second level heading and one-half line
space after the second level heading.

\subsubsection{Third Level Heading}

Third level headings are flush left, initial caps, bold, and in point
size 10. Use one line space before the third level heading and one-half line
space after the third level heading.

\paragraph{Fourth Level Heading}

Fourth level headings must be flush left, initial caps, bold, and
Roman type.  Use one line space before the fourth level heading, and
place the section text immediately after the heading with no line
break, but an 11 point horizontal space.

%%%
\subsection{Citations, Figure, References}

\subsubsection{Citations in Text}

Citations within the text should include the author's last name and
year, e.g., (Cheesman, 1985). 
%Apart from including the author's last name and year, citations can follow any style, as long as the style is consistent throughout the paper.  
Be sure that the sentence reads
correctly if the citation is deleted: e.g., instead of ``As described
by (Cheesman, 1985), we first frobulate the widgets,'' write ``As
described by Cheesman (1985), we first frobulate the widgets.''

The references listed at the end of the paper can follow any style as long as it is used consistently.

%Be sure to avoid
%accidentally disclosing author identities through citations.

\subsubsection{Footnotes}

Indicate footnotes with a number\footnote{Sample of the first
  footnote.} in the text. Use 8 point type for footnotes. Place the
footnotes at the bottom of the column in which their markers appear,
continuing to the next column if required. Precede the footnote
section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
long.\footnote{Sample of the second footnote.}

\subsubsection{Figures}

All artwork must be centered, neat, clean, and legible.  All lines
should be very dark for purposes of reproduction, and art work should
not be hand-drawn.  Figures may appear at the top of a column, at the
top of a page spanning multiple columns, inline within a column, or
with text wrapped around them, but the figure number and caption
always appear immediately below the figure.  Leave 2 line spaces
between the figure and the caption. The figure caption is initial caps
and each figure should be numbered consecutively.

Make sure that the figure caption does not get separated from the
figure. Leave extra white space at the bottom of the page rather than
splitting the figure and figure caption.
\begin{figure}[h]
\vspace{.3in}
\centerline{\fbox{This figure intentionally left non-blank}}
\vspace{.3in}
\caption{Sample Figure Caption}
\end{figure}

\subsubsection{Tables}

All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
Table number and title always appear above the table.
See Table~\ref{sample-table}.

Use one line space before the table title, one line space after the table title,
and one line space after the table. The table title must be
initial caps and each table numbered consecutively.

\begin{table}[h]
\caption{Sample Table Title} \label{sample-table}
\begin{center}
\begin{tabular}{ll}
\textbf{PART}  &\textbf{DESCRIPTION} \\
\hline \\
Dendrite         &Input terminal \\
Axon             &Output terminal \\
Soma             &Cell body (contains cell nucleus) \\
\end{tabular}
\end{center}
\end{table}

\section{SUPPLEMENTARY MATERIAL}

If you need to include additional appendices during submission, you can include them in the supplementary material file.
You can submit a single file of additional supplementary material which may be either a pdf file (such as proof details) or a zip file for other formats/more files (such as code or videos). 
Note that reviewers are under no obligation to examine your supplementary material. 
If you have only one supplementary pdf file, please upload it as is; otherwise gather everything to the single zip file.

You must use \texttt{aistats2023.sty} as a style file for your supplementary pdf file and follow the same formatting instructions as in the main paper. 
The only difference is that it must be in a \emph{single-column} format.
You can use \texttt{supplement.tex} in our starter pack as a starting point.
Alternatively, you may append the supplementary content to the main paper and split the final PDF into two separate files.

\section{SUBMISSION INSTRUCTIONS}

To submit your paper to AISTATS 2023, please follow these instructions.

\begin{enumerate}
    \item Download \texttt{aistats2023.sty}, \texttt{fancyhdr.sty}, and \texttt{sample\_paper.tex} provided in our starter pack. 
    Please, do not modify the style files as this might result in a formatting violation.

    \item Use \texttt{sample\_paper.tex} as a starting point.
    \item Begin your document with
    \begin{flushleft}
    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
    \texttt{\textbackslash usepackage\{aistats2023\}}
    \end{flushleft}
    The \texttt{twoside} option for the class article allows the
    package \texttt{fancyhdr.sty} to include headings for even and odd
    numbered pages.
    \item When you are ready to submit the manuscript, compile the latex file to obtain the pdf file.
    \item Check that the content of your submission, \emph{excluding} references, is limited to \textbf{8 pages}. The number of pages containing references alone is not limited.
    \item Upload the PDF file along with other supplementary material files to the CMT website.
\end{enumerate}

\subsection{Camera-ready Papers}

%For the camera-ready paper, if you are using \LaTeX, please make sure
%that you follow these instructions.  
% (If you are not using \LaTeX,
%please make sure to achieve the same effect using your chosen
%typesetting package.)

If your papers are accepted, you will need to submit the camera-ready version. Please make sure that you follow these instructions:
\begin{enumerate}
    %\item Download \texttt{fancyhdr.sty} -- the
    %\texttt{aistats2023.sty} file will make use of it.
    \item Change the beginning of your document to
    \begin{flushleft}
    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
    \texttt{\textbackslash usepackage[accepted]\{aistats2023\}}
    \end{flushleft}
    The option \texttt{accepted} for the package
    \texttt{aistats2023.sty} will write a copyright notice at the end of
    the first column of the first page. This option will also print
    headings for the paper.  For the \emph{even} pages, the title of
    the paper will be used as heading and for \emph{odd} pages the
    author names will be used as heading.  If the title of the paper
    is too long or the number of authors is too large, the style will
    print a warning message as heading. If this happens additional
    commands can be used to place as headings shorter versions of the
    title and the author names. This is explained in the next point.
    \item  If you get warning messages as described above, then
    immediately after $\texttt{\textbackslash
    begin\{document\}}$, write
    \begin{flushleft}
    \texttt{\textbackslash runningtitle\{Provide here an alternative
    shorter version of the title of your paper\}}\\
    \texttt{\textbackslash runningauthor\{Provide here the surnames of
    the authors of your paper, all separated by commas\}}
    \end{flushleft}
    Note that the text that appears as argument in \texttt{\textbackslash
      runningtitle} will be printed as a heading in the \emph{even}
    pages. The text that appears as argument in \texttt{\textbackslash
      runningauthor} will be printed as a heading in the \emph{odd}
    pages.  If even the author surnames do not fit, it is acceptable
    to give a subset of author names followed by ``et al.''

    %\item Use the file sample\_paper.tex as an example.

    \item The camera-ready versions of the accepted papers are 8
      pages, plus any additional pages needed for references.

    \item If you need to include additional appendices,
      you can include them in the supplementary
      material file.

    \item Please, do not change the layout given by the above
      instructions and by the style file.

\end{enumerate}

\subsubsection*{Acknowledgements}
All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support. 
To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.

\subsubsection*{References}

\bibliography{biblio}

References follow the acknowledgements.  Use an unnumbered third level
heading for the references section.  Please use the same font
size for references as for the body of the paper---remember that
references do not count against your page length total.

% \begin{thebibliography}{}
% \setlength{\itemindent}{-\leftmargin}
% \makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
% \bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
%     \newblock Performance of a stochastic learning microchip.
%     \newblock In D. S. Touretzky (ed.),
%     \textit{Advances in Neural Information Processing Systems 1}, 748--760.
%     San Mateo, Calif.: Morgan Kaufmann.

% \bibitem{} F.~Rosenblatt (1962).
%     \newblock \textit{Principles of Neurodynamics.}
%     \newblock Washington, D.C.: Spartan Books.

% \bibitem{} G.~Tesauro (1989).
%     \newblock Neurogammon wins computer Olympiad.
%     \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
% \end{thebibliography}

\end{document}

### 2022-09-05 AISTATS2023/supplement.tex
initialized AISTATS2023 submission

--- 2022-09-05 AISTATS2023/sample_paper.tex
@@
 
-\usepackage{aistats2023}
+\usepackage{aistats2022}
 % If your paper is accepted, change the options for the package
-% aistats2023 as follows:
+% aistats2022 as follows:
 %
-%\usepackage[accepted]{aistats2023}
+%\usepackage[accepted]{aistats2022}
 %
@@
 % If you use natbib package, activate the following three lines:
-\usepackage[round]{natbib}
-\renewcommand{\bibname}{References}
-\renewcommand{\bibsection}{\subsubsection*{\bibname}}
-\bibliographystyle{plainnat}
+%\usepackage[round]{natbib}
+%\renewcommand{\bibname}{References}
+%\renewcommand{\bibsection}{\subsubsection*{\bibname}}
 
@@
 
-\twocolumn[
+% Supplementary material: To improve readability, you must use a single-column format for the supplementary material.
+\onecolumn
+\aistatstitle{Instructions for Paper Submissions to AISTATS 2022: \\
+Supplementary Materials}
 
-\aistatstitle{Instructions for Paper Submissions to AISTATS 2023}
+\section{FORMATTING INSTRUCTIONS}
 
-\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
+To prepare a supplementary pdf file, we ask the authors to use \texttt{aistats2022.sty} as a style file and to follow the same formatting instructions as in the main paper.
+The only difference is that the supplementary material must be in a \emph{single-column} format.
+You can use \texttt{supplement.tex} in our starter pack as a starting point, or append the supplementary content to the main paper and split the final PDF into two separate files.
 
-\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]
+Note that reviewers are under no obligation to examine your supplementary material.
 
-\begin{abstract}
-  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
-  both left and right-hand margins. Use 10~point type, with a vertical
-  spacing of 11~points. The \textbf{Abstract} heading must be centered,
-  bold, and in point size 12. Two line spaces precede the
-  Abstract. The Abstract must be limited to one paragraph.
-\end{abstract}
+\section{MISSING PROOFS}
 
-\section{INTRODUCTION}
+The supplementary materials may contain detailed proofs of the results that are missing in the main paper.
 
-XAI literature distinguishes between local and global interpretation methods~\citep{Molnar2020}.
-Local methods explain a specific prediction, whereas global methods explain the entire model behavior.
-Global methods provide a universal explanation, summarizing the numerous local explanations into a single interpretable outcome (number or plot).
-For example, if a user wants to know which features are significant (feature importance) or whether a particular feature has a positive or negative effect on the output (feature effect), they should opt for a global explainability technique.
-Aggregating the individual explanations for producing a global one comes at a cost.
-In cases where feature interactions are strong, the global explanation may obfuscate heterogeneous effects~\citep{Herbinger2022} that exist under the hood, a phenomenon called aggregation bias~\citep{Mehrabi2019}.
+\subsection{Proof of Lemma 3}
 
-Feature effect forms a fundamental category of global explainability methods.
-The goal of the feature effect is to isolate the average impact of a single feature on the output.
-Feature effect methods suffer from aggregation bias because the rationale behind the average effect might be unclear.
-For example, a feature with zero average effect may indicate that the feature has no effect on the output or has a highly positive effect in some cases and a highly negative one in others.
+\textit{In this section, we present the detailed proof of Lemma 3 and then [ ... ]}
 
-There are two widely-used feature effect methods; Partial Dependence Plots (PDPlots) (Friedman, 2001) and Aggregated Local Effects (ALE)\citep{Apley2020}.
-PDPlots have been criticized for producing erroneous feature effect plots when the input features are correlated due to marginalizing over out-of-distribution synthetic instances.
-Therefore, ALE has been established as the state-of-the-art feature effect method since it can isolate feature effects in situations where input features are highly correlated.
+\section{ADDITIONAL EXPERIMENTS}
 
-However, ALE faces two crucial drawbacks.
-First, it lacks a way to inform the user about potential heterogeneous effects that hide behind the average effect.
-In contrast, in Partial Dependence Plots (PDP), the heterogeneous effects can be spotted by exploring the Individual Conditional Expectations (ICE).
-Second, ALE requires an additional step, where the axis of the feature of interest is split in \(K\) fixed-size non-overlapping intervals, where \(K\) is a hyperparameter provided by the user.
-This splitting is done blindly, which can lead to inconsistent explanations.
+If you have additional experimental results, you may include them in the supplementary materials.
 
-In this paper, we extend ALE with a probabilistic component for measuring the uncertainty of the global explanation.
-The uncertainty of the global explanation expresses how certain we are that the global (expected) explanation is valid if applied to an instance drawn at random, or, in other words, the level of heterogeneous effects hidden behind the expected explanation.
-The uncertainty of the global explanation emerges from the natural characteristics of the experiment, i.e.,~the data generating distribution and the black-box function.
+\subsection{The Effect of Regularization Parameter}
 
-Furthermore, we automate the step of splitting the axis into non-overlapping intervals.
-The need for non-overlapping bins emerges from the ignorance of the data-generating distribution, enforcing all estimations to be based on the limited instances of the training set.
-Therefore, there is an implicit trade-off behind the formation of bins.
-Each bin must include enough instances for a robust estimation of the bin feature effect (expected value), and the uncertainty of the explanation (variance).
-On the other hand, each bin should include points with similar local effects.
-Therefore, we transform the bin splitting step into an unsupervised clustering problem, encoding the trade-off mentioned above in the objective function.
-We formally show that the objective of the clustering problem has lower-bound the (unavoidable) heterogeneity, i.e., the aggregated uncertainty of the global explanation.
-Therefore, we aim to find the optimal grouping of samples that adds the slightest uncertainty over the unavoidable heterogeneity.
-We finally solve the minimization problem by finding the global optimum using dynamic programming.
-Our method works out of the box without requiring any input by the user.
-We provide a theoretical and empirical evaluation of our method.
+\textit{Our algorithm depends on the regularization parameter $\lambda$. Figure 1 below illustrates the effect of this parameter on the performance of our algorithm. As we can see, [ ... ]}
 
-\section{BACKGROUND AND RELATED WORK}
-
-\section{THE U-DALE METHOD}
-
-\subsection{Uncertainty Quantification}
-
-\subsubsection{Methodology}
-
-\subsection{Bin Spliting as a Clustering Problem}
-
-\subsubsection{Methodology}
-
-\subsubsection{Algorithms}
-
-\section{SYNTHETIC EXAMPLES}
-
-\section{REAL-WORLD EXAMPLES}
-
-\section{GENERAL FORMATTING INSTRUCTIONS}
-
-The camera-ready versions of the accepted papers are 8 pages,
-plus any additional pages needed for references.
-
-Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
-Each column is 3.25~inches wide (19.5~picas).  The space
-between the columns is .25~inches wide (1.5~picas).  The left margin is 0.88~inches (5.28~picas).
-Use 10~point type with a vertical spacing of
-11~points. Please use US Letter size paper instead of A4.
-
-Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
-Top rule is 4~points thick and bottom rule is 1~point thick.
-Allow 1/4~inch space above and below title to rules.
-
-Author descriptions are center-justified, initial caps.  The lead
-author is to be listed first (left-most), and the Co-authors are set
-to follow.  If up to three authors, use a single row of author
-descriptions, each one center-justified, and all set side by side;
-with more authors or unusually long names or institutions, use more
-rows.
-
-Use one-half line space between paragraphs, with no indent.
-
-\section{FIRST LEVEL HEADINGS}
-
-First level headings are all caps, flush left, bold, and in point size
-12. Use one line space before the first level heading and one-half line space
-after the first level heading.
-
-\subsection{Second Level Heading}
-
-Second level headings are initial caps, flush left, bold, and in point
-size 10. Use one line space before the second level heading and one-half line
-space after the second level heading.
-
-\subsubsection{Third Level Heading}
-
-Third level headings are flush left, initial caps, bold, and in point
-size 10. Use one line space before the third level heading and one-half line
-space after the third level heading.
-
-\paragraph{Fourth Level Heading}
-
-Fourth level headings must be flush left, initial caps, bold, and
-Roman type.  Use one line space before the fourth level heading, and
-place the section text immediately after the heading with no line
-break, but an 11 point horizontal space.
-
-%%%
-\subsection{Citations, Figure, References}
-
-\subsubsection{Citations in Text}
-
-Citations within the text should include the author's last name and
-year, e.g., (Cheesman, 1985). 
-%Apart from including the author's last name and year, citations can follow any style, as long as the style is consistent throughout the paper.  
-Be sure that the sentence reads
-correctly if the citation is deleted: e.g., instead of ``As described
-by (Cheesman, 1985), we first frobulate the widgets,'' write ``As
-described by Cheesman (1985), we first frobulate the widgets.''
-
-The references listed at the end of the paper can follow any style as long as it is used consistently.
-
-%Be sure to avoid
-%accidentally disclosing author identities through citations.
-
-\subsubsection{Footnotes}
-
-Indicate footnotes with a number\footnote{Sample of the first
-  footnote.} in the text. Use 8 point type for footnotes. Place the
-footnotes at the bottom of the column in which their markers appear,
-continuing to the next column if required. Precede the footnote
-section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
-long.\footnote{Sample of the second footnote.}
-
-\subsubsection{Figures}
-
-All artwork must be centered, neat, clean, and legible.  All lines
-should be very dark for purposes of reproduction, and art work should
-not be hand-drawn.  Figures may appear at the top of a column, at the
-top of a page spanning multiple columns, inline within a column, or
-with text wrapped around them, but the figure number and caption
-always appear immediately below the figure.  Leave 2 line spaces
-between the figure and the caption. The figure caption is initial caps
-and each figure should be numbered consecutively.
-
-Make sure that the figure caption does not get separated from the
-figure. Leave extra white space at the bottom of the page rather than
-splitting the figure and figure caption.
-\begin{figure}[h]
-\vspace{.3in}
-\centerline{\fbox{This figure intentionally left non-blank}}
-\vspace{.3in}
-\caption{Sample Figure Caption}
-\end{figure}
-
-\subsubsection{Tables}
-
-All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
-Table number and title always appear above the table.
-See Table~\ref{sample-table}.
-
-Use one line space before the table title, one line space after the table title,
-and one line space after the table. The table title must be
-initial caps and each table numbered consecutively.
-
-\begin{table}[h]
-\caption{Sample Table Title} \label{sample-table}
-\begin{center}
-\begin{tabular}{ll}
-\textbf{PART}  &\textbf{DESCRIPTION} \\
-\hline \\
-Dendrite         &Input terminal \\
-Axon             &Output terminal \\
-Soma             &Cell body (contains cell nucleus) \\
-\end{tabular}
-\end{center}
-\end{table}
-
-\section{SUPPLEMENTARY MATERIAL}
-
-If you need to include additional appendices during submission, you can include them in the supplementary material file.
-You can submit a single file of additional supplementary material which may be either a pdf file (such as proof details) or a zip file for other formats/more files (such as code or videos). 
-Note that reviewers are under no obligation to examine your supplementary material. 
-If you have only one supplementary pdf file, please upload it as is; otherwise gather everything to the single zip file.
-
-You must use \texttt{aistats2023.sty} as a style file for your supplementary pdf file and follow the same formatting instructions as in the main paper. 
-The only difference is that it must be in a \emph{single-column} format.
-You can use \texttt{supplement.tex} in our starter pack as a starting point.
-Alternatively, you may append the supplementary content to the main paper and split the final PDF into two separate files.
-
-\section{SUBMISSION INSTRUCTIONS}
-
-To submit your paper to AISTATS 2023, please follow these instructions.
-
-\begin{enumerate}
-    \item Download \texttt{aistats2023.sty}, \texttt{fancyhdr.sty}, and \texttt{sample\_paper.tex} provided in our starter pack. 
-    Please, do not modify the style files as this might result in a formatting violation.
-
-    \item Use \texttt{sample\_paper.tex} as a starting point.
-    \item Begin your document with
-    \begin{flushleft}
-    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
-    \texttt{\textbackslash usepackage\{aistats2023\}}
-    \end{flushleft}
-    The \texttt{twoside} option for the class article allows the
-    package \texttt{fancyhdr.sty} to include headings for even and odd
-    numbered pages.
-    \item When you are ready to submit the manuscript, compile the latex file to obtain the pdf file.
-    \item Check that the content of your submission, \emph{excluding} references, is limited to \textbf{8 pages}. The number of pages containing references alone is not limited.
-    \item Upload the PDF file along with other supplementary material files to the CMT website.
-\end{enumerate}
-
-\subsection{Camera-ready Papers}
-
-%For the camera-ready paper, if you are using \LaTeX, please make sure
-%that you follow these instructions.  
-% (If you are not using \LaTeX,
-%please make sure to achieve the same effect using your chosen
-%typesetting package.)
-
-If your papers are accepted, you will need to submit the camera-ready version. Please make sure that you follow these instructions:
-\begin{enumerate}
-    %\item Download \texttt{fancyhdr.sty} -- the
-    %\texttt{aistats2023.sty} file will make use of it.
-    \item Change the beginning of your document to
-    \begin{flushleft}
-    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
-    \texttt{\textbackslash usepackage[accepted]\{aistats2023\}}
-    \end{flushleft}
-    The option \texttt{accepted} for the package
-    \texttt{aistats2023.sty} will write a copyright notice at the end of
-    the first column of the first page. This option will also print
-    headings for the paper.  For the \emph{even} pages, the title of
-    the paper will be used as heading and for \emph{odd} pages the
-    author names will be used as heading.  If the title of the paper
-    is too long or the number of authors is too large, the style will
-    print a warning message as heading. If this happens additional
-    commands can be used to place as headings shorter versions of the
-    title and the author names. This is explained in the next point.
-    \item  If you get warning messages as described above, then
-    immediately after $\texttt{\textbackslash
-    begin\{document\}}$, write
-    \begin{flushleft}
-    \texttt{\textbackslash runningtitle\{Provide here an alternative
-    shorter version of the title of your paper\}}\\
-    \texttt{\textbackslash runningauthor\{Provide here the surnames of
-    the authors of your paper, all separated by commas\}}
-    \end{flushleft}
-    Note that the text that appears as argument in \texttt{\textbackslash
-      runningtitle} will be printed as a heading in the \emph{even}
-    pages. The text that appears as argument in \texttt{\textbackslash
-      runningauthor} will be printed as a heading in the \emph{odd}
-    pages.  If even the author surnames do not fit, it is acceptable
-    to give a subset of author names followed by ``et al.''
-
-    %\item Use the file sample\_paper.tex as an example.
-
-    \item The camera-ready versions of the accepted papers are 8
-      pages, plus any additional pages needed for references.
-
-    \item If you need to include additional appendices,
-      you can include them in the supplementary
-      material file.
-
-    \item Please, do not change the layout given by the above
-      instructions and by the style file.
-
-\end{enumerate}
-
-\subsubsection*{Acknowledgements}
-All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support.
-To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.
-
-\subsubsection*{References}
-
-\bibliography{biblio}
-
-References follow the acknowledgements.  Use an unnumbered third level
-heading for the references section.  Please use the same font
-size for references as for the body of the paper---remember that
-references do not count against your page length total.
-
-% \begin{thebibliography}{}
-% \setlength{\itemindent}{-\leftmargin}
-% \makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
-% \bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
-%     \newblock Performance of a stochastic learning microchip.
-%     \newblock In D. S. Touretzky (ed.),
-%     \textit{Advances in Neural Information Processing Systems 1}, 748--760.
-%     San Mateo, Calif.: Morgan Kaufmann.
-
-% \bibitem{} F.~Rosenblatt (1962).
-%     \newblock \textit{Principles of Neurodynamics.}
-%     \newblock Washington, D.C.: Spartan Books.
-
-% \bibitem{} G.~Tesauro (1989).
-%     \newblock Neurogammon wins computer Olympiad.
-%     \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
-% \end{thebibliography}
+\vfill
 

### 2022-09-05 paper/summary.tex
initialized AISTATS2023 submission

@@
 
+An introductory paragraph for interpretable ML
+
 XAI literature distinguishes between local and global interpretation methods~\cite{Molnar2020}.
-Global interpretation methods aim at explaining the overall behavior of an ML model.
-Many of these global interpretation methods are confounded by feature interactions,
-meaning that they can produce misleading explanations when feature interactions are present.
-In these cases, they often average individual effects of local interpretation methods and thereby obfuscate heterogeneous effects induced by feature interactions~\cite{Herbinger2022}.
+Local methods explain a specific prediction, whereas global methods explain the entire model behavior.
+Global methods can provide a universal explanation, summarizing the numerous local explanations into a single interpretable outcome (number or plot).
+For example, if a user wants to know which features are significant (feature importance) or whether a particular feature has a positive or negative effect on the output (feature effect), they should opt for a global explainability technique.
+Aggregating the individual explanations for producing a global one comes at a cost.
+In cases where feature interactions are strong, the global explanation may obfuscate heterogeneous effects~\cite{Herbinger2022} that exist under the hood, a phenomenon called aggregation bias~\cite{Mehrabi2019}.
 
-This so-called aggregation bias~\cite{Mehrabi2019} is responsible for producing global explanations that often conceal that individual instances may deviate from the global explanation.
-Therefore, global methods must quantify and inform about \textit{the uncertainty of the global explanation}, i.e., how certain we are that a global explanation is valid if applied to a local individual drawn at random.
+Feature effect forms a fundamental category of global explainability methods.
+The goal of the feature effect is to isolate the average impact of a single feature on the output.
+Feature effect methods suffer from aggregation bias because the rationale behind the average effect might be unclear.
+For example, a feature with zero average effect may indicate that the feature has no effect on the output or has a highly positive effect in some cases and a highly negative one in others.
+There are two widely-used feature effect methods; Partial Dependence Plots (PDPlots) (Friedman, 2001) and Aggregated Local Effects (ALE) (Apley and Zhu, 2020).
+PDPlots have been criticized for producing erroneous feature effect plots when the input features are correlated due to marginalizing over out-of-distribution synthetic instances.
+Therefore, ALE has been established as the state-of-the-art feature effect method due to isolating feature effects in situations where input features are highly correlated.
+
+However, ALE faces two crucial drawbacks.
+First, it lacks a way to inform the user about potential heterogeneous effects that hide behind the average effect.
+In contrast, in Partial Dependence Plots (PDP), the heterogeneous effects can be spotted by exploring the Individual Conditional Expectations (ICE).
+Second, ALE require an additional step, where the input space is split in \(K\) fixed-size non-overlapping intervals, where \(K\) is a hyperparameter provided by the user.
+This splitting is done blindly, which can lead to inconsistent results.
+
+In this paper, we extend ALE with a probabilistic component for measuring the uncertainty of the global explanation.
+The uncertainty of the global explanation expresses how certain we are that the global (expected) explanation is valid if applied to a local individual drawn at random, or, in other words, to what extent heterogeneous effects are hidden behind the expected explanation.
 The uncertainty of the global explanation emerges from the natural characteristics of the experiment, i.e.,~the data generating distribution and the black-box function.
 
-In real ML scenarios, we do not know the data generating distribution for computing the expectations and the uncertainty.
-The local effects are estimated from the training set's limited instances, and the global effect is computed by aggregating them.
-Many methods, such as ALE, require an appropriate grouping of samples (partitioning of the feature space) for aggregating local effects that are as homogeneous as possible.
-If this grouping is not done correctly, it may erroneously mix heterogeneous local effects, not due to the unavoidable heterogeneity of the data generating mechanism but due to improper grouping.
-
-ALE is a SotA method for measuring the global feature effect, but so far, it has two limitations.
-First, it does not quantify the uncertainty of the global explanation.
-Second, like most global explanation methods, it requires grouping together samples before computing the global effect.
-So far, this grouping is done by blindly splitting the feature space in \(K\) fixed-size non-overlapping intervals, where \(K\) is a hyperparameter provided by the user.
-
-In this paper, first, we extend the ALE definition for quantifying the uncertainty of the global explanation.
-Second, we readjust ALE to work for variable-size bins and formulate the partitioning in non-overlapping intervals as an unsupervised clustering problem.
-In the unsupervised clustering case, we minimize an objective which has as lower-bound the (unavoidable) heterogeneity, i.e.\ the aggregated uncertainty of the global explanation.
-Therefore, we aim to minimize the added uncertainty induced by the wrong grouping of samples.
-In other words, we aim to find the optimal grouping that adds no uncertainty over the unavoidable heterogeneity.
+Furthermore, we automate the step of splitting the axis into non-overlapping intervals, transforming it into an unsupervised clustering problem, which we solve automatically using dynamic programming.
+The objective of the clustering problem has as lower-bound the (unavoidable) heterogeneity, i.e., the aggregated uncertainty of the global explanation.
+The restriction of the clustering is to create groups of samples for a robust estimation estimating the feature effect  (expected value) and the uncertainty of the explanation (variance), we reside on the limited instances of the training set.
+Therefore, we aim to find the optimal grouping of samples that adds the slightest uncertainty over the unavoidable heterogeneity, but with enough samples.
 We finally solve the minimization problem by finding the global optimum using dynamic programming.
@@
 We provide a theoretical and empirical evaluation of our method.
+
 % In all cases, to estimate
@@
 
-\paragraph{Problem Statement - Contribution.} The contribution of our
+\paragraph{Contribution.} The contribution of our
 paper in bullets:
@@
 
-In real scenarios, we have ignorance about the generative distribution \(\p(x_s, \mathbf{x}_c)\),
+In real scenarios, we have ignorance about the generative distribution \(p(x_s, \mathbf{x}_c)\),
 residing to Monte-Carlo approximations of \(\mu(x_s)\), \(\sigma^2(x_s)\) using the samples of the training set.
@@
 \[
-\(\rho(x_s) = \mu(x_s) - \mu_k\)
+\rho(x_s) = \mu(x_s) - \mu_k
 \]

### 2022-09-06 AISTATS2023/sample_paper.tex
added Intro to the submission

@@
 
-XAI literature distinguishes between local and global interpretation methods~\citep{Molnar2020}.
+Recently, ML has flourished in critical domains, such as healthcare and finance.
+In these areas, we need ML models that predict accurately but also with the ability to explain their predictions.
+Therefore, Explainable AI (XAI) is a rapidly growing field due to the interest in interpreting black box machine learning (ML) models.
+XAI literature distinguishes between local and global interpretation methods~\citep{Molnar2020interpretable}.
 Local methods explain a specific prediction, whereas global methods explain the entire model behavior.
@@
 Aggregating the individual explanations for producing a global one comes at a cost.
-In cases where feature interactions are strong, the global explanation may obfuscate heterogeneous effects~\citep{Herbinger2022} that exist under the hood, a phenomenon called aggregation bias~\citep{Mehrabi2019}.
+In cases where feature interactions are strong, the global explanation may obfuscate heterogeneous effects~\citep{Herbinger2022repid} that exist under the hood, a phenomenon called aggregation bias~\citep{mehrabi2021survey}.
 
-Feature effect forms a fundamental category of global explainability methods.
-The goal of the feature effect is to isolate the average impact of a single feature on the output.
+Feature effect forms a fundamental category of global explainability methods, isolating a single feature's average impact on the output.
 Feature effect methods suffer from aggregation bias because the rationale behind the average effect might be unclear.
-For example, a feature with zero average effect may indicate that the feature has no effect on the output or has a highly positive effect in some cases and a highly negative one in others.
+For example, a feature with zero average effect may indicate that the feature has no effect on the output or, contrarily, it has a highly positive effect in some cases and a highly negative one in others.
 
-There are two widely-used feature effect methods; Partial Dependence Plots (PDPlots) (Friedman, 2001) and Aggregated Local Effects (ALE)\citep{Apley2020}.
+There are two widely-used feature effect methods; Partial Dependence Plots (PDPlots)\citep{friedman2001greedy} and Aggregated Local Effects (ALE)\citep{apley2020visualizing}.
 PDPlots have been criticized for producing erroneous feature effect plots when the input features are correlated due to marginalizing over out-of-distribution synthetic instances.
@@
 However, ALE faces two crucial drawbacks.
-First, it lacks a way to inform the user about potential heterogeneous effects that hide behind the average effect.
-In contrast, in Partial Dependence Plots (PDP), the heterogeneous effects can be spotted by exploring the Individual Conditional Expectations (ICE).
+First, it does not provide a way to inform the user about potential heterogeneous effects that are hidden behind the average effect.
+In contrast, in the case of PDPlots, the heterogeneous effects can be spotted by exploring the Individual Conditional Expectations (ICE)\citep{goldstein2015peeking}.
 Second, ALE requires an additional step, where the axis of the feature of interest is split in \(K\) fixed-size non-overlapping intervals, where \(K\) is a hyperparameter provided by the user.
@@
 In this paper, we extend ALE with a probabilistic component for measuring the uncertainty of the global explanation.
-The uncertainty of the global explanation expresses how certain we are that the global (expected) explanation is valid if applied to an instance drawn at random, or, in other words, the level of heterogeneous effects hidden behind the expected explanation.
+The uncertainty of the global explanation expresses how certain we are that the global (expected) explanation is valid if applied to an instance drawn at random and informs the user about the level of heterogeneous effects hidden behind the expected explanation.
+Our method completes ALE, as ICE plots complement PDPlots, for revealing the heterogeneous effects.
+
+Our method also automates the step of axis splitting into non-overlapping intervals.
+We, firstly, transform the bin splitting step into an unsupervised clustering problem and, second, find the optimal bin splitting for a robust estimation of (a) the global (expected) effect and (b) the uncertainty of the explanation from the limited samples of the training set.
+We formally prove that the objective of the clustering problem has as lower-bound the aggregated uncertainty of the global explanation.
+Our method works out of the box without requiring any input from the user.
+
+\paragraph{Contributions.} The contributions of this paper are the following:
+
+\begin{itemize}
+  \item We introduce Uncertainty DALE (UDALE), an extension of DALE that quantifies the uncertainty of the global explanation, i.e.~the level of heterogeneous effects hidden behind the global explanation.\item We provide an algorithm that automatically computes the optimal bin splitting for robustly estimating the explanatory quantities, i.e., the global effect and the uncertainty.
+  \item We formally prove that our method finds the optimal grouping of samples, minimizing the added uncertainty over the unavoidable heterogeneity that is the lower-bound of the objective.
+  \item We provide empirical evaluation of the method in artificial and real datasets.
+\end{itemize}
+
+The implementation of our method and the code for reproducing all the
+experiments is provided in the submission and will become publicly
+available upon acceptance.
+
+\section{BACKGROUND AND RELATED WORK}
+
+\section{THE U-DALE METHOD}
+
+\subsection{Uncertainty Quantification}
+
 The uncertainty of the global explanation emerges from the natural characteristics of the experiment, i.e.,~the data generating distribution and the black-box function.
+
+\subsubsection{Methodology}
+
+\subsection{Bin Spliting as a Clustering Problem}
+
+\subsubsection{Methodology}
 
@@
 
-\section{BACKGROUND AND RELATED WORK}
-
-\section{THE U-DALE METHOD}
-
-\subsection{Uncertainty Quantification}
-
-\subsubsection{Methodology}
-
-\subsection{Bin Spliting as a Clustering Problem}
-
-\subsubsection{Methodology}
-
 \subsubsection{Algorithms}
@@
 \section{REAL-WORLD EXAMPLES}
-
-\section{GENERAL FORMATTING INSTRUCTIONS}
-
-The camera-ready versions of the accepted papers are 8 pages,
-plus any additional pages needed for references.
-
-Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
-Each column is 3.25~inches wide (19.5~picas).  The space
-between the columns is .25~inches wide (1.5~picas).  The left margin is 0.88~inches (5.28~picas).
-Use 10~point type with a vertical spacing of
-11~points. Please use US Letter size paper instead of A4.
-
-Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
-Top rule is 4~points thick and bottom rule is 1~point thick.
-Allow 1/4~inch space above and below title to rules.
-
-Author descriptions are center-justified, initial caps.  The lead
-author is to be listed first (left-most), and the Co-authors are set
-to follow.  If up to three authors, use a single row of author
-descriptions, each one center-justified, and all set side by side;
-with more authors or unusually long names or institutions, use more
-rows.
-
-Use one-half line space between paragraphs, with no indent.
-
-\section{FIRST LEVEL HEADINGS}
-
-First level headings are all caps, flush left, bold, and in point size
-12. Use one line space before the first level heading and one-half line space
-after the first level heading.
-
-\subsection{Second Level Heading}
-
-Second level headings are initial caps, flush left, bold, and in point
-size 10. Use one line space before the second level heading and one-half line
-space after the second level heading.
-
-\subsubsection{Third Level Heading}
-
-Third level headings are flush left, initial caps, bold, and in point
-size 10. Use one line space before the third level heading and one-half line
-space after the third level heading.
-
-\paragraph{Fourth Level Heading}
-
-Fourth level headings must be flush left, initial caps, bold, and
-Roman type.  Use one line space before the fourth level heading, and
-place the section text immediately after the heading with no line
-break, but an 11 point horizontal space.
-
-%%%
-\subsection{Citations, Figure, References}
-
-\subsubsection{Citations in Text}
-
-Citations within the text should include the author's last name and
-year, e.g., (Cheesman, 1985). 
-%Apart from including the author's last name and year, citations can follow any style, as long as the style is consistent throughout the paper.  
-Be sure that the sentence reads
-correctly if the citation is deleted: e.g., instead of ``As described
-by (Cheesman, 1985), we first frobulate the widgets,'' write ``As
-described by Cheesman (1985), we first frobulate the widgets.''
-
-The references listed at the end of the paper can follow any style as long as it is used consistently.
-
-%Be sure to avoid
-%accidentally disclosing author identities through citations.
-
-\subsubsection{Footnotes}
-
-Indicate footnotes with a number\footnote{Sample of the first
-  footnote.} in the text. Use 8 point type for footnotes. Place the
-footnotes at the bottom of the column in which their markers appear,
-continuing to the next column if required. Precede the footnote
-section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
-long.\footnote{Sample of the second footnote.}
-
-\subsubsection{Figures}
-
-All artwork must be centered, neat, clean, and legible.  All lines
-should be very dark for purposes of reproduction, and art work should
-not be hand-drawn.  Figures may appear at the top of a column, at the
-top of a page spanning multiple columns, inline within a column, or
-with text wrapped around them, but the figure number and caption
-always appear immediately below the figure.  Leave 2 line spaces
-between the figure and the caption. The figure caption is initial caps
-and each figure should be numbered consecutively.
-
-Make sure that the figure caption does not get separated from the
-figure. Leave extra white space at the bottom of the page rather than
-splitting the figure and figure caption.
-\begin{figure}[h]
-\vspace{.3in}
-\centerline{\fbox{This figure intentionally left non-blank}}
-\vspace{.3in}
-\caption{Sample Figure Caption}
-\end{figure}
-
-\subsubsection{Tables}
-
-All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
-Table number and title always appear above the table.
-See Table~\ref{sample-table}.
-
-Use one line space before the table title, one line space after the table title,
-and one line space after the table. The table title must be
-initial caps and each table numbered consecutively.
-
-\begin{table}[h]
-\caption{Sample Table Title} \label{sample-table}
-\begin{center}
-\begin{tabular}{ll}
-\textbf{PART}  &\textbf{DESCRIPTION} \\
-\hline \\
-Dendrite         &Input terminal \\
-Axon             &Output terminal \\
-Soma             &Cell body (contains cell nucleus) \\
-\end{tabular}
-\end{center}
-\end{table}
-
-\section{SUPPLEMENTARY MATERIAL}
-
-If you need to include additional appendices during submission, you can include them in the supplementary material file.
-You can submit a single file of additional supplementary material which may be either a pdf file (such as proof details) or a zip file for other formats/more files (such as code or videos). 
-Note that reviewers are under no obligation to examine your supplementary material. 
-If you have only one supplementary pdf file, please upload it as is; otherwise gather everything to the single zip file.
-
-You must use \texttt{aistats2023.sty} as a style file for your supplementary pdf file and follow the same formatting instructions as in the main paper. 
-The only difference is that it must be in a \emph{single-column} format.
-You can use \texttt{supplement.tex} in our starter pack as a starting point.
-Alternatively, you may append the supplementary content to the main paper and split the final PDF into two separate files.
-
-\section{SUBMISSION INSTRUCTIONS}
-
-To submit your paper to AISTATS 2023, please follow these instructions.
-
-\begin{enumerate}
-    \item Download \texttt{aistats2023.sty}, \texttt{fancyhdr.sty}, and \texttt{sample\_paper.tex} provided in our starter pack. 
-    Please, do not modify the style files as this might result in a formatting violation.
-
-    \item Use \texttt{sample\_paper.tex} as a starting point.
-    \item Begin your document with
-    \begin{flushleft}
-    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
-    \texttt{\textbackslash usepackage\{aistats2023\}}
-    \end{flushleft}
-    The \texttt{twoside} option for the class article allows the
-    package \texttt{fancyhdr.sty} to include headings for even and odd
-    numbered pages.
-    \item When you are ready to submit the manuscript, compile the latex file to obtain the pdf file.
-    \item Check that the content of your submission, \emph{excluding} references, is limited to \textbf{8 pages}. The number of pages containing references alone is not limited.
-    \item Upload the PDF file along with other supplementary material files to the CMT website.
-\end{enumerate}
-
-\subsection{Camera-ready Papers}
-
-%For the camera-ready paper, if you are using \LaTeX, please make sure
-%that you follow these instructions.  
-% (If you are not using \LaTeX,
-%please make sure to achieve the same effect using your chosen
-%typesetting package.)
-
-If your papers are accepted, you will need to submit the camera-ready version. Please make sure that you follow these instructions:
-\begin{enumerate}
-    %\item Download \texttt{fancyhdr.sty} -- the
-    %\texttt{aistats2023.sty} file will make use of it.
-    \item Change the beginning of your document to
-    \begin{flushleft}
-    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
-    \texttt{\textbackslash usepackage[accepted]\{aistats2023\}}
-    \end{flushleft}
-    The option \texttt{accepted} for the package
-    \texttt{aistats2023.sty} will write a copyright notice at the end of
-    the first column of the first page. This option will also print
-    headings for the paper.  For the \emph{even} pages, the title of
-    the paper will be used as heading and for \emph{odd} pages the
-    author names will be used as heading.  If the title of the paper
-    is too long or the number of authors is too large, the style will
-    print a warning message as heading. If this happens additional
-    commands can be used to place as headings shorter versions of the
-    title and the author names. This is explained in the next point.
-    \item  If you get warning messages as described above, then
-    immediately after $\texttt{\textbackslash
-    begin\{document\}}$, write
-    \begin{flushleft}
-    \texttt{\textbackslash runningtitle\{Provide here an alternative
-    shorter version of the title of your paper\}}\\
-    \texttt{\textbackslash runningauthor\{Provide here the surnames of
-    the authors of your paper, all separated by commas\}}
-    \end{flushleft}
-    Note that the text that appears as argument in \texttt{\textbackslash
-      runningtitle} will be printed as a heading in the \emph{even}
-    pages. The text that appears as argument in \texttt{\textbackslash
-      runningauthor} will be printed as a heading in the \emph{odd}
-    pages.  If even the author surnames do not fit, it is acceptable
-    to give a subset of author names followed by ``et al.''
-
-    %\item Use the file sample\_paper.tex as an example.
-
-    \item The camera-ready versions of the accepted papers are 8
-      pages, plus any additional pages needed for references.
-
-    \item If you need to include additional appendices,
-      you can include them in the supplementary
-      material file.
-
-    \item Please, do not change the layout given by the above
-      instructions and by the style file.
-
-\end{enumerate}
 
@@
 
-References follow the acknowledgements.  Use an unnumbered third level
-heading for the references section.  Please use the same font
-size for references as for the body of the paper---remember that
-references do not count against your page length total.
-
-% \begin{thebibliography}{}
-% \setlength{\itemindent}{-\leftmargin}
-% \makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
-% \bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
-%     \newblock Performance of a stochastic learning microchip.
-%     \newblock In D. S. Touretzky (ed.),
-%     \textit{Advances in Neural Information Processing Systems 1}, 748--760.
-%     San Mateo, Calif.: Morgan Kaufmann.
-
-% \bibitem{} F.~Rosenblatt (1962).
-%     \newblock \textit{Principles of Neurodynamics.}
-%     \newblock Washington, D.C.: Spartan Books.
-
-% \bibitem{} G.~Tesauro (1989).
-%     \newblock Neurogammon wins computer Olympiad.
-%     \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
-% \end{thebibliography}
-
 \end{document}

### 2022-09-06 AISTATS2023/paper.tex
started Related Work

--- 2022-09-06 AISTATS2023/sample_paper.tex
@@
 \bibliographystyle{plainnat}
+\usepackage{amsmath}
+\usepackage{amssymb}
 
@@
 %\bibliographystyle{apalike}
+
+\newcommand{\dfdx}{\frac{\partial f}{\partial x_s}}
+\newcommand{\xc}{\mathbf{x_c}}
+\newcommand{\DY}{\mathbf{\Delta Y}}
 
@@
 
-\section{THE U-DALE METHOD}
+Some notation for describing the methods afterwards.
+
+It is crucial for feature effect methods to inform about the
+heterogeneous effects. Elaborate.
+
+There are two established feture effect methods PDPlots and ALE. ALE
+has some important advantages. Elaborate.
+
+Interpretation of the heterogeneous effects behind the global effect
+is available only for PDP, with three different approaches; (a) ICE
+and d-ICE plots provide a visual understanding of the heterogeneous
+effects. (b) grouping of ICE in homogeneous clusters, for spliting the
+input space into subspace(s) with homogeneous effects (c) Feature
+Interaction strength indexes, like H-statistic, provide a value
+indicating how much a feature interacts with the others (not the type
+of interaction).
+
+There is no method for quantifying the heterogeneous effects, based on
+ALE. Therefore, no method to exploit the advantages of ALE while, on
+the same time, informing about the heterogeneous effects. We present
+it in the next section.
+
+ALE also has the peculiarity of spliting the axis into intervals,
+allocating the instances of the training set in the intervals and
+compute a single (constant) effect in each interval. With DALE
+extension, bin spliting is decoupled from instant effect
+estimation. With our extension for measuring the heterogeneous
+effects, we transfrom interval spliting from a step to a clustering
+problem with a meaningful objective to minimise. We provide a thorough
+analysis, where we show that our objective has a consistent
+meaning. It can be split in two parts; the first part is the
+unavoidable uncertainty due to the natural characteristics of the
+experiment, i.e.,~the data generating distribution and the black-box
+function. The second part is an added uncertainty due to the
+limited-samples estimation, that enforces to create groups with
+constant main effect. We opt for minimizing the objective, i.e. sum of
+the two uncertainties, that given that the first uncertainty is
+independent of the bin spliting, therefore we want to minimize the
+added uncertainty. To conclude, we transform the axis-spliting into an
+unsupervised clustering problem with a principled objective. We a
+computationally-grounded solution that works out-of-the-box, relaxing
+the user from providing a hyperparameter without any indication which
+one is the correct. This step can be used independently of whether the
+user wants to explore the heterogeneous effects or not.
+
+\section{OUR METHOD}
 
@@
 
-The uncertainty of the global explanation emerges from the natural characteristics of the experiment, i.e.,~the data generating distribution and the black-box function.
+ALE is a state-of-the-art method when it comes to feature effect
+estimation. Unfortunately, there is no method for quantifying the
+heterogeneous effects, based on ALE.
+
+% The uncertainty of the global explanation emerges from the natural characteristics of the experiment, i.e.,~the data generating distribution and the black-box function.
 
 \subsubsection{Methodology}
+
+% \paragraph{Effect at point \(x_s\).} In the intro, we described the
+% \textit{uncertainty of the global explanation} as a metric of how
+% certain we are that the global explanation is valid if applied to a
+% randomly-drawn individual.
+% ALE plots measure the \(s\)-th feature
+% effect at point \(x_s\) as the \textbf{expected} change in the output \(y\), if
+% we slightly change the value of the feature of interest \(x_s\):
+
+% \begin{equation}
+%   \label{eq:ALE_mean}
+%   \mu(x_s) = \mathbb{E}_{\xc|x_s}\left [\dfdx (x_s, \xc)\right ]
+% \end{equation}
+
+% \noindent
+% We model the s-th feature effect at point \(x_s\) (change in the output \(y\) wrt a slight change in the feature of interest \(x_s\)) as the random variable $\DY ; x_s $.
+% For notation convenience, we will refer to the s-th feature effect as \(\DY\), ommiting the \(;xs\) part.
+% The randomness has its origins in the ignorance of the values of the rest of the features, denoted with \(\mathbf{X}_c\).
+% Therefore, the s-th feature effect is defined as:
+
+% \begin{equation}
+%   \label{eq:fe_rv}
+%   \DY = g(x_s) = \dfdx (x_s, \mathbf{X}_c)
+% \end{equation}
+% %
+% As shown in Eq.~\eqref{eq:ALE_mean}, ALE is only interested in the expected value of \(\DY\).
+% Instead, we are also interested in the variance of \(\DY\) for measuring the uncertainty of the local change:
+
+% \begin{equation}
+%   \label{eq:ALE_var}
+%   \sigma^2(x_s) = \mathrm{Var}_{\xc|x_s}\left [\dfdx (x_s, \xc) \right ]
+% \end{equation}
+
+% \noindent
+% The variance in Eq.~\eqref{eq:ALE_var} informs us about the heterogeneous
+% effects hiding behind the explanation.
+
+% \paragraph{Effect at interval \([z_{k-1}, z_k]\).}
+
+% In real scenarios, we have ignorance about the generative distribution \(p(x_s, \mathbf{x}_c)\),
+% residing to Monte-Carlo approximations of \(\mu(x_s)\), \(\sigma^2(x_s)\) using the samples of the training set.
+% Therefore, it is impossible to estimate Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
+% point \(x_s\)\, since the  possibility to observe a sample in the interval
+% \([x_s - h, x_s + h]\) is zero, in the limit where \(h \to 0\).
+% Therefore, we are obliged to estimate the local effect in larger intervals (\(h>0\)).
+% We refer to the expected value and the variance of the feature effect at the interval \([z_{k-1}, z_k)\), as:
+
+% \begin{equation}
+%   \label{eq:mu_bin}
+%   \mu_k = \mu(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k}
+%   \mathbb{E}_{\xc|x_s=z}\left [\frac{\partial f}{\partial x_s} \right ] \partial z
+% \end{equation}
+
+% \noindent
+
+% \begin{equation}
+%   \label{eq:var_bin}
+%   \sigma^2_k = \sigma^2(z_{k-1}, z_k) = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k}
+%   \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
+%       f}{\partial x_s}(x_s, \xc) - \mu_k )^2 \right] \partial z
+% \end{equation}
+
+% \noindent
+% We prove that the interval-variance \(\sigma^2_k \) is:
+
+% \begin{equation}
+%  \sigma^2_k = \frac{1}{z_k - z_{k-1}} \int_{z_{k-1}}^{z_k} \sigma^2(z) + \rho^2(z) \partial z
+% \end{equation}
+% %
+% where \(\rho(x_s) = \mu(x_s) - \mu_k\).
+% The proof is at Section \ref{sec:proofs}.
+% We observe that the interval-variance is the mean point-variance of the points
+% inside the interval \([z_{k-1}, z_k]\), plus the mean of the residual term \(\rho\).
+% The mean point-variance is the unavoidable variance, due to the uncertainty of the global explanation.
+% The mean of the residual term is an extra variance (uncertainty) due to limiting the granularity of the effect at
+% the bin level.
+
+% \noindent
+
+% \paragraph{Approximation of effect at interval \([z_{k-1}, z_k]\).}
+% Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} are
+% approximated by the instances of the training set that lie inside the
+% \(k\)-th interval, i.e.
+% \( \mathcal{S}_k = \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_{k} \} \):
+
+% \begin{equation}
+%   \label{eq:mean_estimation}
+%   \hat{\mu}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
+%     \mathcal{S}_k} \left [ \dfdx(\mathbf{x}^i) \right ]
+% \end{equation}
+
+% \begin{equation}
+%   \label{eq:variance_estimation}
+%   \hat{\sigma}_k(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
+%   \mathcal{S}_k} \left [ \dfdx(\mathbf{x}^i) - \hat{\mu}_k(z_{k-1}, z_k) \right ]^2
+% \end{equation}
+
+% \paragraph{Uncertainty of the global effect.}
+
+% Eq.~\eqref{eq:variance_estimation} gives an approximation of the
+% uncertainty of the bin effect.
+% The uncertainty of the global effect is
+% simply the sum of the uncertainties in the bin effects.
+% The
+% approximation is unbiased only if the points are uniformly distributed
+% in \([z_{k-1}, z_k]\). (TODOs: Check what happens otherwise).
+
+% \paragraph{Minimizing the uncertainty}
+
+% Solving the problem of finding (a) the optimal number of bins \(K\) and (b) the optimal bin limits for each bin \([z_{k-1}, z_k] \forall k\) to minimize:
+
+% \begin{equation}
+%   \label{eq:1}
+%   \mathcal{L} = \sum_{k=0}^K \hat{\sigma}_k(z_{k-1}, z_k)
+% \end{equation}
+% %
+% The constraints are that all bins must include more than \(\tau\)
+% points, i.e., \(|\mathcal{S}_k| \geq \tau\).
+
+% \noindent
+% TODOS. Show theoretically that \(\mathcal{L} \geq \int_{x_{s, \min}}^{x_{s, \max}}\sigma^2(x_s) \partial x_s\)
+
+% \paragraph{Uncertainty of the approximation.}
+
+% In all experiments, it also important to measure the uncertainty of
+% the approximation.
+% The uncertainty of the approximation can be
+% quantified with two approaches:
+
+% \begin{itemize}
+% \item Splitting the dataset in many folds and (re)estimating
+%   \(\hat{\mu}(z_{k-1}, z_k), \hat{\sigma}_k(z_{k-1}, z_k)\)
+% \item Using the central limit theorem, we can (under assumptions) say
+%   that the standard error of the approximation in
+%   eq.~\eqref{eq:mean_estimation} is
+%   \(\mathrm{std\_error} =
+%   \frac{\hat{\sigma}_k}{\sqrt{|\mathcal{S}_k|}}\).
+% \end{itemize}
 
@@
 
-Furthermore, we automate the step of splitting the axis into non-overlapping intervals.
+% Furthermore, we automate the step of splitting the axis into non-overlapping intervals.
 The need for non-overlapping bins emerges from the ignorance of the data-generating distribution, enforcing all estimations to be based on the limited instances of the training set.

### 2022-09-07 AISTATS2023/paper.tex
started method description

--- 2022-09-06 AISTATS2023/sample_paper.tex
@@
 %\bibliographystyle{apalike}
+
+\newcommand{\dfdx}{\frac{\partial f}{\partial x_s}}
+\newcommand{\xc}{\mathbf{x_c}}
+\newcommand{\DY}{\mathbf{\Delta Y}}
+\newcommand{\xb}{\mathbf{x}}
+\newcommand{\Xcb}{\mathcal{X}_c}
+\newcommand{\Xb}{\mathcal{X}}
 
@@
 
-\section{THE U-DALE METHOD}
+\paragraph{Notation.}
 
-\subsection{Uncertainty Quantification}
+We refer to random variables (rv) using uppercase \( X \), whereas to
+simple variables with plain lowercase \( x \). Bold denotes a vector;
+\( \xb \) for simple variables or \(\mathbf{X}\) for rvs. Often, we
+partition the input vector \(\xb \in \mathbb{R}^D\) to the feature of
+interest \(x_s \in \mathbb{R} \) and the rest of the features
+\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
+\((x_s, \mathbb{x}_c)\), but we clarify that it corresponds to the vector
+\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
+corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
+function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
+feature effect of the \(s\)-th feature is
+\(f_{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
+name of the feature effect method. \footnote{An extensive list of all
+  symbols used in the paper is provided in the helping material.}
 
-The uncertainty of the global explanation emerges from the natural characteristics of the experiment, i.e.,~the data generating distribution and the black-box function.
+\paragraph{Feature Effect Methods.} PDPlots formulate the feature
+effect of the \(s\)-th attribute as an expectation over the marginal
+distribution \(\mathbf{X}_c\), i.e.,
+\(f_{\mathtt{PDP}}(x_s) =
+\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPlots formulate
+it as an expectation over the conditional
+\(\mathcal{X}_c|\mathcal{X}_s\), i.e.,
+\(f_{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathcal{X}_c|\mathcal{X}_s = x_s}[f(x_s,
+\mathcal{X}_c)]\). ALE computes the global effect at \(x_s\) as an
+integration of local effects. The local effects are measured as the
+expected change on the output
+\( \frac{\partial f(x_s, \mathcal{X}_c)}{\partial x_s} \) over the
+conditional distribution \( \Xcb|\mathcal{X}_s\). The formula that
+defines ALE is presented below:
+
+\begin{equation}
+  \label{eq:ALE}
+  f_{\mathtt{ALE}}(x_s) = c + \int_{-\infty}^{x_s} \mathbb{E}_{\Xcb|\mathcal{X}_s=z}\left[\frac{\partial f(z, \mathcal{X}_c)}{\partial z}\right] \partial z
+\end{equation}
+%
+The constant \(c\) is used for centering the ALE plot. PDPlots
+integrate over unrealistic instances due to the use of the marginal
+distribution \( p(\mathcal{X}_1) \). Therefore, they incorrectly
+result in a quadratic effect in the region \(x_1 \in [0, 1]\). MPlots
+resolve this issue using the conditional distribution
+\( \mathcal{X}_2|\mathcal{X}_1 \) but suffer from computing combined
+effects. ALE plots resolve, both they have two drawbacks TODO add more
+info.
+
+\paragraph{Qunatify the Heterogeneous Effects.}
+It is crucial for feature effect methods to inform about the
+heterogeneous effects. Elaborate. Interpretation of the heterogeneous
+effects behind the global effect is available only for PDP, with three
+different approaches; (a) ICE and d-ICE plots provide a visual
+understanding of the heterogeneous effects. (b) grouping of ICE in
+homogeneous clusters, for spliting the input space into subspace(s)
+with homogeneous effects (c) Feature Interaction strength indexes,
+like H-statistic, provide a value indicating how much a feature
+interacts with the others (not the type of interaction). There is no
+method for quantifying the heterogeneous effects, based on
+ALE. Therefore, no method to exploit the advantages of ALE while, on
+the same time, informing about the heterogeneous effects.
+
+\paragraph{Bin Spliting.} ALE also has the peculiarity of spliting the
+axis into intervals, allocating the instances of the training set in
+the intervals and compute a single (constant) effect in each
+interval. With DALE extension, bin spliting is decoupled from instant
+effect estimation. With our extension for measuring the heterogeneous
+effects, we transfrom interval spliting from a step to a clustering
+problem with a meaningful objective to minimise. We provide a thorough
+analysis, where we show that our objective has a consistent
+meaning. It can be split in two parts; the first part is the
+unavoidable uncertainty due to the natural characteristics of the
+experiment, i.e.,~the data generating distribution and the black-box
+function. The second part is an added uncertainty due to the
+limited-samples estimation, that enforces to create groups with
+constant main effect. We opt for minimizing the objective, i.e. sum of
+the two uncertainties, that given that the first uncertainty is
+independent of the bin spliting, therefore we want to minimize the
+added uncertainty. To conclude, we transform the axis-spliting into an
+unsupervised clustering problem with a principled objective. We a
+computationally-grounded solution that works out-of-the-box, relaxing
+the user from providing a hyperparameter without any indication which
+one is the correct. This step can be used independently of whether the
+user wants to explore the heterogeneous effects or not.
+
+\section{THE ... METHOD}
+
+\subsection{ALE with Uncertainty Quantification}
+
+ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at
+\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). Given that the black-box
+function \(f(\cdot)\) is a deterministic predictor, the local effect
+is also a deterministic quantity. Global methods summarize the local
+explanations into a single quantity. Therefore, ALE summarizes all the
+local explanations at \(x_s\) by averaging the local effects accross
+all values of \(\xc\) weighting them by \(p(\xc|x_s)\). In other
+words, ALE globalizes the local explanations at \(x_s\) through the
+expected change in the output \(y\) wrt \(x_s\):
+
+\begin{equation}
+  \label{eq:ALE_mean}
+  \mu(x_s) = \mathbb{E}_{\xc|x_s}\left [\dfdx (x_s, \xc)\right ]
+\end{equation}
+%
+As described at the Introduction, limiting the explanation at this
+level does not shed light to possible heterogeneous effects behind the
+averaged explanation. Therefore, we model the model the uncertainty of
+the effect at \(x_s\) as the variance of the local explanations:
+
+\begin{equation}
+  \label{eq:ALE_var}
+  \sigma^2(x_s) = \mathrm{Var}_{\xc|x_s}\left [\dfdx (x_s, \xc) \right ]
+\end{equation}
+
+\noindent
+%
+The uncertainty of the explanation emerges from the natural
+characteristics of the experiment, i.e.,~the data generating
+distribution and the black-box function. It is important to clarify
+that the variance is only a way to model the uncertainty. Other
+statistical properties can also be used. Elaborate.
+
+ALE computes the final effect at \(x_s\) by accumulating/integrating
+the averaged local effects \(\mu (x_s=z) \) over all values of \(z\)
+from \(x_{min}\) until \(x_s\), as show in Eq.~(\ref{eq:ALE}). The
+choice for \(x_{min}\) is not important, as it only affects the
+centering of ALE plot along the vertical axis. Equivalently, we define
+the accumulated uncertainty until the point \(x_s\), as the integral
+of the variance of local effects:
+
+\begin{equation}
+  \label{eq:3}
+  f_{\mathtt{ALE}}^{\sigma^2}(x_s) = \int_{z_1, min}^{x_s} \sigma^2(z) \partial z
+\end{equation}
+
+In ALE plots the absolute value at a specific point is not an
+interpretable quantity (add ref to paper). The meaningful
+interpretation is the effect at a specific point, i.e. what happens to
+the output given a small change in the feature of interest. The
+integration takes place only for making the visual interpretation
+smooth, for example, to spot larger intervals where the effect of a
+\textbf{local} change is continuously negative. The same stands for
+the accumulated uncertainty. It is meaningful to measure, only, the
+uncertainty at at a specific point to understand the heterogeneous
+effects behind the averaged effect. In the experimental Section, we
+propose appropriate visualizations to concentrate the user to this
+effect. However, the accumulated uncertainty, i.e. the aggregate
+uncertainty allong all the points of the plot, will help us define a
+meaningfull objective for the interval spliting step. (TODO: check
+here if the idea about whether the accumulated uncertainty is a good
+metric for modeling the interaction strength)
+
+\subsection{Uncertainty Quantification at an interval}
+
+In real scenarios, we have ignorance about the data-generating
+distribution \(p(x_s, \mathbf{x}_c)\) and on estimations based on the
+limited instances of the training set. Therefore, it is impossible to
+estimate Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the
+granularity of a point \(x_s\)\, since the possibility to observe a
+sample in the interval \([x_s - h, x_s + h]\) is zero, when
+\(h \to 0\). Therefore, for estimation purposes, we are obliged to
+create larger intervals (\(h>0\)). The mean effect at the interval
+\([z_1, z_2)\) is the mean of the expected effects:
+
+\begin{equation}
+  \label{eq:mu_bin}
+  \mu(z_1, z_2) = \frac{1}{z_2 - z_1} \int_{z_1}^{z_2}
+  \mathbb{E}_{\xc|x_s=z}\left [\frac{\partial f}{\partial x_s} \right ] \partial z
+\end{equation}
+
+\noindent
+Accordingly, the accumulated uncertainty (variance) at the interval
+\([z_1, z_2)\) is:
+
+\begin{equation}
+  \label{eq:var_bin}
+  \sigma^2(z_1, z_2) = \int_{z_1}^{z_2}
+  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
+      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z
+\end{equation}
+
+\noindent
+
+\paragraph{Theorem 1.}
+
+The accumulated variance at the interval \([z_1, z_2)\) is the
+accumulation of the all variances inside the interval plus the
+accumulated squared residuals:
+
+\begin{equation}
+ \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
+\end{equation}
+%
+where \(\rho(z) = \mu(z) - \mu(z_1, z_2)\). The proof is at the
+Appendix. Theorem 1 decouples the (observable) variance in the
+interval \([z_1, z_2)\) into two terms. The first term
+\(\int_{z_1}^{z_2} \sigma^2(z) \partial z\) accumulates the variance
+of the local explanations (Eq.~(\ref{eq:ALE_var})) for all points
+inside the interval. The second term accumulates the squared
+differences between the mean effect at the interval and mean of the
+local explanations for all points inside the interval. The first term
+accumulates the the uncertainty due to the natural characteristics of
+the experiment, whereas, the second term adds extra uncertainty due to
+limiting the granularity of the effect at the bin level.
+
+\subsection{Estimation of Effect at an Interval}
+
+As stated before, Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be
+estimated from the set \(\mathcal{S}\) of the instances of the
+training set with the \(s\)-th feature lying inside the interval,
+i.e., \( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \}
+\). The mean effect at the interval, Eq.~(\ref{eq:mu_bin}), can be approximated
+by:
+
+\begin{equation}
+  \label{eq:mean_estimation}
+  \hat{\mu}(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
+    \mathcal{S}} \left [ \dfdx(\mathbf{x}^i) \right ]
+\end{equation}
+
+and the accumulated variance at the interval, Eq.~(\ref{eq:var_bin})
+can be approximated by
+
+\begin{equation}
+  \label{eq:variance_estimation}
+  \hat{\sigma}_k(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
+  \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}_k(z_1, z_2) \right )^2
+\end{equation}
+
+The approximation is unbiased only if the points are uniformly
+distributed in \([z_{k-1}, z_k]\). (TODOs: Check what happens
+otherwise).
+
+\subsection{Bin Spliting as a Clustering Problem}
+
+\paragraph{Uncertainty of the global effect.}
+
+Eq.~\eqref{eq:variance_estimation} gives an approximation of the
+uncertainty of the bin effect.
+The uncertainty of the global effect is
+simply the sum of the uncertainties in the bin effects.
+
+\paragraph{Minimizing the uncertainty}
+
+Solving the problem of finding (a) the optimal number of bins \(K\) and (b) the optimal bin limits for each bin \([z_{k-1}, z_k] \forall k\) to minimize:
+
+\begin{equation}
+  \label{eq:1}
+  \mathcal{L} = \sum_{k=0}^K \hat{\sigma}_k(z_{k-1}, z_k)
+\end{equation}
+%
+The constraints are that all bins must include more than \(\tau\)
+points, i.e., \(|\mathcal{S}_k| \geq \tau\).
+
+\noindent
+TODOS. Show theoretically that \(\mathcal{L} \geq \int_{x_{s, \min}}^{x_{s, \max}}\sigma^2(x_s) \partial x_s\)
 
@@
 
-\subsection{Bin Spliting as a Clustering Problem}
-
-\subsubsection{Methodology}
-
-Furthermore, we automate the step of splitting the axis into non-overlapping intervals.
+% Furthermore, we automate the step of splitting the axis into non-overlapping intervals.
 The need for non-overlapping bins emerges from the ignorance of the data-generating distribution, enforcing all estimations to be based on the limited instances of the training set.
@@
 
-\subsubsection*{References}
-
 \bibliography{biblio}
 
+\section*{Appendix}
+
+\subsection{Proof for variance of the bin}
+
 \end{document}

### 2022-09-08 AISTATS2023/paper.tex
continued method description

@@
 \newcommand{\xb}{\mathbf{x}}
-\newcommand{\Xcb}{\mathcal{X}_c}
+\newcommand{\Xcb}{\mathbf{X}_c}
 \newcommand{\Xb}{\mathcal{X}}
@@
 
-Recently, ML has flourished in critical domains, such as healthcare and finance.
-In these areas, we need ML models that predict accurately but also with the ability to explain their predictions.
-Therefore, Explainable AI (XAI) is a rapidly growing field due to the interest in interpreting black box machine learning (ML) models.
-XAI literature distinguishes between local and global interpretation methods~\citep{Molnar2020interpretable}.
-Local methods explain a specific prediction, whereas global methods explain the entire model behavior.
-Global methods provide a universal explanation, summarizing the numerous local explanations into a single interpretable outcome (number or plot).
-For example, if a user wants to know which features are significant (feature importance) or whether a particular feature has a positive or negative effect on the output (feature effect), they should opt for a global explainability technique.
-Aggregating the individual explanations for producing a global one comes at a cost.
-In cases where feature interactions are strong, the global explanation may obfuscate heterogeneous effects~\citep{Herbinger2022repid} that exist under the hood, a phenomenon called aggregation bias~\citep{mehrabi2021survey}.
+Recently, ML has flourished in critical domains, such as healthcare
+and finance. In these areas, we need ML models that predict accurately
+but also with the ability to explain their predictions. Therefore,
+Explainable AI (XAI) is a rapidly growing field due to the interest in
+interpreting black box machine learning (ML) models. XAI literature
+distinguishes between local and global interpretation
+methods~\citep{Molnar2020interpretable}. Local methods explain a
+specific prediction, whereas global methods explain the entire model
+behavior. Global methods provide a universal explanation, summarizing
+the numerous local explanations into a single interpretable outcome
+(number or plot). For example, if a user wants to know which features
+are significant (feature importance) or whether a particular feature
+has a positive or negative effect on the output (feature effect), they
+should opt for a global explainability technique. Aggregating the
+individual explanations for producing a global one comes at a cost. In
+cases where feature interactions are strong, the global explanation
+may obfuscate heterogeneous effects~\citep{Herbinger2022repid} that
+exist under the hood, a phenomenon called aggregation
+bias~\citep{mehrabi2021survey}.
 
-Feature effect forms a fundamental category of global explainability methods, isolating a single feature's average impact on the output.
-Feature effect methods suffer from aggregation bias because the rationale behind the average effect might be unclear.
-For example, a feature with zero average effect may indicate that the feature has no effect on the output or, contrarily, it has a highly positive effect in some cases and a highly negative one in others.
+Feature effect forms a fundamental category of global explainability
+methods, isolating a single feature's average impact on the
+output. Feature effect methods suffer from aggregation bias because
+the rationale behind the average effect might be unclear. For example,
+a feature with zero average effect may indicate that the feature has
+no effect on the output or, contrarily, it has a highly positive
+effect in some cases and a highly negative one in others.
 
-There are two widely-used feature effect methods; Partial Dependence Plots (PDPlots)\citep{friedman2001greedy} and Aggregated Local Effects (ALE)\citep{apley2020visualizing}.
-PDPlots have been criticized for producing erroneous feature effect plots when the input features are correlated due to marginalizing over out-of-distribution synthetic instances.
-Therefore, ALE has been established as the state-of-the-art feature effect method since it can isolate feature effects in situations where input features are highly correlated.
+There are two widely-used feature effect methods; Partial Dependence
+Plots (PDPlots)\citep{friedman2001greedy} and Aggregated Local Effects
+(ALE)\citep{apley2020visualizing}. PDPlots have been criticized for
+producing erroneous feature effect plots when the input features are
+correlated due to marginalizing over out-of-distribution synthetic
+instances. Therefore, ALE has been established as the state-of-the-art
+feature effect method since it can isolate feature effects in
+situations where input features are highly correlated.
 
-However, ALE faces two crucial drawbacks.
-First, it does not provide a way to inform the user about potential heterogeneous effects that are hidden behind the average effect.
-In contrast, in the case of PDPlots, the heterogeneous effects can be spotted by exploring the Individual Conditional Expectations (ICE)\citep{goldstein2015peeking}.
-Second, ALE requires an additional step, where the axis of the feature of interest is split in \(K\) fixed-size non-overlapping intervals, where \(K\) is a hyperparameter provided by the user.
-This splitting is done blindly, which can lead to inconsistent explanations.
+However, ALE faces two crucial drawbacks. First, it does not provide a
+way to inform the user about potential heterogeneous effects that are
+hidden behind the average effect. In contrast, in the case of PDPlots,
+the heterogeneous effects can be spotted by exploring the Individual
+Conditional Expectations (ICE)\citep{goldstein2015peeking}. Second,
+ALE requires an additional step, where the axis of the feature of
+interest is split in \(K\) fixed-size non-overlapping intervals, where
+\(K\) is a hyperparameter provided by the user. This splitting is done
+blindly, which can lead to inconsistent explanations.
 
-In this paper, we extend ALE with a probabilistic component for measuring the uncertainty of the global explanation.
-The uncertainty of the global explanation expresses how certain we are that the global (expected) explanation is valid if applied to an instance drawn at random and informs the user about the level of heterogeneous effects hidden behind the expected explanation.
-Our method completes ALE, as ICE plots complement PDPlots, for revealing the heterogeneous effects.
+In this paper, we extend ALE with a probabilistic component for
+measuring the uncertainty of the global explanation. The uncertainty
+of the global explanation expresses how certain we are that the global
+(expected) explanation is valid if applied to an instance drawn at
+random and informs the user about the level of heterogeneous effects
+hidden behind the expected explanation. Our method completes ALE, as
+ICE plots complement PDPlots, for revealing the heterogeneous effects.
 
-Our method also automates the step of axis splitting into non-overlapping intervals.
-We, firstly, transform the bin splitting step into an unsupervised clustering problem and, second, find the optimal bin splitting for a robust estimation of (a) the global (expected) effect and (b) the uncertainty of the explanation from the limited samples of the training set.
-We formally prove that the objective of the clustering problem has as lower-bound the aggregated uncertainty of the global explanation.
-Our method works out of the box without requiring any input from the user.
+Our method also automates the step of axis splitting into
+non-overlapping intervals. We, firstly, transform the bin splitting
+step into an unsupervised clustering problem and, second, find the
+optimal bin splitting for a robust estimation of (a) the global
+(expected) effect and (b) the uncertainty of the explanation from the
+limited samples of the training set. We formally prove that the
+objective of the clustering problem has as lower-bound the aggregated
+uncertainty of the global explanation. Our method works out of the box
+without requiring any input from the user.
 
@@
 \begin{itemize}
-  \item We introduce Uncertainty DALE (UDALE), an extension of DALE that quantifies the uncertainty of the global explanation, i.e.~the level of heterogeneous effects hidden behind the global explanation.\item We provide an algorithm that automatically computes the optimal bin splitting for robustly estimating the explanatory quantities, i.e., the global effect and the uncertainty.
-  \item We formally prove that our method finds the optimal grouping of samples, minimizing the added uncertainty over the unavoidable heterogeneity that is the lower-bound of the objective.
-  \item We provide empirical evaluation of the method in artificial and real datasets.
+\item We introduce Uncertainty DALE (UDALE), an extension of DALE that
+  quantifies the uncertainty of the global explanation, i.e.~the level
+  of heterogeneous effects hidden behind the global explanation.
+\item
+  We provide an algorithm that automatically computes the optimal bin
+  splitting for robustly estimating the explanatory quantities, i.e.,
+  the global effect and the uncertainty.
+\item We formally prove that our method finds the optimal grouping of
+  samples, minimizing the added uncertainty over the unavoidable
+  heterogeneity that is the lower-bound of the objective.
+\item We provide empirical evaluation of the method in artificial and
+  real datasets.
 \end{itemize}
@@
 \(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
-\((x_s, \mathbb{x}_c)\), but we clarify that it corresponds to the vector
-\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
-corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
+\((x_s, \mathbf{x}_c)\), but we clarify that it corresponds to the
+vector \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote
+the corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
 function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
 feature effect of the \(s\)-th feature is
-\(f_{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
-name of the feature effect method. \footnote{An extensive list of all
+\(f^{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
+name of the feature effect method.\footnote{An extensive list of all
   symbols used in the paper is provided in the helping material.}
@@
 distribution \(\mathbf{X}_c\), i.e.,
-\(f_{\mathtt{PDP}}(x_s) =
-\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPlots formulate
-it as an expectation over the conditional
-\(\mathcal{X}_c|\mathcal{X}_s\), i.e.,
-\(f_{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathcal{X}_c|\mathcal{X}_s = x_s}[f(x_s,
-\mathcal{X}_c)]\). ALE computes the global effect at \(x_s\) as an
+\(f^{\mathtt{PDP}}(x_s) =
+\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPlots formulate it
+as an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
+\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
+\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as an
 integration of local effects. The local effects are measured as the
 expected change on the output
-\( \frac{\partial f(x_s, \mathcal{X}_c)}{\partial x_s} \) over the
-conditional distribution \( \Xcb|\mathcal{X}_s\). The formula that
+\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \) over the
+conditional distribution \( \Xcb|X_s\). The formula that
 defines ALE is presented below:
@@
 \begin{equation}
-  \label{eq:ALE}
-  f_{\mathtt{ALE}}(x_s) = c + \int_{-\infty}^{x_s} \mathbb{E}_{\Xcb|\mathcal{X}_s=z}\left[\frac{\partial f(z, \mathcal{X}_c)}{\partial z}\right] \partial z
+  \label{eq:ALE_accumulated_mean}
+  f^{\mathtt{ALE}}(x_s) = c + \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
 \end{equation}
@@
 
-\paragraph{Qunatify the Heterogeneous Effects.}
-It is crucial for feature effect methods to inform about the
-heterogeneous effects. Elaborate. Interpretation of the heterogeneous
-effects behind the global effect is available only for PDP, with three
-different approaches; (a) ICE and d-ICE plots provide a visual
-understanding of the heterogeneous effects. (b) grouping of ICE in
-homogeneous clusters, for spliting the input space into subspace(s)
-with homogeneous effects (c) Feature Interaction strength indexes,
-like H-statistic, provide a value indicating how much a feature
-interacts with the others (not the type of interaction). There is no
-method for quantifying the heterogeneous effects, based on
-ALE. Therefore, no method to exploit the advantages of ALE while, on
-the same time, informing about the heterogeneous effects.
+\paragraph{Qunatify the Heterogeneous Effects.} It is crucial for
+feature effect methods to inform about the heterogeneous
+effects. Elaborate. Interpretation of the heterogeneous effects behind
+the global effect is available only for PDP, with three different
+approaches; (a) ICE and d-ICE plots provide a visual understanding of
+the heterogeneous effects. (b) grouping of ICE in homogeneous
+clusters, for spliting the input space into subspace(s) with
+homogeneous effects (c) Feature Interaction strength indexes, like
+H-statistic, provide a value indicating how much a feature interacts
+with the others (not the type of interaction). There is no method for
+quantifying the heterogeneous effects, based on ALE. Therefore, no
+method to exploit the advantages of ALE while, on the same time,
+informing about the heterogeneous effects.
 
@@
 
+The terms \textit{uncertainty of explanation} and \textit{level of
+  heterogeneous effects} refer to the same effect
+
 \subsection{ALE with Uncertainty Quantification}
@@
 explanations into a single quantity. Therefore, ALE summarizes all the
-local explanations at \(x_s\) by averaging the local effects accross
+local explanations at \(x_s\) by averaging the local effects across
 all values of \(\xc\) weighting them by \(p(\xc|x_s)\). In other
-words, ALE globalizes the local explanations at \(x_s\) through the
-expected change in the output \(y\) wrt \(x_s\):
+words, ALE globalizes the local explanations at \(x_s\), computing the
+expected effect at \(x_s\):
 
@@
   \label{eq:ALE_mean}
-  \mu(x_s) = \mathbb{E}_{\xc|x_s}\left [\dfdx (x_s, \xc)\right ]
+  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
 \end{equation}
-%
+\noindent
 As described at the Introduction, limiting the explanation at this
 level does not shed light to possible heterogeneous effects behind the
-averaged explanation. Therefore, we model the model the uncertainty of
-the effect at \(x_s\) as the variance of the local explanations:
+averaged explanation. Therefore, we model the uncertainty of the local
+effects at \(x_s\) as the variance of the local explanations:
 
@@
   \label{eq:ALE_var}
-  \sigma^2(x_s) = \mathrm{Var}_{\xc|x_s}\left [\dfdx (x_s, \xc) \right ]
+  \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
 \end{equation}
-
 \noindent
-%
 The uncertainty of the explanation emerges from the natural
@@
 the averaged local effects \(\mu (x_s=z) \) over all values of \(z\)
-from \(x_{min}\) until \(x_s\), as show in Eq.~(\ref{eq:ALE}). The
+from \(x_{min}\) until \(x_s\), as show in Eq.~(\ref{eq:ALE_accumulated_mean}). The
 choice for \(x_{min}\) is not important, as it only affects the
 centering of ALE plot along the vertical axis. Equivalently, we define
-the accumulated uncertainty until the point \(x_s\), as the integral
-of the variance of local effects:
+the accumulated uncertainty (variance) until the point \(x_s\), as the
+integral of the variance of local effects:
 
 \begin{equation}
-  \label{eq:3}
-  f_{\mathtt{ALE}}^{\sigma^2}(x_s) = \int_{z_1, min}^{x_s} \sigma^2(z) \partial z
+  \label{eq:ALE_accumulated_var}
+  f^{\mathtt{ALE}}_{\sigma^2}(x_s) = \int_{z_{s, min}}^{x_s} \sigma^2(z) \partial z
 \end{equation}
+\noindent
+In ALE plots the accumulated effect as absolute value is not an
+interpretable quantity and is only used for better visual
+interpretation\citep{Gromping2020MAEP}. The meaningful interpretation
+is the effect at a specific point, i.e., what happens to the output
+given a small change in the feature of interest. The integration takes
+place only for making the visual interpretation easier, for example,
+to spot larger intervals where the effect of a \textbf{local} change
+is continuously constant. The same stands for the accumulated
+uncertainty. The uncertainty has a meaninful interpretation as the
+level of heterogeneous effects, only at a specific point. In the
+experimental Section, we propose appropriate visualizations to
+communicate to this effect to the user. However, the accumulated
+uncertainty, i.e., the aggregated variance of the local effects, helps
+us define a sensible objective for the interval spliting step. (TODO:
+check here if the idea about whether the accumulated uncertainty is a
+good metric for modeling the interaction strength)
 
-In ALE plots the absolute value at a specific point is not an
-interpretable quantity (add ref to paper). The meaningful
-interpretation is the effect at a specific point, i.e. what happens to
-the output given a small change in the feature of interest. The
-integration takes place only for making the visual interpretation
-smooth, for example, to spot larger intervals where the effect of a
-\textbf{local} change is continuously negative. The same stands for
-the accumulated uncertainty. It is meaningful to measure, only, the
-uncertainty at at a specific point to understand the heterogeneous
-effects behind the averaged effect. In the experimental Section, we
-propose appropriate visualizations to concentrate the user to this
-effect. However, the accumulated uncertainty, i.e. the aggregate
-uncertainty allong all the points of the plot, will help us define a
-meaningfull objective for the interval spliting step. (TODO: check
-here if the idea about whether the accumulated uncertainty is a good
-metric for modeling the interaction strength)
-
-\subsection{Uncertainty Quantification at an interval}
+\subsection{Uncertainty Quantification and Estimation at an Interval}
 
 In real scenarios, we have ignorance about the data-generating
-distribution \(p(x_s, \mathbf{x}_c)\) and on estimations based on the
-limited instances of the training set. Therefore, it is impossible to
-estimate Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the
-granularity of a point \(x_s\)\, since the possibility to observe a
-sample in the interval \([x_s - h, x_s + h]\) is zero, when
-\(h \to 0\). Therefore, for estimation purposes, we are obliged to
-create larger intervals (\(h>0\)). The mean effect at the interval
+distribution \(p(x_s, \mathbf{x}_c)\) and all estimations are based on
+the limited instances of the training set. Unfortunately, it is
+impossible to estimate Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at
+the granularity of a point \(x_s\), because The possibility to observe
+a sample in the interval \([x_s - h, x_s + h]\) is zero, when
+\(h \to 0\). Therefore, we are obliged split the axis \(x_s\) into a
+sequence of larger non-overlaping intervals and estimate mean and
+variance at the interval resolution. The mean effect at an interval
 \([z_1, z_2)\) is the mean of the expected effects:
@@
 \noindent
-Accordingly, the accumulated uncertainty (variance) at the interval
-\([z_1, z_2)\) is:
+Accordingly, the accumulated variance at an interval \([z_1, z_2)\)
+is:
 
@@
 
-\noindent
-
-\paragraph{Theorem 1.}
-
-The accumulated variance at the interval \([z_1, z_2)\) is the
-accumulation of the all variances inside the interval plus the
-accumulated squared residuals:
-
-\begin{equation}
- \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
-\end{equation}
-%
-where \(\rho(z) = \mu(z) - \mu(z_1, z_2)\). The proof is at the
-Appendix. Theorem 1 decouples the (observable) variance in the
-interval \([z_1, z_2)\) into two terms. The first term
-\(\int_{z_1}^{z_2} \sigma^2(z) \partial z\) accumulates the variance
-of the local explanations (Eq.~(\ref{eq:ALE_var})) for all points
-inside the interval. The second term accumulates the squared
-differences between the mean effect at the interval and mean of the
-local explanations for all points inside the interval. The first term
-accumulates the the uncertainty due to the natural characteristics of
-the experiment, whereas, the second term adds extra uncertainty due to
-limiting the granularity of the effect at the bin level.
-
-\subsection{Estimation of Effect at an Interval}
-
 As stated before, Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be
@@
 i.e., \( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \}
-\). The mean effect at the interval, Eq.~(\ref{eq:mu_bin}), can be approximated
-by:
+\). The mean effect at the interval, Eq.~(\ref{eq:mu_bin}), can be
+approximated by:
 
@@
   \label{eq:variance_estimation}
-  \hat{\sigma}_k(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
+  \hat{\sigma}(z_1, z_2) = \frac{z_2 - z_1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
     \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}_k(z_1, z_2) \right )^2
@@
 The approximation is unbiased only if the points are uniformly
-distributed in \([z_{k-1}, z_k]\). (TODOs: Check what happens
+distributed in \([z_1, z_2]\). (TODOs: Check what happens
 otherwise).
@@
 \subsection{Bin Spliting as a Clustering Problem}
+
+ALE authors\citep{apley2020visualizing} proposed an approximation of
+(\ref{eq:ALE_accumulated_mean}) by partitioning the axis of the
+feature of interest into a sequence of non-overlaping bin and
+computing a single effect in each bin. Later, (TODO: add citation to
+us), proposed the DALE (Differential ALE) approximation that decouples
+the two steps. The local effects are computed once using the partial
+derivatives on the available instances of the training set,
+i.e. \(\frac{\partial f}{\partial x_s}\), and the bin-spliting step,
+simply allocates the precomputed local effects into bins, for
+estimating the expected effect in each bin. The formula that estimates
+the accumulate effects is:
+
+\begin{equation}
+  f^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x}
+  \frac{1}{|\mathcal{S}_k|} \sum_{i:\xb^i \in \mathcal{S}_k}
+  [f_s(\xb^i)] = \Delta x \sum_{k=1}^{k_x} \hat{\mu}_k
+ \label{eq:DALE}
+\end{equation}
+
+We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
+i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
+the set of training instance that lie in the \(k\)-th bin, i.e.
+\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \} \). Both
+methods face the limitation that the partitioning into non-overlaping
+intervals is done blindly. The user pass the total number of bins
+\(K\) as a hyperparameter, the bins are defined with equal-size
+spliting, and the training instances are allocated accordingly. This
+approach is vulnerable to non-robust estimations. The mean effect is
+often poorly approximated from a very small number of samples and the
+mean effect of empty bins is interpolated from their
+neighboors. Furthermore, in our case, we need sufficient sample
+populations for estimating the variance of the approximation, apart
+from the mean effect.
+
+\subsubsection{Methodology}
+
+For overcoming this limitations, we reformulate the partitioning as a
+clustering of the training instances into a sequence variable-size
+intervals. The objective of the clustering problem is inspired
+
+ALE requires splittingthe estimation of the
+
+In this section, we introduce a framework
+
+\paragraph{Theorem 1.} If we define the residual \(\rho(z)\) as the
+difference between the expected effect at \(x_s\) and the mean
+expected effect at the interval, i.e
+\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
+at an interval \([z_1, z_2)\) is the accumulation of the all variances
+plus the accumulation of squared residuals inside the interval:
+
+\begin{equation}
+ \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
+\end{equation}
+%
+The proof is at the Appendix. Theorem 1 decouples the accumulated
+variance at an interval, the only quantity we can estimate, into two
+terms. The first term \(\int_{z_1}^{z_2} \sigma^2(z) \partial z\),
+quantifies the uncertainty due to the natural characteristics of the
+experiment and the second term adds extra uncertainty due to the
+limited resolution.
 
@@
 
-\subsubsection{Methodology}
-
 % Furthermore, we automate the step of splitting the axis into non-overlapping intervals.
@@
 
-\subsubsection{Algorithms}
+\subsection{Visualization of ALE with Uncertainty}
 

### 2022-09-09 AISTATS2023/paper.tex
continued method description

@@
 
+TODO: Discuss the interpretation of ALE. The differences from PDP. Analyze both methods as processes.
+
 \paragraph{Qunatify the Heterogeneous Effects.} It is crucial for
@@
 
-The terms \textit{uncertainty of explanation} and \textit{level of
-  heterogeneous effects} refer to the same effect
-
 \subsection{ALE with Uncertainty Quantification}
 
-ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at
-\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). Given that the black-box
-function \(f(\cdot)\) is a deterministic predictor, the local effect
-is also a deterministic quantity. Global methods summarize the local
-explanations into a single quantity. Therefore, ALE summarizes all the
-local explanations at \(x_s\) by averaging the local effects across
-all values of \(\xc\) weighting them by \(p(\xc|x_s)\). In other
-words, ALE globalizes the local explanations at \(x_s\), computing the
-expected effect at \(x_s\):
+ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at a single
+point\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
+explanations at \(x_s\) are weigthed by the conditional distribution
+\(p(\xc|x_s)\) and are averaged, to produce the summarized effect:
 
@@
 averaged explanation. Therefore, we model the uncertainty of the local
-effects at \(x_s\) as the variance of the local explanations:
+effects at \(\mathcal{H}(x_s)\) as the variance of the local
+explanations:
 
@@
   \label{eq:ALE_var}
-  \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
+  \mathcal{H}(x_s) := \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
 \end{equation}
@@
 characteristics of the experiment, i.e.,~the data generating
-distribution and the black-box function. It is important to clarify
-that the variance is only a way to model the uncertainty. Other
-statistical properties can also be used. Elaborate.
-
-ALE computes the final effect at \(x_s\) by accumulating/integrating
-the averaged local effects \(\mu (x_s=z) \) over all values of \(z\)
-from \(x_{min}\) until \(x_s\), as show in Eq.~(\ref{eq:ALE_accumulated_mean}). The
-choice for \(x_{min}\) is not important, as it only affects the
-centering of ALE plot along the vertical axis. Equivalently, we define
-the accumulated uncertainty (variance) until the point \(x_s\), as the
-integral of the variance of local effects:
+distribution and the properties of the black-box function. In Section
+(TODO), we propose appropriate visualizations for easier
+interpretation of Eq.~(\ref{eq:ALE_var}). In ALE, the feature effect
+at \(x_s\) is the accumulation of the averaged local effects from
+\(x_{min}\) until \(x_s\), as show in
+Eq.~(\ref{eq:ALE_accumulated_mean}). Equivalently, we define the
+accumulated uncertainty (variance) until the point \(x_s\), as the
+integral of the variances of local effects:
 
@@
 \noindent
-In ALE plots the accumulated effect as absolute value is not an
-interpretable quantity and is only used for better visual
-interpretation\citep{Gromping2020MAEP}. The meaningful interpretation
-is the effect at a specific point, i.e., what happens to the output
-given a small change in the feature of interest. The integration takes
-place only for making the visual interpretation easier, for example,
-to spot larger intervals where the effect of a \textbf{local} change
-is continuously constant. The same stands for the accumulated
-uncertainty. The uncertainty has a meaninful interpretation as the
-level of heterogeneous effects, only at a specific point. In the
-experimental Section, we propose appropriate visualizations to
-communicate to this effect to the user. However, the accumulated
-uncertainty, i.e., the aggregated variance of the local effects, helps
-us define a sensible objective for the interval spliting step. (TODO:
-check here if the idea about whether the accumulated uncertainty is a
-good metric for modeling the interaction strength)
+
+The accumulated uncertainty is not a directly interpretable
+quantity. It only helps us define a sensible objective for the
+interval spliting step, as we discuss in Section TODO: adde ref.
 
@@
 distribution \(p(x_s, \mathbf{x}_c)\) and all estimations are based on
-the limited instances of the training set. Unfortunately, it is
-impossible to estimate Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at
-the granularity of a point \(x_s\), because The possibility to observe
-a sample in the interval \([x_s - h, x_s + h]\) is zero, when
-\(h \to 0\). Therefore, we are obliged split the axis \(x_s\) into a
-sequence of larger non-overlaping intervals and estimate mean and
-variance at the interval resolution. The mean effect at an interval
-\([z_1, z_2)\) is the mean of the expected effects:
+the limited instances of the training set. Estimating
+Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
+point \(x_s\) is impossible, because the probability of observing a
+sample inside the interval \([x_s - h, x_s + h]\) tends to zero, when
+\(h \to 0\). Therefore, we are obliged to split the axis of \(x_s\)
+into a sequence of non-overlaping intervals (bins) and estimate the
+mean and the variance from the samples that lie inside each bin. The
+mean effect at an interval \([z_1, z_2)\) is defined as the mean of
+the expected effects:
 
@@
 Accordingly, the accumulated variance at an interval \([z_1, z_2)\)
-is:
+is defined as:
 
@@
 
-As stated before, Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be
-estimated from the set \(\mathcal{S}\) of the instances of the
-training set with the \(s\)-th feature lying inside the interval,
-i.e., \( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \}
-\). The mean effect at the interval, Eq.~(\ref{eq:mu_bin}), can be
-approximated by:
+Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be directly estimated
+from the set \(\mathcal{S}\) of the dataset instances with the
+\(s\)-th feature lying inside the interval, i.e.,
+\( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \} \). The mean
+effect at the interval, Eq.~(\ref{eq:mu_bin}) is approximated by:
 
@@
 
-and the accumulated variance at the interval, Eq.~(\ref{eq:var_bin})
-can be approximated by
+and the accumulated variance, Eq.~(\ref{eq:var_bin}) can be
+approximated by
 
@@
   \label{eq:variance_estimation}
-  \hat{\sigma}(z_1, z_2) = \frac{z_2 - z_1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
-    \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}_k(z_1, z_2) \right )^2
+  \hat{\sigma}^2(z_1, z_2) = \frac{z_2 - z_1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
+    \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}(z_1, z_2) \right )^2
 \end{equation}
@@
 
-ALE authors\citep{apley2020visualizing} proposed an approximation of
-(\ref{eq:ALE_accumulated_mean}) by partitioning the axis of the
-feature of interest into a sequence of non-overlaping bin and
-computing a single effect in each bin. Later, (TODO: add citation to
-us), proposed the DALE (Differential ALE) approximation that decouples
-the two steps. The local effects are computed once using the partial
-derivatives on the available instances of the training set,
-i.e. \(\frac{\partial f}{\partial x_s}\), and the bin-spliting step,
-simply allocates the precomputed local effects into bins, for
-estimating the expected effect in each bin. The formula that estimates
-the accumulate effects is:
+ALE, Eq.(\ref{eq:ALE_accumulated_mean}), is estimated by spliting the
+axis \(x_s\) into a sequence of non-overlaping bins (TODO: add some
+discussion for ALE, DALE):
 
 \begin{equation}
-  f^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x}
-  \frac{1}{|\mathcal{S}_k|} \sum_{i:\xb^i \in \mathcal{S}_k}
-  [f_s(\xb^i)] = \Delta x \sum_{k=1}^{k_x} \hat{\mu}_k
- \label{eq:DALE}
+  \label{eq:ALE_accumulated_mean_est}
+  \hat{f}^{\mathtt{ALE}}_{\mu}(x_s) = \Delta x \sum_{k=1}^{k_x} \hat{\mu}(z_{k-1}, z_k)
+\end{equation}
+
+and
+
+\begin{equation}
+  \label{eq:ALE_accumulated_var_est}
+  \hat{f}^{\mathtt{ALE}}_{\sigma^2}(x_s) = \sum_{k=1}^{k_x} \hat{\sigma}^2(z_{k-1}, z_k)
 \end{equation}

### 2022-09-12 AISTATS2023/paper.tex
updated background

@@
 
-\paragraph{Notation.}
-
-We refer to random variables (rv) using uppercase \( X \), whereas to
-simple variables with plain lowercase \( x \). Bold denotes a vector;
-\( \xb \) for simple variables or \(\mathbf{X}\) for rvs. Often, we
-partition the input vector \(\xb \in \mathbb{R}^D\) to the feature of
-interest \(x_s \in \mathbb{R} \) and the rest of the features
+\paragraph{Notation.} We refer to random variables (rv) using
+uppercase \( X \), whereas to simple variables with plain lowercase
+\( x \). Bold denotes a vector; \( \xb \) for simple variables or
+\(\mathbf{X}\) for rvs. Often, we partition the input vector
+\(\xb \in \mathbb{R}^D\) to the feature of interest
+\(x_s \in \mathbb{R} \) and the rest of the features
 \(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
@@
 
-\paragraph{Feature Effect Methods.} PDPlots formulate the feature
-effect of the \(s\)-th attribute as an expectation over the marginal
-distribution \(\mathbf{X}_c\), i.e.,
+\paragraph{Feature Effect Methods.} There are three well-known feature
+effect methods: PDPlots, MPlots and ALE. PDPlots formulate the feature effect of the \(s\)-th
+attribute as an expectation over the marginal distribution
+\(\mathbf{X}_c\), i.e.,
 \(f^{\mathtt{PDP}}(x_s) =
@@
 \mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as an
-integration of local effects. The local effects are measured as the
-expected change on the output
-\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \) over the
-conditional distribution \( \Xcb|X_s\). The formula that
-defines ALE is presented below:
+accumulation (integration) of the expected value of the local
+effects:
+\begin{equation}
+  \label{eq:ALE_accumulated_mean}
+  f^{\mathtt{ALE}}(x_s) = \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
+\end{equation}
+%
+ALE has specific advantages which gain particular value in cases of
+corelated input features. In this cases, PDPlots integrate over
+unrealistic instances, due to the use of the marginal distribution
+\( p(\mathbf{X}_c) \), and MPlots compute aggregated effects, i.e.,
+impute the combined effect of sets of features to a single
+feature. ALE manages to resolve both issues, and is therefore the only
+trustable method in cases of correlated features.
+
+\paragraph{Quantify the Heterogeneous Effects.}
+
+Feature effect methods answer the question \textit{what happens
+  (effect) to the output, if I increase/decrease the value of a
+  specific feature}. Having answered the question above, it comes
+naturally to also wonder \textit{how certain we are about the change
+  on the output}. For this reason, a lot of interest is given lately
+for quantifying the level of uncertainty, along with the expected
+effect. The level of uncertainty is mostly quantified by measuring the
+existence of heterogeneous effects, i.e. whether there are local
+explanations that deviate from the expected global effect. ICE and
+d-ICE plots provide a visual understanding of the heterogeneous
+effects on top of PDPs. Another approach targets on grouping the
+heterogeneous effects, e.g., allocating ICE plots in homogeneous
+clusters, by dividing the input space. Some other approaches, like
+H-Statistic, Greenwel, move a step behind and try to quantify the
+level of interaction between the input features, a possible cause of
+heterogeneous effects. In this case, the interpretation is indirect,
+since a strong interaction index indicates the possibility of
+heterogeneous effects. The aforementioned approaches face two
+pathogenies; They either do not quantify the uncertainty of the
+feature effect directly or they are based on PDPs, and, therefore,
+they are subject to the failure modes of PDPs in cases of correlated
+features. To the best of our knowledge, no method so far targets on
+quantify the uncertainty of the feature effect as it is modelled by
+ALE.
+
+\paragraph{Cluster Instances with homogeneous effects.}
+
+In real ML scenarios, the expected feature effect and the uncertainty
+are estimated from the limited instances of the training set. ALE
+approximation requires an additional step, where the axis of the
+\(s\)-th feature is split into a sequence of non-overlaping bins and a
+single effect (expectation and uncertainty) is computed for from the
+population of instances that lie inside each
+bin. \citep{apley2020visualizing} proposed estimating the local
+effects in each bin by evaluating the black box-funtion at the bin
+limits:
 
 \begin{equation}
-  \label{eq:ALE_accumulated_mean}
-  f^{\mathtt{ALE}}(x_s) = c + \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
+  \label{eq:ALE_accumulated_mean_est}
+  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
+    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
+\end{equation}
+
+In contrast, (cite) proposed the Differential ALE (DALE) estimation
+for quantifying the local effects on the training-set instances,
+instead of the bin limits:
+
+\begin{equation}
+  \label{eq:DALE_accumulated_mean_est}
+  \hat{f}^{\mathtt{ALE}}_{\mu}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
+    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
 \end{equation}
 %
-The constant \(c\) is used for centering the ALE plot. PDPlots
-integrate over unrealistic instances due to the use of the marginal
-distribution \( p(\mathcal{X}_1) \). Therefore, they incorrectly
-result in a quadratic effect in the region \(x_1 \in [0, 1]\). MPlots
-resolve this issue using the conditional distribution
-\( \mathcal{X}_2|\mathcal{X}_1 \) but suffer from computing combined
-effects. ALE plots resolve, both they have two drawbacks TODO add more
-info.
+Their method has the advantages of remaining on-distribution even when
+bins become wider and, most importantly, it allows the recomputation
+of the accumulated effect with different bin-splitting with near-zero
+computational overhead. However, none of the approximations above
+deals with the crucial problem of the optimal bin-spliting. They split
+the axis in a blind-way, partition in \(K\) equally-sized which can
+lead to erroneous approximations.
 
-TODO: Discuss the interpretation of ALE. The differences from PDP. Analyze both methods as processes.
-
-\paragraph{Qunatify the Heterogeneous Effects.} It is crucial for
-feature effect methods to inform about the heterogeneous
-effects. Elaborate. Interpretation of the heterogeneous effects behind
-the global effect is available only for PDP, with three different
-approaches; (a) ICE and d-ICE plots provide a visual understanding of
-the heterogeneous effects. (b) grouping of ICE in homogeneous
-clusters, for spliting the input space into subspace(s) with
-homogeneous effects (c) Feature Interaction strength indexes, like
-H-statistic, provide a value indicating how much a feature interacts
-with the others (not the type of interaction). There is no method for
-quantifying the heterogeneous effects, based on ALE. Therefore, no
-method to exploit the advantages of ALE while, on the same time,
-informing about the heterogeneous effects.
-
-\paragraph{Bin Spliting.} ALE also has the peculiarity of spliting the
-axis into intervals, allocating the instances of the training set in
-the intervals and compute a single (constant) effect in each
-interval. With DALE extension, bin spliting is decoupled from instant
-effect estimation. With our extension for measuring the heterogeneous
-effects, we transfrom interval spliting from a step to a clustering
-problem with a meaningful objective to minimise. We provide a thorough
-analysis, where we show that our objective has a consistent
-meaning. It can be split in two parts; the first part is the
-unavoidable uncertainty due to the natural characteristics of the
-experiment, i.e.,~the data generating distribution and the black-box
-function. The second part is an added uncertainty due to the
-limited-samples estimation, that enforces to create groups with
-constant main effect. We opt for minimizing the objective, i.e. sum of
-the two uncertainties, that given that the first uncertainty is
-independent of the bin spliting, therefore we want to minimize the
-added uncertainty. To conclude, we transform the axis-spliting into an
-unsupervised clustering problem with a principled objective. We a
-computationally-grounded solution that works out-of-the-box, relaxing
-the user from providing a hyperparameter without any indication which
-one is the correct. This step can be used independently of whether the
-user wants to explore the heterogeneous effects or not.
+Instead, we propose treating the axis-spliting step as an unsupervised
+clustering problem. The objective of the clustering problem should
+fulfill in the best way to contradictory ojectives. First, secure
+robust estimations of the expected effect and the uncertainty inside
+each bin given the limited instances of the training set and, second,
+create bins with as homogeneous local effects as possible, for not
+losing fine-grain resolution feature effects due to wide bins.
 
@@
 
-ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at a single
-point\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
-explanations at \(x_s\) are weigthed by the conditional distribution
-\(p(\xc|x_s)\) and are averaged, to produce the summarized effect:
+ALE defines the local effect of the \(s\)-th feature on \(f(\cdot)\)
+at point \((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
+explanations at \(x_s\) are, then, weigthed by the conditional
+distribution \(p(\xc|x_s)\) and are averaged, to produce the
+summarized effect at \(x_s\):
 
@@
 \noindent
-As described at the Introduction, limiting the explanation at this
-level does not shed light to possible heterogeneous effects behind the
-averaged explanation. Therefore, we model the uncertainty of the local
-effects at \(\mathcal{H}(x_s)\) as the variance of the local
-explanations:
+As described at the Introduction, limiting the explanation to the
+exepected value level does not shed light to possible heterogeneous
+effects behind the averaged explanation. Therefore, we model the
+uncertainty of the local effects at \(\mathcal{H}(x_s)\) as the
+variance of the local explanations:
 
@@
 \end{equation}
-
 and

### 2022-09-12 AISTATS2023/paper.tex
updated Intro

--- 2022-09-09 AISTATS2023/paper.tex
@@
 \usepackage{amssymb}
+\usepackage{hyperref}
 
@@
 Recently, ML has flourished in critical domains, such as healthcare
-and finance. In these areas, we need ML models that predict accurately
-but also with the ability to explain their predictions. Therefore,
-Explainable AI (XAI) is a rapidly growing field due to the interest in
-interpreting black box machine learning (ML) models. XAI literature
-distinguishes between local and global interpretation
+and finance. In these areas, we need ML models that not only predict
+accurately but also explain their predictions. Therefore, Explainable
+AI (XAI), the field that provides interpretations for the prediction
+mechanism of complex black-box models, has increased interest. XAI
+literature distinguishes between local and global interpretation
 methods~\citep{Molnar2020interpretable}. Local methods explain a
@@
 behavior. Global methods provide a universal explanation, summarizing
-the numerous local explanations into a single interpretable outcome
-(number or plot). For example, if a user wants to know which features
-are significant (feature importance) or whether a particular feature
-has a positive or negative effect on the output (feature effect), they
+the numerous local explanations into a single interpretable outcome,
+usually a number or a plot. If a user wants to know which features are
+significant (feature importance) or whether a particular feature has a
+positive or negative effect on the output (feature effect), they
 should opt for a global explainability technique. Aggregating the
-individual explanations for producing a global one comes at a cost. In
-cases where feature interactions are strong, the global explanation
-may obfuscate heterogeneous effects~\citep{Herbinger2022repid} that
-exist under the hood, a phenomenon called aggregation
+individual explanations for producing a concise global one comes at a
+cost. Under strong feature interactions, the global explanation may
+obfuscate heterogeneous effects~\citep{Herbinger2022repid} that exist
+under the hood; a phenomenon called aggregation
 bias~\citep{mehrabi2021survey}.
 
-Feature effect forms a fundamental category of global explainability
-methods, isolating a single feature's average impact on the
-output. Feature effect methods suffer from aggregation bias because
-the rationale behind the average effect might be unclear. For example,
-a feature with zero average effect may indicate that the feature has
-no effect on the output or, contrarily, it has a highly positive
-effect in some cases and a highly negative one in others.
-
-There are two widely-used feature effect methods; Partial Dependence
-Plots (PDPlots)\citep{friedman2001greedy} and Aggregated Local Effects
-(ALE)\citep{apley2020visualizing}. PDPlots have been criticized for
-producing erroneous feature effect plots when the input features are
-correlated due to marginalizing over out-of-distribution synthetic
-instances. Therefore, ALE has been established as the state-of-the-art
-feature effect method since it can isolate feature effects in
-situations where input features are highly correlated.
+Feature effect\citep{Gromping2020MAEP} forms a fundamental category of
+global explainability methods, isolating a single feature's average
+impact on the output. Feature effect methods suffer from aggregation
+bias because the rationale behind the average effect might be
+unclear. For example, a feature with zero average effect may indicate
+that the feature has no effect on the output or, contrarily, it has a
+highly positive effect in some cases and a highly negative in
+others. There are three widely-used feature effect methods; Partial
+Dependence Plots (PDP)\citep{friedman2001greedy}, Marignal Plots
+(MP)\citep{apley2020visualizing} and Aggregated Local Effects
+(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
+computing erroneous effects when the input features are (highly)
+correlated, which is the case in most ML setups. Therefore, ALE has
+been established as the state-of-the-art feature effect method.
 
@@
 way to inform the user about potential heterogeneous effects that are
-hidden behind the average effect. In contrast, in the case of PDPlots,
-the heterogeneous effects can be spotted by exploring the Individual
+hidden behind the average effect. In contrast, in the case of PDP, the
+heterogeneous effects can be spotted by exploring the Individual
 Conditional Expectations (ICE)\citep{goldstein2015peeking}. Second,
@@
 interest is split in \(K\) fixed-size non-overlapping intervals, where
-\(K\) is a hyperparameter provided by the user. This splitting is done
-blindly, which can lead to inconsistent explanations.
+\(K\) is a hyperparameter provided by the user. So far, this splitting
+is done in a blind way, i.e., the user does not have an indication
+about the parameter K, which often leads to inconsistent explanations.
 
@@
 random and informs the user about the level of heterogeneous effects
-hidden behind the expected explanation. Our method completes ALE, as
-ICE plots complement PDPlots, for revealing the heterogeneous effects.
-
-Our method also automates the step of axis splitting into
-non-overlapping intervals. We, firstly, transform the bin splitting
-step into an unsupervised clustering problem and, second, find the
-optimal bin splitting for a robust estimation of (a) the global
-(expected) effect and (b) the uncertainty of the explanation from the
-limited samples of the training set. We formally prove that the
-objective of the clustering problem has as lower-bound the aggregated
-uncertainty of the global explanation. Our method works out of the box
-without requiring any input from the user.
+behind the expected explanation. The probabilistic extension completes
+ALE, as ICE plots complement PDP, for revealing the heterogeneous
+effects. Our method also transforms the axis-splitting step into an
+unsupervised clustering problem, i.e., we search for the optimal
+splitting given the instances of the training set. We, therefore,
+relieve the user from the obligation of defining a parameter \(K\) by
+providing an automated method for finding the optimal bin splitting to
+robustly estimate (a) the global (expected) effect and (b) the
+uncertainty of the explanation, from the limited samples of the
+training set.
 
@@
   of heterogeneous effects hidden behind the global explanation.
-\item
-  We provide an algorithm that automatically computes the optimal bin
-  splitting for robustly estimating the explanatory quantities, i.e.,
-  the global effect and the uncertainty.
+\item We provide an algorithm that automatically computes the optimal
+  axis-splitting for robustly estimating the the global effect and the
+  uncertainty.
 \item We formally prove that our method finds the optimal grouping of
@@
 
-\paragraph{Feature Effect Methods.} PDPlots formulate the feature
-effect of the \(s\)-th attribute as an expectation over the marginal
+\paragraph{Feature Effect Methods.} PDP formulates the feature effect
+of the \(s\)-th attribute as an expectation over the marginal
 distribution \(\mathbf{X}_c\), i.e.,
 \(f^{\mathtt{PDP}}(x_s) =
-\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPlots formulate it
-as an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
+\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPs formulate it as
+an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
 \(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
-\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as an
-integration of local effects. The local effects are measured as the
-expected change on the output
-\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \) over the
-conditional distribution \( \Xcb|X_s\). The formula that
-defines ALE is presented below:
+\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as the
+accumulation of the averaged local effects, which are defined as the
+change on the output, i.e.
+\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \):
 
@@
 %
-The constant \(c\) is used for centering the ALE plot. PDPlots
-integrate over unrealistic instances due to the use of the marginal
-distribution \( p(\mathcal{X}_1) \). Therefore, they incorrectly
-result in a quadratic effect in the region \(x_1 \in [0, 1]\). MPlots
-resolve this issue using the conditional distribution
-\( \mathcal{X}_2|\mathcal{X}_1 \) but suffer from computing combined
-effects. ALE plots resolve, both they have two drawbacks TODO add more
-info.
+The constant \(c\) is used for centering the ALE plot. PDP integrates
+over unrealistic instances due to the use of the marginal distribution
+\( \mathbf{X}_c\). MPs resolve this issue using the conditional
+distribution \( \math{C}_2|\mathcal{X}_1 \) but suffer from
+computing combined effects. ALE plots resolve, both they have two
+drawbacks TODO add more info.
 

### 2022-09-12 AISTATS2023/paper.tex

@@
 
-\paragraph{Feature Effect Methods.} PDP formulates the feature effect
-of the \(s\)-th attribute as an expectation over the marginal
-distribution \(\mathbf{X}_c\), i.e.,
+\paragraph{Feature Effect Methods.} As stated at the Introduction, the
+three well-known feature effect methods are: PDPs, MPs and ALE. PDPs
+formulate the feature effect of the \(s\)-th attribute as an
+expectation over the marginal distribution \(\mathbf{X}_c\), i.e.,
 \(f^{\mathtt{PDP}}(x_s) =
@@
   \label{eq:ALE_accumulated_mean}
-  f^{\mathtt{ALE}}(x_s) = c + \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
+  f^{\mathtt{ALE}}(x_s) = \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
 \end{equation}
 %
-The constant \(c\) is used for centering the ALE plot. PDP integrates
-over unrealistic instances due to the use of the marginal distribution
-\( \mathbf{X}_c\). MPs resolve this issue using the conditional
-distribution \( \math{C}_2|\mathcal{X}_1 \) but suffer from
-computing combined effects. ALE plots resolve, both they have two
-drawbacks TODO add more info.
+ALE has specific advantages which gain particular value in cases of
+corelated input features. In these cases, PDPs integrate over
+unrealistic instances, due to the use of the marginal distribution
+\( p(\mathbf{X}_c) \), and MPlots compute aggregated effects, i.e.,
+impute the combined effect of sets of features to a single
+feature. ALE manages to resolve both issues, and is therefore the only
+trustable method in cases of correlated features.
 
-TODO: Discuss the interpretation of ALE. The differences from PDP. Analyze both methods as processes.
+\paragraph{Quantify the Heterogeneous Effects.}
 
-\paragraph{Qunatify the Heterogeneous Effects.} It is crucial for
-feature effect methods to inform about the heterogeneous
-effects. Elaborate. Interpretation of the heterogeneous effects behind
-the global effect is available only for PDP, with three different
-approaches; (a) ICE and d-ICE plots provide a visual understanding of
-the heterogeneous effects. (b) grouping of ICE in homogeneous
-clusters, for spliting the input space into subspace(s) with
-homogeneous effects (c) Feature Interaction strength indexes, like
-H-statistic, provide a value indicating how much a feature interacts
-with the others (not the type of interaction). There is no method for
-quantifying the heterogeneous effects, based on ALE. Therefore, no
-method to exploit the advantages of ALE while, on the same time,
-informing about the heterogeneous effects.
+Feature effect methods answer the question \textit{what happens
+  (effect) to the output, if I increase/decrease the value of a
+  specific feature}. Having answered the question above, it comes
+naturally to also wonder \textit{how certain we are about the change
+  on the output}. For this reason, a lot of interest is given lately
+for quantifying the level of uncertainty, along with the expected
+effect. The level of uncertainty is mostly quantified by measuring the
+existence of heterogeneous effects, i.e. whether there are local
+explanations that deviate from the expected global effect. ICE and
+d-ICE plots provide a visual understanding of the heterogeneous
+effects on top of PDPs. Another approach targets on grouping the
+heterogeneous effects, e.g., allocating ICE plots in homogeneous
+clusters, by dividing the input space. Some other approaches, like
+H-Statistic, Greenwel, move a step behind and try to quantify the
+level of interaction between the input features, a possible cause of
+heterogeneous effects. In this case, the interpretation is indirect,
+since a strong interaction index is only an indicator of heterogeneous
+effects. The aforementioned approaches face two pathogenies; They
+either do not quantify the uncertainty of the feature effect directly
+or they are based on PDPs, and, therefore, they are subject to the
+failure modes of PDPs in cases of correlated features. To the best of
+our knowledge, no method so far targets on quantify the uncertainty of
+the feature effect as it is modelled by ALE.
 
-\paragraph{Bin Spliting.} ALE also has the peculiarity of spliting the
-axis into intervals, allocating the instances of the training set in
-the intervals and compute a single (constant) effect in each
-interval. With DALE extension, bin spliting is decoupled from instant
-effect estimation. With our extension for measuring the heterogeneous
-effects, we transfrom interval spliting from a step to a clustering
-problem with a meaningful objective to minimise. We provide a thorough
-analysis, where we show that our objective has a consistent
-meaning. It can be split in two parts; the first part is the
-unavoidable uncertainty due to the natural characteristics of the
-experiment, i.e.,~the data generating distribution and the black-box
-function. The second part is an added uncertainty due to the
-limited-samples estimation, that enforces to create groups with
-constant main effect. We opt for minimizing the objective, i.e. sum of
-the two uncertainties, that given that the first uncertainty is
-independent of the bin spliting, therefore we want to minimize the
-added uncertainty. To conclude, we transform the axis-spliting into an
-unsupervised clustering problem with a principled objective. We a
-computationally-grounded solution that works out-of-the-box, relaxing
-the user from providing a hyperparameter without any indication which
-one is the correct. This step can be used independently of whether the
-user wants to explore the heterogeneous effects or not.
+\paragraph{Cluster Instances with homogeneous effects.}
 
-\section{THE ... METHOD}
+In real ML scenarios, the expected feature effect and the uncertainty
+are estimated from the limited instances of the training set. ALE
+approximation requires an additional step, where the axis of the
+\(s\)-th feature is split into a sequence of non-overlaping bins and a
+single effect (expectation and uncertainty) is computed from the
+population of instances that lie inside each
+bin. \citep{apley2020visualizing} proposed estimating the local
+effects in each bin by evaluating the black box-funtion at the bin
+limits:
 
-\subsection{ALE with Uncertainty Quantification}
+\begin{equation}
+  \label{eq:ALE_accumulated_mean_est}
+  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
+    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
+\end{equation}
 
-ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at a single
-point\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
-explanations at \(x_s\) are weigthed by the conditional distribution
-\(p(\xc|x_s)\) and are averaged, to produce the summarized effect:
+We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
+i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
+the set of training instance that lie in the \(k\)-th bin, i.e.
+\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \} \). In
+contrast, (cite) proposed the Differential ALE (DALE) estimation for
+quantifying the local effects on the training-set instances, instead
+of the bin limits:
+
+\begin{equation}
+  \label{eq:DALE_accumulated_mean_est}
+  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
+    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
+\end{equation}
+%
+Their method has the advantages of remaining on-distribution even when
+bins become wider and, most importantly, it allows the recomputation
+of the accumulated effect with different bin-splitting in near-zero
+computational overhead. However, none of the approximations above
+deals with the crucial problem of the optimal bin-spliting. They
+partition the axis in \(K\) equally-sized bins without considering the
+properties of the underlying data, which can lead to erroneous
+approximations.
+
+Instead, we propose treating the axis-spliting step as an unsupervised
+clustering problem. The objective of the clustering problem should
+fulfill in the best way to contradictory ojectives. First, secure
+robust estimations of the expected effect and the uncertainty inside
+each bin given the limited instances of the training set and, second,
+create bins with as homogeneous local effects as possible, for not
+losing fine-grain resolution feature effects due to wide bins.
+
+\section{THE NAME METHOD}
+\label{sec:NAME-method}
+
+The NAME method extends the traditional ALE method with a component
+for uncertainty quantification and improves the ALE estimation by
+automatically discovering the optimal bin-splitting. In
+Section~\ref{sec:NAME-definition} we define the component for the
+uncertainty quantification. In
+Section~\ref{sec:interval-based-estimation}, we show how to estimate
+NAME from the limited samples of the training set and we make an
+important proof about the aggregated variance defined over a bin. In
+Section~\ref{sec:interval-spliting}, we define and solve the problem
+of optimal bin-splitting. Finally, in Section~\ref{sec:visualization},
+we illustrate the appropriate visualization of NAME for facilitating
+its interpretation by a non-expert and we discuss important aspects of
+the method
+
+\subsection{Uncertainty Quantification}
+\label{sec:NAME-definition}
+
+ALE defines the local effect of the \(s\)-th feature on \(f(\cdot)\)
+at point \((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
+explanations at \(x_s\) are, then, weigthed by the conditional
+distribution \(p(\xc|x_s)\) and are averaged, to produce the
+summarized effect at \(x_s\):
 
@@
 \end{equation}
-\noindent
-As described at the Introduction, limiting the explanation at this
-level does not shed light to possible heterogeneous effects behind the
-averaged explanation. Therefore, we model the uncertainty of the local
-effects at \(\mathcal{H}(x_s)\) as the variance of the local
-explanations:
+%
+The feature effect at \(x_s\) is the accumulation of the averaged
+local effects from \(x_{s, min}\) until \(x_s\), i.e.
+\(f^{\mathtt{ALE}}(x_s) = \int_{x_{s, min}}^{x_s} \mu(z) \partial z
+\). As described at the Introduction, limiting the explanation to the
+exepected value level does not shed light to possible heterogeneous
+effects behind the averaged explanation. Therefore, we model the
+uncertainty of the local effects at \(\mathcal{H}(x_s)\) as the
+variance of the local explanations:
 
@@
 characteristics of the experiment, i.e.,~the data generating
-distribution and the properties of the black-box function. In Section
-(TODO), we propose appropriate visualizations for easier
-interpretation of Eq.~(\ref{eq:ALE_var}). In ALE, the feature effect
-at \(x_s\) is the accumulation of the averaged local effects from
-\(x_{min}\) until \(x_s\), as show in
+distribution and the properties of the black-box function. In
+Section~\ref{sec:visualization}, we propose appropriate visualizations
+for easier interpretation of Eq.~(\ref{eq:ALE_var}). In ALE, the
+feature effect at \(x_s\) is the accumulation of the averaged local
+effects from \(x_{min}\) until \(x_s\), as shown in
 Eq.~(\ref{eq:ALE_accumulated_mean}). Equivalently, we define the
-accumulated uncertainty (variance) until the point \(x_s\), as the
-integral of the variances of local effects:
+accumulated uncertainty (variance) until \(x_s\), as the integral of
+the variances of the local effects:
 
@@
 quantity. It only helps us define a sensible objective for the
-interval spliting step, as we discuss in Section TODO: adde ref.
+interval spliting step, as we discuss in
+Section~\ref{sec:interval-spliting}.
 
-\subsection{Uncertainty Quantification and Estimation at an Interval}
+\subsection{Interval-Based Estimation}
+\label{sec:interval-based-estimation}
 
 In real scenarios, we have ignorance about the data-generating
-distribution \(p(x_s, \mathbf{x}_c)\) and all estimations are based on
-the limited instances of the training set. Estimating
+distribution \(p(x_s, \mathbf{x}_c)\), so, the estimations are based
+on the limited instances of the training set. Estimating
 Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
-point \(x_s\) is impossible, because the probability of observing a
-sample inside the interval \([x_s - h, x_s + h]\) tends to zero, when
-\(h \to 0\). Therefore, we are obliged to split the axis of \(x_s\)
+point is impossible, because the probability of observing a sample
+inside the interval \([x_s - h, x_s + h]\) tends to zero, when
+\(h \to 0\). We are, therefore, obliged to split the axis of \(x_s\)
 into a sequence of non-overlaping intervals (bins) and estimate the
@@
 \noindent
-Accordingly, the accumulated variance at an interval \([z_1, z_2)\)
-is defined as:
+Similarly, the accumulated variance at an interval \([z_1, z_2)\) is
+defined as:
 
@@
       f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z
+\end{equation}
+\paragraph{Theorem 1.} If we define the residual \(\rho(z)\) as the
+difference between the expected effect at \(x_s\) and the mean
+expected effect at the interval, i.e
+\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
+at an interval \([z_1, z_2)\) is the accumulation of the all variances
+plus the accumulation of squared residuals inside the interval:
+
+\begin{equation}
+ \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
+\end{equation}
+%
+The proof is in the Appendix. Theorem 1 decouples the accumulated
+variance at an interval into two terms. The first term
+\(\int_{z_1}^{z_2} \sigma^2(z) \partial z\), quantifies the aggregated
+uncertainty due to the natural characteristics of the experiment
+\(\mathcal{H}\) and the second term adds extra nuisance uncertainty
+due to the limited resolution \(\mathcal{H}_n\). In other words,
+enforcing the computation of a single effect for all points in
+\([z_1, z_2)]\), burdens the estimation inside the interval with a
+nuisance uncertainty of:
+
+\begin{equation}
+  \label{eq:uncertainty_bin}
+  \mathcal{H}_{bin}(z_1, z_2) = \int_{z_1}^{z_2} \mathcal{H}(z) + \mathcal{H}_n(z) \partial z
 \end{equation}
@@
 The approximation is unbiased only if the points are uniformly
-distributed in \([z_1, z_2]\). (TODOs: Check what happens
-otherwise).
+distributed in \([z_1, z_2]\). Elaborate.
 
-\subsection{Bin Spliting as a Clustering Problem}
+\subsection{Interval Spliting As A Clustering Problem}
+\label{sec:interval-spliting}
 
-ALE, Eq.(\ref{eq:ALE_accumulated_mean}), is estimated by spliting the
-axis \(x_s\) into a sequence of non-overlaping bins (TODO: add some
-discussion for ALE, DALE):
+We formulate the axis-splitting as an unsupervised clustering
+problem. We search for the optimal bin splitting, i.e., the number and
+size of consecutive non-overlapping intervals that minimizes the
+accumulated variance. The optimization problem is defined as follows:
 
 \begin{equation}
-  \label{eq:ALE_accumulated_mean_est}
-  \hat{f}^{\mathtt{ALE}}_{\mu}(x_s) = \Delta x \sum_{k=1}^{k_x} \hat{\mu}(z_{k-1}, z_k)
+  \label{eq:opt}
+\begin{aligned}
+\min_{ \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \\
+\textrm{s.t.} \quad & |\mathcal{S}_k| \geq N\\
+\end{aligned}
 \end{equation}
 
-and
+The objective of the optimization problem is the sum of the
+accumulated variances in each bin, estimated by the instances of the
+training set. As we prove in Theorem 1,
+\(\mathcal{L} = \sum_{k=1}^K \mathcal{H}_{bin}(z_{k-1}, z_k)\) is the
+sum of the uncertainty of the local explanations \(\mathcal{H}\) due
+to heterogeneous effects and the added nuisance uncertainty due to
+bin-splitting \(\mathcal{H}_n\). Given that \(\mathcal{H}\) is not
+affected by interval splitting, optimizing \(\mathcal{L}\) equals to
+finding the sequence of non-overlaping bins that add the minimum
+nuisance uncertainty. The restriction of (\ref{eq:opt}) secures that
+each bin is populated with enough samples for a robust estimation,
+which is set by the parameter \(N\). In other words, the user makes
+the following proposal; \textit{find the bin sequence that adds the
+  minimum nuisance uncertainty, given that I want at least \(N\)
+  points per bin}.
 
-\begin{equation}
-  \label{eq:ALE_accumulated_var_est}
-  \hat{f}^{\mathtt{ALE}}_{\sigma^2}(x_s) = \sum_{k=1}^{k_x} \hat{\sigma}^2(z_{k-1}, z_k)
-\end{equation}
+\subsubsection{Algorithm For Solving The Clustering Problem}
+\label{sec:dynamic-programing}
 
-We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
-i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
-the set of training instance that lie in the \(k\)-th bin, i.e.
-\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \} \). Both
-methods face the limitation that the partitioning into non-overlaping
-intervals is done blindly. The user pass the total number of bins
-\(K\) as a hyperparameter, the bins are defined with equal-size
-spliting, and the training instances are allocated accordingly. This
-approach is vulnerable to non-robust estimations. The mean effect is
-often poorly approximated from a very small number of samples and the
-mean effect of empty bins is interpolated from their
-neighboors. Furthermore, in our case, we need sufficient sample
-populations for estimating the variance of the approximation, apart
-from the mean effect.
-
-\subsubsection{Methodology}
-
-For overcoming this limitations, we reformulate the partitioning as a
-clustering of the training instances into a sequence variable-size
-intervals. The objective of the clustering problem is inspired
-
-ALE requires splittingthe estimation of the
-
-In this section, we introduce a framework
-
-\paragraph{Theorem 1.} If we define the residual \(\rho(z)\) as the
-difference between the expected effect at \(x_s\) and the mean
-expected effect at the interval, i.e
-\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
-at an interval \([z_1, z_2)\) is the accumulation of the all variances
-plus the accumulation of squared residuals inside the interval:
-
-\begin{equation}
- \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
-\end{equation}
-%
-The proof is at the Appendix. Theorem 1 decouples the accumulated
-variance at an interval, the only quantity we can estimate, into two
-terms. The first term \(\int_{z_1}^{z_2} \sigma^2(z) \partial z\),
-quantifies the uncertainty due to the natural characteristics of the
-experiment and the second term adds extra uncertainty due to the
-limited resolution.
-
-\paragraph{Uncertainty of the global effect.}
-
-Eq.~\eqref{eq:variance_estimation} gives an approximation of the
-uncertainty of the bin effect.
-The uncertainty of the global effect is
-simply the sum of the uncertainties in the bin effects.
-
-\paragraph{Minimizing the uncertainty}
-
-Solving the problem of finding (a) the optimal number of bins \(K\) and (b) the optimal bin limits for each bin \([z_{k-1}, z_k] \forall k\) to minimize:
+Solving the problem of finding (a) the optimal number of bins \(K\)
+and (b) the optimal bin limits for each bin
+\([z_{k-1}, z_k] \forall k\) to minimize:
 
@@
 
-\subsection{Visualization of ALE with Uncertainty}
+\subsection{Visualization of NAME Method and Discussion}
+\label{sec:visualization}
+
+\begin{itemize}
+\item Discuss about the meaning of ALE, to find intervals with some effect
+\end{itemize}
 
@@
 
-\section*{Appendix}
-
-\subsection{Proof for variance of the bin}
-
 \end{document}

### 2022-09-12 AISTATS2023/paper.tex
updated method description

@@
 Recently, ML has flourished in critical domains, such as healthcare
-and finance. In these areas, we need ML models that not only predict
-accurately but also explain their predictions. Therefore, Explainable
-AI (XAI), the field that provides interpretations for the prediction
-mechanism of complex black-box models, has increased interest. XAI
-literature distinguishes between local and global interpretation
-methods~\citep{Molnar2020interpretable}. Local methods explain a
+and finance. In these areas, we need ML system with the capability to
+explain their predictions, apart from predicting with accuracy. For
+this reason there is an increased interest in Explainable AI (XAI),
+the field that provides interpretations about the behavior of complex
+black-box models. XAI literature distinguishes between local and
+global explainability
+techniques~\citep{Molnar2020interpretable}. Local methods explain a
 specific prediction, whereas global methods explain the entire model
 behavior. Global methods provide a universal explanation, summarizing
-the numerous local explanations into a single interpretable outcome,
-usually a number or a plot. If a user wants to know which features are
-significant (feature importance) or whether a particular feature has a
-positive or negative effect on the output (feature effect), they
-should opt for a global explainability technique. Aggregating the
-individual explanations for producing a concise global one comes at a
-cost. Under strong feature interactions, the global explanation may
-obfuscate heterogeneous effects~\citep{Herbinger2022repid} that exist
-under the hood; a phenomenon called aggregation
-bias~\citep{mehrabi2021survey}.
+the numerous local ones into a single interpretable outcome, usually a
+number or a plot. If a user wants to get a rough overview about which
+features are significant (feature importance) or whether a particular
+feature has a positive or negative effect on the output (feature
+effect), they should opt for a global explainability technique. On
+ther other hand, aggregating the individual explanations for producing
+a concise global one is vulnerable to misinterpretations; Under strong
+feature interactions, the global explanation may obfuscate
+heterogeneous effects~\citep{Herbinger2022repid} that exist under the
+hood; a phenomenon called aggregation bias~\citep{mehrabi2021survey}.
 
-Feature effect\citep{Gromping2020MAEP} forms a fundamental category of
-global explainability methods, isolating a single feature's average
-impact on the output. Feature effect methods suffer from aggregation
-bias because the rationale behind the average effect might be
-unclear. For example, a feature with zero average effect may indicate
-that the feature has no effect on the output or, contrarily, it has a
-highly positive effect in some cases and a highly negative in
-others. There are three widely-used feature effect methods; Partial
-Dependence Plots (PDP)\citep{friedman2001greedy}, Marignal Plots
+Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category
+of global explainability methods. The objective of FE is to the
+isolate and visulalize the impact of a single feature on the
+output.~\footnote{FE methods also isolate the effect of a pair of
+  features to the output. Combinations of more than two features are
+  not usual, because they encounter, among others, visualization
+  difficulties.} FE methods suffer from aggregation bias because,
+often, the rationale behind the average effect might be unclear. For
+example, a feature with zero average effect may indicate that the
+feature has no effect on the output or, contrarily, it has a highly
+positive effect in some cases and a highly negative in others. There
+are three widely-used FE methods; Partial Dependence Plots
+(PDP)\citep{friedman2001greedy}, Marignal Plots
 (MP)\citep{apley2020visualizing} and Aggregated Local Effects
@@
 computing erroneous effects when the input features are (highly)
-correlated, which is the case in most ML setups. Therefore, ALE has
-been established as the state-of-the-art feature effect method.
+correlated, which is a frequent scenario in many ML
+problems. Therefore, ALE has been established as the state-of-the-art
+FE method.
 
@@
 Conditional Expectations (ICE)\citep{goldstein2015peeking}. Second,
-ALE requires an additional step, where the axis of the feature of
-interest is split in \(K\) fixed-size non-overlapping intervals, where
-\(K\) is a hyperparameter provided by the user. So far, this splitting
-is done in a blind way, i.e., the user does not have an indication
-about the parameter K, which often leads to inconsistent explanations.
+ALE requires an additional step where the axis of the feature of
+interest is split in \(K\) fixed-size non-overlapping intervals. So
+far, the user is asked to provide a value for the parameter \(K\)
+blindly, i.e without an indication about whether a big or a small
+value could be more appropriate, which often leads to unstable
+explanations. For convenience, in the rest of the paper we will refer
+to this step using the term bin-splitting problem.
 
@@
 of the global explanation expresses how certain we are that the global
-(expected) explanation is valid if applied to an instance drawn at
-random and informs the user about the level of heterogeneous effects
-behind the expected explanation. The probabilistic extension completes
-ALE, as ICE plots complement PDP, for revealing the heterogeneous
-effects. Our method also transforms the axis-splitting step into an
-unsupervised clustering problem, i.e., we search for the optimal
-splitting given the instances of the training set. We, therefore,
-relieve the user from the obligation of defining a parameter \(K\) by
-providing an automated method for finding the optimal bin splitting to
-robustly estimate (a) the global (expected) effect and (b) the
-uncertainty of the explanation, from the limited samples of the
-training set.
+(averaged) explanation is valid if applied to an instance drawn at
+random. The probabilistic extension completes ALE, as ICE plots
+complement PDP, for revealing the heterogeneous effects. Furthermore,
+we transform the bin-splitting step into a data-driven clustering
+problem, i.e., we search for the optimal splitting given available
+instances of the training set. We, also, present a
+computationally-grounded algorithm for finding the optimal
+solution.
 
@@
 \begin{itemize}
-\item We introduce Uncertainty DALE (UDALE), an extension of DALE that
-  quantifies the uncertainty of the global explanation, i.e.~the level
-  of heterogeneous effects hidden behind the global explanation.
-\item We provide an algorithm that automatically computes the optimal
-  axis-splitting for robustly estimating the the global effect and the
-  uncertainty.
+\item We introduce NAME, an extension of ALE that quantifies the
+  uncertainty of the global explanation, i.e.~the level of
+  heterogeneous effects hidden behind the global explanation.
+\item We present an algorithm that automatically computes the optimal
+  bin-splitting
 \item We formally prove that our method finds the optimal grouping of
   samples, minimizing the added uncertainty over the unavoidable
-  heterogeneity that is the lower-bound of the objective.
+  heterogeneity.
 \item We provide empirical evaluation of the method in artificial and
@@
 function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
-feature effect of the \(s\)-th feature is
+FE of the \(s\)-th feature is
 \(f^{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
-name of the feature effect method.\footnote{An extensive list of all
+name of the FE method.\footnote{An extensive list of all
   symbols used in the paper is provided in the helping material.}
 
-\paragraph{Feature Effect Methods.} As stated at the Introduction, the
-three well-known feature effect methods are: PDPs, MPs and ALE. PDPs
-formulate the feature effect of the \(s\)-th attribute as an
-expectation over the marginal distribution \(\mathbf{X}_c\), i.e.,
+\paragraph{Feature Effect Methods.} The three well-known feature
+effect methods are: PDP, MP and ALE. PDP formulates the FE
+of the \(s\)-th attribute as an expectation over the marginal
+distribution \(\mathbf{X}_c\), i.e.,
 \(f^{\mathtt{PDP}}(x_s) =
-\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPs formulate it as
-an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
+\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
+formulates it as an expectation over the conditional
+\(\mathbf{X}_c|X_s\), i.e.,
 \(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
 \mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as the
-accumulation of the averaged local effects, which are defined as the
-change on the output, i.e.
-\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \):
+accumulation of the averaged local effects:
 
@@
 ALE has specific advantages which gain particular value in cases of
-corelated input features. In these cases, PDPs integrate over
+corelated input features. In these cases, PDP integrates over
 unrealistic instances, due to the use of the marginal distribution
-\( p(\mathbf{X}_c) \), and MPlots compute aggregated effects, i.e.,
-impute the combined effect of sets of features to a single
-feature. ALE manages to resolve both issues, and is therefore the only
-trustable method in cases of correlated features.
+\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
+the combined effect of sets of features to a single feature. ALE
+manages to resolve both issues, and is therefore the only trustable
+FE method in cases of correlated features.
 
@@
 
-Feature effect methods answer the question \textit{what happens
-  (effect) to the output, if I increase/decrease the value of a
-  specific feature}. Having answered the question above, it comes
-naturally to also wonder \textit{how certain we are about the change
-  on the output}. For this reason, a lot of interest is given lately
-for quantifying the level of uncertainty, along with the expected
-effect. The level of uncertainty is mostly quantified by measuring the
-existence of heterogeneous effects, i.e. whether there are local
-explanations that deviate from the expected global effect. ICE and
-d-ICE plots provide a visual understanding of the heterogeneous
-effects on top of PDPs. Another approach targets on grouping the
-heterogeneous effects, e.g., allocating ICE plots in homogeneous
-clusters, by dividing the input space. Some other approaches, like
-H-Statistic, Greenwel, move a step behind and try to quantify the
-level of interaction between the input features, a possible cause of
+FE methods answer the question \textit{what is expected to happen to
+  the output (expected effect), if the value of a specific feature is
+  increased/decreased}. Having an answer to the question above, it
+comes naturally to also wonder \textit{how certain we are about the
+  expected change (uncertainty)}. For this reason, the quantification
+of the uncertainty along with the expected effect has attracted a lot
+of interest. The level of uncertainty is mostly quantified by
+measuring the existence of heterogeneous effects, i.e. whether there
+are local explanations that deviate from the expected global
+effect. ICE and d-ICE plots provide a visual understanding of the
+heterogeneous effects on top of PDPs. Another approach targets on
+grouping the heterogeneous effects, e.g., allocating ICE plots in
+homogeneous clusters, by dividing the input
+space.\citep{molnar2020model} Some other approaches, like H-Statistic,
+Greenwel, move a step behind and try to quantify the level of
+interaction between the input features, a possible cause of
 heterogeneous effects. In this case, the interpretation is indirect,
@@
 effects. The aforementioned approaches face two pathogenies; They
-either do not quantify the uncertainty of the feature effect directly
-or they are based on PDPs, and, therefore, they are subject to the
-failure modes of PDPs in cases of correlated features. To the best of
-our knowledge, no method so far targets on quantify the uncertainty of
-the feature effect as it is modelled by ALE.
+either do not quantify the uncertainty of the FE directly or they are
+based on PDPs, and, therefore, they are subject to the failure modes
+of PDPs in cases of correlated features. To the best of our knowledge,
+no method so far targets on quantifying the uncertainty of ALE.
 
-\paragraph{Cluster Instances with homogeneous effects.}
+\paragraph{Bin-Splitting for ALE estimation.}
 
-In real ML scenarios, the expected feature effect and the uncertainty
-are estimated from the limited instances of the training set. ALE
-approximation requires an additional step, where the axis of the
-\(s\)-th feature is split into a sequence of non-overlaping bins and a
-single effect (expectation and uncertainty) is computed from the
-population of instances that lie inside each
-bin. \citep{apley2020visualizing} proposed estimating the local
-effects in each bin by evaluating the black box-funtion at the bin
-limits:
-
+In real ML scenarios, the expected FE and the uncertainty are
+estimated from the limited instances of the training set.
+\citep{apley2020visualizing} proposed estimating the local effects in
+each bin by evaluating the black box-funtion at the bin limits:
 \begin{equation}
@@
 \end{equation}
-
 We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
@@
 the set of training instance that lie in the \(k\)-th bin, i.e.
-\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \} \). In
-contrast, (cite) proposed the Differential ALE (DALE) estimation for
-quantifying the local effects on the training-set instances, instead
-of the bin limits:
-
+\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
+\). Afterwards, (cite) proposed the Differential ALE (DALE) that
+exploits differentiaton for computing the local effects on the instances
+training-set, instead of the bin limits:
 \begin{equation}
@@
 Their method has the advantages of remaining on-distribution even when
-bins become wider and, most importantly, it allows the recomputation
-of the accumulated effect with different bin-splitting in near-zero
-computational overhead. However, none of the approximations above
-deals with the crucial problem of the optimal bin-spliting. They
-partition the axis in \(K\) equally-sized bins without considering the
-properties of the underlying data, which can lead to erroneous
-approximations.
-
-Instead, we propose treating the axis-spliting step as an unsupervised
-clustering problem. The objective of the clustering problem should
-fulfill in the best way to contradictory ojectives. First, secure
-robust estimations of the expected effect and the uncertainty inside
-each bin given the limited instances of the training set and, second,
-create bins with as homogeneous local effects as possible, for not
-losing fine-grain resolution feature effects due to wide bins.
+bins become wider and, most importantly, allows the recomputation of
+the accumulated effect with different bin-splitting with near-zero
+computational cost. However, none of the approximations above deals
+with the crucial problem of bin-spliting. As indicated
+by~\citep{molnar2022}, in ALE the effects are computed per interval
+(region) and the interpretation of the effect can only be
+local.
 
@@
 
-The NAME method extends the traditional ALE method with a component
-for uncertainty quantification and improves the ALE estimation by
-automatically discovering the optimal bin-splitting. In
-Section~\ref{sec:NAME-definition} we define the component for the
+In Section~\ref{sec:NAME-definition} we define the component for the
 uncertainty quantification. In
 Section~\ref{sec:interval-based-estimation}, we show how to estimate
-NAME from the limited samples of the training set and we make an
-important proof about the aggregated variance defined over a bin. In
-Section~\ref{sec:interval-spliting}, we define and solve the problem
-of optimal bin-splitting. Finally, in Section~\ref{sec:visualization},
-we illustrate the appropriate visualization of NAME for facilitating
-its interpretation by a non-expert and we discuss important aspects of
-the method
+the average effect and the uncertainty from the limited samples of the
+training set and we make an important proof about the aggregated
+variance defined over a bin. In Section~\ref{sec:interval-spliting},
+we define and solve the problem of optimal bin-splitting. Finally, in
+Section~\ref{sec:visualization}, we illustrate the appropriate
+visualization of NAME for facilitating its interpretation by a
+non-expert and we discuss important aspects of the method
 
@@
 %
-The feature effect at \(x_s\) is the accumulation of the averaged
+The FE at \(x_s\) is the accumulation of the averaged
 local effects from \(x_{s, min}\) until \(x_s\), i.e.
@@
 for easier interpretation of Eq.~(\ref{eq:ALE_var}). In ALE, the
-feature effect at \(x_s\) is the accumulation of the averaged local
+FE at \(x_s\) is the accumulation of the averaged local
 effects from \(x_{min}\) until \(x_s\), as shown in
@@
 
+% Instead, we propose treating the axis-spliting step as an unsupervised
+% clustering problem. The objective of the clustering problem should
+% fulfill in the best way to contradictory ojectives. First, secure
+% robust estimations of the expected effect and the uncertainty inside
+% each bin given the limited instances of the training set and, second,
+% create bins with as homogeneous local effects as possible, for not
+% losing fine-grain resolution FEs due to wide bins.
+
 We formulate the axis-splitting as an unsupervised clustering
@@
 \noindent
-TODOS. Show theoretically that \(\mathcal{L} \geq \int_{x_{s, \min}}^{x_{s, \max}}\sigma^2(x_s) \partial x_s\)
+
+\begin{itemize}
+\item Computational complexity
+\end{itemize}
 
@@
 Therefore, there is an implicit trade-off behind the formation of bins.
-Each bin must include enough instances for a robust estimation of the bin feature effect (expected value), and the uncertainty of the explanation (variance).
+Each bin must include enough instances for a robust estimation of the bin FE (expected value), and the uncertainty of the explanation (variance).
 On the other hand, each bin should include points with similar local effects.

### 2022-09-13 AISTATS2023/paper.tex
reached v1 of method description

@@
 effect), they should opt for a global explainability technique. On
-ther other hand, aggregating the individual explanations for producing
+the other hand, aggregating the individual explanations for producing
 a concise global one is vulnerable to misinterpretations; Under strong
@@
 of global explainability methods. The objective of FE is to the
-isolate and visulalize the impact of a single feature on the
+isolate and visualize the impact of a single feature on the
 output.~\footnote{FE methods also isolate the effect of a pair of
@@
 
-However, ALE faces two crucial drawbacks. First, it does not provide a
-way to inform the user about potential heterogeneous effects that are
-hidden behind the average effect. In contrast, in the case of PDP, the
-heterogeneous effects can be spotted by exploring the Individual
-Conditional Expectations (ICE)\citep{goldstein2015peeking}. Second,
-ALE requires an additional step where the axis of the feature of
-interest is split in \(K\) fixed-size non-overlapping intervals. So
-far, the user is asked to provide a value for the parameter \(K\)
-blindly, i.e without an indication about whether a big or a small
-value could be more appropriate, which often leads to unstable
-explanations. For convenience, in the rest of the paper we will refer
-to this step using the term bin-splitting problem.
+However, ALE faces two crucial drawbacks. First, it does not inform
+the user about potential heterogeneous effects hidden behind the
+average effect. In contrast, in the case of PDP, the heterogeneous
+effects can be spotted by exploring the Individual Conditional
+Expectations (ICE)\citep{goldstein2015peeking}. Second, ALE
+approximation from the samples of the training set requires
+partitioning the axis of the feature of interest in a sequence of
+non-overlapping intervals and estimating a single effect from the
+population of samples in each interval. We refer to this step using
+the term \textit{bin-splitting problem}. Deciding on the sequence of
+intervals has important implications because ALE's interpretation in
+cases of correlated features is meaningful only inside each interval
+\citep{molnar2022}. However, this crucial step has not been given the
+appropriate attention, making the approximation vulnerable to
+potential misinterpretations.
 
 In this paper, we extend ALE with a probabilistic component for
-measuring the uncertainty of the global explanation. The uncertainty
-of the global explanation expresses how certain we are that the global
-(averaged) explanation is valid if applied to an instance drawn at
-random. The probabilistic extension completes ALE, as ICE plots
-complement PDP, for revealing the heterogeneous effects. Furthermore,
-we transform the bin-splitting step into a data-driven clustering
-problem, i.e., we search for the optimal splitting given available
-instances of the training set. We, also, present a
-computationally-grounded algorithm for finding the optimal
-solution.
+measuring the uncertainty of the explanation. The uncertainty of the
+global explanation expresses how certain we are that the averaged
+explanation is valid if applied to an instance drawn at random and
+reveals the heterogeneous effects. Given that ALE's interpretation
+(expected effect and uncertainty) is meaningful only inside each
+interval, we design a principled framework, where we treat the
+bin-splitting step as a data-driven clustering problem, searching for
+the optimal splitting given available instances of the training
+set. We, also, present a computationally-grounded algorithm for
+finding the optimal solution.
 
-\paragraph{Contributions.} The contributions of this paper are the following:
+\paragraph{Contributions.} The contribution of this paper is NAME, a
+feature effect method that:
 
 \begin{itemize}
-\item We introduce NAME, an extension of ALE that quantifies the
-  uncertainty of the global explanation, i.e.~the level of
-  heterogeneous effects hidden behind the global explanation.
-\item We present an algorithm that automatically computes the optimal
-  bin-splitting
-\item We formally prove that our method finds the optimal grouping of
-  samples, minimizing the added uncertainty over the unavoidable
-  heterogeneity.
-\item We provide empirical evaluation of the method in artificial and
-  real datasets.
+\item Extends ALE with a probabilistic component to quantify the
+  heterogeneous effects behind the global explanation.
+\item Automatically extracts regions with similar effects, improving
+  the estimation and the interpretability of ALE plots.
 \end{itemize}
 
-The implementation of our method and the code for reproducing all the
-experiments is provided in the submission and will become publicly
-available upon acceptance.
+We provide empirical evaluation of the method in artificial and real
+datasets. The implementation of our method and the code for
+reproducing all the experiments is provided in the submission and will
+become publicly available upon acceptance.
 
@@
 accumulation of the averaged local effects:
-
 \begin{equation}
@@
 ALE has specific advantages which gain particular value in cases of
-corelated input features. In these cases, PDP integrates over
+correlated input features. In these cases, PDP integrates over
 unrealistic instances, due to the use of the marginal distribution
@@
 the combined effect of sets of features to a single feature. ALE
-manages to resolve both issues, and is therefore the only trustable
-FE method in cases of correlated features.
+manages to resolve both issues, and is therefore a trust able FE method
+in cases of correlated features.
 
@@
 since a strong interaction index is only an indicator of heterogeneous
-effects. The aforementioned approaches face two pathogenies; They
+effects. The aforementioned approaches are under two limitations; They
 either do not quantify the uncertainty of the FE directly or they are
@@
 
-In real ML scenarios, the expected FE and the uncertainty are
-estimated from the limited instances of the training set.
-\citep{apley2020visualizing} proposed estimating the local effects in
-each bin by evaluating the black box-funtion at the bin limits:
+In real ML scenarios, the FE is estimated from the limited instances
+of the training set.  \citep{apley2020visualizing} proposed dividing
+the \(s\)-th axis in \(K\) bins and estimating the local effects in
+each bin by evaluating the black box-function at the bin limits:
 \begin{equation}
@@
 \( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
-\). Afterwards, (cite) proposed the Differential ALE (DALE) that
-exploits differentiaton for computing the local effects on the instances
-training-set, instead of the bin limits:
+\). Afterwards, (cite) proposed the Differential ALE (DALE) for
+computing the local effects on the training instances using auto-differentiation:
 \begin{equation}
@@
 computational cost. However, none of the approximations above deals
-with the crucial problem of bin-spliting. As indicated
+with the crucial problem of bin-splitting. As indicated
 by~\citep{molnar2022}, in ALE the effects are computed per interval
@@
 
-In Section~\ref{sec:NAME-definition} we define the component for the
-uncertainty quantification. In
+The NAME method extends ALE for measuring the uncertainty of the
+expected effect and automatically extracts regions with homogeneous
+effects. In Section~\ref{sec:NAME-definition} we define the component
+for the uncertainty quantification. In
 Section~\ref{sec:interval-based-estimation}, we show how to estimate
-the average effect and the uncertainty from the limited samples of the
-training set and we make an important proof about the aggregated
-variance defined over a bin. In Section~\ref{sec:interval-spliting},
-we define and solve the problem of optimal bin-splitting. Finally, in
-Section~\ref{sec:visualization}, we illustrate the appropriate
-visualization of NAME for facilitating its interpretation by a
-non-expert and we discuss important aspects of the method
+the average effect and the uncertainty on a specific interval from the
+limited samples of the training set. We, also, make an important proof
+about the aggregated variance defined over a bin. In
+Section~\ref{sec:interval-spliting}, we define and solve the problem
+of optimal bin-splitting for automatically extract regions with
+homogeneous effects. Finally, in Section~\ref{sec:visualization}, we
+illustrate the appropriate visualization of NAME for facilitating its
+interpretation by a non-expert and we discuss important aspects of the
+method.
 
@@
 
-ALE defines the local effect of the \(s\)-th feature on \(f(\cdot)\)
-at point \((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
-explanations at \(x_s\) are, then, weigthed by the conditional
-distribution \(p(\xc|x_s)\) and are averaged, to produce the
-summarized effect at \(x_s\):
-
+ALE defines the local effect of the \(s\)-th feature at point
+\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local explanations at
+\(x_s\) are, then, weighted by the conditional distribution
+\(p(\xc|x_s)\) and are averaged, to produce the summarized effect at
+\(x_s\):
 \begin{equation}
@@
 %
-The FE at \(x_s\) is the accumulation of the averaged
-local effects from \(x_{s, min}\) until \(x_s\), i.e.
+The FE at \(x_s\) is the accumulation of the averaged local effects
+from \(x_{s, min}\)\footnote{\(x_{s,min}\) is the minimum value the
+  \(s\)-th attribute can take. In estimations, it is the minimum
+  \(s\)-th attribute value among the training set instances.} until
+\(x_s\), i.e.
 \(f^{\mathtt{ALE}}(x_s) = \int_{x_{s, min}}^{x_s} \mu(z) \partial z
-\). As described at the Introduction, limiting the explanation to the
-exepected value level does not shed light to possible heterogeneous
-effects behind the averaged explanation. Therefore, we model the
-uncertainty of the local effects at \(\mathcal{H}(x_s)\) as the
-variance of the local explanations:
-
+\). Limiting the explanation to the expected value level does not
+shed light to potential heterogeneous effects behind the averaged
+explanation. Therefore, we model the uncertainty of the local effects
+\(\mathcal{H}(x_s)\) with the variance of the local explanations:
 \begin{equation}
@@
 Section~\ref{sec:visualization}, we propose appropriate visualizations
-for easier interpretation of Eq.~(\ref{eq:ALE_var}). In ALE, the
-FE at \(x_s\) is the accumulation of the averaged local
-effects from \(x_{min}\) until \(x_s\), as shown in
-Eq.~(\ref{eq:ALE_accumulated_mean}). Equivalently, we define the
-accumulated uncertainty (variance) until \(x_s\), as the integral of
-the variances of the local effects:
-
+for easier interpretation of Eq.~(\ref{eq:ALE_var}). ALE defines the
+FE at \(x_s\) as the accumulation of the averaged local effects from
+\(x_{min}\) until \(x_s\). Equivalently, we define the accumulated
+uncertainty (variance) until \(x_s\), as the integral of the variances
+of the local effects:
 \begin{equation}
@@
 \noindent
-
-The accumulated uncertainty is not a directly interpretable
-quantity. It only helps us define a sensible objective for the
-interval spliting step, as we discuss in
+The accumulated uncertainty is not a directly interpretable quantity,
+since it measures the total uncertainty that characterizes the
+\(s\)-th feature effect plot. As stated in the Introduction, ALE
+outcomes should be interpreted at a local level, therefore, the only
+\(\mathcal{H}(x_s)\) is the directly interpretable quantity. However,
+the accumulated uncertainty will help us define a sensible objective
+for the bin-splitting step, as we will discuss in
 Section~\ref{sec:interval-spliting}.
@@
 \(h \to 0\). We are, therefore, obliged to split the axis of \(x_s\)
-into a sequence of non-overlaping intervals (bins) and estimate the
+into a sequence of non-overlapping intervals (bins) and estimate the
 mean and the variance from the samples that lie inside each bin. The
@@
 the expected effects:
-
 \begin{equation}
@@
 \end{equation}
-
 \noindent
@@
 defined as:
-
 \begin{equation}
@@
 \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
-at an interval \([z_1, z_2)\) is the accumulation of the all variances
-plus the accumulation of squared residuals inside the interval:
-
+\(\sigma^2(z_1, z_2)\) equals to the accumulation of the point-wise
+variances plus the squared residuals inside the interval:
 \begin{equation}
@@
 %
-The proof is in the Appendix. Theorem 1 decouples the accumulated
-variance at an interval into two terms. The first term
-\(\int_{z_1}^{z_2} \sigma^2(z) \partial z\), quantifies the aggregated
-uncertainty due to the natural characteristics of the experiment
-\(\mathcal{H}\) and the second term adds extra nuisance uncertainty
-due to the limited resolution \(\mathcal{H}_n\). In other words,
-enforcing the computation of a single effect for all points in
-\([z_1, z_2)]\), burdens the estimation inside the interval with a
-nuisance uncertainty of:
-
+The proof is in the Appendix. Theorem 1 decouples the total
+uncertainty inside a bin
+\(\mathcal{H}_{bin}(z_1, z_2) := \sigma^2(z_1, z_2)\) into two terms. The first term quantifies
+the accumulated uncertainty due to the natural characteristics of the
+experiment, i.e.,
+\(\mathcal{H}(z_1, z_2) = \int_{z_1}^{z_2}\mathcal{H}(z) \partial z\),
+and the second term adds extra nuisance uncertainty due to limiting
+the resolution
+\(\mathcal{H}_n(z_1, z_2) := \int_{z_1}^{z_2}\rho^2(z) \partial z
+\). In other words, the creation of a bin burdens the estimation with
+a nuisance uncertainty:
 \begin{equation}
   \label{eq:uncertainty_bin}
-  \mathcal{H}_{bin}(z_1, z_2) = \int_{z_1}^{z_2} \mathcal{H}(z) + \mathcal{H}_n(z) \partial z
+  \mathcal{H}_{bin}(z_1, z_2) = \mathcal{H}(z_1, z_2) + \mathcal{H}_n(z_1, z_2)
 \end{equation}
-
 Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be directly estimated
@@
 effect at the interval, Eq.~(\ref{eq:mu_bin}) is approximated by:
-
 \begin{equation}
@@
 approximated by
-
 \begin{equation}
@@
 \end{equation}
-
 The approximation is unbiased only if the points are uniformly
@@
 
-\subsection{Interval Spliting As A Clustering Problem}
+\subsection{Bin-Splitting: Finding Regions With Homogeneous Effects}
 \label{sec:interval-spliting}
 
-% Instead, we propose treating the axis-spliting step as an unsupervised
-% clustering problem. The objective of the clustering problem should
-% fulfill in the best way to contradictory ojectives. First, secure
-% robust estimations of the expected effect and the uncertainty inside
-% each bin given the limited instances of the training set and, second,
-% create bins with as homogeneous local effects as possible, for not
-% losing fine-grain resolution FEs due to wide bins.
+As we discussed in the Introduction, when the features are (strongly)
+correlated, ALE provides meaningful interpretations only inside a
+bin. The accumulation of the mean effects inside each bin is done only
+for a producing smooth curve for easier visual exploration. Therefore,
+the extraction of regions with common effects is particularly
+important for a correct interpretation.
 
-We formulate the axis-splitting as an unsupervised clustering
-problem. We search for the optimal bin splitting, i.e., the number and
-size of consecutive non-overlapping intervals that minimizes the
-accumulated variance. The optimization problem is defined as follows:
-
+We formulate the bin-splitting as an unsupervised clustering
+problem. We want to partition the axis of the \(s\)-th feature into a
+sequence of variable-size non-overlapping intervals, so that the
+regions that will be created to provide meaningful explanations. A
+region with a meaningful explanation means that it clusters together
+points with similar expected (averaged) effects. On the other hand,
+since we estimate all the quantities from the limited samples of the
+training set, we need enough samples inside each bin for robustly
+estimate the expected effect and the variance. Therefore, we transform
+the bin-splitting problem to the following optimization problem:
 \begin{equation}
@@
   \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N\\
+                                     & z_0 = x_{s,min}\\
+                                     & z_K = x_{s, max}
 \end{aligned}
 \end{equation}
+%
+We search for the sequence of intervals \(z_0, \ldots, z_K\) covering
+the axis of the \(s\)-th feature, \(z_0 = x_{s,min}, z_K = x_{s,max}\)
+that minimize the sum of bin variances. The constraint of at least
+\(N\) points in each bin sets the limit for the minimum number of
+points inside each bin, to robustly estimate the mean effect and the
+variance. There is no constraint about the bin size and the number of
+bins \(K\).
 
-The objective of the optimization problem is the sum of the
-accumulated variances in each bin, estimated by the instances of the
-training set. As we prove in Theorem 1,
-\(\mathcal{L} = \sum_{k=1}^K \mathcal{H}_{bin}(z_{k-1}, z_k)\) is the
-sum of the uncertainty of the local explanations \(\mathcal{H}\) due
-to heterogeneous effects and the added nuisance uncertainty due to
-bin-splitting \(\mathcal{H}_n\). Given that \(\mathcal{H}\) is not
-affected by interval splitting, optimizing \(\mathcal{L}\) equals to
-finding the sequence of non-overlaping bins that add the minimum
-nuisance uncertainty. The restriction of (\ref{eq:opt}) secures that
-each bin is populated with enough samples for a robust estimation,
-which is set by the parameter \(N\). In other words, the user makes
-the following proposal; \textit{find the bin sequence that adds the
-  minimum nuisance uncertainty, given that I want at least \(N\)
-  points per bin}.
+The objective function is the sum of the approximations of the bin
+variances. As we prove in Theorem 1, the observable bin variance can
+can be split in the accumulation of point-wise variances plus the
+residuals. In Corollary 1, we show that the accumulation of the
+point-wise variances is independent of the bin-splitting
+procedure. Therefore by minimizing \(L\), we minimize the accumulation
+of the point-wise residuals in each bin, searching for bins that
+cluster together points with a common expected effect. An additional
+interpretation is that we search for the bin-splitting that adds the
+minimum the nuisance uncertainty over the lower-bound of the
+uncertainty due to the natural characteristics of the experiment.
 
-\subsubsection{Algorithm For Solving The Clustering Problem}
+\subsubsection{Solve Bin-Splitting with Dynamic Programming}
 \label{sec:dynamic-programing}
 
-Solving the problem of finding (a) the optimal number of bins \(K\)
-and (b) the optimal bin limits for each bin
-\([z_{k-1}, z_k] \forall k\) to minimize:
+For achieving a computationally-grounded solution we set a threshold
+\(K_{max}\) on the maximum number of bins which also discretizes the
+solution space. The width of the bin can take discrete values that are
+multiple of the minimum step
+\(u = \frac{x_{s, max} - x_{s, min}}{K}\). For defining the solution,
+we use two indexes. The index \(i \in \{0, \ldots, K_{max}\}\) denotes
+the \(i\)-th limit \((z_i)\) and the index
+\(j \in \{0, \ldots, K_{max}\} \) denotes the \(j\)-th position along
+the axis \((x_j = x_{s,min} + j \cdot u)\). The recursive cost
+function \(T(i,j)\) is the cost of setting the \(i\)-th limit \(z_i\)
+to \(x_j\):
+\begin{equation}
+  \label{eq:recursive_cost}
+  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
+\end{equation}
+%
+where \(\mathcal{T}(0,j)\) equals zero if \(j=0\) and \(\infty\) in
+any other case. \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin
+with limits \([x_l, x_j)]\):
 
 \begin{equation}
-  \label{eq:1}
-  \mathcal{L} = \sum_{k=0}^K \hat{\sigma}_k(z_{k-1}, z_k)
+  \label{eq:cost_step}
+  \mathcal{B}(x_l, x_j) = \begin{cases}
+                            \infty, & \text{if $x_j > x_l$ or \(|\mathcal{S}_{(x_j, x_l)}| < N\)}\\
+                            0, & \text{if $x_j = x_l$}\\
+                            \hat{\sigma}^2(x_j, x_l), &\text{if $x_j \leq x_l$}
+  \end{cases}
 \end{equation}
-%
-The constraints are that all bins must include more than \(\tau\)
-points, i.e., \(|\mathcal{S}_k| \geq \tau\).
+
+The optimal solution is given by solving
+\(\mathcal{L} = \mathcal{T}(K_{max}, K_{max})\) and keeping track of the sequence of
+steps.
+
+% \begin{equation}
+%   \label{eq:1}
+%   \mathcal{L} = \sum_{k=0}^K \hat{\sigma}_k(z_{k-1}, z_k)
+% \end{equation}
+% %
+% The constraints are that all bins must include more than \(\tau\)
+% points, i.e., \(|\mathcal{S}_k| \geq \tau\).
+
+% \(\mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k)\) is the sum
+% of the uncertainty of the local explanations \(\mathcal{H}\) due to
+% heterogeneous effects and the added nuisance uncertainty due to
+% bin-splitting \(\mathcal{H}_n\). Given that \(\mathcal{H}\) is not
+% affected by interval splitting, optimizing \(\mathcal{L}\) equals to
+% finding the sequence of non-overlaping bins that add the minimum
+% nuisance uncertainty. The restriction of (\ref{eq:opt}) secures that
+% each bin is populated with enough samples for a robust estimation,
+% which is set by the parameter \(N\). In other words, the user makes
+% the following proposal; \textit{find the bin sequence that adds the
+%   minimum nuisance uncertainty, given that I want at least \(N\)
+%   points per bin}.
 
@@
 
-% Furthermore, we automate the step of splitting the axis into non-overlapping intervals.
-The need for non-overlapping bins emerges from the ignorance of the data-generating distribution, enforcing all estimations to be based on the limited instances of the training set.
-Therefore, there is an implicit trade-off behind the formation of bins.
-Each bin must include enough instances for a robust estimation of the bin FE (expected value), and the uncertainty of the explanation (variance).
-On the other hand, each bin should include points with similar local effects.
-Therefore, we transform the bin splitting step into an unsupervised clustering problem, encoding the trade-off mentioned above in the objective function.
-We formally show that the objective of the clustering problem has lower-bound the (unavoidable) heterogeneity, i.e., the aggregated uncertainty of the global explanation.
-Therefore, we aim to find the optimal grouping of samples that adds the slightest uncertainty over the unavoidable heterogeneity.
-We finally solve the minimization problem by finding the global optimum using dynamic programming.
-Our method works out of the box without requiring any input by the user.
-We provide a theoretical and empirical evaluation of our method.
-
-\subsection{Visualization of NAME Method and Discussion}
+\subsection{Visualization and Interpretation}
 \label{sec:visualization}
@@
 
-\subsubsection*{Acknowledgements}
+\subsubsection*{Acknowledgments}
 All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support.

### 2022-09-19 AISTATS2023/paper.tex
added examples of 4.1.1

@@
 \usepackage{hyperref}
-
+\usepackage{graphicx}
+\graphicspath{ {./../examples/} }
 % If you use BibTeX in apalike style, activate the following line:
@@
 
-\section{SYNTHETIC EXAMPLES}
+\clearpage
+\section{EXPERIMENTS}
+
+The experimental section is divided in simulation and real-world
+examples. The simulation examples, where the data-generating
+distribution \(p(\mathbf{X})\) and the predictive function
+\(f(\cdot)\) are defined by us, are used for comparing all methods
+against a solid ground-truth. Having the freedom to experiment with
+\(p(\mathbf{X}), f(\cdot)\), we explore the behavior of each method in
+different settings for highlighting important aspects
+(advantages/disadvantages) of each one. In particular, we design two
+end-to-end simulation examples; At the first one, we focus on
+comparing NAME method versus PDP-ICE in terms of quantifying the
+heterogeneous effects for a range of different black-box functions. At
+the second one, we provide empirical clues that NAME method achieves a
+better approximation compared to traditional ALE, due to appropriate
+bin-splitting. Afterwards, we also apply the methods on some
+real-world datasets. In these cases, it is infeasible to obtain ground
+truth information, therefore, we perform evaluation at two
+levels. First, we compare visually NAME versus PDP with ICE
+plots. Second, we hold as ground-truth the effects computed on the
+full training-set and we, then, perform subsampling to evaluate the
+robustness of the bin-splitting method.
+
+\paragraph{Metrics.} Describe the metrics used for comparing methods.
+
+\subsection{Synthetic Example 1: Uncertainty Quantification}
+
+In general, there is no a single unchallenged aggreement about the
+correct feature effect; each method models it in a slightly different
+way. Therefore, for creating a universal evaluation framework we
+reside on variations of the simple piecewise-linear function, where in
+most cases, we can have an homophony about the correct effect.
+
+In all the experiments of the example, we will use the following
+generative distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\)
+where \(x_1 \sim \mathcal{U}(0,1)\),
+\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2)\),
+\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). 
+%
+The predictive function is:
+
+\begin{equation}
+  \label{eq:2}
+  f(\mathbf{x}) = \begin{cases}
+                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
+                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
+                    \alpha f_{\mathtt{int}} &\text{otherwise}
+                  \end{cases}
+\end{equation}
+%
+where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
+\(f_{inter}(\mathbf{x}) = x_1x_3\). The term \(f_{\mathtt{lin}}\) is a
+linear function with \(a_1, a_2\) the weights of the corresponding
+variables. The term \(f_{\mathtt{lin}}\) is the interaction between
+the independent variables \(x_1\) and \(x_3\), with \(\alpha\) setting
+the weight of the interaction term.
+
+\subsubsection{Effect Without Uncertainty}
+
+We test the feature effect when no interaction term is apparent,
+i.e. \(\alpha=0\), there is no uncertainty about the value of \(x_2\),
+i.e. \(x_2 = x_1\) and the weights are equal weights \(a_1=a_2=1\).
+In this case, ground truth effect for \(f^{\mathtt{GT}}(x_1)\) is
+\(x_1\) when \(0 \leq x_1 < 0.25\), \(-x_1\) when
+\(0.25 \leq x_1 < 0.5\) and zero otherwise. In each position \(x_1\),
+there is complete absence of heterogeneous effects, and therefore, the
+uncertainty of the global explanation is zero,
+i.e. \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In figure we present,
+the effect of each method;
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.23\textwidth]{example_3/dale_auto_bins.png}
+  \includegraphics[width=.23\textwidth]{example_3/pdp_ice.png}
+  \caption{Figure 1}
+  \label{fig:ex-synth-1-1}
+\end{figure}
+
+\subsubsection{Effect With Uncertainty}
+
+\subsection{Synthetic Example 2: Bin-Splitting}
+
+The perfect bin-splitting according to the definition of the
+minimization problem, is the one that creates:
+
+\begin{itemize}
+\item the biggest bins in regions with constant effect
+\item the smallest bins with \(N_k > \tau\), in regions with non-constant effect
+\end{itemize}
+
+Prove that (a) this is a good objective, i.e., prove that on average,
+this splitting gives the minimum distance from the ground truth ALE
+and (b) that DP finds the optimal.
+
+Set-ups:
+
+For limited instances, measure the mean error and std (many seeds)
+between gt ALE with uncertainty and approximation with bin-splitting
+using (a) fixed-size bins (b) variable-size bins for many \(K\)
+
+for many different \(N_k\)
+
+\begin{itemize}
+\item In case where the perfect bin-splitting is served by
+  equally-sized bins, DP finds it automatically. Without DP, the user
+  should try many different \(K\). and compare them the user's overhead of
+  testing many differ
+\end{itemize}
+
+For many model-agnostic interpretation techniques
+– including interaction detection methods – ground
+truth information is usually not available on real-world
+data. Therefore, well-constructed simulation experi-
+ments with a known ground truth are often used for
+empirical evaluations and comparisons, while only one
+or few real-world datasets are used to demonstrate
+practical applicability (e.g., see Friedman et al. (2008),
+Fisher et al. (2019), Goldstein et al. (2015), Greenwell
+et al. (2018), or Aas et al. (2021)). Hence, we follow
+this commonly used approach to evaluate our method
+using various simulation settings.
 

### 2022-09-19 example_models/models.py

@@ -2,7 +2,7 @@ import matplotlib.pyplot as plt
 import typing
 import numpy as np
 from functools import partial
-
+import copy
 
 class ModelBase:
     def __init__(self):
@@ -114,12 +114,17 @@ class Example3(ModelBase):
     f(x1, x2) = 0.5 - x1 - x2    , if x1 + x2 >= 0.5 and x1 + x2 < 1
                 0                , otherwise
     """
-    def __init__(self):
-        pass
+    def __init__(self, a1=1, a2=1, a=0):
+        self.a1 = a1
+        self.a2 = a2
+        self.a = a
 
     def predict(self, x: np.ndarray) -> np.ndarray:
         """Callable (x: np.ndarray (N,D)) -> np.ndarray (N)
         """
+        x = copy.deepcopy(x)
+        x[:,0] = self.a1 * x[:,0]
+        x[:,1] = self.a2 * x[:,1]
 
         # find indices
         ind1 = x[:,0] + x[:,1] < .5
@@ -129,12 +134,16 @@ class Example3(ModelBase):
         y = np.zeros_like(x[:,0])
         y[ind1] = x[ind1,0] + x[ind1,1]
         y[ind2] = .5 - x[ind2,0] - x[ind2,1]
-        y += x[:, 0]*x[:, 2]
+        y += self.a * x[:, 0]*x[:, 2]
         return y
 
     def jacobian(self, x: np.ndarray) -> np.ndarray:
         """Callable (x: np.ndarray (N,D)) -> np.ndarray (N)
         """
+
+        x = copy.deepcopy(x)
+        x[:,0] = self.a1 * x[:,0]
+        x[:,1] = self.a2 * x[:,1]
 
         # find indices
         ind1 = x[:,0] + x[:,1] < .5
@@ -142,11 +151,19 @@ class Example3(ModelBase):
 
         # set values
         y = np.zeros_like(x)
-        y[ind1] = 1
-        y[ind2] = -1
-        y[:, 0] += x[:, 2]
-        return y
 
+        # for df/dx1
+        y[ind1, 0] = self.a1
+        y[ind2, 0] = -self.a1
+        y[:, 0] += self.a * x[:, 2]
+
+        # for df/dx2
+        y[ind1, 1] = self.a2
+        y[ind2, 1] = -self.a2
+
+        # for df/dx3
+        y[:, 2] = self.a * x[:, 2]
+        return y
 
 class LinearWithInteraction(ModelBase):
[…]

### 2022-09-19 examples/example_1.py

@@ -6,28 +6,34 @@ import feature_effect as fe
 import example_models.distributions as dist
 import example_models.models as models
 
+savefig = True
+np.random.seed(21)
 
 # gen dist
-gen_dist = dist.Correlated1(D=2, x1_min=0, x1_max=1, x2_sigma=0)
-X = gen_dist.generate(N=500)
+gen_dist = dist.Correlated_3D_1(D=3, x1_min=0, x1_max=1, x2_sigma=0, x3_sigma=0.5)
+X = gen_dist.generate(N=200)
 axis_limits = gen_dist.axis_limits
 
 # model
-model = models.Example1()
+model = models.Example3(a1=1, a2=1, a=0)
 # model.plot(axis_limits=axis_limits, nof_points=30)
 
-# ALE with equal bin sizes
-dale = fe.DALE(data=X,
-               model=model.predict,
-               model_jac=model.jacobian,
-               axis_limits=axis_limits)
+# # ALE with equal bin sizes
+# dale = fe.DALE(data=X,
+#                model=model.predict,
+#                model_jac=model.jacobian,
+#                axis_limits=axis_limits)
 
-alg_params = {"bin_method" : "fixed", "nof_bins" : 20, "min_points_per_bin": 5}
-dale.fit(alg_params=alg_params)
-y, var, stderr = dale.eval(x=np.linspace(axis_limits[0,0], axis_limits[1,0], 100),
-                           s=0,
-                           uncertainty=True)
-dale.plot(s=0)
+# alg_params = {"bin_method" : "fixed", "nof_bins" : 20, "min_points_per_bin": 5}
+# dale.fit(features=[0], alg_params=alg_params)
+# y, var, stderr = dale.eval(x=np.linspace(axis_limits[0,0], axis_limits[1,0], 100),
+#                            s=0,
+#                            uncertainty=True)
+# if savefig:
+#     pathname = os.path.join(path, "example_1", "dale_fixed_bins.png")
+#     dale.plot(s=0, error="std", savefig=pathname)
+# else:
+#     dale.plot(s=0, error="std")
 
 
 # ALE with variable bin sizes
@@ -37,25 +43,34 @@ dale = fe.DALE(data=X,
                axis_limits=axis_limits)
 
 alg_params = {"bin_method" : "dp", "max_nof_bins" : 20, "min_points_per_bin": 10}
-dale.fit(alg_params=alg_params)
[…]

### 2022-09-19 examples/example_2.py

@@ -6,28 +6,34 @@ import feature_effect as fe
 import example_models.distributions as dist
 import example_models.models as models
 
+savefig = True
+np.random.seed(21)
 
 # gen dist
-gen_dist = dist.Correlated1(D=2, x1_min=0, x1_max=1, x2_sigma=0)
+gen_dist = dist.Correlated_3D_1(D=3, x1_min=0, x1_max=1, x2_sigma=0, x3_sigma=0.5)
 X = gen_dist.generate(N=500)
 axis_limits = gen_dist.axis_limits
 
 # model
-model = models.Example2()
+model = models.Example3(a1=2, a2=.5, a=0)
 # model.plot(axis_limits=axis_limits, nof_points=30)
 
-# ALE with equal bin sizes
-dale = fe.DALE(data=X,
-               model=model.predict,
-               model_jac=model.jacobian,
-               axis_limits=axis_limits)
+# # ALE with equal bin sizes
+# dale = fe.DALE(data=X,
+#                model=model.predict,
+#                model_jac=model.jacobian,
+#                axis_limits=axis_limits)
 
-alg_params = {"bin_method" : "fixed", "nof_bins" : 20, "min_points_per_bin": 5}
-dale.fit(alg_params=alg_params)
-y, var, stderr = dale.eval(x=np.linspace(axis_limits[0,0], axis_limits[1,0], 100),
-                           s=0,
-                           uncertainty=True)
-dale.plot(s=0)
+# alg_params = {"bin_method" : "fixed", "nof_bins" : 20, "min_points_per_bin": 5}
+# dale.fit(features=[0], alg_params=alg_params)
+# y, var, stderr = dale.eval(x=np.linspace(axis_limits[0,0], axis_limits[1,0], 100),
+#                            s=0,
+#                            uncertainty=True)
+# if savefig:
+#     pathname = os.path.join(path, "example_2", "dale_fixed_bins.png")
+#     dale.plot(s=0, error="std", savefig=pathname)
+# else:
+#     dale.plot(s=0, error="std")
 
 
 # ALE with variable bin sizes
@@ -36,26 +42,35 @@ dale = fe.DALE(data=X,
                model_jac=model.jacobian,
                axis_limits=axis_limits)
 
-alg_params = {"bin_method" : "dp", "max_nof_bins" : 20, "min_points_per_bin": 10}
-dale.fit(alg_params=alg_params)
[…]

### 2022-09-19 examples/example_3.py

@@ -6,29 +6,34 @@ import feature_effect as fe
 import example_models.distributions as dist
 import example_models.models as models
 
+savefig = True
+np.random.seed(23)
 
 # gen dist
-sigma = .5
-gen_dist = dist.Correlated_3D_1(D=3, x1_min=0, x1_max=1, x2_sigma=0, x3_sigma=.5)
+gen_dist = dist.Correlated_3D_1(D=3, x1_min=0, x1_max=1, x2_sigma=0, x3_sigma=0.5)
 X = gen_dist.generate(N=500)
 axis_limits = gen_dist.axis_limits
 
 # model
-model = models.Example3()
+model = models.Example3(a1=1, a2=1, a=1)
 # model.plot(axis_limits=axis_limits, nof_points=30)
 
-# ALE with equal bin sizes
-dale = fe.DALE(data=X,
-               model=model.predict,
-               model_jac=model.jacobian,
-               axis_limits=axis_limits)
+# # ALE with equal bin sizes
+# dale = fe.DALE(data=X,
+#                model=model.predict,
+#                model_jac=model.jacobian,
+#                axis_limits=axis_limits)
 
-alg_params = {"bin_method" : "fixed", "nof_bins" : 20, "min_points_per_bin": 5}
-dale.fit(features=[0], alg_params=alg_params)
-y, var, stderr = dale.eval(x=np.linspace(axis_limits[0,0], axis_limits[1,0], 100),
-                           s=0,
-                           uncertainty=True)
-dale.plot(s=0, error="std")
+# alg_params = {"bin_method" : "fixed", "nof_bins" : 20, "min_points_per_bin": 5}
+# dale.fit(features=[0], alg_params=alg_params)
+# y, var, stderr = dale.eval(x=np.linspace(axis_limits[0,0], axis_limits[1,0], 100),
+#                            s=0,
+#                            uncertainty=True)
+# if savefig:
+#     pathname = os.path.join(path, "example_3", "dale_fixed_bins.png")
+#     dale.plot(s=0, error="std", savefig=pathname)
+# else:
+#     dale.plot(s=0, error="std")
 
 
 # ALE with variable bin sizes
@@ -38,25 +43,34 @@ dale = fe.DALE(data=X,
                axis_limits=axis_limits)
 
 alg_params = {"bin_method" : "dp", "max_nof_bins" : 20, "min_points_per_bin": 10}
-dale.fit(features=[0], alg_params=alg_params)
[…]

### 2022-09-19 feature_effect/dale.py

@@ -169,13 +169,12 @@ class DALE(FeatureEffectBase):
 
 
     def plot(self, s: int = 0, error="standard error", block=False, gt=None, gt_bins=None, savefig=False):
-        title = "DALE: Effect of feature %d" % (s + 1)
         vis.feature_effect_plot(self.feature_effect["feature_"+str(s)],
                                 self.eval,
                                 s,
                                 error=error,
                                 min_points_per_bin=self.feature_effect["feature_"+str(s)]["alg_params"]["min_points_per_bin"],
-                                title=title,
+                                title=None,
                                 block=block,
                                 gt=gt,
                                 gt_bins=gt_bins,

### 2022-09-19 feature_effect/pdp.py

@@ -325,14 +325,14 @@ class PDPwithICE:
         axis_limits = self.axis_limits
         X = self.data
         model = self.model
-        x = np.linspace(axis_limits[0,0], axis_limits[1,0], nof_points)
+        x = np.linspace(axis_limits[0, s], axis_limits[1, s], nof_points)
         self.x = x
         # pdp
         pdp = PDP(data=X, model=model, axis_limits=axis_limits)
         if normalized:
-            y_pdp = pdp.eval(x=x, s=0, uncertainty=False)
+            y_pdp = pdp.eval(x=x, s=s, uncertainty=False)
         else:
-            y_pdp = pdp.eval_unnorm(x=x, s=0, uncertainty=False)
+            y_pdp = pdp.eval_unnorm(x=x, s=s, uncertainty=False)
         self.y_pdp = y_pdp
 
         # ice curves
@@ -340,17 +340,18 @@ class PDPwithICE:
         for i in range(X.shape[0]):
             ice = ICE(data=X, model=model, axis_limits=axis_limits, instance=i)
             if normalized:
-                y = ice.eval(x=x, s=0, uncertainty=False)
+                y = ice.eval(x=x, s=s, uncertainty=False)
             else:
-                y = ice.eval_unnorm(x=x, s=0, uncertainty=False)
+                y = ice.eval_unnorm(x=x, s=s, uncertainty=False)
             y_ice.append(y)
         self.y_ice = np.array(y_ice)
 
-    def plot(self, s: int, normalized: bool = True, nof_points: int = 30) -> None:
+    def plot(self, s: int, normalized: bool = True,
+             nof_points: int = 30,
+             savefig = None) -> None:
         """Plot the s-th feature
         """
-        if self.y_pdp is None:
-            self.fit(s, normalized, nof_points)
+        self.fit(s, normalized, nof_points)
 
         axis_limits = self.axis_limits
-        vis.plot_PDP_ICE(s, self.x, self.y_pdp, self.y_ice)
+        vis.plot_PDP_ICE(s, self.x, self.y_pdp, self.y_ice, savefig)

### 2022-09-19 feature_effect/visualization.py

@@ -24,9 +24,9 @@ def feature_effect_plot(params, eval, feature, error, min_points_per_bin, title=
     fig, (ax1, ax2) = plt.subplots(2, 1, sharex=True)
 
     if title is None:
-        fig.suptitle("Effect of feature %d" % (feature + 1))
+        ax1.set_title("ALE with Uncertainty: $x_{" + str(feature + 1) + "}$")
     else:
-        fig.suptitle(title)
+        ax1.set_title(title)
 
     # first subplot
     feature_effect(ax1, x, y, estimator_var, std, rem_eff, first_empty_bin, limits, min_points_per_bin, error=error, gt=gt)
@@ -46,8 +46,7 @@ def feature_effect_plot(params, eval, feature, error, min_points_per_bin, title=
 
 def feature_effect(ax1, x, y, estimator_var, std, rem_eff, first_empty, limits, point_limit, error=True, gt=None):
     # first subplot
-    ax1.set_title("Plot")
-    ax1.plot(x, y, "b--", label="estimation")\
+    ax1.plot(x, y, "b--", label="$f_{\mu}$")\
 
     if first_empty is not None:
         added_line = .3*(np.max(y) - np.min(y))
@@ -56,7 +55,7 @@ def feature_effect(ax1, x, y, estimator_var, std, rem_eff, first_empty, limits, 
                    alpha=.7,
                    label="first bin with < " + str(point_limit) + " points")
     if error == "std":
-        ax1.fill_between(x, y-std, y+std, color='red', alpha=0.2, label="std")
+        ax1.fill_between(x, y-std, y+std, color='red', alpha=0.2, label="$f_{\sigma^2}$")
         # ax1.fill_between(x, y-rem_eff, y+rem_eff, color='blue', alpha=0.7, label="std")
     elif error == "standard error":
         ax1.fill_between(x, y-2*np.sqrt(estimator_var), y+2*np.sqrt(estimator_var), color='red', alpha=0.6, label="standard error")
@@ -74,19 +73,20 @@ def feature_effect(ax1, x, y, estimator_var, std, rem_eff, first_empty, limits, 
 
 
 def effects_per_bin(ax2, bin_effects, bin_variance, error, is_bin_empty, limits, dx, gt_bins=None, point_limit=10):
-    ax2.set_title("Effects per bin")
+    # ax2.set_title("Effects per bin")
     bin_centers = (limits[:-1] + limits[1:]) / 2
     is_bin_full = ~np.array(is_bin_empty)
     if error:
[…]

### 2022-09-20 AISTATS2023/paper.tex
added bin spliting script

@@
 \(f(\cdot)\) are defined by us, are used for comparing all methods
-against a solid ground-truth. Having the freedom to experiment with
-\(p(\mathbf{X}), f(\cdot)\), we explore the behavior of each method in
-different settings for highlighting important aspects
-(advantages/disadvantages) of each one. In particular, we design two
-end-to-end simulation examples; At the first one, we focus on
-comparing NAME method versus PDP-ICE in terms of quantifying the
-heterogeneous effects for a range of different black-box functions. At
-the second one, we provide empirical clues that NAME method achieves a
-better approximation compared to traditional ALE, due to appropriate
+against a solid ground-truth. In particular, we design two end-to-end
+simulation examples; At the first one, we focus on comparing NAME
+method versus PDP-ICE in terms of quantifying the heterogeneous
+effects for a range of different black-box functions. At the second
+one, we provide empirical clues that NAME method achieves a better
+approximation compared to traditional ALE, due to appropriate
 bin-splitting. Afterwards, we also apply the methods on some
@@
 
-In general, there is no a single unchallenged aggreement about the
-correct feature effect; each method models it in a slightly different
-way. Therefore, for creating a universal evaluation framework we
-reside on variations of the simple piecewise-linear function, where in
-most cases, we can have an homophony about the correct effect.
-
-In all the experiments of the example, we will use the following
-generative distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\)
-where \(x_1 \sim \mathcal{U}(0,1)\),
-\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2)\),
-\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). 
-%
-The predictive function is:
+This synthetic example aims to showcase the superiority of NAME method
+in quantifying the main effect and the uncertainty compared to PDP
+with ICE plots. For being able to compare the two methods against a
+solid ground-truth, we limit ourserlves a simple evaluation framework based
+on following piecewise linear function:
 
 \begin{equation}
-  \label{eq:2}
+  \label{eq:synth-ex-1-function}
   f(\mathbf{x}) = \begin{cases}
@@
 where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
-\(f_{inter}(\mathbf{x}) = x_1x_3\). The term \(f_{\mathtt{lin}}\) is a
-linear function with \(a_1, a_2\) the weights of the corresponding
-variables. The term \(f_{\mathtt{lin}}\) is the interaction between
-the independent variables \(x_1\) and \(x_3\), with \(\alpha\) setting
-the weight of the interaction term.
+\(f_{int}(\mathbf{x}) = x_1x_3\). As we observe, the linear term
+\(f_{\mathtt{lin}}\) includes the two correlated features and the term
+\(f_{\mathtt{int}}\) interacts the two non-correlated variables. The
+samples that we use in all examples are coming from the following
+distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
+\(x_1 \sim \mathcal{U}(0,1)\),
+\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0)\) and
+\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test three variations
+by changing the parameters of the black-box function and of the
+generative distribution. Despite the model's simplicity, as we will
+see below, PDP with ICE plot fail completely to capture the
+ground-truth effect, due to the correlation between features \(X_1\)
+and \(X_2\). Furthermore, besides not the focus of the current
+example, NAME method manages to automatically extract the regions with
+constant effect. Details for obtaining the ground truth effects and
+the experimental set-up are in the Appendix.\footnote{The Appendix
+  contains the effects for all the features, some of them are ommited
+  here for compact presentation.}
 
-\subsubsection{Effect Without Uncertainty}
+\paragraph{No Interaction, Equal weights.}
 
 We test the feature effect when no interaction term is apparent,
-i.e. \(\alpha=0\), there is no uncertainty about the value of \(x_2\),
-i.e. \(x_2 = x_1\) and the weights are equal weights \(a_1=a_2=1\).
-In this case, ground truth effect for \(f^{\mathtt{GT}}(x_1)\) is
-\(x_1\) when \(0 \leq x_1 < 0.25\), \(-x_1\) when
-\(0.25 \leq x_1 < 0.5\) and zero otherwise. In each position \(x_1\),
-there is complete absence of heterogeneous effects, and therefore, the
-uncertainty of the global explanation is zero,
-i.e. \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In figure we present,
-the effect of each method;
+i.e. \(\alpha=0\), and the weights are equal \(a_1=a_2=1\).  In this
+case, ground truth effect for \(f^{\mathtt{GT}}(x_1)\) is \(x_1\) when
+\(0 \leq x_1 < 0.25\), \(-x_1\) when \(0.25 \leq x_1 < 0.5\) and zero
+otherwise. In each position \(x_1\), there is complete absence of
+heterogeneous effects, and therefore, the uncertainty of the global
+explanation is zero, i.e. \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
+Figure~\ref{fig:ex-synth-1-1} we see that PDP effect is completely
+wrong and ICE plots imply heterogeneous effects. In contrast, ALE
+captures correctly the average effect and the zero uncertainty.
 
@@
   \centering
-  \includegraphics[width=.23\textwidth]{example_3/dale_auto_bins.png}
-  \includegraphics[width=.23\textwidth]{example_3/pdp_ice.png}
+  \includegraphics[width=.23\textwidth]{example_1/dale_feat_0.png}
+  \includegraphics[width=.23\textwidth]{example_1/pdp_ice_feat_0.png}
+  \caption{Feature effect for \(x_1\); NAME method at the left and PDP
+    with ICE at the right.}
+  \label{fig:synth-ex-1-case-1}
+\end{figure}
+
+\paragraph{No Uncertainty, Different weights.}
+
+As before, there is no interaction term, i.e. \(\alpha=0\), and
+uncertainty about the value of \(x_2\), i.e.  \(x_2 = x_1\), but we
+give different weight to each feture \(a_1=2, a_2=0.5\). Again, we
+observe that ALE quantifies perfectly the ground truth effect and the
+zero-uncertainty in contrast to PDP with ICE.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.23\textwidth]{example_2/dale_feat_0.png}
+  \includegraphics[width=.23\textwidth]{example_2/pdp_ice_feat_0.png}
   \caption{Figure 1}
@@
 
-\subsubsection{Effect With Uncertainty}
+\paragraph{Uncertainty, Equal weights.}
+
+In this case we activate the interaction term, i.e. \(a=1\), and we
+set equal weights \(a_1=a_2=1\). In contrast with above, the
+interaction term provokes heterogeneous effects for features
+\(x_1, x_3\); at any position \(x_1\), the local effects dependend on
+the unknown value of \(X_3\) and vice-versa. The effect of \(x_2\)
+continues to have zero-uncertainty, since it does not appear in any
+interaction term. NAME captures perfectly the effect of all features,
+and the uncertainty in the case of \(x_1, x_2\). On the contrary,
+PDP-ICE quantify the effect and the uncertainty correctly only in the
+case of \(x_3\). For all other features, both the average effect and
+the uncertainty implied by ICE plots are wrong.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.23\textwidth]{example_3/dale_feat_0.png}
+  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_0.png}\\
+  \includegraphics[width=.23\textwidth]{example_3/dale_feat_1.png}
+  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_1.png}\\
+  \includegraphics[width=.23\textwidth]{example_3/dale_feat_2.png}
+  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_2.png}\\
+  \caption{Figure 1}
+  \label{fig:ex-synth-1-1}
+\end{figure}
+
+\paragraph{Discussion.}
+
+The examples presented above, do not cover all cases. Therefore, they
+are not sufficiect to prove that NAME method quantifies the effect
+correctly in all cases. Their main objective is to highlight that PDP
+indifference for correlated features provoques erroneous effects, even
+in very simple cases. Unfortunately, the methods that are based on
+PDP and ICE to quantify heterogeneity are also vulnerable to the same
+errors.
+
+The examples above do not cover the case where the interaction term
+include correlated features, \(x_1, x_2\) for our example. In this
+case, there is controversy about the ground-truth
+effect~\citep{Gromping2020MAEP}.
 
@@
 
-The perfect bin-splitting according to the definition of the
-minimization problem, is the one that creates:
+The aim of this example is to illustrate the advantages of automatic
+bin-splitting based on the objective of Eq.(\ref{eq:opt}). We explore
+different setups (black-box functios and dististributions) where, in
+each one, we treat as ground-truth the ALE obtained with very dense
+sampling. Afterwards, we measure the error of the ALE approximation
+with \(K\) fixed-size bins versus the approximation of NAME with
+automatic bin-spltting.
 

### 2022-09-20 example_models/models.py

@@ -27,7 +27,7 @@ class ModelBase:
         XX, YY = np.meshgrid(x1, x2)
         x = np.vstack([XX.ravel(), YY.ravel()]).T
         Z = self.predict(x)
-        ZZ = Z.reshape([30, 30])
+        ZZ = Z.reshape([nof_points, nof_points])
 
         plt.figure()
         plt.contourf(XX, YY, ZZ, levels=100)
@@ -110,9 +110,9 @@ class Example2(ModelBase):
 
 class Example3(ModelBase):
     """
-                x1 + x2          , if x1 + x2 < 0.5
-    f(x1, x2) = 0.5 - x1 - x2    , if x1 + x2 >= 0.5 and x1 + x2 < 1
-                0                , otherwise
+                x1 + x2 + x1x3          , if x1 + x2 < 0.5
+    f(x1, x2) = 0.5 - x1 - x2 + x1x3    , if x1 + x2 >= 0.5 and x1 + x2 < 1
+                x1 + x3                 , otherwise
     """
     def __init__(self, a1=1, a2=1, a=0):
         self.a1 = a1
@@ -162,7 +162,7 @@ class Example3(ModelBase):
         y[ind2, 1] = -self.a2
 
         # for df/dx3
-        y[:, 2] = self.a * x[:, 2]
+        y[:, 2] = self.a * x[:, 0]
         return y
 
 class LinearWithInteraction(ModelBase):
@@ -204,6 +204,7 @@ class SquareWithInteraction(ModelBase):
 class PiecewiseLinear(ModelBase):
     def __init__(self, params):
         self.params = params
+        self.nof_parts = len(params)
 
     @staticmethod
     def _linear_part(x, a, b, x0):
@@ -223,8 +224,8 @@ class PiecewiseLinear(ModelBase):
     def predict(self, x):
         """f(x1, x2) = a + b*x1 + x1x2
         """
-        condlist = [self._create_cond(x, i, s=0) for i in range(4)]
-        funclist = [self._create_func(i, self._linear_part) for i in range(4)]
+        condlist = [self._create_cond(x, i, s=0) for i in range(self.nof_parts)]
+        funclist = [self._create_func(i, self._linear_part) for i in range(self.nof_parts)]
 
         y = np.zeros(x.shape[0])
         for i, cond in enumerate(condlist):
@@ -232,7 +233,7 @@ class PiecewiseLinear(ModelBase):
         return y
 
     def jacobian(self, x):
[…]

### 2022-09-20 examples/example_bin_splitting_1.py

@@ -1,0 +1,157 @@
+import examples.utils as utils
+path = utils.add_parent_path()
+import matplotlib.pyplot as plt
+import numpy as np
+import feature_effect as fe
+import example_models.distributions as dist
+import example_models.models as models
+
+
+def compare(dale_gt, dale):
+    limits_gt = dale_gt.feature_effect["feature_0"]["limits"]
+    limits = dale.feature_effect["feature_0"]["limits"]
+    ind = np.digitize(limits, limits_gt) - 1
+
+
+    bins_res_err = []
+    bins_mu_err = []
+    bins_var_err = []
+    for i in range(limits.shape[0] - 1):
+        means = dale_gt.feature_effect["feature_0"]["bin_effect"][ind[i]+1:ind[i+1]]
+        variances = dale_gt.feature_effect["feature_0"]["bin_variance"][ind[i]+1:ind[i+1]]
+
+        res_err = means.var()
+        bins_res_err.append(res_err)
+
+        mu_gt = means.mean()
+        mu_est = dale.feature_effect["feature_0"]["bin_effect"][i]
+        bins_mu_err.append(np.abs(mu_gt - mu_est))
+
+        var_gt = variances.mean()
+        var_est = dale.feature_effect["feature_0"]["bin_variance"][i]
+        bins_var_err.append(np.abs(var_gt - var_est))
+
+    res = np.mean(bins_res_err)
+    mu = np.mean(bins_mu_err)
+    var = np.mean(bins_var_err)
+    return res, mu, var
+
+
+def measure_fixed_error(dale_gt, K_list, nof_iterations):
+    # fit approximation
+    rho_mean = []
+    rho_std = []
+
+    mu_mean = []
+    mu_std = []
+
+    var_mean = []
+    var_std = []
+
+    # for fixed
+    for i, k in enumerate(K_list):
+        rho_tmp = []
+        mu_tmp = []
+        var_tmp = []
+        for l in range(nof_iterations):
+            X = gen_dist.generate(N=500)
+            dale = fe.DALE(data=X,
+                           model=model.predict,
+                           model_jac=model.jacobian,
+                           axis_limits=axis_limits)
+            alg_params = {"bin_method" : "fixed", "nof_bins" : k, "min_points_per_bin": 2}
+            try:
[…]

### 2022-09-20 feature_effect/utils.py

@@ -248,7 +248,6 @@ def compute_accumulated_effect(x: np.ndarray, limits: np.ndarray, bin_effect: np
 
     # find where each point belongs to
     ind = np.digitize(x, limits)
-
     # find for each point, the accumulated full-bin effect
     if square:
         x_cumsum = (bin_effect * dx ** 2).cumsum()

### 2022-09-20 feature_effect/visualization.py

@@ -36,7 +36,7 @@ def feature_effect_plot(params, eval, feature, error, min_points_per_bin, title=
                     limits, dx, gt_bins, min_points_per_bin)
 
     if savefig:
-        plt.savefig(savefig)
+        plt.savefig(savefig, bbox_inches="tight")
 
     if block is False:
         plt.show(block=False)
@@ -198,5 +198,5 @@ def plot_PDP_ICE(s, x, y_pdp, y_ice, savefig):
     plt.plot(x, y_pdp, color="blue", label="PDP")
     plt.legend()
     if savefig is not None:
-        plt.savefig(savefig)
+        plt.savefig(savefig, bbox_inches="tight")
     plt.show(block=False)

### 2022-09-21 AISTATS2023/paper.tex
added synthetic exampls

@@
 \usepackage{amssymb}
+\usepackage{xfrac}
 \usepackage{hyperref}
@@
 The aim of this example is to illustrate the advantages of automatic
-bin-splitting based on the objective of Eq.(\ref{eq:opt}). We explore
-different setups (black-box functios and dististributions) where, in
-each one, we treat as ground-truth the ALE obtained with very dense
-sampling. Afterwards, we measure the error of the ALE approximation
-with \(K\) fixed-size bins versus the approximation of NAME with
-automatic bin-spltting.
+bin-splitting based on the objective of Eq.(\ref{eq:opt}). For
+evaluating our approach, we create the following validation framework;
+We generate a big dataset using dense sampling \(N=10^6\) and we treat
+as ground-truth the ALE estimation using dense fixed-size binning
+\(K=10^3\). Afterwards, we generate less samples \(X=500\) and we
+compare (a) the fixed-size ALE estimation for many different \(K\)
+versus (b) the auto-bin splitting algorithm. In all cases, the data
+generating distribution is \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
+\(x_1 \sim \mathcal{U}(0,1)\),
+\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=2)\).
 
-\begin{itemize}
-\item the biggest bins in regions with constant effect
-\item the smallest bins with \(N_k > \tau\), in regions with non-constant effect
-\end{itemize}
+We denote as
+\(\mathcal{Z^{\mathtt{DP}}} = \{z^{\mathtt{DP}}_{k-1}, \cdots,
+z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic-splitting
+based on the optimisation problem of Eq.~(\ref{eq:opt}) and with
+\(\mathcal{Z^{\mathtt{K}}}\}\) the corresponding fixed-sized
+splitting. The comparison is done using three metrics; (a)
+\(\mathcal{L}_{\mathtt{DP|K}}^{\mu} =
+\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k=1}^K | \mu_k -
+\hat{\mu}_k | \) quantifies the average error in estimating the the
+expected effect, (b)
+\(\mathcal{L}_{\mathtt{DP|K}}^{\sigma^2} =
+\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k=1}^K | \sigma_k^2 -
+\hat{\sigma}^2_k | \) the average error in estimating the variance of
+the effect and (c)
+\(\mathcal{L}^{\rho^2}_{\mathtt{DP|K}} =
+\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k=1}^K \rho^2_k \) the
+average extra nuisance uncertainty.
 
-Prove that (a) this is a good objective, i.e., prove that on average,
-this splitting gives the minimum distance from the ground truth ALE
-and (b) that DP finds the optimal.
+\paragraph{Example 1.}
 
-Set-ups:
+In this example, we defin a black-box function
+\(f_1(x_1) = a_1x_1 + x_1x_2\) with 6 piecewise-linear regions of
+different-size, i.e., \(a_1\) equals to \(=\{2, -2, 5, -10, 0.5\}\) in
+the regions defined by the sequence \(\{0, 0.2, 0.4, 0.45, 0.5,
+1\}\). The example highlights the importance of variable-size
+splitting. Given that the smalest intervals size is \(0.05\) any
+fixed-size splitting with less than \( \frac{1}{\Delta x} = 20\) bins,
+will mix together effects and increase the variance error
+\(\mathcal{L}_{\mathtt{K}}^{\sigma^2}\) due to the increase of the
+residual \(\mathcal{L}_{\mathtt{K}}^{\rho^2} > 0\). This is evident in
+figure. On the other hand, a large number of bins divides homogeneuos
+regions, i.e. region \([0.5, 1)\), into smaller ones with more
+unstable estimations which increases
+\(\mathcal{L}_{\mathtt{K}}^{\mu}\) Figure. Variable-size binning leads
+to the optimal solution. As we see in figure automatic-bin splitting
+small bins in areas with fine-grain effects and merges together
+regions with similar effects, leading to a near optimal bin splitting.
 
-For limited instances, measure the mean error and std (many seeds)
-between gt ALE with uncertainty and approximation with bin-splitting
-using (a) fixed-size bins (b) variable-size bins for many \(K\)
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_1.jpg}
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_2.jpg}\\
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_3.jpg}
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.jpg}
+  \caption{Figure 1}
+  \label{fig:ex-synth-1-1}
+\end{figure}
 
-for many different \(N_k\)
+\paragraph{Example 2.}
 
-\begin{itemize}
-\item In case where the perfect bin-splitting is served by
-  equally-sized bins, DP finds it automatically. Without DP, the user
-  should try many different \(K\). and compare them the user's overhead of
-  testing many differ
-\end{itemize}
+The black-box function \(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\)
 
-For many model-agnostic interpretation techniques
-– including interaction detection methods – ground
-truth information is usually not available on real-world
-data. Therefore, well-constructed simulation experi-
-ments with a known ground truth are often used for
-empirical evaluations and comparisons, while only one
-or few real-world datasets are used to demonstrate
-practical applicability (e.g., see Friedman et al. (2008),
-Fisher et al. (2019), Goldstein et al. (2015), Greenwell
-et al. (2018), or Aas et al. (2021)). Hence, we follow
-this commonly used approach to evaluate our method
-using various simulation settings.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_1.jpg}
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_2.jpg}\\
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_3.jpg}
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_4.jpg}
+  \caption{Figure 1}
+  \label{fig:ex-synth-1-1}
+\end{figure}
 

### 2022-09-21 examples/example_bin_splitting_2.py

@@ -7,20 +7,15 @@ import example_models.distributions as dist
 import example_models.models as models
 
 savefig = True
-folder_name =  "example_bin_splitting_1"
+folder_name =  "example_bin_splitting_2"
 np.random.seed(23)
 
 # gen dist
-gen_dist = dist.Correlated1(D=2, x1_min=0, x1_max=1, x2_sigma=.5)
+gen_dist = dist.Correlated1(D=2, x1_min=0, x1_max=1, x2_sigma=.4)
 axis_limits = gen_dist.axis_limits
 
 # model
-params = [{"from": 0., "to": .2, "a": 0, "b": 2},
-          {"from": .2, "to": .4, "a": .2, "b": -2},
-          {"from": .4, "to": .45, "a": 0, "b": 5},
-          {"from": .45, "to": .5, "a": 2.5, "b": -10},
-          {"from": .5, "to": 1., "a": -2.5, "b": .5}]
-model = models.PiecewiseLinear(params)
+model = models.SquareWithInteraction(b0=0, b1=4, b2=1, b3=1)
 
 # fit ground-truth
 X_0 = gen_dist.generate(N=1000000)
@@ -31,6 +26,7 @@ dale_gt = fe.DALE(data=X_0,
 
 alg_params = {"bin_method" : "fixed", "nof_bins" : 1000, "min_points_per_bin": 5}
 dale_gt.fit(features=0, alg_params=alg_params)
+
 
 # fit approx with auto binning
 X = gen_dist.generate(N=500)
@@ -48,9 +44,8 @@ else:
     dale.plot()
 
 
-
 # fit approximation
-K_list = list(range(2,101))
+K_list = list(range(2,11)) + list(range(15, 101, 5))
 stats_fixed = utils.measure_fixed_error(dale_gt,
                                         gen_dist,
                                         model,

### 2022-09-21 examples/utils.py

@@ -1,5 +1,8 @@
 import sys
 import os
+import feature_effect as fe
+import numpy as np
+import matplotlib.pyplot as plt
 
 def add_parent_path():
     cwd = os.getcwd()
@@ -13,3 +16,131 @@ def add_parent_path():
         raise OSError
 
     return path_to_examples
+
+
+def compare(dale_gt, dale):
+    """Returns the mean error (error per bin) for mu, var, res
+
+    :param dale_gt:
+    :param dale:
+    :returns:
+
+    """
+    limits_gt = dale_gt.feature_effect["feature_0"]["limits"]
+    limits = dale.feature_effect["feature_0"]["limits"]
+    ind = np.digitize(limits, limits_gt) - 1
+
+    rho_list = []
+    mu_err_list = []
+    var_err_list = []
+    for i in range(limits.shape[0] - 1):
+        means = dale_gt.feature_effect["feature_0"]["bin_effect"][ind[i]+1:ind[i+1]]
+        variances = dale_gt.feature_effect["feature_0"]["bin_variance"][ind[i]+1:ind[i+1]]
+
+        rho_bin = means.var()
+        rho_list.append(rho_bin)
+
+        mu_gt = means.mean()
+        mu_est = dale.feature_effect["feature_0"]["bin_effect"][i]
+        mu_err_list.append(np.abs(mu_gt - mu_est))
+
+        var_gt = variances.mean()
+        var_est = dale.feature_effect["feature_0"]["bin_variance"][i]
+        var_err_list.append(np.abs(var_gt - var_est))
+
+    mean_rho = np.mean(rho_list)
+    mean_mu_err = np.mean(mu_err_list)
+    mean_var_err = np.mean(var_err_list)
+    return mean_rho, mean_mu_err, mean_var_err
+
+
+def measure_fixed_error(dale_gt, gen_dist, model, axis_limits, K_list, nof_iterations, nof_points):
+    # fit approximation
+    rho_mean = []
+    rho_std = []
+
+    mu_mean = []
+    mu_std = []
+
+    var_mean = []
+    var_std = []
+
+    # for fixed
+    for i, k in enumerate(K_list):
+        rho_tmp = []
+        mu_tmp = []
+        var_tmp = []
+        for l in range(nof_iterations):
+            X = gen_dist.generate(N=nof_points)
+            dale = fe.DALE(data=X,
+                           model=model.predict,
+                           model_jac=model.jacobian,
+                           axis_limits=axis_limits)
[…]

### 2022-09-21 examples/real_dataset_1.py
added real worls examples

@@ -1,0 +1,175 @@
+import examples.utils as utils
+path = utils.add_parent_path()
+import numpy as np
+import pandas as pd
+import random as python_random
+import tensorflow as tf
+from tensorflow import keras
+import copy
+import feature_effect as fe
+import matplotlib.pyplot as plt
+
+def create_nn(X):
+    nn = keras.Sequential([keras.layers.Input(shape=[X.shape[1]]),
+                           keras.layers.Dense(1024, activation='relu', use_bias=True),
+                           keras.layers.Dense(512, activation='relu', use_bias=True),
+                           keras.layers.Dense(256, activation='relu', use_bias=True),
+                           keras.layers.Dense(128, activation='relu', use_bias=True),
+                           keras.layers.Dense(64, activation='relu', use_bias=True),
+                           keras.layers.Dense(32, activation='relu', use_bias=True),
+                           keras.layers.Dense(1, use_bias=True)])
+
+    nn.compile(optimizer=tf.keras.optimizers.Adam(learning_rate=0.01),
+              loss="mean_squared_error",
+              metrics="mean_absolute_error")
+
+    return nn
+
+
+def create_model(nn):
+    def model(X):
+        return nn(X).numpy()[:, 0]
+    return model
+
+
+def create_model_grad(nn):
+    """Computes the gradients of outputs w.r.t input image.
+
+    Args:
+        img_input: 4D image tensor
+        top_pred_idx: Predicted label for the input image
+
+    Returns:
+        Gradients of the predictions w.r.t img_input
+    """
+    def model_grad(inp):
+        x_inp = tf.cast(inp, tf.float32)
+
+        with tf.GradientTape() as tape:
+            tape.watch(x_inp)
+            preds = nn(x_inp)
+            grads = tape.gradient(preds, x_inp)
+        return grads.numpy()
+    return model_grad
+
+
+def scale_x(x, s):
+    x = copy.deepcopy(x)
+    if s == 7:
+        x = x*(39 - (-8)) + (-8)
+    elif s==8:
+        x = x*(50 - (-16)) + (-16)
+    elif s==9:
+        x = x*100
+    elif s==10:
+        x = x*67
[…]

### 2022-09-21 examples/real_dataset_2.py

@@ -1,0 +1,108 @@
+import examples.utils as utils
+path = utils.add_parent_path()
+import numpy as np
+import pandas as pd
+import random as python_random
+import tensorflow as tf
+from tensorflow import keras
+import copy
+import feature_effect as fe
+import matplotlib.pyplot as plt
+from sklearn.preprocessing import Normalizer
+
+def create_nn(X):
+    nn = keras.Sequential([keras.layers.Input(shape=[X.shape[1]]),
+                           keras.layers.Dense(256, activation='relu', use_bias=True),
+                           keras.layers.Dense(256, activation='relu', use_bias=True),                                     keras.layers.Dense(256, activation='relu', use_bias=True),                                     keras.layers.Dense(256, activation='relu', use_bias=True),
+                           keras.layers.Dense(128, activation='relu', use_bias=True),
+                           keras.layers.Dense(32, activation='relu', use_bias=True),
+                           keras.layers.Dense(1, use_bias=True)])
+
+    nn.compile(optimizer=tf.keras.optimizers.Adam(learning_rate=0.01),
+              loss=tf.keras.losses.BinaryCrossentropy(from_logits=True),
+              metrics=[tf.keras.metrics.BinaryAccuracy(),
+                       tf.keras.metrics.BinaryCrossentropy()])
+
+    return nn
+
+
+def create_model(nn):
+    def model(X):
+        return nn(X).numpy()[:, 0]
+    return model
+
+
+def create_model_grad(nn):
+    """Computes the gradients of outputs w.r.t input image.
+
+    Args:
+        img_input: 4D image tensor
+        top_pred_idx: Predicted label for the input image
+
+    Returns:
+        Gradients of the predictions w.r.t img_input
+    """
+    def model_grad(inp):
+        x_inp = tf.cast(inp, tf.float32)
+
+        with tf.GradientTape() as tape:
+            tape.watch(x_inp)
+            preds = nn(x_inp)
+            grads = tape.gradient(preds, x_inp)
+        return grads.numpy()
+    return model_grad
+
+
+def plot_local_effects(s, X, dfdx, mu, sigma):
+    plt.figure()
+    plt.title("Local Effects of feature: " + cols[s])
[…]

### 2022-09-21 examples/real_dataset_3.py

@@ -1,0 +1,121 @@
+import examples.utils as utils
+path = utils.add_parent_path()
+import numpy as np
+import pandas as pd
+import random as python_random
+import tensorflow as tf
+from tensorflow import keras
+import copy
+import feature_effect as fe
+import matplotlib.pyplot as plt
+from sklearn.preprocessing import Normalizer
+
+def create_nn(X):
+    nn = keras.Sequential([keras.layers.Input(shape=[X.shape[1]]),
+                           keras.layers.Dense(256, activation='relu', use_bias=True),
+                           keras.layers.Dense(256, activation='relu', use_bias=True),                                     keras.layers.Dense(128, activation='relu', use_bias=True),
+                           keras.layers.Dense(32, activation='relu', use_bias=True),
+                           keras.layers.Dense(1, use_bias=True)])
+
+    nn.compile(optimizer=tf.keras.optimizers.Adam(learning_rate=0.01),
+              loss=tf.keras.losses.MeanSquaredError(),
+              metrics=[tf.keras.metrics.MeanAbsoluteError()])
+
+    return nn
+
+
+def create_model(nn):
+    def model(X):
+        return nn(X).numpy()[:, 0]
+    return model
+
+
+def create_model_grad(nn):
+    """Computes the gradients of outputs w.r.t input image.
+
+    Args:
+        img_input: 4D image tensor
+        top_pred_idx: Predicted label for the input image
+
+    Returns:
+        Gradients of the predictions w.r.t img_input
+    """
+    def model_grad(inp):
+        x_inp = tf.cast(inp, tf.float32)
+
+        with tf.GradientTape() as tape:
+            tape.watch(x_inp)
+            preds = nn(x_inp)
+            grads = tape.gradient(preds, x_inp)
+        return grads.numpy()
+    return model_grad
+
+
+def plot_local_effects(s, X, dfdx, mu, sigma):
+    plt.figure()
+    plt.title("Local Effects of feature: " + cols[s])
+    plt.plot(X[:, s]*sigma+ mu, dfdx[:, s], "rx")
+    plt.show(block=False)
+
+
+# freeze seed
+np.random.seed(138232)
+python_random.seed(1244343)
+tf.random.set_seed(12384)
+
+# init data
[…]

### 2022-09-21 feature_effect/bin_estimation.py

@@ -354,12 +354,16 @@ class FixedSizeBase:
     def bin_valid(self, start, stop):
         return NotImplementedError
 
-    def solve(self, min_points, K):
+    def solve(self, min_points, K, enforce_bin_creation):
         self.min_points = min_points
         self.K = K
         assert min_points >= 2, "We need at least two points per bin to estimate the variance"
 
         limits, dx = np.linspace(self.xs_min, self.xs_max, num=K + 1, endpoint=True, retstep=True)
+        if enforce_bin_creation:
+            self.limits = limits
+            return limits
+
         valid_binning = True
         for i in range(K):
             start = limits[i]

### 2022-09-21 feature_effect/dale.py

@@ -116,7 +116,8 @@ class DALE(FeatureEffectBase):
         if alg_params["bin_method"] == "fixed":
             bin_est = be.FixedSize(self.data, self.data_effect, feature=s)
             bin_est.solve(min_points = alg_params["min_points_per_bin"],
-                          K = alg_params["nof_bins"])
+                          K = alg_params["nof_bins"],
+                          enforce_bin_creation = alg_params["enforce_bin_creation"])
         elif alg_params["bin_method"] == "greedy":
             bin_est = be.Greedy(self.data, self.data_effect, feature=s)
             bin_est.solve(min_points = alg_params["min_points_per_bin"],
@@ -168,7 +169,7 @@ class DALE(FeatureEffectBase):
     #     vis.plot_1D(x, y, title="ALE (Monte Carlo) for feature %d" % (s+1))
 
 
-    def plot(self, s: int = 0, error="standard error", block=False, gt=None, gt_bins=None, savefig=False):
+    def plot(self, s: int = 0, error="std", block=False, gt=None, gt_bins=None, savefig=False):
         vis.feature_effect_plot(self.feature_effect["feature_"+str(s)],
                                 self.eval,
                                 s,

### 2022-09-22 AISTATS2023/paper.tex
updated synthetic examples 1

@@
 %\bibliographystyle{apalike}
+
+\usepackage{tikz}
+\usetikzlibrary{matrix,positioning,arrows.meta,arrows,fit,backgrounds,decorations.pathreplacing}
+\tikzset{ mymat/.style={ matrix of math nodes, text height=2.5ex, text
+depth=0.75ex, text width=6.00ex, align=center, column
+sep=-\pgflinewidth, nodes={minimum height=5.0ex} }, mymats/.style={
+mymat, nodes={draw,fill=#1} }, mymat2/.style={ matrix of math nodes,
+text height=1.0ex, text depth=0.0ex, minimum width=5ex, % text
+width=7.00ex, align=center, column sep=-\pgflinewidth }, }
+\usetikzlibrary{shapes.geometric, arrows, backgrounds, scopes}
+\usepackage{pgfplots} \pgfplotsset{width=6.75cm, compat=newest}
+\usepackage[utf8]{inputenc} \DeclareUnicodeCharacter{2212}{−}
+\usepgfplotslibrary{groupplots,dateplot}
+\usetikzlibrary{patterns,shapes.arrows}
 
@@
 feature interactions, the global explanation may obfuscate
-heterogeneous effects~\citep{Herbinger2022repid} that exist under the
+heterogeneous effects~\citep{herbinger2022repid} that exist under the
 hood; a phenomenon called aggregation bias~\citep{mehrabi2021survey}.
@@
 \clearpage
-\section{EXPERIMENTS}
+\section{SIMULATION EXAMPLES}
+\label{sec:simulation-examples}
 
-The experimental section is divided in simulation and real-world
-examples. The simulation examples, where the data-generating
-distribution \(p(\mathbf{X})\) and the predictive function
-\(f(\cdot)\) are defined by us, are used for comparing all methods
-against a solid ground-truth. In particular, we design two end-to-end
-simulation examples; At the first one, we focus on comparing NAME
-method versus PDP-ICE in terms of quantifying the heterogeneous
-effects for a range of different black-box functions. At the second
-one, we provide empirical clues that NAME method achieves a better
-approximation compared to traditional ALE, due to appropriate
-bin-splitting. Afterwards, we also apply the methods on some
-real-world datasets. In these cases, it is infeasible to obtain ground
-truth information, therefore, we perform evaluation at two
-levels. First, we compare visually NAME versus PDP with ICE
-plots. Second, we hold as ground-truth the effects computed on the
-full training-set and we, then, perform subsampling to evaluate the
-robustness of the bin-splitting method.
+The simulation examples, where the data-generating distribution
+\(p(\mathbf{X})\) and the predictive function \(f(\cdot)\) are defined
+by us, enable the evaluation of competitive approaches against a solid
+ground-truth. We follow this common XAI
+practice~\citep{aas2021explaining, herbinger2022repid} for providing
+secure empirical evidence about the superiority of NAME method.
 
-\paragraph{Metrics.} Describe the metrics used for comparing methods.
+We split the simulation examples in two groups. The first group,
+Section~\ref{sec:simulation-examples-1}, aims at showcasing that NAME
+method is more accurate than PDP-ICE in quantifying the average effect
+and the level of heterogeneous effects (uncertainty), when input
+features are correlated. The second group,
+Section~\ref{sec:simulation-examples-2}, illustrates that NAME method
+achieves a better approximation (average effect and uncertainty) in
+cases of limited samples, due to automatic bin-splitting. In both
+groups, we choose to illustrate the most indicative examples; a more
+extensive evaluation is provided in the Appendix.
 
-\subsection{Synthetic Example 1: Uncertainty Quantification}
+\subsection{Case 1: Uncertainty Quantification}
+\label{sec:simulation-examples-1}
 
-This synthetic example aims to showcase the superiority of NAME method
-in quantifying the main effect and the uncertainty compared to PDP
-with ICE plots. For being able to compare the two methods against a
-solid ground-truth, we limit ourserlves a simple evaluation framework based
-on following piecewise linear function:
+In this simulation, we will compare NAME method against PDP-ICE in
+quantifying the main effect \(f_\mu\) and the uncertainty
+\(f_\sigma^2\). Since there is ambiguity about the ground-truth
+first-order effect in cases of correlated
+features, e.g.~\citep{apley2020visualizing,Gromping2020MAEP}, we limit
+ourselves to the following piecewise linear function,
 
@@
 distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
-\(x_1 \sim \mathcal{U}(0,1)\),
-\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0)\) and
-\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test three variations
-by changing the parameters of the black-box function and of the
-generative distribution. Despite the model's simplicity, as we will
-see below, PDP with ICE plot fail completely to capture the
-ground-truth effect, due to the correlation between features \(X_1\)
-and \(X_2\). Furthermore, besides not the focus of the current
-example, NAME method manages to automatically extract the regions with
-constant effect. Details for obtaining the ground truth effects and
-the experimental set-up are in the Appendix.\footnote{The Appendix
-  contains the effects for all the features, some of them are ommited
-  here for compact presentation.}
+\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
+\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test the effect
+computed by NAME and PDP-ICE in three cases; (a) no interaction
+(\(\alpha=0\)) and equal weights (\(a_1=a_2\)), (b) no interaction
+(\(\alpha=0\)) and different weigths (\( a_1 \neq a_2 \)) and (c) with
+interaction (\(\alpha \neq 0\)) with equal weights (\(a_1=a_2\)).
+
+In all cases, we firstly compute the ground-truth average effect and
+the uncertainty analytically (proofs in the Appendix) and then we
+compare it against the approximation provided by each method. The
+approximation is computed after generating \(N=300\) samples. As we
+will see, despite the model's simplicity, PDP-ICE fail in quantifying
+the effect correctly, due to correlation between features \(X_1\) and
+\(X_2\). Finally, besides it is not the focus of the current examples,
+we will also observe that NAME manages to correctly extract the
+regions with constant effect. Details for obtaining the ground truth
+effects and the experimental set-up are in the Appendix.
 
@@
 i.e. \(\alpha=0\), and the weights are equal \(a_1=a_2=1\).  In this
-case, ground truth effect for \(f^{\mathtt{GT}}(x_1)\) is \(x_1\) when
-\(0 \leq x_1 < 0.25\), \(-x_1\) when \(0.25 \leq x_1 < 0.5\) and zero
-otherwise. In each position \(x_1\), there is complete absence of
-heterogeneous effects, and therefore, the uncertainty of the global
-explanation is zero, i.e. \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
-Figure~\ref{fig:ex-synth-1-1} we see that PDP effect is completely
-wrong and ICE plots imply heterogeneous effects. In contrast, ALE
-captures correctly the average effect and the zero uncertainty.
+case, the ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\)
+when \(0 \leq x_1 < 0.25\), \(-x_1\) when \(0.25 \leq x_1 < 0.5\) and
+zero otherwise. In each position, there is total absence of
+heterogeneous effects; therefore, the uncertainty of the global
+explanation is zero \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
+Figure~\ref{fig:ex-synth-1-1}, we observe that PDP main effect is
+wrong and ICE plots imply the existence of heterogeneous effects. In
+contrast, ALE captures correctly the average effect and the zero
+uncertainty and it also groups perfectly the regions with
+constant-effect.
 
@@
   \centering
-  \includegraphics[width=.23\textwidth]{example_1/dale_feat_0.png}
-  \includegraphics[width=.23\textwidth]{example_1/pdp_ice_feat_0.png}
-  \caption{Feature effect for \(x_1\); NAME method at the left and PDP
-    with ICE at the right.}
+  \includegraphics[width=.23\textwidth]{example_1/dale_feat_0.pdf}
+  \includegraphics[width=.23\textwidth]{example_1/pdp_ice_feat_0.pdf}
+  \caption{No interaction, Equal weights: Feature effect for \(x_1\);
+    Ground-truth vs (a) NAME method at the left and (b) PDP-ICE at the
+    right.}
   \label{fig:synth-ex-1-case-1}
@@
 
-\paragraph{No Uncertainty, Different weights.}
+\paragraph{No Interaction, Different weights.}
 
-As before, there is no interaction term, i.e. \(\alpha=0\), and
-uncertainty about the value of \(x_2\), i.e.  \(x_2 = x_1\), but we
-give different weight to each feture \(a_1=2, a_2=0.5\). Again, we
-observe that ALE quantifies perfectly the ground truth effect and the
-zero-uncertainty in contrast to PDP with ICE.
+As before, there is no interaction term and, therefore, the
+ground-truth is zero, i.e.  \(f^{\mathcal{GT}}_{\sigma^2}(x_1) =
+0\). The weigts are different \(a_1=2, a_2=0.5\), therefore, the
+ground-truth effect is \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) when
+\(0 \leq x_1 < 0.2\), \(-2x_1\) when \(0.2 \leq x_1 < 0.4\) and zero
+otherwise. In Figure~\ref{fig:ex-synth-1-2}, we observe that PDP
+estimation is completely opposite to the ground-truth effect,
+i.e.~negative in \([0, 0.2)\) and positive in \([0.2, 0.4)\), and the
+ICE erroneously implies heterogeneous effects.  As before, ALE
+quantifies perfectly the ground truth effect and the zero-uncertainty,
+extracting correcltly the constant effect regions.
 
@@
   \centering
-  \includegraphics[width=.23\textwidth]{example_2/dale_feat_0.png}
-  \includegraphics[width=.23\textwidth]{example_2/pdp_ice_feat_0.png}
-  \caption{Figure 1}
-  \label{fig:ex-synth-1-1}
+  \includegraphics[width=.23\textwidth]{example_2/dale_feat_0.pdf}
+  \includegraphics[width=.23\textwidth]{example_2/pdp_ice_feat_0.pdf}
+  \caption{No interaction, Different weights: Feature effect for \(x_1\);
+    Ground-truth vs (a) NAME method at the left and (b) PDP-ICE at the
+    right.}
+  \label{fig:ex-synth-1-2}
 \end{figure}
@@
 In this case we activate the interaction term, i.e. \(a=1\), and we
-set equal weights \(a_1=a_2=1\). In contrast with above, the
-interaction term provokes heterogeneous effects for features
-\(x_1, x_3\); at any position \(x_1\), the local effects dependend on
-the unknown value of \(X_3\) and vice-versa. The effect of \(x_2\)
-continues to have zero-uncertainty, since it does not appear in any
-interaction term. NAME captures perfectly the effect of all features,
-and the uncertainty in the case of \(x_1, x_2\). On the contrary,
-PDP-ICE quantify the effect and the uncertainty correctly only in the
-case of \(x_3\). For all other features, both the average effect and
-the uncertainty implied by ICE plots are wrong.
+set equal weights \(a_1=a_2=1\). The interaction term provokes
+heterogeneous effects in features \(x_1, x_3\), i.e., at any position
+\(x_1\), the local effects dependend on the unknown value of \(X_3\)
+and vice-versa. The effect of \(x_2\) continues to have
+zero-uncertainty, since it does not appear in any interaction term. As
+we observe in Figure~\ref{fig:ex-synth-1-3}, NAME captures perfectly
+the effect of all features, and the uncertainty in all cases. As
+expected, PDP-ICE quantify the effect and the uncertainty correctly
+only in the case of \(x_3\), since \(X_3\) is independent from other
+features. For the correlated features, i.e. \(x_1, x_2\), the average
+effect computed by PDP and the uncertainty implied by ICE plots are
+wrong.
 
@@
   \centering
-  \includegraphics[width=.23\textwidth]{example_3/dale_feat_0.png}
-  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_0.png}\\
-  \includegraphics[width=.23\textwidth]{example_3/dale_feat_1.png}
-  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_1.png}\\
-  \includegraphics[width=.23\textwidth]{example_3/dale_feat_2.png}
-  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_2.png}\\
-  \caption{Figure 1}
-  \label{fig:ex-synth-1-1}
+  \includegraphics[width=.23\textwidth]{example_3/dale_feat_0.pdf}
+  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_0.pdf}\\
+  \includegraphics[width=.23\textwidth]{example_3/dale_feat_1.pdf}
+  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_1.pdf}\\
+  \includegraphics[width=.23\textwidth]{example_3/dale_feat_2.pdf}
+  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_2.pdf}\\
+  \caption{With interaction, equal weights: Feature effect for all
+    features, \(x_1\) to \(x_3\) from top to bottom; Ground-truth vs
+    (a) NAME method at the left columns and (b) PDP-ICE at the right
+    column.}
+  \label{fig:ex-synth-1-3}
 \end{figure}
@@
 
-The examples presented above, do not cover all cases. Therefore, they
-are not sufficiect to prove that NAME method quantifies the effect
-correctly in all cases. Their main objective is to highlight that PDP
-indifference for correlated features provoques erroneous effects, even
-in very simple cases. Unfortunately, the methods that are based on
-PDP and ICE to quantify heterogeneity are also vulnerable to the same
-errors.
-
-The examples above do not cover the case where the interaction term
-include correlated features, \(x_1, x_2\) for our example. In this
-case, there is controversy about the ground-truth
+In purpose, the examples above do not cover the case where there is an
+interaction term includes correlated features, i.e. \(x_1x_2\). In
+this scenario, there is an open debate about the ground-truth
 effect~\citep{Gromping2020MAEP}.
 
-\subsection{Synthetic Example 2: Bin-Splitting}
+\subsection{Examples 2: Bin-Splitting}
+\label{sec:simulation-examples-2}
 
@@
   \centering
-  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_1.jpg}
-  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_2.jpg}\\
-  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_3.jpg}
-  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.jpg}
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_1.pdf}
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_2.pdf}\\
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_3.pdf}
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.pdf}
   \caption{Figure 1}
@@
   \centering
-  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_1.jpg}
-  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_2.jpg}\\
-  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_3.jpg}
-  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_4.jpg}
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_1.pdf}
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_2.pdf}\\
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_3.pdf}
+  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_4.pdf}
   \caption{Figure 1}
@@
 \section{REAL-WORLD EXAMPLES}
+
+we also
+apply the methods on some real-world datasets. In these cases, it is
+infeasible to obtain ground truth information, therefore, we perform
+evaluation at two levels. First, we compare visually NAME versus PDP
+with ICE plots. Second, we hold as ground-truth the effects computed
+on the full training-set and we, then, perform subsampling to evaluate
+the robustness of the bin-splitting method.
 

### 2022-09-22 examples/example_1.py

@@ -30,7 +30,7 @@ model = models.Example3(a1=1, a2=1, a=0)
 #                            s=0,
 #                            uncertainty=True)
 # if savefig:
-#     pathname = os.path.join(path, "example_1", "dale_fixed_bins.png")
+#     pathname = os.path.join(path, "example_1", "dale_fixed_bins.pdf")
 #     dale.plot(s=0, error="std", savefig=pathname)
 # else:
 #     dale.plot(s=0, error="std")
@@ -48,11 +48,11 @@ y, var, stderr = dale.eval(x=np.linspace(axis_limits[0,0], axis_limits[1,0], 100
                            s=0,
                            uncertainty=True)
 if savefig:
-    pathname = os.path.join(path, "example_1", "dale_feat_0.png")
+    pathname = os.path.join(path, "example_1", "dale_feat_0.pdf")
     dale.plot(s=0, error="std", savefig=pathname)
-    pathname = os.path.join(path, "example_1", "dale_feat_1.png")
+    pathname = os.path.join(path, "example_1", "dale_feat_1.pdf")
     dale.plot(s=1, error="std", savefig=pathname)
-    pathname = os.path.join(path, "example_1", "dale_feat_2.png")
+    pathname = os.path.join(path, "example_1", "dale_feat_2.pdf")
     dale.plot(s=2, error="std", savefig=pathname)
 else:
     dale.plot(s=0, error="std")
@@ -64,11 +64,11 @@ pdp_ice = fe.PDPwithICE(data=X,
                         model=model.predict,
                         axis_limits=axis_limits)
 if savefig:
-    pathname = os.path.join(path, "example_1", "pdp_ice_feat_0.png")
+    pathname = os.path.join(path, "example_1", "pdp_ice_feat_0.pdf")
     pdp_ice.plot(s=0, normalized=True, nof_points=300, savefig=pathname)
-    pathname = os.path.join(path, "example_1", "pdp_ice_feat_1.png")
+    pathname = os.path.join(path, "example_1", "pdp_ice_feat_1.pdf")
     pdp_ice.plot(s=1, normalized=True, nof_points=300, savefig=pathname)
-    pathname = os.path.join(path, "example_1", "pdp_ice_feat_2.png")
+    pathname = os.path.join(path, "example_1", "pdp_ice_feat_2.pdf")
     pdp_ice.plot(s=2, normalized=True, nof_points=300, savefig=pathname)
 else:
     pdp_ice.plot(s=0, normalized=True, nof_points=300)

### 2022-09-22 examples/example_2.py

@@ -30,7 +30,7 @@ model = models.Example3(a1=2, a2=.5, a=0)
 #                            s=0,
 #                            uncertainty=True)
 # if savefig:
-#     pathname = os.path.join(path, "example_2", "dale_fixed_bins.png")
+#     pathname = os.path.join(path, "example_2", "dale_fixed_bins.pdf")
 #     dale.plot(s=0, error="std", savefig=pathname)
 # else:
 #     dale.plot(s=0, error="std")
@@ -48,11 +48,11 @@ y, var, stderr = dale.eval(x=np.linspace(axis_limits[0,0], axis_limits[1,0], 100
                            s=0,
                            uncertainty=True)
 if savefig:
-    pathname = os.path.join(path, "example_2", "dale_feat_0.png")
+    pathname = os.path.join(path, "example_2", "dale_feat_0.pdf")
     dale.plot(s=0, error="std", savefig=pathname)
-    pathname = os.path.join(path, "example_2", "dale_feat_1.png")
+    pathname = os.path.join(path, "example_2", "dale_feat_1.pdf")
     dale.plot(s=1, error="std", savefig=pathname)
-    pathname = os.path.join(path, "example_2", "dale_feat_2.png")
+    pathname = os.path.join(path, "example_2", "dale_feat_2.pdf")
     dale.plot(s=2, error="std", savefig=pathname)
 else:
     dale.plot(s=0, error="std")
@@ -64,11 +64,11 @@ pdp_ice = fe.PDPwithICE(data=X,
                         model=model.predict,
                         axis_limits=axis_limits)
 if savefig:
-    pathname = os.path.join(path, "example_2", "pdp_ice_feat_0.png")
+    pathname = os.path.join(path, "example_2", "pdp_ice_feat_0.pdf")
     pdp_ice.plot(s=0, normalized=True, nof_points=300, savefig=pathname)
-    pathname = os.path.join(path, "example_2", "pdp_ice_feat_1.png")
+    pathname = os.path.join(path, "example_2", "pdp_ice_feat_1.pdf")
     pdp_ice.plot(s=1, normalized=True, nof_points=300, savefig=pathname)
-    pathname = os.path.join(path, "example_2", "pdp_ice_feat_2.png")
+    pathname = os.path.join(path, "example_2", "pdp_ice_feat_2.pdf")
     pdp_ice.plot(s=2, normalized=True, nof_points=300, savefig=pathname)
 else:
     pdp_ice.plot(s=0, normalized=True, nof_points=300)

### 2022-09-22 examples/example_3.py

@@ -48,11 +48,11 @@ y, var, stderr = dale.eval(x=np.linspace(axis_limits[0,0], axis_limits[1,0], 100
                            s=0,
                            uncertainty=True)
 if savefig:
-    pathname = os.path.join(path, "example_3", "dale_feat_0.png")
+    pathname = os.path.join(path, "example_3", "dale_feat_0.pdf")
     dale.plot(s=0, error="std", savefig=pathname)
-    pathname = os.path.join(path, "example_3", "dale_feat_1.png")
+    pathname = os.path.join(path, "example_3", "dale_feat_1.pdf")
     dale.plot(s=1, error="std", savefig=pathname)
-    pathname = os.path.join(path, "example_3", "dale_feat_2.png")
+    pathname = os.path.join(path, "example_3", "dale_feat_2.pdf")
     dale.plot(s=2, error="std", savefig=pathname)
 else:
     dale.plot(s=0, error="std")
@@ -64,11 +64,11 @@ pdp_ice = fe.PDPwithICE(data=X,
                         model=model.predict,
                         axis_limits=axis_limits)
 if savefig:
-    pathname = os.path.join(path, "example_3", "pdp_ice_feat_0.png")
+    pathname = os.path.join(path, "example_3", "pdp_ice_feat_0.pdf")
     pdp_ice.plot(s=0, normalized=True, nof_points=300, savefig=pathname)
-    pathname = os.path.join(path, "example_3", "pdp_ice_feat_1.png")
+    pathname = os.path.join(path, "example_3", "pdp_ice_feat_1.pdf")
     pdp_ice.plot(s=1, normalized=True, nof_points=300, savefig=pathname)
-    pathname = os.path.join(path, "example_3", "pdp_ice_feat_2.png")
+    pathname = os.path.join(path, "example_3", "pdp_ice_feat_2.pdf")
     pdp_ice.plot(s=2, normalized=True, nof_points=300, savefig=pathname)
 else:
     pdp_ice.plot(s=0, normalized=True, nof_points=300)

### 2022-09-22 examples/example_bin_splitting_1.py

@@ -42,7 +42,7 @@ alg_params = {"bin_method" : "dp", "max_nof_bins" : 40, "min_points_per_bin": 10
 dale.fit(features=0, alg_params=alg_params)
 if savefig:
     path2dir = os.path.join(path, folder_name)
-    savepath = os.path.join(path2dir, "fig_1.jpg") if savefig else None
+    savepath = os.path.join(path2dir, "fig_1.pdf") if savefig else None
     dale.plot(savefig=savepath)
 else:
     dale.plot()
@@ -67,7 +67,7 @@ stats_auto = utils.measure_auto_error(dale_gt,
 
 # plots
 path2dir = os.path.join(path, folder_name)
-savepath = os.path.join(path2dir, "fig_2.jpg") if savefig else None
+savepath = os.path.join(path2dir, "fig_2.pdf") if savefig else None
 utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["mu_err_mean"],
                          stats_fixed["mu_err_std"],
@@ -76,7 +76,7 @@ utils.plot_fixed_vs_auto(K_list,
                          "mu error per bin",
                          savefig=savepath)
 
-savepath = os.path.join(path2dir, "fig_3.jpg") if savefig else None
+savepath = os.path.join(path2dir, "fig_3.pdf") if savefig else None
 utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["var_err_mean"],
                          stats_fixed["var_err_std"],
@@ -85,7 +85,7 @@ utils.plot_fixed_vs_auto(K_list,
                          "var error per bin",
                          savefig=savepath)
 
-savepath = os.path.join(path2dir, "fig_4.jpg") if savefig else None
+savepath = os.path.join(path2dir, "fig_4.pdf") if savefig else None
 utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["rho_mean"],
                          stats_fixed["rho_std"],

### 2022-09-22 examples/example_bin_splitting_2.py

@@ -38,7 +38,7 @@ alg_params = {"bin_method" : "dp", "max_nof_bins" : 40, "min_points_per_bin": 10
 dale.fit(features=0, alg_params=alg_params)
 if savefig:
     path2dir = os.path.join(path, folder_name)
-    savepath = os.path.join(path2dir, "fig_1.jpg") if savefig else None
+    savepath = os.path.join(path2dir, "fig_1.pdf") if savefig else None
     dale.plot(savefig=savepath)
 else:
     dale.plot()
@@ -62,7 +62,7 @@ stats_auto = utils.measure_auto_error(dale_gt,
 
 # plots
 path2dir = os.path.join(path, folder_name)
-savepath = os.path.join(path2dir, "fig_2.jpg") if savefig else None
+savepath = os.path.join(path2dir, "fig_2.pdf") if savefig else None
 utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["mu_err_mean"],
                          stats_fixed["mu_err_std"],
@@ -71,7 +71,7 @@ utils.plot_fixed_vs_auto(K_list,
                          "mu error per bin",
                          savefig=savepath)
 
-savepath = os.path.join(path2dir, "fig_3.jpg") if savefig else None
+savepath = os.path.join(path2dir, "fig_3.pdf") if savefig else None
 utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["var_err_mean"],
                          stats_fixed["var_err_std"],
@@ -80,7 +80,7 @@ utils.plot_fixed_vs_auto(K_list,
                          "var error per bin",
                          savefig=savepath)
 
-savepath = os.path.join(path2dir, "fig_4.jpg") if savefig else None
+savepath = os.path.join(path2dir, "fig_4.pdf") if savefig else None
 utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["rho_mean"],
                          stats_fixed["rho_std"],

### 2022-09-22 examples/utils.py

@@ -146,5 +146,5 @@ def plot_fixed_vs_auto(K_list, fixed_mean, fixed_std, auto_mean, auto_std, title
              label="auto-bin")
     plt.legend()
     if savefig is not None:
-        plt.savefig(savefig)
+        plt.savefig(savefig, bbox_inches="tight")
     plt.show(block=False)

### 2022-09-22 feature_effect/dale.py

@@ -107,7 +107,7 @@ class DALE(FeatureEffectBase):
             pass
 
     def fit_feature(self, s: int, alg_params: typing.Dict = None) -> typing.Dict:
-        alg_params = helpers.prep_dale_fit_params(alg_params)
+        alg_params = helpers.prep_dale_fit_bparams(alg_params)
 
         if self.data_effect is None:
             self.compile()

### 2022-09-22 feature_effect/visualization.py

@@ -2,6 +2,7 @@ import matplotlib.pyplot as plt
 import numpy as np
 from feature_effect.utils import compute_remaining_effect
 
+
 def feature_effect_plot(params, eval, feature, error, min_points_per_bin, title=None, block=False, gt=False, gt_bins=None, savefig=False):
     assert all(name in params for name in ["first_empty_bin", "limits", "dx", "is_bin_empty", "bin_estimator_variance", "bin_effect"])
     limits = params["limits"]
@@ -17,7 +18,7 @@ def feature_effect_plot(params, eval, feature, error, min_points_per_bin, title=
     else:
         first_empty_bin = None
 
-    x = np.linspace(params["limits"][0], params["limits"][-1], 10000)
+    x = np.linspace(params["limits"][0], params["limits"][-1], 1000)
     y, std, estimator_var = eval(x, feature, True)
     rem_eff = compute_remaining_effect(x, limits, np.sqrt(bin_variance), square=False)
 
@@ -35,6 +36,11 @@ def feature_effect_plot(params, eval, feature, error, min_points_per_bin, title=
     effects_per_bin(ax2, bin_effect, bin_variance, error, is_bin_empty,
                     limits, dx, gt_bins, min_points_per_bin)
 
+    ax1.set_ylabel("$y$")
+
+    ax2.set_xlabel("$x_{%d}$" % (feature+1))
+    ax2.set_ylabel("$\partial y / \partial x_{%d}$" % (feature+1))
+
     if savefig:
         plt.savefig(savefig, bbox_inches="tight")
 
@@ -46,7 +52,7 @@ def feature_effect_plot(params, eval, feature, error, min_points_per_bin, title=
 
 def feature_effect(ax1, x, y, estimator_var, std, rem_eff, first_empty, limits, point_limit, error=True, gt=None):
     # first subplot
-    ax1.plot(x, y, "b--", label="$f_{\mu}$")\
+    ax1.plot(x, y, "b--", label="$\hat{f}_{\mu}$")
 
     if first_empty is not None:
         added_line = .3*(np.max(y) - np.min(y))
@@ -55,7 +61,7 @@ def feature_effect(ax1, x, y, estimator_var, std, rem_eff, first_empty, limits, 
                    alpha=.7,
                    label="first bin with < " + str(point_limit) + " points")
     if error == "std":
[…]

### 2022-09-22 AISTATS2023/paper.tex
added content in Synthetic Examples

@@
 
-\subsection{Examples 2: Bin-Splitting}
+\subsection{Case 2: Bin-Splitting}
 \label{sec:simulation-examples-2}
 
-The aim of this example is to illustrate the advantages of automatic
-bin-splitting based on the objective of Eq.(\ref{eq:opt}). For
-evaluating our approach, we create the following validation framework;
-We generate a big dataset using dense sampling \(N=10^6\) and we treat
-as ground-truth the ALE estimation using dense fixed-size binning
-\(K=10^3\). Afterwards, we generate less samples \(X=500\) and we
-compare (a) the fixed-size ALE estimation for many different \(K\)
-versus (b) the auto-bin splitting algorithm. In all cases, the data
-generating distribution is \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
-\(x_1 \sim \mathcal{U}(0,1)\),
-\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=2)\).
+In this simulation, we aim to to quantify the advantages of automatic
+bin-splitting based on the objective of Eq.(\ref{eq:opt}), through the
+following validation framework. We generate a big dataset with dense
+sampling (\(N=10^6\)) and we treat the DALE estimation with dense
+fixed-size bins (\(K=10^3\)) as ground-truth. Afterwards, we generate
+less samples (\(N=500\)) and we compare (a) the fixed-size DALE
+estimation for many different \(K\) versus (b) the auto-bin splitting
+algorithm. In all cases, we use a bivariate black-box function
+\(f(\cdot)\) where the samples are instances of the distribution
+\(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
+\(x_1 \sim \mathcal{U}(0,1)\) and
+\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\).
 
@@
 \(\mathcal{Z^{\mathtt{DP}}} = \{z^{\mathtt{DP}}_{k-1}, \cdots,
-z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic-splitting
-based on the optimisation problem of Eq.~(\ref{eq:opt}) and with
-\(\mathcal{Z^{\mathtt{K}}}\}\) the corresponding fixed-sized
-splitting. The comparison is done using three metrics; (a)
+z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic
+bin-splitting based on the optimisation problem of Eq.~(\ref{eq:opt})
+and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size splitting with
+\(K\) bins. The evaluation is done with regard to two metrics;
 \(\mathcal{L}_{\mathtt{DP|K}}^{\mu} =
-\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k=1}^K | \mu_k -
-\hat{\mu}_k | \) quantifies the average error in estimating the the
-expected effect, (b)
+\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
+  \mathcal{Z}^{\mathtt{DP|K}}} | \mu_k - \hat{\mu}_k | \) quantifies
+the average error in estimating the the expected effect and
 \(\mathcal{L}_{\mathtt{DP|K}}^{\sigma^2} =
-\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k=1}^K | \sigma_k^2 -
-\hat{\sigma}^2_k | \) the average error in estimating the variance of
-the effect and (c)
+\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
+  \mathcal{Z}^{\mathtt{DP|K}}} | \sigma_k^2 - \hat{\sigma}^2_k | \)
+the average error in estimating the variance of the effect. The metric
 \(\mathcal{L}^{\rho^2}_{\mathtt{DP|K}} =
-\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k=1}^K \rho^2_k \) the
-average extra nuisance uncertainty.
+\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
+  \mathcal{Z}^{\mathtt{DP|K}}} \rho^2_k \) quantifies the average
+nuisance uncertainty. The optimal bin-splitting is the one that
+minimises, at the same time, both \(\mathcal{L}^{\mu}\) and
+\(\mathcal{L}^{\sigma^2}\) errors. For consistent results, in all the
+examples below, we regenerate samples and repeat the computations for
+\(t = 10\) times, providing the mean value for all metrics.
 
-\paragraph{Example 1.}
+\paragraph{Piecewise-Linear Function.}
 
-In this example, we defin a black-box function
-\(f_1(x_1) = a_1x_1 + x_1x_2\) with 6 piecewise-linear regions of
-different-size, i.e., \(a_1\) equals to \(=\{2, -2, 5, -10, 0.5\}\) in
-the regions defined by the sequence \(\{0, 0.2, 0.4, 0.45, 0.5,
-1\}\). The example highlights the importance of variable-size
-splitting. Given that the smalest intervals size is \(0.05\) any
-fixed-size splitting with less than \( \frac{1}{\Delta x} = 20\) bins,
-will mix together effects and increase the variance error
-\(\mathcal{L}_{\mathtt{K}}^{\sigma^2}\) due to the increase of the
-residual \(\mathcal{L}_{\mathtt{K}}^{\rho^2} > 0\). This is evident in
-figure. On the other hand, a large number of bins divides homogeneuos
-regions, i.e. region \([0.5, 1)\), into smaller ones with more
-unstable estimations which increases
-\(\mathcal{L}_{\mathtt{K}}^{\mu}\) Figure. Variable-size binning leads
-to the optimal solution. As we see in figure automatic-bin splitting
-small bins in areas with fine-grain effects and merges together
-regions with similar effects, leading to a near optimal bin splitting.
+In this example, we define \(f(\mathbf{x}) = a_1x_1 + x_1x_2\) with
+\(5\) piecewise-linear regions of different-size, i.e., \(a_1\) equals
+to \(=\{2, -2, 5, -10, 0.5\}\) in the intervals defined by the
+sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). As we observe, in
+Figure~\ref{fig:ex-synth-2-1}, NAME method extracts a sequence of
+intervals with better \(\mathcal{L}_{\mathtt{DP}}^{\mu}\) and
+\(\mathcal{L}_{\mathtt{DP}}^{\sigma^2}\) error compared to any
+fixed-size splitting. Analyzing the fixed-size errors helps us
+understand the importance of variable-size splitting. In
+Figure~\ref{fig:ex-synth-2-1}(b), we observe a positive trend between
+\(\mathcal{L}^{\mu}_{\mathtt{K}}\) and \(K\), concluding that bin
+effect estimation is more incosistent as \(\Delta x\) becomes smaller,
+due to less points contributing to each bin. The interpretation of
+variance error is slightly more complex. Given that the smallest
+interval is \(\Delta x = 0.05 \Rightarrow K = 20\) and all intervals
+are multiples of the smallest interval, any \(K\) that is not a
+multiple of \(20\) adds nuisance uncertainty
+\(\mathcal{L}^{\rho^2}_{\mathtt{K}}\) leading to a high variance error
+\(\mathcal{L}^{\sigma^2}_{\mathtt{K}}\). In these cases, the variance
+error reduces as \(K\) grows bigger because the length of the bins
+that lie in the limit between two piecewise linear regions becomes
+smaller. For \(K=\{20, 40, 60, 80, 100\}\) where
+\(\mathcal{L}^{\rho^2}_{\mathtt{K}} \approx 0\), we conclude the same
+as with with the mean effect error, i.e. the estimation becomes more
+incosistent as \(K\) grows larger.
+
+Variable-size extracts correctly the fine-grain bins, e.g., intervals
+\([0.4, 0.45]\), \([0.45, 0.5]\), and acts as a merging mechanism to
+create wider bins when the effect remains constant, e.g. interval
+\([0.5, 1]\), leading to an optimal solution.
 
@@
   \caption{Figure 1}
-  \label{fig:ex-synth-1-1}
+  \label{fig:ex-synth-2-1}
 \end{figure}
 
-\paragraph{Example 2.}
+\paragraph{Non-Linear Function.}
 
-The black-box function \(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\)
+In this example, we define a black-box function
+\(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), where the effect is
+non-linear in all the range of \(x\). This case has two
+specialties. First, there is no obvious advantage of variable-size
+versus fixed-size splitting, and, second, there is not an apriori
+optimal bin-size. Widening a bin will increase the resolution error
+\(\mathcal{L}^{\rho^2}\) and narrowing will make less robust. In
+Figure~\ref{fig:ex-synth-2-2}, we observe that automatic bin splitting
+finds a solution that compromises the conflicting objectives, i.e., it
+keeps as low as possible both the main effect
+\(\mathcal{L}^{\mu}_{\mathtt{K}}\) and the variance error
+\(\mathcal{L}^{\sigma^2}_{\mathtt{K}}\).
 
@@
   \caption{Figure 1}
-  \label{fig:ex-synth-1-1}
+  \label{fig:ex-synth-2-2}
 \end{figure}
+
+\paragraph{Discussion.}
 
@@
 
+\section{CONCLUSION}
+
 \subsubsection*{Acknowledgments}

### 2022-09-22 examples/example_bin_splitting_1.py

@@ -73,7 +73,7 @@ utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["mu_err_std"],
                          stats_auto["mu_err_mean"],
                          stats_auto["mu_err_mean"],
-                         "mu error per bin",
+                         "mu",
                          savefig=savepath)
 
 savepath = os.path.join(path2dir, "fig_3.pdf") if savefig else None
@@ -82,7 +82,7 @@ utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["var_err_std"],
                          stats_auto["var_err_mean"],
                          stats_auto["var_err_mean"],
-                         "var error per bin",
+                         "var",
                          savefig=savepath)
 
 savepath = os.path.join(path2dir, "fig_4.pdf") if savefig else None
@@ -91,5 +91,5 @@ utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["rho_std"],
                          stats_auto["rho_mean"],
                          stats_auto["rho_mean"],
-                         "rho per bin",
+                         "rho",
                          savefig=savepath)

### 2022-09-22 examples/example_bin_splitting_2.py

@@ -68,7 +68,7 @@ utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["mu_err_std"],
                          stats_auto["mu_err_mean"],
                          stats_auto["mu_err_mean"],
-                         "mu error per bin",
+                         "mu",
                          savefig=savepath)
 
 savepath = os.path.join(path2dir, "fig_3.pdf") if savefig else None
@@ -77,7 +77,7 @@ utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["var_err_std"],
                          stats_auto["var_err_mean"],
                          stats_auto["var_err_mean"],
-                         "var error per bin",
+                         "var",
                          savefig=savepath)
 
 savepath = os.path.join(path2dir, "fig_4.pdf") if savefig else None
@@ -86,5 +86,5 @@ utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["rho_std"],
                          stats_auto["rho_mean"],
                          stats_auto["rho_mean"],
-                         "rho per bin",
+                         "rho",
                          savefig=savepath)

### 2022-09-22 examples/utils.py

@@ -136,14 +136,27 @@ def measure_auto_error(dale_gt, gen_dist, model, axis_limits, nof_iterations, no
     return stats
 
 
-def plot_fixed_vs_auto(K_list, fixed_mean, fixed_std, auto_mean, auto_std, title, savefig=None):
+def plot_fixed_vs_auto(K_list, fixed_mean, fixed_std, auto_mean, auto_std, metric, savefig=None):
     plt.figure()
-    plt.title(title)
-    plt.plot(K_list, fixed_mean, 'x', label="fixed-bin")
+    assert metric in ["rho", "mu", "var"]
+    if metric == "rho":
+        pass
+        plt.ylabel(r"$\log \mathcal{L}_{\mathtt{K}}^{\rho^2}$")
+        plt.title(r"Log of Nuisance Uncertainty ($\log \mathcal{L}_{\mathtt{K}}^{\rho^2}$)")
+    elif metric == "var":
+        plt.ylabel("$\log \mathcal{L}_{\mathtt{K}}^{\sigma^2}$")
+        plt.title("Log Error in Uncertainty ($\log \mathcal{L}_{\mathtt{K}}^{\sigma^2}$)")
+    elif metric == "mu":
+        plt.ylabel("$\log \mathcal{L}_{\mathtt{K}}^{\mu}$")
+        plt.title("Log Error in Mean Effect ($\log \mathcal{L}_{\mathtt{K}}^{\mu}$)")
+
+    plt.plot(K_list, np.log(fixed_mean), 'x', label="fixed-bin")
     plt.plot(K_list,
-             np.repeat(auto_mean, len(K_list)),
+             np.log(np.repeat(auto_mean, len(K_list))),
              "r--",
              label="auto-bin")
+    plt.xlabel("$K$")
+
     plt.legend()
     if savefig is not None:
         plt.savefig(savefig, bbox_inches="tight")

### 2022-09-22 feature_effect/dale.py

@@ -107,7 +107,7 @@ class DALE(FeatureEffectBase):
             pass
 
     def fit_feature(self, s: int, alg_params: typing.Dict = None) -> typing.Dict:
-        alg_params = helpers.prep_dale_fit_bparams(alg_params)
+        alg_params = helpers.prep_dale_fit_params(alg_params)
 
         if self.data_effect is None:
             self.compile()

### 2022-09-25 AISTATS2023/paper.tex
updated method description

@@
 
-\newcommand{\dfdx}{\frac{\partial f}{\partial x_s}}
-\newcommand{\xc}{\mathbf{x_c}}
+\newcommand{\dfdx}{f^s}
+\newcommand{\xc}{\mathbf{x}_c}
 \newcommand{\DY}{\mathbf{\Delta Y}}
@@
 Recently, ML has flourished in critical domains, such as healthcare
-and finance. In these areas, we need ML system with the capability to
-explain their predictions, apart from predicting with accuracy. For
+and finance. In these areas, we need a combination of accurate
+predictions along with meaningful explanations to support them. For
 this reason there is an increased interest in Explainable AI (XAI),
@@
 behavior. Global methods provide a universal explanation, summarizing
-the numerous local ones into a single interpretable outcome, usually a
-number or a plot. If a user wants to get a rough overview about which
-features are significant (feature importance) or whether a particular
-feature has a positive or negative effect on the output (feature
-effect), they should opt for a global explainability technique. On
-the other hand, aggregating the individual explanations for producing
-a concise global one is vulnerable to misinterpretations; Under strong
-feature interactions, the global explanation may obfuscate
-heterogeneous effects~\citep{herbinger2022repid} that exist under the
-hood; a phenomenon called aggregation bias~\citep{mehrabi2021survey}.
+the various local explanations into a single interpretable outcome,
+usually a number or a plot. If a user wants to get a rough overview
+about which features are significant (feature importance) or whether a
+particular feature has a positive or negative effect on the output
+(feature effect), they should opt for a global explainability
+technique. On the other hand, aggregating the individual explanations
+for producing a concise global one is vulnerable to
+misinterpretations. Under strong interactions and correlations between
+features, the global explanation may obfuscate heterogeneous effects
+that exist under the hood~\citep{herbinger2022repid}; a phenomenon
+called aggregation bias~\citep{mehrabi2021survey}.
 
@@
 
-However, ALE faces two crucial drawbacks. First, it does not inform
-the user about potential heterogeneous effects hidden behind the
-average effect. In contrast, in the case of PDP, the heterogeneous
-effects can be spotted by exploring the Individual Conditional
-Expectations (ICE)\citep{goldstein2015peeking}. Second, ALE
-approximation from the samples of the training set requires
-partitioning the axis of the feature of interest in a sequence of
-non-overlapping intervals and estimating a single effect from the
-population of samples in each interval. We refer to this step using
-the term \textit{bin-splitting problem}. Deciding on the sequence of
-intervals has important implications because ALE's interpretation in
-cases of correlated features is meaningful only inside each interval
-\citep{molnar2022}. However, this crucial step has not been given the
-appropriate attention, making the approximation vulnerable to
-potential misinterpretations.
+However, ALE faces two crucial limitations, the first concerns ALE
+definition and the second one the ALE approximation. Regarding the
+definition, i.e., the formulation of the feature effect, ALE does not
+inform the user about the level of heterogeneous effects hidden behind
+the average effect (global explanation). In contrast, in the case of
+PDP, the heterogeneous effects can be (partially) spotted by exploring
+the Individual Conditional Expectations
+(ICE)\citep{goldstein2015peeking}. Regarding the approximation,
+e.g.~the estimation of ALE from the limited samples of the training
+set, ALE requires an important additional step, called
+\textit{bin-splitting}. Bin-splitting consists of partitioning the
+axis of the feature of interest in a sequence of non-overlapping
+intervals and estimating a single effect from the population of
+samples in each interval. Specifying an approriate sequence of
+intervals is of particular importance, since ALE's interpretation is
+meaningful only inside each interval \citep{molnar2022}. However, this
+crucial step has not raised the appropriate attention, making the
+approximation vulnerable to potential misinterpretations.
 
@@
 measuring the uncertainty of the explanation. The uncertainty of the
-global explanation expresses how certain we are that the averaged
-explanation is valid if applied to an instance drawn at random and
-reveals the heterogeneous effects. Given that ALE's interpretation
-(expected effect and uncertainty) is meaningful only inside each
-interval, we design a principled framework, where we treat the
-bin-splitting step as a data-driven clustering problem, searching for
-the optimal splitting given available instances of the training
+global explanation, i.e.~how certain we are that the averaged
+explanation is valid if applied to an instance drawn at random,
+quantifies the level of heterogeneous effects. Given that ALE's
+interpretation (expected effect and uncertainty) is meaningful only
+inside each interval, we design a principled framework, where we treat
+the bin-splitting step as a data-driven clustering problem, searching
+for the optimal splitting given available instances of the training
 set. We, also, present a computationally-grounded algorithm for
@@
 
+At Section~\ref{sec:feat-effect-meth}, we present the basic Feature
+Effect (FE) methods and we review the ALE
+definition\citep{apley2020visualizing}. At
+Section~\ref{sec:quant-heter-effects}, we present the techniques for
+quantifying the heterogeneous effects and we discuss some of their
+limitations. Finally, at Section~\ref{sec:ale-approximation}, we
+review the ALE approximation\citep{apley2020visualizing}(add us),
+describing why it is vulnerable to a poor approximation due to
+inappropriate bin-splitting.
+
 \paragraph{Notation.} We refer to random variables (rv) using
-uppercase \( X \), whereas to simple variables with plain lowercase
-\( x \). Bold denotes a vector; \( \xb \) for simple variables or
-\(\mathbf{X}\) for rvs. Often, we partition the input vector
-\(\xb \in \mathbb{R}^D\) to the feature of interest
+uppercase \( X \), to simple variables with plain lowercase \( x \)
+and to vectors with bold \( \xb \). Often, we partition the input
+vector \(\xb \in \mathbb{R}^D\) to the feature of interest
 \(x_s \in \mathbb{R} \) and the rest of the features
 \(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
-\((x_s, \mathbf{x}_c)\), but we clarify that it corresponds to the
-vector \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote
-the corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
-function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
-FE of the \(s\)-th feature is
-\(f^{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
-name of the FE method.\footnote{An extensive list of all
-  symbols used in the paper is provided in the helping material.}
+\((x_s, \mathbf{x}_c)\), but we clarify that it implies the vector
+\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
+corresponding rv as \(\mathbf{X} = (X_s, \mathbf{X}_c)\). The
+black-box function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and
+the FE of the \(s\)-th feature is \(f^{\mathtt{<method>}}(x_s)\),
+where \(\mathtt{<method>}\) is the name of the FE method.\footnote{An
+  extensive list of all symbols used in the paper is provided in the
+  helping material.}
 
-\paragraph{Feature Effect Methods.} The three well-known feature
-effect methods are: PDP, MP and ALE. PDP formulates the FE
-of the \(s\)-th attribute as an expectation over the marginal
-distribution \(\mathbf{X}_c\), i.e.,
+\subsection{Feature Effect Methods And ALE Definition.}
+\label{sec:feat-effect-meth}
+
+The three well-known feature effect methods are: PDP, MP and ALE. PDP
+formulates the FE of the \(s\)-th attribute as an expectation over the
+marginal distribution \(\mathbf{X}_c\), i.e.,
 \(f^{\mathtt{PDP}}(x_s) =
@@
 \(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
-\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as the
-accumulation of the averaged local effects:
+\mathbf{X}_c)]\). ALE defines the local effect of the \(s\)-th feature
+at point \(x_s = z\) as \(f^s(z, \xc) = \dfdx (z, \xc)\). All the
+local explanations at \(z\) are, then, weighted by the conditional
+distribution \(p(\xc|x_s=z)\) and are averaged, to produce the
+averaged effect \(\mu(z)\). Finally, ALE is the accumulation of the
+averaged local effects:
+
 \begin{equation}
-  \label{eq:ALE_accumulated_mean}
-  f^{\mathtt{ALE}}(x_s) = \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
+  \label{eq:ALE}
+  f^{\mathtt{ALE}}(x_s) = \int_{x_{s,min}}^{x_s} \underbrace{\mathbb{E}_{\Xcb|X_s=z}\left [f^s (z, \Xcb)\right ]}_{\mu(z)} \partial z
 \end{equation}
-%
+
 ALE has specific advantages which gain particular value in cases of
 correlated input features. In these cases, PDP integrates over
-unrealistic instances, due to the use of the marginal distribution
+unrealistic instances due to the use of the marginal distribution
 \(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
 the combined effect of sets of features to a single feature. ALE
-manages to resolve both issues, and is therefore a trust able FE method
-in cases of correlated features.
+manages to resolve both issues, and is therefore the most trustable
+method in cases of correlated features.
 
-\paragraph{Quantify the Heterogeneous Effects.}
+\subsection{Quantification Of Heterogeneous Effects.}
+\label{sec:quant-heter-effects}
 
-FE methods answer the question \textit{what is expected to happen to
-  the output (expected effect), if the value of a specific feature is
-  increased/decreased}. Having an answer to the question above, it
-comes naturally to also wonder \textit{how certain we are about the
+FE methods pose the question \textit{what is expected to happen to the
+  output (expected effect), if the value of a specific feature is
+  increased/decreased}. Having quantified the expected effect, it
+comes naturally to also ask \textit{how certain we are about the
   expected change (uncertainty)}. For this reason, the quantification
-of the uncertainty along with the expected effect has attracted a lot
-of interest. The level of uncertainty is mostly quantified by
-measuring the existence of heterogeneous effects, i.e. whether there
-are local explanations that deviate from the expected global
-effect. ICE and d-ICE plots provide a visual understanding of the
-heterogeneous effects on top of PDPs. Another approach targets on
-grouping the heterogeneous effects, e.g., allocating ICE plots in
-homogeneous clusters, by dividing the input
-space.\citep{molnar2020model} Some other approaches, like H-Statistic,
-Greenwel, move a step behind and try to quantify the level of
-interaction between the input features, a possible cause of
-heterogeneous effects. In this case, the interpretation is indirect,
-since a strong interaction index is only an indicator of heterogeneous
-effects. The aforementioned approaches are under two limitations; They
-either do not quantify the uncertainty of the FE directly or they are
-based on PDPs, and, therefore, they are subject to the failure modes
-of PDPs in cases of correlated features. To the best of our knowledge,
-no method so far targets on quantifying the uncertainty of ALE.
+of the uncertainty, whether there are local explanations that deviate
+from the expected global effect, has attracted a lot of interest.
 
-\paragraph{Bin-Splitting for ALE estimation.}
+ICE and d-ICE\citep{goldstein2015peeking} provide a set of curves that
+are produce a plot plot on top-of PDP. Both methods produce one curve
+for each instance of the training set, i.e. ICE the instance effect
+\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) and d-ICE the derivative
+effect, i.e.
+\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
+\xc^i)\). The user can visually examine the level of homogeneity of
+the set of curves as an indicator of the level of heterogeneous
+effects. Some methods try to automate the aforementioned visual
+exploration, by grouping (d-)ICE into clusters~\citep{molnar2020model,
+  herbinger2022repid}. Some other approaches, like H-Statistic,
+Greenweal, quantify the level of interaction between the input
+features. A strong interaction index is an indirect indicator for the
+existence of heterogeneous effects.
+
+The aforementioned approaches are under two limitations; They either
+do not quantify the uncertainty of the FE directly or they are based
+on PDPs, and, therefore, they are subject to the failure modes of PDPs
+in cases of correlated features, as we will analyze later. To the best
+of our knowledge, no method so far quantifies the heterogeneous
+effects on-top of ALE.
+
+\subsection{ALE Approximation.}
+\label{sec:ale-approximation}
 
 In real ML scenarios, the FE is estimated from the limited instances
-of the training set.  \citep{apley2020visualizing} proposed dividing
-the \(s\)-th axis in \(K\) bins and estimating the local effects in
-each bin by evaluating the black box-function at the bin limits:
+of the training set. Therefore, \citep{apley2020visualizing} proposed
+dividing the \(s\)-th axis in \(K\) bins and estimating the local
+effects in each bin by evaluating the black box-function at the bin
+limits:
 \begin{equation}
@@
 \( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
-\). Afterwards, (cite) proposed the Differential ALE (DALE) for
-computing the local effects on the training instances using auto-differentiation:
+\). Afterwards, (cite) proposed the Differential ALE (DALE)
+approximation for computing the local effects on the training
+instances using auto-differentiation:
 \begin{equation}
@@
 the accumulated effect with different bin-splitting with near-zero
-computational cost. However, none of the approximations above deals
-with the crucial problem of bin-splitting. As indicated
-by~\citep{molnar2022}, in ALE the effects are computed per interval
-(region) and the interpretation of the effect can only be
-local.
+computational cost.
+
+Both approximations ask from the user to blindly decide the number of
+bins \(K\) for splitting the axis into K equally-sized bins, an
+approach with crucial limitations. Firstly, it is vulnerable to poor
+bin-effect approximations that lead to a misleading ALE plot. On the
+one hand, setting a \(K\) to a small value may hide fine-grain effects
+due to large bins and setting \(K\) to a high value leads to
+estimations of bin-effect from very limited samples. In general, the
+user may face contradictory explanations for differnt \(K\) without a
+clue to decide which one to trust. Secondly, as indicated
+by~\citep{molnar2022}, in ALE the interpretation of the effect can
+only be local. Preseting a fixed-size intervals, does not permit the
+extraction of local regions (intervals) ng of data-driven local
+regions
 
@@
 
-The NAME method extends ALE for measuring the uncertainty of the
-expected effect and automatically extracts regions with homogeneous
-effects. In Section~\ref{sec:NAME-definition} we define the component
-for the uncertainty quantification. In
-Section~\ref{sec:interval-based-estimation}, we show how to estimate
-the average effect and the uncertainty on a specific interval from the
-limited samples of the training set. We, also, make an important proof
-about the aggregated variance defined over a bin. In
-Section~\ref{sec:interval-spliting}, we define and solve the problem
-of optimal bin-splitting for automatically extract regions with
-homogeneous effects. Finally, in Section~\ref{sec:visualization}, we
-illustrate the appropriate visualization of NAME for facilitating its
-interpretation by a non-expert and we discuss important aspects of the
-method.
+The NAME method extends ALE for quantifying the level of heterogeneous
+effects (uncertainty) and automating the bin-splitting step for robust
+estimations. At Section~\ref{sec:NAME-definition}, we define NAME,
+presenting the component that extends ALE with uncertainty
+quantification. In Section~\ref{sec:interval-based-estimation}, we
+provide an alternative definition of NAME based on variable-size bins,
+providing important remarks and proofs for connecting it with the
+initial definition. In Section~\ref{sec:interval-spliting}, we define
+the problem of optimal bin-splitting, i.e. how to optimally
+approximate METHOD based on limited samples, and we provide an
+algorithmic solution to the problem. Finally, in
+Section~\ref{sec:visualization}, we illustrate the appropriate
+visualization of NAME for facilitating its interpretation by a
+non-expert and we discuss important aspects of the method.
 
-\subsection{Uncertainty Quantification}
+\subsection{NAME: ALE With Uncertainty Quantification}
 \label{sec:NAME-definition}
 
-ALE defines the local effect of the \(s\)-th feature at point
-\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local explanations at
-\(x_s\) are, then, weighted by the conditional distribution
-\(p(\xc|x_s)\) and are averaged, to produce the summarized effect at
-\(x_s\):
-\begin{equation}
-  \label{eq:ALE_mean}
-  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
-\end{equation}
-%
-The FE at \(x_s\) is the accumulation of the averaged local effects
-from \(x_{s, min}\)\footnote{\(x_{s,min}\) is the minimum value the
-  \(s\)-th attribute can take. In estimations, it is the minimum
-  \(s\)-th attribute value among the training set instances.} until
-\(x_s\), i.e.
-\(f^{\mathtt{ALE}}(x_s) = \int_{x_{s, min}}^{x_s} \mu(z) \partial z
-\). Limiting the explanation to the expected value level does not
-shed light to potential heterogeneous effects behind the averaged
-explanation. Therefore, we model the uncertainty of the local effects
-\(\mathcal{H}(x_s)\) with the variance of the local explanations:
+ALE does not shed light to potential heterogeneous effects that
+contribute to the average explanation. Therefore, we extend ALE
+definition of Eq.~(\ref{eq:ALE}) with a component that quantifies the
+level of heterogeneous effects. We denote \(\mathcal{H}(z)\) the
+uncertainty of the local effects at a specific point \(x_s=z\) and we
+quantify it as the standard deviation of the local explanations:
 \begin{equation}
   \label{eq:ALE_var}
-  \mathcal{H}(x_s) := \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
+  \mathcal{H}(z) := \sigma(z) = \sqrt{\mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] }
 \end{equation}
 \noindent
-The uncertainty of the explanation emerges from the natural
-characteristics of the experiment, i.e.,~the data generating
-distribution and the properties of the black-box function. In
-Section~\ref{sec:visualization}, we propose appropriate visualizations
-for easier interpretation of Eq.~(\ref{eq:ALE_var}). ALE defines the
-FE at \(x_s\) as the accumulation of the averaged local effects from
-\(x_{min}\) until \(x_s\). Equivalently, we define the accumulated
-uncertainty (variance) until \(x_s\), as the integral of the variances
-of the local effects:
+The uncertainty emerges from the natural characteristics of the
+experiment, i.e.,~the feature correlations existent in the data
+generating distribution and the implicit interactions of the black-box
+function. We also define the accumulated uncertainty at \(x_s\), as
+the accumulation of the standard deviation of the local effects along
+the axis:
 \begin{equation}
-  \label{eq:ALE_accumulated_var}
-  f^{\mathtt{ALE}}_{\sigma^2}(x_s) = \int_{z_{s, min}}^{x_s} \sigma^2(z) \partial z
+  \label{eq:ALE_acc_unc}
+  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
 \end{equation}
 \noindent
-The accumulated uncertainty is not a directly interpretable quantity,
-since it measures the total uncertainty that characterizes the
-\(s\)-th feature effect plot. As stated in the Introduction, ALE
-outcomes should be interpreted at a local level, therefore, the only
-\(\mathcal{H}(x_s)\) is the directly interpretable quantity. However,
-the accumulated uncertainty will help us define a sensible objective
-for the bin-splitting step, as we will discuss in
-Section~\ref{sec:interval-spliting}.
+NAME method formulates the effect at a specific point \(x_s\) with a
+compact entity that consists of the average effect and the uncertainty
+\(\mu(z) \pm \sigma(z)\) and visualizes them as a continuous curve
+with a confidence region, i.e.
+\(f^{\mathtt{METHOD}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
+f^{\mathtt{ALE}}_{\sigma}(x_s)\). In Section~\ref{sec:visualization},
+we propose an appropriate visualization for easier interpretation of
+NAME.
 
-\subsection{Interval-Based Estimation}
+\subsection{Interval-Based Formulation of NAME}
 \label{sec:interval-based-estimation}
 
-In real scenarios, we have ignorance about the data-generating
-distribution \(p(x_s, \mathbf{x}_c)\), so, the estimations are based
-on the limited instances of the training set. Estimating
-Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
-point is impossible, because the probability of observing a sample
-inside the interval \([x_s - h, x_s + h]\) tends to zero, when
-\(h \to 0\). We are, therefore, obliged to split the axis of \(x_s\)
-into a sequence of non-overlapping intervals (bins) and estimate the
-mean and the variance from the samples that lie inside each bin. The
-mean effect at an interval \([z_1, z_2)\) is defined as the mean of
-the expected effects:
+In real scenarios, the estimations are based on the limited instances
+of the training set. Estimating \(\mu(z), \sigma(z)\) at the
+granularity of a point is impossible, because the probability of
+observing a sample inside the interval \([x_s - h, x_s + h]\) tends to
+zero, when \(h \to 0\). For this reason, we firstly define the mean
+effect (bin-effect) and the uncertainty (bin-uncertainty) inside an interval \([z_1, z_2)\):
+
 \begin{equation}
@@
   \mu(z_1, z_2) = \frac{1}{z_2 - z_1} \int_{z_1}^{z_2}
-  \mathbb{E}_{\xc|x_s=z}\left [\frac{\partial f}{\partial x_s} \right ] \partial z
+  \mathbb{E}_{\xc|z}\left [f^s(z, \xc) \right ] \partial z
 \end{equation}
+
 \noindent
-Similarly, the accumulated variance at an interval \([z_1, z_2)\) is
-defined as:
+
 \begin{equation}
   \label{eq:var_bin}
-  \sigma^2(z_1, z_2) = \int_{z_1}^{z_2}
+  \sigma(z_1, z_2) = \frac{\int_{z_1}^{z_2}
   \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
-      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z
+      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 - z_1}
 \end{equation}
-\paragraph{Theorem 1.} If we define the residual \(\rho(z)\) as the
+
+%
+We also define as \(\mathcal{Z}_s\) a sequence of \(K+1\) points that
+partitions the \(s\)-th feature axis into a sequence \(K\)
+variable-size intervals, i.e. \(\mathcal{Z}_s = \{z_0, z_K\}\), where
+\(z_0=x_{s,min}\) and \(z_K=x_{s,max}\). Based on the above, we
+redefine NAME using an interval-based formulation
+\(\tilde{f}^{\mathtt{METHOD}}(x_s):=
+\tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) \pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:
+
+\begin{equation}
+  \label{eq:ALE_2}
+  \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
+\end{equation}
+
+\begin{equation}
+  \label{eq:ALE_accumulated_var}
+  \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
+\end{equation}
+%
+\paragraph{Theorem 1.}
+\label{sec:theorem-1}
+
+If we define the residual \(\rho(z)\) as the
 difference between the expected effect at \(x_s\) and the mean
@@
 %
-The proof is in the Appendix. Theorem 1 decouples the total
-uncertainty inside a bin
-\(\mathcal{H}_{bin}(z_1, z_2) := \sigma^2(z_1, z_2)\) into two terms. The first term quantifies
-the accumulated uncertainty due to the natural characteristics of the
-experiment, i.e.,
+The proof is in the Appendix.
+
+\paragraph{Theorem 2.}
+\label{sec:theorem-2}
+
+The two definitions of NAME are equivalent, i.e.,
+\(f^{\mathtt{ALE}}_{\mu}(x_s) = \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)\)
+and
+\(f^{\mathtt{ALE}}_{\sigma}(x_s) =
+\tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), if and only if the sequence
+of limits \(\mathcal{Z}\) is such that
+\(\sum_{k=1}^K \rho^2(z_{k-1}, z_{k}) = 0\), where \(\rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial z\).
+
+\paragraph{Discussion}
+
+Theorem 1 decouples the aggregated uncertainty inside a bin, i.e.,
+\(\mathcal{H}_{bin}(z_1, z_2) := \sigma^2(z_1, z_2)\), into two
+terms. The first term quantifies the accumulated uncertainty due to
+the natural characteristics of the experiment, i.e.,
 \(\mathcal{H}(z_1, z_2) = \int_{z_1}^{z_2}\mathcal{H}(z) \partial z\),
@@
 \(\mathcal{H}_n(z_1, z_2) := \int_{z_1}^{z_2}\rho^2(z) \partial z
-\). In other words, the creation of a bin burdens the estimation with
-a nuisance uncertainty:
+\). Therefore, it holds that:
 \begin{equation}
@@
 \end{equation}
-Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be directly estimated
-from the set \(\mathcal{S}\) of the dataset instances with the
-\(s\)-th feature lying inside the interval, i.e.,
-\( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \} \). The mean
-effect at the interval, Eq.~(\ref{eq:mu_bin}) is approximated by:
-\begin{equation}
-  \label{eq:mean_estimation}
-  \hat{\mu}(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
-    \mathcal{S}} \left [ \dfdx(\mathbf{x}^i) \right ]
-\end{equation}
-
-and the accumulated variance, Eq.~(\ref{eq:var_bin}) can be
-approximated by
-\begin{equation}
-  \label{eq:variance_estimation}
-  \hat{\sigma}^2(z_1, z_2) = \frac{z_2 - z_1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
-    \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}(z_1, z_2) \right )^2
-\end{equation}
-The approximation is unbiased only if the points are uniformly
-distributed in \([z_1, z_2]\). Elaborate.
+%
+In Theorem 2 we prove that if we find a sequence of bins, where each
+bin adds no nuisance uncertainty, then we will get exactly the same
+NAME explanation as in the original definition.
 
@@
 
-As we discussed in the Introduction, when the features are (strongly)
-correlated, ALE provides meaningful interpretations only inside a
-bin. The accumulation of the mean effects inside each bin is done only
-for a producing smooth curve for easier visual exploration. Therefore,
-the extraction of regions with common effects is particularly
-important for a correct interpretation.
-
-We formulate the bin-splitting as an unsupervised clustering
-problem. We want to partition the axis of the \(s\)-th feature into a
-sequence of variable-size non-overlapping intervals, so that the
-regions that will be created to provide meaningful explanations. A
-region with a meaningful explanation means that it clusters together
-points with similar expected (averaged) effects. On the other hand,
-since we estimate all the quantities from the limited samples of the
-training set, we need enough samples inside each bin for robustly
-estimate the expected effect and the variance. Therefore, we transform
-the bin-splitting problem to the following optimization problem:
+In this section we formulate bin-splitting as an unsupervised
+clustering problem. The mean effect (Eq.~\eqref{eq:mu_bin}) and the
+uncertainty (Eq.~\eqref{eq:var_bin}) can be directly estimated from
+the set \(\mathcal{S}\) of the dataset instances with the \(s\)-th
+feature lying inside the interval
+\( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \} \), i.e.
+\(\hat{\mu}(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i
+  \in \mathcal{S}} \left [ \dfdx(\mathbf{x}^i) \right ] \) and
+\( \hat{\sigma}^2(z_1, z_2) = \frac{1}{|\mathcal{S}|}
+\sum_{i:\mathbf{x}^i \in \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) -
+  \hat{\mu}(z_1, z_2) \right )^2 \). Both approximations are unbiased
+only under the assumption that the points are uniformly distributed in
+\([z_1, z_2]\). Under this assumption, and due to the law of large
+numbers, the approximations become more robust, as the bins become
+wider. On the other hand, bins should group local explanations with
+similar effects and enlarging a bin is under the danger of adding
+nuisance uncertainty. Therefore, we set-up the following optimization
+problem:
 \begin{equation}
@@
 \begin{aligned}
-  \min_{ \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \\
-  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N\\
+  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
+  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
+  & \tau_k = 1 - 0.2 \frac{|S_k|}{N} \\
+  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                      & z_0 = x_{s,min}\\
@@
 %
-We search for the sequence of intervals \(z_0, \ldots, z_K\) covering
-the axis of the \(s\)-th feature, \(z_0 = x_{s,min}, z_K = x_{s,max}\)
-that minimize the sum of bin variances. The constraint of at least
-\(N\) points in each bin sets the limit for the minimum number of
-points inside each bin, to robustly estimate the mean effect and the
-variance. There is no constraint about the bin size and the number of
-bins \(K\).
+We search for the sequence of intervals \(z_0, \ldots, z_K\) that
+minimizes the sum of the cost of each bin. The cost of each bin is the
+mean variance \(\sigma^2_k\) scaled by the bin length \(\Delta z_k\)
+and discounted by the term \(\tau_k\). The term
+\(0.8 \leq \tau_K \leq 1\) acts as a tie-braker in cases of two
+candidate bins with equal mean variance, favoring the selection of the
+bigger bin. The constraint of at least \(N_{\mathtt{PPB}}\) points per
+bin sets the lowest-limit for a \textit{robust} estimation.
 
-The objective function is the sum of the approximations of the bin
-variances. As we prove in Theorem 1, the observable bin variance can
-can be split in the accumulation of point-wise variances plus the
-residuals. In Corollary 1, we show that the accumulation of the
-point-wise variances is independent of the bin-splitting
-procedure. Therefore by minimizing \(L\), we minimize the accumulation
-of the point-wise residuals in each bin, searching for bins that
-cluster together points with a common expected effect. An additional
-interpretation is that we search for the bin-splitting that adds the
-minimum the nuisance uncertainty over the lower-bound of the
-uncertainty due to the natural characteristics of the experiment.
+According to Eq.~(\ref{eq:uncertainty_bin}), the uncertainty of a bin
+\(\mathcal{H}_{bin}\) is the sum of the uncertainty due to the
+characteristics of the experiment (data and black-box model)
+\(\mathcal{H}\) and the nuisance uncertainty \(\mathcal{H}_n\). Since
+\(\mathcal{H}\) is independent of the bin-spliting procedure, by
+minimizing \(\mathcal{L}\) we minimize the nuiscane uncertainty
+\(\mathcal{H}_n\) in each bin.
 
@@
 
-\clearpage
 \section{SIMULATION EXAMPLES}

### 2022-09-27 AISTATS2023/paper.tex
updated method description

@@
 definition and the second one the ALE approximation. Regarding the
-definition, i.e., the formulation of the feature effect, ALE does not
-inform the user about the level of heterogeneous effects hidden behind
-the average effect (global explanation). In contrast, in the case of
-PDP, the heterogeneous effects can be (partially) spotted by exploring
-the Individual Conditional Expectations
-(ICE)\citep{goldstein2015peeking}. Regarding the approximation,
-e.g.~the estimation of ALE from the limited samples of the training
-set, ALE requires an important additional step, called
+definition ALE does not inform the user about the level of
+heterogeneous effects hidden behind the average effect (global
+explanation). In contrast, in the case of PDP, the heterogeneous
+effects can be (partially) spotted by exploring the Individual
+Conditional Expectations (ICE)\citep{goldstein2015peeking}. Regarding
+the approximation, e.g.~the estimation of ALE from the limited samples
+of the training set, ALE requires an important additional step, called
 \textit{bin-splitting}. Bin-splitting consists of partitioning the
 axis of the feature of interest in a sequence of non-overlapping
-intervals and estimating a single effect from the population of
-samples in each interval. Specifying an approriate sequence of
-intervals is of particular importance, since ALE's interpretation is
-meaningful only inside each interval \citep{molnar2022}. However, this
-crucial step has not raised the appropriate attention, making the
-approximation vulnerable to potential misinterpretations.
+intervals and estimating a single effect in each one, from the
+population of samples that fall inside. Specifying an approriate
+sequence of intervals is of particular importance, since ALE's
+interpretation is meaningful only inside each interval
+\citep{molnar2022}. However, this crucial step has not raised the
+appropriate attention, making the approximation vulnerable to
+potential misinterpretations.
 
@@
 \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
-corresponding rv as \(\mathbf{X} = (X_s, \mathbf{X}_c)\). The
-black-box function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and
-the FE of the \(s\)-th feature is \(f^{\mathtt{<method>}}(x_s)\),
-where \(\mathtt{<method>}\) is the name of the FE method.\footnote{An
-  extensive list of all symbols used in the paper is provided in the
-  helping material.}
+corresponding rv as \(\mathbf{X} = (X_s, \mathbf{X}_c)\). When we
+refer to an instance of the training set, we use
+\(\xb^i= (\xc^i, x_s^i) \). The black-box function is
+\(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the FE of the \(s\)-th
+feature is \(f^{\mathtt{<method>}}(x_s)\), with \(\mathtt{<method>}\)
+indicating the particular method in use.\footnote{An extensive list of all
+  symbols used in the paper is provided at the Appendix.}
 
-\subsection{Feature Effect Methods And ALE Definition.}
+\subsection{Feature Effect Methods And ALE Definition}
 \label{sec:feat-effect-meth}
@@
 \(\mathbf{X}_c|X_s\), i.e.,
-\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
+\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|x_s}[f(x_s,
 \mathbf{X}_c)]\). ALE defines the local effect of the \(s\)-th feature
-at point \(x_s = z\) as \(f^s(z, \xc) = \dfdx (z, \xc)\). All the
+at point \(x_s = z\) as \(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
 local explanations at \(z\) are, then, weighted by the conditional
@@
 
-FE methods pose the question \textit{what is expected to happen to the
-  output (expected effect), if the value of a specific feature is
-  increased/decreased}. Having quantified the expected effect, it
-comes naturally to also ask \textit{how certain we are about the
-  expected change (uncertainty)}. For this reason, the quantification
-of the uncertainty, whether there are local explanations that deviate
-from the expected global effect, has attracted a lot of interest.
+FE methods pose the question \textit{what is average (global) effect
+  on the output, if the value of a specific feature is
+  increased/decreased}. After that, it comes naturally to also ask
+about \textit{to what extent the individual (local) effects deviate
+  from the average}. For this reason, the quantification of the
+uncertainty has attracted a lot of interest.
 
 ICE and d-ICE\citep{goldstein2015peeking} provide a set of curves that
-are produce a plot plot on top-of PDP. Both methods produce one curve
-for each instance of the training set, i.e. ICE the instance effect
-\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) and d-ICE the derivative
-effect, i.e.
+are plot on top-of the PDP. Both methods produce one curve for each
+instance of the training set;
+\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) for ICE and
 \(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
-\xc^i)\). The user can visually examine the level of homogeneity of
-the set of curves as an indicator of the level of heterogeneous
-effects. Some methods try to automate the aforementioned visual
+\xc^i)\) for d-ICE. The user then visually observes if the (d)-ICE
+curves are homogeneous and to what extent they deviate from the
+PDP. Some methods try to automate the aforementioned visual
 exploration, by grouping (d-)ICE into clusters~\citep{molnar2020model,
-  herbinger2022repid}. Some other approaches, like H-Statistic,
-Greenweal, quantify the level of interaction between the input
-features. A strong interaction index is an indirect indicator for the
-existence of heterogeneous effects.
+  herbinger2022repid, britton2019vine}. Some other approaches, like
+H-Statistic\citep{friedman2008predictive}, Greenwell's interaction
+index\citep{greenwell2018simple} or SHAP interaction
+values\citep{lundberg2018consistent}, quantify the level of
+interaction between the input features, with an interaction value. A
+strong interaction index is an indicator for the existence of
+heterogeneous effects.
 
@@
 on PDPs, and, therefore, they are subject to the failure modes of PDPs
-in cases of correlated features, as we will analyze later. To the best
-of our knowledge, no method so far quantifies the heterogeneous
-effects on-top of ALE.
+in cases of correlated features\citep{baniecki2021fooling}. To the
+best of our knowledge, no work exist so far that quantifies the
+heterogeneous effects based on the formulation of ALE.
 
@@
   \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
-    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
+    \mathcal{S}_k} f^s(\mathbf{x}^i)
 \end{equation}
@@
 Both approximations ask from the user to blindly decide the number of
-bins \(K\) for splitting the axis into K equally-sized bins, an
-approach with crucial limitations. Firstly, it is vulnerable to poor
-bin-effect approximations that lead to a misleading ALE plot. On the
-one hand, setting a \(K\) to a small value may hide fine-grain effects
-due to large bins and setting \(K\) to a high value leads to
-estimations of bin-effect from very limited samples. In general, the
-user may face contradictory explanations for differnt \(K\) without a
-clue to decide which one to trust. Secondly, as indicated
-by~\citep{molnar2022}, in ALE the interpretation of the effect can
-only be local. Preseting a fixed-size intervals, does not permit the
-extraction of local regions (intervals) ng of data-driven local
-regions
+bins \(K\) for splitting the axis into equally-sized bins, an approach
+with crucial limitations. Firstly, it is vulnerable to poor bin-effect
+approximations that lead to a misleading ALE plot. Setting \(K\) to a
+small value may hide fine-grain effects due to large bins and setting
+\(K\) to a high value leads to estimations of many bin-effects from
+very limited samples. In general, the user may face contradictory
+explanations for differnt \(K\) without a clue to decide which one to
+trust. Secondly, as indicated by~\citep{molnar2022}, the ALE's feature
+effect can only be interpreted in local regions. Therefore, for better
+interpretation, the extend of these regions should be determined
+in a data-driven way.
 
@@
 initial definition. In Section~\ref{sec:interval-spliting}, we define
-the problem of optimal bin-splitting, i.e. how to optimally
-approximate METHOD based on limited samples, and we provide an
-algorithmic solution to the problem. Finally, in
-Section~\ref{sec:visualization}, we illustrate the appropriate
-visualization of NAME for facilitating its interpretation by a
-non-expert and we discuss important aspects of the method.
+the problem of optimal bin-splitting and we provide an algorithmic
+solution to the problem.
 
@@
 \(f^{\mathtt{METHOD}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
-f^{\mathtt{ALE}}_{\sigma}(x_s)\). In Section~\ref{sec:visualization},
-we propose an appropriate visualization for easier interpretation of
-NAME.
+f^{\mathtt{ALE}}_{\sigma}(x_s)\).
 
@@
 observing a sample inside the interval \([x_s - h, x_s + h]\) tends to
-zero, when \(h \to 0\). For this reason, we firstly define the mean
-effect (bin-effect) and the uncertainty (bin-uncertainty) inside an interval \([z_1, z_2)\):
+zero, when \(h \to 0\). We define the mean effect (bin-effect) and the
+uncertainty (bin-uncertainty) inside an interval \([z_1, z_2)\) as:
 
@@
 %
-We also define as \(\mathcal{Z}_s\) a sequence of \(K+1\) points that
-partitions the \(s\)-th feature axis into a sequence \(K\)
+We also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points that
+partition the \(s\)-th feature axis into a sequence \(K\)
 variable-size intervals, i.e. \(\mathcal{Z}_s = \{z_0, z_K\}\), where
@@
 
-If we define the residual \(\rho(z)\) as the
+\textit{If we define the residual \(\rho(z)\) as the
 difference between the expected effect at \(x_s\) and the mean
@@
  \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
-\end{equation}
+\end{equation}}
 %
-The proof is in the Appendix.
 
@@
 
-The two definitions of NAME are equivalent, i.e.,
-\(f^{\mathtt{ALE}}_{\mu}(x_s) = \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)\)
-and
-\(f^{\mathtt{ALE}}_{\sigma}(x_s) =
-\tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), if and only if the sequence
-of limits \(\mathcal{Z}\) is such that
-\(\sum_{k=1}^K \rho^2(z_{k-1}, z_{k}) = 0\), where \(\rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial z\).
+\textit{If we define as \(\rho^2(z_{k-1}, z_{k})\) the accumulated residual of
+a bin, i.e.
+\(= \rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial
+z\), then the two definitions of NAME are equivalent, i.e.,
+\(f^{\mathtt{NAME}} = \tilde{f}^{\mathtt{NAME}}\), if and only if the
+sequence of limits \(\mathcal{Z}\) is such that
+\(\sum_{k=1}^K \rho^2(z_{k-1}, z_{k}) = 0\).}
 
@@
 uncertainty (Eq.~\eqref{eq:var_bin}) can be directly estimated from
-the set \(\mathcal{S}\) of the dataset instances with the \(s\)-th
+the set \(\mathcal{S}_k\) of the dataset instances with the \(s\)-th
 feature lying inside the interval
-\( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \} \), i.e.
-\(\hat{\mu}(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i
-  \in \mathcal{S}} \left [ \dfdx(\mathbf{x}^i) \right ] \) and
-\( \hat{\sigma}^2(z_1, z_2) = \frac{1}{|\mathcal{S}|}
-\sum_{i:\mathbf{x}^i \in \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) -
-  \hat{\mu}(z_1, z_2) \right )^2 \). Both approximations are unbiased
-only under the assumption that the points are uniformly distributed in
-\([z_1, z_2]\). Under this assumption, and due to the law of large
-numbers, the approximations become more robust, as the bins become
-wider. On the other hand, bins should group local explanations with
-similar effects and enlarging a bin is under the danger of adding
-nuisance uncertainty. Therefore, we set-up the following optimization
-problem:
+\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \}
+\):
+
+\begin{equation}
+  \label{eq:2}
+  \hat{\mu}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
+  \sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left [ \dfdx(\mathbf{x}^i)
+  \right ]
+\end{equation}
+%
+\begin{equation}
+  \label{eq:3}
+  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
+\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
+  \hat{\mu}(z_1, z_2) \right )^2
+\end{equation}
+
+Both approximations are unbiased only under the assumption that the
+points are uniformly distributed inside the intervals. Under this
+assumption, and due to the law of large numbers, the approximations
+become more robust, as the bins become wider. On the other hand, bins
+should group local explanations with similar effects and enlarging a
+bin is under the danger of adding nuisance uncertainty. Therefore, we
+set-up the following optimization problem:
 \begin{equation}
@@
 mean variance \(\sigma^2_k\) scaled by the bin length \(\Delta z_k\)
-and discounted by the term \(\tau_k\). The term
-\(0.8 \leq \tau_K \leq 1\) acts as a tie-braker in cases of two
-candidate bins with equal mean variance, favoring the selection of the
-bigger bin. The constraint of at least \(N_{\mathtt{PPB}}\) points per
-bin sets the lowest-limit for a \textit{robust} estimation.
+and discounted by the term \(\tau_k\). The term \(\tau_K\) acts as a
+tie-braker in cases of two candidate bins with equal mean variance,
+favoring the selection of the bigger bin. The constraint of at least
+\(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
+\textit{robust} estimation.
 
@@
 multiple of the minimum step
-\(u = \frac{x_{s, max} - x_{s, min}}{K}\). For defining the solution,
-we use two indexes. The index \(i \in \{0, \ldots, K_{max}\}\) denotes
-the \(i\)-th limit \((z_i)\) and the index
-\(j \in \{0, \ldots, K_{max}\} \) denotes the \(j\)-th position along
-the axis \((x_j = x_{s,min} + j \cdot u)\). The recursive cost
-function \(T(i,j)\) is the cost of setting the \(i\)-th limit \(z_i\)
-to \(x_j\):
+\(u = \frac{x_{s, max} - x_{s, min}}{K_{max}}\). For defining the
+solution, we use two indexes. The index
+\(i \in \{0, \ldots, K_{max}\}\) denotes the point \((z_i)\) and the
+index \(j \in \{0, \ldots, K_{max}\} \) denotes the position of the
+\(j\)-th multiple of the minimum step, i.e., 
+\(x_j = x_{s,min} + j \cdot u\). The recursive cost function
+\(T(i,j)\) is the cost of setting \(z_i=x_j\):
 \begin{equation}
@@
 
-% \begin{equation}
-%   \label{eq:1}
-%   \mathcal{L} = \sum_{k=0}^K \hat{\sigma}_k(z_{k-1}, z_k)
-% \end{equation}
-% %
-% The constraints are that all bins must include more than \(\tau\)
-% points, i.e., \(|\mathcal{S}_k| \geq \tau\).
-
-% \(\mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k)\) is the sum
-% of the uncertainty of the local explanations \(\mathcal{H}\) due to
-% heterogeneous effects and the added nuisance uncertainty due to
-% bin-splitting \(\mathcal{H}_n\). Given that \(\mathcal{H}\) is not
-% affected by interval splitting, optimizing \(\mathcal{L}\) equals to
-% finding the sequence of non-overlaping bins that add the minimum
-% nuisance uncertainty. The restriction of (\ref{eq:opt}) secures that
-% each bin is populated with enough samples for a robust estimation,
-% which is set by the parameter \(N\). In other words, the user makes
-% the following proposal; \textit{find the bin sequence that adds the
-%   minimum nuisance uncertainty, given that I want at least \(N\)
-%   points per bin}.
-
 \noindent
@@
 \begin{itemize}
-\item Computational complexity
-\end{itemize}
-
-\subsection{Visualization and Interpretation}
-\label{sec:visualization}
-
-\begin{itemize}
-\item Discuss about the meaning of ALE, to find intervals with some effect
+\item Discuss more aspects (e.g. Computational complexity)
 \end{itemize}
@@
 In this simulation, we will compare NAME method against PDP-ICE in
-quantifying the main effect \(f_\mu\) and the uncertainty
-\(f_\sigma^2\). Since there is ambiguity about the ground-truth
-first-order effect in cases of correlated
-features, e.g.~\citep{apley2020visualizing,Gromping2020MAEP}, we limit
-ourselves to the following piecewise linear function,
+quantifying the main effect and the uncertainty. Since there is
+ambiguity about the ground-truth first-order effect in cases of
+correlated features,
+e.g.~\citep{apley2020visualizing,Gromping2020MAEP}, we limit ourselves
+to the following piecewise linear function,
 
@@
 explanation is zero \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
-Figure~\ref{fig:ex-synth-1-1}, we observe that PDP main effect is
+Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP main effect is
 wrong and ICE plots imply the existence of heterogeneous effects. In
@@
 
-\paragraph{Discussion.}
+% \paragraph{Discussion.}
 
-In purpose, the examples above do not cover the case where there is an
-interaction term includes correlated features, i.e. \(x_1x_2\). In
-this scenario, there is an open debate about the ground-truth
-effect~\citep{Gromping2020MAEP}.
+% In purpose, the examples above do not cover the case where there is an
+% interaction term includes correlated features, i.e. \(x_1x_2\). In
+% this scenario, there is an open debate about the ground-truth
+% effect~\citep{Gromping2020MAEP}.
 
@@
 the average error in estimating the the expected effect and
-\(\mathcal{L}_{\mathtt{DP|K}}^{\sigma^2} =
+\(\mathcal{L}_{\mathtt{DP|K}}^{\sigma} =
 \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
-  \mathcal{Z}^{\mathtt{DP|K}}} | \sigma_k^2 - \hat{\sigma}^2_k | \)
+  \mathcal{Z}^{\mathtt{DP|K}}} | \sigma_k - \hat{\sigma}_k | \)
 the average error in estimating the variance of the effect. The metric
-\(\mathcal{L}^{\rho^2}_{\mathtt{DP|K}} =
+\(\mathcal{L}^{\rho}_{\mathtt{DP|K}} =
 \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
-  \mathcal{Z}^{\mathtt{DP|K}}} \rho^2_k \) quantifies the average
+  \mathcal{Z}^{\mathtt{DP|K}}} \rho_k \) quantifies the average
 nuisance uncertainty. The optimal bin-splitting is the one that
@@
 
-\paragraph{Discussion.}
-
 \section{REAL-WORLD EXAMPLES}
 
-we also
-apply the methods on some real-world datasets. In these cases, it is
-infeasible to obtain ground truth information, therefore, we perform
-evaluation at two levels. First, we compare visually NAME versus PDP
-with ICE plots. Second, we hold as ground-truth the effects computed
-on the full training-set and we, then, perform subsampling to evaluate
-the robustness of the bin-splitting method.
+In Real-World Datasets it is infeasible to obtain ground truth
+information, therefore, we perform evaluation at two levels. First, we
+compare visually NAME versus PDP with ICE plots. Second, we hold as
+ground-truth the effects computed on the full training-set and we,
+then, perform subsampling to evaluate the robustness of the
+bin-splitting method.
 

### 2022-09-30 AISTATS2023/paper.tex
updated real dataset part

@@
 
-\section{REAL-WORLD EXAMPLES}
+\section{REAL-WORLD EXAMPLE}
 
-In Real-World Datasets it is infeasible to obtain ground truth
-information, therefore, we perform evaluation at two levels. First, we
-compare visually NAME versus PDP with ICE plots. Second, we hold as
-ground-truth the effects computed on the full training-set and we,
-then, perform subsampling to evaluate the robustness of the
-bin-splitting method.
+Here, we aim at demonstrating the usefuleness of uncertainty
+quantification and the advantages of automatic bin-splitting, on the
+real-world California Housing dataset~\citep{pace1997sparse}.
 
+\paragraph{ML setup}
+
+The California Housing is a largely-studied dataset with approximately
+\(20000\) training instances, making it appropriate for robust
+Monte-Carlo approximations. The dataset contains \(D=8\) features
+describing building blocks of California, like latitude, longitude,
+total number of people residing in the block etc., and the target is
+the median house value inside the block. We exclude instances with
+missing or outlier values, i.e. \(x_d^i > \hat{sigma}_j\) in any
+feature and normalize them to have
+\(\hat{\mu}_d = 0, \hat{\sigma_d}\). We split the dataset into
+\(N_{tr} = 15639\) training and \(N_test = 3910\) test examples (80/20
+split) and we fit a Neural Network with 3 hidden layers of 256, 128
+and 36 units respectively. After 15 epochs using the Adam optimizer
+with learning rate \(\eta = 0.02\), the model achieves a normalized
+mean squared error (R-Squared) of \(0.25\) (0.75). In depth information
+about preprocessing, training and evaluation are provided in the
+Appendix.
+
+\paragraph{Uncertainty Quantification}
+
+In real-world datasets, it is infeasible to obtain the ground truth FE
+for seamlessly evaluating the competitive methods. Therefore, the main
+purpose of this part is to demonstrate the usefulness of uncertainty
+quantification in NAME, i.e., illustrate that modeling the
+heterogeneous effects is important for the interpretation of an ALE
+plot. For this reason, Secondly, we discuss about the FE plots obtaing with NAME and
+PDP-ICE.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_ale_auto.pdf}
+  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_pdp_ice.pdf}\\
+  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_ale_auto.pdf}
+  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_pdp_ice.pdf}\\
+  \caption{Figure 1}
+  \label{fig:ex-synth-2-1}
+\end{figure}
+
+\paragraph{Bin Splitting}
+
+For evaluating NAME in terms of bin-splitting we set-up the following
+evaluation framework. We treat as ground-truth the effects computed on
+the full training-set \(N=20000\) with dense fixed-size bin-splitting
+(\(K=80\)). Given the big number of samples, we make the hypothesis
+that the approximation with dense binning is close to the ground
+truth. Afterwards we select less samples \(N=1000\) and we compare
+NAME approximation with all possible fixed-size alternatives, in terms
+of accuracy, i.e. we provide the quantities
+\(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma^2}, \mathcal{L}^{\rho^2}\).
 \section{CONCLUSION}

### 2022-10-03 AISTATS2023/paper.tex
added some content

@@
 
-\aistatstitle{Instructions for Paper Submissions to AISTATS 2023}
+\aistatstitle{Accumulated Regional Effects: An ALE-based method for regional effects with uncertainty quantification}
 
@@
 \begin{abstract}
-  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
-  both left and right-hand margins. Use 10~point type, with a vertical
-  spacing of 11~points. The \textbf{Abstract} heading must be centered,
-  bold, and in point size 12. Two line spaces precede the
-  Abstract. The Abstract must be limited to one paragraph.
+  This paper is on Feature Effect (FE), a class of global explainability techniques that create a plot with the average effect of a single feature on the output.
+The task is particularly challenging when the black-box function contains non-linear interactions between correlated features.
+In these cases, FE should provide, first, the expected effect and, second, the uncertainty of the effect, i.e.~the level of heterogeneous effects due to implicit feature interactions.
+Accumulated Local Effects (ALE) is a novel method that models accurately the expected effect in cases of correlated features, overcoming limitations of previous approaches.
+In this paper, we define Accumulated Regional Effects (ARE) a FE method that extends ALE to quantify the heterogeneous effects.
+We also provide an alternative definition of ARE based on a sequence of interpretable variable-size regions and we formally prove that under some conditions both definitions are equivalent.
+Using that, we convert the typical bin-splitting step of ALE to a principled optimization problem and we present an algorithm for finding the optimal sequence of regions.
+We demonstrate through synthetic and real-datasets, that ARE quantifies correctly the average and heterogeneous effects and approximates them robustly by extracting regions with similar-effect.
 \end{abstract}
@@
 
-Recently, ML has flourished in critical domains, such as healthcare
-and finance. In these areas, we need a combination of accurate
-predictions along with meaningful explanations to support them. For
-this reason there is an increased interest in Explainable AI (XAI),
-the field that provides interpretations about the behavior of complex
-black-box models. XAI literature distinguishes between local and
-global explainability
+Recently, Machine Learning (ML) has flourished in critical domains,
+such as healthcare and finance. In these areas, we need a combination
+of accurate predictions along with meaningful explanations to support
+them. For this reason there is an increased interest in Explainable AI
+(XAI), the field that provides interpretations about the behavior of
+complex black-box models. XAI literature distinguishes between local
+and global explainability
 techniques~\citep{Molnar2020interpretable}. Local methods explain a
@@
 isolate and visualize the impact of a single feature on the
-output.~\footnote{FE methods also isolate the effect of a pair of
-  features to the output. Combinations of more than two features are
-  not usual, because they encounter, among others, visualization
-  difficulties.} FE methods suffer from aggregation bias because,
-often, the rationale behind the average effect might be unclear. For
-example, a feature with zero average effect may indicate that the
-feature has no effect on the output or, contrarily, it has a highly
-positive effect in some cases and a highly negative in others. There
-are three widely-used FE methods; Partial Dependence Plots
-(PDP)\citep{friedman2001greedy}, Marignal Plots
-(MP)\citep{apley2020visualizing} and Aggregated Local Effects
+output.~\footnote{Often, FE methods also isolate the combined effect
+  of a pair of features to the output. Combinations of more than two
+  features are not usual, because they encounter, among others,
+  visualization difficulties.} FE methods suffer from aggregation bias
+because, often, the rationale behind the average effect is
+unclear. For example, a feature with zero average effect may indicate
+no effect on the output or, in contrast, highly positive in some cases
+and highly negative in others. There are three widely-used FE methods;
+Partial Dependence Plots (PDP)\citep{friedman2001greedy}, Marignal
+Plots (MP)\citep{apley2020visualizing} and Aggregated Local Effects
 (ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
 computing erroneous effects when the input features are (highly)
-correlated, which is a frequent scenario in many ML
-problems. Therefore, ALE has been established as the state-of-the-art
-FE method.
+correlated, which is a frequent scenario in many ML cases. Therefore,
+ALE has been established as the state-of-the-art FE method.
 
 However, ALE faces two crucial limitations, the first concerns ALE
-definition and the second one the ALE approximation. Regarding the
+definition and the second one ALE approximation. Regarding the
 definition ALE does not inform the user about the level of
-heterogeneous effects hidden behind the average effect (global
-explanation). In contrast, in the case of PDP, the heterogeneous
-effects can be (partially) spotted by exploring the Individual
-Conditional Expectations (ICE)\citep{goldstein2015peeking}. Regarding
-the approximation, e.g.~the estimation of ALE from the limited samples
-of the training set, ALE requires an important additional step, called
-\textit{bin-splitting}. Bin-splitting consists of partitioning the
-axis of the feature of interest in a sequence of non-overlapping
-intervals and estimating a single effect in each one, from the
-population of samples that fall inside. Specifying an approriate
-sequence of intervals is of particular importance, since ALE's
-interpretation is meaningful only inside each interval
-\citep{molnar2022}. However, this crucial step has not raised the
-appropriate attention, making the approximation vulnerable to
+uncertainty, i.e., the heterogeneous effects hidden behing the average
+effect due to implicit feature interactions. In contrast, in PDPs the
+heterogeneous effects can be (partially) spotted by exploring the
+Individual Conditional Expectations
+(ICE)\citep{goldstein2015peeking}. Regarding the approximation,
+i.e.~the estimation usign the (limited) samples of the training set,
+ALE requires the additional \textit{bin-splitting} step. Bin-splitting
+consists of partitioning the axis of the feature of interest in a
+sequence of non-overlapping intervals (regions) and estimating a
+constant effect in each one from the population of samples that fall
+inside. Specifying an approriate sequence of regions is of particular
+importance, since ALE's interpretation is meaningful only inside each
+region \citep{molnar2022}. However, this crucial step has not raised
+the appropriate attention, making the approximation vulnerable to
 potential misinterpretations.
 
-In this paper, we extend ALE with a probabilistic component for
-measuring the uncertainty of the explanation. The uncertainty of the
-global explanation, i.e.~how certain we are that the averaged
-explanation is valid if applied to an instance drawn at random,
-quantifies the level of heterogeneous effects. Given that ALE's
-interpretation (expected effect and uncertainty) is meaningful only
-inside each interval, we design a principled framework, where we treat
-the bin-splitting step as a data-driven clustering problem, searching
-for the optimal splitting given available instances of the training
-set. We, also, present a computationally-grounded algorithm for
-finding the optimal solution.
+In this paper, we present Accumulated Regional Effects (ARE) an
+extension of ALE for measuring the uncertainty of the explanation,
+i.e.~how certain we are that the average explanation is valid if
+applied to an instance drawn at random. We also provide an alternative
+definition of ARE based on a sequence of constant-effect regions and
+we define a metric of the dissimilarity of the two definitions. We
+formally prove that under some conditions the dissimilarity is zero
+and both definitions are equivalent. We, therefore, set up an
+optimization problem where we search for the sequence of region that
+minimizes the disimilarity, given enough training points iniside each
+region for a robust estimation. Finally, we propose a
+computationally-grounded algorithm for finding the optimal solution.
 
-\paragraph{Contributions.} The contribution of this paper is NAME, a
-feature effect method that:
+\paragraph{Contributions.} The contribution of this paper are:
 
 \begin{itemize}
-\item Extends ALE with a probabilistic component to quantify the
-  heterogeneous effects behind the global explanation.
-\item Automatically extracts regions with similar effects, improving
-  the estimation and the interpretability of ALE plots.
+\item ARE, a FE method that extends ALE for quantifying the
+  heterogeneous effects (uncertainty).
+\item An interval-based definition of ARE and a formal proof that
+  under some conditions it is equivalent to the initial definition.
+\item A principled framework for automatically extracting regions with
+  similar effects, improving the estimation and the interpretability
+  of ALE plots.
 \end{itemize}
@@
 
-At Section~\ref{sec:feat-effect-meth}, we present the basic Feature
-Effect (FE) methods and we review the ALE
-definition\citep{apley2020visualizing}. At
-Section~\ref{sec:quant-heter-effects}, we present the techniques for
-quantifying the heterogeneous effects and we discuss some of their
-limitations. Finally, at Section~\ref{sec:ale-approximation}, we
-review the ALE approximation\citep{apley2020visualizing}(add us),
-describing why it is vulnerable to a poor approximation due to
-inappropriate bin-splitting.
+In this section, we describe the basic methods for FE and for
+uncertainty quantification, focusing on ALE definition. We, then,
+review the ALE approximation\citep{apley2020visualizing, gkolemis22},
+describing some of its vulnerabilities.
 
@@
 \mathbf{X}_c)]\). ALE defines the local effect of the \(s\)-th feature
-at point \(x_s = z\) as \(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
+at point \(x_s = z\) as
+\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
 local explanations at \(z\) are, then, weighted by the conditional
 distribution \(p(\xc|x_s=z)\) and are averaged, to produce the
-averaged effect \(\mu(z)\). Finally, ALE is the accumulation of the
-averaged local effects:
+averaged effect \(\mu(z)\). ALE is the accumulation of the averaged
+local effects:
 
@@
 
-ALE has specific advantages which gain particular value in cases of
+Eq.~(\ref{eq:ALE}) has specific advantages which gain particular value in cases of
 correlated input features. In these cases, PDP integrates over
@@
 
-FE methods pose the question \textit{what is average (global) effect
-  on the output, if the value of a specific feature is
-  increased/decreased}. After that, it comes naturally to also ask
-about \textit{to what extent the individual (local) effects deviate
-  from the average}. For this reason, the quantification of the
-uncertainty has attracted a lot of interest.
-
-ICE and d-ICE\citep{goldstein2015peeking} provide a set of curves that
-are plot on top-of the PDP. Both methods produce one curve for each
+FE methods reply to the question \textit{what is the average (global)
+  effect on the output, if the value of a specific feature is
+  increased/decreased}. It comes naturally to also ask \textit{to what
+  extent the local effects deviate from the global
+  explanation}. Quantifying the uncertainty of the global explanation
+has attracted a lot of interest. ICE and
+d-ICE\citep{goldstein2015peeking} provide a set of curves that are
+plot on top-of the PDP. Both methods produce one curve for each
 instance of the training set;
@@
 \(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
-\xc^i)\) for d-ICE. The user then visually observes if the (d)-ICE
-curves are homogeneous and to what extent they deviate from the
-PDP. Some methods try to automate the aforementioned visual
-exploration, by grouping (d-)ICE into clusters~\citep{molnar2020model,
+\xc^i)\) for d-ICE. The user then visually observes if the curves are
+homogeneous and to what extent they deviate from the PDP. Some methods
+try to automate the aforementioned visual exploration, by grouping
+(d-)ICE plots into clusters~\citep{molnar2020model,
   herbinger2022repid, britton2019vine}. Some other approaches, like
@@
 on PDPs, and, therefore, they are subject to the failure modes of PDPs
-in cases of correlated features\citep{baniecki2021fooling}. To the
-best of our knowledge, no work exist so far that quantifies the
+in cases of correlated features\citep{baniecki2021fooling}, as we will
+show in-depth in Section~\ref{sec:simulation-examples-1}. To the best
+of our knowledge, no work exist so far that quantifies the
 heterogeneous effects based on the formulation of ALE.
@@
 
-In real ML scenarios, the FE is estimated from the limited instances
-of the training set. Therefore, \citep{apley2020visualizing} proposed
-dividing the \(s\)-th axis in \(K\) bins and estimating the local
-effects in each bin by evaluating the black box-function at the bin
-limits:
+In real ML scenarios, the FE is estimated from the training set
+examples. Therefore, \citep{apley2020visualizing} proposed dividing
+the \(s\)-th axis in \(K\) bins and estimating the local effects in
+each bin by evaluating the black box-function at the bin limits:
 \begin{equation}
@@
 \( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
-\). Afterwards, (cite) proposed the Differential ALE (DALE)
-approximation for computing the local effects on the training
-instances using auto-differentiation:
+\). Afterwards,~\citep{gkolemis22} proposed the Differential ALE
+(DALE) that computes the local effects on the training instances using
+auto-differentiation:
 \begin{equation}
@@
 Both approximations ask from the user to blindly decide the number of
-bins \(K\) for splitting the axis into equally-sized bins, an approach
-with crucial limitations. Firstly, it is vulnerable to poor bin-effect
-approximations that lead to a misleading ALE plot. Setting \(K\) to a
-small value may hide fine-grain effects due to large bins and setting
-\(K\) to a high value leads to estimations of many bin-effects from
-very limited samples. In general, the user may face contradictory
-explanations for differnt \(K\) without a clue to decide which one to
-trust. Secondly, as indicated by~\citep{molnar2022}, the ALE's feature
-effect can only be interpreted in local regions. Therefore, for better
-interpretation, the extend of these regions should be determined
-in a data-driven way.
+bins, denoted with \(K\), for splitting the axis into equally-sized
+bins, an approach with crucial limitations. Setting \(K\) to a small
+value may hide fine-grain effects due to large bins and setting \(K\)
+to a high value leads to poor bin-effect estimations from limited
+samples. In general, the user may face contradictory explanations for
+different \(K\) without a clue to decide which one to trust.
 
-\section{THE NAME METHOD}
+\section{Accumulated Regional Effects (ARE)}
 \label{sec:NAME-method}
 
-The NAME method extends ALE for quantifying the level of heterogeneous
-effects (uncertainty) and automating the bin-splitting step for robust
-estimations. At Section~\ref{sec:NAME-definition}, we define NAME,
-presenting the component that extends ALE with uncertainty
-quantification. In Section~\ref{sec:interval-based-estimation}, we
-provide an alternative definition of NAME based on variable-size bins,
-providing important remarks and proofs for connecting it with the
-initial definition. In Section~\ref{sec:interval-spliting}, we define
-the problem of optimal bin-splitting and we provide an algorithmic
+ARE extends ALE for quantifying the level of heterogeneous effects
+(uncertainty) and automates the bin-splitting step for robust
+estimations. At Section~\ref{sec:NAME-definition}, we define ARE and
+at Section~\ref{sec:interval-based-estimation} we provide an
+alternative definition of ARE based on variable-size bins, providing
+important remarks and proofs for connecting it with the initial
+definition. In Section~\ref{sec:interval-spliting}, we define the
+problem of optimal bin-splitting and we propose an algorithmic
 solution to the problem.
 
-\subsection{NAME: ALE With Uncertainty Quantification}
+\subsection{ARE: ALE With Uncertainty Quantification}
 \label{sec:NAME-definition}
 
-ALE does not shed light to potential heterogeneous effects that
-contribute to the average explanation. Therefore, we extend ALE
-definition of Eq.~(\ref{eq:ALE}) with a component that quantifies the
-level of heterogeneous effects. We denote \(\mathcal{H}(z)\) the
+We extend ALE definition of Eq.~(\ref{eq:ALE}) with a component for
+quantifying the uncertainty. We denote as \(\mathcal{H}(z)\) the
 uncertainty of the local effects at a specific point \(x_s=z\) and we
@@
 \noindent
-NAME method formulates the effect at a specific point \(x_s\) with a
-compact entity that consists of the average effect and the uncertainty
-\(\mu(z) \pm \sigma(z)\) and visualizes them as a continuous curve
+ARE method formulates the effect at a specific point \(x_s\) with a
+compact tuple that consists of the average effect and the uncertainty
+\((\mu(z), \sigma(z))\) and visualizes them as a continuous curve
 with a confidence region, i.e.
-\(f^{\mathtt{METHOD}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
+\(f^{\mathtt{ARE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
 f^{\mathtt{ALE}}_{\sigma}(x_s)\).
 
-\subsection{Interval-Based Formulation of NAME}
+\subsection{Interval-Based Formulation of ARE}
 \label{sec:interval-based-estimation}
@@
 granularity of a point is impossible, because the probability of
-observing a sample inside the interval \([x_s - h, x_s + h]\) tends to
-zero, when \(h \to 0\). We define the mean effect (bin-effect) and the
-uncertainty (bin-uncertainty) inside an interval \([z_1, z_2)\) as:
+observing a sample inside the region \([x_s - h, x_s + h]\) tends to
+zero, when \(h \to 0\). We define the region-effect and the
+region-uncertainty inside the region \([z_1, z_2)\) as:
 
@@
   \label{eq:var_bin}
-  \sigma(z_1, z_2) = \frac{\int_{z_1}^{z_2}
+  \sigma(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
   \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
-      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 - z_1}
+      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 - z_1}}
 \end{equation}
@@
 %
-We also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points that
-partition the \(s\)-th feature axis into a sequence \(K\)
-variable-size intervals, i.e. \(\mathcal{Z}_s = \{z_0, z_K\}\), where
+We also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points
+that partition the \(s\)-th feature axis into \(K\) variable-size
+intervals, i.e. \(\mathcal{Z}_s = \{z_0, z_K\}\), where
 \(z_0=x_{s,min}\) and \(z_K=x_{s,max}\). Based on the above, we
-redefine NAME using an interval-based formulation
-\(\tilde{f}^{\mathtt{METHOD}}(x_s):=
-\tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) \pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:
+redefine ARE using an interval-based formulation
+\(\tilde{f}^{\mathtt{ARE}}(x_s):=
+\tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) \pm
+\tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:
 
@@
 
-\textit{If we define the residual \(\rho(z)\) as the
-difference between the expected effect at \(x_s\) and the mean
-expected effect at the interval, i.e
-\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
-\(\sigma^2(z_1, z_2)\) equals to the accumulation of the point-wise
+\textit{If we define the residual \(\rho(z)\) as the difference
+  between the expected effect at \(x_s\) and the region-effect, i.e
+  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the region-variance
+  \(\sigma^2(z_1, z_2)\) equals to the integration of the point-wise
   variances plus the squared residuals inside the interval:
 \begin{equation}
- \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
+ \sigma^2(z_1, z_2) = \frac{\int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z}{z_2 - z_1}
 \end{equation}}
@@
 
-\textit{If we define as \(\rho^2(z_{k-1}, z_{k})\) the accumulated residual of
-a bin, i.e.
-\(= \rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial
-z\), then the two definitions of NAME are equivalent, i.e.,
-\(f^{\mathtt{NAME}} = \tilde{f}^{\mathtt{NAME}}\), if and only if the
-sequence of limits \(\mathcal{Z}\) is such that
+\textit{If we define as \(\rho^2(z_{k-1}, z_{k})\) the
+  region-residual, where
+  \(\rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial
+  z\), then the two definitions of ARE,
+  \(f^{\mathtt{ARE}}, \tilde{f}^{\mathtt{ARE}}\) are equivalent, if
+  and only if the sequence of limits \(\mathcal{Z}_s\) are such that
   \(\sum_{k=1}^K \rho^2(z_{k-1}, z_{k}) = 0\).}
@@
 the natural characteristics of the experiment, i.e.,
-\(\mathcal{H}(z_1, z_2) = \int_{z_1}^{z_2}\mathcal{H}(z) \partial z\),
+\(\mathcal{H}(z_1, z_2) = \int_{z_1}^{z_2}\mathcal{H}^2(z) \partial z\),
 and the second term adds extra nuisance uncertainty due to limiting
@@
 bin adds no nuisance uncertainty, then we will get exactly the same
-NAME explanation as in the original definition.
+ARE explanation as in the original definition. In general, we use the
+quantity \(Q_{\mathcal{Z}} = \sum_{k=1}^K \rho^2(z_{k-1}, z_{k})\) to
+measure the dissimilarity between the actual definition of ARE
+\(f^{\mathtt{ARE}}\) and the region-based definition
+\(\tilde{f}^{\mathtt{ARE}}\) using the sequence of intervals
+\(\mathcal{Z}_s\). In this way, we access a quantified metric for
+evaluating how good is a particular splitting.
 
@@
 feature lying inside the interval
-\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \}
-\):
+\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \} \):
 
@@
 
+% Under this
+% assumption, and due to the law of large numbers, the approximations
+% become more robust, as the bins become wider. On the other hand, bins
+% should group local explanations with similar effects and enlarging a
+% bin is under the danger of adding nuisance uncertainty.
+
 Both approximations are unbiased only under the assumption that the
-points are uniformly distributed inside the intervals. Under this
-assumption, and due to the law of large numbers, the approximations
-become more robust, as the bins become wider. On the other hand, bins
-should group local explanations with similar effects and enlarging a
-bin is under the danger of adding nuisance uncertainty. Therefore, we
-set-up the following optimization problem:
+points are uniformly distributed inside the intervals. We set-up the
+following optimization problem:
+
 \begin{equation}
@@
 \begin{aligned}
-  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
+  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \tau_k \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
   \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
@@
 We search for the sequence of intervals \(z_0, \ldots, z_K\) that
-minimizes the sum of the cost of each bin. The cost of each bin is the
-mean variance \(\sigma^2_k\) scaled by the bin length \(\Delta z_k\)
+minimizes the sum of bin-costs. The cost of each bin is the mean
+variance \(\hat{sigma}^2_k\) scaled by the bin length \(\Delta z_k\)
 and discounted by the term \(\tau_k\). The term \(\tau_K\) acts as a
@@
 practice~\citep{aas2021explaining, herbinger2022repid} for providing
-secure empirical evidence about the superiority of NAME method.
+secure empirical evidence about the superiority of ARE method.
 
 We split the simulation examples in two groups. The first group,
-Section~\ref{sec:simulation-examples-1}, aims at showcasing that NAME
+Section~\ref{sec:simulation-examples-1}, aims at showcasing that ARE
 method is more accurate than PDP-ICE in quantifying the average effect
@@
 features are correlated. The second group,
-Section~\ref{sec:simulation-examples-2}, illustrates that NAME method
+Section~\ref{sec:simulation-examples-2}, illustrates that ARE method
 achieves a better approximation (average effect and uncertainty) in
@@
 
-In this simulation, we will compare NAME method against PDP-ICE in
+In this simulation, we will compare ARE method against PDP-ICE in
 quantifying the main effect and the uncertainty. Since there is
@@
 \(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test the effect
-computed by NAME and PDP-ICE in three cases; (a) no interaction
+computed by ARE and PDP-ICE in three cases; (a) no interaction
 (\(\alpha=0\)) and equal weights (\(a_1=a_2\)), (b) no interaction
@@
 \(X_2\). Finally, besides it is not the focus of the current examples,
-we will also observe that NAME manages to correctly extract the
+we will also observe that ARE manages to correctly extract the
 regions with constant effect. Details for obtaining the ground truth
@@
   \caption{No interaction, Equal weights: Feature effect for \(x_1\);
-    Ground-truth vs (a) NAME method at the left and (b) PDP-ICE at the
+    Ground-truth vs (a) ARE method at the left and (b) PDP-ICE at the
     right.}
@@
   \caption{No interaction, Different weights: Feature effect for \(x_1\);
-    Ground-truth vs (a) NAME method at the left and (b) PDP-ICE at the
+    Ground-truth vs (a) ARE method at the left and (b) PDP-ICE at the
     right.}
@@
 zero-uncertainty, since it does not appear in any interaction term. As
-we observe in Figure~\ref{fig:ex-synth-1-3}, NAME captures perfectly
+we observe in Figure~\ref{fig:ex-synth-1-3}, ARE captures perfectly
 the effect of all features, and the uncertainty in all cases. As
@@
     features, \(x_1\) to \(x_3\) from top to bottom; Ground-truth vs
-    (a) NAME method at the left columns and (b) PDP-ICE at the right
+    (a) ARE method at the left columns and (b) PDP-ICE at the right
     column.}
@@
 sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). As we observe, in
-Figure~\ref{fig:ex-synth-2-1}, NAME method extracts a sequence of
+Figure~\ref{fig:ex-synth-2-1}, ARE method extracts a sequence of
 intervals with better \(\mathcal{L}_{\mathtt{DP}}^{\mu}\) and
@@
 purpose of this part is to demonstrate the usefulness of uncertainty
-quantification in NAME, i.e., illustrate that modeling the
+quantification in ARE, i.e., illustrate that modeling the
 heterogeneous effects is important for the interpretation of an ALE
-plot. For this reason, Secondly, we discuss about the FE plots obtaing with NAME and
+plot. For this reason, Secondly, we discuss about the FE plots obtaing with ARE and
 PDP-ICE.
@@
 
-For evaluating NAME in terms of bin-splitting we set-up the following
+For evaluating ARE in terms of bin-splitting we set-up the following
 evaluation framework. We treat as ground-truth the effects computed on
@@
 truth. Afterwards we select less samples \(N=1000\) and we compare
-NAME approximation with all possible fixed-size alternatives, in terms
+ARE approximation with all possible fixed-size alternatives, in terms
 of accuracy, i.e. we provide the quantities

### 2022-10-04 AISTATS2023/paper.tex
explained better the variance part

@@
 
-\subsection{Interval-Based Formulation of ARE}
+\subsection{Interval-Based Formulation and Approximation}
 \label{sec:interval-based-estimation}
@@
 observing a sample inside the region \([x_s - h, x_s + h]\) tends to
-zero, when \(h \to 0\). We define the region-effect and the
-region-uncertainty inside the region \([z_1, z_2)\) as:
+zero, when \(h \to 0\). Therefore we define the regional-effect
+\(\mu(z_1, z_2)\) and the regional-uncertainty \(\sigma(z_1, z_2)\),
+as the summary statistics that characterize the effect and the
+uncertainty inside the interval \([z_1, z_2)\):
 
@@
   \sigma(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
-  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
-      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 - z_1}}
+  \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z) )^2 \right] \partial z}{z_2 - z_1}}
 \end{equation}
@@
 %
-We also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points
+The regional-effect and the regional-uncertainty are the mean values
+of the local effects and local uncertainty inside an interval. In
+other words, they quantify what is the expected effect and the
+expected uncertainty of a sample drawn at random from \([z_1, z_2]\)
+given uniform distribution \(\mathcal{U}(z_1, z_2)\).
+
+If we also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points
 that partition the \(s\)-th feature axis into \(K\) variable-size
-intervals, i.e. \(\mathcal{Z}_s = \{z_0, z_K\}\), where
-\(z_0=x_{s,min}\) and \(z_K=x_{s,max}\). Based on the above, we
+intervals, i.e. \(\mathcal{Z}_s = \{z_0, \ldots, z_K\}\), we can
 redefine ARE using an interval-based formulation
-\(\tilde{f}^{\mathtt{ARE}}(x_s):=
-\tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) \pm
-\tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:
+\(\tilde{f}^{\mathtt{ARE}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)
+\pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:
 
@@
 %
-\paragraph{Theorem 1.}
-\label{sec:theorem-1}
 
-\textit{If we define the residual \(\rho(z)\) as the difference
-  between the expected effect at \(x_s\) and the region-effect, i.e
-  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the region-variance
-  \(\sigma^2(z_1, z_2)\) equals to the integration of the point-wise
-  variances plus the squared residuals inside the interval:
-\begin{equation}
- \sigma^2(z_1, z_2) = \frac{\int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z}{z_2 - z_1}
-\end{equation}}
-%
+\subsection{Interval-Based Approximation}
 
-\paragraph{Theorem 2.}
-\label{sec:theorem-2}
-
-\textit{If we define as \(\rho^2(z_{k-1}, z_{k})\) the
-  region-residual, where
-  \(\rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial
-  z\), then the two definitions of ARE,
-  \(f^{\mathtt{ARE}}, \tilde{f}^{\mathtt{ARE}}\) are equivalent, if
-  and only if the sequence of limits \(\mathcal{Z}_s\) are such that
-  \(\sum_{k=1}^K \rho^2(z_{k-1}, z_{k}) = 0\).}
-
-\paragraph{Discussion}
-
-Theorem 1 decouples the aggregated uncertainty inside a bin, i.e.,
-\(\mathcal{H}_{bin}(z_1, z_2) := \sigma^2(z_1, z_2)\), into two
-terms. The first term quantifies the accumulated uncertainty due to
-the natural characteristics of the experiment, i.e.,
-\(\mathcal{H}(z_1, z_2) = \int_{z_1}^{z_2}\mathcal{H}^2(z) \partial z\),
-and the second term adds extra nuisance uncertainty due to limiting
-the resolution
-\(\mathcal{H}_n(z_1, z_2) := \int_{z_1}^{z_2}\rho^2(z) \partial z
-\). Therefore, it holds that:
-\begin{equation}
-  \label{eq:uncertainty_bin}
-  \mathcal{H}_{bin}(z_1, z_2) = \mathcal{H}(z_1, z_2) + \mathcal{H}_n(z_1, z_2)
-\end{equation}
-%
-In Theorem 2 we prove that if we find a sequence of bins, where each
-bin adds no nuisance uncertainty, then we will get exactly the same
-ARE explanation as in the original definition. In general, we use the
-quantity \(Q_{\mathcal{Z}} = \sum_{k=1}^K \rho^2(z_{k-1}, z_{k})\) to
-measure the dissimilarity between the actual definition of ARE
-\(f^{\mathtt{ARE}}\) and the region-based definition
-\(\tilde{f}^{\mathtt{ARE}}\) using the sequence of intervals
-\(\mathcal{Z}_s\). In this way, we access a quantified metric for
-evaluating how good is a particular splitting.
-
-\subsection{Bin-Splitting: Finding Regions With Homogeneous Effects}
-\label{sec:interval-spliting}
-
-In this section we formulate bin-splitting as an unsupervised
-clustering problem. The mean effect (Eq.~\eqref{eq:mu_bin}) and the
+The mean effect (Eq.~\eqref{eq:mu_bin}) and the
 uncertainty (Eq.~\eqref{eq:var_bin}) can be directly estimated from
-the set \(\mathcal{S}_k\) of the dataset instances with the \(s\)-th
+the set \(\mathcal{S}_k\) of dataset instances with the \(s\)-th
 feature lying inside the interval
@@
 
-% Under this
-% assumption, and due to the law of large numbers, the approximations
-% become more robust, as the bins become wider. On the other hand, bins
-% should group local explanations with similar effects and enlarging a
-% bin is under the danger of adding nuisance uncertainty.
+Under the assumption that the points are uniformly distributed inside
+the intervals Eq.\eqref{eq:2} is an unbiased estimatior of
+Eq.~(\ref{eq:mu_bin}). The same does not hold in the general case for
+Eq.\eqref{eq:3}. In the general case, \(\hat{\sigma}^2(z_{k-1}, z_k)\)
+is an unbiased estimator of the observable standard deviation
+\(\sigma_{obs}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
+    \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2
+    \right] \partial z}{z_2 - z_1}}\), which as we prove in Theorem 1
+is an overestimation of the real regional variance.
 
-Both approximations are unbiased only under the assumption that the
-points are uniformly distributed inside the intervals. We set-up the
+\paragraph{Theorem 1.}
+\label{sec:theorem-1}
+
+\textit{If we define the residual \(\rho(z)\) as the difference
+  between the expected effect at \(x_s\) and the regional effect, i.e
+  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the regional variance
+  \(\sigma^2(z_1, z_2)\) equals to:}
+
+\begin{equation}
+    \label{eq:bin-uncertainty-proof}
+ \sigma_{obs}^2(z_1, z_2) = \sigma^2(z) + \frac{\int_{z_1}^{z_2}\rho^2(z) \partial z}{z_2 - z_1}
+\end{equation}
+\noindent
+Theorem 1 reveals an important connection between the ground-truth and
+the observable uncertainty inside a region. Setting as ground-truth
+the regional uncertainty
+\(\mathcal{H}_{true}(z_1, z_2) = \sigma(z_1, z_2)}\) and as observable
+uncertainty the one we estimate with Eq.~(\ref{eq:3}), i.e.
+\(\mathcal{H}_{obs}(z_1, z_2) := \sigma_{obs}(z_1, z_2)\), then there
+is an error term quantified by the residual inside the interval, i.e.
+\(\mathcal{E}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}\rho^2(z)
+    \partial z}{z_2 - z_1}\)}, such that:
+
+\begin{equation}
+  \label{eq:uncertainty-bin}
+  \mathcal{H}^2_{obs}(z_1, z_2) = \mathcal{H}^2_{true}(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
+\end{equation}
+%
+Eq.~(\ref{eq:uncertainty-bin}) shows that the error term quantifies
+how much the observable uncertainty of the region \([z_1, z_2]\)
+deviates for the one we would like to measure and provides a
+quantified metric to evaluate a particular splitting. As we prove in
+the Appendix, UALE uncertainty approximation is unbiased if and only if the
+partitioning is such that ther error term is zero for all bins, i.e. 
+\(\sum_{k=1}^K \mathcal{E}^2(z_{k-1}, z_{k}) = 0\)
+
+\subsection{Bin-Splitting: Finding Regions With Homogeneous Effects}
+\label{sec:interval-spliting}
+
+In this section we formulate bin-splitting as an unsupervised
+clustering problem. The main idea of the clustering problem is to find
+the best bin-splitting solution, handling appropriately two
+conflicting objectives. On the one hand, we want to minimize the error
+term \(\mathcal{E}\) defined above for minimizing the error in the
+uncertainty estimation and on the other hand, we want wide-enough bins
+for including a fair population of samples for a robust estimation of
+\(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\). We set-up the
 following optimization problem:
@@
   \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
-  & \tau_k = 1 - 0.2 \frac{|S_k|}{N} \\
+  & \tau_k = 1 - \alpha \frac{|S_k|}{N} \\
   \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
@@
 We search for the sequence of intervals \(z_0, \ldots, z_K\) that
-minimizes the sum of bin-costs. The cost of each bin is the mean
-variance \(\hat{sigma}^2_k\) scaled by the bin length \(\Delta z_k\)
-and discounted by the term \(\tau_k\). The term \(\tau_K\) acts as a
-tie-braker in cases of two candidate bins with equal mean variance,
-favoring the selection of the bigger bin. The constraint of at least
-\(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
-\textit{robust} estimation.
-
-According to Eq.~(\ref{eq:uncertainty_bin}), the uncertainty of a bin
-\(\mathcal{H}_{bin}\) is the sum of the uncertainty due to the
-characteristics of the experiment (data and black-box model)
-\(\mathcal{H}\) and the nuisance uncertainty \(\mathcal{H}_n\). Since
-\(\mathcal{H}\) is independent of the bin-spliting procedure, by
-minimizing \(\mathcal{L}\) we minimize the nuiscane uncertainty
-\(\mathcal{H}_n\) in each bin.
+minimizes the sum of bin costs. The cost of each bin is the
+approximated variance \(\hat{\sigma}^2_k\) scaled by the bin length
+\(\Delta z_k\) and discounted by the term \(\tau_k\). The term
+\(\tau_K\) acts as a tie-braker that favors the selection of a bigger
+bin in case it has similar variance with the aggregate variance of
+many smaller ones. The constraint of at least \(N_{\mathtt{PPB}}\)
+points per bin sets the lowest-limit for a \textit{robust}
+estimation. The user can choose to what extent they favor the creation
+of wide bins through the discount parameter by setting the parameter
+\(\alpha\) and where they set the minimum for robust approximation
+with the parameter \(N_{\mathtt{PPB}}\). For providing a rough idea of
+some sensible value, in our experiments we set \(\alpha = 0.2\) which
+means that the discount ranges in \([0\%, 20\%]\) and
+\(N_{\mathtt{PPB}} = \frac{N}{20}\). It is also important to clarify
+that by minimizing \(\hat{\sigma}^2_k \approx \mathcal{H}_{obs}^2\),
+we actually minimize the squared error term \(\mathcal{E}^2\), since
+the term \(\mathcal{H}^2_{true}\) is independent of the bin splitting
+procedure.
 

### 2022-10-04 AISTATS2023/paper.tex

@@
 
-\aistatstitle{Accumulated Regional Effects: An ALE-based method for regional effects with uncertainty quantification}
+\aistatstitle{Uncertainty-aware Accumulated Local Effects (UALE) for
+  quantifying the heterogeneous effects}
 
@@
 \begin{abstract}
-  This paper is on Feature Effect (FE), a class of global explainability techniques that create a plot with the average effect of a single feature on the output.
-The task is particularly challenging when the black-box function contains non-linear interactions between correlated features.
-In these cases, FE should provide, first, the expected effect and, second, the uncertainty of the effect, i.e.~the level of heterogeneous effects due to implicit feature interactions.
-Accumulated Local Effects (ALE) is a novel method that models accurately the expected effect in cases of correlated features, overcoming limitations of previous approaches.
-In this paper, we define Accumulated Regional Effects (ARE) a FE method that extends ALE to quantify the heterogeneous effects.
-We also provide an alternative definition of ARE based on a sequence of interpretable variable-size regions and we formally prove that under some conditions both definitions are equivalent.
-Using that, we convert the typical bin-splitting step of ALE to a principled optimization problem and we present an algorithm for finding the optimal sequence of regions.
-We demonstrate through synthetic and real-datasets, that ARE quantifies correctly the average and heterogeneous effects and approximates them robustly by extracting regions with similar-effect.
+Accumulated Local Effects (ALE) is a popular approach to explainable AI that describes how a feature influences the decisions of a model on average, taking into account potential correlations between the features.
+However, ALE does not provide an explicit component for modeling the uncertainty of the effect, i.e., the level of heterogeneous effects behind the average explanation, which is crucial for a complete interpretation when the black-box function contains interactions between the correlated features.
+In this work, we propose Uncertainty-aware ALE (UALE), an extension of ALE for quantifying the heterogeneous effects.
+We show the importance of bin-splitting for a robust approximation of the uncertainty and we formally prove the conditions for an unbiased estimation.
+We finally propose a computationally-grounded algorithm for finding the optimal sequence of bins given the limited instances of the training set.
+We demonstrate through synthetic and real datasets, that UALE quantifies the average and heterogeneous effects correctly and approximates them robustly by optimizing the bin-splitting procedure.
 \end{abstract}
@@
 the regional uncertainty
-\(\mathcal{H}_{true}(z_1, z_2) = \sigma(z_1, z_2)}\) and as observable
+\(\mathcal{H}_{true}(z_1, z_2) = \sigma(z_1, z_2)\) and as observable
 uncertainty the one we estimate with Eq.~(\ref{eq:3}), i.e.
@@
 \(\mathcal{E}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}\rho^2(z)
-    \partial z}{z_2 - z_1}\)}, such that:
+    \partial z}{z_2 - z_1}}\), such that:
 

### 2022-10-04 AISTATS2023/paper.tex
Updates from Overleaf
changed the name

@@
 technique. On the other hand, aggregating the individual explanations
-for producing a concise global one is vulnerable to
-misinterpretations. Under strong interactions and correlations between
-features, the global explanation may obfuscate heterogeneous effects
-that exist under the hood~\citep{herbinger2022repid}; a phenomenon
-called aggregation bias~\citep{mehrabi2021survey}.
+is vulnerable to misinterpretations. Under strong interactions and
+correlations between features, the global explanation may obfuscate
+heterogeneous effects that exist under the
+hood~\citep{herbinger2022repid}; a phenomenon called aggregation
+bias~\citep{mehrabi2021survey}.
 
-Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category
-of global explainability methods. The objective of FE is to the
-isolate and visualize the impact of a single feature on the
-output.~\footnote{Often, FE methods also isolate the combined effect
-  of a pair of features to the output. Combinations of more than two
-  features are not usual, because they encounter, among others,
-  visualization difficulties.} FE methods suffer from aggregation bias
-because, often, the rationale behind the average effect is
-unclear. For example, a feature with zero average effect may indicate
-no effect on the output or, in contrast, highly positive in some cases
-and highly negative in others. There are three widely-used FE methods;
-Partial Dependence Plots (PDP)\citep{friedman2001greedy}, Marignal
-Plots (MP)\citep{apley2020visualizing} and Aggregated Local Effects
+Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental global
+explainability category. The objective of FE is to isolate the impact
+of a single feature on the output.~\footnote{Often, FE methods also
+  isolate the combined effect of a pair of features to the
+  output. Combinations of more than two features are not usual,
+  because they encounter, among others, visualization difficulties.}
+FE methods suffer from aggregation bias because, often, the rationale
+behind the average effect is unclear. For example, a feature with zero
+average effect may indicate no effect on the output or, in contrast,
+highly positive in some cases and highly negative in others. There are
+three widely-used FE methods; Partial Dependence Plots
+(PDP)\citep{friedman2001greedy}, Marignal Plots
+(MP)\citep{apley2020visualizing} and Aggregated Local Effects
 (ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
@@
 definition and the second one ALE approximation. Regarding the
-definition ALE does not inform the user about the level of
-uncertainty, i.e., the heterogeneous effects hidden behing the average
-effect due to implicit feature interactions. In contrast, in PDPs the
+definition ALE does not inform the user about the uncertainty, i.e.,
+the level of heterogeneous effects hidden behing the average effect
+due to implicit feature interactions. In contrast, in PDPs the
 heterogeneous effects can be (partially) spotted by exploring the
@@
 region \citep{molnar2022}. However, this crucial step has not raised
-the appropriate attention, making the approximation vulnerable to
+the appropriate attention and the approximation is vulnerable to
 potential misinterpretations.
 
-In this paper, we present Accumulated Regional Effects (ARE) an
-extension of ALE for measuring the uncertainty of the explanation,
-i.e.~how certain we are that the average explanation is valid if
-applied to an instance drawn at random. We also provide an alternative
-definition of ARE based on a sequence of constant-effect regions and
-we define a metric of the dissimilarity of the two definitions. We
-formally prove that under some conditions the dissimilarity is zero
-and both definitions are equivalent. We, therefore, set up an
-optimization problem where we search for the sequence of region that
-minimizes the disimilarity, given enough training points iniside each
-region for a robust estimation. Finally, we propose a
-computationally-grounded algorithm for finding the optimal solution.
+In this paper, we present Uncertainty-awary Accumulated Local Effects
+(UALE) an extension of ALE for measuring the uncertainty of the
+explanation, i.e.~how certain we are that the average explanation is
+valid if applied to an instance drawn at random. We show that choosing
+an appropriate sequence of intervals is particularly important for a
+robust estimation of the uncertainty, providing an algorithm that
+automates this step. We also provide theoretical proofs for an
+unbiased estimation of the uncertainty. We apply our method in multiple
+synthetic and real world datasets for evaluating its performance.
 
@@
 \begin{itemize}
-\item ARE, a FE method that extends ALE for quantifying the
+\item UALE, a FE method that extends ALE for quantifying the
   heterogeneous effects (uncertainty).
-\item An interval-based definition of ARE and a formal proof that
-  under some conditions it is equivalent to the initial definition.
-\item A principled framework for automatically extracting regions with
-  similar effects, improving the estimation and the interpretability
-  of ALE plots.
+\item A formal proof of the bin-splitting requirements for an unbiased
+  approximation of the heterogeneous effects.
+\item A framework for automatically extracting regions with similar
+  effects, improving the estimation and the interpretability of ALE
+  plots.
 \end{itemize}
@@
 
-In this section, we describe the basic methods for FE and for
-uncertainty quantification, focusing on ALE definition. We, then,
-review the ALE approximation\citep{apley2020visualizing, gkolemis22},
-describing some of its vulnerabilities.
+In this section, we describe the basic methods for FE and uncertainty
+quantification, focusing on ALE definition. We, then, review the ALE
+approximation\citep{apley2020visualizing, gkolemis22}, describing some
+of its vulnerabilities.
 
@@
 
-In real ML scenarios, the FE is estimated from the training set
+In ML scenarios, ALE is estimated from the training set
 examples. Therefore, \citep{apley2020visualizing} proposed dividing
@@
 
-\section{Accumulated Regional Effects (ARE)}
-\label{sec:NAME-method}
+\section{Uncertainty-Aware ALE (UALE)}
+\label{sec:UALE}
 
-ARE extends ALE for quantifying the level of heterogeneous effects
-(uncertainty) and automates the bin-splitting step for robust
-estimations. At Section~\ref{sec:NAME-definition}, we define ARE and
-at Section~\ref{sec:interval-based-estimation} we provide an
-alternative definition of ARE based on variable-size bins, providing
-important remarks and proofs for connecting it with the initial
-definition. In Section~\ref{sec:interval-spliting}, we define the
-problem of optimal bin-splitting and we propose an algorithmic
-solution to the problem.
+UALE extends ALE for quantifying the level of heterogeneous effects
+(uncertainty) and automates the bin-splitting. At
+Section~\ref{sec:UALE-definition-1} we define UALE, at
+Section~\ref{sec:UALE-definition-2} we redefine it using an
+interval-based formulation, at
+Section~\ref{sec:UALE-approximation} we provide an
+approximation and at Section~\ref{sec:bin-spliting} we define and
+solve the problem of optimal bin-splitting.
 
-\subsection{ARE: ALE With Uncertainty Quantification}
-\label{sec:NAME-definition}
+\subsection{UALE: ALE With Uncertainty Quantification}
+\label{sec:UALE-definition-1}
 
@@
 \noindent
-ARE method formulates the effect at a specific point \(x_s\) with a
-compact tuple that consists of the average effect and the uncertainty
-\((\mu(z), \sigma(z))\) and visualizes them as a continuous curve
-with a confidence region, i.e.
-\(f^{\mathtt{ARE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
+UALE formulates the effect at a specific point \(x_s\) with a compact
+tuple that consists of the average effect and the uncertainty
+\((\mu(z), \sigma(z))\) and visualizes them as a continuous curve with
+a confidence region, i.e.
+\(f^{\mathtt{UALE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
 f^{\mathtt{ALE}}_{\sigma}(x_s)\).
@@
 \subsection{Interval-Based Formulation and Approximation}
-\label{sec:interval-based-estimation}
+\label{sec:UALE-definition-2}
 
-In real scenarios, the estimations are based on the limited instances
-of the training set. Estimating \(\mu(z), \sigma(z)\) at the
-granularity of a point is impossible, because the probability of
-observing a sample inside the region \([x_s - h, x_s + h]\) tends to
-zero, when \(h \to 0\). Therefore we define the regional-effect
-\(\mu(z_1, z_2)\) and the regional-uncertainty \(\sigma(z_1, z_2)\),
-as the summary statistics that characterize the effect and the
-uncertainty inside the interval \([z_1, z_2)\):
+In real scenarios, all estimations are based on the instances of the
+training set. It is impossible to estimate \(\mu(z), \sigma(z)\) at
+the granularity of a point, because the probability of observing a
+sample inside the region \([x_s - h, x_s + h]\) tends to zero, when
+\(h \to 0\). Therefore we define the regional-effect \(\mu(z_1, z_2)\)
+and the regional-uncertainty \(\sigma(z_1, z_2)\), as the summary
+statistics that characterize the effect and the uncertainty inside the
+interval \([z_1, z_2)\):
 
@@
 %
-The regional-effect and the regional-uncertainty are the mean values
-of the local effects and local uncertainty inside an interval. In
-other words, they quantify what is the expected effect and the
-expected uncertainty of a sample drawn at random from \([z_1, z_2]\)
-given uniform distribution \(\mathcal{U}(z_1, z_2)\).
+Intuitively, the regional-effect and the regional-uncertainty are the
+average effect \(\mu(z^*)\) and the uncertainty \(\sigma(z^*)\) that I
+expect to get if draw a sample at random from the region
+\([z_1, z_2]\) given a uniform distribution
+\(z^* \sim \mathcal{U}(z_1, z_2)\).
 
@@
 intervals, i.e. \(\mathcal{Z}_s = \{z_0, \ldots, z_K\}\), we can
-redefine ARE using an interval-based formulation
-\(\tilde{f}^{\mathtt{ARE}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)
+redefine UALE using an interval-based formulation
+\(\tilde{f}^{\mathtt{UALE}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)
 \pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:
@@
 \subsection{Interval-Based Approximation}
+\label{sec:UALE-approximation}
 
-The mean effect (Eq.~\eqref{eq:mu_bin}) and the
-uncertainty (Eq.~\eqref{eq:var_bin}) can be directly estimated from
-the set \(\mathcal{S}_k\) of dataset instances with the \(s\)-th
-feature lying inside the interval
+For approximating the mean effect (Eq.~\eqref{eq:mu_bin}) and the
+uncertainty (Eq.~\eqref{eq:var_bin}) we use the set \(\mathcal{S}_k\)
+of dataset instances with the \(s\)-th feature lying inside the
+interval
 \( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \} \):
@@
 
-Under the assumption that the points are uniformly distributed inside
-the intervals Eq.\eqref{eq:2} is an unbiased estimatior of
-Eq.~(\ref{eq:mu_bin}). The same does not hold in the general case for
+Eq.\eqref{eq:2} is an unbiased estimator of Eq.~(\ref{eq:mu_bin})
+under the assumption that the points are uniformly distributed inside
+the intervals. The same does not hold in the general case for
 Eq.\eqref{eq:3}. In the general case, \(\hat{\sigma}^2(z_{k-1}, z_k)\)
-is an unbiased estimator of the observable standard deviation
+is an unbiased estimator of the \textit{observable} standard deviation
 \(\sigma_{obs}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
@@
     \right] \partial z}{z_2 - z_1}}\), which as we prove in Theorem 1
-is an overestimation of the real regional variance.
+is an overestimation of \(\sigma(z_1, z_2)\).
 
@@
   between the expected effect at \(x_s\) and the regional effect, i.e
-  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the regional variance
+  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then the regional variance
   \(\sigma^2(z_1, z_2)\) equals to:}
@@
 \subsection{Bin-Splitting: Finding Regions With Homogeneous Effects}
-\label{sec:interval-spliting}
+\label{sec:bin-spliting}
 
@@
 practice~\citep{aas2021explaining, herbinger2022repid} for providing
-secure empirical evidence about the superiority of ARE method.
+secure empirical evidence about the superiority of UALE method.
 
 We split the simulation examples in two groups. The first group,
-Section~\ref{sec:simulation-examples-1}, aims at showcasing that ARE
+Section~\ref{sec:simulation-examples-1}, aims at showcasing that UALE
 method is more accurate than PDP-ICE in quantifying the average effect
@@
 features are correlated. The second group,
-Section~\ref{sec:simulation-examples-2}, illustrates that ARE method
+Section~\ref{sec:simulation-examples-2}, illustrates that UALE method
 achieves a better approximation (average effect and uncertainty) in
@@
 
-In this simulation, we will compare ARE method against PDP-ICE in
+In this simulation, we will compare UALE method against PDP-ICE in
 quantifying the main effect and the uncertainty. Since there is
@@
 \(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test the effect
-computed by ARE and PDP-ICE in three cases; (a) no interaction
+computed by UALE and PDP-ICE in three cases; (a) no interaction
 (\(\alpha=0\)) and equal weights (\(a_1=a_2\)), (b) no interaction
@@
 \(X_2\). Finally, besides it is not the focus of the current examples,
-we will also observe that ARE manages to correctly extract the
+we will also observe that UALE manages to correctly extract the
 regions with constant effect. Details for obtaining the ground truth
@@
   \caption{No interaction, Equal weights: Feature effect for \(x_1\);
-    Ground-truth vs (a) ARE method at the left and (b) PDP-ICE at the
+    Ground-truth vs (a) UALE method at the left and (b) PDP-ICE at the
     right.}
@@
   \caption{No interaction, Different weights: Feature effect for \(x_1\);
-    Ground-truth vs (a) ARE method at the left and (b) PDP-ICE at the
+    Ground-truth vs (a) UALE method at the left and (b) PDP-ICE at the
     right.}
@@
 zero-uncertainty, since it does not appear in any interaction term. As
-we observe in Figure~\ref{fig:ex-synth-1-3}, ARE captures perfectly
+we observe in Figure~\ref{fig:ex-synth-1-3}, UALE captures perfectly
 the effect of all features, and the uncertainty in all cases. As
@@
     features, \(x_1\) to \(x_3\) from top to bottom; Ground-truth vs
-    (a) ARE method at the left columns and (b) PDP-ICE at the right
+    (a) UALE method at the left columns and (b) PDP-ICE at the right
     column.}
@@
 sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). As we observe, in
-Figure~\ref{fig:ex-synth-2-1}, ARE method extracts a sequence of
+Figure~\ref{fig:ex-synth-2-1}, UALE method extracts a sequence of
 intervals with better \(\mathcal{L}_{\mathtt{DP}}^{\mu}\) and
@@
 purpose of this part is to demonstrate the usefulness of uncertainty
-quantification in ARE, i.e., illustrate that modeling the
+quantification in UALE, i.e., illustrate that modeling the
 heterogeneous effects is important for the interpretation of an ALE
-plot. For this reason, Secondly, we discuss about the FE plots obtaing with ARE and
+plot. For this reason, Secondly, we discuss about the FE plots obtaing with UALE and
 PDP-ICE.
@@
 
-For evaluating ARE in terms of bin-splitting we set-up the following
+For evaluating UALE in terms of bin-splitting we set-up the following
 evaluation framework. We treat as ground-truth the effects computed on
@@
 truth. Afterwards we select less samples \(N=1000\) and we compare
-ARE approximation with all possible fixed-size alternatives, in terms
+UALE approximation with all possible fixed-size alternatives, in terms
 of accuracy, i.e. we provide the quantities

### 2022-10-05 AISTATS2023/paper.tex
resolved Christos changes in chapters 1, 2

@@
 Accumulated Local Effects (ALE) is a popular approach to explainable AI that describes how a feature influences the decisions of a model on average, taking into account potential correlations between the features.
-However, ALE does not provide an explicit component for modeling the uncertainty of the effect, i.e., the level of heterogeneous effects behind the average explanation, which is crucial for a complete interpretation when the black-box function contains interactions between the correlated features.
+However, ALE does not provide an explicit component for modeling the uncertainty of the effect, i.e., the level of heterogeneity of the effects behind the average explanation, which is crucial for a complete interpretation when the black-box function contains interactions between the correlated features.
 In this work, we propose Uncertainty-aware ALE (UALE), an extension of ALE for quantifying the heterogeneous effects.
@@
 
-Recently, Machine Learning (ML) has flourished in critical domains,
+Recently, Machine Learning (ML) has been adopted in critical domains,
 such as healthcare and finance. In these areas, we need a combination
@@
 and global explainability
-techniques~\citep{Molnar2020interpretable}. Local methods explain a
-specific prediction, whereas global methods explain the entire model
-behavior. Global methods provide a universal explanation, summarizing
-the various local explanations into a single interpretable outcome,
-usually a number or a plot. If a user wants to get a rough overview
+techniques~\citep{Molnar2020interpretable}. Local methods explain the prediction of a
+specific instance, whereas global methods provide an explanation that summarizes the entire model behavior.
+Global methods provide a universal explanation, aggregating the various local explanations into a single interpretable outcome,
+usually a number or a plot. If users want to get a rough overview
 about which features are significant (feature importance) or whether a
@@
 
-Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental global
-explainability category. The objective of FE is to isolate the impact
+Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category of global
+explainability methods. The objective of FE is to isolate the impact
 of a single feature on the output.~\footnote{Often, FE methods also
   isolate the combined effect of a pair of features to the
-  output. Combinations of more than two features are not usual,
-  because they encounter, among others, visualization difficulties.}
-FE methods suffer from aggregation bias because, often, the rationale
-behind the average effect is unclear. For example, a feature with zero
+  output. Combinations of more than two features are uncommon, since they are difficult to estimate and visualize.}
+FE methods suffer from aggregation bias because, often, implications of the average effect to the model behavior are unclear. For example, a feature with zero
 average effect may indicate no effect on the output or, in contrast,
-highly positive in some cases and highly negative in others. There are
+highly positive in some cases and highly negative in others.
+
+There are
 three widely-used FE methods; Partial Dependence Plots
@@
 correlated, which is a frequent scenario in many ML cases. Therefore,
-ALE has been established as the state-of-the-art FE method.
-
-However, ALE faces two crucial limitations, the first concerns ALE
+ALE has been established as the state-of-the-art FE method. However, ALE faces two crucial limitations, the first concerns ALE
 definition and the second one ALE approximation. Regarding the
 definition ALE does not inform the user about the uncertainty, i.e.,
-the level of heterogeneous effects hidden behing the average effect
+the level of heterogeneous effects hidden behind the average effect
 due to implicit feature interactions. In contrast, in PDPs the
-heterogeneous effects can be (partially) spotted by exploring the
+heterogeneous effects can be identified by exploring the
 Individual Conditional Expectations
 (ICE)\citep{goldstein2015peeking}. Regarding the approximation,
-i.e.~the estimation usign the (limited) samples of the training set,
+i.e.~the estimation using the (limited) samples of the training set,
 ALE requires the additional \textit{bin-splitting} step. Bin-splitting
-consists of partitioning the axis of the feature of interest in a
-sequence of non-overlapping intervals (regions) and estimating a
-constant effect in each one from the population of samples that fall
-inside. Specifying an approriate sequence of regions is of particular
+consists of partitioning the domain of the feature of interest into consecutive non-overlapping intervals (regions) and estimating a
+constant effect in each region. The effect is estimated from the population of samples that fall
+inside the region. Specifying an appropriate set of bins is of particular
 importance, since ALE's interpretation is meaningful only inside each
@@
 
-In this paper, we present Uncertainty-awary Accumulated Local Effects
+In this paper, we present Uncertainty-aware Accumulated Local Effects
 (UALE) an extension of ALE for measuring the uncertainty of the
-explanation, i.e.~how certain we are that the average explanation is
-valid if applied to an instance drawn at random. We show that choosing
+explanation, i.e.~how certain we are that the average explanation would predict the feature effect to an instance drawn at random. We show that choosing
 an appropriate sequence of intervals is particularly important for a
@@
 
-\paragraph{Contributions.} The contribution of this paper are:
+\paragraph{Contributions.} The contributions of this paper are:
 
@@
 examples. Therefore, \citep{apley2020visualizing} proposed dividing
-the \(s\)-th axis in \(K\) bins and estimating the local effects in
+the feature domain in \(K\) equal-sized bins and estimating the local effects in
 each bin by evaluating the black box-function at the bin limits:
@@
 bins, denoted with \(K\), for splitting the axis into equally-sized
-bins, an approach with crucial limitations. Setting \(K\) to a small
-value may hide fine-grain effects due to large bins and setting \(K\)
+bins. This approach introduces significant limitations. Specifically, setting \(K\) to a small
+value may hide fine-grain effects due to large bins while setting \(K\)
 to a high value leads to poor bin-effect estimations from limited
 samples. In general, the user may face contradictory explanations for
-different \(K\) without a clue to decide which one to trust.
+different \(K\) without a clue for which one to trust.
 

### 2022-10-05 AISTATS2023/paper.tex
updated method description

@@
 \begin{abstract}
-Accumulated Local Effects (ALE) is a popular approach to explainable AI that describes how a feature influences the decisions of a model on average, taking into account potential correlations between the features.
-However, ALE does not provide an explicit component for modeling the uncertainty of the effect, i.e., the level of heterogeneity of the effects behind the average explanation, which is crucial for a complete interpretation when the black-box function contains interactions between the correlated features.
-In this work, we propose Uncertainty-aware ALE (UALE), an extension of ALE for quantifying the heterogeneous effects.
-We show the importance of bin-splitting for a robust approximation of the uncertainty and we formally prove the conditions for an unbiased estimation.
-We finally propose a computationally-grounded algorithm for finding the optimal sequence of bins given the limited instances of the training set.
-We demonstrate through synthetic and real datasets, that UALE quantifies the average and heterogeneous effects correctly and approximates them robustly by optimizing the bin-splitting procedure.
+  Accumulated Local Effects (ALE) is a popular approach to explainable
+  AI that describes how a feature influences the decisions of a model
+  on average, taking into account potential correlations between the
+  features.  However, ALE does not provide an explicit component for
+  modeling the uncertainty of the effect, i.e., the level of
+  heterogeneity of the effects behind the average explanation, which
+  is crucial for a complete interpretation when the black-box function
+  contains interactions between the correlated features. In this work,
+  we define Uncertainty-aware ALE (UALE), an extension of ALE for
+  quantifying the heterogeneous effects and we propose a method for
+  estimating it. We formally prove the conditions the bin-splitting,
+  i.e.~the partitioning the domain of the feature of interest into
+  consecutive non-overlapping intervals, must respect for an an
+  unbiased estimation of the uncertainty. We finally propose a
+  computationally-grounded algorithm for finding the optimal sequence
+  of bins given the limited instances of the training set. We
+  demonstrate through synthetic and real datasets, that UALE
+  quantifies the average and heterogeneous effects correctly and
+  approximates them robustly by optimizing the bin-splitting
+  procedure.
 \end{abstract}
@@
 (UALE) an extension of ALE for measuring the uncertainty of the
-explanation, i.e.~how certain we are that the average explanation would predict the feature effect to an instance drawn at random. We show that choosing
-an appropriate sequence of intervals is particularly important for a
-robust estimation of the uncertainty, providing an algorithm that
-automates this step. We also provide theoretical proofs for an
-unbiased estimation of the uncertainty. We apply our method in multiple
+explanation, i.e.~how certain we are that the average explanation
+would predict the feature effect to an instance drawn at random. We
+show that choosing an appropriate sequence of intervals is
+particularly important for a robust and unbiased estimation of the
+uncertainty and we provide an algorithm that automates this step. We
+also formally prove the bin-splitting conditions for an unbiased
+estimation of the uncertainty. We apply our method in multiple
 synthetic and real world datasets for evaluating its performance.
@@
 uncertainty of the local effects at a specific point \(x_s=z\) and we
-quantify it as the standard deviation of the local explanations:
+quantify it as the standard deviation of the local explanations
+\(\mathcal{H}(z) = \sigma(z)\), where:
+
 \begin{equation}
   \label{eq:ALE_var}
-  \mathcal{H}(z) := \sigma(z) = \sqrt{\mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] }
+  \sigma^2(z) = \mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] 
 \end{equation}
@@
 In real scenarios, all estimations are based on the instances of the
-training set. It is impossible to estimate \(\mu(z), \sigma(z)\) at
+training set. Therefore, we cannot estimate \(\mu(z), \sigma(z)\) at
 the granularity of a point, because the probability of observing a
 sample inside the region \([x_s - h, x_s + h]\) tends to zero, when
-\(h \to 0\). Therefore we define the regional-effect \(\mu(z_1, z_2)\)
-and the regional-uncertainty \(\sigma(z_1, z_2)\), as the summary
-statistics that characterize the effect and the uncertainty inside the
-interval \([z_1, z_2)\):
+\(h \to 0\). We define the regional-effect \(\mu(z_1, z_2)\) and the
+regional-uncertainty \(\mathcal{H}(z_1, z_2) = \sigma(z_1, z_2)\), as the summary statistics
+that characterize the effect and the uncertainty inside the interval
+\([z_1, z_2)\):
 
@@
   \label{eq:mu_bin}
-  \mu(z_1, z_2) = \frac{1}{z_2 - z_1} \int_{z_1}^{z_2}
-  \mathbb{E}_{\xc|z}\left [f^s(z, \xc) \right ] \partial z
+    \mu(z_1, z_2) = \mathbb{E}_{z \sim \mathcal{U}(z_1,z_2)} [\mu(z)]
+    = \frac{\int_{z_1}^{z_2} \mu(z) \partial z}{z_2 - z_1}
 \end{equation}
-
-\noindent
 
@@
   \label{eq:var_bin}
-  \sigma(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
-  \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z) )^2 \right] \partial z}{z_2 - z_1}}
+  \sigma^2(z_1, z_2) = \mathbb{E}_{z \sim \mathcal{U}(z_1,z_2)} [\sigma^2(z)] =  \frac{\int_{z_1}^{z_2} \sigma^2(z)  \partial z}{z_2 - z_1}
 \end{equation}
@@
 %
-Intuitively, the regional-effect and the regional-uncertainty are the
-average effect \(\mu(z^*)\) and the uncertainty \(\sigma(z^*)\) that I
-expect to get if draw a sample at random from the region
-\([z_1, z_2]\) given a uniform distribution
-\(z^* \sim \mathcal{U}(z_1, z_2)\).
-
-If we also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points
-that partition the \(s\)-th feature axis into \(K\) variable-size
-intervals, i.e. \(\mathcal{Z}_s = \{z_0, \ldots, z_K\}\), we can
-redefine UALE using an interval-based formulation
-\(\tilde{f}^{\mathtt{UALE}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)
-\pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:
+Intuitively, the regional-effect and the regional-uncertainty quantify
+the expected average effect and the expected uncertainty if we
+randomly draw a point \(z^*\) given a uniform distribution
+\(z^* \sim \mathcal{U}(z_1, z_2)\). If we also define as
+\(\mathcal{Z}\) the sequence of \(K+1\) points that partition the
+domain of the \(s\)-th feature into \(K\) variable-size intervals,
+i.e.  \(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can redefine UALE
+using an interval-based formulation
+\(\tilde{f}^{\mathtt{UALE}}_{\mathcal{Z}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s)
+\pm \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s)\), where:
 
@@
   \label{eq:ALE_2}
-  \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
+  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
 \end{equation}
@@
   \label{eq:ALE_accumulated_var}
-  \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
+  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
 \end{equation}
@@
 Eq.\eqref{eq:3}. In the general case, \(\hat{\sigma}^2(z_{k-1}, z_k)\)
-is an unbiased estimator of the \textit{observable} standard deviation
-\(\sigma_{obs}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
+is an unbiased estimator of the \textit{observable} variance
+\(\sigma^2_{obs}(z_1, z_2) = \frac{\int_{z_1}^{z_2}
     \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2
-    \right] \partial z}{z_2 - z_1}}\), which as we prove in Theorem 1
-is an overestimation of \(\sigma(z_1, z_2)\).
+    \right] \partial z}{z_2 - z_1}\), which as we prove in Theorem 1
+is an overestimation of \(\sigma^2(z_1, z_2)\).
 
@@
     \label{eq:bin-uncertainty-proof}
- \sigma_{obs}^2(z_1, z_2) = \sigma^2(z) + \frac{\int_{z_1}^{z_2}\rho^2(z) \partial z}{z_2 - z_1}
+ \sigma_{obs}^2(z_1, z_2) = \sigma^2(z_1, z_2) + \frac{\int_{z_1}^{z_2}\rho^2(z) \partial z}{z_2 - z_1}
 \end{equation}
@@
 Theorem 1 reveals an important connection between the ground-truth and
-the observable uncertainty inside a region. Setting as ground-truth
-the regional uncertainty
-\(\mathcal{H}_{true}(z_1, z_2) = \sigma(z_1, z_2)\) and as observable
-uncertainty the one we estimate with Eq.~(\ref{eq:3}), i.e.
-\(\mathcal{H}_{obs}(z_1, z_2) := \sigma_{obs}(z_1, z_2)\), then there
-is an error term quantified by the residual inside the interval, i.e.
-\(\mathcal{E}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}\rho^2(z)
-    \partial z}{z_2 - z_1}}\), such that:
+the observable uncertainty inside a region. For convenience, we denote
+as
+\(\mathcal{E}^2(z_1, z_2) = \frac{\int_{z_1}^{z_2}\rho^2(z) \partial
+  z}{z_2 - z_1}\) the error term that models the expected square
+residual inside the interval and as
+\(\mathcal{H}_{obs}(z_1, z_2) := \sigma_{obs}(z_1, z_2)\) the
+observable uncertainty. It holds that the observable uncertainty is an
+overestimation of the correct uncertainty
 
@@
   \label{eq:uncertainty-bin}
-  \mathcal{H}^2_{obs}(z_1, z_2) = \mathcal{H}^2_{true}(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
+  \mathcal{H}^2_{obs}(z_1, z_2) = \mathcal{H}^2(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
 \end{equation}
 %
-Eq.~(\ref{eq:uncertainty-bin}) shows that the error term quantifies
-how much the observable uncertainty of the region \([z_1, z_2]\)
-deviates for the one we would like to measure and provides a
-quantified metric to evaluate a particular splitting. As we prove in
-the Appendix, UALE uncertainty approximation is unbiased if and only if the
-partitioning is such that ther error term is zero for all bins, i.e. 
-\(\sum_{k=1}^K \mathcal{E}^2(z_{k-1}, z_{k}) = 0\)
+and, therefore, the estimation is unbiased only in case
+\(\mathcal{E}^2(z_1, z_2) = 0\).
 
@@
 In this section we formulate bin-splitting as an unsupervised
-clustering problem. The main idea of the clustering problem is to find
-the best bin-splitting solution, handling appropriately two
-conflicting objectives. On the one hand, we want to minimize the error
-term \(\mathcal{E}\) defined above for minimizing the error in the
-uncertainty estimation and on the other hand, we want wide-enough bins
-for including a fair population of samples for a robust estimation of
+clustering problem, searching for a solution that compromises two
+contradictory objectives. On the one hand, we want to minimize the
+error term \(\mathcal{E}\) for approaching an unbiased estimation and
+on the other hand, we want wide-enough bins for including a fair
+population of samples for a robust estimation of
 \(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\). We set-up the
@@
 We search for the sequence of intervals \(z_0, \ldots, z_K\) that
-minimizes the sum of bin costs. The cost of each bin is the
+minimizes the sum of the bin costs. The cost of each bin is the
 approximated variance \(\hat{\sigma}^2_k\) scaled by the bin length
 \(\Delta z_k\) and discounted by the term \(\tau_k\). The term
-\(\tau_K\) acts as a tie-braker that favors the selection of a bigger
-bin in case it has similar variance with the aggregate variance of
-many smaller ones. The constraint of at least \(N_{\mathtt{PPB}}\)
-points per bin sets the lowest-limit for a \textit{robust}
-estimation. The user can choose to what extent they favor the creation
-of wide bins through the discount parameter by setting the parameter
-\(\alpha\) and where they set the minimum for robust approximation
-with the parameter \(N_{\mathtt{PPB}}\). For providing a rough idea of
-some sensible value, in our experiments we set \(\alpha = 0.2\) which
-means that the discount ranges in \([0\%, 20\%]\) and
-\(N_{\mathtt{PPB}} = \frac{N}{20}\). It is also important to clarify
-that by minimizing \(\hat{\sigma}^2_k \approx \mathcal{H}_{obs}^2\),
-we actually minimize the squared error term \(\mathcal{E}^2\), since
-the term \(\mathcal{H}^2_{true}\) is independent of the bin splitting
-procedure.
+\(\tau_K\) favors the selection of a bigger bin in case it has similar
+variance with the aggregate variance of many smaller ones. The
+constraint of at least \(N_{\mathtt{PPB}}\) points per bin sets the
+lowest-limit for a \textit{robust} estimation. The user can choose to
+what extent they favor the creation of wide bins through the discount
+parameter by setting the parameter \(\alpha\) and where they set the
+minimum for robust approximation with the parameter
+\(N_{\mathtt{PPB}}\). For providing a rough idea, in our experiments
+we set \(\alpha = 0.2\) which means that the discount ranges between
+\([0\%, 20\%]\) and \(N_{\mathtt{PPB}} = \frac{N}{20}\). It is also
+important to clarify that by minimizing
+\(\hat{\sigma}^2_k \approx \mathcal{H}_{obs}^2\), we actually minimize
+the squared error term \(\mathcal{E}^2\), since the term
+\(\mathcal{H}^2_{true}\) is independent of the bin splitting.
 

### 2022-10-05 examples/real_dataset_3.py

@@ -172,7 +172,7 @@ for s in range(8):
                                                          model_grad,
                                                          axis_limits,
                                                          K_list,
-                                                         nof_iterations=10,
+                                                         nof_iterations=30,
                                                          nof_points=1000,
                                                          feature=s)
     stats_fixed_list.append(stats_fixed)
@@ -181,7 +181,7 @@ for s in range(8):
                                                        model,
                                                        model_grad,
                                                        axis_limits,
-                                                       nof_iterations=10,
+                                                       nof_iterations=30,
                                                        nof_points=1000,
                                                        feature=s)
     stats_auto_list.append(stats_auto)

### 2022-10-05 examples/utils.py

@@ -37,15 +37,15 @@ def compare(dale_gt, dale):
         means = dale_gt["bin_effect"][ind[i]+1:ind[i+1]]
         variances = dale_gt["bin_variance"][ind[i]+1:ind[i+1]]
 
-        rho_bin = means.var()
+        rho_bin = means.std()
         rho_list.append(rho_bin)
 
         mu_gt = means.mean()
         mu_est = dale["bin_effect"][i]
         mu_err_list.append(np.abs(mu_gt - mu_est))
 
-        var_gt = variances.mean()
-        var_est = dale["bin_variance"][i]
+        var_gt = np.sqrt(variances.mean())
+        var_est = np.sqrt(dale["bin_variance"][i])
         var_err_list.append(np.abs(var_gt - var_est))
 
     mean_rho = np.mean(rho_list)

### 2022-10-05 feature_effect/pdp.py

@@ -41,7 +41,7 @@ class FeatureEffectBase:
         func = partial(self.eval_unnorm, s=s, uncertainty=False)
         start = self.axis_limits[0, s]
         stop = self.axis_limits[1, s]
-        z = utils_integrate.normalization_constant_1D(func, start, stop)
+        z = utils_integrate.normalization_constant_1D_linspace(func, start, stop)
         return z
 
     def fit(self,

### 2022-10-05 feature_effect/utils_integrate.py

@@ -17,6 +17,17 @@ def normalization_constant_1D(func, start, stop):
     y = integrate.quad(wrapper, start, stop)[0]
     return y
 
+def normalization_constant_1D_linspace(func, start, stop):
+    """Computes normalization constant of PDP of feature
+
+    :param s: index of feature
+    :returns:
+
+    """
+    x = np.linspace(start, stop, 1000)
+    z = np.mean(func(x))
+    return z
+
 
 def expectation_1D(xs, func, p_xc, s, start=-np.inf, stop=np.inf):
     """

### 2022-10-05 AISTATS2023/paper.tex
updated real example part

@@
 
-The simulation examples, where the data-generating distribution
-\(p(\mathbf{X})\) and the predictive function \(f(\cdot)\) are defined
-by us, enable the evaluation of competitive approaches against a solid
-ground-truth. We follow this common XAI
-practice~\citep{aas2021explaining, herbinger2022repid} for providing
-secure empirical evidence about the superiority of UALE method.
-
-We split the simulation examples in two groups. The first group,
-Section~\ref{sec:simulation-examples-1}, aims at showcasing that UALE
-method is more accurate than PDP-ICE in quantifying the average effect
-and the level of heterogeneous effects (uncertainty), when input
-features are correlated. The second group,
-Section~\ref{sec:simulation-examples-2}, illustrates that UALE method
-achieves a better approximation (average effect and uncertainty) in
-cases of limited samples, due to automatic bin-splitting. In both
-groups, we choose to illustrate the most indicative examples; a more
-extensive evaluation is provided in the Appendix.
+Simulation examples provide the freedom to select the data-generating
+distribution \(p(\mathbf{X})\) and the predictive function
+\(f(\cdot)\). We follow this common XAI
+practice~\citep{aas2021explaining, herbinger2022repid} for empirically
+evaluating UALE against a solid ground-truth. The first group of
+synthetic examples (Section~\ref{sec:simulation-examples-1}) compares
+UALE with PDP-ICE plots in quantifying the average and the uncertainy
+of the effect. The second group
+(Section~\ref{sec:simulation-examples-2}) evaluates the accuracy of
+the approximation (average effect and uncertainty) using the automatic
+bin-splitting algorithm. In both groups, we illustrate the most
+indicative examples; a more extensive evaluation is provided in the
+Appendix.
 
@@
 In this simulation, we will compare UALE method against PDP-ICE in
-quantifying the main effect and the uncertainty. Since there is
-ambiguity about the ground-truth first-order effect in cases of
-correlated features,
-e.g.~\citep{apley2020visualizing,Gromping2020MAEP}, we limit ourselves
-to the following piecewise linear function,
+quantifying the main effect and the uncertainty. The samples that we
+use in all examples are coming from the following distribution:
+\(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
+\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
+\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We create the following
+piecewise linear function:
+
+% Since there is
+% ambiguity about the ground-truth first-order effect in cases of
+% correlated features,
+% e.g.~\citep{apley2020visualizing,Gromping2020MAEP},
 
@@
 where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
-\(f_{int}(\mathbf{x}) = x_1x_3\). As we observe, the linear term
-\(f_{\mathtt{lin}}\) includes the two correlated features and the term
-\(f_{\mathtt{int}}\) interacts the two non-correlated variables. The
-samples that we use in all examples are coming from the following
-distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
-\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
-\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test the effect
-computed by UALE and PDP-ICE in three cases; (a) no interaction
-(\(\alpha=0\)) and equal weights (\(a_1=a_2\)), (b) no interaction
-(\(\alpha=0\)) and different weigths (\( a_1 \neq a_2 \)) and (c) with
-interaction (\(\alpha \neq 0\)) with equal weights (\(a_1=a_2\)).
+\(f_{\mathtt{int}}(\mathbf{x}) = x_1x_3\). The linear term
+\(f_{\mathtt{lin}}\) includes the two correlated features and
+\(f_{\mathtt{int}}\) interacts the two non-correlated features.  We
+evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
+without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
+(b) without interaction (\(\alpha=0\)) and different weigths
+(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha \neq 0\)) with
+equal weights (\(a_1=a_2\)).
 
-In all cases, we firstly compute the ground-truth average effect and
-the uncertainty analytically (proofs in the Appendix) and then we
-compare it against the approximation provided by each method. The
-approximation is computed after generating \(N=300\) samples. As we
-will see, despite the model's simplicity, PDP-ICE fail in quantifying
-the effect correctly, due to correlation between features \(X_1\) and
-\(X_2\). Finally, besides it is not the focus of the current examples,
-we will also observe that UALE manages to correctly extract the
-regions with constant effect. Details for obtaining the ground truth
-effects and the experimental set-up are in the Appendix.
+In all cases, we provide the ground-truth average effect and
+uncertainty computed with analytical derivations (proofs at the
+Appendix) and we compare it against the approximation provided by each
+method. The approximation is computed after generating \(N=500\)
+samples.
 
@@
 
-We test the feature effect when no interaction term is apparent,
-i.e. \(\alpha=0\), and the weights are equal \(a_1=a_2=1\).  In this
-case, the ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\)
-when \(0 \leq x_1 < 0.25\), \(-x_1\) when \(0.25 \leq x_1 < 0.5\) and
-zero otherwise. In each position, there is total absence of
-heterogeneous effects; therefore, the uncertainty of the global
-explanation is zero \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
+We test the feature effect when no interaction exits \(\alpha=0\) and
+the weights are equal \(a_1=a_2=1\). The ground truth effect
+\(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\) when \(0 \leq x_1 < 0.25\),
+\(-x_1\) when \(0.25 \leq x_1 < 0.5\) and zero otherwise. There zero
+heterogeneous effects, therefore
+\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
 Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP main effect is
@@
 contrast, ALE captures correctly the average effect and the zero
-uncertainty and it also groups perfectly the regions with
-constant-effect.
+uncertainty and it also groups perfectly the constant-effect regions.
 
@@
 
-As before, there is no interaction term and, therefore, the
-ground-truth is zero, i.e.  \(f^{\mathcal{GT}}_{\sigma^2}(x_1) =
-0\). The weigts are different \(a_1=2, a_2=0.5\), therefore, the
-ground-truth effect is \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) when
-\(0 \leq x_1 < 0.2\), \(-2x_1\) when \(0.2 \leq x_1 < 0.4\) and zero
-otherwise. In Figure~\ref{fig:ex-synth-1-2}, we observe that PDP
-estimation is completely opposite to the ground-truth effect,
-i.e.~negative in \([0, 0.2)\) and positive in \([0.2, 0.4)\), and the
-ICE erroneously implies heterogeneous effects.  As before, ALE
-quantifies perfectly the ground truth effect and the zero-uncertainty,
-extracting correcltly the constant effect regions.
+As before, no interaction exits \(\alpha=0\), therefore, the
+ground-truth uncertainty is zero, i.e.
+\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). The weigts are
+\(a_1=2, a_2=0.5\), therefore, the ground-truth effect is
+\(f_\mu^{\mathtt{GT}}(x_1)\) is, \(2x_1\) when \(0 \leq x_1 < 0.2\),
+\(-2x_1\) when \(0.2 \leq x_1 < 0.4\) and zero otherwise. In
+Figure~\ref{fig:ex-synth-1-2}, we observe that PDP estimation is
+completely opposite to the ground-truth effect, i.e.~negative in the
+region \([0, 0.2)\) and positive in \([0.2, 0.4)\), and the ICE
+erroneously implies the existence of heterogeneous effects. As before,
+ALE quantifies correctly the ground truth effect, the
+zero-uncertainty and extracts the constant effect regions.
 
@@
 In this case we activate the interaction term, i.e. \(a=1\), and we
-set equal weights \(a_1=a_2=1\). The interaction term provokes
-heterogeneous effects in features \(x_1, x_3\), i.e., at any position
-\(x_1\), the local effects dependend on the unknown value of \(X_3\)
-and vice-versa. The effect of \(x_2\) continues to have
-zero-uncertainty, since it does not appear in any interaction term. As
-we observe in Figure~\ref{fig:ex-synth-1-3}, UALE captures perfectly
-the effect of all features, and the uncertainty in all cases. As
-expected, PDP-ICE quantify the effect and the uncertainty correctly
-only in the case of \(x_3\), since \(X_3\) is independent from other
-features. For the correlated features, i.e. \(x_1, x_2\), the average
-effect computed by PDP and the uncertainty implied by ICE plots are
-wrong.
+set the weights to \(a_1=a_2=1\). The interaction term provokes
+heterogeneous effects for features \(x_1, x_3\), because the local
+effects at \(x_1\) depend on the unknown value of \(X_3\) and
+vice-versa. The effect of \(x_2\) has zero-uncertainty since it does
+not appear in any interaction term. As we observe in
+Figure~\ref{fig:ex-synth-1-3}, UALE correcly models the average effect
+and the uncertainty in all cases. PDP-ICE quantifies correctly the
+effect and the uncertainty only in the case of feature \(x_3\),
+because \(X_3\) is independent from other features. For features
+\(x_1, x_2\) that are correlated the average effect of PDP
+and the uncertainty of ICE plots are wrong.
 
@@
 
-% \paragraph{Discussion.}
+\paragraph{Discussion.}
 
-% In purpose, the examples above do not cover the case where there is an
-% interaction term includes correlated features, i.e. \(x_1x_2\). In
-% this scenario, there is an open debate about the ground-truth
-% effect~\citep{Gromping2020MAEP}.
+Despite the model's simplicity, PDP-ICE fail in quantifying the
+average effect and the uncertainty due to correlation between features
+\(X_1\) and \(X_2\). UALE quantifies them correctly and estimates them
+accurately extracting the regions with constant effect. The examples
+above do not cover the case of an interaction term between correlated
+features, e.g. \(x_1x_2\), because there is an open debate about the
+ground-truth effect in this case~\citep{Gromping2020MAEP}.
 
@@
 In this simulation, we aim to to quantify the advantages of automatic
-bin-splitting based on the objective of Eq.(\ref{eq:opt}), through the
-following validation framework. We generate a big dataset with dense
-sampling (\(N=10^6\)) and we treat the DALE estimation with dense
-fixed-size bins (\(K=10^3\)) as ground-truth. Afterwards, we generate
-less samples (\(N=500\)) and we compare (a) the fixed-size DALE
-estimation for many different \(K\) versus (b) the auto-bin splitting
-algorithm. In all cases, we use a bivariate black-box function
-\(f(\cdot)\) where the samples are instances of the distribution
-\(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
+bin-splitting. We set-up the following validation framework. We
+generate a big dataset with dense sampling (\(N=10^6\)) and we treat
+the DALE estimation with dense fixed-size bins (\(K=10^3\)) as
+ground-truth. Afterwards, we generate less samples (\(N=500\)) and we
+compare the fixed-size DALE estimation (for many different \(K\))
+versus the automatic bin-splitting. In all cases, we use a bivariate
+black-box function \(f(\cdot)\) where the samples are instances of the
+distribution \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
 \(x_1 \sim \mathcal{U}(0,1)\) and
-\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\).
-
-We denote as
+\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
 \(\mathcal{Z^{\mathtt{DP}}} = \{z^{\mathtt{DP}}_{k-1}, \cdots,
 z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic
-bin-splitting based on the optimisation problem of Eq.~(\ref{eq:opt})
-and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size splitting with
-\(K\) bins. The evaluation is done with regard to two metrics;
+bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size
+splitting with \(K\) bins.  For consistent results, in all the
+examples below, we regenerate samples and repeat the computations for
+\(t = 30\) times, providing the mean value for all metrics.
+
+\paragraph{Metrics}
+
+The evaluation is done with regard to two metrics that count the mean
+bin error, where bin error is the absolute difference of the
+approximation from the ground-truth. The metric
 \(\mathcal{L}_{\mathtt{DP|K}}^{\mu} =
 \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
-  \mathcal{Z}^{\mathtt{DP|K}}} | \mu_k - \hat{\mu}_k | \) quantifies
-the average error in estimating the the expected effect and
+  \mathcal{Z}^{\mathtt{DP|K}}} | \mu_k - \hat{\mu}_k | \) counts the
+mean bin error on the average effect and
 \(\mathcal{L}_{\mathtt{DP|K}}^{\sigma} =
 \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
-  \mathcal{Z}^{\mathtt{DP|K}}} | \sigma_k - \hat{\sigma}_k | \)
-the average error in estimating the variance of the effect. The metric
+  \mathcal{Z}^{\mathtt{DP|K}}} | \sigma_k - \hat{\sigma}_k | \) the
+mean bin error on the uncertainty. In both cases the ground truth is
+computed from the dense bins of ground-truth that overlapse with the
+wider bin of the approximation. We also provide the mean (per bin)
+error term
 \(\mathcal{L}^{\rho}_{\mathtt{DP|K}} =
 \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
-  \mathcal{Z}^{\mathtt{DP|K}}} \rho_k \) quantifies the average
-nuisance uncertainty. The optimal bin-splitting is the one that
-minimises, at the same time, both \(\mathcal{L}^{\mu}\) and
-\(\mathcal{L}^{\sigma^2}\) errors. For consistent results, in all the
-examples below, we regenerate samples and repeat the computations for
-\(t = 10\) times, providing the mean value for all metrics.
+  \mathcal{Z}^{\mathtt{DP|K}}} \rho_k \) for a complete understanding
+of the uncertainty error.
 
@@
 the median house value inside the block. We exclude instances with
-missing or outlier values, i.e. \(x_d^i > \hat{sigma}_j\) in any
+missing or outlier values, i.e. \(x_d^i > \hat{\sigma}_j\) in any
 feature and normalize them to have
 \(\hat{\mu}_d = 0, \hat{\sigma_d}\). We split the dataset into
-\(N_{tr} = 15639\) training and \(N_test = 3910\) test examples (80/20
+\(N_{tr} = 15639\) training and \(N_{test} = 3910\) test examples (80/20
 split) and we fit a Neural Network with 3 hidden layers of 256, 128
@@
 with learning rate \(\eta = 0.02\), the model achieves a normalized
-mean squared error (R-Squared) of \(0.25\) (0.75). In depth information
-about preprocessing, training and evaluation are provided in the
-Appendix.
+mean squared error (R-Squared) of \(0.25\) (0.75). In depth
+information about preprocessing, training and evaluation are provided
+in the Appendix.
 
@@
 In real-world datasets, it is infeasible to obtain the ground truth FE
-for seamlessly evaluating the competitive methods. Therefore, the main
-purpose of this part is to demonstrate the usefulness of uncertainty
-quantification in UALE, i.e., illustrate that modeling the
-heterogeneous effects is important for the interpretation of an ALE
-plot. For this reason, Secondly, we discuss about the FE plots obtaing with UALE and
-PDP-ICE.
+for seamlessly evaluating the competitive methods. For this reason, we
+selected the California Dataset, where, in broad-terms, UALE and
+PDP-ICE aggree in the quantification of the average effect and of the
+uncertainty. In Figure\ref{fig:ex-real-1}, we observe that UALE
+complements ALE for the quantification of the heterogeneous effects,
+similarly to as ICE complements PDP. Add some content..
 
@@
   \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_pdp_ice.pdf}\\
-  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_ale_auto.pdf}
-  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_pdp_ice.pdf}\\
+  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_4_ale_auto.pdf}
+  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_4_pdp_ice.pdf}\\
+  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_5_ale_auto.pdf}
+  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_5_pdp_ice.pdf}\\
   \caption{Figure 1}
-  \label{fig:ex-synth-2-1}
+  \label{fig:ex-real-1}
 \end{figure}
@@
 
-For evaluating UALE in terms of bin-splitting we set-up the following
-evaluation framework. We treat as ground-truth the effects computed on
-the full training-set \(N=20000\) with dense fixed-size bin-splitting
-(\(K=80\)). Given the big number of samples, we make the hypothesis
-that the approximation with dense binning is close to the ground
-truth. Afterwards we select less samples \(N=1000\) and we compare
-UALE approximation with all possible fixed-size alternatives, in terms
-of accuracy, i.e. we provide the quantities
-\(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma^2}, \mathcal{L}^{\rho^2}\).
+As in the Syntetic Example (Case 2), we treat as ground-truth the
+effects computed on the full training-set \(N=20000\) with dense
+fixed-size bin-splitting (\(K=80\)). Given the big number of samples,
+we make the hypothesis that the approximation with dense binning is
+close to the ground truth. Afterwards, we randomly select \(N=1000\)
+samples and we compare UALE approximation with all possible fixed-size
+alternatives using \(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\). Add discussion
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_1.pdf}
+  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_1.pdf}\\
+  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_4.pdf}
+  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_4.pdf}\\
+  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_5.pdf}
+  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_5.pdf}\\
+  \caption{Figure 1}
+  \label{fig:ex-real-2}
+\end{figure}
+
+\paragraph{Discussion}
+
 \section{CONCLUSION}

### 2022-10-06 AISTATS2023/paper.tex
added analysis to the real example

@@
 \(20000\) training instances, making it appropriate for robust
-Monte-Carlo approximations. The dataset contains \(D=8\) features
-describing building blocks of California, like latitude, longitude,
-total number of people residing in the block etc., and the target is
-the median house value inside the block. We exclude instances with
-missing or outlier values, i.e. \(x_d^i > \hat{\sigma}_j\) in any
-feature and normalize them to have
-\(\hat{\mu}_d = 0, \hat{\sigma_d}\). We split the dataset into
-\(N_{tr} = 15639\) training and \(N_{test} = 3910\) test examples (80/20
-split) and we fit a Neural Network with 3 hidden layers of 256, 128
-and 36 units respectively. After 15 epochs using the Adam optimizer
-with learning rate \(\eta = 0.02\), the model achieves a normalized
-mean squared error (R-Squared) of \(0.25\) (0.75). In depth
-information about preprocessing, training and evaluation are provided
-in the Appendix.
+Monte-Carlo approximations. The dataset contains \(D=8\) numerical
+features with characteristics about the building blocks of California,
+e.g. latitude, longitude, population of the block or median age of
+houses in the block. The target variable is the median value of the
+houses inside the block in dollars that ranges between
+\([15, 500] \cdots 10^3\), with a mean value of
+\(\mu_Y \approx 201 \cdot 10^3 \) and a standard deviation of
+\(\sigma_Y \approx 110 \cdot 10^3\).
+
+We exclude instances with missing values or with outlier values. As
+outlier we define the feature values which ar over three standard
+deviations away from the mean feature value. We also normalize all
+features to zero-mean and unit standard deviation. We split the
+dataset into \(N_{tr} = 15639\) training and \(N_{test} = 3910\) test
+examples (80/20 split) and we fit a Neural Network with 3 hidden
+layers of 256, 128 and 36 units respectively. After 15 epochs using
+the Adam optimizer with learning rate \(\eta = 0.02\), the model
+achieves a normalized mean squared error (R-Squared) of \(0.25\)
+(0.75), which corresponds to a MAE of \(37 \cdot 10^3\) dollars.
+
+Below, we illustrate the feature effect for three features: latitude
+\(x_2\), population \(x_6\) and median income \(x_8\). The particular
+features cover the main FE cases, e.g.~positive/negative trend and
+linear/non-linear curve, and they are appropriate for illustration
+purposes. The complete results for all features, along with in-depth
+information about the preprocessing, training and evaluation parts are
+provided in the Appendix.
 
@@
 In real-world datasets, it is infeasible to obtain the ground truth FE
-for seamlessly evaluating the competitive methods. For this reason, we
-selected the California Dataset, where, in broad-terms, UALE and
-PDP-ICE aggree in the quantification of the average effect and of the
-uncertainty. In Figure\ref{fig:ex-real-1}, we observe that UALE
-complements ALE for the quantification of the heterogeneous effects,
-similarly to as ICE complements PDP. Add some content..
+for seamlessly evaluating the competitive methods. We selected the
+particular experiment, because, in broad terms, UALE and PDP-ICE plots
+aggree in the estimation of the average effect and
+uncertainty. Therefore, we can focus on juding the quality of the
+information provided by the two methods.
+
+In Figure\ref{fig:ex-real-1}, we observe, from top to bottom, the
+effects for the latitude, population and the median income. The effect
+of UALE and PDP-ICE are similar for the population and the median
+income. The population has a negative impact that progressively
+decreases: from 400 to 1500 people the house value decreases with a
+rate of \(-150 (\pm 140)\) dollars per added person, a rate that
+decreases from \(-80 (\pm 80)\) to \(-60 (\pm 60)\) dollars per added
+person as we move from 1500 to 2800 people. The level of uncertainty
+indicates significant variance in absolute value of the rate, but in
+the grant majority of instances the rate is negative. With the same
+inspection, we observe that the median annual income has a positive
+impact on the value (all numbers are thousands of dollars):
+\(20\pm 15\) per \(10\) of added median income for incomes in
+\([8, 15]\), \(32 \pm 20\) per \(10\) added income in \([15, 60]\) and
+\(40 \pm 15\) per \(10\) added income in \([60, 70]\). The uncertainty
+indicates that there are less heterogeneous effects about the median
+income compared to the number of people. In both cases, we can end-up
+to the same conclusion by inspecting the PDP-ICE plots. For the
+latitude, there is a small difference in the explanations for the
+region \([32, 35]\), where UALE estimates a less negative slope with
+less uncertainty than PDP, while the explanations are similar for the
+range \([35,39]\), where both methods reveal an increase in the
+uncertainty around the feature value \(37.5\).
+
+In general, we observe that UALE complements ALE in the quantification
+of the heterogeneous effects, similarly to as ICE complements PDP. The
+automatic extraction of constant-effect and constant-uncertainty
+regions provided by UALE is helpful for an easier interpretation. On
+the other hand, ICE plots sometimes are more descritive on locating
+the type of heterogeneity, whereas UALE quantifies only the level (not
+the type) of heterogeneity.
 
@@
   \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_pdp_ice.pdf}\\
-  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_4_ale_auto.pdf}
-  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_4_pdp_ice.pdf}\\
   \includegraphics[width=.23\textwidth]{real_dataset_3/feature_5_ale_auto.pdf}
   \includegraphics[width=.23\textwidth]{real_dataset_3/feature_5_pdp_ice.pdf}\\
+  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_ale_auto.pdf}
+  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_pdp_ice.pdf}\\
   \caption{Figure 1}
@@
 
-As in the Syntetic Example (Case 2), we treat as ground-truth the
+In this part, we evaluate the robustness of the approximation using
+automatic bin-splitting. Following the evaluation framework of
+Section~\ref{sec:simulation-examples-2}, we treat as ground-truth the
 effects computed on the full training-set \(N=20000\) with dense
@@
 we make the hypothesis that the approximation with dense binning is
-close to the ground truth. Afterwards, we randomly select \(N=1000\)
-samples and we compare UALE approximation with all possible fixed-size
-alternatives using \(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\). Add discussion
+close to the ground truth. Afterwards, we randomly select less samples
+\(N=1000\) and we compare UALE approximation with all possible
+fixed-size alternatives, repeating this process for 30 times for
+robust results. In Figure~\ref{fig:ex-real-2}, we illustrate the mean
+values for \(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\) of the 30
+repetitions. We observe that automatic bin-spliting provides (close to
+the) best approximation in the three features. In the Appendix, we
+provide the same evaluation for all features.
 
@@
   \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_1.pdf}\\
-  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_4.pdf}
-  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_4.pdf}\\
   \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_5.pdf}
   \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_5.pdf}\\
+  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_7.pdf}
+  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_7.pdf}\\
   \caption{Figure 1}
@@
 
-\paragraph{Discussion}
+\section{DISCUSSION}
 
-\section{CONCLUSION}
+add limitations.
 

### 2022-10-06 examples/real_dataset_3.py

@@ -147,13 +147,16 @@ stats_auto_list = []
 
 # fit feature
 for s in range(8):
+    scale_x = {"mean": X_mu[s], "std": X_sigma[s]}
+    scale_y = {"mean": Y_mu, "std": Y_sigma}
+
     print("Feature " + str(s))
     dale = fe.DALE(X_tr, model, model_grad, axis_limits)
     alg_params = {"bin_method" : "dp",
                   "max_nof_bins" : 20,
                   "min_points_per_bin": 30}
     dale.fit(features=s, alg_params=alg_params)
-    dale.plot(s=s)
+    # dale.plot(s=s)
 
 
     dale_gt = fe.DALE(X_tr, model, model_grad, axis_limits)
@@ -161,7 +164,7 @@ for s in range(8):
                   "nof_bins" : 80,
                   "min_points_per_bin": 30}
     dale_gt.fit(features=s, alg_params=alg_params)
-    dale_gt.plot(s)
+    # dale_gt.plot(s)
 
 
     # get statistics
@@ -195,6 +198,7 @@ for s in range(8):
                              stats_auto["mu_err_mean"],
                              stats_auto["mu_err_std"],
                              "mu",
+                             scale_y,
                              savefig=savepath)
 
     savepath = os.path.join(path2dir, "compare_var_err_feature_" + str(s) + ".pdf") if savefig else None
@@ -204,6 +208,7 @@ for s in range(8):
                              stats_auto["var_err_mean"],
                              stats_auto["var_err_std"],
                              "var",
+                             scale_y,
                              savefig=savepath)
 
     savepath = os.path.join(path2dir, "compare_rho_feature_" + str(s) + ".pdf") if savefig else None
@@ -213,14 +218,21 @@ for s in range(8):
                              stats_auto["rho_mean"],
                              stats_auto["rho_std"],
                              "rho",
+                             scale_y,
                              savefig=savepath)
 
 
+path2dir = os.path.join(path, folder_name)
 for s in range(8):
+    scale_x = {"mean": X_mu[s], "std": X_sigma[s]}
+    scale_y = {"mean": Y_mu, "std": Y_sigma}
+
+    print("Feature " + str(s))
     ind = np.random.choice(X_tr.shape[0], size=200, replace=False)
[…]

### 2022-10-06 examples/utils.py

@@ -1,6 +1,7 @@
 import sys
 import os
 import feature_effect as fe
+from feature_effect import visualization as vis
 import numpy as np
 import matplotlib.pyplot as plt
 
@@ -240,7 +241,26 @@ def measure_auto_error_real_dataset(dale_gt,
     return stats
 
 
-def plot_fixed_vs_auto(K_list, fixed_mean, fixed_std, auto_mean, auto_std, metric, savefig=None):
+def plot_fixed_vs_auto(K_list, fixed_mean, fixed_std, auto_mean, auto_std, metric,
+                       scale_x = None,
+                       scale_y = None,
+                       savefig=None):
+
+    def scale(x, scale_x, scale_y):
+        return x * scale_y / scale_x
+
+    fixed_mean = np.array(fixed_mean)
+    fixed_mean = fixed_mean if scale_y is None else scale(fixed_mean, scale_x, scale_y)
+
+    fixed_std = np.array(fixed_std)
+    fixed_std = fixed_std if scale_y is None else scale(fixed_std, scale_x, scale_y)
+
+    auto_mean = np.array(auto_mean)
+    auto_mean = auto_mean if scale_y is None else scale(auto_mean, scale_x, scale_y)
+
+    auto_std = np.array(auto_std)
+    auto_std = auto_std if scale_y is None else scale(auto_std, scale_x, scale_y)
+
     plt.figure()
     assert metric in ["rho", "mu", "var"]
     if metric == "rho":

### 2022-10-06 feature_effect/dale.py

@@ -182,7 +182,14 @@ class DALE(FeatureEffectBase):
     #     vis.plot_1D(x, y, title="ALE (Monte Carlo) for feature %d" % (s+1))
 
 
-    def plot(self, s: int = 0, error="std", block=False, gt=None, gt_bins=None, savefig=False):
+    def plot(self, s: int = 0,
+             error="std",
+             scale_x=None,
+             scale_y=None,
+             block=False,
+             gt=None,
+             gt_bins=None,
+             savefig=False):
         vis.feature_effect_plot(self.feature_effect["feature_"+str(s)],
                                 self.eval,
                                 s,
@@ -192,6 +199,8 @@ class DALE(FeatureEffectBase):
                                 block=block,
                                 gt=gt,
                                 gt_bins=gt_bins,
+                                scale_x=scale_x,
+                                scale_y=scale_y,
                                 savefig=savefig)
 
     def plot_local_effects(self, s: int = 0, limits=True, block=False):

### 2022-10-06 feature_effect/pdp.py

@@ -346,7 +346,10 @@ class PDPwithICE:
             y_ice.append(y)
         self.y_ice = np.array(y_ice)
 
-    def plot(self, s: int, normalized: bool = True,
+    def plot(self, s: int,
+             scale_x=None,
+             scale_y=None,
+             normalized: bool = True,
              nof_points: int = 30,
              savefig = None) -> None:
         """Plot the s-th feature
@@ -354,4 +357,4 @@ class PDPwithICE:
         self.fit(s, normalized, nof_points)
 
         axis_limits = self.axis_limits
-        vis.plot_PDP_ICE(s, self.x, self.y_pdp, self.y_ice, savefig)
+        vis.plot_PDP_ICE(s, self.x, self.y_pdp, self.y_ice, scale_x, scale_y, savefig)

### 2022-10-06 feature_effect/visualization.py

@@ -3,12 +3,56 @@ import numpy as np
 from feature_effect.utils import compute_remaining_effect
 
 
-def feature_effect_plot(params, eval, feature, error, min_points_per_bin, title=None, block=False, gt=False, gt_bins=None, savefig=False):
+def affine(x, transform):
+     return x*transform["std"] + transform["mean"]
+
+def scale(x, transform, square=False):
+    if square:
+        y = x*transform["std"]**2
+    else:
+        y = x*transform["std"]
+    return y
+
+def scale_bin(x, scale_x, scale_y, square=False):
+    if square:
+        y = x*(scale_y["std"]/scale_x["std"])**2
+    else:
+        y = x*scale_y["std"]/scale_x["std"]
+    return y
+
+
+def feature_effect_plot(params,
+                        eval,
+                        feature,
+                        error,
+                        min_points_per_bin,
+                        title=None,
+                        block=False,
+                        gt=False,
+                        gt_bins=None,
+                        scale_x = None,
+                        scale_y = None,
+                        savefig=False):
     assert all(name in params for name in ["first_empty_bin", "limits", "dx", "is_bin_empty", "bin_estimator_variance", "bin_effect"])
-    limits = params["limits"]
-    dx = params["dx"]
-    bin_variance = params["bin_variance"]
-    bin_effect = params["bin_effect"]
+
+    x = np.linspace(params["limits"][0], params["limits"][-1], 1000)
+    y, std, estimator_var = eval(x, feature, True)
+    rem_eff = compute_remaining_effect(x,
+                                       params["limits"],
+                                       np.sqrt(params["bin_variance"]),
+                                       square=False)
+
+    x = x if scale_x is None else affine(x, scale_x)
+    y = y if scale_y is None else affine(y, scale_y)
+    std = std if scale_y is None else scale(std, scale_y)
+    estimator_var = estimator_var if scale_y is None else scale(estimator_var, scale_y, True)
+
+
+    limits = params["limits"] if scale_x is None else affine(params["limits"], scale_x)
+    dx = params["dx"] if scale_x is None else scale(params["dx"], scale_x)
+
+    bin_variance = params["bin_variance"] if scale_y is None else scale_bin(params["bin_variance"], scale_x, scale_y, True)
[…]

### 2022-10-08 AISTATS2023/paper.tex

@@
   Accumulated Local Effects (ALE) is a popular approach to explainable
-  AI that describes how a feature influences the decisions of a model
+  AI (XAI) that describes how a feature influences the decisions of a model
   on average, taking into account potential correlations between the
@@
 them. For this reason there is an increased interest in Explainable AI
-(XAI), the field that provides interpretations about the behavior of
+(XAI), to understand the decision mechanism of 
+%the field that provides interpretations about the behavior of
 complex black-box models. XAI literature distinguishes between local
@@
 heterogeneous effects that exist under the
-hood~\citep{herbinger2022repid}; a phenomenon called aggregation
-bias~\citep{mehrabi2021survey}.
+hood~\citep{herbinger2022repid}; a phenomenon called \emph{aggregation
+bias}~\citep{mehrabi2021survey}.
 
 Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category of global
-explainability methods. The objective of FE is to isolate the impact
-of a single feature on the output.~\footnote{Often, FE methods also
+explainability methods aiming at isolating the impact
+of a single feature on the output\footnote{Often, FE methods also
   isolate the combined effect of a pair of features to the
-  output. Combinations of more than two features are uncommon, since they are difficult to estimate and visualize.}
-FE methods suffer from aggregation bias because, often, implications of the average effect to the model behavior are unclear. For example, a feature with zero
-average effect may indicate no effect on the output or, in contrast,
-highly positive in some cases and highly negative in others.
-
+  output. Combinations of more than two features are uncommon, since they are difficult to estimate and visualize.}.
+FE methods suffer from aggregation bias because, often, implications of the average effect on the model behavior are unclear.
+For example, a feature with zero average effect may indicate no effect on the output or,
+a highly positive effect in some cases and a highly negative effect in others. 
 There are
-three widely-used FE methods; Partial Dependence Plots
-(PDP)\citep{friedman2001greedy}, Marignal Plots
-(MP)\citep{apley2020visualizing} and Aggregated Local Effects
-(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
+three widely-used FE methods: \emph{Partial Dependence Plots}
+(PDP)\citep{friedman2001greedy}, \emph{Marignal Plots}
+(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
+(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized \todo{Add a reference}for
 computing erroneous effects when the input features are (highly)
-correlated, which is a frequent scenario in many ML cases. Therefore,
+correlated, which is typical in many ML cases. Therefore,
 ALE has been established as the state-of-the-art FE method. However, ALE faces two crucial limitations, the first concerns ALE
 definition and the second one ALE approximation. Regarding the
-definition ALE does not inform the user about the uncertainty, i.e.,
+definition, ALE does not inform the user about the uncertainty, i.e.,
 the level of heterogeneous effects hidden behind the average effect
@@
 Individual Conditional Expectations
-(ICE)\citep{goldstein2015peeking}. Regarding the approximation,
+(ICE)~\citep{goldstein2015peeking}. Regarding the approximation,
 i.e.~the estimation using the (limited) samples of the training set,
-ALE requires the additional \textit{bin-splitting} step. Bin-splitting
-consists of partitioning the domain of the feature of interest into consecutive non-overlapping intervals (regions) and estimating a
+ALE requires an additional \textit{bin-splitting} step to partition the domain of the feature of interest into consecutive non-overlapping intervals (regions) and to estimate a
 constant effect in each region. The effect is estimated from the population of samples that fall
@@
 importance, since ALE's interpretation is meaningful only inside each
-region \citep{molnar2022}. However, this crucial step has not raised
+region~\citep{molnar2022}. However, this crucial step has not raised
 the appropriate attention and the approximation is vulnerable to
-potential misinterpretations.
+potential misinterpretations.\todo{I miss what exactly is the simplication of existing mehtods here. Should we add e.g., typically equisized bins?}
 

### 2022-10-08 AISTATS2023/paper.tex
temporary vasilis commit

--- 2022-10-06 AISTATS2023/paper.tex
@@
 \aistatstitle{Uncertainty-aware Accumulated Local Effects (UALE) for
-  quantifying the heterogeneous effects}
+  quantifying the hetegeneity of the local effects}
 
@@
 \begin{abstract}
-  Accumulated Local Effects (ALE) is a popular approach to explainable
-  AI that describes how a feature influences the decisions of a model
-  on average, taking into account potential correlations between the
-  features.  However, ALE does not provide an explicit component for
-  modeling the uncertainty of the effect, i.e., the level of
-  heterogeneity of the effects behind the average explanation, which
-  is crucial for a complete interpretation when the black-box function
-  contains interactions between the correlated features. In this work,
-  we define Uncertainty-aware ALE (UALE), an extension of ALE for
-  quantifying the heterogeneous effects and we propose a method for
-  estimating it. We formally prove the conditions the bin-splitting,
-  i.e.~the partitioning the domain of the feature of interest into
-  consecutive non-overlapping intervals, must respect for an an
-  unbiased estimation of the uncertainty. We finally propose a
-  computationally-grounded algorithm for finding the optimal sequence
-  of bins given the limited instances of the training set. We
-  demonstrate through synthetic and real datasets, that UALE
-  quantifies the average and heterogeneous effects correctly and
-  approximates them robustly by optimizing the bin-splitting
-  procedure.
+Accumulated Local Effects (ALE) is a popular approach to explainable AI that quantifies how a feature influences the decisions of a model taking into account feature correlations.
+However, in case of strong interactions between features, instance-level feature effects may deviate from the average curve provided by ALE.
+Therefore, it is crucial to quantify this deviation, namely, the uncertainty of the effect.
+In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and the uncertainty.
+Furthermore, as in ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting).
+The average effect and the uncertainty are then computed from the instances that lie in each bin.
+In this work, we formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints.
+Based on this, we propose a method for automatically finding the optimal intervals, balancing the trade-off between estimation bias and variance.
+We demonstrate through synthetic and real datasets (a) the advantages of modelling the uncertainty with UALE compared to other methods and (b) the importance of appropriate bin splitting for an accurate approximation of the mean effect and uncertainty.
 \end{abstract}

### 2022-10-08 AISTATS2023/paper.tex

--- 2022-10-08 AISTATS2023/paper.tex
@@
 \begin{abstract}
-  Accumulated Local Effects (ALE) is a popular approach to explainable
-  AI (XAI) that describes how a feature influences the decisions of a model
-  on average, taking into account potential correlations between the
-  features.  However, ALE does not provide an explicit component for
-  modeling the uncertainty of the effect, i.e., the level of
-  heterogeneity of the effects behind the average explanation, which
-  is crucial for a complete interpretation when the black-box function
-  contains interactions between the correlated features. In this work,
-  we define Uncertainty-aware ALE (UALE), an extension of ALE for
-  quantifying the heterogeneous effects and we propose a method for
-  estimating it. We formally prove the conditions the bin-splitting,
-  i.e.~the partitioning the domain of the feature of interest into
-  consecutive non-overlapping intervals, must respect for an an
-  unbiased estimation of the uncertainty. We finally propose a
-  computationally-grounded algorithm for finding the optimal sequence
-  of bins given the limited instances of the training set. We
-  demonstrate through synthetic and real datasets, that UALE
-  quantifies the average and heterogeneous effects correctly and
-  approximates them robustly by optimizing the bin-splitting
-  procedure.
+
+Accumulated Local Effects (ALE) is a popular approach to explainable AI that quantifies how a feature influences the decisions of a model taking into account feature correlations.
+However, in case of strong interactions between features, instance-level feature effects may deviate from the average curve provided by ALE.
+Therefore, it is crucial to quantify this deviation, namely, the uncertainty of the effect.
+In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and the uncertainty.
+Furthermore, as in ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting).
+The average effect and the uncertainty are then computed from the instances that lie in each bin.
+In this work, we formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints.
+Based on this, we propose a method for automatically finding the optimal intervals, balancing the trade-off between estimation bias and variance.
+We demonstrate through synthetic and real datasets (a) the advantages of modelling the uncertainty with UALE compared to other methods and (b) the importance of appropriate bin splitting for an accurate approximation of the mean effect and uncertainty.
 \end{abstract}

### 2022-10-08 AISTATS2023/paper.tex
updated title
Intro ready

@@
 \newcommand{\Xb}{\mathcal{X}}
+\newcommand{\todo}[1]{[\textcolor{red}{#1}]}
 
@@
 
-\aistatstitle{Uncertainty-aware Accumulated Local Effects (UALE) for
-  quantifying the hetegeneity of the local effects}
+\aistatstitle{Uncertainty-aware Accumulated Local Effects (UALE) for quantifying the heterogeneity of instance-level feature effects}
 
@@
 In this work, we formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints.
-Based on this, we propose a method for automatically finding the optimal intervals, balancing the trade-off between estimation bias and variance.
-We demonstrate through synthetic and real datasets (a) the advantages of modelling the uncertainty with UALE compared to other methods and (b) the importance of appropriate bin splitting for an accurate approximation of the mean effect and uncertainty.
+Based on this, we propose a method for automatically finding the optimal intervals, balancing the trade-off between estimation bias and variance~\todo{check if this phrase explains well the concept.} We demonstrate through synthetic and real datasets (a) the advantages of modeling the uncertainty with UALE compared to other methods and (b) the importance of appropriate bin splitting for an accurate approximation of the average effect and uncertainty.
 \end{abstract}
@@
 them. For this reason there is an increased interest in Explainable AI
-(XAI), to understand the decision mechanism of 
-%the field that provides interpretations about the behavior of
-complex black-box models. XAI literature distinguishes between local
-and global explainability
-techniques~\citep{Molnar2020interpretable}. Local methods explain the prediction of a
-specific instance, whereas global methods provide an explanation that summarizes the entire model behavior.
-Global methods provide a universal explanation, aggregating the various local explanations into a single interpretable outcome,
-usually a number or a plot. If users want to get a rough overview
-about which features are significant (feature importance) or whether a
-particular feature has a positive or negative effect on the output
-(feature effect), they should opt for a global explainability
-technique. On the other hand, aggregating the individual explanations
-is vulnerable to misinterpretations. Under strong interactions and
-correlations between features, the global explanation may obfuscate
-heterogeneous effects that exist under the
-hood~\citep{herbinger2022repid}; a phenomenon called \emph{aggregation
-bias}~\citep{mehrabi2021survey}.
+(XAI), to understand the decision mechanism of complex black-box
+models. XAI literature distinguishes between local and global
+techniques~\citep{Molnar2020interpretable}. Local methods provide
+instance-level explanations, i.e.~explain the prediction for a
+specific instance, whereas global methods summarize the entire model
+behavior.
 
-Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category of global
-explainability methods aiming at isolating the impact
-of a single feature on the output\footnote{Often, FE methods also
-  isolate the combined effect of a pair of features to the
-  output. Combinations of more than two features are uncommon, since they are difficult to estimate and visualize.}.
-FE methods suffer from aggregation bias because, often, implications of the average effect on the model behavior are unclear.
-For example, a feature with zero average effect may indicate no effect on the output or,
-a highly positive effect in some cases and a highly negative effect in others. 
-There are
-three widely-used FE methods: \emph{Partial Dependence Plots}
-(PDP)\citep{friedman2001greedy}, \emph{Marignal Plots}
+Global methods create a global explanation by aggregating the
+instance-level explanations into a single interpretable outcome,
+usually a number or a plot. When strong interactions between features
+exist, the instance-level explanations may significantly deviate from
+the aggregated outcome~\citep{herbinger2022repid}, a phenomenon called
+\emph{aggregation bias}~\citep{mehrabi2021survey}. Therefore, global
+methods should quantify not only the aggregated (averaged) effect, but
+also the level of heterogeneity of instance-level explanations. The
+same holds for Feature Effect (FE)~\citep{Gromping2020MAEP}, a class
+of global XAI methods that isolate the impact of a single feature on
+the output.\footnote{FE methods also isolate the combined effect of a
+  pair of features to the output. Combinations of more than two
+  features are uncommon, since they are difficult to estimate and
+  visualize.} For example, a feature with zero average effect may
+indicate no effect on the output or a highly positive effect in some
+cases and a highly negative effect in others. The ambiguity about the
+correct interpretation could be resolved, if the FE method informs
+about the heterogeneity of instance-level explanations, a quantity we
+refer to as the \emph{uncertainty of the effect}.
+
+There are three widely-used FE methods: \emph{Partial Dependence
+  Plots} (PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
 (MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
-(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized \todo{Add a reference}for
-computing erroneous effects when the input features are (highly)
-correlated, which is typical in many ML cases. Therefore,
-ALE has been established as the state-of-the-art FE method. However, ALE faces two crucial limitations, the first concerns ALE
-definition and the second one ALE approximation. Regarding the
-definition, ALE does not inform the user about the uncertainty, i.e.,
-the level of heterogeneous effects hidden behind the average effect
-due to implicit feature interactions. In contrast, in PDPs the
-heterogeneous effects can be identified by exploring the
-Individual Conditional Expectations
-(ICE)~\citep{goldstein2015peeking}. Regarding the approximation,
-i.e.~the estimation using the (limited) samples of the training set,
-ALE requires an additional \textit{bin-splitting} step to partition the domain of the feature of interest into consecutive non-overlapping intervals (regions) and to estimate a
-constant effect in each region. The effect is estimated from the population of samples that fall
-inside the region. Specifying an appropriate set of bins is of particular
-importance, since ALE's interpretation is meaningful only inside each
-region~\citep{molnar2022}. However, this crucial step has not raised
-the appropriate attention and the approximation is vulnerable to
-potential misinterpretations.\todo{I miss what exactly is the simplication of existing mehtods here. Should we add e.g., typically equisized bins?}
+(ALE)\citep{apley2020visualizing}. ALE has been established as the
+state-of-the-art for measuring the average effect, since PDP and MP
+have been criticized~\citep{Gromping2020MAEP} to compute erroneous
+effects when the input features are (highly) correlated. However, ALE
+faces two limitations, the first concerns ALE definition and the
+second ALE approximation. Regarding the definition, ALE does not
+inform the user about the uncertainty. In contrast, in PDP the
+heterogeneity of instance-level effects can be identified by exploring
+the Individual Conditional Expectations
+(ICE)~\citep{goldstein2015peeking}. Regarding the approximation, ALE
+requires partitioning the feature domain into non-overlapping
+intervals (bin-splitting) and estimate the effect inside each interval
+(bin-effect) from the training set instances that fall inside. Finding
+an appropriate sequence of bins is of particular importance, since
+ALE's interpretation is meaningful only inside each
+bin~\citep{molnar2022}. However, this crucial step has not raised the
+appropriate attention. With the current
+approaches~\citep{apley2020visualizing, gkolemis22}, the domain is
+split \emph{blindly}, i.e.~without considering the characteristics of
+the instance-level effects, in \(K\) equisized bins.
 
 In this paper, we present Uncertainty-aware Accumulated Local Effects
-(UALE) an extension of ALE for measuring the uncertainty of the
-explanation, i.e.~how certain we are that the average explanation
-would predict the feature effect to an instance drawn at random. We
-show that choosing an appropriate sequence of intervals is
-particularly important for a robust and unbiased estimation of the
-uncertainty and we provide an algorithm that automates this step. We
-also formally prove the bin-splitting conditions for an unbiased
-estimation of the uncertainty. We apply our method in multiple
-synthetic and real world datasets for evaluating its performance.
-
-\paragraph{Contributions.} The contributions of this paper are:
+(UALE), a method that visualizes in a single plot both the average
+feature effect and the uncertainty, i.e.~the level of heterogeneity of
+instance-level effects. The average effect is defined as in ALE and
+the uncertainty inherits ALE's advantages, i.e.~takes into account the
+correlations between features. We also propose an accurate UALE
+approximation. We formally prove that for an unbiased estimation of
+the uncertainty, the partitioning of the feature domain must respect
+specific conditions. Using that, we set-up an optimization problem
+where we search for a partitioning, a sequence of \emph{variable-size}
+intervals, (a) with enough instances (accurate estimation) and (b)
+that respect the above conditions (unbiased estimation). Finally, we
+propose a method for automatically finding the optimal
+partitioning. We apply our method in multiple synthetic and real world
+datasets for evaluating its performance. The contributions of this
+paper are:
 
 \begin{itemize}
-\item UALE, a FE method that extends ALE for quantifying the
-  heterogeneous effects (uncertainty).
-\item A formal proof of the bin-splitting requirements for an unbiased
-  approximation of the heterogeneous effects.
-\item A framework for automatically extracting regions with similar
-  effects, improving the estimation and the interpretability of ALE
-  plots.
+\item UALE, a feature effect method that quantifies both the average
+  effect and the uncertainty, i.e.,~the heterogeneity of instance level
+  effects.
+\item A formal proof of the conditions the feature domain partitioning
+  must fulfill for an unbiased approximation of the uncertainty.
+\item An accurate UALE approximation by automatically partitioning
+  into variable-size intervals that respect the above conditions.
 \end{itemize}
 
-We provide empirical evaluation of the method in artificial and real
-datasets. The implementation of our method and the code for
+The implementation of our method and the code for
 reproducing all the experiments is provided in the submission and will

### 2022-10-09 AISTATS2023/paper.tex
update

@@
 of accurate predictions along with meaningful explanations to support
-them. For this reason there is an increased interest in Explainable AI
-(XAI), to understand the decision mechanism of complex black-box
+them. For this reason, there is an increased interest in Explainable
+AI (XAI) to understand the decision mechanism of complex black-box
 models. XAI literature distinguishes between local and global
@@
 cases and a highly negative effect in others. The ambiguity about the
-correct interpretation could be resolved, if the FE method informs
-about the heterogeneity of instance-level explanations, a quantity we
-refer to as the \emph{uncertainty of the effect}.
+correct interpretation can be resolved, if the FE method informs about
+the heterogeneity of instance-level explanations, a quantity we refer
+to as the \emph{uncertainty of the effect}.
 
@@
 
-In this section, we describe the basic methods for FE and uncertainty
-quantification, focusing on ALE definition. We, then, review the ALE
-approximation\citep{apley2020visualizing, gkolemis22}, describing some
-of its vulnerabilities.
+We here review the existing methods for modeling the average effect
+and the heterogeneity of instance-level effects and we focus on ALE
+definition~\citep{apley2020visualizing} and
+approximation~\citep{apley2020visualizing, gkolemis22}.
 
@@
 \(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
-\((x_s, \mathbf{x}_c)\), but we clarify that it implies the vector
+\((x_s, \mathbf{x}_c)\), but we it refers the vector
 \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
@@
 refer to an instance of the training set, we use
-\(\xb^i= (\xc^i, x_s^i) \). The black-box function is
-\(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the FE of the \(s\)-th
-feature is \(f^{\mathtt{<method>}}(x_s)\), with \(\mathtt{<method>}\)
-indicating the particular method in use.\footnote{An extensive list of all
-  symbols used in the paper is provided at the Appendix.}
+\(\xb^i= (x_s^i, \xc^i) \).
 
-\subsection{Feature Effect Methods And ALE Definition}
+% The black-box function is
+% \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the FE of the \(s\)-th
+% feature is \(f^{\mathtt{<method>}}(x_s)\), with \(\mathtt{<method>}\)
+% indicating the particular method in use.\footnote{An extensive list of
+%   all symbols used in the paper is provided at the Appendix.}
+
+\subsection{Feature Effect Methods}
 \label{sec:feat-effect-meth}
@@
 The three well-known feature effect methods are: PDP, MP and ALE. PDP
-formulates the FE of the \(s\)-th attribute as an expectation over the
+formulates the FE of the \(s\)-th feature as an expectation over the
 marginal distribution \(\mathbf{X}_c\), i.e.,
@@
 \(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|x_s}[f(x_s,
-\mathbf{X}_c)]\). ALE defines the local effect of the \(s\)-th feature
-at point \(x_s = z\) as
-\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
-local explanations at \(z\) are, then, weighted by the conditional
-distribution \(p(\xc|x_s=z)\) and are averaged, to produce the
-averaged effect \(\mu(z)\). ALE is the accumulation of the averaged
-local effects:
+\mathbf{X}_c)]\). ALE defines the local effect at \(x_s = z\) as
+\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). The local
+explanations are then weighted by the conditional distribution
+\(p(\xc|x_s=z)\) to compute the averaged effect \(\mu(z)\). ALE is the
+accumulation of the averaged local effects:
 
@@
 
-Eq.~(\ref{eq:ALE}) has specific advantages which gain particular value in cases of
-correlated input features. In these cases, PDP integrates over
+Eq.~(\ref{eq:ALE}) has important advantages compared to PDP and MP in
+cases of correlated features. In these cases, PDP integrates over
 unrealistic instances due to the use of the marginal distribution
@@
 the combined effect of sets of features to a single feature. ALE
-manages to resolve both issues, and is therefore the most trustable
-method in cases of correlated features.
+resolves both issues.
 
-\subsection{Quantification Of Heterogeneous Effects.}
+\subsection{Heterogeneity Of Instance-Level Effects.}
 \label{sec:quant-heter-effects}
 
-FE methods reply to the question \textit{what is the average (global)
-  effect on the output, if the value of a specific feature is
-  increased/decreased}. It comes naturally to also ask \textit{to what
-  extent the local effects deviate from the global
-  explanation}. Quantifying the uncertainty of the global explanation
+FE methods show what is the average (global) effect on the output, if
+the value of a specific feature is increased/decreased. For completing
+the explanation about a feature, we need to also know to what extent
+the local effects deviate from the global explanation, i.e.~to
+quantify the uncertainty of the average effect.
+
+Heterogeneity It comes naturally to also ask \textit{to
+  what extent }. Quantifying the uncertainty of the global explanation
 has attracted a lot of interest. ICE and

### 2022-10-09 AISTATS2023/paper.tex
small modifications

@@
 
-\subsection{Heterogeneity Of Instance-Level Effects.}
+\subsection{Heterogeneity Of Instance-Level Effects}
 \label{sec:quant-heter-effects}
@@
 
-Heterogeneity It comes naturally to also ask \textit{to
-  what extent }. Quantifying the uncertainty of the global explanation
-has attracted a lot of interest. ICE and
-d-ICE\citep{goldstein2015peeking} provide a set of curves that are
-plot on top-of the PDP. Both methods produce one curve for each
+ICE and d-ICE\citep{goldstein2015peeking} provide a set of curves that
+are plot on top-of PDP. Both methods produce one curve for each
 instance of the training set;
@@
 (d-)ICE plots into clusters~\citep{molnar2020model,
-  herbinger2022repid, britton2019vine}. Some other approaches, like
+  herbinger2022repid, britton2019vine}. Other approaches, like
 H-Statistic\citep{friedman2008predictive}, Greenwell's interaction
 index\citep{greenwell2018simple} or SHAP interaction
-values\citep{lundberg2018consistent}, quantify the level of
-interaction between the input features, with an interaction value. A
-strong interaction index is an indicator for the existence of
-heterogeneous effects.
+values\citep{lundberg2018consistent} inform indirectly about
+heterogeneous effects, quantifying interactions between the input
+features. A strong interaction index is an indicator for the existence
+of heterogeneous effects.
 
-The aforementioned approaches are under two limitations; They either
+The aforementioned approaches are under two limitations. They either
 do not quantify the uncertainty of the FE directly or they are based
@@
 
-\subsection{ALE Approximation.}
+\subsection{ALE Approximation}
 \label{sec:ale-approximation}
 
-In ML scenarios, ALE is estimated from the training set
-examples. Therefore, \citep{apley2020visualizing} proposed dividing
-the feature domain in \(K\) equal-sized bins and estimating the local effects in
-each bin by evaluating the black box-function at the bin limits:
+In ML scenarios, ALE is estimated from the instances of the training
+set.~\citep{apley2020visualizing} proposed dividing the feature domain
+in \(K\) equisized bins and estimating the local effects in each bin
+by evaluating the black box-function at the bin limits:
 \begin{equation}
@@
 i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
-the set of training instance that lie in the \(k\)-th bin, i.e.
+the set of training instances of the \(k\)-th bin, i.e.
 \( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
 \). Afterwards,~\citep{gkolemis22} proposed the Differential ALE
-(DALE) that computes the local effects on the training instances using
-auto-differentiation:
+(DALE) that uses the partial derivatives:
 \begin{equation}
@@
 %
-Their method has the advantages of remaining on-distribution even when
-bins become wider and, most importantly, allows the recomputation of
+Their method has the advantages of remaining on-distribution when
+bins become wide and, most importantly, allows the recomputation of
 the accumulated effect with different bin-splitting with near-zero
@@
 
-Both approximations ask from the user to blindly decide the number of
-bins, denoted with \(K\), for splitting the axis into equally-sized
-bins. This approach introduces significant limitations. Specifically, setting \(K\) to a small
-value may hide fine-grain effects due to large bins while setting \(K\)
-to a high value leads to poor bin-effect estimations from limited
-samples. In general, the user may face contradictory explanations for
+Both approximations ask from the user to blindly decide the bin
+density for splitting the feature axis into \(K\) equisized bins, an
+approach that is error prone. Setting \(K\) to a small value may hide
+fine-grain effects due to large bins while setting \(K\) to a high
+value leads to poor bin-effect estimations due to few samples per
+bin. In general, the user may receive contradictory explanations for
 different \(K\) without a clue for which one to trust.
@@
 Section~\ref{sec:UALE-definition-1} we define UALE, at
-Section~\ref{sec:UALE-definition-2} we redefine it using an
-interval-based formulation, at
-Section~\ref{sec:UALE-approximation} we provide an
-approximation and at Section~\ref{sec:bin-spliting} we define and
-solve the problem of optimal bin-splitting.
+Section~\ref{sec:UALE-definition-2} we reformulate it using an
+interval-based formulation, at Section~\ref{sec:UALE-approximation} we
+provide UALE approximation and at Section~\ref{sec:bin-spliting} we
+define and solve the problem of optimal bin-splitting.
 
@@
 
-We extend ALE definition of Eq.~(\ref{eq:ALE}) with a component for
+UALE extends ALE definition of Eq.~(\ref{eq:ALE}) with a component for
 quantifying the uncertainty. We denote as \(\mathcal{H}(z)\) the
 uncertainty of the local effects at a specific point \(x_s=z\) and we
-quantify it as the standard deviation of the local explanations
+quantify it with the standard deviation of the local explanations
 \(\mathcal{H}(z) = \sigma(z)\), where:
@@
 \noindent
-UALE formulates the effect at a specific point \(x_s\) with a compact
-tuple that consists of the average effect and the uncertainty
-\((\mu(z), \sigma(z))\) and visualizes them as a continuous curve with
-a confidence region, i.e.
+UALE formulates the effect at a specific point \(x_s\) with a tuple
+that consists of the average effect and the uncertainty
+\((\mu(z), \sigma(z))\) and visualizes it with a continuous curve and
+a confidence region
 \(f^{\mathtt{UALE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
@@
 
-In real scenarios, all estimations are based on the instances of the
-training set. Therefore, we cannot estimate \(\mu(z), \sigma(z)\) at
-the granularity of a point, because the probability of observing a
+In real ML scenarios, all estimations are based on the instances of
+the training set. Therefore, we cannot estimate \(\mu(z), \sigma(z)\)
+at the granularity of a point, because the probability of observing a
 sample inside the region \([x_s - h, x_s + h]\) tends to zero, when
 \(h \to 0\). We define the regional-effect \(\mu(z_1, z_2)\) and the
-regional-uncertainty \(\mathcal{H}(z_1, z_2) = \sigma(z_1, z_2)\), as the summary statistics
-that characterize the effect and the uncertainty inside the interval
-\([z_1, z_2)\):
+regional-uncertainty \(\mathcal{H}(z_1, z_2) = \sigma(z_1, z_2)\), as
+the summary statistics that characterize the effect and the
+uncertainty inside the interval \([z_1, z_2)\):
 

### 2022-10-10 AISTATS2023/paper.tex
review on simulation experiments

@@
 
+\todo{Review all the description}
 For achieving a computationally-grounded solution we set a threshold
@@
 
-Simulation examples provide the freedom to select the data-generating
-distribution \(p(\mathbf{X})\) and the predictive function
-\(f(\cdot)\). We follow this common XAI
-practice~\citep{aas2021explaining, herbinger2022repid} for empirically
-evaluating UALE against a solid ground-truth. The first group of
-synthetic examples (Section~\ref{sec:simulation-examples-1}) compares
-UALE with PDP-ICE plots in quantifying the average and the uncertainy
-of the effect. The second group
-(Section~\ref{sec:simulation-examples-2}) evaluates the accuracy of
-the approximation (average effect and uncertainty) using the automatic
-bin-splitting algorithm. In both groups, we illustrate the most
-indicative examples; a more extensive evaluation is provided in the
-Appendix.
+In XAI it is common~\citep{aas2021explaining, herbinger2022repid} to
+use simulation examples for compering methods, because the
+data-generating distribution \(p(\mathbf{X})\) and the predictive
+function \(f(\cdot)\) are known, and, therefore, the ground-truth is
+accesible. We follow this common XAI practice and we split the
+evaluation in two groups of examples. The first group
+(Section~\ref{sec:simulation-examples-1}) compares UALE definition
+against PDP-ICE in modeling the average effect and the
+uncertainty. The second group
+(Section~\ref{sec:simulation-examples-2}) compares UALE approximation
+using the automatic bin-splitting against the fixed-size alternative,
+in estimating the average effect and the uncertainty.
 
-\subsection{Case 1: Uncertainty Quantification}
+\subsection{Case 1: UALE vs PDP-ICE}
 \label{sec:simulation-examples-1}
 
-In this simulation, we will compare UALE method against PDP-ICE in
-quantifying the main effect and the uncertainty. The samples that we
-use in all examples are coming from the following distribution:
+In this simulation, we compare UALE against PDP-ICE. The aim of this
+example is to highlight that when features are correlated, PDP and ICE
+fail in quantifying the average effect and the heterogeneity, even
+when the black-box function is relatatively simple.
+
+We use the following distribution:
 \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
 \(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
-\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We create the following
-piecewise linear function:
-
-% Since there is
-% ambiguity about the ground-truth first-order effect in cases of
-% correlated features,
-% e.g.~\citep{apley2020visualizing,Gromping2020MAEP},
+\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\) and the following piecewise
+linear function:
 
@@
 \(f_{\mathtt{int}}(\mathbf{x}) = x_1x_3\). The linear term
-\(f_{\mathtt{lin}}\) includes the two correlated features and
-\(f_{\mathtt{int}}\) interacts the two non-correlated features.  We
+\(f_{\mathtt{lin}}\) includes the two correlated features and the
+interaction term \(f_{\mathtt{int}}\) the two non-correlated.  We
 evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
@@
 (\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha \neq 0\)) with
-equal weights (\(a_1=a_2\)).
-
-In all cases, we provide the ground-truth average effect and
-uncertainty computed with analytical derivations (proofs at the
-Appendix) and we compare it against the approximation provided by each
-method. The approximation is computed after generating \(N=500\)
-samples.
+equal weights (\(a_1=a_2\)). In all cases, we provide the ground-truth
+average effect and uncertainty computed with analytical derivations
+(proofs at the Appendix) and we compare it against the approximations
+provided by each method, after generating \(N=500\) samples.
 
@@
 
-We test the feature effect when no interaction exits \(\alpha=0\) and
-the weights are equal \(a_1=a_2=1\). The ground truth effect
-\(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\) when \(0 \leq x_1 < 0.25\),
-\(-x_1\) when \(0.25 \leq x_1 < 0.5\) and zero otherwise. There zero
-heterogeneous effects, therefore
-\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
-Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP main effect is
+In this set-up, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\)
+(equal weights). The ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\)
+is: \(x_1\) in \([0, 0.25]\), \(-x_1\) in \([0.25, 0.5]\) and \(0\) in
+\([0.5, 1]\). The uncertainty is
+\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) in \([0,1]\) because \(x_1\)
+does not interact with ohter variables. In
+Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP effect is
 wrong and ICE plots imply the existence of heterogeneous effects. In
-contrast, ALE captures correctly the average effect and the zero
-uncertainty and it also groups perfectly the constant-effect regions.
+contrast, UALE models correctly both the average effect and the zero
+uncertainty and approximates it accurately finding a zero aggregate
+residual error partitioning of \(x_1\) domain.
 
@@
   \includegraphics[width=.23\textwidth]{example_1/pdp_ice_feat_0.pdf}
-  \caption{No interaction, Equal weights: Feature effect for \(x_1\);
-    Ground-truth vs (a) UALE method at the left and (b) PDP-ICE at the
-    right.}
+  \caption{No interaction, Equal weights: Feature effect for \(x_1\)
+    using UALE (Left) and PDP-ICE (Right).}
   \label{fig:synth-ex-1-case-1}
@@
 
-\paragraph{No Interaction, Different weights.}
+\paragraph{No Interaction, Non-Equal Weights.}
 
-As before, no interaction exits \(\alpha=0\), therefore, the
-ground-truth uncertainty is zero, i.e.
-\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). The weigts are
-\(a_1=2, a_2=0.5\), therefore, the ground-truth effect is
-\(f_\mu^{\mathtt{GT}}(x_1)\) is, \(2x_1\) when \(0 \leq x_1 < 0.2\),
-\(-2x_1\) when \(0.2 \leq x_1 < 0.4\) and zero otherwise. In
-Figure~\ref{fig:ex-synth-1-2}, we observe that PDP estimation is
-completely opposite to the ground-truth effect, i.e.~negative in the
-region \([0, 0.2)\) and positive in \([0.2, 0.4)\), and the ICE
-erroneously implies the existence of heterogeneous effects. As before,
-ALE quantifies correctly the ground truth effect, the
-zero-uncertainty and extracts the constant effect regions.
+In this set-up, \(\alpha=0\) (no interaction), \(a_1 = 2\) and
+\(a_2= 0.5\) (non-equal weights). The non-equal weights have
+implications at both the gradient and the interval of the piece-wise
+linear regions, i.e., \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) in
+\([0, 0.2]\), \(-2x_1\) in \([0.2, 0.4]\) and \(0\) in \([0.4,
+1]\). As before, the ground-truth uncertainty is
+\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) because \(x_1\) does not
+interact with other features.  In Figure~\ref{fig:ex-synth-1-2}, we
+observe that PDP estimation is completely opposite to the ground-truth
+effect, i.e.~negative in the region \([0, 0.2)\), positive in
+\([0.2, 0.4)\), and the ICE erroneously implies the existence of
+heterogeneous effects. As before, ALE quantifies correctly the ground
+truth effect, the zero-uncertainty and partioning correctly the
+\(x_1\) domain.
 
@@
   \includegraphics[width=.23\textwidth]{example_2/pdp_ice_feat_0.pdf}
-  \caption{No interaction, Different weights: Feature effect for \(x_1\);
-    Ground-truth vs (a) UALE method at the left and (b) PDP-ICE at the
-    right.}
+  \caption{No interaction, Different weights: Feature effect for \(x_1\)
+    using UALE (Left) and PDP-ICE (Right).}
   \label{fig:ex-synth-1-2}
@@
 
-In this case we activate the interaction term, i.e. \(a=1\), and we
-set the weights to \(a_1=a_2=1\). The interaction term provokes
-heterogeneous effects for features \(x_1, x_3\), because the local
-effects at \(x_1\) depend on the unknown value of \(X_3\) and
+In this set-up, we activate the interaction term, i.e.  \(a=1\),
+keeping the weights equal \(a_1=a_2=1\). The interaction term provokes
+heterogeneous effects for features \(x_1\) and \(x_3\), because the
+local effects at \(x_1\) depend on the unknown value of \(X_3\) and
 vice-versa. The effect of \(x_2\) has zero-uncertainty since it does
@@
 Figure~\ref{fig:ex-synth-1-3}, UALE correcly models the average effect
-and the uncertainty in all cases. PDP-ICE quantifies correctly the
-effect and the uncertainty only in the case of feature \(x_3\),
-because \(X_3\) is independent from other features. For features
-\(x_1, x_2\) that are correlated the average effect of PDP
-and the uncertainty of ICE plots are wrong.
+and the uncertainty in all cases.  On the other hand, PDP-ICE
+quantifies correctly the average effect and the uncertainty only in
+the case of feature \(x_3\), because \(X_3\) is independent from other
+features. For the correlated features (\(x_1, x_2\)) both the average
+effect (PDP) and the uncertainty (ICE) are wrong.
 
@@
   \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_2.pdf}\\
-  \caption{With interaction, equal weights: Feature effect for all
-    features, \(x_1\) to \(x_3\) from top to bottom; Ground-truth vs
-    (a) UALE method at the left columns and (b) PDP-ICE at the right
-    column.}
+  \caption{With interaction, equal weights: From top to bottom,
+    feature effect for features \(\{x_1, x_2, x_3\}\) using UALE (left
+    column) and PDP-ICE (right column).}
   \label{fig:ex-synth-1-3}
@@
 
-Despite the model's simplicity, PDP-ICE fail in quantifying the
-average effect and the uncertainty due to correlation between features
-\(X_1\) and \(X_2\). UALE quantifies them correctly and estimates them
-accurately extracting the regions with constant effect. The examples
-above do not cover the case of an interaction term between correlated
-features, e.g. \(x_1x_2\), because there is an open debate about the
-ground-truth effect in this case~\citep{Gromping2020MAEP}.
+Despite the model's simplicity, PDP-ICE fail in modeling both the
+average effect and the uncertainty. The error is due to the method's
+formulation, i.e.~the use of the marginal distribution that ignores
+correlations between features, and not due to poor approximation
+because of limited samples. In contrast, UALE models both correctly
+and estimates them accurately extracting the regions with constant
+effect.  The examples above do not cover the case of an interaction
+term between correlated features, for example a term \(x_1x_2\),
+because there is an open debate about the ground-truth effect in this
+case~\citep{Gromping2020MAEP}.
 
@@
 
-In this simulation, we aim to to quantify the advantages of automatic
-bin-splitting. We set-up the following validation framework. We
-generate a big dataset with dense sampling (\(N=10^6\)) and we treat
-the DALE estimation with dense fixed-size bins (\(K=10^3\)) as
-ground-truth. Afterwards, we generate less samples (\(N=500\)) and we
-compare the fixed-size DALE estimation (for many different \(K\))
-versus the automatic bin-splitting. In all cases, we use a bivariate
-black-box function \(f(\cdot)\) where the samples are instances of the
-distribution \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
+In this simulation, we illustrate the advantages of automatic
+bin-splitting against the fixed-size alternative. For this reason, we
+generate a very big dataset applying dense sampling (\(N=10^6\)) and
+we treat the estimation with dense fixed-size bins (\(K=10^3\)) as the
+ground-truth UALE. Afterwards, we generate less samples (\(N=500\))
+and we compare the fixed-size estimation (for many different \(K\))
+against UALE automatic bin-splitting. In all set-ups, we sample from
+\(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
 \(x_1 \sim \mathcal{U}(0,1)\) and
@@
 bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size
-splitting with \(K\) bins.  For consistent results, in all the
-examples below, we regenerate samples and repeat the computations for
-\(t = 30\) times, providing the mean value for all metrics.
+splitting with \(K\) bins. The evaluation metrics we report are the
+average number of \(t = 30\) independent runs, using each time
+\(N=500\) different samples.
 
@@
 
-The evaluation is done with regard to two metrics that count the mean
-bin error, where bin error is the absolute difference of the
-approximation from the ground-truth. The metric
-\(\mathcal{L}_{\mathtt{DP|K}}^{\mu} =
-\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
-  \mathcal{Z}^{\mathtt{DP|K}}} | \mu_k - \hat{\mu}_k | \) counts the
-mean bin error on the average effect and
-\(\mathcal{L}_{\mathtt{DP|K}}^{\sigma} =
-\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
-  \mathcal{Z}^{\mathtt{DP|K}}} | \sigma_k - \hat{\sigma}_k | \) the
-mean bin error on the uncertainty. In both cases the ground truth is
-computed from the dense bins of ground-truth that overlapse with the
-wider bin of the approximation. We also provide the mean (per bin)
-error term
+The evaluation is done with regard to two metrics counting the mean
+error per bin, where the bin error is definde as the absolute
+difference of the approximation from the ground-truth. The first
+metric, Eq.~\eqref{eq:eval_met_1}, counts the mean bin error wrt
+average effect and the second, Eq.~\eqref{eq:eval_met_2} the mean bin
+error wrt uncertainty:
+
+\begin{equation}
+  \label{eq:eval_met_1}
+  \mathcal{L}_{\mathtt{DP|K}}^{\mu} = \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
+  \mathcal{Z}^{\mathtt{DP|K}}} | \mu(z_k- z_{k-1}) - \hat{\mu}(z_k-
+  z_{k-1}) |
+\end{equation}
+
+\begin{equation}
+  \label{eq:eval_met_2}
+  \mathcal{L}_{\mathtt{DP|K}}^{\sigma} = \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
+    \mathcal{Z}^{\mathtt{DP|K}}} | \sigma(z_k- z_{k-1}) - \hat{\sigma}(z_k-
+  z_{k-1}) |
+\end{equation}
+
+The ground truth \(\mu_k\) (\(\sigma_k\)) is the mean value of the
+bin-effects (bin-uncertainties) of the bins inside the interval
+defined by the wide bin. For better interpretation of the uncertainty
+error, we also provide the mean (per bin) error term
 \(\mathcal{L}^{\rho}_{\mathtt{DP|K}} =
 \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
-  \mathcal{Z}^{\mathtt{DP|K}}} \rho_k \) for a complete understanding
-of the uncertainty error.
+  \mathcal{Z}^{\mathtt{DP|K}}} \mathcal{\Epsilon}(z_{k-1}, z_k) \).
 
@@
 
-In this example, we define \(f(\mathbf{x}) = a_1x_1 + x_1x_2\) with
-\(5\) piecewise-linear regions of different-size, i.e., \(a_1\) equals
-to \(=\{2, -2, 5, -10, 0.5\}\) in the intervals defined by the
-sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). As we observe, in
-Figure~\ref{fig:ex-synth-2-1}, UALE method extracts a sequence of
-intervals with better \(\mathcal{L}_{\mathtt{DP}}^{\mu}\) and
-\(\mathcal{L}_{\mathtt{DP}}^{\sigma^2}\) error compared to any
-fixed-size splitting. Analyzing the fixed-size errors helps us
-understand the importance of variable-size splitting. In
-Figure~\ref{fig:ex-synth-2-1}(b), we observe a positive trend between
-\(\mathcal{L}^{\mu}_{\mathtt{K}}\) and \(K\), concluding that bin
-effect estimation is more incosistent as \(\Delta x\) becomes smaller,
-due to less points contributing to each bin. The interpretation of
-variance error is slightly more complex. Given that the smallest
-interval is \(\Delta x = 0.05 \Rightarrow K = 20\) and all intervals
-are multiples of the smallest interval, any \(K\) that is not a
-multiple of \(20\) adds nuisance uncertainty
-\(\mathcal{L}^{\rho^2}_{\mathtt{K}}\) leading to a high variance error
-\(\mathcal{L}^{\sigma^2}_{\mathtt{K}}\). In these cases, the variance
-error reduces as \(K\) grows bigger because the length of the bins
-that lie in the limit between two piecewise linear regions becomes
-smaller. For \(K=\{20, 40, 60, 80, 100\}\) where
-\(\mathcal{L}^{\rho^2}_{\mathtt{K}} \approx 0\), we conclude the same
-as with with the mean effect error, i.e. the estimation becomes more
-incosistent as \(K\) grows larger.
-
-Variable-size extracts correctly the fine-grain bins, e.g., intervals
-\([0.4, 0.45]\), \([0.45, 0.5]\), and acts as a merging mechanism to
-create wider bins when the effect remains constant, e.g. interval
-\([0.5, 1]\), leading to an optimal solution.
+In this set-up, \(f(\mathbf{x}) = a_1x_1 + x_1x_2\) is a
+piecewise-linear function with \(5\) different-width regions, i.e.,
+\(a_1\) equals to \(=\{2, -2, 5, -10, 0.5\}\) in the intervals defined
+by the sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). As we observe, in
+Figure~\ref{fig:ex-synth-2-1}, UALE's approximation is better than all
+fixed-size alternatives, in both mean effect (\(\mathcal{L}^{\mu}\))
+and uncertainty (\(\mathcal{L}^{\sigma}\)). For understanding the
+importance of variable-size splitting, we analyze the fixed-bin
+error. In the top-right of Figure~\ref{fig:ex-synth-2-1}, we observe
+the Law of Large Numbers (LLN); \(\mathcal{L}^{\mu}_{\mathtt{K}}\) and
+\(K\) have positive correlation because as \(\Delta x\) becomes
+smaller less points contributing to each estimation. The
+interpretation of \(\mathcal{L}^{\sigma}\) is more complex. Given that
+the smallest piecewise-linear region is \(\Delta x = 0.05 \) any
+\(K\) that is not a multiple of \(\frac{1}{\Delta x} = 20\), adds
+nuisance uncertainty \(\mathcal{L}^{\rho}_{\mathtt{K}} > 0\) leading
+to a biased uncertainty approximation. For these \(K\), uncertainy
+error decreases for bigger \(K\) because the length of the bins that
+lie in the limit between two piecewise linear regions becomes
+smaller. For \(K\) that are multiple of \(20\), i.e.
+\(K=\{20, 40, 60, 80, 100\}\), the uncertainty approximation is
+unbiased \(\mathcal{L}^{\rho}_{\mathtt{K}} \approx 0\). Here, as in
+\(\mathcal{L}^{\mu}\), uncertainty estimation is worse as \(K\) grows
+larger (LLN). As we observe in the top left of
+Figure~\ref{fig:ex-synth-2-1}, automatic bin-splitting separates the
+fine-grain bins, e.g.~regions \([0.4, 0.45]\), \([0.45, 0.5]\), and
+unites constant-effect regions into a single bin, e.g.~region
+\([0.5, 1]\).
 
@@
 
-In this example, we define a black-box function
-\(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), where the effect is
-non-linear in all the range of \(x\). This case has two
-specialties. First, there is no obvious advantage of variable-size
-versus fixed-size splitting, and, second, there is not an apriori
-optimal bin-size. Widening a bin will increase the resolution error
-\(\mathcal{L}^{\rho^2}\) and narrowing will make less robust. In
-Figure~\ref{fig:ex-synth-2-2}, we observe that automatic bin splitting
-finds a solution that compromises the conflicting objectives, i.e., it
-keeps as low as possible both the main effect
-\(\mathcal{L}^{\mu}_{\mathtt{K}}\) and the variance error
-\(\mathcal{L}^{\sigma^2}_{\mathtt{K}}\).
+In this set-up, \(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), so the
+effect is non-linear in \(0 \leq x_1 \leq 1\). In this set-up, wide
+bins increase \(\mathcal{L}^{\rho^2}\) making the uncertainty
+approximation more biased and narrow bins lead to a worse
+approximation. In Figure~\ref{fig:ex-synth-2-2}, we observe that
+automatic bin splitting manages to compromise the
+conflicting objectives, \todo{Explain better}
 
@@
 
-We exclude instances with missing values or with outlier values. As
-outlier we define the feature values which ar over three standard
-deviations away from the mean feature value. We also normalize all
-features to zero-mean and unit standard deviation. We split the
-dataset into \(N_{tr} = 15639\) training and \(N_{test} = 3910\) test
-examples (80/20 split) and we fit a Neural Network with 3 hidden
-layers of 256, 128 and 36 units respectively. After 15 epochs using
-the Adam optimizer with learning rate \(\eta = 0.02\), the model
-achieves a normalized mean squared error (R-Squared) of \(0.25\)
-(0.75), which corresponds to a MAE of \(37 \cdot 10^3\) dollars.
+We exclude instances with missing and outlier values. As outlier we
+define the feature values which ar over three standard deviations away
+from the mean feature value. We also normalize all features to
+zero-mean and unit standard deviation. We split the dataset into
+\(N_{tr} = 15639\) training and \(N_{test} = 3910\) test examples
+(80/20 split) and we fit a Neural Network with 3 hidden layers of 256,
+128 and 36 units respectively. After 15 epochs using the Adam
+optimizer with learning rate \(\eta = 0.02\), the model achieves a
+normalized mean squared error (R-Squared) of \(0.25\) (0.75), which
+corresponds to a MAE of \(37 \cdot 10^3\) dollars.
 
@@
 
-add limitations.
+\todo{Add Conclusion, limitations, and future work}
 

### 2022-10-11 AISTATS2023/paper.tex
with Eirini's review until chapter 1

--- 2022-10-09 AISTATS2023/paper.tex
@@
 of accurate predictions along with meaningful explanations to support
-them. For this reason, there is an increased interest in Explainable
-AI (XAI) to understand the decision mechanism of complex black-box
+them. For this reason there is an increased interest in Explainable AI
+(XAI), to understand the decision mechanism of complex black-box
 models. XAI literature distinguishes between local and global
@@
 instance-level explanations into a single interpretable outcome,
-usually a number or a plot. When strong interactions between features
-exist, the instance-level explanations may significantly deviate from
-the aggregated outcome~\citep{herbinger2022repid}, a phenomenon called
-\emph{aggregation bias}~\citep{mehrabi2021survey}. Therefore, global
-methods should quantify not only the aggregated (averaged) effect, but
-also the level of heterogeneity of instance-level explanations. The
-same holds for Feature Effect (FE)~\citep{Gromping2020MAEP}, a class
-of global XAI methods that isolate the impact of a single feature on
-the output.\footnote{FE methods also isolate the combined effect of a
+usually a number or a plot.
+A popular class of global methods are the so-called Feature Effect (FE) methods~\citep{Gromping2020MAEP} that isolate the impact of a single feature on
+the output\footnote{FE methods also isolate the combined effect of a
   pair of features to the output. Combinations of more than two
   features are uncommon, since they are difficult to estimate and
-  visualize.} For example, a feature with zero average effect may
-indicate no effect on the output or a highly positive effect in some
-cases and a highly negative effect in others. The ambiguity about the
-correct interpretation can be resolved, if the FE method informs about
-the heterogeneity of instance-level explanations, a quantity we refer
-to as the \emph{uncertainty of the effect}.
-
-There are three widely-used FE methods: \emph{Partial Dependence
+  visualize.}. There are three widely-used FE methods: \emph{Partial Dependence
   Plots} (PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
 (MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
-(ALE)\citep{apley2020visualizing}. ALE has been established as the
+(ALE)\citep{apley2020visualizing} with ALE being established as the
 state-of-the-art for measuring the average effect, since PDP and MP
 have been criticized~\citep{Gromping2020MAEP} to compute erroneous
-effects when the input features are (highly) correlated. However, ALE
-faces two limitations, the first concerns ALE definition and the
-second ALE approximation. Regarding the definition, ALE does not
-inform the user about the uncertainty. In contrast, in PDP the
+effects when the input features are (highly) correlated.
+
+However,
+when strong interactions between features exist,
+the instance-level feature effects may significantly deviate from
+the aggregated outcome of ALE~\citep{herbinger2022repid}, a phenomenon called
+\emph{aggregation bias}~\citep{mehrabi2021survey}. 
+Aggregation bias leads to ambiguities; for example, a feature with zero average effect may
+indicate no effect on the output or a highly positive effect for some instances and a highly negative effect for others.
+Such ambiguities could be resolved if not only the average effect but also the deviation of instance-level feature effects, a quantity we refer to as the \emph{uncertainty of the effect}, from the average effect is provided.
+
+%Therefore, global methods should quantify not only the aggregated (averaged) effect, but also the level of heterogeneity of instance-level explanations.
+
+%There are three widely-used FE methods: \emph{Partial Dependence Plots} (PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
+%(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
+%(ALE)\citep{apley2020visualizing}. ALE has been established as the
+%state-of-the-art for measuring the average effect, since PDP and MP
+%have been criticized~\citep{Gromping2020MAEP} to compute erroneous
+%effects when the input features are (highly) correlated.
+
+In this work, we propose UALE,  a global feature-effect method that extends ALE by quantifying the uncertainty associated with the average effect. 
+Similarly to ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting).
+In contrast to ALE that requires a fixed equi-width partitioning without considering the characteristics of
+the instance-level effects, we propose an automatic partitioning that minimizes the uncertainty of the resulting bins.
+
+%However, ALE faces two limitations, the first concerns ALE definition and the second ALE approximation. Regarding the definition, ALE does not inform the user about the uncertainty. 
+\color{blue}\todo{This should go to related work}In contrast, in PDP the
 heterogeneity of instance-level effects can be identified by exploring
 the Individual Conditional Expectations
-(ICE)~\citep{goldstein2015peeking}. Regarding the approximation, ALE
-requires partitioning the feature domain into non-overlapping
-intervals (bin-splitting) and estimate the effect inside each interval
-(bin-effect) from the training set instances that fall inside. Finding
+(ICE)~\citep{goldstein2015peeking}. \color{black}
+%Regarding the approximation, ALE requires partitioning the feature domain into non-overlapping intervals (bin-splitting) and estimate the effect inside each interval (bin-effect) from the training set instances that fall inside.
+\color{blue}\todo{to be moved in related work}
+Finding
 an appropriate sequence of bins is of particular importance, since
@@
 split \emph{blindly}, i.e.~without considering the characteristics of
-the instance-level effects, in \(K\) equisized bins.
+the instance-level effects, in \(K\) equisized bins.\color{black}
 
-In this paper, we present Uncertainty-aware Accumulated Local Effects
-(UALE), a method that visualizes in a single plot both the average
-feature effect and the uncertainty, i.e.~the level of heterogeneity of
-instance-level effects. The average effect is defined as in ALE and
-the uncertainty inherits ALE's advantages, i.e.~takes into account the
-correlations between features. We also propose an accurate UALE
-approximation. We formally prove that for an unbiased estimation of
-the uncertainty, the partitioning of the feature domain must respect
-specific conditions. Using that, we set-up an optimization problem
-where we search for a partitioning, a sequence of \emph{variable-size}
-intervals, (a) with enough instances (accurate estimation) and (b)
-that respect the above conditions (unbiased estimation). Finally, we
-propose a method for automatically finding the optimal
-partitioning. We apply our method in multiple synthetic and real world
-datasets for evaluating its performance. The contributions of this
+%In this paper, we present Uncertainty-aware Accumulated Local Effects(UALE), a method that visualizes in a single plot both the av erage feature effect and the uncertainty, i.e.~the level of heterogeneity of instance-level effects.
+The average effect is defined as in ALE and the uncertainty inherits ALE's advantages, i.e.~takes into account the correlations between features.
+We also propose an accurate UALE approximation.
+We formally prove that for an unbiased estimation of the uncertainty, the partitioning of the feature domain must respect specific conditions.
+Using that, we set-up an optimization problem where we search for a partitioning, a sequence of \emph{variable-size}intervals, (a) with enough instances (accurate estimation) and (b) that respect the above conditions (unbiased estimation).
+Finally, we propose a method for automatically finding the optimal partitioning.
+We apply our method in multiple synthetic and real world datasets for evaluating its performance.
+The contributions of this
 paper are:
-
 \begin{itemize}
-\item UALE, a feature effect method that quantifies both the average
+\item A global feature effect method that quantifies both the average
   effect and the uncertainty, i.e.,~the heterogeneity of instance level
   effects.
-\item A formal proof of the conditions the feature domain partitioning
-  must fulfill for an unbiased approximation of the uncertainty.
-\item An accurate UALE approximation by automatically partitioning
-  into variable-size intervals that respect the above conditions.
+\item An unbasied approximation of the uncertainty through a dynamic uncertainty-driven bin partitioning of the feature domain.
+%A formal proof of the conditions the feature domain partitioning must fulfill for an unbiased approximation of the uncertainty.
+\item A thorough evaluation on both synthetic and real datasets.
+%An accurate UALE approximation by automatically partitioning into variable-size intervals that respect the above conditions.
 \end{itemize}
@@
 \section{BACKGROUND AND RELATED WORK}
+The necessary background for ALE is presented in Section~\ref{sec:feat-effect-meth}), whereas ALE approximations are presented in Section~\ref{sec:ale-approximation}.
 
-We here review the existing methods for modeling the average effect
-and the heterogeneity of instance-level effects and we focus on ALE
-definition~\citep{apley2020visualizing} and
-approximation~\citep{apley2020visualizing, gkolemis22}.
+The approximation 
+In this section, we describe the basic methods for FE and uncertainty
+quantification, focusing on ALE definition. We, then, review the ALE
+approximation\citep{apley2020visualizing, gkolemis22}, describing some
+of its vulnerabilities.
 
-\paragraph{Notation.} We refer to random variables (rv) using
-uppercase \( X \), to simple variables with plain lowercase \( x \)
-and to vectors with bold \( \xb \). Often, we partition the input
-vector \(\xb \in \mathbb{R}^D\) to the feature of interest
-\(x_s \in \mathbb{R} \) and the rest of the features
-\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
-\((x_s, \mathbf{x}_c)\), but we it refers the vector
-\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
-corresponding rv as \(\mathbf{X} = (X_s, \mathbf{X}_c)\). When we
-refer to an instance of the training set, we use
-\(\xb^i= (x_s^i, \xc^i) \).
+\paragraph{Notation}
 
-% The black-box function is
-% \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the FE of the \(s\)-th
-% feature is \(f^{\mathtt{<method>}}(x_s)\), with \(\mathtt{<method>}\)
-% indicating the particular method in use.\footnote{An extensive list of
-%   all symbols used in the paper is provided at the Appendix.}
+Let \(\mathcal{X} \in \mathbb{R}^d\) be the d-dimensional feature space, \(\mathcal{Y} \in \mathbb{R}\) the target space and \(f(\cdot) : \mathcal{X} \rightarrow \mathcal{Y}\) the black-box function.
+We use index \(s \in \{1, \ldots, d\}\) for the feature of interest and \(c \in \{1, \ldots, d\} - s\) for the rest feature indexes.
+For convenience, we denote the feature vector \(\xb = (x_1, \cdots , x_s, \cdots, x_D)\) with \((x_s, \xc)\) and the corresponding random variables \(X = (X_1, \cdots , X_s, \cdots, X_D)\) with \((X_s, X_c)\).
+The training set is \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) sampled i.i.d.
+from the distribution \(\mathbb{P}_{X,Y}\).
+Finally, we use \(f^{\mathtt{<method>}}(x_s)\) for the FE of the \(s\)-th feature, with \(\mathtt{<method>}\)
+indicating the particular method in use, for example ALE.\footnote{An extensive list of all
+  symbols used in the paper is provided at the Appendix.}
 
-\subsection{Feature Effect Methods}
+\subsection{Feature Effect Methods and ALE}
 \label{sec:feat-effect-meth}
-
-The three well-known feature effect methods are: PDP, MP and ALE. PDP
-formulates the FE of the \(s\)-th feature as an expectation over the
+The three well-known feature effect methods are: \emph{Partial Dependence
+  Plots} (PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
+(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
+(ALE)\citep{apley2020visualizing}
+PDP
+formulates the FE of the \(s\)-th attribute as an expectation over the
 marginal distribution \(\mathbf{X}_c\), i.e.,
@@
 \(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|x_s}[f(x_s,
-\mathbf{X}_c)]\). ALE defines the local effect at \(x_s = z\) as
-\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). The local
-explanations are then weighted by the conditional distribution
-\(p(\xc|x_s=z)\) to compute the averaged effect \(\mu(z)\). ALE is the
-accumulation of the averaged local effects:
+\mathbf{X}_c)]\). ALE defines the local effect of the \(s\)-th feature
+at instance \(x_s = z\) as
+\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
+local explanations at \(z\) are, then, weighted by the conditional
+distribution \(p(\xc|x_s=z)\) and are averaged, to produce the
+averaged effect \(\mu(z)\). ALE is the accumulation of the averaged
+local effects:
 
@@
 
-Eq.~(\ref{eq:ALE}) has important advantages compared to PDP and MP in
-cases of correlated features. In these cases, PDP integrates over
+Eq.~(\ref{eq:ALE}) has specific advantages which gain particular value in cases of
+correlated input features. In these cases, PDP integrates over
 unrealistic instances due to the use of the marginal distribution
@@
 the combined effect of sets of features to a single feature. ALE
-resolves both issues.
+manages to resolve both issues, and is therefore the most trustable
+method in cases of correlated features. \todo{cite someone}
 
-\subsection{Heterogeneity Of Instance-Level Effects.}
+\subsection{Quantification Of Heterogeneous Effects.}
 \label{sec:quant-heter-effects}
 
-FE methods show what is the average (global) effect on the output, if
-the value of a specific feature is increased/decreased. For completing
-the explanation about a feature, we need to also know to what extent
-the local effects deviate from the global explanation, i.e.~to
-quantify the uncertainty of the average effect.
-
-Heterogeneity It comes naturally to also ask \textit{to
-  what extent }. Quantifying the uncertainty of the global explanation
+FE methods reply to the question \textit{what is the average (global)
+  effect on the output, if the value of a specific feature is
+  increased/decreased}. It comes naturally to also ask \textit{to what
+  extent the local effects \todo{add what local means here} deviate from the global (average)
+  effect/explanation}. 
+  \todo{A connector line is needed here on the connection between local-global and uncertainty}
+  Quantifying the uncertainty of the global explanation
 has attracted a lot of interest. ICE and
@@
 \xc^i)\) for d-ICE. The user then visually observes if the curves are
-homogeneous and to what extent they deviate from the PDP. Some methods
+homogeneous, i.e., all instances have similar values, and to what extent they deviate from the PDP. Some methods
 try to automate the aforementioned visual exploration, by grouping
@@
 values\citep{lundberg2018consistent}, quantify the level of
-interaction between the input features, with an interaction value. A
+interaction between the input features, with an interaction value \todo{not clear what htis value is, maybe add a few words}. A
 strong interaction index is an indicator for the existence of
@@
 The aforementioned approaches are under two limitations; They either
-do not quantify the uncertainty of the FE directly or they are based
+do not quantify the uncertainty of the FE directly \todo{I dont get this argument. What do you mean directly?} or they are based
 on PDPs, and, therefore, they are subject to the failure modes of PDPs
-in cases of correlated features\citep{baniecki2021fooling}, as we will
-show in-depth in Section~\ref{sec:simulation-examples-1}. To the best
+in cases of correlated features\citep{baniecki2021fooling}, as we will also show in our experimental analysis (c.f., Section~\ref{sec:simulation-examples-1}). To the best
 of our knowledge, no work exist so far that quantifies the
 heterogeneous effects based on the formulation of ALE.
+\todo{My understanding of the contribution is also the last sentence.
+None has done it for ALE, but it has been done f for PDPs.
+For this I think we need some more related owrk on the limiations of PDPs.
+Also, is this work a straight-forward transfer of the d-ICE etc methods decribed above or transfering these ideas to ALE requires a different approach?}
 
@@
 the feature domain in \(K\) equal-sized bins and estimating the local effects in
-each bin by evaluating the black box-function at the bin limits:
+each bin by evaluating the black box-function $f()$ at the bin limits: \todo{k or K? I prefer k for the cardinality and it seems that k is used afterwards. }
 \begin{equation}
@@
 \( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
-\). Afterwards,~\citep{gkolemis22} proposed the Differential ALE
+\).
+
+\citep{gkolemis22} proposed the Differential ALE
 (DALE) that computes the local effects on the training instances using
 auto-differentiation:
-\begin{equation}
-  \label{eq:DALE_accumulated_mean_est}
+\begin{equation}  \label{eq:DALE_accumulated_mean_est}
   \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
@@
 bins become wider and, most importantly, allows the recomputation of
-the accumulated effect with different bin-splitting with near-zero
+the accumulated effect with different bin-splitting with near-zero additional
 computational cost.
 
-Both approximations ask from the user to blindly decide the number of
-bins, denoted with \(K\), for splitting the axis into equally-sized
-bins. This approach introduces significant limitations. Specifically, setting \(K\) to a small
+Both approximations, however, require the number of bins \(K\) as input and perform an  equally-sized bin splitting of the attribute domain.
+The choice of \(K\) is therefore critical and introduces significant limitations.
+Specifically,
+setting \(K\) to a small
 value may hide fine-grain effects due to large bins while setting \(K\)
@@
 samples. In general, the user may face contradictory explanations for
-different \(K\) without a clue for which one to trust.
+different \(K\) values without a clue for which one to trust.
 
@@
 ground-truth uncertainty is zero, i.e.
-\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). The weigts are
+\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). The weights are
 \(a_1=2, a_2=0.5\), therefore, the ground-truth effect is

### 2022-10-12 AISTATS2023/paper.tex
revised method description

--- 2022-10-10 AISTATS2023/paper.tex
@@
 \usepackage{amssymb}
+\DeclareMathOperator*{\argmin}{arg\,min}
+\usepackage{amsthm}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Corollary}[theorem]
+\newtheorem{lemma}[theorem]{Lemma}
+
 \usepackage{xfrac}
@@
 \begin{abstract}
-
-Accumulated Local Effects (ALE) is a popular approach to explainable AI that quantifies how a feature influences the decisions of a model taking into account feature correlations.
-However, in case of strong interactions between features, instance-level feature effects may deviate from the average curve provided by ALE.
-Therefore, it is crucial to quantify this deviation, namely, the uncertainty of the effect.
-In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and the uncertainty.
-Furthermore, as in ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting).
-The average effect and the uncertainty are then computed from the instances that lie in each bin.
-In this work, we formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints.
-Based on this, we propose a method for automatically finding the optimal intervals, balancing the trade-off between estimation bias and variance~\todo{check if this phrase explains well the concept.} We demonstrate through synthetic and real datasets (a) the advantages of modeling the uncertainty with UALE compared to other methods and (b) the importance of appropriate bin splitting for an accurate approximation of the average effect and uncertainty.
+Accumulated Local Effects (ALE) is a popular explainable AI method that quantifies how a feature influences the decisions of a model, handling well feature correlations.
+In case of complex interactions between features, instance-level feature effects may deviate from the ALE curve.
+It is therefore crucial to quantify this deviation, namely, the uncertainty of the effect.
+In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and its uncertainty.
+We show that UALE quantifies uncertainty effectively, even in case of correlated features.
+We also note that as in ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting).
+The average effect and the uncertainty are computed from the instances that lie in each bin.
+We formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints.
+Based on this, we propose a method to determine the optimal intervals, \todo{balancing the estimation bias and variance.} We demonstrate, through synthetic and real datasets, (a) the advantages of modeling the uncertainty with UALE compared to alternative methods and (b) the effectiveness of UALE's appropriate bin splitting for a \todo{good} approximating of the average effect and its uncertainty.
 \end{abstract}
@@
 
-Recently, Machine Learning (ML) has been adopted in critical domains,
-such as healthcare and finance. In these areas, we need a combination
-of accurate predictions along with meaningful explanations to support
-them. For this reason, there is an increased interest in Explainable
-AI (XAI) to understand the decision mechanism of complex black-box
-models. XAI literature distinguishes between local and global
-techniques~\citep{Molnar2020interpretable}. Local methods provide
-instance-level explanations, i.e.~explain the prediction for a
+Recently, Machine Learning (ML) has been widely adopted in mission critical domains,
+such as healthcare and finance.
+In such high-stakes areas,  it is important not only to provide accurate predictions but also meaningful explanations.
+For this reason, there is an increased interest in Explainable
+AI (XAI). XAI literature distinguishes between local and global
+methods~\citep{Molnar2020interpretable}. Local methods provide
+instance-level explanations, i.e., explain the prediction for a
 specific instance, whereas global methods summarize the entire model
-behavior.
+behavior. Global methods create a global explanation by aggregating the
+instance-level explanations into a single interpretable outcome,
+usually a number or a plot.
 
-Global methods create a global explanation by aggregating the
-instance-level explanations into a single interpretable outcome,
-usually a number or a plot. When strong interactions between features
-exist, the instance-level explanations may significantly deviate from
-the aggregated outcome~\citep{herbinger2022repid}, a phenomenon called
-\emph{aggregation bias}~\citep{mehrabi2021survey}. Therefore, global
-methods should quantify not only the aggregated (averaged) effect, but
-also the level of heterogeneity of instance-level explanations. The
-same holds for Feature Effect (FE)~\citep{Gromping2020MAEP}, a class
-of global XAI methods that isolate the impact of a single feature on
-the output.\footnote{FE methods also isolate the combined effect of a
-  pair of features to the output. Combinations of more than two
-  features are uncommon, since they are difficult to estimate and
-  visualize.} For example, a feature with zero average effect may
-indicate no effect on the output or a highly positive effect in some
-cases and a highly negative effect in others. The ambiguity about the
-correct interpretation can be resolved, if the FE method informs about
-the heterogeneity of instance-level explanations, a quantity we refer
-to as the \emph{uncertainty of the effect}.
+A popular class of global XAI are feature effect (FE)
+methods~\citep{Gromping2020MAEP} that quantify the average (over all
+instances) effect of a single feature on the output\footnote{FE
+  methods also isolate the combined effect of a pair of features to
+  the output. Combinations of more than two features are uncommon,
+  since they are difficult to estimate and visualize.}. There are
+three widely-used FE methods: \emph{Partial Dependence Plots}
+(PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
+(MP)\citep{apley2020visualizing} and \emph{Accumulated Local Effects}
+(ALE)\citep{apley2020visualizing}. ALE is established as the
+state-of-the-art in quantifying the average effect, since PDP and MP
+have been criticized~\citep{Gromping2020MAEP} of providing misleading explanations in case of  correlated input features.
 
-There are three widely-used FE methods: \emph{Partial Dependence
-  Plots} (PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
-(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
-(ALE)\citep{apley2020visualizing}. ALE has been established as the
-state-of-the-art for measuring the average effect, since PDP and MP
-have been criticized~\citep{Gromping2020MAEP} to compute erroneous
-effects when the input features are (highly) correlated. However, ALE
-faces two limitations, the first concerns ALE definition and the
-second ALE approximation. Regarding the definition, ALE does not
-inform the user about the uncertainty. In contrast, in PDP the
-heterogeneity of instance-level effects can be identified by exploring
-the Individual Conditional Expectations
-(ICE)~\citep{goldstein2015peeking}. Regarding the approximation, ALE
-requires partitioning the feature domain into non-overlapping
-intervals (bin-splitting) and estimate the effect inside each interval
-(bin-effect) from the training set instances that fall inside. Finding
-an appropriate sequence of bins is of particular importance, since
-ALE's interpretation is meaningful only inside each
-bin~\citep{molnar2022}. However, this crucial step has not raised the
-appropriate attention. With the current
-approaches~\citep{apley2020visualizing, gkolemis22}, the domain is
-split \emph{blindly}, i.e.~without considering the characteristics of
-the instance-level effects, in \(K\) equisized bins.
+When complex interactions between features exist, the instance-level
+(local) feature effects may exhibit significant deviation, aka heterogeneity, from the aggregated (global)
+explanation, a phenomenon known as aggregation
+  bias~\citep{mehrabi2021survey}. We call this deviation \emph{the uncertainty of the global effect}. Aggregation bias leads to
+ambiguities; for example, a feature with zero average effect may
+indicate (a) no effect on the output or (b) an effect that is highly
+positive for some instances and highly negative for some others. As a
+result, there is a need for FE methods to quantify the deviation (uncertainty) of
+instance-level explanations, in addition to the average effect. In the
+case of PDP, such deviation is visualized via the Individual
+Conditional Explanations (ICE)~\citep{goldstein2015peeking}. ICE
+plots, however, have the same limitations as PDP in case of correlated
+features. No method to quantify the deviation of instance-level effects has been proposed for ALE.
 
-In this paper, we present Uncertainty-aware Accumulated Local Effects
-(UALE), a method that visualizes in a single plot both the average
-feature effect and the uncertainty, i.e.~the level of heterogeneity of
-instance-level effects. The average effect is defined as in ALE and
-the uncertainty inherits ALE's advantages, i.e.~takes into account the
-correlations between features. We also propose an accurate UALE
-approximation. We formally prove that for an unbiased estimation of
-the uncertainty, the partitioning of the feature domain must respect
-specific conditions. Using that, we set-up an optimization problem
-where we search for a partitioning, a sequence of \emph{variable-size}
-intervals, (a) with enough instances (accurate estimation) and (b)
-that respect the above conditions (unbiased estimation). Finally, we
-propose a method for automatically finding the optimal
-partitioning. We apply our method in multiple synthetic and real world
-datasets for evaluating its performance. The contributions of this
+% Such ambiguities could be resolved if the heterogeneity of instance-level feature effects is provided in addition to the average effect. 
+% In this paper, we measure the heterogeneity as the standard deviation of the instance-level effects and we refer to it using as the \emph{uncertainty of the effect}.
+
+%Therefore, global methods should quantify not only the aggregated (averaged) effect, but also the level of heterogeneity of instance-level explanations.
+
+%There are three widely-used FE methods: \emph{Partial Dependence Plots} (PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
+%(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
+%(ALE)\citep{apley2020visualizing}. ALE has been established as the
+%state-of-the-art for measuring the average effect, since PDP and MP
+%have been criticized~\citep{Gromping2020MAEP} to compute erroneous
+%effects when the input features are (highly) correlated.
+
+In this work, we propose UALE, a global feature-effect method that
+extends ALE for modeling the uncertainty of the feature effect.
+Our method handles well cases with correlated features. 
+As in ALE, we provide an interval-based formulation of UALE. 
+This formulation partitions the feature domain into non-overlapping 
+intervals (bin-splitting) and defines the average effect (bin-effect) and its uncertainty (bin-uncertainty). We exploit this formulation to provide an UALE approximation. 
+As we formally prove, a fixed equi-width partitioning that does not consider
+the characteristics of the instance-level effects, which is the case in ALE, 
+may lead to an erroneous (biased) estimation of the uncertainty. 
+Therefore, we propose a variable-width partitioning that leads to an unbiased approximation.
+
+% %However, ALE faces two limitations, the first concerns ALE definition and the second ALE approximation. Regarding the definition, ALE does not inform the user about the uncertainty. 
+% \color{blue}\todo{This should go to related work}In contrast, in PDP the
+% heterogeneity of instance-level effects can be identified by exploring
+% the Individual Conditional Expectations
+% (ICE)~\citep{goldstein2015peeking}. \color{black}
+% %Regarding the approximation, ALE requires partitioning the feature domain into non-overlapping intervals (bin-splitting) and estimate the effect inside each interval (bin-effect) from the training set instances that fall inside.
+% \color{blue}\todo{to be moved in related work}
+% Finding
+% an appropriate sequence of bins is of particular importance, since
+% ALE's interpretation is meaningful only inside each
+% bin~\citep{molnar2022}. However, this crucial step has not raised the
+% appropriate attention. With the current
+% approaches~\citep{apley2020visualizing, gkolemis22}, the domain is
+% split \emph{blindly}, i.e.~without considering the characteristics of
+% the instance-level effects, in \(K\) equisized bins.\color{black}
+
+%In this paper, we present Uncertainty-aware Accumulated Local Effects(UALE), a method that visualizes in a single plot both the av erage feature effect and the uncertainty, i.e.~the level of heterogeneity of instance-level effects.
+The average effect is defined as in ALE and the uncertainty inherits ALE's advantages, i.e.~takes into account the correlations between features.
+We also propose an accurate UALE approximation.
+We formally prove that for an unbiased estimation of the uncertainty, the partitioning of the feature domain must respect specific conditions.
+Using that, we set-up an optimization problem where we search for a partitioning, a sequence of \emph{variable-size}intervals, (a) with enough instances (accurate estimation) and (b) that respect the above conditions (unbiased estimation).
+Finally, we propose a method for automatically finding the optimal partitioning.
+We apply our method in multiple synthetic and real world datasets for evaluating its performance.
+The contributions of this
 paper are:
-
 \begin{itemize}
-\item UALE, a feature effect method that quantifies both the average
-  effect and the uncertainty, i.e.,~the heterogeneity of instance level
-  effects.
-\item A formal proof of the conditions the feature domain partitioning
-  must fulfill for an unbiased approximation of the uncertainty.
-\item An accurate UALE approximation by automatically partitioning
-  into variable-size intervals that respect the above conditions.
+\item A global feature effect method (UALE) that quantifies both the
+  average effect and its uncertainty, i.e.,~the deviation of
+  instance-level effects.
+\item An unbiased approximation of the uncertainty through a dynamic
+  variable-size bin partitioning of the feature domain.
+%A formal proof of the conditions the feature domain partitioning must fulfill for an unbiased approximation of the uncertainty.
+\item The evaluation of UALE on both synthetic and real datasets.
+%An accurate UALE approximation by automatically partitioning into variable-size intervals that respect the above conditions.
 \end{itemize}
 
-The implementation of our method and the code for
-reproducing all the experiments is provided in the submission and will
-become publicly available upon acceptance.
+The implementation of our method and the code for reproducing all the
+experiments is provided along with the manuscript and will become
+publicly available upon acceptance.
 
@@
 
-We here review the existing methods for modeling the average effect
-and the heterogeneity of instance-level effects and we focus on ALE
-definition~\citep{apley2020visualizing} and
-approximation~\citep{apley2020visualizing, gkolemis22}.
-
-\paragraph{Notation.} We refer to random variables (rv) using
-uppercase \( X \), to simple variables with plain lowercase \( x \)
-and to vectors with bold \( \xb \). Often, we partition the input
-vector \(\xb \in \mathbb{R}^D\) to the feature of interest
-\(x_s \in \mathbb{R} \) and the rest of the features
-\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
-\((x_s, \mathbf{x}_c)\), but we it refers the vector
-\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
-corresponding rv as \(\mathbf{X} = (X_s, \mathbf{X}_c)\). When we
-refer to an instance of the training set, we use
-\(\xb^i= (x_s^i, \xc^i) \).
-
-% The black-box function is
-% \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the FE of the \(s\)-th
-% feature is \(f^{\mathtt{<method>}}(x_s)\), with \(\mathtt{<method>}\)
-% indicating the particular method in use.\footnote{An extensive list of
-%   all symbols used in the paper is provided at the Appendix.}
+In the rest of the paper, we use following notation. Let \(\mathcal{X} \in \mathbb{R}^d\) be the \(d\)-dimensional feature
+space, \(\mathcal{Y} \in \mathbb{R}\) the target space and
+\(f(\cdot) : \mathcal{X} \rightarrow \mathcal{Y}\) the black-box
+function. We use index \(s \in \{1, \ldots, d\}\) for the feature of
+interest and \(c = \{1, \ldots, d\} - s\) for the set with all other
+indexes. For convenience, we denote the feature vector
+\(\xb = (x_1, \cdots , x_s, \cdots, x_D)\) with \((x_s, \xc)\) and the
+corresponding random variables
+\(X = (X_1, \cdots , X_s, \cdots, X_D)\) with \((X_s, X_c)\). The
+training set is \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) sampled
+i.i.d. from the distribution \(\mathbb{P}_{X,Y}\). Finally, we use
+\(f^{\mathtt{<method>}}(x_s)\) for denoting the \(s\)-th FE, where
+\(\mathtt{<method>}\) indicates the particular method in use, for
+example ALE. An extensive list with all symbols used in this paper is
+provided at Appendix~\ref{sec:list-with-symbols}.
 
@@
 \label{sec:feat-effect-meth}
+The three well-known feature effect methods are: \emph{Partial
+  Dependence Plots} (PDP), \emph{Marginal Plots} (MP) and
+\emph{Accumulated Local Effects} (ALE).  PDP formulates the global FE
+as an expectation over the distribution of \(X_c\), i.e.,
+\(f^{\mathtt{PDP}}(x_s) = \mathbb{E}_{X_c}[f(x_s,X_c)]\), whereas MP
+as an expectation over the distribution of \(X_c|X_s\), i.e.,
+\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{X_c|x_s}[f(x_s, X_c)]\). Both
+methods suffer from misestimations in case of correlated features. PDP
+integrates over unrealistic instances and MP computes aggregated
+effects, i.e., imputes the combined effect of sets of features to a
+single feature~\citep{apley2020visualizing}.
 
-The three well-known feature effect methods are: PDP, MP and ALE. PDP
-formulates the FE of the \(s\)-th feature as an expectation over the
-marginal distribution \(\mathbf{X}_c\), i.e.,
-\(f^{\mathtt{PDP}}(x_s) =
-\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
-formulates it as an expectation over the conditional
-\(\mathbf{X}_c|X_s\), i.e.,
-\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|x_s}[f(x_s,
-\mathbf{X}_c)]\). ALE defines the local effect at \(x_s = z\) as
-\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). The local
-explanations are then weighted by the conditional distribution
-\(p(\xc|x_s=z)\) to compute the averaged effect \(\mu(z)\). ALE is the
-accumulation of the averaged local effects:
+ALE overcomes these problems. Specifically, ALE defines the local
+effect of the \(s\)-th feature at a specific point \((z, \xc)\) of the
+input space \(\mathcal{X}\) with the partial derivative
+\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
+local explanations at \(z\) are then weighted by the conditional
+distribution \(p(\xc|z)\) and are averaged, to produce the averaged
+effect \(\mu(z)\). ALE plot is the accumulation of the averaged local
+effects:
 
@@
   \label{eq:ALE}
-  f^{\mathtt{ALE}}(x_s) = \int_{x_{s,min}}^{x_s} \underbrace{\mathbb{E}_{\Xcb|X_s=z}\left [f^s (z, \Xcb)\right ]}_{\mu(z)} \partial z
+  f^{\mathtt{ALE}}(x_s) = \int_{x_{s,min}}^{x_s} \underbrace{\mathbb{E}_{X_c|X_s=z}\left [f^s (z, X_c)\right ]}_{\mu(z)} \partial z
 \end{equation}
-
-Eq.~(\ref{eq:ALE}) has important advantages compared to PDP and MP in
-cases of correlated features. In these cases, PDP integrates over
-unrealistic instances due to the use of the marginal distribution
-\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
-the combined effect of sets of features to a single feature. ALE
-resolves both issues.
+%
+where \(x_{s,min}\) is the minimum value of the \(s\)-th feature.
 
@@
 
-FE methods show what is the average (global) effect on the output, if
-the value of a specific feature is increased/decreased. For completing
-the explanation about a feature, we need to also know to what extent
-the local effects deviate from the global explanation, i.e.~to
-quantify the uncertainty of the average effect.
+The global effect is computed as an expectation over local
+(instance-level) effects. In addition to this global effect, it is
+important to know to what extent the local effects deviate from the
+global explanation.
 
-ICE and d-ICE\citep{goldstein2015peeking} provide a set of curves that
-are plot on top-of PDP. Both methods produce one curve for each
-instance of the training set;
-\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) for ICE and
-\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
-\xc^i)\) for d-ICE. The user then visually observes if the curves are
-homogeneous and to what extent they deviate from the PDP. Some methods
-try to automate the aforementioned visual exploration, by grouping
-(d-)ICE plots into clusters~\citep{molnar2020model,
-  herbinger2022repid, britton2019vine}. Other approaches, like
-H-Statistic\citep{friedman2008predictive}, Greenwell's interaction
-index\citep{greenwell2018simple} or SHAP interaction
-values\citep{lundberg2018consistent} inform indirectly about
-heterogeneous effects, quantifying interactions between the input
-features. A strong interaction index is an indicator for the existence
-of heterogeneous effects.
+ICE plots and similar methods (e.g., d-ICE
+plots~\citep{goldstein2015peeking}) provide a set of curves
+illustrated on top-of PDP. Each curve corresponds to one instance of
+the dataset, \(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\). The user
+then visually observes if the curves are homogeneous, i.e., all
+instances have similar effect plots, and to what extent they deviate
+from the PDP. There are methods try to automate the aforementioned
+visual exploration, by grouping ICE plots into
+clusters~\citep{molnar2020model, herbinger2022repid,
+  britton2019vine}. Unfortunately, these methods are subject to the
+failure modes of PDPs in cases of correlated
+features~\citep{baniecki2021fooling}, as we also confirm in our
+experimental analysis (c.f., Section~\ref{sec:simulation-examples-1}).
 
-The aforementioned approaches are under two limitations. They either
-do not quantify the uncertainty of the FE directly or they are based
-on PDPs, and, therefore, they are subject to the failure modes of PDPs
-in cases of correlated features\citep{baniecki2021fooling}, as we will
-show in-depth in Section~\ref{sec:simulation-examples-1}. To the best
-of our knowledge, no work exist so far that quantifies the
-heterogeneous effects based on the formulation of ALE.
+Other approaches, like H-Statistic \citep{friedman2008predictive},
+Greenwell's interaction index \citep{greenwell2018simple} or SHAP
+interaction values \citep{lundberg2018consistent} quantify
+interactions between the input features. A strong interaction is
+  an implicit indication of the existence of heterogeneous effects. These
+  methods, however, do not quantify directly the level of heterogeneity.
+
+To the best of our knowledge, no work exist so far that quantifies the
+heterogeneous effects (uncertainty) based on the formulation of ALE.
 
@@
 
-In ML scenarios, ALE is estimated from the instances of the training
-set.~\citep{apley2020visualizing} proposed dividing the feature domain
-in \(K\) equisized bins and estimating the local effects in each bin
-by evaluating the black box-function at the bin limits:
+ALE is estimated from the available dataset instances. \citep{apley2020visualizing} proposed to divide the feature
+domain in \(K\) bins of equal size and to estimate the local effects
+in each bin by evaluating the model \(f\) at the bin limits:
+
 \begin{equation}
@@
 \end{equation}
-We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
+%
+where \(k_x\) the index of the bin that \(x_s\) belongs to,
 i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
@@
 \( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
-\). Afterwards,~\citep{gkolemis22} proposed the Differential ALE
-(DALE) that uses the partial derivatives:
-\begin{equation}
-  \label{eq:DALE_accumulated_mean_est}
+\). \citep{gkolemis22} proposed the Differential ALE (DALE) that
+computes the local effects on the training instances using
+auto-differentiation:
+\begin{equation}  \label{eq:DALE_accumulated_mean_est}
   \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
@@
 %
-Their method has the advantages of remaining on-distribution when
-bins become wide and, most importantly, allows the recomputation of
-the accumulated effect with different bin-splitting with near-zero
-computational cost.
+Their method has significant computational advantages and allows the
+recomputation of the accumulated effect with different bin-splitting
+with near-zero additional computational cost. Moreover, by avoiding
+the use of artificial samples at the bin limits DALE allows the use of
+wider bins without out-of-distribution sampling. \todo{Maybe connect with our work}. In both cases, the
+approximations partition the feature domain in \(K\) equally-sized
+bins, without considering the underlying local effects.
 
-Both approximations ask from the user to blindly decide the bin
-density for splitting the feature axis into \(K\) equisized bins, an
-approach that is error prone. Setting \(K\) to a small value may hide
-fine-grain effects due to large bins while setting \(K\) to a high
-value leads to poor bin-effect estimations due to few samples per
-bin. In general, the user may receive contradictory explanations for
-different \(K\) without a clue for which one to trust.
+% as
+% input and perform an equally-sized bin splitting of the attribute
+% domain. The choice of \(K\) is therefore critical and introduces
+% significant limitations. Specifically, setting \(K\) to a small value
+% may hide fine-grain effects due to large bins while setting \(K\) to a
+% high value leads to poor bin-effect estimations due to few samples per
+% bin. In general, the user may receive contradictory explanations for
+% different \(K\) values without a clue for which one to trust.
 
@@
 
-UALE extends ALE for quantifying the level of heterogeneous effects
-(uncertainty) and automates the bin-splitting. At
-Section~\ref{sec:UALE-definition-1} we define UALE, at
-Section~\ref{sec:UALE-definition-2} we reformulate it using an
-interval-based formulation, at Section~\ref{sec:UALE-approximation} we
-provide UALE approximation and at Section~\ref{sec:bin-spliting} we
-define and solve the problem of optimal bin-splitting.
+In this Section~\ref{sec:UALE-definition-1}, we define UALE. Then, we
+provide an interval-based formulation that partitions the feature
+domain into non-overlapping intervals (bin-splitting) and defines the
+bin-effect and the bin-uncertainty
+(Section~\ref{sec:UALE-definition-2}). Then, we present the UALE
+approximation (Section~\ref{sec:UALE-approximation}), formally proving
+the bin-splitting requirements for an unbiased estimation of the
+uncertainty. Based on this, we present an optimization method to
+determine an appropriate variable-width partitioning
+(Section~\ref{sec:bin-spliting}).
 
-\subsection{UALE: ALE With Uncertainty Quantification}
+\subsection{UALE Definition And Interval-Based Formulation}
 \label{sec:UALE-definition-1}
 
-UALE extends ALE definition of Eq.~(\ref{eq:ALE}) with a component for
-quantifying the uncertainty. We denote as \(\mathcal{H}(z)\) the
-uncertainty of the local effects at a specific point \(x_s=z\) and we
-quantify it with the standard deviation of the local explanations
-\(\mathcal{H}(z) = \sigma(z)\), where:
+We quantify the uncertainty of the local effects at \(x_s=z\) with the
+standard deviation of the local explanations, \(\sigma(z)\), where:
 
@@
   \label{eq:ALE_var}
-  \sigma^2(z) = \mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] 
+  \sigma^2(z) = \mathbb{E}_{X_c|X_s=z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] 
 \end{equation}
 \noindent
-The uncertainty emerges from the natural characteristics of the
-experiment, i.e.,~the feature correlations existent in the data
-generating distribution and the implicit interactions of the black-box
-function. We also define the accumulated uncertainty at \(x_s\), as
-the accumulation of the standard deviation of the local effects along
-the axis:
+The uncertainty emerges from (a) the feature correlations and (b) the
+implicit feature interactions of the black-box function. We also
+define the accumulated uncertainty at \(x_s\), as the accumulation of
+the standard deviation of the local effects:
+
 \begin{equation}
@@
 \noindent
-UALE formulates the effect at a specific point \(x_s\) with a tuple
-that consists of the average effect and the uncertainty
-\((\mu(z), \sigma(z))\) and visualizes it with a continuous curve and
-a confidence region
-\(f^{\mathtt{UALE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
-f^{\mathtt{ALE}}_{\sigma}(x_s)\).
 
-\subsection{Interval-Based Formulation and Approximation}
-\label{sec:UALE-definition-2}
+UALE defines the effect as a pair that consists of the average effect
+and the uncertainty \((\mu(z), \sigma(z))\). For visualizing it, we
+use the accumulated average effect, as defined in Eq.~(\ref{eq:ALE}),
+and the accumulated uncertainty, as defined in
+Eq.~(\ref{eq:ALE_acc_unc}). Figure~\ref{fig:UALE-figure} illustrates
+an example \todo{Describe after putting the figure}.
 
-In real ML scenarios, all estimations are based on the instances of
-the training set. Therefore, we cannot estimate \(\mu(z), \sigma(z)\)
-at the granularity of a point, because the probability of observing a
-sample inside the region \([x_s - h, x_s + h]\) tends to zero, when
-\(h \to 0\). We define the regional-effect \(\mu(z_1, z_2)\) and the
-regional-uncertainty \(\mathcal{H}(z_1, z_2) = \sigma(z_1, z_2)\), as
-the summary statistics that characterize the effect and the
-uncertainty inside the interval \([z_1, z_2)\):
+\begin{figure}
+  \centering
+  \caption{UALE-concept-figure}
+  \label{fig:UALE-figure}
+\end{figure}
+
+For the interval-based formulation, we define the bin-effect
+\(\mu(z_1, z_2)\) and the corresponding bin-uncertainty
+\(\sigma(z_1, z_2)\) as:
 
@@
 %
-Intuitively, the regional-effect and the regional-uncertainty quantify
-the expected average effect and the expected uncertainty if we
-randomly draw a point \(z^*\) given a uniform distribution
-\(z^* \sim \mathcal{U}(z_1, z_2)\). If we also define as
-\(\mathcal{Z}\) the sequence of \(K+1\) points that partition the
-domain of the \(s\)-th feature into \(K\) variable-size intervals,
-i.e.  \(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can redefine UALE
-using an interval-based formulation
-\(\tilde{f}^{\mathtt{UALE}}_{\mathcal{Z}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s)
-\pm \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s)\), where:
-
+Intuitively, if we randomly draw a point \(z^*\) given a uniform
+distribution \(z^* \sim \mathcal{U}(z_1, z_2)\), the bin-effect
+(Eq.~(\ref{eq:mu_bin})) and the bin-uncertainty
+(Eq.~(\ref{eq:var_bin})) are the expected average effect and the
+expected uncertainty, respectively. Also, if we define \(\mathcal{Z}\)
+to be the sequence of \(K+1\) points that partition the domain of the
+\(s\)-th feature into \(K\) variable-size intervals, i.e.
+\(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can reformulate UALE to:
 \begin{equation}
@@
 %
+where \(k_x\) is the index of the bin such that
+\( z_{k_x - 1} \leq x_s < z_{k_x}\).
 
-\subsection{Interval-Based Approximation}
-\label{sec:UALE-approximation}
+\subsection{UALE Interval-Based Approximation}
+~\label{sec:UALE-approximation}
 
-For approximating the mean effect (Eq.~\eqref{eq:mu_bin}) and the
-uncertainty (Eq.~\eqref{eq:var_bin}) we use the set \(\mathcal{S}_k\)
-of dataset instances with the \(s\)-th feature lying inside the
-interval
-\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \} \):
+For approximating the bin-effect and the bin-uncertainty, we use the
+set \(\mathcal{S}_k\) of dataset instances with the \(s\)-th feature
+lying inside the \(k\)-th bin, i.e.,
+\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \}
+\). The bin-effect is approximated with,
 
 \begin{equation}
-  \label{eq:2}
+  \label{eq:mu_bin_approx}
   \hat{\mu}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
@@
 \end{equation}
+
+%
+which, as shown by\citep{gkolemis22}, is an unbiased estimator of
+Eq.~(\ref{eq:mu_bin}) under the assumption that the points are
+uniformly distributed inside the intervals. The approximation for the
+uncertainty (Eq.~(\ref{eq:var_bin})) from the available instances is
+defined as follows:
+
 %
 \begin{equation}
-  \label{eq:3}
+  \label{eq:var_bin_approx}
   \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
@@
 \end{equation}
+%
+In the Appendix~\ref{sec:proof-1}, we show that
+\(\hat{\sigma}^2(z_{k-1}, z_k)\) is an unbiased estimator of
+\(\sigma^2_*(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{\xc|x_s=z}
+  \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
+  z_1} \)
+%
+which, in the general case, overestimates
+\(\sigma^2(z_1, z_2)\), as shown in Theorem~\ref{sec:theorem-1}.
 
-Eq.\eqref{eq:2} is an unbiased estimator of Eq.~(\ref{eq:mu_bin})
-under the assumption that the points are uniformly distributed inside
-the intervals. The same does not hold in the general case for
-Eq.\eqref{eq:3}. In the general case, \(\hat{\sigma}^2(z_{k-1}, z_k)\)
-is an unbiased estimator of the \textit{observable} variance
-\(\sigma^2_{obs}(z_1, z_2) = \frac{\int_{z_1}^{z_2}
-    \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2
-    \right] \partial z}{z_2 - z_1}\), which as we prove in Theorem 1
-is an overestimation of \(\sigma^2(z_1, z_2)\).
-
-\paragraph{Theorem 1.}
+\begin{theorem}
   \label{sec:theorem-1}
-
-\textit{If we define the residual \(\rho(z)\) as the difference
-  between the expected effect at \(x_s\) and the regional effect, i.e
-  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then the regional variance
-  \(\sigma^2(z_1, z_2)\) equals to:}
-
+  If we define (a) the residual \(\rho(z)\) as the difference between
+  the expected effect at \(x_s\) and the regional effect, i.e
+  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), and (b) \(\mathcal{E}(z_1, z_2)\) as the
+  mean squared residual of the bin, i.e.
+  \(\mathcal{E}(z_1, z_2) = \frac{\int_{z_1}^{z_2}\rho^2(z) \partial
+    z}{z_2 - z_1}\), then it holds
 \begin{equation}
     \label{eq:bin-uncertainty-proof}
- \sigma_{obs}^2(z_1, z_2) = \sigma^2(z_1, z_2) + \frac{\int_{z_1}^{z_2}\rho^2(z) \partial z}{z_2 - z_1}
+ \sigma_*^2(z_1, z_2) = \sigma^2(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
 \end{equation}
+  \end{theorem}
+
+\begin{proof}
+The proof is at Appendix~\ref{sec:proof-of-3-1}.
+\end{proof}
+
 \noindent
-Theorem 1 reveals an important connection between the ground-truth and
-the observable uncertainty inside a region. For convenience, we denote
-as
-\(\mathcal{E}^2(z_1, z_2) = \frac{\int_{z_1}^{z_2}\rho^2(z) \partial
-  z}{z_2 - z_1}\) the error term that models the expected square
-residual inside the interval and as
-\(\mathcal{H}_{obs}(z_1, z_2) := \sigma_{obs}(z_1, z_2)\) the
-observable uncertainty. It holds that the observable uncertainty is an
-overestimation of the correct uncertainty
 
-\begin{equation}
-  \label{eq:uncertainty-bin}
-  \mathcal{H}^2_{obs}(z_1, z_2) = \mathcal{H}^2(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
-\end{equation}
-%
-and, therefore, the estimation is unbiased only in case
-\(\mathcal{E}^2(z_1, z_2) = 0\).
+Based on Theorem\ref{sec:theorem-1}, the estimation is unbiased only
+when \(\mathcal{E}^2(z_1, z_2) = 0\).
 
-\subsection{Bin-Splitting: Finding Regions With Homogeneous Effects}
+\subsection{Bin-Splitting}
 \label{sec:bin-spliting}
 
-In this section we formulate bin-splitting as an unsupervised
-clustering problem, searching for a solution that compromises two
-contradictory objectives. On the one hand, we want to minimize the
-error term \(\mathcal{E}\) for approaching an unbiased estimation and
-on the other hand, we want wide-enough bins for including a fair
-population of samples for a robust estimation of
-\(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\). We set-up the
-following optimization problem:
+The quality of interval-based approximation is affected by (a) the
+population of samples in each bin and (b) the error term
+\(\mathcal{E}(z_1, z_2)\) of each bin. On the one hand, we favor a
+partitioning of the feature domain in large-bins to allow a robust
+estimation of \(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\) from a
+sufficient population of samples. On the other hand, we should
+minimize the aggregation of bin errors, i.e.,
+\( \mathcal{E}^2_{\mathcal{Z}} = \sum_{k=1}^K\mathcal{E}^2(z_1, z_2)
+\Delta z_k\), where \(\mathcal{Z} = \{z_0, \cdots, z_K\}\) and
+\(\Delta z_k = z_k - z_{k-1}\). For convenience, we refer to
+\(\mathcal{E}^2_{\mathcal{Z}}\) as aggregated error. We search for a
+partitioning \(\mathcal{Z}_*\) that balances this trade-off.
+
+Due to the Corollary~\ref{sec:coroll}, minimizing
+\( \mathcal{E}^2_{\mathcal{Z}} \) is equivalent to minimizing
+\(\sigma_*^2(z_{k-1}, z_k)\Delta z_k\), which can be directly
+estimated by \(\hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \).
+
+\begin{corollary}
+  \label{sec:coroll}
+  The minimization of the aggregation of bin variances is equivalent
+  to the minimization of the aggregated error:
+  \begin{equation}
+  \begin{gathered}
+    \label{eq:2}
+    \argmin_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \mathcal{E}^2_{\mathcal{Z}}  = \argmin_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \sum_{k=1}^K\sigma_*^2(z_1, z_2) \Delta z_k 
+  \end{gathered}
+  \end{equation}
+\end{corollary}
+
+\begin{proof}
+~\label{sec:coroll-1}
+  The proof is at Appendix~\ref{sec:proof-of-coroll}
+  \end{proof}
+
+Based on the above, we set-up the following optimization problem:
 
@@
 %
-We search for the sequence of intervals \(z_0, \ldots, z_K\) that
-minimizes the sum of the bin costs. The cost of each bin is the
-approximated variance \(\hat{\sigma}^2_k\) scaled by the bin length
-\(\Delta z_k\) and discounted by the term \(\tau_k\). The term
-\(\tau_K\) favors the selection of a bigger bin in case it has similar
-variance with the aggregate variance of many smaller ones. The
-constraint of at least \(N_{\mathtt{PPB}}\) points per bin sets the
-lowest-limit for a \textit{robust} estimation. The user can choose to
-what extent they favor the creation of wide bins through the discount
-parameter by setting the parameter \(\alpha\) and where they set the
-minimum for robust approximation with the parameter
-\(N_{\mathtt{PPB}}\). For providing a rough idea, in our experiments
-we set \(\alpha = 0.2\) which means that the discount ranges between
-\([0\%, 20\%]\) and \(N_{\mathtt{PPB}} = \frac{N}{20}\). It is also
-important to clarify that by minimizing
-\(\hat{\sigma}^2_k \approx \mathcal{H}_{obs}^2\), we actually minimize
-the squared error term \(\mathcal{E}^2\), since the term
-\(\mathcal{H}^2_{true}\) is independent of the bin splitting.
+The objective \(\mathcal{L}\) searches for the partitioning
+\(\mathcal{Z}_* = \{ z_0, \ldots, z_K \} \) that minimizes the
+aggregated error \(\mathcal{E}_{\mathcal{Z}}^2\), while the term
+\(\tau_K\) favors the partitioning that contains bigger bins, in case
+many partitionings have similar aggregated error. The constraint of at
+least \(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
+\textit{robust} estimation. The user can choose to what extent they
+favor the creation of wide bins through (a) the parameter \(\alpha\)
+that controls the discount \(\tau_k\) and (b) the parameter
+\(N_{\mathtt{PPB}}\) that sets the minimum population per bin. For
+providing a rough idea, in our experiments we set \(\alpha = 0.2\)
+which means that the discount ranges between \([0\%, 20\%]\) and
+\(N_{\mathtt{PPB}} = \frac{N}{20}\).
 
@@
 \(\mathcal{L}^{\rho}_{\mathtt{DP|K}} =
-\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
-  \mathcal{Z}^{\mathtt{DP|K}}} \mathcal{\Epsilon}(z_{k-1}, z_k) \).
+\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in \mathcal{Z}^{\mathtt{DP|K}}} \mathcal{E}(z_{k-1}, z_k) \).
 
@@
 
+\appendix
+
+\section{List with Symbols}
+\label{sec:list-with-symbols}
+
+\section{Show that is unbiased estimator}
+\label{sec:proof-1}
+
+\section{Proof of Theorem~\ref{sec:theorem-1}}
+\label{sec:proof-of-3-1}
+
+\section{Proof of Corollary~\ref{sec:coroll-1}}
+\label{sec:proof-of-coroll}
+
 \end{document}

### 2022-10-12 AISTATS2023/paper.tex
Updates from Overleaf

@@
 
-In this Section~\ref{sec:UALE-definition-1}, we define UALE. Then, we
+In Section~\ref{sec:UALE-definition-1} we define UALE and we
 provide an interval-based formulation that partitions the feature
 domain into non-overlapping intervals (bin-splitting) and defines the
-bin-effect and the bin-uncertainty
-(Section~\ref{sec:UALE-definition-2}). Then, we present the UALE
+bin-effect and the bin-uncertainty. Then, we present UALE
 approximation (Section~\ref{sec:UALE-approximation}), formally proving

### 2022-10-12 AISTATS2023/paper.tex
latest

@@
 \begin{document}
-
-% If your paper is accepted and the title of your paper is very long,
-% the style will print as headings an error message. Use the following
-% command to supply a shorter title of your paper so that it can be
-% used as headings.
-%
-%\runningtitle{I use this title instead because the last one was very long}
-
-% If your paper is accepted and the number of authors is large, the
-% style will print as headings an error message. Use the following
-% command to supply a shorter version of the authors names so that
-% they can be used as headings (for example, use only the surnames)
-%
-%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}
 
@@
 
-% Such ambiguities could be resolved if the heterogeneity of instance-level feature effects is provided in addition to the average effect. 
-% In this paper, we measure the heterogeneity as the standard deviation of the instance-level effects and we refer to it using as the \emph{uncertainty of the effect}.
-
-%Therefore, global methods should quantify not only the aggregated (averaged) effect, but also the level of heterogeneity of instance-level explanations.
-
-%There are three widely-used FE methods: \emph{Partial Dependence Plots} (PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
-%(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
-%(ALE)\citep{apley2020visualizing}. ALE has been established as the
-%state-of-the-art for measuring the average effect, since PDP and MP
-%have been criticized~\citep{Gromping2020MAEP} to compute erroneous
-%effects when the input features are (highly) correlated.
-
 In this work, we propose UALE, a global feature-effect method that
@@
 
-% %However, ALE faces two limitations, the first concerns ALE definition and the second ALE approximation. Regarding the definition, ALE does not inform the user about the uncertainty. 
-% \color{blue}\todo{This should go to related work}In contrast, in PDP the
-% heterogeneity of instance-level effects can be identified by exploring
-% the Individual Conditional Expectations
-% (ICE)~\citep{goldstein2015peeking}. \color{black}
-% %Regarding the approximation, ALE requires partitioning the feature domain into non-overlapping intervals (bin-splitting) and estimate the effect inside each interval (bin-effect) from the training set instances that fall inside.
-% \color{blue}\todo{to be moved in related work}
-% Finding
-% an appropriate sequence of bins is of particular importance, since
-% ALE's interpretation is meaningful only inside each
-% bin~\citep{molnar2022}. However, this crucial step has not raised the
-% appropriate attention. With the current
-% approaches~\citep{apley2020visualizing, gkolemis22}, the domain is
-% split \emph{blindly}, i.e.~without considering the characteristics of
-% the instance-level effects, in \(K\) equisized bins.\color{black}
-
-%In this paper, we present Uncertainty-aware Accumulated Local Effects(UALE), a method that visualizes in a single plot both the av erage feature effect and the uncertainty, i.e.~the level of heterogeneity of instance-level effects.
-The average effect is defined as in ALE and the uncertainty inherits ALE's advantages, i.e.~takes into account the correlations between features.
-We also propose an accurate UALE approximation.
-We formally prove that for an unbiased estimation of the uncertainty, the partitioning of the feature domain must respect specific conditions.
-Using that, we set-up an optimization problem where we search for a partitioning, a sequence of \emph{variable-size}intervals, (a) with enough instances (accurate estimation) and (b) that respect the above conditions (unbiased estimation).
-Finally, we propose a method for automatically finding the optimal partitioning.
-We apply our method in multiple synthetic and real world datasets for evaluating its performance.
 The contributions of this
@@
   variable-size bin partitioning of the feature domain.
-%A formal proof of the conditions the feature domain partitioning must fulfill for an unbiased approximation of the uncertainty.
 \item The evaluation of UALE on both synthetic and real datasets.
-%An accurate UALE approximation by automatically partitioning into variable-size intervals that respect the above conditions.
 \end{itemize}
@@
 
-% as
-% input and perform an equally-sized bin splitting of the attribute
-% domain. The choice of \(K\) is therefore critical and introduces
-% significant limitations. Specifically, setting \(K\) to a small value
-% may hide fine-grain effects due to large bins while setting \(K\) to a
-% high value leads to poor bin-effect estimations due to few samples per
-% bin. In general, the user may receive contradictory explanations for
-% different \(K\) values without a clue for which one to trust.
-
 \section{Uncertainty-Aware ALE (UALE)}
@@
 
-In Section~\ref{sec:UALE-definition-1} we define UALE and we
-provide an interval-based formulation that partitions the feature
-domain into non-overlapping intervals (bin-splitting) and defines the
-bin-effect and the bin-uncertainty. Then, we present UALE
-approximation (Section~\ref{sec:UALE-approximation}), formally proving
+We define UALE and provide an interval-based formulation that
+partitions the feature domain into non-overlapping intervals
+(bin-splitting) and defines the bin-effect and the
+bin-uncertainty. Then, we present UALE approximation, formally proving
 the bin-splitting requirements for an unbiased estimation of the
 uncertainty. Based on this, we present an optimization method to
-determine an appropriate variable-width partitioning
-(Section~\ref{sec:bin-spliting}).
+determine an appropriate variable-width partitioning.
 
@@
 For the interval-based formulation, we define the bin-effect
-\(\mu(z_1, z_2)\) and the corresponding bin-uncertainty
-\(\sigma(z_1, z_2)\) as:
+\(\mu(z_1, z_2)\) and the bin-uncertainty \(\sigma(z_1, z_2)\) as:
 
@@
 %
-Intuitively, if we randomly draw a point \(z^*\) given a uniform
+Intuitively, if we randomly draw a point \(z^*\) from a uniform
 distribution \(z^* \sim \mathcal{U}(z_1, z_2)\), the bin-effect
@@
 (Eq.~(\ref{eq:var_bin})) are the expected average effect and the
-expected uncertainty, respectively. Also, if we define \(\mathcal{Z}\)
-to be the sequence of \(K+1\) points that partition the domain of the
-\(s\)-th feature into \(K\) variable-size intervals, i.e.
-\(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can reformulate UALE to:
+expected uncertainty, respectively. Also, if we denote as
+\(\mathcal{Z}\) the sequence of \(K+1\) points that partition the
+domain of the \(s\)-th feature into \(K\) variable-size intervals,
+i.e.  \(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can reformulate UALE
+to:
 \begin{equation}
@@
 where \(k_x\) is the index of the bin such that
-\( z_{k_x - 1} \leq x_s < z_{k_x}\).
+\( z_{k_x - 1} \leq x_s < z_{k_x}\). Eq.~(\ref{eq:ALE_2}) and
+Eq.~(\ref{eq:ALE_accumulated_var}) are the interval-based formulation
+of UALE.
 
@@
 In the Appendix~\ref{sec:proof-1}, we show that
-\(\hat{\sigma}^2(z_{k-1}, z_k)\) is an unbiased estimator of
+\(\hat{\sigma}^2(z_1, z_2)\) is an unbiased estimator of
 \(\sigma^2_*(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{\xc|x_s=z}
   \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
-  z_1} \)
-%
-which, in the general case, overestimates
-\(\sigma^2(z_1, z_2)\), as shown in Theorem~\ref{sec:theorem-1}.
+  z_1} \). In Theorem~\ref{sec:theorem-1} we prove that in the general
+case, \(\sigma_*^2(z_1, z_2) \geq \sigma^2(z_1, z_2)\), so
+\(\hat{\sigma}^2(z_1, z_2)\) is an overestimation of the bin
+uncertainty.
 
@@
   If we define (a) the residual \(\rho(z)\) as the difference between
-  the expected effect at \(x_s\) and the regional effect, i.e
-  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), and (b) \(\mathcal{E}(z_1, z_2)\) as the
-  mean squared residual of the bin, i.e.
+  the expected effect at \(z\) and the bin-effect, i.e
+  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), and (b)
+  \(\mathcal{E}(z_1, z_2)\) as the mean squared residual of the bin,
+  i.e.
   \(\mathcal{E}(z_1, z_2) = \frac{\int_{z_1}^{z_2}\rho^2(z) \partial
@@
 
-Based on Theorem\ref{sec:theorem-1}, the estimation is unbiased only
-when \(\mathcal{E}^2(z_1, z_2) = 0\).
+We refer to \(\mathcal{E}^2(z_1, z_2)\) as bin-error. Based on
+Theorem\ref{sec:theorem-1}, the estimation is unbiased only when
+\(\mathcal{E}^2(z_1, z_2) = 0\).
 
@@
 
-The quality of interval-based approximation is affected by (a) the
-population of samples in each bin and (b) the error term
-\(\mathcal{E}(z_1, z_2)\) of each bin. On the one hand, we favor a
-partitioning of the feature domain in large-bins to allow a robust
-estimation of \(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\) from a
-sufficient population of samples. On the other hand, we should
-minimize the aggregation of bin errors, i.e.,
+The quality of UALE approximation is affected by (a) the population of
+samples in each bin and (b) the error term \(\mathcal{E}(z_1, z_2)\)
+of each bin. On the one hand, we favor a partitioning of the feature
+domain in large-bins to allow a robust estimation of
+\(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\) from a sufficient
+population of samples. On the other hand, we want to minimize the
+cummulative bin-error, i.e.,
 \( \mathcal{E}^2_{\mathcal{Z}} = \sum_{k=1}^K\mathcal{E}^2(z_1, z_2)
@@
 \(\Delta z_k = z_k - z_{k-1}\). For convenience, we refer to
-\(\mathcal{E}^2_{\mathcal{Z}}\) as aggregated error. We search for a
+\(\mathcal{E}^2_{\mathcal{Z}}\) as accumulated error. We search for a
 partitioning \(\mathcal{Z}_*\) that balances this trade-off.
@@
   \label{sec:coroll}
-  The minimization of the aggregation of bin variances is equivalent
-  to the minimization of the aggregated error:
+  A bin-splitting \(\mathcal{Z}\) minimizes the accumulated if and
+  only if it also minimizes the accumulated error.
   \begin{equation}
@@
 %
-The objective \(\mathcal{L}\) searches for the partitioning
-\(\mathcal{Z}_* = \{ z_0, \ldots, z_K \} \) that minimizes the
-aggregated error \(\mathcal{E}_{\mathcal{Z}}^2\), while the term
-\(\tau_K\) favors the partitioning that contains bigger bins, in case
-many partitionings have similar aggregated error. The constraint of at
-least \(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
+The objective \(\mathcal{L}\) searches for a partitioning
+\(\mathcal{Z}_* = \{ z_0, \ldots, z_K \} \) with low aggregated error
+\(\mathcal{E}_{\mathcal{Z}}^2\). In case of many partitionings with
+similar aggregate-error, the term \(\tau_K\) favors the partitioning
+with more points per bin (with wider bins). The constraint of at least
+\(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
 \textit{robust} estimation. The user can choose to what extent they
@@
 which means that the discount ranges between \([0\%, 20\%]\) and
-\(N_{\mathtt{PPB}} = \frac{N}{20}\).
+\(N_{\mathtt{PPB}} = \frac{N}{20}\), where \(N\) is the dataset size.
+
+For solving the optimization problem of~Eq.\ref{eq:opt} we add some
+constraints. First, we set a threshold \(K_{max}\) on the maximum
+number of bins which, in turn, defines the minimum bin-width,
+i.e. \(\Delta x_{min} = \frac{x_{s, max} - x_{s,min}}{K_{max}}\).
+Based on that, we restrict the bin-limits to be multiples of the
+minimum width, i.e.
+\(z_k = k\cdot \Delta x_{min}, \text{where } k \in \{0 , \cdots,
+K_{max}\} \). In this discretized solution space, we find the global
+optimal using dynamic programming.
 
@@
 \label{sec:dynamic-programing}
-
-\todo{Review all the description}
-For achieving a computationally-grounded solution we set a threshold
-\(K_{max}\) on the maximum number of bins which also discretizes the
-solution space. The width of the bin can take discrete values that are
-multiple of the minimum step
-\(u = \frac{x_{s, max} - x_{s, min}}{K_{max}}\). For defining the
-solution, we use two indexes. The index
-\(i \in \{0, \ldots, K_{max}\}\) denotes the point \((z_i)\) and the
-index \(j \in \{0, \ldots, K_{max}\} \) denotes the position of the
-\(j\)-th multiple of the minimum step, i.e., 
-\(x_j = x_{s,min} + j \cdot u\). The recursive cost function
-\(T(i,j)\) is the cost of setting \(z_i=x_j\):
-\begin{equation}
-  \label{eq:recursive_cost}
-  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
-\end{equation}
-%
-where \(\mathcal{T}(0,j)\) equals zero if \(j=0\) and \(\infty\) in
-any other case. \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin
-with limits \([x_l, x_j)]\):
-
-\begin{equation}
-  \label{eq:cost_step}
-  \mathcal{B}(x_l, x_j) = \begin{cases}
-                            \infty, & \text{if $x_j > x_l$ or \(|\mathcal{S}_{(x_j, x_l)}| < N\)}\\
-                            0, & \text{if $x_j = x_l$}\\
-                            \hat{\sigma}^2(x_j, x_l), &\text{if $x_j \leq x_l$}
-  \end{cases}
-\end{equation}
-
-The optimal solution is given by solving
-\(\mathcal{L} = \mathcal{T}(K_{max}, K_{max})\) and keeping track of the sequence of
-steps.
-
-\noindent
-
-\begin{itemize}
-\item Discuss more aspects (e.g. Computational complexity)
-\end{itemize}
 
@@
 
+\section{Dynamic Programming}
+\label{sec:dynamic-programming}
+
+\todo{Review all the description}
+For achieving a computationally-grounded solution we set a threshold
+\(K_{max}\) on the maximum number of bins which also discretizes the
+solution space. The width of the bin can take discrete values that are
+multiple of the minimum step
+\(u = \frac{x_{s, max} - x_{s, min}}{K_{max}}\). For defining the
+solution, we use two indexes. The index
+\(i \in \{0, \ldots, K_{max}\}\) denotes the point \((z_i)\) and the
+index \(j \in \{0, \ldots, K_{max}\} \) denotes the position of the
+\(j\)-th multiple of the minimum step, i.e., 
+\(x_j = x_{s,min} + j \cdot u\). The recursive cost function
+\(T(i,j)\) is the cost of setting \(z_i=x_j\):
+\begin{equation}
+  \label{eq:recursive_cost}
+  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
+\end{equation}
+%
+where \(\mathcal{T}(0,j)\) equals zero if \(j=0\) and \(\infty\) in
+any other case. \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin
+with limits \([x_l, x_j)]\):
+
+\begin{equation}
+  \label{eq:cost_step}
+  \mathcal{B}(x_l, x_j) = \begin{cases}
+                            \infty, & \text{if $x_j > x_l$ or \(|\mathcal{S}_{(x_j, x_l)}| < N\)}\\
+                            0, & \text{if $x_j = x_l$}\\
+                            \hat{\sigma}^2(x_j, x_l), &\text{if $x_j \leq x_l$}
+  \end{cases}
+\end{equation}
+
+The optimal solution is given by solving
+\(\mathcal{L} = \mathcal{T}(K_{max}, K_{max})\) and keeping track of the sequence of
+steps.
+
+\noindent
+
+\begin{itemize}
+\item Discuss more aspects (e.g. Computational complexity)
+\end{itemize}
+
 \end{document}

### 2022-10-12 AISTATS2023/paper.tex
Review simulation experiments

@@
 
-\subsection{UALE Definition And Interval-Based Formulation}
+\subsection{UALE Definition and Interval-Based Formulation}
 \label{sec:UALE-definition-1}
@@
 domain of the \(s\)-th feature into \(K\) variable-size intervals,
-i.e.  \(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can reformulate UALE
-to:
+i.e.  \(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we provide the
+interval-based formulation of UALE:
+
 \begin{equation}
@@
 \( z_{k_x - 1} \leq x_s < z_{k_x}\). Eq.~(\ref{eq:ALE_2}) and
-Eq.~(\ref{eq:ALE_accumulated_var}) are the interval-based formulation
-of UALE.
+Eq.~(\ref{eq:ALE_accumulated_var}) are piecewise-linear approximation
+of Eq.~(\ref{eq:ALE}) and Eq.~(\ref{eq:ALE_acc_unc}).
 
@@
 %
-which, as shown by\citep{gkolemis22}, is an unbiased estimator of
-Eq.~(\ref{eq:mu_bin}) under the assumption that the points are
-uniformly distributed inside the intervals. The approximation for the
-uncertainty (Eq.~(\ref{eq:var_bin})) from the available instances is
-defined as follows:
+which is an unbiased estimator of Eq.~(\ref{eq:mu_bin}) under the
+assumption that the points are uniformly distributed inside the
+interval. The approximation for the uncertainty
+(Eq.~(\ref{eq:var_bin})) from the available instances is defined as
+follows:
 
@@
 \(\hat{\sigma}^2(z_1, z_2)\) is an unbiased estimator of
-\(\sigma^2_*(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{\xc|x_s=z}
-  \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
+\(\sigma^2_*(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{X_c|X_s=z}
+  \left [ (f^s(z, X_c) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
   z_1} \). In Theorem~\ref{sec:theorem-1} we prove that in the general
@@
   the expected effect at \(z\) and the bin-effect, i.e
-  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), and (b)
+  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\) and (b)
   \(\mathcal{E}(z_1, z_2)\) as the mean squared residual of the bin,
@@
 population of samples. On the other hand, we want to minimize the
-cummulative bin-error, i.e.,
+cumulative bin-error, i.e.,
 \( \mathcal{E}^2_{\mathcal{Z}} = \sum_{k=1}^K\mathcal{E}^2(z_1, z_2)
 \Delta z_k\), where \(\mathcal{Z} = \{z_0, \cdots, z_K\}\) and
-\(\Delta z_k = z_k - z_{k-1}\). For convenience, we refer to
-\(\mathcal{E}^2_{\mathcal{Z}}\) as accumulated error. We search for a
-partitioning \(\mathcal{Z}_*\) that balances this trade-off.
+\(\Delta z_k = z_k - z_{k-1}\). We search for a partitioning that
+balances this trade-off.
 
-Due to the Corollary~\ref{sec:coroll}, minimizing
+Corollary~\ref{sec:coroll} shows that minimizing
 \( \mathcal{E}^2_{\mathcal{Z}} \) is equivalent to minimizing
-\(\sigma_*^2(z_{k-1}, z_k)\Delta z_k\), which can be directly
-estimated by \(\hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \).
+\(\sum_{k=1}^K \sigma_*^2(z_{k-1}, z_k)\Delta z_k\), which can be directly
+estimated from \(\sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \).
 
@@
   \label{sec:coroll}
-  A bin-splitting \(\mathcal{Z}\) minimizes the accumulated if and
-  only if it also minimizes the accumulated error.
-  \begin{equation}
-  \begin{gathered}
-    \label{eq:2}
-    \argmin_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \mathcal{E}^2_{\mathcal{Z}}  = \argmin_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \sum_{k=1}^K\sigma_*^2(z_1, z_2) \Delta z_k 
-  \end{gathered}
-  \end{equation}
+  A bin-splitting \(\mathcal{Z}\) that minimizes the accumulated error
+  if and only if it also minimizes
+  \(\sum_{k=1}^K\sigma_*^2(z_1, z_2) \Delta z_k \)
 \end{corollary}
@@
 \begin{proof}
-~\label{sec:coroll-1}
-  The proof is at Appendix~\ref{sec:proof-of-coroll}
+  ~\label{sec:coroll-1} The proof is based on the observation that
+  \(\sum_{k=1}^K \sigma^2(z_{k-1}, z_k) \Delta z_k = \sigma^2(z_0,
+  z_K) (z_K - z_0)\) which is independent of the bin-splitting. A more
+  detailed proof is provided in the Appendix~\ref{sec:proof-of-coroll}.
  \end{proof}
@@
 similar aggregate-error, the term \(\tau_K\) favors the partitioning
-with more points per bin (with wider bins). The constraint of at least
-\(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
+with more points per bin, i.e., with wider bins. The constraint of at
+least \(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
 \textit{robust} estimation. The user can choose to what extent they
@@
 i.e. \(\Delta x_{min} = \frac{x_{s, max} - x_{s,min}}{K_{max}}\).
-Based on that, we restrict the bin-limits to be multiples of the
+Based on that, we restrict the bin-limits to the multiples of the
 minimum width, i.e.
@@
 K_{max}\} \). In this discretized solution space, we find the global
-optimal using dynamic programming.
+optimum using dynamic programming. \todo{Add description} See also
+Appendix~\ref{sec:dynamic-programming}.
 
-\subsubsection{Solve Bin-Splitting with Dynamic Programming}
-\label{sec:dynamic-programing}
+% \begin{algorithm}
+%   \end{algorithm}
 
@@
 
-In XAI it is common~\citep{aas2021explaining, herbinger2022repid} to
-use simulation examples for compering methods, because the
-data-generating distribution \(p(\mathbf{X})\) and the predictive
-function \(f(\cdot)\) are known, and, therefore, the ground-truth is
-accesible. We follow this common XAI practice and we split the
-evaluation in two groups of examples. The first group
-(Section~\ref{sec:simulation-examples-1}) compares UALE definition
-against PDP-ICE in modeling the average effect and the
-uncertainty. The second group
-(Section~\ref{sec:simulation-examples-2}) compares UALE approximation
-using the automatic bin-splitting against the fixed-size alternative,
-in estimating the average effect and the uncertainty.
+Similarly to~\citep{aas2021explaining, herbinger2022repid} we first
+use simulation examples for evaluation. The first group
+(Section~\ref{sec:simulation-examples-1}) compares UALE against
+PDP-ICE in modeling the average effect and the uncertainty. The second
+group (Section~\ref{sec:simulation-examples-2}) compares UALE
+approximation with optimal bin-splitting against the fixed-size
+alternative.
 
@@
 
-In this simulation, we compare UALE against PDP-ICE. The aim of this
-example is to highlight that when features are correlated, PDP and ICE
-fail in quantifying the average effect and the heterogeneity, even
-when the black-box function is relatatively simple.
+The aim of this example is to highlight that when features are
+correlated, PDP and ICE fail in quantifying the average effect and the
+heterogeneity, even when the black-box function is relatatively
+simple.
 
@@
   f(\mathbf{x}) = \begin{cases}
-                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
-                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
-                    \alpha f_{\mathtt{int}} &\text{otherwise}
+                    f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $f_1(\mathbf{x}) < \frac{1}{2}$ }\\
+                    \frac{1}{2} - f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $\frac{1}{2} \leq f_2(\mathbf{x}) < 1$}\\
+                    \alpha f_1(\mathbf{x}) &\text{otherwise}
                   \end{cases}
@@
 %
-where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
-\(f_{\mathtt{int}}(\mathbf{x}) = x_1x_3\). The linear term
-\(f_{\mathtt{lin}}\) includes the two correlated features and the
-interaction term \(f_{\mathtt{int}}\) the two non-correlated.  We
-evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
-without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
-(b) without interaction (\(\alpha=0\)) and different weigths
-(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha \neq 0\)) with
-equal weights (\(a_1=a_2\)). In all cases, we provide the ground-truth
-average effect and uncertainty computed with analytical derivations
-(proofs at the Appendix) and we compare it against the approximations
-provided by each method, after generating \(N=500\) samples.
+where \(f_1(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
+\(f_2(\mathbf{x}) = x_1x_3\). The linear term \(f_1\) includes the two
+correlated features and the interaction term \(f_2\) the two
+non-correlated.  We evaluate the effect computed by UALE and PDP-ICE
+in three cases; (a) without interaction (\(\alpha=0\)) and equal
+weights (\(a_1=a_2\)), (b) without interaction (\(\alpha=0\)) and
+different weigths (\( a_1 \neq a_2 \)) and (c) with interaction
+(\(\alpha \neq 0\)) with equal weights (\(a_1=a_2\)).
+
+In all cases, we assume that because the \todo{give ground truth
+  effect}. We compare it against the approximations provided by each
+method, after generating \(N=500\) samples.
 
@@
 Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP effect is
-wrong and ICE plots imply the existence of heterogeneous effects. In
-contrast, UALE models correctly both the average effect and the zero
-uncertainty and approximates it accurately finding a zero aggregate
-residual error partitioning of \(x_1\) domain.
+wrong and ICE plots show heterogeneous effects. In contrast, UALE
+models correctly both the average effect and the
+uncertainty. Furthermore, we observe that bin-splitting is accurate
+(zero cumulative error \(\mathcal{E}_{\mathcal{Z}}\)).
 
@@
 interact with other features.  In Figure~\ref{fig:ex-synth-1-2}, we
-observe that PDP estimation is completely opposite to the ground-truth
-effect, i.e.~negative in the region \([0, 0.2)\), positive in
-\([0.2, 0.4)\), and the ICE erroneously implies the existence of
-heterogeneous effects. As before, ALE quantifies correctly the ground
-truth effect, the zero-uncertainty and partioning correctly the
-\(x_1\) domain.
+observe that PDP estimation is opposite to the ground-truth effect,
+i.e.~negative in the region \([0, 0.2)\), positive in \([0.2, 0.4)\),
+and the ICE erroneously implies the existence of heterogeneous
+effects. As before, ALE quantifies correctly the ground truth effect,
+the zero-uncertainty and partioning correctly the \(x_1\) domain.
 
@@
 
-\paragraph{Uncertainty, Equal weights.}
+\paragraph{With Interaction, Equal weights.}
 
@@
 heterogeneous effects for features \(x_1\) and \(x_3\), because the
-local effects at \(x_1\) depend on the unknown value of \(X_3\) and
+local effects at \(x_1\) depend on the unknown value of \(x_3\) and
 vice-versa. The effect of \(x_2\) has zero-uncertainty since it does
@@
 quantifies correctly the average effect and the uncertainty only in
-the case of feature \(x_3\), because \(X_3\) is independent from other
-features. For the correlated features (\(x_1, x_2\)) both the average
-effect (PDP) and the uncertainty (ICE) are wrong.
+the case of feature \(x_3\), because it is independent from other
+features. For the correlated features \(x_1\) and \(x_2\) both the
+average effect (PDP) and the uncertainty (ICE) are wrong.
 
@@
 because of limited samples. In contrast, UALE models both correctly
-and estimates them accurately extracting the regions with constant
-effect.  The examples above do not cover the case of an interaction
-term between correlated features, for example a term \(x_1x_2\),
-because there is an open debate about the ground-truth effect in this
+and estimates them accurately using an optimal bin splitting in terms
+of Eq.~(\ref{eq:opt}). The examples above do not cover the case of an interaction term
+between correlated features, for example a term \(x_1x_2\), because
+there is an open debate about the ground-truth effect in this
 case~\citep{Gromping2020MAEP}.
@@
 we treat the estimation with dense fixed-size bins (\(K=10^3\)) as the
-ground-truth UALE. Afterwards, we generate less samples (\(N=500\))
-and we compare the fixed-size estimation (for many different \(K\))
-against UALE automatic bin-splitting. In all set-ups, we sample from
+ground-truth UALE. Afterwards, we generate fewer samples (\(N=500\))
+and we compare the fixed-size estimation (for several \(K\)) against
+UALE automatic bin-splitting. In all set-ups, we sample from
 \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
@@
 \(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
-\(\mathcal{Z^{\mathtt{DP}}} = \{z^{\mathtt{DP}}_{k-1}, \cdots,
-z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic
+\(\mathcal{Z^*} = \{z^*_0, \cdots,
+z^*_K\}\) the sequence obtained by automatic
 bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size
 splitting with \(K\) bins. The evaluation metrics we report are the
-average number of \(t = 30\) independent runs, using each time
+average number of independent runs \(t = 30\), using each time
 \(N=500\) different samples.
@@
 
-The evaluation is done with regard to two metrics counting the mean
-error per bin, where the bin error is definde as the absolute
-difference of the approximation from the ground-truth. The first
-metric, Eq.~\eqref{eq:eval_met_1}, counts the mean bin error wrt
-average effect and the second, Eq.~\eqref{eq:eval_met_2} the mean bin
-error wrt uncertainty:
+Evaluation is perform in terms of the Mean Absolute Error (MAE) of the
+estimation of \(\mu\) and \(\sigma\) accross bins, i.e.,
+
+% The evaluation is done with regard to two metrics counting the mean
+% error per bin, where the bin error is definde as the absolute
+% difference of the approximation from the ground-truth. The first
+% metric, Eq.~\eqref{eq:eval_met_1}, counts the mean bin error wrt
+% average effect and the second, Eq.~\eqref{eq:eval_met_2} the mean bin
+% error wrt uncertainty:
 
@@
   \label{eq:eval_met_1}
-  \mathcal{L}_{\mathtt{DP|K}}^{\mu} = \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
-  \mathcal{Z}^{\mathtt{DP|K}}} | \mu(z_k- z_{k-1}) - \hat{\mu}(z_k-
-  z_{k-1}) |
+  \mathcal{L}^{\mu} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
+  \mathcal{Z}} | \mu(z_{k-1}, z_k) - \hat{\mu}(z_{k-1}, z_k) |
 \end{equation}
@@
   \label{eq:eval_met_2}
-  \mathcal{L}_{\mathtt{DP|K}}^{\sigma} = \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
-    \mathcal{Z}^{\mathtt{DP|K}}} | \sigma(z_k- z_{k-1}) - \hat{\sigma}(z_k-
-  z_{k-1}) |
+  \mathcal{L}^{\sigma} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
+    \mathcal{Z}} | \sigma(z_{k-1}, z_k) - \hat{\sigma}(z_{k-1}, z_k) |
 \end{equation}
 
-The ground truth \(\mu_k\) (\(\sigma_k\)) is the mean value of the
-bin-effects (bin-uncertainties) of the bins inside the interval
-defined by the wide bin. For better interpretation of the uncertainty
-error, we also provide the mean (per bin) error term
-\(\mathcal{L}^{\rho}_{\mathtt{DP|K}} =
-\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in \mathcal{Z}^{\mathtt{DP|K}}} \mathcal{E}(z_{k-1}, z_k) \).
+The ground truth UALE is the average of the dense fixed-size bins that
+are within in the interval \([z_{k-1}, z_k]\). For better
+interpretation of \(\mathcal{L}^{\sigma}\), we also provide the mean
+error term
+\(\mathcal{L}^{\rho} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
+  \mathcal{Z}} \mathcal{E}(z_{k-1}, z_k) \).
 
@@
 piecewise-linear function with \(5\) different-width regions, i.e.,
-\(a_1\) equals to \(=\{2, -2, 5, -10, 0.5\}\) in the intervals defined
-by the sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). As we observe, in
-Figure~\ref{fig:ex-synth-2-1}, UALE's approximation is better than all
-fixed-size alternatives, in both mean effect (\(\mathcal{L}^{\mu}\))
-and uncertainty (\(\mathcal{L}^{\sigma}\)). For understanding the
-importance of variable-size splitting, we analyze the fixed-bin
-error. In the top-right of Figure~\ref{fig:ex-synth-2-1}, we observe
-the Law of Large Numbers (LLN); \(\mathcal{L}^{\mu}_{\mathtt{K}}\) and
-\(K\) have positive correlation because as \(\Delta x\) becomes
-smaller less points contributing to each estimation. The
-interpretation of \(\mathcal{L}^{\sigma}\) is more complex. Given that
-the smallest piecewise-linear region is \(\Delta x = 0.05 \) any
-\(K\) that is not a multiple of \(\frac{1}{\Delta x} = 20\), adds
-nuisance uncertainty \(\mathcal{L}^{\rho}_{\mathtt{K}} > 0\) leading
-to a biased uncertainty approximation. For these \(K\), uncertainy
-error decreases for bigger \(K\) because the length of the bins that
-lie in the limit between two piecewise linear regions becomes
-smaller. For \(K\) that are multiple of \(20\), i.e.
-\(K=\{20, 40, 60, 80, 100\}\), the uncertainty approximation is
-unbiased \(\mathcal{L}^{\rho}_{\mathtt{K}} \approx 0\). Here, as in
-\(\mathcal{L}^{\mu}\), uncertainty estimation is worse as \(K\) grows
-larger (LLN). As we observe in the top left of
-Figure~\ref{fig:ex-synth-2-1}, automatic bin-splitting separates the
-fine-grain bins, e.g.~regions \([0.4, 0.45]\), \([0.45, 0.5]\), and
-unites constant-effect regions into a single bin, e.g.~region
-\([0.5, 1]\).
+\(a_1\) equals to \(\{2, -2, 5, -10, 0.5\}\) in the intervals defined
+by the sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\).
+
+As we observe in the top left of Figure~\ref{fig:ex-synth-2-1}, UALE's
+bin-splitting separates the fine-grained bins, e.g.~regions
+\([0.4, 0.45]\), \([0.45, 0.5]\), and unites (most) constant-effect
+regions into a single bin, e.g.~region \([0.5, 1]\). This explains the
+fact that UALE achieves lower MAE \(\mathcal{L}^{\mu}\),
+\(\mathcal{L}^{\sigma}\) compared to fixed-size bins for all \(K\).
+
+% We observe that in
+% this case all intervals are multiples of \(0.05\). As we observe, in
+% Figure~\ref{fig:ex-synth-2-1}, UALE's approximation is better than all
+% fixed-size alternatives (for all \(K\)), in both mean effect
+% \(\mathcal{L}^{\mu}\) and uncertainty \(\mathcal{L}^{\sigma}\). For
+% understanding the importance of variable-size splitting, we analyze
+% the fixed-bin error. In the top-right of
+% Figure~\ref{fig:ex-synth-2-1}, we observe a positive trend between
+% \(\mathcal{L}^{\mu}\) and \(K\), because as \(K\) becomes larger fewer
+% points contribute to the bin estimations, leading to a worse
+% approximation. The interpretation of \(\mathcal{L}^{\sigma}\) is more
+% complex. For \(K\) that are multiples of \(20\), i.e.
+% \(K=\{20, 40, 60, 80, 100\}\), the mean error term
+% \(\mathcal{L}^{\rho}\) is low as we see in the bottom-right of
+% Figure~\ref{fig:ex-synth-2-2}. In these cases, we also observe in the
+% bottom-left figure, that \(L\) is increasing with \(K\) due to fewer
+% points in each bin. For all other values of \(K\), the \(L\) is mainly
+% affected by the mean error term \(\mathcal{E}\).
 
@@
 In this set-up, \(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), so the
-effect is non-linear in \(0 \leq x_1 \leq 1\). In this set-up, wide
-bins increase \(\mathcal{L}^{\rho^2}\) making the uncertainty
-approximation more biased and narrow bins lead to a worse
-approximation. In Figure~\ref{fig:ex-synth-2-2}, we observe that
-automatic bin splitting manages to compromise the
-conflicting objectives, \todo{Explain better}
+effect is non-linear in \(0 \leq x_1 \leq 1\). Due to this non
+linearity, wide bins increase \(\mathcal{L}^{\rho}\) making the
+uncertainty approximation biased. On the other hand, narrow bins lead
+to worse approximation due to the limited number of
+samples. Interestingly, in Figure~\ref{fig:ex-synth-2-2}, we observe
+that UALE's bin splitting manages to compromise these conflicting
+objectives.
 
@@
 Here, we aim at demonstrating the usefuleness of uncertainty
-quantification and the advantages of automatic bin-splitting, on the
+quantification and the advantages of UALE's approximation, on the
 real-world California Housing dataset~\citep{pace1997sparse}.
@@
 \(20000\) training instances, making it appropriate for robust
-Monte-Carlo approximations. The dataset contains \(D=8\) numerical
+approximation with large \(K\). The dataset contains \(D=8\) numerical
 features with characteristics about the building blocks of California,
@@
 houses inside the block in dollars that ranges between
-\([15, 500] \cdots 10^3\), with a mean value of
+\([15, 500] \cdot 10^3\), with a mean value of
 \(\mu_Y \approx 201 \cdot 10^3 \) and a standard deviation of
@@
 128 and 36 units respectively. After 15 epochs using the Adam
-optimizer with learning rate \(\eta = 0.02\), the model achieves a
-normalized mean squared error (R-Squared) of \(0.25\) (0.75), which
-corresponds to a MAE of \(37 \cdot 10^3\) dollars.
+optimizer with learning rate \(\eta = 0.02\), the model achieves a MAE
+of \(37 \cdot 10^3\) dollars.
 
@@
 features cover the main FE cases, e.g.~positive/negative trend and
-linear/non-linear curve, and they are appropriate for illustration
-purposes. The complete results for all features, along with in-depth
+linear/non-linear curve, and are therefore appropriate for
+illustration purposes. Results for all features, along with in-depth
 information about the preprocessing, training and evaluation parts are
@@
 In real-world datasets, it is infeasible to obtain the ground truth FE
-for seamlessly evaluating the competitive methods. We selected the
-particular experiment, because, in broad terms, UALE and PDP-ICE plots
-aggree in the estimation of the average effect and
-uncertainty. Therefore, we can focus on juding the quality of the
-information provided by the two methods.
+for evaluation. We observe in Figure \ref{fig:ex-real-1} that in broad
+terms UALE and PDP-ICE plots aggree on the average effect and the
+uncertainty. It is interesting to note that UALE has selected bins of
+varying size as the optimal partiotioning. This demonstrates an
+important advantage of UALE; through the bin splitting process we can
+identify wide intervals of the feature domain with similar level of
+uncertainty.
 
-In Figure\ref{fig:ex-real-1}, we observe, from top to bottom, the
-effects for the latitude, population and the median income. The effect
-of UALE and PDP-ICE are similar for the population and the median
-income. The population has a negative impact that progressively
-decreases: from 400 to 1500 people the house value decreases with a
-rate of \(-150 (\pm 140)\) dollars per added person, a rate that
-decreases from \(-80 (\pm 80)\) to \(-60 (\pm 60)\) dollars per added
-person as we move from 1500 to 2800 people. The level of uncertainty
-indicates significant variance in absolute value of the rate, but in
-the grant majority of instances the rate is negative. With the same
-inspection, we observe that the median annual income has a positive
-impact on the value (all numbers are thousands of dollars):
-\(20\pm 15\) per \(10\) of added median income for incomes in
-\([8, 15]\), \(32 \pm 20\) per \(10\) added income in \([15, 60]\) and
-\(40 \pm 15\) per \(10\) added income in \([60, 70]\). The uncertainty
-indicates that there are less heterogeneous effects about the median
-income compared to the number of people. In both cases, we can end-up
-to the same conclusion by inspecting the PDP-ICE plots. For the
-latitude, there is a small difference in the explanations for the
-region \([32, 35]\), where UALE estimates a less negative slope with
-less uncertainty than PDP, while the explanations are similar for the
-range \([35,39]\), where both methods reveal an increase in the
-uncertainty around the feature value \(37.5\).
+% We selected the particular experiment, because, in
+% broad terms, UALE and PDP-ICE plots aggree in the estimation of the
+% average effect and uncertainty. Therefore, we can focus on juding the
+% quality of the information provided by the two methods.
 
-In general, we observe that UALE complements ALE in the quantification
-of the heterogeneous effects, similarly to as ICE complements PDP. The
-automatic extraction of constant-effect and constant-uncertainty
-regions provided by UALE is helpful for an easier interpretation. On
-the other hand, ICE plots sometimes are more descritive on locating
-the type of heterogeneity, whereas UALE quantifies only the level (not
-the type) of heterogeneity.
+% In Figure\ref{fig:ex-real-1}, we observe, from top to bottom, the
+% effects for the latitude, population and the median income. The effect
+% of UALE and PDP-ICE are similar for the population and the median
+% income. The population has a negative impact that progressively
+% decreases: from 400 to 1500 people the house value decreases with a
+% rate of \(-150 (\pm 140)\) dollars per added person, a rate that
+% decreases from \(-80 (\pm 80)\) to \(-60 (\pm 60)\) dollars per added
+% person as we move from 1500 to 2800 people. The level of uncertainty
+% indicates significant variance in absolute value of the rate, but in
+% the grant majority of instances the rate is negative. With the same
+% inspection, we observe that the median annual income has a positive
+% impact on the value (all numbers are thousands of dollars):
+% \(20\pm 15\) per \(10\) of added median income for incomes in
+% \([8, 15]\), \(32 \pm 20\) per \(10\) added income in \([15, 60]\) and
+% \(40 \pm 15\) per \(10\) added income in \([60, 70]\). The uncertainty
+% indicates that there are less heterogeneous effects about the median
+% income compared to the number of people. In both cases, we can end-up
+% to the same conclusion by inspecting the PDP-ICE plots. For the
+% latitude, there is a small difference in the explanations for the
+% region \([32, 35]\), where UALE estimates a less negative slope with
+% less uncertainty than PDP, while the explanations are similar for the
+% range \([35,39]\), where both methods reveal an increase in the
+% uncertainty around the feature value \(37.5\).
+
+% In general, we observe that UALE complements ALE in the quantification
+% of the heterogeneous effects, similarly to as ICE complements PDP. The
+% automatic extraction of constant-effect and constant-uncertainty
+% regions provided by UALE is helpful for an easier interpretation.
+
+% On
+% the other hand, ICE plots sometimes are more descritive due to
+% illustrating the type of heterogeneity, whereas UALE quantifies only the
+% level (not the type) of heterogeneity.
 
@@
 
-In this part, we evaluate the robustness of the approximation using
-automatic bin-splitting. Following the evaluation framework of
-Section~\ref{sec:simulation-examples-2}, we treat as ground-truth the
-effects computed on the full training-set \(N=20000\) with dense
-fixed-size bin-splitting (\(K=80\)). Given the big number of samples,
-we make the hypothesis that the approximation with dense binning is
-close to the ground truth. Afterwards, we randomly select less samples
-\(N=1000\) and we compare UALE approximation with all possible
-fixed-size alternatives, repeating this process for 30 times for
-robust results. In Figure~\ref{fig:ex-real-2}, we illustrate the mean
-values for \(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\) of the 30
-repetitions. We observe that automatic bin-spliting provides (close to
-the) best approximation in the three features. In the Appendix, we
-provide the same evaluation for all features.
+We evaluate the robustness of UALE approximation. Following the
+evaluation framework of Section~\ref{sec:simulation-examples-2}, we
+treat as ground-truth the effects computed on the full training-set
+\(N=20000\) with dense fixed-size bin-splitting (\(K=80\)). Given the
+big number of samples, we make the hypothesis that the approximation
+with dense binning is close to the ground truth. Afterwards, we
+randomly select fewer samples, \(N=1000\), and we compare UALE
+approximation with all possible fixed-size approximation. We repeat
+this process \(t=30\) times. In Figure~\ref{fig:ex-real-2}, we
+illustrate the mean values of
+\(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\) across the 30
+repetitions. We observe that UALE achieves (near) optimal
+approximation in all cases. \todo{add explanation for near-constant
+  feature effect vs the first case that the plot }
 

### 2022-10-13 AISTATS2023/paper.tex
latest

@@
 
-Similarly to~\citep{aas2021explaining, herbinger2022repid} we first
-use simulation examples for evaluation. The first group
-(Section~\ref{sec:simulation-examples-1}) compares UALE against
-PDP-ICE in modeling the average effect and the uncertainty. The second
-group (Section~\ref{sec:simulation-examples-2}) compares UALE
-approximation with optimal bin-splitting against the fixed-size
-alternative.
+We perform a two-fold evaluation of UALE. The first group
+(Section~\ref{sec:simulation-examples-1}) compares UALE with PDP-ICE,
+the main feature effect method that also quantifies the heterogeneity
+of instance-based effects. The second group
+(Section~\ref{sec:simulation-examples-2}) evaluates UALE
+approximation, i.e., compares UALE's optimal bin-splitting against the
+fixed-size alternative.
 
@@
 
-The aim of this example is to highlight that when features are
-correlated, PDP and ICE fail in quantifying the average effect and the
-heterogeneity, even when the black-box function is relatatively
+This a qualitative example with the aim to highlight that PDP-ICE is
+vulnerable to misleading estimations of both the average effect and
+the uncertainty, even when the black-box function is relatatively
 simple.
 
-We use the following distribution:
+We use the data generating distribution
 \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
@@
 \(x_3 \sim \mathcal{N}(0, \sigma_3^2)\) and the following piecewise
-linear function:
+linear function
 
@@
                     f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $f_1(\mathbf{x}) < \frac{1}{2}$ }\\
-                    \frac{1}{2} - f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $\frac{1}{2} \leq f_2(\mathbf{x}) < 1$}\\
-                    \alpha f_1(\mathbf{x}) &\text{otherwise}
+                    \frac{1}{2} - f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $\frac{1}{2} \leq f_1(\mathbf{x}) < 1$}\\
+                    \alpha f_2(\mathbf{x}) &\text{otherwise}
                   \end{cases}
@@
 where \(f_1(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
-\(f_2(\mathbf{x}) = x_1x_3\). The linear term \(f_1\) includes the two
-correlated features and the interaction term \(f_2\) the two
-non-correlated.  We evaluate the effect computed by UALE and PDP-ICE
-in three cases; (a) without interaction (\(\alpha=0\)) and equal
-weights (\(a_1=a_2\)), (b) without interaction (\(\alpha=0\)) and
-different weigths (\( a_1 \neq a_2 \)) and (c) with interaction
-(\(\alpha \neq 0\)) with equal weights (\(a_1=a_2\)).
+\(f_2(\mathbf{x}) = x_1x_3\). The part \(f_1\) is a linear function of
+the two correlated features, \(x_1, x_2\), and \(f_2\) is an
+interaction term between the two non-correlated, \(x_1, x_3\).  We
+evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
+without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
+(b) without interaction (\(\alpha=0\)) and different weigths
+(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha > 0\)) with
+equal weights (\(a_1=a_2\)).
 
-In all cases, we assume that because the \todo{give ground truth
-  effect}. We compare it against the approximations provided by each
-method, after generating \(N=500\) samples.
+In the general case, there is no unique ground-truth feature effect
+and uncertainty and each method defines the effect in a different
+way. We select this simple case, to determine the ground-truth,
+exploiting two facts. First, in \(p(\xb)\) it holds that
+\(x_1=x_2\). Therefore, knowing the value of either \(x_1\) or
+\(x_2\), we know the closed-form of \(f(\xb)\), too. For example, if
+\(a_1=a_2=1\) and \(x_1 = \frac{1}{6}\), then
+\(f_1(\xb) = \frac{1}{3}\), so
+\(f(\xb) = f_1(\xb) + \alpha f_2(\xb) \). Afterwards, we use the
+properties of interpretable models~\citep{molnar2022} to determine the
+ground-truth. In the example above, if also \(a=0\), then
+\(f(\xb) = x_1 + x_2\) which is an interpretable linear model. For the
+uncertainty, we use the fact that under no-interactions, \(a=0\), the
+uncertainty must be zero (no heteregoneity). We discuss separately the
+case with \(a>0\). Finally, we want to also show that UALE's
+approximation is accurate, due to appropriate bin splitting.
 
@@
 In this set-up, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\)
-(equal weights). The ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\)
-is: \(x_1\) in \([0, 0.25]\), \(-x_1\) in \([0.25, 0.5]\) and \(0\) in
-\([0.5, 1]\). The uncertainty is
-\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) in \([0,1]\) because \(x_1\)
-does not interact with ohter variables. In
-Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP effect is
-wrong and ICE plots show heterogeneous effects. In contrast, UALE
-models correctly both the average effect and the
+(equal weights). We examine only the effect of feature \(x_1\),
+because it is exactly the same with \(x_2\). Based on the discussion
+above, the ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is:
+\(x_1\) in \([0, \frac{1}{4}]\), \(-x_1\) in
+\([\frac{1}{4}, \frac{1}{2}]\) and \(0\) in \([\frac{1}{2},
+1]\). Because \(x_1\) does not interact with any other variable, the
+uncertainty is \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) in
+\([0,1]\). In Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP
+effect is wrong and ICE plots show heterogeneous effects. In contrast,
+UALE models correctly both the average effect and the absence of
 uncertainty. Furthermore, we observe that bin-splitting is accurate

### 2022-10-13 AISTATS2023/paper.tex
latest

@@
 
-We perform a two-fold evaluation of UALE. The first group
+We perform a two-fold evaluation of UALE. First,
 (Section~\ref{sec:simulation-examples-1}) compares UALE with PDP-ICE,
 the main feature effect method that also quantifies the heterogeneity
-of instance-based effects. The second group
+of instance-level effects. Second,
 (Section~\ref{sec:simulation-examples-2}) evaluates UALE
@@
 
-\subsection{Case 1: UALE vs PDP-ICE}
+\subsection{UALE vs PDP-ICE}
 \label{sec:simulation-examples-1}
 
-This a qualitative example with the aim to highlight that PDP-ICE is
-vulnerable to misleading estimations of both the average effect and
-the uncertainty, even when the black-box function is relatatively
-simple.
+This a qualitative example to highlight that PDP-ICE is vulnerable to
+misleading estimations of both the average effect and the uncertainty,
+even in case of a simple black-box function, as opposed to UALE.
 
+\paragraph{Example set-up.}
 We use the data generating distribution
-\(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
+\(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\), where
 \(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
-\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\) and the following piecewise
-linear function
+\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\), and the following function,
 
@@
 (b) without interaction (\(\alpha=0\)) and different weigths
-(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha > 0\)) with
+(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha > 0\)) and
 equal weights (\(a_1=a_2\)).
 
-In the general case, there is no unique ground-truth feature effect
-and uncertainty and each method defines the effect in a different
-way. We select this simple case, to determine the ground-truth,
-exploiting two facts. First, in \(p(\xb)\) it holds that
-\(x_1=x_2\). Therefore, knowing the value of either \(x_1\) or
-\(x_2\), we know the closed-form of \(f(\xb)\), too. For example, if
-\(a_1=a_2=1\) and \(x_1 = \frac{1}{6}\), then
-\(f_1(\xb) = \frac{1}{3}\), so
-\(f(\xb) = f_1(\xb) + \alpha f_2(\xb) \). Afterwards, we use the
-properties of interpretable models~\citep{molnar2022} to determine the
-ground-truth. In the example above, if also \(a=0\), then
-\(f(\xb) = x_1 + x_2\) which is an interpretable linear model. For the
-uncertainty, we use the fact that under no-interactions, \(a=0\), the
-uncertainty must be zero (no heteregoneity). We discuss separately the
-case with \(a>0\). Finally, we want to also show that UALE's
-approximation is accurate, due to appropriate bin splitting.
+In the general case, evaluating feature effect methods is a
+challenging task, because there is no unique ground-truth feature
+effect and uncertainty and each method defines the effect in a
+different way. To deal with this, we exploit two characteristics of
+the above set-up. First, in \(p(\xb)\) it holds that
+\(x_1=x_2\). Therefore, knowing \(x_1\) or \(x_2\), we know the
+closed-form of \(f(\xb)\), too. For example, if \(a_1=a_2=1\) and
+\( 0 \leq x_1 < \frac{1}{4}\), then \(f_1(\xb) < \frac{1}{2}\), so
+\(f(\xb) = f_1(\xb) + \alpha f_2(\xb) \), i.e., the first branch of
+Eq.~\ref{eq:synth-ex-1-function}. Afterwards, we can determine the
+ground-truth. In the example above, if \(a=0\), then
+\(f(\xb) = x_1 + x_2\) and therefore the ground-truth effect of
+\(x_1\) is \(f^{\mathcal{GT}}(x_1) = x_1\) in the corresponding
+interval, i.e. \(x_1 \in [0, \frac{1}{4})\). For the uncertainty, we
+use the fact that under no-interactions, \(a=0\), the uncertainty must
+be zero (no heteregoneity). We discuss separately the case with
+\(a>0\). Finally, we also demonstrate that UALE's approximation is
+accurate, due to the bin splitting that we propose.
 
-\paragraph{No Interaction, Equal weights.}
+\paragraph{(a) No Interaction, Equal weights.}
 
-In this set-up, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\)
-(equal weights). We examine only the effect of feature \(x_1\),
-because it is exactly the same with \(x_2\). Based on the discussion
-above, the ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is:
-\(x_1\) in \([0, \frac{1}{4}]\), \(-x_1\) in
-\([\frac{1}{4}, \frac{1}{2}]\) and \(0\) in \([\frac{1}{2},
-1]\). Because \(x_1\) does not interact with any other variable, the
-uncertainty is \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) in
-\([0,1]\). In Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP
-effect is wrong and ICE plots show heterogeneous effects. In contrast,
-UALE models correctly both the average effect and the absence of
-uncertainty. Furthermore, we observe that bin-splitting is accurate
-(zero cumulative error \(\mathcal{E}_{\mathcal{Z}}\)).
+Here, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\) (equal
+weights). We examine only the effect of feature \(x_1\), because it is
+exactly the same as with \(x_2\). Based on the discussion above, the
+ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\) in
+\([0, \frac{1}{4}]\), \(-x_1\) in \([\frac{1}{4}, \frac{1}{2}]\) and
+\(0\) in \([\frac{1}{2}, 1]\). Because \(x_1\) does not interact with
+any other feature, the uncertainty is
+\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
+Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP effect is
+wrong and ICE plots show heterogeneous effects. In contrast, UALE
+models correctly both the average effect and the absence of
+uncertainty. Finally, we observe that UALE's bin-splitting leads (a)
+to an accurate uncertainty estimation (near-zero cumulative bin-error
+\(\mathcal{E}_{\mathcal{Z}}\)) and (b) an easier interpretation,
+creating three wide bins, i.e.
+\([0, \frac{1}{4}), [\frac{1}{4}, \frac{1}{2}), [\frac{1}{2}, 1)\),
+that correspond to the piecewise-linear regions of the effect.
 
@@
 
-\paragraph{No Interaction, Non-Equal Weights.}
+\paragraph{(b) No Interaction, Non-Equal Weights.}
 
-In this set-up, \(\alpha=0\) (no interaction), \(a_1 = 2\) and
-\(a_2= 0.5\) (non-equal weights). The non-equal weights have
+Here, \(\alpha=0\) (no interaction), \(a_1 = 2\) and
+\(a_2= \frac{1}{2}\) (non-equal weights). The non-equal weights have
 implications at both the gradient and the interval of the piece-wise
 linear regions, i.e., \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) in
-\([0, 0.2]\), \(-2x_1\) in \([0.2, 0.4]\) and \(0\) in \([0.4,
-1]\). As before, the ground-truth uncertainty is
-\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) because \(x_1\) does not
-interact with other features.  In Figure~\ref{fig:ex-synth-1-2}, we
+\([0, \frac{1}{5}]\), \(-2x_1\) in \([\frac{1}{5}, \frac{2}{5}]\) and
+\(0\) in \([\frac{2}{5}, 1]\). As before, the ground-truth uncertainty
+is \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) because \(x_1\) does not
+interact with any other feature. In Figure~\ref{fig:ex-synth-1-2}, we
 observe that PDP estimation is opposite to the ground-truth effect,
-i.e.~negative in the region \([0, 0.2)\), positive in \([0.2, 0.4)\),
-and the ICE erroneously implies the existence of heterogeneous
-effects. As before, ALE quantifies correctly the ground truth effect,
-the zero-uncertainty and partioning correctly the \(x_1\) domain.
+i.e.~negative in the region \([0, \frac{1}{5})\), positive in
+\([\frac{1}{5}, \frac{2}{5})\), and the ICE erroneously implies the
+existence of heterogeneous effects. As before, UALE quantifies
+correctly the ground truth effect, the zero-uncertainty and partitions
+the \(x_1\) domain into wide bins that facilitate the interpretation
+and create a zero cummulative bin-error \(\mathcal{E}_{\mathcal{Z}}\)
+approximation.
 
@@
 
-\paragraph{With Interaction, Equal weights.}
+\paragraph{(c) With Interaction, Equal weights.}
 
-In this set-up, we activate the interaction term, i.e.  \(a=1\),
-keeping the weights equal \(a_1=a_2=1\). The interaction term provokes
-heterogeneous effects for features \(x_1\) and \(x_3\), because the
-local effects at \(x_1\) depend on the unknown value of \(x_3\) and
-vice-versa. The effect of \(x_2\) has zero-uncertainty since it does
-not appear in any interaction term.  As we observe in
-Figure~\ref{fig:ex-synth-1-3}, UALE correcly models the average effect
-and the uncertainty in all cases.  On the other hand, PDP-ICE
-quantifies correctly the average effect and the uncertainty only in
-the case of feature \(x_3\), because it is independent from other
-features. For the correlated features \(x_1\) and \(x_2\) both the
-average effect (PDP) and the uncertainty (ICE) are wrong.
+Here, we activate the interaction term, i.e., \(a=1\), keeping the
+weights equal \(a_1=a_2=1\) and \(\sigma_3^2=\frac{1}{4}\). In this
+case, it is not straightforward to define the ground-truth effect for
+features \(x_1, x_3\), because the interaction term provokes
+heterogeneous instance-level effects, i.e., the instance-level effect
+of \(x_1\) depend on the unknown value of \(x_3\) and vice-versa. We
+observe that UALE's feature effects are quite intuitive. The average
+effect of \(x_1\) is the same with the Example (a) with an added
+uncertainty, reflecting the uncertainty about the instance-level
+effects which is the standard deviation of \(x_3\), i.e.,
+\(\sigma_3 = \frac{1}{2}\). The effect of \(x_3\) is only due to the
+interaction term, therefore, the instance-level effects
+\(\frac{\partial f}{\partial x_3} = x_1 \) follow
+\(p(x_1) \sim \mathcal{U}(0,1)\) which has \(\mu_{x_1}=\frac{1}{2}\)
+and \(\sigma_{x_1}=\frac{1}{4}\). This is reflected, in UALE's
+estimations of Figure~\ref{fig:ex-synth-1-3}.
+
+For \(x_2\), as in Example (a), the effect is
+\(f_\mu^{\mathtt{GT}}(x_2)\) is: \(x_2\) in \([0, \frac{1}{4}]\),
+\(-x_2\) in \([\frac{1}{4}, \frac{1}{2}]\) and \(0\) in
+\([\frac{1}{2}, 1]\)\(x_2\) and it has zero-uncertainty,
+\(f_\sigma^{\mathtt{GT}}(x_2) = 0\), since \(x_2\) does not appear in
+any interaction term. We confirm that UALE computes it correctly
+whereas PDP-ICE fail in both the average effect and the uncertainty.
 
@@
 
-Despite the model's simplicity, PDP-ICE fail in modeling both the
-average effect and the uncertainty. The error is due to the method's
-formulation, i.e.~the use of the marginal distribution that ignores
-correlations between features, and not due to poor approximation
-because of limited samples. In contrast, UALE models both correctly
-and estimates them accurately using an optimal bin splitting in terms
-of Eq.~(\ref{eq:opt}). The examples above do not cover the case of an interaction term
-between correlated features, for example a term \(x_1x_2\), because
-there is an open debate about the ground-truth effect in this
-case~\citep{Gromping2020MAEP}.
+The above experiments show that UALE models correctly and estimates
+accurately both the average effect and its uncertainty. We also
+observed that UALE's variable-size bin-splitting method that we
+propose, leads to (a) an accurate approximation of the uncertainty and
+(b) favoring wider bins we facilitate the interpretation of both the
+average effect and the uncertainty. In contrast, PDP-ICE provides
+misleading explanations, due to ignoring correlations between
+features. The examples above do not cover the case of an interaction
+term between correlated features, for example a term \(x_1x_2\),
+because in this case there is an open debate about the ground-truth
+effect~\citep{Gromping2020MAEP}.
 

### 2022-10-13 AISTATS2023/paper.tex
latest

@@
 
-\subsection{Case 2: Bin-Splitting}
+\subsection{UALE vs Fixed-Size Approximation}
 \label{sec:simulation-examples-2}
 
-In this simulation, we illustrate the advantages of automatic
-bin-splitting against the fixed-size alternative. For this reason, we
-generate a very big dataset applying dense sampling (\(N=10^6\)) and
-we treat the estimation with dense fixed-size bins (\(K=10^3\)) as the
+In this simulation, we focuse on illustrating the advantages of UALE's
+variable size bin-splitting against fixed-size bin-splitting.
+
+\paragraph{Example set-up.}
+The purpose of this experiment is two-fold. First, we show that UALE
+automatically provides an approximation that is as accurate (or even
+more accurate) than the best fixed-size approximation, among all
+possible \(K\). Second, we illustrate that UALE favors the creation of
+wide intervals when they are not adding cumulative bin-error
+\(\mathcal{E}^2_{\mathcal{Z}}\). Wide intervals facilitate the
+interpretation of the average effect and its uncertainty.
+
+For measuring the accuracy of the approximation, we generate a very
+big dataset with dense sampling (\(N=10^6\)) and we use dense
+fixed-size bins (\(K=10^3\)). We treat this estimation as the
 ground-truth UALE. Afterwards, we generate fewer samples (\(N=500\))
 and we compare the fixed-size estimation (for several \(K\)) against
-UALE automatic bin-splitting. In all set-ups, we sample from
-\(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
+UALE's. We sample from \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
 \(x_1 \sim \mathcal{U}(0,1)\) and
-\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
-\(\mathcal{Z^*} = \{z^*_0, \cdots,
-z^*_K\}\) the sequence obtained by automatic
-bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size
-splitting with \(K\) bins. The evaluation metrics we report are the
-average number of independent runs \(t = 30\), using each time
-\(N=500\) different samples.
-
-\paragraph{Metrics}
-
-Evaluation is perform in terms of the Mean Absolute Error (MAE) of the
-estimation of \(\mu\) and \(\sigma\) accross bins, i.e.,
-
-% The evaluation is done with regard to two metrics counting the mean
-% error per bin, where the bin error is definde as the absolute
-% difference of the approximation from the ground-truth. The first
-% metric, Eq.~\eqref{eq:eval_met_1}, counts the mean bin error wrt
-% average effect and the second, Eq.~\eqref{eq:eval_met_2} the mean bin
-% error wrt uncertainty:
+\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We, also, denote as
+\(\mathcal{Z^*} = \{z^*_0, \cdots, z^*_K\}\) the sequence obtained by
+automatic bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the
+fixed-size splitting with \(K\) bins. The evaluation is based on the
+Mean Absolute Error (MAE) in estimating the bin effect \(\mu\) and the
+uncertainty \(\sigma\) accross bins, i.e.,
 
@@
 
-The ground truth UALE is the average of the dense fixed-size bins that
-are within in the interval \([z_{k-1}, z_k]\). For better
-interpretation of \(\mathcal{L}^{\sigma}\), we also provide the mean
-error term
+We obtaint the ground truth bin-effect, \(\mu(z_{k-1}, z_k)\), and the
+ground truth uncertainty, \(\sigma(z_{k-1}, z_k)\), using the average
+of the dense fixed-size bins that are within in the interval
+\([z_{k-1}, z_k]\). We also provide
 \(\mathcal{L}^{\rho} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
-  \mathcal{Z}} \mathcal{E}(z_{k-1}, z_k) \).
+  \mathcal{Z}} \mathcal{E}(z_{k-1}, z_k) \) for facilitating the
+interpretation of the uncertainty error, and we call it mean residual error.
+
+We compare UALE vs fixed-size approximation, when (a) \(f\) is
+piecewise-linear and (b) \(f\) is non-linear. In all examples, we
+execute \(t = 30\) independent runs, using each time \(N=500\)
+different samples, and we report the mean values of the metrics.
 
@@
 
-In this set-up, \(f(\mathbf{x}) = a_1x_1 + x_1x_2\) is a
-piecewise-linear function with \(5\) different-width regions, i.e.,
-\(a_1\) equals to \(\{2, -2, 5, -10, 0.5\}\) in the intervals defined
-by the sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\).
+Here, the black-box function is \(f(\mathbf{x}) = a_1x_1 + x_1x_2\),
+with \(5\) different-width regions, i.e., \(a_1\) equals to
+\(\{2, -2, 5, -10, 0.5\}\) in the intervals defined by the sequence
+\(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). We are interested in the effect of
+\(x_1\) which is \(f^{\mathcal{GT}}_{\mu}(x_1) = a_1x_1\) in the
+regions defined above and the uncertainty is
+\(f^{\mathcal{GT}}_{\sigma}(x_1) = \sigma_2 x_1\).
 
@@
 
-% We observe that in
-% this case all intervals are multiples of \(0.05\). As we observe, in
-% Figure~\ref{fig:ex-synth-2-1}, UALE's approximation is better than all
-% fixed-size alternatives (for all \(K\)), in both mean effect
-% \(\mathcal{L}^{\mu}\) and uncertainty \(\mathcal{L}^{\sigma}\). For
-% understanding the importance of variable-size splitting, we analyze
-% the fixed-bin error. In the top-right of
-% Figure~\ref{fig:ex-synth-2-1}, we observe a positive trend between
-% \(\mathcal{L}^{\mu}\) and \(K\), because as \(K\) becomes larger fewer
-% points contribute to the bin estimations, leading to a worse
-% approximation. The interpretation of \(\mathcal{L}^{\sigma}\) is more
-% complex. For \(K\) that are multiples of \(20\), i.e.
-% \(K=\{20, 40, 60, 80, 100\}\), the mean error term
-% \(\mathcal{L}^{\rho}\) is low as we see in the bottom-right of
-% Figure~\ref{fig:ex-synth-2-2}. In these cases, we also observe in the
-% bottom-left figure, that \(L\) is increasing with \(K\) due to fewer
-% points in each bin. For all other values of \(K\), the \(L\) is mainly
-% affected by the mean error term \(\mathcal{E}\).
-
 \begin{figure}[h]
@@
   \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.pdf}
-  \caption{Figure 1}
+  \caption{Bin-Splitting, piecewise-linear function: UALE's approximation
+    (Top-Left). UALE vs fixed-size approximations in terms of:
+    \(\mathcal{L}^{\mu}\) (Top-Right), \(\mathcal{L}^{\sigma}\)
+    (Bottom-Left), \(\mathcal{L}^{\rho}\) (Bottom-Right).}
   \label{fig:ex-synth-2-1}
@@
 
-In this set-up, \(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), so the
-effect is non-linear in \(0 \leq x_1 \leq 1\). Due to this non
-linearity, wide bins increase \(\mathcal{L}^{\rho}\) making the
-uncertainty approximation biased. On the other hand, narrow bins lead
-to worse approximation due to the limited number of
-samples. Interestingly, in Figure~\ref{fig:ex-synth-2-2}, we observe
-that UALE's bin splitting manages to compromise these conflicting
-objectives.
+Here, the black-box function is
+\(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), so the effect is
+\(f^{\mathcal{GT}}_{\mu}(x_1) = 4 x_1^2\) and the uncertainty
+\(f^{\mathcal{GT}}_{\sigma}(x_1) = \sigma_2 x_1\). Due to the
+non-linear effect, wide bins increase the mean residual error,
+\(\mathcal{L}^{\rho}\) making the uncertainty approximation
+biased. This can be clearly observed in in the bottom-right of
+Figure~\ref{fig:ex-synth-2-2}. On the other hand, narrow bins lead to
+worse approximation due to the limited number of
+samples. Interestingly, UALE's bin splitting manages to compromise
+these conflicting objectives and provide an accurate estimation, as we
+see in the bottom-left of Figure~\ref{fig:ex-synth-2-2}.
 
@@
   \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_4.pdf}
-  \caption{Figure 1}
+  \caption{Bin-Splitting, non-linear function: UALE's approximation
+    (Top-Left). UALE vs fixed-size approximations in terms of:
+    \(\mathcal{L}^{\mu}\) (Top-Right), \(\mathcal{L}^{\sigma}\)
+    (Bottom-Left), \(\mathcal{L}^{\rho}\) (Bottom-Right).}
   \label{fig:ex-synth-2-2}
@@
 
-We exclude instances with missing and outlier values. As outlier we
-define the feature values which ar over three standard deviations away
-from the mean feature value. We also normalize all features to
-zero-mean and unit standard deviation. We split the dataset into
-\(N_{tr} = 15639\) training and \(N_{test} = 3910\) test examples
-(80/20 split) and we fit a Neural Network with 3 hidden layers of 256,
-128 and 36 units respectively. After 15 epochs using the Adam
-optimizer with learning rate \(\eta = 0.02\), the model achieves a MAE
-of \(37 \cdot 10^3\) dollars.
+We exclude instances with missing or outlier values and we normalize
+all features to zero-mean and unit standard deviation. We split the
+dataset into \(N_{tr} = 15639\) training and \(N_{test} = 3910\) test
+examples (80/20 split) and we fit a Neural Network with 3 hidden
+layers of 256, 128 and 36 units respectively. After 15 epochs using
+the Adam optimizer with learning rate \(\eta = 0.02\), the model
+achieves a MAE of \(37 \cdot 10^3\) dollars.
 
@@
   \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_pdp_ice.pdf}\\
-  \caption{Figure 1}
+  \caption{Real-dataset, UALE (left column) vs PDP-ICE (right column)
+    for features \(x_2, x_6, x_8\) (top to bottom).}
   \label{fig:ex-real-1}
@@
 
-We evaluate the robustness of UALE approximation. Following the
-evaluation framework of Section~\ref{sec:simulation-examples-2}, we
+We evaluate the robustness of UALE approximation, following the
+evaluation framework of Section~\ref{sec:simulation-examples-2}. We
 treat as ground-truth the effects computed on the full training-set
@@
 approximation with all possible fixed-size approximation. We repeat
-this process \(t=30\) times. In Figure~\ref{fig:ex-real-2}, we
-illustrate the mean values of
-\(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\) across the 30
-repetitions. We observe that UALE achieves (near) optimal
-approximation in all cases. \todo{add explanation for near-constant
-  feature effect vs the first case that the plot }
+this process for \(t=30\) independent runs and we report the mean
+values for \(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\) accross the
+repetitions. In Figure~\ref{fig:ex-real-2}, we observe that UALE
+achieves accurate approximations in all cases; the MAE of the
+uncertainty \(\mathcal{L}_\sigma\) is (near) the best of all
+fixed-size approximations.
 
@@
   \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_7.pdf}\\
-  \caption{Figure 1}
+  \caption{Real-dataset, UALE vs fixed-size approximation in terms of
+    \(\mathcal{L}^{\mu}\) (left column), \(\mathcal{L}^{\sigma}\)
+    (right column) for features \(x_2, x_6, x_8\) (top to bottom).}
   \label{fig:ex-real-2}

### 2022-10-13 AISTATS2023/paper.tex
latest

@@
 uncertainty. Based on this, we present an optimization method to
-determine an appropriate variable-width partitioning.
+determine an appropriate variable-width
+partitioning. Figure~\ref{fig:UALE-figure} illustrates an UALE
+visualization. The upper plot shows the ALE curve and its uncertainty
+as the shaded area around the curve. The lower plot shows the bins
+obtained after UALE's partitioning and the effects and the uncertainty
+per bin.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.5\textwidth]{concept_figure/fig-1.pdf}
+  \caption{Illustration of UALE feature effect plot.}
+  \label{fig:UALE-figure}
+\end{figure}
 
@@
 and the accumulated uncertainty, as defined in
-Eq.~(\ref{eq:ALE_acc_unc}). Figure~\ref{fig:UALE-figure} illustrates
-an example \todo{Describe after putting the figure}.
-
-\begin{figure}
-  \centering
-  \caption{UALE-concept-figure}
-  \label{fig:UALE-figure}
-\end{figure}
+Eq.~(\ref{eq:ALE_acc_unc}). Figure~\ref{fig:UALE-figure} illustrates a
+toy example, where \(x_1 \sim \mathcal{U}(0,1)\) and
+\(x_2 \sim \mathcal{N}(0,\sigma^2_2)\) and the black-box function is
+\(f(\xb)= \sin(2 \pi x_1) + x_1x_2\).
 
@@
 
-In this simulation, we focuse on illustrating the advantages of UALE's
-variable size bin-splitting against fixed-size bin-splitting.
+In this simulation, we focus on illustrating the advantages of UALE's
+variable-size bin-splitting against fixed-size one.
 
@@
 The purpose of this experiment is two-fold. First, we show that UALE
-automatically provides an approximation that is as accurate (or even
-more accurate) than the best fixed-size approximation, among all
-possible \(K\). Second, we illustrate that UALE favors the creation of
-wide intervals when they are not adding cumulative bin-error
-\(\mathcal{E}^2_{\mathcal{Z}}\). Wide intervals facilitate the
-interpretation of the average effect and its uncertainty.
+provides an approximation which is close to the best fixed-size
+approximation, among all possible \(K\). Second, we illustrate that
+UALE favors the creation of wider intervals when possible to
+facilitate the interpretation of the average effect and its
+uncertainty.
 
-For measuring the accuracy of the approximation, we generate a very
-big dataset with dense sampling (\(N=10^6\)) and we use dense
-fixed-size bins (\(K=10^3\)). We treat this estimation as the
-ground-truth UALE. Afterwards, we generate fewer samples (\(N=500\))
-and we compare the fixed-size estimation (for several \(K\)) against
-UALE's. We sample from \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
+To measure the accuracy of the approximation, we generate a dataset
+with dense sampling (\(N=10^6\)) and we use dense fixed-size bins
+(\(K=10^3\)). We treat this estimation as the ground-truth
+UALE. Afterwards, we generate fewer samples (\(N=500\)) and we compare
+the fixed-size estimation (for several \(K\)) against UALE's.
+
+We sample from \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
 \(x_1 \sim \mathcal{U}(0,1)\) and
-\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We, also, denote as
-\(\mathcal{Z^*} = \{z^*_0, \cdots, z^*_K\}\) the sequence obtained by
-automatic bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the
-fixed-size splitting with \(K\) bins. The evaluation is based on the
-Mean Absolute Error (MAE) in estimating the bin effect \(\mu\) and the
-uncertainty \(\sigma\) accross bins, i.e.,
+\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
+\(\mathcal{Z^*} = \{z^*_0, \cdots, z^*_K\}\) UALE's partitioning, and
+with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size splitting with \(K\)
+bins. The evaluation is based on the Mean Absolute Error (MAE) of the
+bin effect \(\mu\) and of the uncertainty \(\sigma\) accross bins,
+i.e.,
 
@@
   \label{eq:eval_met_1}
-  \mathcal{L}^{\mu} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
+  \mathcal{L}^{\mu} = \frac{1}{|\mathcal{Z}| - 1} \sum_{k \in
   \mathcal{Z}} | \mu(z_{k-1}, z_k) - \hat{\mu}(z_{k-1}, z_k) |
@@
   \label{eq:eval_met_2}
-  \mathcal{L}^{\sigma} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
+  \mathcal{L}^{\sigma} = \frac{1}{|\mathcal{Z}| -1} \sum_{k \in
     \mathcal{Z}} | \sigma(z_{k-1}, z_k) - \hat{\sigma}(z_{k-1}, z_k) |
@@
 
-We obtaint the ground truth bin-effect, \(\mu(z_{k-1}, z_k)\), and the
+We obtain the ground truth bin-effect, \(\mu(z_{k-1}, z_k)\) and the
 ground truth uncertainty, \(\sigma(z_{k-1}, z_k)\), using the average
 of the dense fixed-size bins that are within in the interval
-\([z_{k-1}, z_k]\). We also provide
+\([z_{k-1}, z_k]\). We also provide the mean residual error
 \(\mathcal{L}^{\rho} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
   \mathcal{Z}} \mathcal{E}(z_{k-1}, z_k) \) for facilitating the
-interpretation of the uncertainty error, and we call it mean residual error.
+interpretation of the uncertainty error.
 
@@
 \(\{2, -2, 5, -10, 0.5\}\) in the intervals defined by the sequence
-\(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). We are interested in the effect of
-\(x_1\) which is \(f^{\mathcal{GT}}_{\mu}(x_1) = a_1x_1\) in the
-regions defined above and the uncertainty is
-\(f^{\mathcal{GT}}_{\sigma}(x_1) = \sigma_2 x_1\).
+\(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). We are interested in the effect and
+the uncertainty of \(x_1\) which are
+\(f^{\mathcal{GT}}_{\mu}(x_1) = a_1x_1\) in the regions defined above
+and \(f^{\mathcal{GT}}_{\sigma}(x_1) = \sigma_2 x_1\), respectively.
 
 As we observe in the top left of Figure~\ref{fig:ex-synth-2-1}, UALE's
-bin-splitting separates the fine-grained bins, e.g.~regions
-\([0.4, 0.45]\), \([0.45, 0.5]\), and unites (most) constant-effect
+bin-splitting separates the fine-grained bins, e.g., intervals
+\([0.4, 0.45]\), \([0.45, 0.5]\), and unites (most of) constant-effect
 regions into a single bin, e.g.~region \([0.5, 1]\). This explains the
@@
 Here, the black-box function is
-\(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), so the effect is
-\(f^{\mathcal{GT}}_{\mu}(x_1) = 4 x_1^2\) and the uncertainty
+\(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), so for \(x_1\) the effect
+is \(f^{\mathcal{GT}}_{\mu}(x_1) = 4 x_1^2\) and the uncertainty is
 \(f^{\mathcal{GT}}_{\sigma}(x_1) = \sigma_2 x_1\). Due to the
-non-linear effect, wide bins increase the mean residual error,
-\(\mathcal{L}^{\rho}\) making the uncertainty approximation
+non-linear effect, wide bins (low \(K\)) increase the mean residual
+error, \(\mathcal{L}^{\rho}\), making the uncertainty approximation
 biased. This can be clearly observed in in the bottom-right of
-Figure~\ref{fig:ex-synth-2-2}. On the other hand, narrow bins lead to
-worse approximation due to the limited number of
+Figure~\ref{fig:ex-synth-2-2}. On the other hand, narrow bins (high
+\(K\)) lead to worse approximation due to the limited number of
 samples. Interestingly, UALE's bin splitting manages to compromise
-these conflicting objectives and provide an accurate estimation, as we
+these conflicting objectives and provides an accurate estimation, as we
 see in the bottom-left of Figure~\ref{fig:ex-synth-2-2}.
@@
 In real-world datasets, it is infeasible to obtain the ground truth FE
-for evaluation. We observe in Figure \ref{fig:ex-real-1} that in broad
-terms UALE and PDP-ICE plots aggree on the average effect and the
-uncertainty. It is interesting to note that UALE has selected bins of
-varying size as the optimal partiotioning. This demonstrates an
-important advantage of UALE; through the bin splitting process we can
-identify wide intervals of the feature domain with similar level of
-uncertainty.
-
-% We selected the particular experiment, because, in
-% broad terms, UALE and PDP-ICE plots aggree in the estimation of the
-% average effect and uncertainty. Therefore, we can focus on juding the
-% quality of the information provided by the two methods.
-
-% In Figure\ref{fig:ex-real-1}, we observe, from top to bottom, the
-% effects for the latitude, population and the median income. The effect
-% of UALE and PDP-ICE are similar for the population and the median
-% income. The population has a negative impact that progressively
-% decreases: from 400 to 1500 people the house value decreases with a
-% rate of \(-150 (\pm 140)\) dollars per added person, a rate that
-% decreases from \(-80 (\pm 80)\) to \(-60 (\pm 60)\) dollars per added
-% person as we move from 1500 to 2800 people. The level of uncertainty
-% indicates significant variance in absolute value of the rate, but in
-% the grant majority of instances the rate is negative. With the same
-% inspection, we observe that the median annual income has a positive
-% impact on the value (all numbers are thousands of dollars):
-% \(20\pm 15\) per \(10\) of added median income for incomes in
-% \([8, 15]\), \(32 \pm 20\) per \(10\) added income in \([15, 60]\) and
-% \(40 \pm 15\) per \(10\) added income in \([60, 70]\). The uncertainty
-% indicates that there are less heterogeneous effects about the median
-% income compared to the number of people. In both cases, we can end-up
-% to the same conclusion by inspecting the PDP-ICE plots. For the
-% latitude, there is a small difference in the explanations for the
-% region \([32, 35]\), where UALE estimates a less negative slope with
-% less uncertainty than PDP, while the explanations are similar for the
-% range \([35,39]\), where both methods reveal an increase in the
-% uncertainty around the feature value \(37.5\).
-
-% In general, we observe that UALE complements ALE in the quantification
-% of the heterogeneous effects, similarly to as ICE complements PDP. The
-% automatic extraction of constant-effect and constant-uncertainty
-% regions provided by UALE is helpful for an easier interpretation.
-
-% On
-% the other hand, ICE plots sometimes are more descritive due to
-% illustrating the type of heterogeneity, whereas UALE quantifies only the
-% level (not the type) of heterogeneity.
+for evaluation. We observe in Figure~\ref{fig:ex-real-1} that UALE and
+PDP-ICE compute, roughly, a similar average effect and
+uncertainty. Note that when the average effect remains (almost)
+constant in a wide interval, i.e., the feature effect plot is linear,
+UALE creates a single bin for this interval, which facilitates the
+interpretation. This can be observed, in the bottom plots of
+Figure~\ref{fig:ex-real-1}, in the interval \(x_8 \in [1.5, 6]\).
 
@@
 evaluation framework of Section~\ref{sec:simulation-examples-2}. We
-treat as ground-truth the effects computed on the full training-set
-\(N=20000\) with dense fixed-size bin-splitting (\(K=80\)). Given the
-big number of samples, we make the hypothesis that the approximation
-with dense binning is close to the ground truth. Afterwards, we
-randomly select fewer samples, \(N=1000\), and we compare UALE
-approximation with all possible fixed-size approximation. We repeat
-this process for \(t=30\) independent runs and we report the mean
-values for \(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\) accross the
-repetitions. In Figure~\ref{fig:ex-real-2}, we observe that UALE
-achieves accurate approximations in all cases; the MAE of the
-uncertainty \(\mathcal{L}_\sigma\) is (near) the best of all
-fixed-size approximations.
+treat as ground-truth the effects computed on the full training set
+(\(N_{tr}=15639\)) with dense fixed-size bin-splitting
+(\(K=80\)). Given the sufficient number of samples, we make the
+hypothesis that the approximation with dense binning is close to the
+ground truth. Afterwards, we randomly select fewer samples,
+\(N=1000\), and we compare UALE approximation against fixed-size
+approximation (for all \(K\)). We repeat this process for \(t=30\)
+independent runs and we report the mean values of
+\(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\). In
+Figure~\ref{fig:ex-real-2}, we observe that UALE achieves accurate
+approximations in all cases; the MAE of the uncertainty
+\(\mathcal{L}^\sigma\) is close to the best of the fixed-size
+approximations.
 

### 2022-10-13 examples/concept_figure.py

@@ -1,0 +1,67 @@
+import examples.utils as utils
+path = utils.add_parent_path()
+import matplotlib.pyplot as plt
+import numpy as np
+import feature_effect as fe
+import example_models.distributions as dist
+import example_models.models as models
+
+import numpy as np
+
+
+
+def f(x):
+    y = np.zeros_like(x[:,0])
+
+    a = 1
+    ind = np.logical_and(x[:, 0] >= 0, x[:, 0] < .5)
+    y[ind] = np.sin(2*np.pi * a * x[ind, 0])
+
+
+    # ind = np.logical_and(x[:, 0] > 0.25, x[:, 0] < 1)
+    # y[ind] = - 2 * (x[ind, 0] - 0.25)
+
+    # ind = np.logical_and(x[:, 0] > 0.625, x[:, 0] <= 1)
+    # y[ind] = 3 * (x[ind, 0] - 0.625) - 0.75
+
+    y = y + x[:,0]*x[:,1]
+    return y
+
+
+def dfdx(x):
+    dydx = np.zeros_like(x)
+
+    a = 1
+    ind = np.logical_and(x[:, 0] >= 0, x[:, 0] <= .5)
+    dydx[ind, 0] = a * 2 * np.pi * np.cos(2*np.pi * a * x[ind, 0])
+
+    dydx[:,0] += x[:,1]
+    return dydx
+
+np.random.seed(seed=21)
+axis_limits = np.array([[0,1], [0,1]]).T
+
+x1 = np.linspace(0, 1, 1000)
+x2 = np.zeros_like(x1)
+x = np.stack([x1, x2], -1)
+
+# low sigma
+sigma = 1
+
+x1 = np.random.uniform(0, 1, 1000)
+x2 = np.random.normal(0, sigma, 1000)
+x = np.stack([x1, x2], -1)
+
+# dale1 = fe.DALE(data=x,
+#                model=f,
+#                model_jac=dfdx,
+#                axis_limits=axis_limits)
+# dale1.fit([0], alg_params = {"bin_method": "fixed", "nof_bins":50})
+# dale1.plot(block=False)
+
+dale2 = fe.DALE(data=x,
+               model=f,
+               model_jac=dfdx,
+               axis_limits=axis_limits)
+dale2.fit([0], alg_params = {"bin_method": "dp", "max_nof_bins":50})
+dale2.plot(title="UALE - Feature Effect Plot", savefig= os.path.join(os.getcwd(), "examples/concept_figure/fig-1.pdf"))

### 2022-10-13 feature_effect/dale.py

@@ -189,13 +189,14 @@ class DALE(FeatureEffectBase):
              block=False,
              gt=None,
              gt_bins=None,
-             savefig=False):
+             savefig=False,
+             title=None):
         vis.feature_effect_plot(self.feature_effect["feature_"+str(s)],
                                 self.eval,
                                 s,
                                 error=error,
                                 min_points_per_bin=self.feature_effect["feature_"+str(s)]["alg_params"]["min_points_per_bin"],
-                                title=None,
+                                title=title,
                                 block=block,
                                 gt=gt,
                                 gt_bins=gt_bins,

### 2022-10-13 feature_effect/visualization.py

@@ -118,7 +118,7 @@ def feature_effect(ax1, x, y, estimator_var, std, rem_eff, first_empty, limits, 
                    alpha=.7,
                    label="first bin with < " + str(point_limit) + " points")
     if error == "std":
-        ax1.fill_between(x, y-std, y+std, color='red', alpha=0.2, label="$\hat{f}_{\sigma^2}$")
+        ax1.fill_between(x, y-std, y+std, color='red', alpha=0.2, label="$\hat{f}_{\sigma}$")
         # ax1.fill_between(x, y-rem_eff, y+rem_eff, color='blue', alpha=0.7, label="std")
     elif error == "standard error":
         ax1.fill_between(x, y-2*np.sqrt(estimator_var), y+2*np.sqrt(estimator_var), color='red', alpha=0.6, label="standard error")
@@ -149,7 +149,7 @@ def effects_per_bin(ax2, bin_effects, bin_variance, error, is_bin_empty, limits,
                            edgecolor='blue',
                            yerr=np.sqrt(bin_variance[is_bin_full]),
                            ecolor='red',
-                           label="$\hat{\mu}_k ( N_k \geq " + str(point_limit) + ")$")
+                           label="$\hat{\mu}_k$")
             # ax2.bar_label(bars, labels=['%.1f' % e for e in np.sqrt(bin_variance[is_bin_full])])
         # bins without enough points
         if np.sum(is_bin_empty):
@@ -169,7 +169,7 @@ def effects_per_bin(ax2, bin_effects, bin_variance, error, is_bin_empty, limits,
                     color=(0.1, 0.1, 0.1, 0.1),
                     edgecolor='blue',
                     ecolor='red',
-                    label="$\hat{\mu}_k ( N_k \geq " + str(point_limit) + ")$")
+                    label="$\hat{\mu}_k$")
         if np.sum(is_bin_empty):
             ax2.bar(x=bin_centers[is_bin_empty],
                     height = bin_effects[is_bin_empty],

### 2022-10-13 AISTATS2023/paper.tex
latest

@@
 The objective \(\mathcal{L}\) searches for a partitioning
-\(\mathcal{Z}_* = \{ z_0, \ldots, z_K \} \) with low aggregated error
+\(\mathcal{Z}^* = \{ z_0^*, \ldots, z_K^* \} \) with low aggregated error
 \(\mathcal{E}_{\mathcal{Z}}^2\). In case of many partitionings with
@@
 K_{max}\} \). In this discretized solution space, we find the global
-optimum using dynamic programming. \todo{Add description} See also
-Appendix~\ref{sec:dynamic-programming}.
+optimum using Dynamic Programming.
+
+To define the solution, we use two indexes. The index
+\(i \in \{0, \ldots, K_{max}\}\) denotes the points of the
+partitioning \((z_i)\) and the index \(j \in \{0, \ldots, K_{max}\} \)
+denotes the \(j\)-th multiple of the minimum step, i.e.,
+\(x_j = x_{s,min} + j \cdot \Delta x_{min}\). The recursive cost
+function \(T(i,j)\) computes the cost of setting \(z_i\) to \(x_j\):
+\begin{equation}
+  \label{eq:recursive_cost}
+  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
+\end{equation}
+%
+% where \(\mathcal{T}(0,j)\) equals to zero if \(j=0\) and \(\infty\) in
+% any other case.
+The term \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin with
+limits \([x_l, x_j)\):
+
+\begin{equation}
+  \label{eq:cost_step}
+  \mathcal{B}(x_l, x_j) = \begin{cases}
+                            \infty, & \text{if $x_j > x_l$ or \(|\mathcal{S}_{(x_j, x_l)}| < N\)}\\
+                            0, & \text{if $x_j = x_l$}\\
+                            \hat{\sigma}^2(x_j, x_l), &\text{if $x_j \leq x_l$}
+  \end{cases}
+\end{equation}
+
+The optimal solution is given by solving
+\(\mathcal{L} = \mathcal{T}(K_{max}, K_{max})\) and keeping track of the sequence of
+steps.
 
@@
 
-\section{DISCUSSION}
+\section{CONCLUSION AND FURTHER WORK}
 
@@
 
+\clearpage
 \subsubsection*{Acknowledgments}

### 2022-10-13 AISTATS2023/paper.tex
added discussion

@@
 
-\todo{Add Conclusion, limitations, and future work}
+This paper introduced UALE, a method that extends ALE, for not only quantifying the global feature effect but also its uncertainty, using the standard deviation of instance-level effects.
+We showed that UALE is more accurate than existent methods, e.g.
+ICE plots, in modeling the uncertainty in cases of correlated features.
+However, it is not trivial to make an unbiased estimate of the uncertainty.
+Unlike ALE, where for estimating the average effect, splitting the feature domain in \(K\) equal-width bins is sufficient, we theoretically proved that, for estimating the uncertainty, the same approach would introduce bias.
+We overcame this limitation by proposing an algorithm for automatically determining the optimal variable-size partitioning considering the instance-level effects.
+Through, synthetic and real-world examples, we show that UALE is an effective method for modeling, both, the average effect and the uncertainty.
+
+There are, though, limitations and open issues.
+First, for efficiently finding the optimal bin partitioning, UALE relies on DALE's approximation, so it is directly applicable only to differentiable models.
+Furthermore, UALE quantifies only the standard deviation of instance-level effects, providing limited intuition about the sources of the uncertainty, i.e., which features and to what degree they contribute to the heterogeneity of instance-level effects.
+Future work could target that.
 

### 2022-10-14 AISTATS2023/paper.tex
latest

@@
 In case of complex interactions between features, instance-level feature effects may deviate from the ALE curve.
-It is therefore crucial to quantify this deviation, namely, the uncertainty of the effect.
+It is crucial to quantify this deviation, namely, the uncertainty of the global effect.
 In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and its uncertainty.
@@
 We formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints.
-Based on this, we propose a method to determine the optimal intervals, \todo{balancing the estimation bias and variance.} We demonstrate, through synthetic and real datasets, (a) the advantages of modeling the uncertainty with UALE compared to alternative methods and (b) the effectiveness of UALE's appropriate bin splitting for a \todo{good} approximating of the average effect and its uncertainty.
+Based on this, we propose a method to determine the optimal intervals, balancing the estimation bias and variance.
+We demonstrate, through synthetic and real datasets, (a) the advantages of modeling the uncertainty with UALE compared to alternative methods and (b) the effectiveness of UALE's appropriate bin splitting for accurately approximating of the average effect and its uncertainty.
 \end{abstract}
@@
 
-Recently, Machine Learning (ML) has been widely adopted in mission critical domains,
-such as healthcare and finance.
+Recently, Machine Learning (ML) has been widely adopted in mission critical domains, such as healthcare and finance.
 In such high-stakes areas, it is important not only to provide accurate predictions but also meaningful explanations.
-For this reason, there is an increased interest in Explainable
-AI (XAI). XAI literature distinguishes between local and global
-methods~\citep{Molnar2020interpretable}. Local methods provide
-instance-level explanations, i.e., explain the prediction for a
-specific instance, whereas global methods summarize the entire model
-behavior. Global methods create a global explanation by aggregating the
-instance-level explanations into a single interpretable outcome,
-usually a number or a plot.
+For this reason, there is an increased interest in Explainable AI (XAI).
+XAI literature distinguishes between local and global methods~\citep{Molnar2020interpretable}.
+Local methods provide instance-level explanations, i.e., explain the prediction for a specific instance, whereas global methods summarize the entire model behavior, aggregating the instance-level explanations into a single interpretable outcome, usually a number or a plot.
 
-A popular class of global XAI are feature effect (FE)
-methods~\citep{Gromping2020MAEP} that quantify the average (over all
-instances) effect of a single feature on the output\footnote{FE
-  methods also isolate the combined effect of a pair of features to
-  the output. Combinations of more than two features are uncommon,
-  since they are difficult to estimate and visualize.}. There are
-three widely-used FE methods: \emph{Partial Dependence Plots}
-(PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
-(MP)\citep{apley2020visualizing} and \emph{Accumulated Local Effects}
-(ALE)\citep{apley2020visualizing}. ALE is established as the
-state-of-the-art in quantifying the average effect, since PDP and MP
-have been criticized~\citep{Gromping2020MAEP} of providing misleading explanations in case of  correlated input features.
+A popular class of global XAI are feature effect (FE) methods~\citep{Gromping2020MAEP} that quantify the average (across all instances) effect of a single feature on the output.
+There are three widely-used FE methods: \emph{Partial Dependence Plots} (PDP)~\citep{friedman2001greedy}, \emph{Marginal Plots} (MP)~\citep{apley2020visualizing} and \emph{Accumulated Local Effects} (ALE)~\citep{apley2020visualizing}.
+ALE is established as the state-of-the-art, since PDP and MP have been criticized~\citep{baniecki2021fooling, Gromping2020MAEP} of providing misleading explanations in case of correlated input features.
 
-When complex interactions between features exist, the instance-level
-(local) feature effects may exhibit significant deviation, aka heterogeneity, from the aggregated (global)
-explanation, a phenomenon known as aggregation
-  bias~\citep{mehrabi2021survey}. We call this deviation \emph{the uncertainty of the global effect}. Aggregation bias leads to
-ambiguities; for example, a feature with zero average effect may
-indicate (a) no effect on the output or (b) an effect that is highly
-positive for some instances and highly negative for some others. As a
-result, there is a need for FE methods to quantify the deviation (uncertainty) of
-instance-level explanations, in addition to the average effect. In the
-case of PDP, such deviation is visualized via the Individual
-Conditional Explanations (ICE)~\citep{goldstein2015peeking}. ICE
-plots, however, have the same limitations as PDP in case of correlated
-features. No method to quantify the deviation of instance-level effects has been proposed for ALE.
+In case of complex interactions between features, the instance-level (local) feature effects may exhibit significant deviation, aka heterogeneity, from the aggregated (global) explanation, a phenomenon known as aggregation bias~\citep{mehrabi2021survey}.
+We call this deviation \emph{the uncertainty of the global effect}\footnote{In the paper, the \emph{uncertainty of the global effect} and the \emph{heterogeneity of instance-level effects} are used interchangeably and refer to the deviation of instance-level (local) effects from the averaged feature effect curve (global).}.
+Aggregation bias leads to ambiguities.
+For example, a feature with zero average effect may indicate (a) no effect on the output or (b) highly positive effect for some instances and highly negative for some others.
+As a result, there is a need for FE methods to quantify the deviation of instance-level explanations, in addition to the average effect.
+In the case of PDP, such deviation is visualized via the Individual Conditional Explanations (ICE)~\citep{goldstein2015peeking}.
+ICE plots, however, have the same limitations as PDP in case of correlated features.
+No method to quantify the deviation of instance-level effects has been proposed for ALE.
 
-In this work, we propose UALE, a global feature-effect method that
-extends ALE for modeling the uncertainty of the feature effect.
+In this work, we propose UALE, a global feature-effect method that extends ALE for modeling the uncertainty of the feature effect.
 Our method handles well cases with correlated features.
 As in ALE, we provide an interval-based formulation of UALE.
-This formulation partitions the feature domain into non-overlapping 
-intervals (bin-splitting) and defines the average effect (bin-effect) and its uncertainty (bin-uncertainty). We exploit this formulation to provide an UALE approximation. 
-As we formally prove, a fixed equi-width partitioning that does not consider
-the characteristics of the instance-level effects, which is the case in ALE, 
-may lead to an erroneous (biased) estimation of the uncertainty. 
+This formulation partitions the feature domain into non-overlapping intervals (bin-splitting) and defines the average effect (bin-effect) and its uncertainty (bin-uncertainty).
+We use this formulation to provide an UALE approximation.
+As we formally prove, a fixed equi-width partitioning that does not consider the characteristics of the instance-level effects, which is the case in ALE approximation, may lead to an erroneous (biased) estimation of the uncertainty.
 Therefore, we propose a variable-width partitioning that leads to an unbiased approximation.
+The contributions of this paper are:
 
-The contributions of this
-paper are:
 \begin{itemize}
-\item A global feature effect method (UALE) that quantifies both the
-  average effect and its uncertainty, i.e.,~the deviation of
-  instance-level effects.
-\item An unbiased approximation of the uncertainty through a dynamic
-  variable-size bin partitioning of the feature domain.
+\item A global feature effect method (UALE) that quantifies both the average effect and its uncertainty, i.e.,~the heterogeneity of instance-level effects.
+\item An unbiased approximation of the uncertainty through an appropriate variable-size bin partitioning of the feature domain.
 \item The evaluation of UALE on both synthetic and real datasets.
@@
 
-In the rest of the paper, we use following notation. Let \(\mathcal{X} \in \mathbb{R}^d\) be the \(d\)-dimensional feature
-space, \(\mathcal{Y} \in \mathbb{R}\) the target space and
-\(f(\cdot) : \mathcal{X} \rightarrow \mathcal{Y}\) the black-box
-function. We use index \(s \in \{1, \ldots, d\}\) for the feature of
-interest and \(c = \{1, \ldots, d\} - s\) for the set with all other
-indexes. For convenience, we denote the feature vector
-\(\xb = (x_1, \cdots , x_s, \cdots, x_D)\) with \((x_s, \xc)\) and the
-corresponding random variables
-\(X = (X_1, \cdots , X_s, \cdots, X_D)\) with \((X_s, X_c)\). The
-training set is \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) sampled
-i.i.d. from the distribution \(\mathbb{P}_{X,Y}\). Finally, we use
-\(f^{\mathtt{<method>}}(x_s)\) for denoting the \(s\)-th FE, where
-\(\mathtt{<method>}\) indicates the particular method in use, for
-example ALE. An extensive list with all symbols used in this paper is
-provided at Appendix~\ref{sec:list-with-symbols}.
+In the rest of the paper, we use following notation.
+Let \(\mathcal{X} \in \mathbb{R}^d\) be the \(d\)-dimensional feature space, \(\mathcal{Y} \in \mathbb{R}\) the target space and \(f(\cdot) : \mathcal{X} \rightarrow \mathcal{Y}\) the black-box function.
+We use index \(s \in \{1, \ldots, d\}\) for the feature of interest and \(c = \{1, \ldots, d\} - s\) for the set with all other indexes.
+For convenience, we denote the feature vector \(\xb = (x_1, \cdots , x_s, \cdots, x_D)\) with \((x_s, \xc)\) and the corresponding random variables \(X = (X_1, \cdots , X_s, \cdots, X_D)\) with \((X_s, X_c)\).
+The training set is \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) sampled i.i.d.
+from the distribution \(\mathbb{P}_{X,Y}\).
+Finally, we use \(f^{\mathtt{<method>}}(x_s)\) for denoting the effect of the \(s\)-th, where \(\mathtt{<method>}\) indicates the particular method in use, e.g., ALE.
 
@@
 \label{sec:feat-effect-meth}
-The three well-known feature effect methods are: \emph{Partial
-  Dependence Plots} (PDP), \emph{Marginal Plots} (MP) and
-\emph{Accumulated Local Effects} (ALE).  PDP formulates the global FE
-as an expectation over the distribution of \(X_c\), i.e.,
-\(f^{\mathtt{PDP}}(x_s) = \mathbb{E}_{X_c}[f(x_s,X_c)]\), whereas MP
-as an expectation over the distribution of \(X_c|X_s\), i.e.,
-\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{X_c|x_s}[f(x_s, X_c)]\). Both
-methods suffer from misestimations in case of correlated features. PDP
-integrates over unrealistic instances and MP computes aggregated
-effects, i.e., imputes the combined effect of sets of features to a
-single feature~\citep{apley2020visualizing}.
 
-ALE overcomes these problems. Specifically, ALE defines the local
-effect of the \(s\)-th feature at a specific point \((z, \xc)\) of the
-input space \(\mathcal{X}\) with the partial derivative
-\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
-local explanations at \(z\) are then weighted by the conditional
-distribution \(p(\xc|z)\) and are averaged, to produce the averaged
-effect \(\mu(z)\). ALE plot is the accumulation of the averaged local
-effects:
+The three well-known feature effect methods are: \emph{Partial Dependence Plots} (PDP), \emph{Marginal Plots} (MP) and \emph{Accumulated Local Effects} (ALE).
+PDP formulates the global FE as an expectation over the distribution of \(X_c\), i.e., \(f^{\mathtt{PDP}}(x_s) = \mathbb{E}_{X_c}[f(x_s,X_c)]\), whereas MP as an expectation over the distribution of \(X_c|X_s\), i.e., \(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{X_c|x_s}[f(x_s, X_c)]\).
+Both methods suffer from misestimations when features are correlated.
+PDP integrates over unrealistic instances and MP computes aggregated effects, i.e., imputes the combined effect of sets of features to a single feature~\citep{apley2020visualizing}.
+
+ALE overcomes these problems.
+Specifically, ALE defines the local effect of the \(s\)-th feature at a specific point \((z, \xc)\) of the input space \(\mathcal{X}\) with the partial derivative \(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\).
+All the local explanations at \(z\) are then weighted by the conditional distribution \(p(\xc|z)\) and are averaged, to produce the averaged effect \(\mu(z)\).
+ALE plot is the accumulation of the averaged local effects:
 
@@
 
-The global effect is computed as an expectation over local
-(instance-level) effects. In addition to this global effect, it is
-important to know to what extent the local effects deviate from the
-global explanation.
+The global effect is computed as an expectation over local (instance-level) effects.
+In addition to this global effect, it is important to know to what extent the local effects deviate from the global explanation.
 
-ICE plots and similar methods (e.g., d-ICE
-plots~\citep{goldstein2015peeking}) provide a set of curves
-illustrated on top-of PDP. Each curve corresponds to one instance of
-the dataset, \(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\). The user
-then visually observes if the curves are homogeneous, i.e., all
-instances have similar effect plots, and to what extent they deviate
-from the PDP. There are methods try to automate the aforementioned
-visual exploration, by grouping ICE plots into
-clusters~\citep{molnar2020model, herbinger2022repid,
-  britton2019vine}. Unfortunately, these methods are subject to the
-failure modes of PDPs in cases of correlated
-features~\citep{baniecki2021fooling}, as we also confirm in our
-experimental analysis (c.f., Section~\ref{sec:simulation-examples-1}).
+ICE plots and similar methods (e.g., d-ICE plots~\citep{goldstein2015peeking}) provide a set of curves illustrated on top-of PDP.
+Each curve corresponds to one instance of the dataset, \(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\).
+The user then visually observes if the curves are homogeneous, i.e., all instances have similar effect plots, and to what extent they deviate from the PDP.
+Some approaches~\citep{herbinger2022repid, britton2019vine, molnar2020model} automate the aforementioned visual exploration, by grouping ICE plots into clusters.
+Unfortunately, these methods are subject to the failure modes of PDPs in cases of correlated features.
 
-Other approaches, like H-Statistic \citep{friedman2008predictive},
-Greenwell's interaction index \citep{greenwell2018simple} or SHAP
-interaction values \citep{lundberg2018consistent} quantify
-interactions between the input features. A strong interaction is
-  an implicit indication of the existence of heterogeneous effects. These
-  methods, however, do not quantify directly the level of heterogeneity.
+Other approaches, like H-Statistic \citep{friedman2008predictive}, Greenwell's interaction index \citep{greenwell2018simple} or SHAP interaction values \citep{lundberg2018consistent} quantify interactions between the input features.
+A strong interaction is an implicit indication of the existence of heterogeneous effects.
+These methods, however, do not quantify directly the level of heterogeneity.
 
-To the best of our knowledge, no work exist so far that quantifies the
-heterogeneous effects (uncertainty) based on the formulation of ALE.
+To the best of our knowledge no work exist so far that quantifies the heterogeneous effects (uncertainty) based on the formulation of ALE.
 
@@
 
-ALE is estimated from the available dataset instances. \citep{apley2020visualizing} proposed to divide the feature
-domain in \(K\) bins of equal size and to estimate the local effects
-in each bin by evaluating the model \(f\) at the bin limits:
+ALE is estimated from the available dataset instances.
+\citep{apley2020visualizing} proposed to divide the domain of the \(s\)-th feature in \(K\) equal-size bins and estimate the bin-effects from the instances that lie in each bin after setting the \(s\)-th feature at the bin limits:
 
@@
 %
-where \(k_x\) the index of the bin that \(x_s\) belongs to,
-i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
-the set of training instances of the \(k\)-th bin, i.e.
-\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
-\). \citep{gkolemis22} proposed the Differential ALE (DALE) that
-computes the local effects on the training instances using
-auto-differentiation:
+where \(k_x\) the index of the bin such that \(z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is the set of the instances of the \(k\)-th bin, i.e.
+\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \} \).
+\citep{gkolemis22} proposed the Differential ALE (DALE) that computes the local effects on the training instances using auto-differentiation:
+
 \begin{equation}  \label{eq:DALE_accumulated_mean_est}
@@
 %
-Their method has significant computational advantages and allows the
-recomputation of the accumulated effect with different bin-splitting
-with near-zero additional computational cost. Moreover, by avoiding
-the use of artificial samples at the bin limits DALE allows the use of
-wider bins without out-of-distribution sampling. \todo{Maybe connect with our work}. In both cases, the
-approximations partition the feature domain in \(K\) equally-sized
-bins, without considering the underlying local effects.
+Their method has significant computational advantages and allows the recomputation of the accumulated effect with a modified bin-splitting with near-zero additional computational cost.
+Moreover, by avoiding the use of artificial samples at the bin limits DALE allows the use of wider bins without out-of-distribution sampling.
+In both cases, the approximations partition the feature domain in \(K\) equally-sized bins, without considering the underlying local effects.
 
@@
 
-We define UALE and provide an interval-based formulation that
-partitions the feature domain into non-overlapping intervals
-(bin-splitting) and defines the bin-effect and the
-bin-uncertainty. Then, we present UALE approximation, formally proving
-the bin-splitting requirements for an unbiased estimation of the
-uncertainty. Based on this, we present an optimization method to
-determine an appropriate variable-width
-partitioning. Figure~\ref{fig:UALE-figure} illustrates an UALE
-visualization. The upper plot shows the ALE curve and its uncertainty
-as the shaded area around the curve. The lower plot shows the bins
-obtained after UALE's partitioning and the effects and the uncertainty
-per bin.
+We define UALE and provide an interval-based formulation that partitions the feature domain into non-overlapping intervals (bin-splitting) and defines the bin-effect and the bin-uncertainty.
+Then, we present UALE approximation, formally proving the bin-splitting requirements for an unbiased estimation of the uncertainty.
+Based on this, we present an optimization method to determine an appropriate variable-width partitioning.
+Figure~\ref{fig:UALE-figure} illustrates an UALE visualization.
+The upper plot shows the ALE curve and its uncertainty as the shaded area around the curve.
+The lower plot shows the bins obtained after UALE's partitioning and the effects and the uncertainty per bin.
 
@@
 
-We quantify the uncertainty of the local effects at \(x_s=z\) with the
-standard deviation of the local explanations, \(\sigma(z)\), where:
+We quantify the uncertainty of the local effects at \(x_s=z\) with the standard deviation of the local explanations, \(\sigma(z)\), where:
 
@@
 \end{equation}
+
 \noindent
-The uncertainty emerges from (a) the feature correlations and (b) the
-implicit feature interactions of the black-box function. We also
-define the accumulated uncertainty at \(x_s\), as the accumulation of
-the standard deviation of the local effects:
+The uncertainty emerges from (a) the feature correlations and (b) the implicit feature interactions of the black-box function.
+We also define the accumulated uncertainty at \(x_s\), as the accumulation of the standard deviation of the local effects:
 
@@
 
-UALE defines the effect as a pair that consists of the average effect
-and the uncertainty \((\mu(z), \sigma(z))\). For visualizing it, we
-use the accumulated average effect, as defined in Eq.~(\ref{eq:ALE}),
-and the accumulated uncertainty, as defined in
-Eq.~(\ref{eq:ALE_acc_unc}). Figure~\ref{fig:UALE-figure} illustrates a
-toy example, where \(x_1 \sim \mathcal{U}(0,1)\) and
-\(x_2 \sim \mathcal{N}(0,\sigma^2_2)\) and the black-box function is
-\(f(\xb)= \sin(2 \pi x_1) + x_1x_2\).
+UALE defines the effect as a pair that consists of the average effect and the uncertainty \((\mu(z), \sigma(z))\).
+For visualizing it, we use the accumulated average effect, as defined in Eq.~(\ref{eq:ALE}), and the accumulated uncertainty, as defined in Eq.~(\ref{eq:ALE_acc_unc}).
+Figure~\ref{fig:UALE-figure} illustrates a toy example, where \(x_1 \sim \mathcal{U}(0,1)\) and \(x_2 \sim \mathcal{N}(0,\sigma^2_2)\) and the black-box function is \(f(\xb)= \sin(2 \pi x_1) + x_1x_2\).
 
-For the interval-based formulation, we define the bin-effect
-\(\mu(z_1, z_2)\) and the bin-uncertainty \(\sigma(z_1, z_2)\) as:
+For the interval-based formulation, we define the bin-effect \(\mu(z_1, z_2)\) and the bin-uncertainty \(\sigma(z_1, z_2)\) as:
 
@@
 %
-Intuitively, if we randomly draw a point \(z^*\) from a uniform
-distribution \(z^* \sim \mathcal{U}(z_1, z_2)\), the bin-effect
-(Eq.~(\ref{eq:mu_bin})) and the bin-uncertainty
-(Eq.~(\ref{eq:var_bin})) are the expected average effect and the
-expected uncertainty, respectively. Also, if we denote as
-\(\mathcal{Z}\) the sequence of \(K+1\) points that partition the
-domain of the \(s\)-th feature into \(K\) variable-size intervals,
-i.e.  \(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we provide the
-interval-based formulation of UALE:
+Intuitively, if we randomly draw a point \(z^*\) from a uniform distribution \(z^* \sim \mathcal{U}(z_1, z_2)\), the bin-effect (Eq.~(\ref{eq:mu_bin})) and the bin-uncertainty (Eq.~(\ref{eq:var_bin})) are the expected average effect and the expected uncertainty, respectively.
+Also, if we denote as \(\mathcal{Z}\) the sequence of \(K+1\) points that partition the domain of the \(s\)-th feature into \(K\) variable-size intervals, i.e.
+\(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we provide the interval-based formulation of UALE:
 
@@
 
-Here, we activate the interaction term, i.e., \(a=1\), keeping the
-weights equal \(a_1=a_2=1\) and \(\sigma_3^2=\frac{1}{4}\). In this
-case, it is not straightforward to define the ground-truth effect for
-features \(x_1, x_3\), because the interaction term provokes
-heterogeneous instance-level effects, i.e., the instance-level effect
-of \(x_1\) depend on the unknown value of \(x_3\) and vice-versa. We
-observe that UALE's feature effects are quite intuitive. The average
-effect of \(x_1\) is the same with the Example (a) with an added
-uncertainty, reflecting the uncertainty about the instance-level
-effects which is the standard deviation of \(x_3\), i.e.,
-\(\sigma_3 = \frac{1}{2}\). The effect of \(x_3\) is only due to the
-interaction term, therefore, the instance-level effects
-\(\frac{\partial f}{\partial x_3} = x_1 \) follow
-\(p(x_1) \sim \mathcal{U}(0,1)\) which has \(\mu_{x_1}=\frac{1}{2}\)
-and \(\sigma_{x_1}=\frac{1}{4}\). This is reflected, in UALE's
-estimations of Figure~\ref{fig:ex-synth-1-3}.
+Here, we activate the interaction term, i.e., \(a=1\), keeping the weights equal \(a_1=a_2=1\) and \(\sigma_3^2=\frac{1}{4}\).
+In this case, it is not straightforward to define the ground-truth effect for features \(x_1, x_3\), because the interaction term provokes heterogeneous instance-level effects, i.e., the instance-level effect of \(x_1\) depend on the unknown value of \(x_3\) and vice-versa.
+We observe that UALE's feature effects are quite intuitive.
+The average effect of \(x_1\) is the same with the Example (a) with an added uncertainty, reflecting the uncertainty about the instance-level effects which is the standard deviation of \(x_3\), i.e., \(\sigma_3 = \frac{1}{2}\).
+The effect of \(x_3\) is only due to the interaction term, therefore, the instance-level effects, \(\frac{\partial f}{\partial x_3} = x_1 \), follow \(p(x_1) \sim \mathcal{U}(0,1)\) which has \(\mu_{x_1}=\frac{1}{2}\) and \(\sigma_{x_1}=\frac{1}{4}\).
+This is reflected, in UALE's estimations of Figure~\ref{fig:ex-synth-1-3}.
+We note that PDP-ICE provide (a) for \(x_1\) a near-zero average effect plot, which is not intuitive and (b) for \(x_3\) a plot that, roughly, matches with UALE.
 
-For \(x_2\), as in Example (a), the effect is
-\(f_\mu^{\mathtt{GT}}(x_2)\) is: \(x_2\) in \([0, \frac{1}{4}]\),
-\(-x_2\) in \([\frac{1}{4}, \frac{1}{2}]\) and \(0\) in
-\([\frac{1}{2}, 1]\)\(x_2\) and it has zero-uncertainty,
-\(f_\sigma^{\mathtt{GT}}(x_2) = 0\), since \(x_2\) does not appear in
-any interaction term. We confirm that UALE computes it correctly
-whereas PDP-ICE fail in both the average effect and the uncertainty.
+For \(x_2\), as in Example (a), the effect is \(f_\mu^{\mathtt{GT}}(x_2)\) is: \(x_2\) in \([0, \frac{1}{4}]\), \(-x_2\) in \([\frac{1}{4}, \frac{1}{2}]\) and \(0\) in \([\frac{1}{2}, 1]\)\(x_2\) and it has zero-uncertainty, \(f_\sigma^{\mathtt{GT}}(x_2) = 0\), since \(x_2\) does not appear in any interaction term.
+We confirm that UALE computes it correctly whereas PDP-ICE fail in both the average effect and the uncertainty.
 
@@
 
-\appendix
+% \appendix
 
-\section{List with Symbols}
-\label{sec:list-with-symbols}
+% \section{List with Symbols}
+% \label{sec:list-with-symbols}
 
-\section{Show that is unbiased estimator}
-\label{sec:proof-1}
+% \section{Show that is unbiased estimator}
+% \label{sec:proof-1}
 
-\section{Proof of Theorem~\ref{sec:theorem-1}}
-\label{sec:proof-of-3-1}
+% \section{Proof of Theorem~\ref{sec:theorem-1}}
+% \label{sec:proof-of-3-1}
 
-\section{Proof of Corollary~\ref{sec:coroll-1}}
-\label{sec:proof-of-coroll}
+% \section{Proof of Corollary~\ref{sec:coroll-1}}
+% \label{sec:proof-of-coroll}
 
-\section{Dynamic Programming}
-\label{sec:dynamic-programming}
+% \section{Dynamic Programming}
+% \label{sec:dynamic-programming}
 
-\todo{Review all the description}
-For achieving a computationally-grounded solution we set a threshold
-\(K_{max}\) on the maximum number of bins which also discretizes the
-solution space. The width of the bin can take discrete values that are
-multiple of the minimum step
-\(u = \frac{x_{s, max} - x_{s, min}}{K_{max}}\). For defining the
-solution, we use two indexes. The index
-\(i \in \{0, \ldots, K_{max}\}\) denotes the point \((z_i)\) and the
-index \(j \in \{0, \ldots, K_{max}\} \) denotes the position of the
-\(j\)-th multiple of the minimum step, i.e., 
-\(x_j = x_{s,min} + j \cdot u\). The recursive cost function
-\(T(i,j)\) is the cost of setting \(z_i=x_j\):
-\begin{equation}
-  \label{eq:recursive_cost}
-  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
-\end{equation}
-%
-where \(\mathcal{T}(0,j)\) equals zero if \(j=0\) and \(\infty\) in
-any other case. \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin
-with limits \([x_l, x_j)]\):
+% \todo{Review all the description}
+% For achieving a computationally-grounded solution we set a threshold
+% \(K_{max}\) on the maximum number of bins which also discretizes the
+% solution space. The width of the bin can take discrete values that are
+% multiple of the minimum step
+% \(u = \frac{x_{s, max} - x_{s, min}}{K_{max}}\). For defining the
+% solution, we use two indexes. The index
+% \(i \in \{0, \ldots, K_{max}\}\) denotes the point \((z_i)\) and the
+% index \(j \in \{0, \ldots, K_{max}\} \) denotes the position of the
+% \(j\)-th multiple of the minimum step, i.e., 
+% \(x_j = x_{s,min} + j \cdot u\). The recursive cost function
+% \(T(i,j)\) is the cost of setting \(z_i=x_j\):
+% \begin{equation}
+%   \label{eq:recursive_cost}
+%   \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
+% \end{equation}
+% %
+% where \(\mathcal{T}(0,j)\) equals zero if \(j=0\) and \(\infty\) in
+% any other case. \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin
+% with limits \([x_l, x_j)]\):
 
-\begin{equation}
-  \label{eq:cost_step}
-  \mathcal{B}(x_l, x_j) = \begin{cases}
-                            \infty, & \text{if $x_j > x_l$ or \(|\mathcal{S}_{(x_j, x_l)}| < N\)}\\
-                            0, & \text{if $x_j = x_l$}\\
-                            \hat{\sigma}^2(x_j, x_l), &\text{if $x_j \leq x_l$}
-  \end{cases}
-\end{equation}
+% \begin{equation}
+%   \label{eq:cost_step}
+%   \mathcal{B}(x_l, x_j) = \begin{cases}
+%                             \infty, & \text{if $x_j > x_l$ or \(|\mathcal{S}_{(x_j, x_l)}| < N\)}\\
+%                             0, & \text{if $x_j = x_l$}\\
+%                             \hat{\sigma}^2(x_j, x_l), &\text{if $x_j \leq x_l$}
+%   \end{cases}
+% \end{equation}
 
-The optimal solution is given by solving
-\(\mathcal{L} = \mathcal{T}(K_{max}, K_{max})\) and keeping track of the sequence of
-steps.
+% The optimal solution is given by solving
+% \(\mathcal{L} = \mathcal{T}(K_{max}, K_{max})\) and keeping track of the sequence of
+% steps.
 
-\noindent
+% \noindent
 
-\begin{itemize}
-\item Discuss more aspects (e.g. Computational complexity)
-\end{itemize}
+% \begin{itemize}
+% \item Discuss more aspects (e.g. Computational complexity)
+% \end{itemize}
 

### 2022-10-14 AISTATS2023/paper.tex
latest

@@
   & \tau_k = 1 - \alpha \frac{|S_k|}{N} \\
-  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
+  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{PPB}}\\
                                      & z_0 = x_{s,min}\\
@@
 
-The above experiments show that UALE models correctly and estimates
-accurately both the average effect and its uncertainty. We also
-observed that UALE's variable-size bin-splitting method that we
-propose, leads to (a) an accurate approximation of the uncertainty and
-(b) favoring wider bins we facilitate the interpretation of both the
-average effect and the uncertainty. In contrast, PDP-ICE provides
-misleading explanations, due to ignoring correlations between
-features. The examples above do not cover the case of an interaction
-term between correlated features, for example a term \(x_1x_2\),
-because in this case there is an open debate about the ground-truth
-effect~\citep{Gromping2020MAEP}.
+The above experiments show that UALE models correctly and estimates accurately both the average effect and its uncertainty.
+We also observed that UALE's variable-size bin-splitting method that we propose, leads to (a) an accurate approximation of the uncertainty and (b) favoring wider bins we facilitate the interpretation of both the average effect and the uncertainty.
+In contrast, PDP-ICE provides misleading explanations in many cases, by ignoring dependence relationships between features.
+The examples above do not cover the case of an interaction term between correlated features, for example a term \(x_1x_2\), because in this case there is an open debate about the ground-truth effect~\citep{Gromping2020MAEP}.
 
@@
 
-This paper introduced UALE, a method that extends ALE, for not only quantifying the global feature effect but also its uncertainty, using the standard deviation of instance-level effects.
-We showed that UALE is more accurate than existent methods, e.g.
-ICE plots, in modeling the uncertainty in cases of correlated features.
-However, it is not trivial to make an unbiased estimate of the uncertainty.
-Unlike ALE, where for estimating the average effect, splitting the feature domain in \(K\) equal-width bins is sufficient, we theoretically proved that, for estimating the uncertainty, the same approach would introduce bias.
+We have introduced UALE, a method that extends ALE by quantifying the uncertainty of the global feature effect.
+To do that, UALE uses the standard deviation of instance-level effects.
+We showed that UALE models the uncertainty of feature effect more accurately than existing methods, such as ICE plots, especially in cases where there are dependencies between features.
+Providing an unbiased estimate of the uncertainty of the feature effect, however, is not trivial.
+In the case of ALE and the approximation of the average global effect, splitting the feature domain in K equal-width bins is sufficient.
+For estimating the uncertainty, however, we proved that this approach would introduce bias.
 We overcame this limitation by proposing an algorithm for automatically determining the optimal variable-size partitioning considering the instance-level effects.
-Through, synthetic and real-world examples, we show that UALE is an effective method for modeling, both, the average effect and the uncertainty.
+We show, through both synthetic and real-world examples, that UALE is an effective method for modeling both the average effect and the uncertainty.
 
-There are, though, limitations and open issues.
-First, for efficiently finding the optimal bin partitioning, UALE relies on DALE's approximation, so it is directly applicable only to differentiable models.
-Furthermore, UALE quantifies only the standard deviation of instance-level effects, providing limited intuition about the sources of the uncertainty, i.e., which features and to what degree they contribute to the heterogeneity of instance-level effects.
-Future work could target that.
+There are certain limitations and open issues in the proposed approach.
+First, finding the optimal bin partitioning can be computationally expensive, if applied directly to ALE.
+By using DALE's approximation this limitation is resolved, however this is directly applicable only for differentiable models.
+Furthermore, UALE quantifies only the standard deviation of instance-level effects, providing limited intuition about the sources of the uncertainty.
+Future work should therefore aim at developing methods for identifying which features and to what degree contribute to the heterogeneity of instance-level effects.
 

### 2022-10-14 examples/example_1.py
latest

@@ -5,6 +5,7 @@ import numpy as np
 import feature_effect as fe
 import example_models.distributions as dist
 import example_models.models as models
+# plt.rcParams['text.usetex'] = True
 
 savefig = True
 np.random.seed(21)

### 2022-10-14 examples/example_bin_splitting_1.py

@@ -72,7 +72,7 @@ utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["mu_err_mean"],
                          stats_fixed["mu_err_std"],
                          stats_auto["mu_err_mean"],
-                         stats_auto["mu_err_mean"],
+                         stats_auto["mu_err_std"],
                          "mu",
                          savefig=savepath)
 
@@ -81,7 +81,7 @@ utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["var_err_mean"],
                          stats_fixed["var_err_std"],
                          stats_auto["var_err_mean"],
-                         stats_auto["var_err_mean"],
+                         stats_auto["var_err_std"],
                          "var",
                          savefig=savepath)
 
@@ -90,6 +90,6 @@ utils.plot_fixed_vs_auto(K_list,
                          stats_fixed["rho_mean"],
                          stats_fixed["rho_std"],
                          stats_auto["rho_mean"],
-                         stats_auto["rho_mean"],
+                         stats_auto["rho_std"],
                          "rho",
                          savefig=savepath)

### 2022-10-14 examples/real_dataset_3.py

@@ -11,6 +11,7 @@ import matplotlib.pyplot as plt
 from sklearn.preprocessing import Normalizer
 from sklearn.model_selection import train_test_split
 
+savefig = True
 def create_nn(X):
     nn = keras.Sequential([keras.layers.Input(shape=[X.shape[1]]),
                            keras.layers.Dense(256, activation='relu', use_bias=True),
@@ -115,7 +116,6 @@ python_random.seed(1244343)
 tf.random.set_seed(12384)
 
 folder_name = "real_dataset_3"
-savefig = True
 
 # load-prepare data
 parent_path = os.path.dirname(os.path.dirname(path))

### 2022-10-14 examples/utils.py

@@ -264,21 +264,26 @@ def plot_fixed_vs_auto(K_list, fixed_mean, fixed_std, auto_mean, auto_std, metri
     plt.figure()
     assert metric in ["rho", "mu", "var"]
     if metric == "rho":
-        pass
-        plt.ylabel(r"$\log \mathcal{L}_{\mathtt{K}}^{\rho^2}$")
-        plt.title(r"Log of Nuisance Uncertainty ($\log \mathcal{L}_{\mathtt{K}}^{\rho^2}$)")
+        plt.ylabel(r"$\log \mathcal{L}_{\mathtt{K}}^{\rho}$")
+        plt.title(r"log mean bin-residual error($\log \mathcal{L}_{\mathtt{K}}^{\rho}$)")
     elif metric == "var":
-        plt.ylabel("$\log \mathcal{L}_{\mathtt{K}}^{\sigma^2}$")
-        plt.title("Log Error in Uncertainty ($\log \mathcal{L}_{\mathtt{K}}^{\sigma^2}$)")
+        plt.ylabel("$\log \mathcal{L}_{\mathtt{K}}^{\sigma}$")
+        plt.title("log mean bin-uncertainty error ($\log \mathcal{L}_{\mathtt{K}}^{\sigma}$)")
     elif metric == "mu":
         plt.ylabel("$\log \mathcal{L}_{\mathtt{K}}^{\mu}$")
-        plt.title("Log Error in Mean Effect ($\log \mathcal{L}_{\mathtt{K}}^{\mu}$)")
+        plt.title("log mean bin-effect error ($\log \mathcal{L}_{\mathtt{K}}^{\mu}$)")
 
     plt.plot(K_list, np.log(fixed_mean), 'x', label="fixed-bin")
     plt.plot(K_list,
              np.log(np.repeat(auto_mean, len(K_list))),
              "r--",
-             label="auto-bin")
+             label="UALE (variable-size)")
+
+    # plt.errorbar(K_list, fixed_mean, yerr=fixed_std, fmt="", linestyle='', label="fixed-size")
+    # plt.plot(K_list,
+    #          np.repeat(auto_mean, len(K_list)),
+    #          "r--",
+    #          label="UALE (variable-size)")
 
     plt.xlabel("$K$")
 

### 2022-10-14 feature_effect/visualization.py

@@ -66,7 +66,7 @@ def feature_effect_plot(params,
     fig, (ax1, ax2) = plt.subplots(2, 1, sharex=True)
 
     if title is None:
-        ax1.set_title("UALE: $x_{" + str(feature + 1) + "}$")
+        ax1.set_title(r"UALE: $x_{" + str(feature + 1) + "}$")
     else:
         ax1.set_title(title)
 
@@ -93,10 +93,10 @@ def feature_effect_plot(params,
                     gt_bins,
                     min_points_per_bin)
 
-    ax1.set_ylabel("$y$")
+    ax1.set_ylabel(r"$y$")
 
-    ax2.set_xlabel("$x_{%d}$" % (feature+1))
-    ax2.set_ylabel("$\partial y / \partial x_{%d}$" % (feature+1))
+    ax2.set_xlabel(r"$x_{%d}$" % (feature+1))
+    ax2.set_ylabel(r"$\partial y / \partial x_{%d}$" % (feature+1))
 
     if savefig:
         plt.savefig(savefig, bbox_inches="tight")
@@ -109,7 +109,7 @@ def feature_effect_plot(params,
 
 def feature_effect(ax1, x, y, estimator_var, std, rem_eff, first_empty, limits, point_limit, error=True, gt=None):
     # first subplot
-    ax1.plot(x, y, "b--", label="$\hat{f}_{\mu}$")
+    ax1.plot(x, y, "b--", label=r"$\hat{f}_{\mu}$")
 
     if first_empty is not None:
         added_line = .3*(np.max(y) - np.min(y))
@@ -118,7 +118,7 @@ def feature_effect(ax1, x, y, estimator_var, std, rem_eff, first_empty, limits, 
                    alpha=.7,
                    label="first bin with < " + str(point_limit) + " points")
     if error == "std":
-        ax1.fill_between(x, y-std, y+std, color='red', alpha=0.2, label="$\hat{f}_{\sigma}$")
+        ax1.fill_between(x, y-std, y+std, color='red', alpha=0.2, label=r"$\hat{f}_{\sigma}$")
         # ax1.fill_between(x, y-rem_eff, y+rem_eff, color='blue', alpha=0.7, label="std")
     elif error == "standard error":
         ax1.fill_between(x, y-2*np.sqrt(estimator_var), y+2*np.sqrt(estimator_var), color='red', alpha=0.6, label="standard error")
[…]

### 2022-10-15 AISTATS2023/paper.tex
ready for submission

@@
 In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and its uncertainty.
-We show that UALE quantifies uncertainty effectively, even in case of correlated features.
 We also note that as in ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting).
@@
 Based on this, we propose a method to determine the optimal intervals, balancing the estimation bias and variance.
-We demonstrate, through synthetic and real datasets, (a) the advantages of modeling the uncertainty with UALE compared to alternative methods and (b) the effectiveness of UALE's appropriate bin splitting for accurately approximating of the average effect and its uncertainty.
+We demonstrate, through synthetic and real datasets, (a) the advantages of modeling the uncertainty with UALE compared to alternative methods, especially in cases of correlated features and (b) the effectiveness of UALE's appropriate bin splitting for accurately approximating of the average effect and its uncertainty.
 \end{abstract}
@@
 In the case of PDP, such deviation is visualized via the Individual Conditional Explanations (ICE)~\citep{goldstein2015peeking}.
-ICE plots, however, have the same limitations as PDP in case of correlated features.
+ICE plots, however, have the same limitations as PDP~\citep{baniecki2021fooling} in case of correlated features.
 No method to quantify the deviation of instance-level effects has been proposed for ALE.
@@
   \centering
-  \includegraphics[width=.5\textwidth]{concept_figure/fig-1.pdf}
-  \caption{Illustration of UALE feature effect plot.}
+  \includegraphics[width=.4\textwidth]{concept_figure/fig-1.pdf}
+  \caption{Illustration of UALE feature effect plot on a toy example,
+    where \(x_1 \sim \mathcal{U}(0,1)\) and
+    \(x_2 \sim \mathcal{N}(0,\sigma^2_2)\) and
+    \(f(\xb)= f_0(\xb) + x_1x_2\), where \(f_0(\xb)\) when
+    \(x_1 < \frac{1}{5}\) and zero otherwise.}
   \label{fig:UALE-figure}
@@
 
-UALE defines the effect as a pair that consists of the average effect and the uncertainty \((\mu(z), \sigma(z))\).
-For visualizing it, we use the accumulated average effect, as defined in Eq.~(\ref{eq:ALE}), and the accumulated uncertainty, as defined in Eq.~(\ref{eq:ALE_acc_unc}).
-Figure~\ref{fig:UALE-figure} illustrates a toy example, where \(x_1 \sim \mathcal{U}(0,1)\) and \(x_2 \sim \mathcal{N}(0,\sigma^2_2)\) and the black-box function is \(f(\xb)= \sin(2 \pi x_1) + x_1x_2\).
+UALE defines the effect as a pair that consists of the average effect
+and the uncertainty \((\mu(z), \sigma(z))\). For visualizing it, we
+use the accumulated average effect, as defined in Eq.~(\ref{eq:ALE}),
+and the accumulated uncertainty, as defined in
+Eq.~(\ref{eq:ALE_acc_unc}).
 
@@
 \( z_{k_x - 1} \leq x_s < z_{k_x}\). Eq.~(\ref{eq:ALE_2}) and
-Eq.~(\ref{eq:ALE_accumulated_var}) are piecewise-linear approximation
+Eq.~(\ref{eq:ALE_accumulated_var}) are piecewise linear approximation
 of Eq.~(\ref{eq:ALE}) and Eq.~(\ref{eq:ALE_acc_unc}).
@@
 %
-In the Appendix~\ref{sec:proof-1}, we show that
+In the Appendix %~\ref{sec:proof-1}
+, we show that
 \(\hat{\sigma}^2(z_1, z_2)\) is an unbiased estimator of
@@
 \begin{proof}
-The proof is at Appendix~\ref{sec:proof-of-3-1}.
+The proof is at Appendix.%~\ref{sec:proof-of-3-1}.
 \end{proof}
@@
   \label{sec:coroll}
-  A bin-splitting \(\mathcal{Z}\) that minimizes the accumulated error
-  if and only if it also minimizes
+  If a bin-splitting \(\mathcal{Z}\) minimizes the accumulated error,
+  then it also minimizes
   \(\sum_{k=1}^K\sigma_*^2(z_1, z_2) \Delta z_k \)
@@
   z_K) (z_K - z_0)\) which is independent of the bin-splitting. A more
-  detailed proof is provided in the Appendix~\ref{sec:proof-of-coroll}.
+  detailed proof is provided in the Appendix.%~\ref{sec:proof-of-coroll}.
  \end{proof}
@@
   \label{eq:recursive_cost}
-  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
+  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(l, j) \right ]
 \end{equation}
@@
 % any other case.
-The term \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin with
-limits \([x_l, x_j)\):
+The term \(\mathcal{B}(l, j)\) denotes the cost of creating a bin with
+limits \([x_l, x_j)\). \(\mathcal{B}\) can be custom-based. In our
+case, following Eq.~(\ref{eq:opt}), we set it to
+\(\tau_k \hat{\sigma}^2(x_l, x_j) (x_j-x_l)\) if the bin is valid,
+i.e., it has enough points \(|\mathcal{S}_k| \geq
+N_{\mathtt{PPB}}\). The
+optimal solution is given by solving
+\(\mathcal{L} = \mathcal{T}(K_{max}, K_{max})\) and keeping track of
+the sequence of steps.
 
-\begin{equation}
-  \label{eq:cost_step}
-  \mathcal{B}(x_l, x_j) = \begin{cases}
-                            \infty, & \text{if $x_j > x_l$ or \(|\mathcal{S}_{(x_j, x_l)}| < N\)}\\
-                            0, & \text{if $x_j = x_l$}\\
-                            \hat{\sigma}^2(x_j, x_l), &\text{if $x_j \leq x_l$}
-  \end{cases}
-\end{equation}
-
-The optimal solution is given by solving
-\(\mathcal{L} = \mathcal{T}(K_{max}, K_{max})\) and keeping track of the sequence of
-steps.
-
-% \begin{algorithm}
-%   \end{algorithm}
+Note that in case the instance level effects are computed using the
+DALE method then there is no need to reevaluate \(f\) for each
+partitioning. The optimization algorithm simply reallocates the
+precomputed instance-level effects at each bin. Therefore, setting the
+maximum number \(K_{max}\) up to (roughly) \(100\) bins, the method
+finds the optimal partitioning in a couple of seconds. Further details
+and evaluation are provided in the Appendix.
 
@@
 without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
-(b) without interaction (\(\alpha=0\)) and different weigths
+(b) without interaction (\(\alpha=0\)) and different weights
 (\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha > 0\)) and
@@
 use the fact that under no-interactions, \(a=0\), the uncertainty must
-be zero (no heteregoneity). We discuss separately the case with
+be zero (no heterogeneity). We discuss separately the case with
 \(a>0\). Finally, we also demonstrate that UALE's approximation is
@@
 \([0, \frac{1}{4}), [\frac{1}{4}, \frac{1}{2}), [\frac{1}{2}, 1)\),
-that correspond to the piecewise-linear regions of the effect.
+that correspond to the piecewise linear regions of the effect.
 
@@
 the \(x_1\) domain into wide bins that facilitate the interpretation
-and create a zero cummulative bin-error \(\mathcal{E}_{\mathcal{Z}}\)
+and create a zero cumulative bin-error \(\mathcal{E}_{\mathcal{Z}}\)
 approximation.
@@
 bins. The evaluation is based on the Mean Absolute Error (MAE) of the
-bin effect \(\mu\) and of the uncertainty \(\sigma\) accross bins,
+bin effect \(\mu\) and of the uncertainty \(\sigma\) across bins,
 i.e.,
@@
 We compare UALE vs fixed-size approximation, when (a) \(f\) is
-piecewise-linear and (b) \(f\) is non-linear. In all examples, we
+piecewise linear and (b) \(f\) is non-linear. In all examples, we
 execute \(t = 30\) independent runs, using each time \(N=500\)
@@
 
-\paragraph{Piecewise-Linear Function.}
+\paragraph{Piecewise Linear Function.}
 
@@
   \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.pdf}
-  \caption{Bin-Splitting, piecewise-linear function: UALE's approximation
+  \caption{Bin-Splitting, piecewise linear function: UALE's approximation
     (Top-Left). UALE vs fixed-size approximations in terms of:
@@
 
-Here, we aim at demonstrating the usefuleness of uncertainty
+Here, we aim at demonstrating the usefulness of uncertainty
 quantification and the advantages of UALE's approximation, on the
@@
 illustration purposes. Results for all features, along with in-depth
-information about the preprocessing, training and evaluation parts are
+information about the reprocessing, training and evaluation parts are
 provided in the Appendix.
@@
 
-We have introduced UALE, a method that extends ALE by quantifying the uncertainty of the global feature effect.
-To do that, UALE uses the standard deviation of instance-level effects.
-We showed that UALE models the uncertainty of feature effect more accurately than existing methods, such as ICE plots, especially in cases where there are dependencies between features.
-Providing an unbiased estimate of the uncertainty of the feature effect, however, is not trivial.
-In the case of ALE and the approximation of the average global effect, splitting the feature domain in K equal-width bins is sufficient.
-For estimating the uncertainty, however, we proved that this approach would introduce bias.
-We overcame this limitation by proposing an algorithm for automatically determining the optimal variable-size partitioning considering the instance-level effects.
-We show, through both synthetic and real-world examples, that UALE is an effective method for modeling both the average effect and the uncertainty.
+In this work we presented the global feature effect method UALE, that
+extends ALE to quantify the uncertainty of the feature effect. We
+showed that UALE models such uncertainty more accurately than existing
+methods, such as ICE plots, especially in cases of highly correlated
+features, where that latest fail. Providing an unbiased estimate of
+this uncertainty, however, is not trivial. In the case of ALE, for the
+approximation of the average effect, splitting the feature domain in
+fixed-width bins is sufficient. For estimating the uncertainty,
+though, we proved that this approach would introduce bias. To overcame
+this limitation, we proposed a method to automatically determine the
+optimal variable-width bin-splitting considering the underlying
+instance-level effects. We demonstrated, through both synthetic and
+real-world examples, that UALE is an effective method to model both
+the average effect and the uncertainty.
 
-There are certain limitations and open issues in the proposed approach.
-First, finding the optimal bin partitioning can be computationally expensive, if applied directly to ALE.
-By using DALE's approximation this limitation is resolved, however this is directly applicable only for differentiable models.
-Furthermore, UALE quantifies only the standard deviation of instance-level effects, providing limited intuition about the sources of the uncertainty.
-Future work should therefore aim at developing methods for identifying which features and to what degree contribute to the heterogeneity of instance-level effects.
+Future work involves exploring other summary statistics than standard
+deviation to reveal further uncertainty aspects. Also, we would like
+to study ways to quantify the contribution of each feature in the
+uncertainty. Finally, we would like to explore alternatives to
+implement UALE's bin splitting (exhaustive vs approximate computation)
+to deal with data-intensive cases.
 
 \clearpage
-\subsubsection*{Acknowledgments}
-All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support.
-To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.
+% \subsubsection*{Acknowledgments}
+% All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support.
+% To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.
 

### 2022-10-18 AISTATS2023/supplement.tex
started method description
added content in helping material

@@
 
-\usepackage{aistats2022}
+\usepackage{aistats2023}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Corollary}[theorem]
+\DeclareMathOperator*{\argmin}{arg\,min}
+
+\newcommand{\dfdx}{\frac{\partial f}{\partial x_s}}
+\newcommand{\xc}{\mathbf{x_c}}
+\newcommand{\DY}{\mathbf{\Delta Y}}
+\newcommand{\xb}{\mathbf{x}}
+\newcommand{\Xcb}{\mathcal{X}_c}
+\newcommand{\Xb}{\mathcal{X}}
+
 % If your paper is accepted, change the options for the package
@@
 \onecolumn
-\aistatstitle{Instructions for Paper Submissions to AISTATS 2022: \\
+\aistatstitle{Instructions for Paper Submissions to AISTATS 2023: \\
 Supplementary Materials}
 
-\section{FORMATTING INSTRUCTIONS}
+\section{THORETICAL DERIVATIONS}
 
-To prepare a supplementary pdf file, we ask the authors to use \texttt{aistats2022.sty} as a style file and to follow the same formatting instructions as in the main paper.
-The only difference is that the supplementary material must be in a \emph{single-column} format.
-You can use \texttt{supplement.tex} in our starter pack as a starting point, or append the supplementary content to the main paper and split the final PDF into two separate files.
+\subsection{Proof that \(\hat{\sigma}^2(z_1, z_2)\) is an unbiased estimator of \(\sigma_*^2(z_1, z_2)\)}
 
-Note that reviewers are under no obligation to examine your supplementary material.
+In Section .. we use that \(\hat{\sigma}^2(z_1, z_2)\) is an unbiased
+estimator of
+\(\sigma^2_*(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{X_c|X_s=z}
+  \left [ (f^s(z, X_c) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
+  z_1} \), under the assumption that the points are uniformly distributed in \([z_1, z_2)\),
+i.e., \(x_s^i \sim \mathcal{U}(z_1, z_2)\).
 
-\section{MISSING PROOFS}
+\paragraph{Proof}
 
-The supplementary materials may contain detailed proofs of the results that are missing in the main paper.
+We have made the hypothesis that the points \(\xb^i\) are
+coming from the distibution
+\(\mathcal{S} = p(X_c|X_s) p(X_s) = p(X_c|X_s) \mathcal{U}(x_s; z_1,
+z_2)= \frac{p(\xc|z)}{z_2 - z_1}\) uniformly distributed in the interval \([z_1, z_2)\) in terms
 
-\subsection{Proof of Lemma 3}
+\begin{align}
+  \mathbb{E}_\mathcal{S}[\hat{\sigma}^2(z_1, z_2)] &= \mathbb{E}_\mathcal{S}[\frac{1}{|\mathcal{S}|} \sum_{i=1}^{|\mathcal{S}|} (f^s(\xb^i) - \mu(z_1, z_2))^2] \\
+  & = \frac{1}{|\mathcal{S}|} \sum_{i=1}^{|\mathcal{S}|} \mathbb{E}_\mathcal{S}[(f^s(\xb^i) - \mu(z_1, z_2))^2]\\
+  & = \frac{1}{|\mathcal{S}|} \sum_{i=1}^{|\mathcal{S}|} \mathbb{E}_\mathcal{S}[(f^s(\xb^i) - \mu(z_1, z_2))^2]\\
+  \end{align}
 
-\textit{In this section, we present the detailed proof of Lemma 3 and then [ ... ]}
+\begin{align}
+  \sigma^2_*(z_1, z_2) & = \frac{\int_{z_1}^{z_2} \mathbb{E}_{X_c|X_s=z}
+  \left [ (f^s(z, X_c) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
+                         z_1} \\
+  & = \mathbb{E}_{z \sim \mathcal{U}(z_1, z_2)}\mathbb{E}_{X_c|X_s=z}
+    \left [ (f^s(z, X_c) - \mu(z_1, z_2) )^2 \right] \\
+  & = \mathbb{E}_{X} \left [ (f^s(X) - \mu(z_1, z_2) )^2 \right]
+  \end{align}
 
-\section{ADDITIONAL EXPERIMENTS}
+\subsection{Proof Of Theorem 3.1}
 
-If you have additional experimental results, you may include them in the supplementary materials.
+\textit{  If we define (a) the residual \(\rho(z)\) as the difference between
+  the expected effect at \(z\) and the bin-effect, i.e
+  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\) and (b)
+  \(\mathcal{E}(z_1, z_2)\) as the mean squared residual of the bin,
+  i.e.
+  \(\mathcal{E}(z_1, z_2) = \frac{\int_{z_1}^{z_2}\rho^2(z) \partial
+    z}{z_2 - z_1}\), then it holds that:
+\begin{equation}
+    \label{eq:bin-uncertainty-proof}
+ \sigma_*^2(z_1, z_2) = \sigma^2(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
+\end{equation}
+}
 
-\subsection{The Effect of Regularization Parameter}
+\paragraph{Proof}
 
-\textit{Our algorithm depends on the regularization parameter $\lambda$. Figure 1 below illustrates the effect of this parameter on the performance of our algorithm. As we can see, [ ... ]}
+\begin{align}
+  \sigma_*^2(z_1, z_2) &= \frac{1}{z_2 - z_1}\int_{z_1}^{z_2} \mathbb{E}_{X_c|z} \left [ \left( f^s( z, X_c) - \mu(z_1, z_2) \right)^2 \right] \partial z \\
+                       &= \frac{1}{z_2 - z_1} \int_{z_1}^{z_2} \mathbb{E}_{X_c|z} \left [ \left ( f^s(z, X_c) - \mu(z) + \rho(z) \right )^2 \right] \partial z \\
+  &= \frac{1}{z_2 - z_1} \int_{z_1}^{z_2} \mathbb{E}_{X_c|z} \left [ (f^s(z, X_c) - \mu(z) )^2 + \rho(z)^2 + 2f^s(z, X_c)\mu(z) ) \right ]\\
+  &= \frac{1}{z_2 - z_1} \int_{z_1}^{z_2} \left (
+  \underbrace{\mathbb{E}_{X_c|z} \left [ (f^s(z, X_c) - \mu(z) )^2 \right ]}_{\sigma^2(z)}  +
+  \underbrace{\mathbb{E}_{X_c|z} \left [ \rho^2(z) \right]}_{\rho^2(z)} +
+  2 (\underbrace{\mathbb{E}_{X_c|z} \left [ (f^s(z, X_c)   \right ]}_{\mu(z)} - \mu(z)) \rho(z) )\right )  \partial z \\
+  &= \underbrace{\frac{1}{z_2 - z_1} \int_{z_1}^{z_2} \sigma^2(z) \partial z}_{\sigma^2(z_1, z_2)} + \underbrace{\frac{1}{z_2 - z_1} \int_{z_1}^{z_2} \rho^2(z) \partial z}_{\mathcal{E}^2(z_1, z_2)} = \sigma^2(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
+\end{align}
+
+\subsection{Proof Of Corollary}
+
+\textit{ If a bin-splitting \(\mathcal{Z}\) minimizes the accumulated error, then it also minimizes
+  \(\sum_{k=1}^K\sigma_*^2(z_1, z_2) \Delta z_k \)}
+
+We want to show that
+
+\[ \mathcal{Z}^* = \argmin_{\mathcal{Z}} \sum_{k=1}^K \sigma_*^2(z_{k-1}, z_k) \Delta z_k \Leftrightarrow \mathcal{Z}^* = \argmin_{\mathcal{Z}} \sum_{k=1}^K \mathcal{E}^2(z_{k-1}, z_k) \Delta z_k \]
+
+\paragraph{Proof}
+
+\begin{align}
+  \mathcal{Z}^* &= \argmin_{\mathcal{Z}} \sum_{k=1}^K \sigma_*^2(z_{k-1}, z_k) \Delta z_k \\
+                & = \argmin_{\mathcal{Z}} \left [ \sum_{k=1}^K (\sigma^2(z_{k-1}, z_k) + \mathcal{E}^2(z_{k-1}, z_k)) \Delta z_k \right ] \\
+                & = \argmin_{\mathcal{Z}} \left [ \sum_{k=1}^K \left ( \frac{\Delta z_k}{\Delta z_k} \int_{z_{k-1}}^{z_k} \sigma^2(z) \partial z)   + \mathcal{E}^2(z_{k-1}, z_k) \Delta z_k \right ) \right ] \\
+                & = \argmin_{\mathcal{Z}} \left [ \underbrace{\int_{z_0}^{z_K} \sigma^2(z) \partial z}_{\text{independent of } \mathcal{Z}}   + \sum_{k=1}^K\mathcal{E}^2(z_{k-1}, z_k) \Delta z_k) \right ] \\
+                & = \argmin_{\mathcal{Z}} \sum_{k=1}^K\mathcal{E}^2(z_{k-1}, z_k) \Delta z_k
+  \end{align}
+
+\section{Dynamic Programming}
+\label{sec:dynamic-programming}
+
+For achieving a computationally-grounded solution we set a threshold
+\(K_{max}\) on the maximum number of bins which also discretizes the
+solution space. The width of the bin can take discrete values that are
+multiple of the minimum step
+\(u = \frac{x_{s, max} - x_{s, min}}{K_{max}}\). For defining the
+solution, we use two indexes. The index
+\(i \in \{0, \ldots, K_{max}\}\) denotes the point \((z_i)\) and the
+index \(j \in \{0, \ldots, K_{max}\} \) denotes the position of the
+\(j\)-th multiple of the minimum step, i.e.,
+\(x_j = x_{s,min} + j \cdot u\). The recursive cost function
+\(T(i,j)\) is the cost of setting \(z_i=x_j\):
+\begin{equation}
+  \label{eq:recursive_cost}
+  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
+\end{equation}
+%
+where \(\mathcal{T}(0,j)\) equals zero if \(j=0\) and \(\infty\) in
+any other case. \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin
+with limits \([x_l, x_j)]\):
+
+\begin{equation}
+  \label{eq:cost_step}
+  \mathcal{B}(x_l, x_j) = \begin{cases}
+                            \infty, & \text{if $x_j > x_l$ or \(|\mathcal{S}_{(x_j, x_l)}| < N\)}\\
+                            0, & \text{if $x_j = x_l$}\\
+                            \hat{\sigma}^2(x_j, x_l), &\text{if $x_j \leq x_l$}
+  \end{cases}
+\end{equation}
+
+The optimal solution is given by solving
+\(\mathcal{L} = \mathcal{T}(K_{max}, K_{max})\) and keeping track of the sequence of
+steps.
+
+\section{EXPERIMENTS APPEARED AT THE MAIN PAPER}
+
+\subsection{Simulation 1}
+
+\subsection{Simulation 2}
+
+\subsection{Real World Example}
+
+\section{Further Experimentation}
+
+% \section{MISSING PROOFS}
+
+% The supplementary materials may contain detailed proofs of the results that are missing in the main paper.
+
+% % \subsection{Proof of Lemma 3}
+
+% % \textit{In this section, we present the detailed proof of Lemma 3 and then [ ... ]}
+
+% % Given that,
+
+% % The mean effect at a specific point \(x_s\) is:
+
+% % \begin{equation}
+% %   \label{eq:3}
+% %   \mu(x_s) = \mathbb{E}_{\xc|x_s}\left [\dfdx (x_s, \xc) \right ]
+% % \end{equation}
+
+% % The variance at a specific point \(x_s\) is:
+
+% % \begin{equation}
+% %   \label{eq:3}
+% %   \sigma^2(x_s) = \mathrm{Var}_{\xc|x_s}\left [\dfdx (x_s, \xc) \right ]
+% % \end{equation}
+
+% % The mean effect inside the interval \([z_1, z_2)\) is:
+
+% % \begin{equation}
+% %   \label{eq:1}
+% %   \mu(z_1, z_2) = \frac{1}{z_2-z_1} \int_{z_1}^{z_2} \mathbb{E}_{\xc|x_s}\left [\dfdx (x_s, \xc)\right ]
+% % \end{equation}
+
+% % The accumulated variance inside the interval is:
+
+% % \begin{equation}
+% %   \label{eq:2}
+% %   \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \mathbb{E}_{\xc|x_s=z} \left [ \left ( \frac{\partial f}{\partial x_s}(x_s, \xc) - \mu(z_1, z_2) \right )^2 \right] \partial z
+% % \end{equation}
+
+% % The residual at a specific point \(x_s\) is:
+
+% % \begin{equation}
+% %   \label{eq:4}
+% %   \rho(x_s) = \mu(x_s) - \mu(z_1, z_2)
+% % \end{equation}
+
+% % We want to prove that:
+
+% % \begin{equation}
+% %   \label{eq:5}
+% %   \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
+% %   = \int_{z_1}^{z_2} \sigma^2(z) + \int_{z_1}^{z_2} \rho^2(z) \partial z
+% % \end{equation}
+
+% % Proof:
+
+% % \begin{gather}
+% %   \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \mathbb{E}_{\xc|x_s=z} \left [ \left( \frac{\partial f}{\partial x_s}( z, \xc) - \mu(z_1, z_2) \right)^2 \right] \partial z \\
+% %   = \int_{z_1}^{z_2} \mathbb{E}_{\xc|x_s=z} \left [ \left ( \frac{\partial f}{\partial x_s} - \mu(z) + \rho(z) \right )^2 \right] \partial z \\
+% %   = \int_{z_1}^{z_2} \left(
+% %   \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial f}{\partial x_s} - \mu(z) )^2 \right ]  +
+% %   \mathbb{E}_{\xc|x_s=z} \left [ \rho(z)^2 \right] +
+% %   \mathbb{E}_{\xc|x_s=z} \left [ 2(\frac{\partial f}{\partial x_s} - \mu(z) )\rho(z) \right ] \right )  \partial z \\
+% %   = \int_{z_1}^{z_2}(\sigma^2(z) + \rho^2(z) + 2(\mu(z)-\mu(z))\rho(z))\partial z \\
+% %   = \int_{z_1}^{z_2} \sigma^2(z)  + \rho^2(z) \partial z \\
+% % \end{gather}
+
+% \section{ADDITIONAL EXPERIMENTS}
+
+% If you have additional experimental results, you may include them in the supplementary materials.
+
+% \subsection{The Effect of Regularization Parameter}
+
+% \textit{Our algorithm depends on the regularization parameter $\lambda$. Figure 1 below illustrates the effect of this parameter on the performance of our algorithm. As we can see, [ ... ]}
 

### 2022-10-19 AISTATS2023/supplement.tex
added content in helping material

@@
 \DeclareMathOperator*{\argmin}{arg\,min}
+
+\usepackage{graphicx}
+\graphicspath{ {./../examples/} }
 
@@
 
-\subsection{Proof that \(\hat{\sigma}^2(z_1, z_2)\) is an unbiased estimator of \(\sigma_*^2(z_1, z_2)\)}
+\subsection{Show that \(\hat{\mu}(z_1, z_2) \approx \mu(z_1, z_2)\)}
 
-In Section .. we use that \(\hat{\sigma}^2(z_1, z_2)\) is an unbiased
+\textit{We want to show that (a)
+  \(\hat{\mu}(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:
+    \mathbf{x}^i \in \mathcal{S}} f^s(\mathbf{x}^i) \) is an unbiased
+  estimator of
+  \(\mu(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{X_c|z} \left [
+      f^s(z, X_c) \right ] \partial z}{z_2 - z_1} \), under the
+  assumption that that (a) \(z\) follows a uniform distribution in
+  \([z_1, z_2)\), i.e., \(z \sim \mathcal{U}(z_1, z_2)\) and (b) that
+  the points are i.i.d.  samples from the distribution
+  \(p(\xb) = p(\xc|z)p(z) = \frac{1}{z_2-z_1}p(\xc|z) \) }
+
+\paragraph{Description}
+  We just use the the fact that the population mean is an
+  unbiased estimator of the expected value. We just show that
+  \(\mu(z_1, z_2) = \mathbb{E}_{\tilde{X}} [f^s(\tilde{X})]\).
+
+\paragraph{Proof}
+\begin{align}
+  \mu(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{X_c|z} f^s(z, X_c)  \partial z}{z_2 - z_1}
+                = \mathbb{E}_{z \sim \mathcal{U}(z_1, z_2)}\mathbb{E}_{X_c|z} f^s(z, X_c)
+                = \mathbb{E}_{\tilde{X}} f^s(X)
+\end{align}
+
+\subsection{Show that \(\hat{\sigma}^2(z_1, z_2) \approx  \sigma_*^2(z_1, z_2)\)}
+
+\textit{We want to show that \(\hat{\sigma}^2(z_1, z_2) = \frac{1}{|\mathcal{S}_k|}
+\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
+  \hat{\mu}(z_1, z_2) \right )^2\) is an unbiased
 estimator of
@@
   \left [ (f^s(z, X_c) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
-  z_1} \), under the assumption that the points are uniformly distributed in \([z_1, z_2)\),
-i.e., \(x_s^i \sim \mathcal{U}(z_1, z_2)\).
+  z_1} \), , under the
+  assumption that that (a) \(z\) follows a uniform distribution in
+  \([z_1, z_2)\), i.e., \(z \sim \mathcal{U}(z_1, z_2)\) and (b) that
+  the points are i.i.d.  samples from the distribution
+  \(p(\xb) = p(\xc|z)p(z) = \frac{1}{z_2-z_1}p(\xc|z) \).}
+
+\paragraph{Description}
+Same as before, we just use the the fact that the population variance
+is an unbiased estimator of the variance. We just show that
+\(\sigma^2_*(z_1, z_2) = \mathbb{V}_{\tilde{X}} [f^s(\tilde{X})]\).
 
 \paragraph{Proof}
-
-We have made the hypothesis that the points \(\xb^i\) are
-coming from the distibution
-\(\mathcal{S} = p(X_c|X_s) p(X_s) = p(X_c|X_s) \mathcal{U}(x_s; z_1,
-z_2)= \frac{p(\xc|z)}{z_2 - z_1}\) uniformly distributed in the interval \([z_1, z_2)\) in terms
-
-\begin{align}
-  \mathbb{E}_\mathcal{S}[\hat{\sigma}^2(z_1, z_2)] &= \mathbb{E}_\mathcal{S}[\frac{1}{|\mathcal{S}|} \sum_{i=1}^{|\mathcal{S}|} (f^s(\xb^i) - \mu(z_1, z_2))^2] \\
-  & = \frac{1}{|\mathcal{S}|} \sum_{i=1}^{|\mathcal{S}|} \mathbb{E}_\mathcal{S}[(f^s(\xb^i) - \mu(z_1, z_2))^2]\\
-  & = \frac{1}{|\mathcal{S}|} \sum_{i=1}^{|\mathcal{S}|} \mathbb{E}_\mathcal{S}[(f^s(\xb^i) - \mu(z_1, z_2))^2]\\
-  \end{align}
 
@@
                          \left [ (f^s(z, X_c) - \mu(z_1, z_2) )^2 \right] \\
-  & = \mathbb{E}_{X} \left [ (f^s(X) - \mu(z_1, z_2) )^2 \right]
+                       & = \mathbb{E}_{\tilde{X}} \left [ (f^s(X) - \mu(z_1, z_2) )^2 \right]\\
+                       & = \mathbb{V}_{\tilde{X}} [f^s(\tilde{X})]
   \end{align}
@@
 
-\textit{  If we define (a) the residual \(\rho(z)\) as the difference between
+\textit{If we define (a) the residual \(\rho(z)\) as the difference between
   the expected effect at \(z\) and the bin-effect, i.e
@@
 
-\section{Dynamic Programming}
+\section{Dynamic Programming Analysis}
 \label{sec:dynamic-programming}
@@
 
-\section{EXPERIMENTS APPEARED AT THE MAIN PAPER}
+\section{Real World Experiment}
 
-\subsection{Simulation 1}
+In this section, we provide further details on the real-world
+example. The real-world example uses the California Housing Dataset,
+which contains 8 numerical features. We exclude instances with missing
+or outlier values. If we denote as \(\mu_s\) (\(\sigma_s\)) the
+average value (standard deviation) of the \(s\)-th feature, we
+consider outliers the instances of the training set with any feature
+value over three standard deviations from the mean, i.e.
+\(|x_s^i - \mu_s| > \sigma_s\). This preprocessing step discards
+\(884\) instancies, and \(N=19549\) remain. We provide their
+description with some basic descriptive statistics in
+Table~\ref{tab:features-description} and their histogram in
+Figure~\ref{fig:feature-histograms}.
 
-\subsection{Simulation 2}
+In Figure 7 of the main paper, we provided the UALE vs PDP-ICE plots
+for features \(x_2\) (latitude), \(x_6\) (total number of people) and
+\(x_8\) (median house value). In figure 8, we compared UALE with
+fixed-size approximation, for the same features. In
+Figure~\ref{fig:ex-real-1}, we provide the same information for the
+rest of the features; \(x_1\) (longitude), \(x_3\) (median age of
+houses), \(x_4\) (total number of rooms), \(x_5\) (total number of
+bedrooms) and \(x_7\) (total number of households). The observation of
+these feautures leads us to similar conclusion. First, UALE and
+PDP-ICE plots compute similar effects and level of heterogeneity and
+UALE's approximation is (almost) as good as the best fixed-size
+approximation. More specifically, we observe that UALE's variable size
+bin splitting correctly creates wide bins for features
+\(x_3, x_4, x_5, x_7\), where the feature effect plot is (piecewise)
+linear, while using narrow bins for feature \(x_2\) where the feature
+effect is not linear.
 
-\subsection{Real World Example}
+\begin{table}
+  \caption{Description of the features apparent in the California-Housing Dataset}
+  \label{tab:features-description}
+  \centering
+  \begin{tabular}{ c|c|c|c|c|c| }
+    \hline
+    & Description & \(min\) & \(\max\) & \(\mu\) & \(\sigma\) \\
+    \hline
+    \(x_1\) & longitude & \(-124.35\) & \(-114.31\) & \(-119.58\) & \(2\) \\
+    \(x_2\) & latitude  & \(32.54\) & \(41.95\) & \(35.65\) & \(2.14\) \\
+    \(x_3\) & median age of houses & \(1\) & \(52\) & \(29.01\) & \(12.42\) \\
+    \(x_4\) & total number of rooms & \(2\) & \(9179\) & \(2390.79\) & \(1433.83\) \\
+    \(x_5\) & total number of bedrooms & \(2\) & \(1797\) & \(493.86\) & \(291\) \\
+    \(x_6\) & total number of people & \(3\) & \(4818\) & \(1310.91\) & \(771.78\) \\
+    \(x_7\) & total number of households & \(2\) & \(1644\) & \(460.3\) & \(267.34\) \\
+    \(x_8\) & median income of households & \(0.5\) & \(9.56\) & \(3.72\) & \(1.60\) \\
+    \hline
+    \(y\) & median house value & \(14.999\) & \(500000\) & \(206864.41\) & \(115435.67\) \\
+    \hline
+  \end{tabular}
+\end{table}
 
-\section{Further Experimentation}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.43\textwidth]{real_dataset_3/hist_feat_1_after.pdf}
+  \includegraphics[width=.43\textwidth]{real_dataset_3/hist_feat_2_after.pdf}\\
+  \includegraphics[width=.43\textwidth]{real_dataset_3/hist_feat_3_after.pdf}
+  \includegraphics[width=.43\textwidth]{real_dataset_3/hist_feat_4_after.pdf}\\
+  \includegraphics[width=.43\textwidth]{real_dataset_3/hist_feat_5_after.pdf}
+  \includegraphics[width=.43\textwidth]{real_dataset_3/hist_feat_6_after.pdf}\\
+  \includegraphics[width=.43\textwidth]{real_dataset_3/hist_feat_7_after.pdf}
+  \includegraphics[width=.43\textwidth]{real_dataset_3/hist_feat_8_after.pdf}\\
+  \label{fig:feature-histograms}
+  \caption{The Histogram of each feature in the California Housing Dataset.}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.24\textwidth]{real_dataset_3/feature_0_ale_auto.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/feature_0_pdp_ice.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/compare_mu_err_feature_0.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/compare_var_err_feature_0.pdf}\\
+  \includegraphics[width=.24\textwidth]{real_dataset_3/feature_2_ale_auto.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/feature_2_pdp_ice.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/compare_mu_err_feature_2.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/compare_var_err_feature_2.pdf}\\
+  \includegraphics[width=.24\textwidth]{real_dataset_3/feature_3_ale_auto.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/feature_3_pdp_ice.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/compare_mu_err_feature_3.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/compare_var_err_feature_3.pdf}\\
+  \includegraphics[width=.24\textwidth]{real_dataset_3/feature_4_ale_auto.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/feature_4_pdp_ice.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/compare_mu_err_feature_4.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/compare_var_err_feature_4.pdf}\\
+  \includegraphics[width=.24\textwidth]{real_dataset_3/feature_6_ale_auto.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/feature_6_pdp_ice.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/compare_mu_err_feature_6.pdf}
+  \includegraphics[width=.24\textwidth]{real_dataset_3/compare_var_err_feature_6.pdf}\\
+  \caption{From left to right: (a) UALE plot, (b) PDP-ICE plot, (c) UALE vs fixed-size \(\mathcal{L}^{\mu}\) and (d) UALE vs fixed-size \(\mathcal{L}^{\sigma}\).
+From top to bottom, features \(x_1, x_3, x_4, x_5, x_7, x_8\).}
+  \label{fig:ex-real-1}
+\end{figure}
+
+% \section{Further Experimentation}
 

### 2022-10-19 examples/real_dataset_3.py

@@ -110,6 +110,7 @@ def find_axislimits(X):
     return axis_limits
 
 
+
 # freeze seed
 np.random.seed(138232)
 python_random.seed(1244343)
@@ -122,9 +123,40 @@ parent_path = os.path.dirname(os.path.dirname(path))
 datapath = os.path.join(parent_path, "data", "California-Housing", "housing.csv")
 X_df, Y_df, X, Y = load_prepare_data(datapath)
 
+# statistics on data
+X_df.describe()
+Y_df.describe()
+
+path2dir = os.path.join(parent_path, "examples/real_dataset_3/")
+for i in range(8):
+    tmp = X_df.iloc[:, i:i+1].to_numpy().squeeze()
+    plt.figure()
+    plt.hist(tmp, bins=50)
+    plt.xlabel(r"$x_" + str(i) + "$")
+    plt.title(r"Feature $x_" + str(i) + "$ histogram")
+    if savefig:
+        savepath = os.path.join(path2dir, "hist_feat_" + str(i+1) + ".pdf")
+        plt.savefig(savepath)
+    plt.show(block=False)
+
 # preprocess data
 X_norm, Y_norm, X_mu, X_sigma, Y_mu, Y_sigma = preprocess_data(X, Y)
 axis_limits = find_axislimits(X_norm)
+
+# statistics on preprocessed
+path2dir = os.path.join(parent_path, "examples/real_dataset_3/")
+for i in range(8):
+    tmp = X_norm[:,i] * X_sigma[i] + X_mu[i]
+    print(tmp.min(), tmp.max(), tmp.mean(), tmp.std())
+    plt.figure()
+    plt.hist(tmp, bins=50)
+    plt.xlabel(r"$x_" + str(i+1) + "$")
+    plt.title(r"Feature $x_" + str(i+1) + "$ histogram (after preprocessing)")
+    if savefig:
+        savepath = os.path.join(path2dir, "hist_feat_" + str(i+1) + "_after.pdf")
+        plt.savefig(savepath)
+    plt.show(block=False)
+
 
 X_tr, X_te, Y_tr, Y_te = split(X_norm, Y_norm)
 

### 2023-04-22 research/ideas.tex
small changes

\documentclass[12pt]{article}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{bm}

\title{Research Ideas}

\newcommand{\xb}{\mathbf{x}}
\newcommand{\xc}{\mathbf{x_c}}
\newcommand{\fxc}{f^{(\xc)}}
\newcommand{\fxs}{f^{(x_s)}}

\author{Vasilis Gkolemis}
\begin{document}
    \maketitle
    \section{Evaluation framework for interaction methods}

    \subsection{Idea 1}

    We may split every $f: \mathbb{R}^D \rightarrow \mathbb{R}$ into
    a model without interaction between $\xc$ and $x_s$,
    i.e., $f_{ni}(\xb) = \fxs(x_s) +  \fxc(\xc)$,
    and the interaction term $\kappa(\xc, x_s)$:

    $$f(\xb) = \underbrace{\fxs(x_s) +  \fxc(\xc)}_{f_{ni}(\xb)} + \kappa(\xc, x_s)$$

    \subsection{Idea 2}
\end{document}

### 2023-04-22 pythia/interaction.py

@@ -1,4 +1,5 @@
 import numpy as np
+import logging
 from pythia import helpers
 import pythia
 from pythia import dPDP
@@ -23,43 +24,46 @@ class HIndex:
         self.fitted_one_vs_all_matrix = False
 
     def fit(self, interaction_matrix=True, one_vs_all=True):
-        print("Fit HIndex - interacation_matrix")
         if interaction_matrix:
+            logging.info("\nH-Index interaction matrix: Start fitting")
             for i in range(self.nof_features):
                 print("Feature: ", i)
                 self.interaction_matrix[i, i] = 0
                 for j in range(i + 1, self.nof_features):
-                    self.pairwise(i, j)
+                    self.eval_pairwise(i, j)
             self.fitted_interaction_matrix = True
+            logging.info("H-Index interaction matrix: Fitting done")
 
-        print("Fit HIndex - one_vs_all matrix")
         if one_vs_all:
+            logging.info("\nH-Index one-vs-all matrix: Start fitting")
             for i in range(self.nof_features):
                 print("Feature: ", i)
-                self.one_vs_all(i)
+                self.eval_one_vs_all(i)
             self.fitted_one_vs_all_matrix = True
+            logging.info("H-Index one-vs-all matrix: Fitting done")
 
     def plot(self, interaction_matrix=True, one_vs_all=True):
+        # if not fitted, fit
         if not self.fitted_interaction_matrix and interaction_matrix:
             self.fit(interaction_matrix=True, one_vs_all=False)
         if not self.fitted_one_vs_all_matrix and one_vs_all:
             self.fit(interaction_matrix=False, one_vs_all=True)
 
+        # plot
         if interaction_matrix:
             self._plot_interaction_matrix()
         if one_vs_all:
             self._plot_one_vs_all_matrix()
 
-
-    def pairwise(self, feat_1, feat_2):
+    def eval_pairwise(self, feat_1, feat_2):
         if self.interaction_matrix[feat_1, feat_2] != self.empty_symbol:
             return self.interaction_matrix[feat_1, feat_2]
 
         x1 = self.data[:, feat_1]
         x2 = self.data[:, feat_2]
-        pdp_1 = self.pdp_1d(feat_1, x1)
-        pdp_2 = self.pdp_1d(feat_2, x2)
[…]

### 2023-05-24 research/ideas.tex
added some research ideas

@@
 
-    $$f(\xb) = \underbrace{\fxs(x_s) +  \fxc(\xc)}_{f_{ni}(\xb)} + \kappa(\xc, x_s)$$
+    \[
+      f(\xb) = \underbrace{\fxs(x_s) +  \fxc(\xc)}_{f_{ni}(\xb)} + \kappa(\xc, x_s)
+    \]
+
+    A simple approach is defining \(f\) to be a Neural Network and \(f_{ni}\) a Neural Additive Model without interaction
+    between \(x_s\) and \(\xc\).
+Then \(\kappa(\xc, x_s) = f(\xb) - f_{ni}(\xb)\) and we quantify the importance of \(\kappa\) as \(\mathbb{E}_{X_c, X_s} \left [ |\kappa(X_c, X_s)| \right ] \approx \sqrt{\frac{1}{N} \sum_i \kappa^2(\xc, x_s)} \).
 

### 2023-05-24 pythia/helpers.py
builiding interaction indices

@@ -78,6 +78,24 @@ def prep_ale_fit_params(par: dict):
 
 
 def prep_nof_instances(nof_instances: typing.Union[int, str], N: int) -> int:
+    """Prepares the argument nof_instances
+
+    Args
+    ---
+        nof_instances (int or str): The number of instances to use for the explanation
+        N (int): The number of instances in the dataset
+
+    Returns
+    ---
+        nof_instances (int): The number of instances to use for the explanation
+        indices (np.ndarray): The indices of the instances to use for the explanation
+    """
+    # assertions
+    assert type(nof_instances) in [int, str]
+    if type(nof_instances) is not int:
+        assert nof_instances == "all"
+
+    # prepare nof_instances
     if nof_instances == "all":
         nof_instances = N
     indices = np.random.choice(N, nof_instances, replace=False) if nof_instances < N else np.arange(N)

### 2023-05-24 pythia/interaction.py

@@ -7,24 +7,37 @@ from pythia.pdp import pdp_1d_vectorized, pdp_1d_non_vectorized, pdp_nd_non_vect
 import tqdm
 
 class HIndex:
+    """
+    H-Index interaction measure, as described in Friedman and Popescu (2008), https://arxiv.org/abs/0811.1679
+    """
+
     empty_symbol = 1e10
 
     def __init__(self, data, model, nof_instances=1000):
+        """
 
+        Args
+        ---
+            data (np.ndarray): The dataset, shape (nof_instances, nof_features)
+            model (function): The model to explain, takes as input data and returns predictions
+            nof_instances (int): The number of instances to use for the explanation
+        """
+        # setters
         self.nof_instances, self.indices = helpers.prep_nof_instances(nof_instances, data.shape[0])
         self.data = data[self.indices]
-
         self.model = model
         self.nof_features = self.data.shape[1]
 
+        # init
         self.interaction_matrix = np.ones((self.nof_features, self.nof_features)) * self.empty_symbol
         self.one_vs_all_matrix = np.ones((self.nof_features)) * self.empty_symbol
 
+        # flags
         self.fitted_interaction_matrix = False
         self.fitted_one_vs_all_matrix = False
 
-    def fit(self, interaction_matrix=True, one_vs_all=True):
-        if interaction_matrix:
+    def fit(self, pairwise_matrix=True, one_vs_all=True):
+        if pairwise_matrix:
             logging.info("\nH-Index interaction matrix: Start fitting")
             for i in range(self.nof_features):
                 print("Feature: ", i)
@@ -45,9 +58,9 @@ class HIndex:
     def plot(self, interaction_matrix=True, one_vs_all=True):
         # if not fitted, fit
         if not self.fitted_interaction_matrix and interaction_matrix:
-            self.fit(interaction_matrix=True, one_vs_all=False)
+            self.fit(pairwise_matrix=True, one_vs_all=False)
         if not self.fitted_one_vs_all_matrix and one_vs_all:
-            self.fit(interaction_matrix=False, one_vs_all=True)
+            self.fit(pairwise_matrix=False, one_vs_all=True)
 
         # plot
         if interaction_matrix:
@@ -89,6 +102,7 @@ class HIndex:
         return interaction
 
     def _pdp_1d(self, feature, x):
+        # eval normalized pdp for feature at x
[…]

### 2023-05-24 pythia/pdp.py

@@ -283,11 +283,34 @@ class PDPwithdICE:
                          scale_x=scale_x, scale_y=scale_y, savefig=savefig)
 
 
-def pdp_1d_non_vectorized(model, data, x, feature, uncertainty, is_jac):
+def pdp_1d_non_vectorized(model: callable,
+                          data: np.ndarray,
+                          x: np.ndarray,
+                          feature: int,
+                          uncertainty: bool,
+                          is_jac: bool
+                          ) -> typing.Union[np.ndarray, typing.Tuple[np.ndarray, np.ndarray, np.ndarray]]:
+    """Computes the unnormalized 1-dimensional PDP, in a non-vectorized way.
+
+    Args:
+        model (callable): model to be explained
+        data (np.ndarray): dataset, shape (N, D)
+        x (np.ndarray): values of the feature to be explained, shape (K,)
+        feature (int): index of the feature to be explained
+        uncertainty (bool): whether to compute the uncertainty of the PDP
+        is_jac (bool): whether the model returns the prediction (False) or the Jacobian wrt the input (True)
+
+    Returns:
+        if uncertainty is False:
+            np.ndarray: unnormalized 1-dimensional PDP
+        if uncertainty is True:
+            a tuple of three np.ndarray: (unnormalized 1-dimensional PDP, standard deviation, standard error)
+    """
+    K = data.shape[0]
     mean_pdp = []
     sigma_pdp = []
     stderr = []
-    for i in range(x.shape[0]):
+    for i in range(K):
         x_new = copy.deepcopy(data)
         x_new[:, feature] = x[i]
         y = model(x_new)[:, feature] if is_jac else model(x_new)
@@ -300,6 +323,23 @@ def pdp_1d_non_vectorized(model, data, x, feature, uncertainty, is_jac):
 
 
 def pdp_1d_vectorized(model, data, x, feature, uncertainty, is_jac):
+    """Computes the unnormalized 1-dimensional PDP, in a vectorized way.
+
+    Args:
+        model (callable): model to be explained
+        data (np.ndarray): dataset, shape (N, D)
+        x (np.ndarray): values of the feature to be explained, shape (K,)
+        feature (int): index of the feature to be explained
+        uncertainty (bool): whether to compute the uncertainty of the PDP
+        is_jac (bool): whether the model returns the prediction (False) or the Jacobian wrt the input (True)
+
+    Returns:
+        if uncertainty is False:
[…]

### 2023-05-24 examples/synthetic_repid.py
added PDP-based interaction

@@ -10,6 +10,11 @@ import matplotlib.pyplot as plt
 from sklearn.model_selection import train_test_split
 from interpret.glassbox import ExplainableBoostingRegressor
 
+# hack to reload modules
+import importlib
+pythia = importlib.reload(pythia)
+
+
 class RepidSimpleDist:
     """
     x1 ~ U(-1, 1)
@@ -76,12 +81,14 @@ Y = model.predict(X)
 
 # check interactions
 h_index = interaction.HIndex(data=X, model=model.predict, nof_instances=950)
-# print(h_index.pairwise(0, 1))
-# print(h_index.one_vs_all(0))
+# print(h_index.eval_pairwise(0, 1))
+print(h_index.eval_one_vs_all(0))
+h_index.plot(interaction_matrix=False, one_vs_all=True)
 
-h_index.plot(interaction_matrix=True, one_vs_all=True)
-
-
+# check interactions with REPID (dPDP based method)
+repid_index = interaction.REPID(data=X, model=model.predict, model_jac=model.jacobian, nof_instances=950)
+print(repid_index.eval_one_vs_all(0))
+repid_index.plot()
 
 
 

### 2023-05-24 pythia/interaction.py

@@ -1,8 +1,9 @@
 import numpy as np
 import logging
+import typing
 from pythia import helpers
 import pythia
-from pythia import dPDP
+from pythia import dPDP, PDP
 from pythia.pdp import pdp_1d_vectorized, pdp_1d_non_vectorized, pdp_nd_non_vectorized, pdp_nd_vectorized
 import tqdm
 
@@ -69,6 +70,17 @@ class HIndex:
             self._plot_one_vs_all_matrix()
 
     def eval_pairwise(self, feat_1, feat_2):
+        """Evaluate the interaction between feature 1 and feature 2
+
+        Args
+        ---
+            feat_1 (int): The index of feature 1
+            feat_2 (int): The index of feature 2
+
+        Returns
+        ---
+            float: The interaction between feature 1 and feature 2
+        """
         if self.interaction_matrix[feat_1, feat_2] != self.empty_symbol:
             return self.interaction_matrix[feat_1, feat_2]
 
@@ -156,11 +168,14 @@ class HIndex:
 
 
 class REPID:
+    """REPID definition of interaction between feature x_j and all the others, i.e. the variance of the dPDP plot
+    """
     empty_symbol = 1e10
 
     def __init__(self, data, model, model_jac, nof_instances):
         self.nof_instances, self.indices = helpers.prep_nof_instances(nof_instances, data.shape[0])
         self.data = data[self.indices]
+        self.dim = self.data.shape[1]
 
         self.model = model
         self.model_jac = model_jac
@@ -173,27 +188,28 @@ class REPID:
         self.fitted_interaction_matrix = False
         self.fitted_one_vs_all_matrix = False
 
-    def fit(self):
-        for i in range(self.nof_features):
-            print("Feature: ", i)
-            self._fit_one_vs_all(i)
+    def fit(self, features: typing.Union[int, str, list] = "all",):
+        features = helpers.prep_features(features, self.dim)
+        for i, feat in enumerate(features):
+            print("Feature: ", feat)
+            self.eval_one_vs_all(feat)
         self.fitted_one_vs_all_matrix = True
 
-    def _fit_one_vs_all(self, feat):
+    def eval_one_vs_all(self, feat):
         if self.one_vs_all_matrix[feat] != self.empty_symbol:
             return self.one_vs_all_matrix[feat]
 
         axis_limits = pythia.helpers.axis_limits_from_data(self.data)
[…]

### 2023-05-24 pythia/pdp.py

@@ -322,7 +322,13 @@ def pdp_1d_non_vectorized(model: callable,
     return (np.array(mean_pdp), np.array(sigma_pdp), np.array(stderr)) if uncertainty else np.array(mean_pdp)
 
 
-def pdp_1d_vectorized(model, data, x, feature, uncertainty, is_jac):
+def pdp_1d_vectorized(model: callable,
+                          data: np.ndarray,
+                          x: np.ndarray,
+                          feature: int,
+                          uncertainty: bool,
+                          is_jac: bool
+                          ) -> typing.Union[np.ndarray, typing.Tuple[np.ndarray, np.ndarray, np.ndarray]]:
     """Computes the unnormalized 1-dimensional PDP, in a vectorized way.
 
     Args:
@@ -358,18 +364,52 @@ def pdp_1d_vectorized(model, data, x, feature, uncertainty, is_jac):
         return mean_pdp
 
 
-def pdp_nd_non_vectorized(model, data, x, features, uncertainty, is_jac):
+def pdp_nd_non_vectorized(model: callable,
+                          data: np.ndarray,
+                          x: np.ndarray,
+                          features: list,
+                          uncertainty: bool,
+                          is_jac: bool
+                          ) -> typing.Union[np.ndarray, typing.Tuple[np.ndarray, np.ndarray, np.ndarray]]:
+    """Computes the unnormalized n-dimensional PDP, in a non-vectorized way.
+
+    Args:
+        model (callable): model to be explained
+        data (np.ndarray): dataset, shape (N, D)
+        x (np.ndarray): values of the features to be explained, shape (K, D)
+        features (list): indices of the features to be explained
+        uncertainty (bool): whether to compute the uncertainty of the PDP
+        is_jac (bool): whether the model returns the prediction (False) or the Jacobian wrt the input (True)
+
+    Returns:
+        if uncertainty is False:
+            np.ndarray: unnormalized n-dimensional PDP
+        if uncertainty is True:
+            a tuple of three np.ndarray: (unnormalized n-dimensional PDP, standard deviation, standard error)
+    """
     assert len(features) == x.shape[1]
+
+    # nof positions to evaluate the PDP
+    K = data.shape[0]
 
     mean_pdp = []
     sigma_pdp = []
     stderr = []
-    for i in range(x.shape[0]):
+    for k in range(K):
+        # take all dataset
         x_new = copy.deepcopy(data)
-        for j in range(len(features)):
[…]

### 2023-05-24 report/report.tex
added some research ideas

\documentclass[12pt]{article}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{bm}

\title{From global to regional effects: a comparison of the different approaches}

\newcommand{\Rd}{\mathbb{R}^d}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\xc}{\mathbf{x_c}}
\newcommand{\fxc}{f^{(\xc)}}
\newcommand{\fxs}{f^{(x_s)}}
\newcommand{\Xcal}{\mathcal{X}}
\newcommand{\Ycal}{\mathcal{Y}}

\author{Vasilis Gkolemis}
\begin{document}
\maketitle
\section{Introduction}

\section{Background}

Let \(\Xcal \in \Rd\) be the \(d\)-dimensional feature space, \(\Ycal\) the target space and \(f(\cdot) : \Xcal \rightarrow \Ycal\) the black-box function.  We use index \(s \in \{1, \ldots, d\}\) for the feature of interest and \(c = \{1, \ldots, d\} - s\) for the rest.  For convenience, to denote the input vector, we use \((x_s, \xc)\) instead
of \((x_1, \cdots , x_s, \cdots, x_D)\) and, for random variables,
\((X_s, X_c)\) instead of \((X_1, \cdots , X_s, \cdots, X_D)\).  The
training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) is sampled
i.i.d.\ from the distribution \(\mathbb{P}_{X,Y}\).  Finally,
\(f^{\mathtt{<method>}}(x_s)\) denotes how \(\mathtt{<method>}\)
defines the feature effect and \(\hat{f}^{\mathtt{<method>}}(x_s)\)
how it estimates it from the training set.

\section{Feature Effect}

The purpose of any feature effect (FE) method is to explain the
`black-box` function $f: \mathbb{R}^D \rightarrow \mathbb{R}$
using a Generalized Additive Model $f_{\mathtt{<method>}}(x) = c + f_1(x_1) + \cdots + f_D(x_D)$, as a global surrogate.

\subsection{Approaches}

\begin{table}[htbp]
    \centering
    \caption{Table Caption}
    \begin{tabular}{c|c|c}
      \hline
      \textbf{Name} & \textbf{Definition} \((f)\) & \textbf{Approximation} \((\hat{f})\)\\
      \hline
      \textbf{PDP} & $\mathbb{E}_{X_c}[f(x_s,X_c)]$ & $\frac{1}{N} \sum f(x_s,x_c^{(i)})$ \\
      \textbf{dPDP} & $\mathbb{E}_{X_c}[ \frac{\partial f(x_s,X_c)}{\partial x_s}]$ & $\frac{1}{N} \sum \frac{\partial f(x_s,x_c^{(i)})}{\partial x_s}$ \\
    \end{tabular}
\end{table}

    \section{Interaction Index}
    \subsection{Approaches}
    \section{Regional Effects}
    \subsection{Approaches}

    \section{Can we evaluate the approaches?}

    \subsection{Idea 1}

    We may split every $f: \mathbb{R}^D \rightarrow \mathbb{R}$ into
    a model without interaction between $\xc$ and $x_s$,
    i.e., $f_{ni}(\xb) = \fxs(x_s) +  \fxc(\xc)$,
    and the interaction term $\kappa(\xc, x_s)$:

    \[
      f(\xb) = \underbrace{\fxs(x_s) +  \fxc(\xc)}_{f_{ni}(\xb)} + \kappa(\xc, x_s)
    \]

    A simple approach is defining \(f\) to be a Neural Network and \(f_{ni}\) a Neural Additive Model without interaction
    between \(x_s\) and \(\xc\). Then \(\kappa(\xc, x_s) = f(\xb) - f_{ni}(\xb)\) and we quantify the importance of \(\kappa\) as \(\mathbb{E}_{X_c, X_s} \left [ |\kappa(X_c, X_s)| \right ] \approx \sqrt{\frac{1}{N} \sum_i \kappa^2(\xc, x_s)} \).

    \subsection{Idea 2}
\end{document}

### 2023-05-24 examples/real_bike_sharing.py

@@ -1,6 +1,5 @@
 import sys, os
 import timeit
-
 sys.path.append(os.path.dirname(os.getcwd()))
 import pythia
 import pythia.interaction as interaction
@@ -141,7 +140,7 @@ model = keras.Sequential([
 ])
 
 model.compile(optimizer="adam", loss="mse", metrics=["mae", keras.metrics.RootMeanSquaredError()])
-model.fit(X_train, Y_train, epochs=1, verbose=1)
+model.fit(X_train, Y_train, epochs=3, verbose=1)
 model.evaluate(X_test, Y_test, verbose=1)
 
 
@@ -157,8 +156,8 @@ def model_forward(x):
     return model(x).numpy().squeeze()
 
 # Explain
-# h_index = interaction.HIndex(data=X_train.to_numpy(), model=model_forward, nof_instances=100)
-# h_index.plot()
+h_index = interaction.HIndex(data=X_train.to_numpy(), model=model_forward, nof_instances=100)
+h_index.plot()
 
 repid = interaction.REPID(data=X_train.to_numpy(), model=model_forward, model_jac=model_jac, nof_instances=100)
 repid.plot()

### 2023-06-06 report/report.tex
updated example

@@
 \author{Vasilis Gkolemis}
+
 \begin{document}
 \maketitle
+
+\begin{abstract}
+Partial dependence (PD) plots are an established tool for visualizing main effects from general black box
+models. In recent years they have been supplemented with individual conditional expectation plots (ICE
+plots). Furthermore, they have been fundamentally criticized as being invalid for correlated features,
+and average local effects plots (ALE plots) have been proposed as a remedy. This paper discusses the
+properties of PD plots, ICE plots and ALE plots both in terms of their estimands for linear models with
+interactions and in terms of their performance for nonparametric models that do not extrapolate well. A
+stratified PD plot is introduced, which is particularly useful for the visualization of interactions between
+correlated features. Recommendations for the use of model-agnostic effects plots are given, with special
+emphasis to nonparametric machine learning models.
+\end{abstract}
+
 \section{Introduction}
@@
 We use index \(s \in \{1, \ldots, d\}\) for the feature of interest and \(c = \{1, \ldots, d\} - s\) for the rest.
-For convenience, to denote the input vector, we use \((x_s, \xc)\) instead
-of \((x_1, \cdots , x_s, \cdots, x_D)\) and, for random variables,
-\((X_s, X_c)\) instead of \((X_1, \cdots , X_s, \cdots, X_D)\).  The
-training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) is sampled
+For convenience, we use \((x_s, \xc)\) to refer to \((x_1, \cdots , x_s, \cdots, x_D)\) and, equivalently, \((X_s, X_c)\) instead of \((X_1, \cdots , X_s, \cdots, X_D)\) when we refer to random variables.
+The training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) is sampled
 i.i.d.\ from the distribution \(\mathbb{P}_{X,Y}\).  Finally,
@@
 
-The purpose of any feature effect (FE) method is to explain the
-`black-box` function $f: \mathbb{R}^D \rightarrow \mathbb{R}$
-using a Generalized Additive Model $f_{\mathtt{<method>}}(x) = c + f_1(x_1) + \cdots + f_D(x_D)$, as a global surrogate.
+Feature Effect (FE) methods explain the `black-box` function $f: \mathbb{R}^D \rightarrow \mathbb{R}$
+with a set of $1D$ mappings $f_s(x_s): \mathbb{R} \rightarrow \mathbb{R}$,
+each one being the effect of the $s$-th feature on the output.
+For example, [running example]..
+
+In this study, we propose interpreting FE as global surrogate models,
+i.e., as algorithms that take as input (a) the black-box function $f$ and (b) the training set $\mathcal{D}$ and
+they output a generalized additive model $f_{\mathtt{FE}}(\xb) = c + \sum_{s=1}^D f_s(x_s)$.
+Using this perspective, we will discuss the different approaches for defining the feature effect and we will
+propose a quantitative framework for evaluating them.
 
 \subsection{Approaches}
+
+At Table~\ref{tab:fe-methods}, we present the most used feature effect methods, namely,
+(a) the Partial Dependence Plot (PDP),
+(b) the derivative of the PDP (d-PDP)
+(c) the M-Plots and
+(d) the Accumulated Local Effects (ALE) plots.
 
@@
     \caption{Table Caption}
+    \label{tab:fe-methods}
     \begin{tabular}{c|c|c}
       \hline
-      \textbf{Name} & \textbf{Definition} \((f)\) & \textbf{Approximation} \((\hat{f})\)\\
+      \textbf{Name} & \textbf{Definition} \(f(x_s)\) & \textbf{Approximation} \(\hat{f}(x_s)\)\\
       \hline
       \textbf{PDP} & $\mathbb{E}_{X_c}[f(x_s,X_c)]$ & $\frac{1}{N} \sum f(x_s,x_c^{(i)})$ \\
-      \textbf{dPDP} & $\mathbb{E}_{X_c}[ \frac{\partial f(x_s,X_c)}{\partial x_s}]$ & $\frac{1}{N} \sum \frac{\partial f(x_s,x_c^{(i)})}{\partial x_s}$ \\
+      \hline
+      \textbf{d-PDP} & $\mathbb{E}_{X_c}[ \frac{\partial f(x_s,X_c)}{\partial x_s}]$ & $\frac{1}{N} \sum \frac{\partial f(x_s,x_c^{(i)})}{\partial x_s}$ \\
+      \hline
+      \textbf{M-Plot} & $\mathbb{E}_{X_c|x_s}[f(x_s,X_c)]$ & $\frac{1}{N} \sum_{i: x_s^{i} \in B(x_s)} f(x_s,x_c^{(i)})$ \\
+      \hline
+      \textbf{ALE} & \(\int_{x_{s,\min}}^{x_s} \mathbb{E}_{X_c|X_s=z}\left [f^s (z, X_c)\right ] \partial z\) & $\Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^{(i)} \in
+                                                                                                                \mathcal{S}_k} f^s(\mathbf{x}^i)$ \\
+      \hline
     \end{tabular}
 \end{table}
+
+Global surrogate methods are evaluated based on their fidelity, i.e.,
+how well they can replicate the underlying black-box function.
 
@@
 
-    \section{Can we evaluate the approaches?}
+    \section{A unified evaluation framework}
 

### 2023-06-06 examples/synthetic_repid.py

@@ -79,41 +79,75 @@ X = dist.generate(N=1000)
 Y = model.predict(X)
 
 
-# check interactions
-h_index = interaction.HIndex(data=X, model=model.predict, nof_instances=950)
-# print(h_index.eval_pairwise(0, 1))
-print(h_index.eval_one_vs_all(0))
-h_index.plot(interaction_matrix=False, one_vs_all=True)
-
-# check interactions with REPID (dPDP based method)
-repid_index = interaction.REPID(data=X, model=model.predict, model_jac=model.jacobian, nof_instances=950)
-print(repid_index.eval_one_vs_all(0))
-repid_index.plot()
-
-
-
-# # find regions
-# reg = pythia.regions.Regions(data=X, model=model.predict, model_jac=model.jacobian, cat_limit=25)
-# reg.search_splits(nof_levels=2, nof_candidate_splits=20, criterion="rhale")
-# opt_splits = reg.choose_important_splits(0.2)
+# # check interactions
+# h_index = interaction.HIndex(data=X, model=model.predict, nof_instances=950)
+# # print(h_index.eval_pairwise(0, 1))
+# print(h_index.eval_one_vs_all(0))
+# h_index.plot(interaction_matrix=False, one_vs_all=True)
 #
-# transf = pythia.regions.DataTransformer(splits=opt_splits)
-# new_X = transf.transform(X)
+# # check interactions with REPID (dPDP based method)
+# repid_index = interaction.REPID(data=X, model=model.predict, model_jac=model.jacobian, nof_instances=950)
+# print(repid_index.eval_one_vs_all(0))
+# repid_index.plot()
+
+# find regions
+reg = pythia.regions.Regions(data=X, model=model.predict, model_jac=model.jacobian, cat_limit=25)
+reg.search_splits(nof_levels=2, nof_candidate_splits=20, criterion="rhale")
+opt_splits = reg.choose_important_splits(0.2)
+
+transf = pythia.regions.DataTransformer(splits=opt_splits)
+new_X = transf.transform(X)
+
+# split data
+seed = 21
+X_train, X_test, y_train, y_test = train_test_split(X, Y, test_size=0.20, random_state=seed)
+X_train_new, X_test_new, y_train_new, y_test_new = train_test_split(new_X, Y, test_size=0.20, random_state=seed)
+
+
+# fit a GAM to the transformed data
[…]

### 2023-06-06 examples/synthetic_repid_third_degree.py
added example where EBM fails

@@ -10,6 +10,7 @@ import matplotlib.pyplot as plt
 from sklearn.model_selection import train_test_split
 from interpret.glassbox import ExplainableBoostingRegressor
 
+
 # hack to reload modules
 import importlib
 pythia = importlib.reload(pythia)
@@ -34,24 +35,21 @@ class RepidSimpleDist:
                              np.random.uniform(-1, 1., size=int(N-2)),
                              np.array([1])))
         x3 = np.random.choice([0, 1], int(N), p=[0.5, 0.5])
+        x4 = np.random.choice([0, 1], int(N), p=[0.5, 0.5])
 
-        x = np.stack((x1, x2, x3), axis=-1)
+        x = np.stack((x1, x2, x3, x4), axis=-1)
         return x
 
 
 class RepidSimpleModel:
-    def __init__(self, a1=8, a2=-16):
+    def __init__(self, a1=100):
         self.a1 = a1
-        self.a2 = a2
 
     def predict(self, x):
         y = np.zeros_like(x[:,0])
 
-        cond = x[:, 0] > 0
+        cond = np.logical_and(x[:, 0] > 0, x[:, 2] == 0)
         y[cond] += self.a1*x[cond, 1]
-
-        cond = x[:, 2] == 0
-        y[cond] += self.a2*x[cond, 1]
 
         eps = np.random.normal(loc=0, scale=0.1, size=y.shape[0])
         y += eps
@@ -60,11 +58,8 @@ class RepidSimpleModel:
     def jacobian(self, x):
         y = np.zeros_like(x)
 
-        cond = x[:, 0] > 0
+        cond = np.logical_and(x[:, 0] > 0, x[:, 2] == 0)
         y[cond, 1] += self.a1
-
-        cond = x[:, 2] == 0
-        y[cond, 1] += self.a2
         return y
 
 
@@ -89,18 +84,8 @@ X = dist.generate(N=1000)
 Y = model.predict(X)
 
 
-# # check interactions
-# h_index = interaction.HIndex(data=X, model=model.predict, nof_instances=950)
-# # print(h_index.eval_pairwise(0, 1))
-# print(h_index.eval_one_vs_all(0))
-# h_index.plot(interaction_matrix=False, one_vs_all=True)
-#
-# # check interactions with REPID (dPDP based method)
-# repid_index = interaction.REPID(data=X, model=model.predict, model_jac=model.jacobian, nof_instances=950)
-# print(repid_index.eval_one_vs_all(0))
[…]

### 2023-06-07 report/report.tex
added examples

--- 2023-05-24 report/report.tex
@@
 \usepackage{bm}
+\usepackage{bbm}
+\usepackage{graphicx}
+\usepackage{geometry}[margin=1in]
 
-\title{From global to regional effects: a comparison of the different approaches}
+\title{Regionally Additive Models: Explainable-by-design models minimizing feature interactions}
 
@@
 \newcommand{\Ycal}{\mathcal{Y}}
+\newcommand{\when}[1]{\mathbbm{1}_{#1}}
 
 \author{Vasilis Gkolemis}
+
 \begin{document}
 \maketitle
+
+\begin{abstract}
+Generalized Additive Models (GAMs) are a popular class of explainable-by-design models that are widely used in practice.
+GAMs are based on the assumption that the effect of each feature on the target is independent of the values of the
+other features, however, in cases where this assumption is violated they may lead to poor performance.
+To address this limitation we propose Regionally Additive Models (RAMs), a novel class of explainable-by-design models,
+that fits multiple GAMs to subregions of the feature space where interactions are minimized.
+Our approach consists of two steps: first, we fit a black-box model and we identify the subregions where the black-box model is nearly locally additive,
+i.e., where the effect of each feature on the target is independent of the values of the other features.
+Secondly, we train a GAM specifically for each identified subregion.
+
+We show that RAMs are more expressive than GAMs while they are still interpretable.
+
+\end{abstract}
+
 \section{Introduction}
+
+To motivate about the use of RAMs, consider the black-box function
+\(f(\xb) = 8x_2\when{x_1 > 0}\when{x_3=0}\) with \(x_1, x_2 \sim \mathcal{U}(-1,1)\) and \(x_3 \sim Bernoulli(0,1)\).
+The black-box function $f$ involves an interaction term between the three features.
+In principle, such interactions cannot be captured by a GAM which is a linear combination of univariate functions.
+However, the black-box if we split the input space in two regions we observe that $f$ is locally additive
+in each one, i.e.,
+
+\begin{equation}
+    f(\xb) = \begin{cases} 8x_2 & \text{if } x_1 > 0 \text{ and } x_3 = 0 \\ 0 & \text{otherwise} \end{cases}
+\end{equation}
+%
+which makes it trivial to fit a separate GAMs in each subregion.
+
+Correctly identifying the regions where the black-box function is locally additive is the key idea behind RAMs.
+This can be handled
+
+\begin{figure}[htbp]
+    \centering
+    \includegraphics[width=0.32\textwidth]{figures/global_GAM.pdf}
+    \includegraphics[width=0.32\textwidth]{figures/regional_gam_subreg_1.pdf}
+    \includegraphics[width=0.32\textwidth]{figures/regional_gam_subreg_2.pdf}
+    \caption{Caption}
+\end{figure}
 
@@
 
-\section{Feature Effect}
+\section{RAM: Regionally Additive Models}
 
-The purpose of any feature effect (FE) method is to explain the
-`black-box` function $f: \mathbb{R}^D \rightarrow \mathbb{R}$
-using a Generalized Additive Model $f_{\mathtt{<method>}}(x) = c + f_1(x_1) + \cdots + f_D(x_D)$, as a global surrogate.
+%Feature Effect (FE) methods explain the `black-box` function $f: \mathbb{R}^D \rightarrow \mathbb{R}$
+%with a set of $1D$ mappings $f_s(x_s): \mathbb{R} \rightarrow \mathbb{R}$,
+%each one being the effect of the $s$-th feature on the output.
+%For example, [running example]..
+%
+%In this study, we propose interpreting FE as global surrogate models,
+%i.e., as algorithms that take as input (a) the black-box function $f$ and (b) the training set $\mathcal{D}$ and
+%they output a generalized additive model $f_{\mathtt{FE}}(\xb) = c + \sum_{s=1}^D f_s(x_s)$.
+%Using this perspective, we will discuss the different approaches for defining the feature effect and we will
+%propose a quantitative framework for evaluating them.
+%
+%\subsection{Approaches}
+%
+%At Table~\ref{tab:fe-methods}, we present the most used feature effect methods, namely,
+%(a) the Partial Dependence Plot (PDP),
+%(b) the derivative of the PDP (d-PDP)
+%(c) the M-Plots and
+%(d) the Accumulated Local Effects (ALE) plots.
+%
+%\begin{table}[htbp]
+%    \centering
+%    \caption{Table Caption}
+%    \label{tab:fe-methods}
+%    \begin{tabular}{c|c|c}
+%      \hline
+%      \textbf{Name} & \textbf{Definition} \(f(x_s)\) & \textbf{Approximation} \(\hat{f}(x_s)\)\\
+%      \hline
+%      \textbf{PDP} & $\mathbb{E}_{X_c}[f(x_s,X_c)]$ & $\frac{1}{N} \sum f(x_s,x_c^{(i)})$ \\
+%      \hline
+%      \textbf{d-PDP} & $\mathbb{E}_{X_c}[ \frac{\partial f(x_s,X_c)}{\partial x_s}]$ & $\frac{1}{N} \sum \frac{\partial f(x_s,x_c^{(i)})}{\partial x_s}$ \\
+%      \hline
+%      \textbf{M-Plot} & $\mathbb{E}_{X_c|x_s}[f(x_s,X_c)]$ & $\frac{1}{N} \sum_{i: x_s^{i} \in B(x_s)} f(x_s,x_c^{(i)})$ \\
+%      \hline
+%      \textbf{ALE} & \(\int_{x_{s,\min}}^{x_s} \mathbb{E}_{X_c|X_s=z}\left [f^s (z, X_c)\right ] \partial z\) & $\Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^{(i)} \in
+%                                                                                                                \mathcal{S}_k} f^s(\mathbf{x}^i)$ \\
+%      \hline
+%    \end{tabular}
+%\end{table}
+%
+%Global surrogate methods are evaluated based on their fidelity, i.e.,
+%how well they can replicate the underlying black-box function.
+%
+%    \section{Interaction Index}
+%    \subsection{Approaches}
+%    \section{Regional Effects}
+%    \subsection{Approaches}
+%
+%    \section{A unified evaluation framework}
+%
+%    \subsection{Idea 1}
+%
+%    We may split every $f: \mathbb{R}^D \rightarrow \mathbb{R}$ into
+%    a model without interaction between $\xc$ and $x_s$,
+%    i.e., $f_{ni}(\xb) = \fxs(x_s) +  \fxc(\xc)$,
+%    and the interaction term $\kappa(\xc, x_s)$:
+%
+%    \[
+%      f(\xb) = \underbrace{\fxs(x_s) +  \fxc(\xc)}_{f_{ni}(\xb)} + \kappa(\xc, x_s)
+%    \]
+%
+%    A simple approach is defining \(f\) to be a Neural Network and \(f_{ni}\) a Neural Additive Model without interaction
+%    between \(x_s\) and \(\xc\).
+Then \(\kappa(\xc, x_s) = f(\xb) - f_{ni}(\xb)\) and we quantify the importance of \(\kappa\) as \(\mathbb{E}_{X_c, X_s} \left [ |\kappa(X_c, X_s)| \right ] \approx \sqrt{\frac{1}{N} \sum_i \kappa^2(\xc, x_s)} \).
+%
+%    \subsection{Idea 2}
 
-\subsection{Approaches}
+\section{Synthetic Examples}
 
-\begin{table}[htbp]
-    \centering
-    \caption{Table Caption}
-    \begin{tabular}{c|c|c}
-      \hline
-      \textbf{Name} & \textbf{Definition} \((f)\) & \textbf{Approximation} \((\hat{f})\)\\
-      \hline
-      \textbf{PDP} & $\mathbb{E}_{X_c}[f(x_s,X_c)]$ & $\frac{1}{N} \sum f(x_s,x_c^{(i)})$ \\
-      \textbf{dPDP} & $\mathbb{E}_{X_c}[ \frac{\partial f(x_s,X_c)}{\partial x_s}]$ & $\frac{1}{N} \sum \frac{\partial f(x_s,x_c^{(i)})}{\partial x_s}$ \\
-    \end{tabular}
-\end{table}
+\section{Real-World Datasets}
 
-    \section{Interaction Index}
-    \subsection{Approaches}
-    \section{Regional Effects}
-    \subsection{Approaches}
-
-    \section{Can we evaluate the approaches?}
-
-    \subsection{Idea 1}
-
-    We may split every $f: \mathbb{R}^D \rightarrow \mathbb{R}$ into
-    a model without interaction between $\xc$ and $x_s$,
-    i.e., $f_{ni}(\xb) = \fxs(x_s) +  \fxc(\xc)$,
-    and the interaction term $\kappa(\xc, x_s)$:
-
-    \[
-      f(\xb) = \underbrace{\fxs(x_s) +  \fxc(\xc)}_{f_{ni}(\xb)} + \kappa(\xc, x_s)
-    \]
-
-    A simple approach is defining \(f\) to be a Neural Network and \(f_{ni}\) a Neural Additive Model without interaction
-    between \(x_s\) and \(\xc\).
-Then \(\kappa(\xc, x_s) = f(\xb) - f_{ni}(\xb)\) and we quantify the importance of \(\kappa\) as \(\mathbb{E}_{X_c, X_s} \left [ |\kappa(X_c, X_s)| \right ] \approx \sqrt{\frac{1}{N} \sum_i \kappa^2(\xc, x_s)} \).
-
-    \subsection{Idea 2}
 \end{document}

### 2023-06-07 examples/synthetic_repid.py

@@ -1,130 +1,0 @@
-import sys, os
-import timeit
-sys.path.append(os.path.dirname(os.getcwd()))
-import numpy as np
-import pythia
-import pythia.regions as regions
-import pythia.interaction as interaction
-import matplotlib.pyplot as plt
-# from nodegam.sklearn import NodeGAMClassifier, NodeGAMRegressor
-from sklearn.model_selection import train_test_split
-from interpret.glassbox import ExplainableBoostingRegressor
-
-# hack to reload modules
-import importlib
-pythia = importlib.reload(pythia)
-
-
-class RepidSimpleDist:
-    """
-    x1 ~ U(-1, 1)
-    x2 ~ U(-1, 1)
-    x3 ~ Bernoulli(0.5)
-    """
-
-    def __init__(self):
-        self.D = 2
-        self.axis_limits = np.array([[-1, 1], [-1, 1], [0, 1]]).T
-
-    def generate(self, N):
-        x1 = np.concatenate((np.array([-1]),
-                             np.random.uniform(-1, 1., size=int(N-2)),
-                             np.array([1])))
-        x2 = np.concatenate((np.array([-1]),
-                             np.random.uniform(-1, 1., size=int(N-2)),
-                             np.array([1])))
-        x3 = np.random.choice([0, 1], int(N), p=[0.5, 0.5])
-
-        x = np.stack((x1, x2, x3), axis=-1)
-        return x
-
-
-class RepidSimpleModel:
-    def __init__(self, a1=8, a2=-16):
-        self.a1 = a1
-        self.a2 = a2
-
-    def predict(self, x):
-        y = np.zeros_like(x[:,0])
-
-        cond = x[:, 0] > 0
-        y[cond] += self.a1*x[cond, 1]
-
-        cond = x[:, 2] == 0
-        y[cond] += self.a2*x[cond, 1]
-
-        eps = np.random.normal(loc=0, scale=0.1, size=y.shape[0])
-        y += eps
-        return y
-
-    def jacobian(self, x):
-        y = np.zeros_like(x)
-
-        cond = x[:, 0] > 0
-        y[cond, 1] += self.a1
-
-        cond = x[:, 2] == 0
-        y[cond, 1] += self.a2
-        return y
-
-
-def plot_effect_ebm(ebm_model, ii):
-    explanation = ebm_model.explain_global()
-    plt.figure(figsize=(10, 6))
[…]

### 2023-06-07 report/synthetic_repid.py

@@ -1,0 +1,190 @@
+import sys, os
+import timeit
+sys.path.append(os.path.dirname(os.getcwd()))
+import numpy as np
+import pythia
+import pythia.regions as regions
+import pythia.interaction as interaction
+import matplotlib.pyplot as plt
+from nodegam.sklearn import NodeGAMClassifier, NodeGAMRegressor
+from sklearn.model_selection import train_test_split
+from interpret.glassbox import ExplainableBoostingRegressor
+
+# hack to reload modules
+import importlib
+pythia = importlib.reload(pythia)
+
+
+class RepidSimpleDist:
+    """
+    x1 ~ U(-1, 1)
+    x2 ~ U(-1, 1)
+    x3 ~ Bernoulli(0.5)
+    """
+
+    def __init__(self):
+        self.D = 2
+        self.axis_limits = np.array([[-1, 1], [-1, 1], [0, 1]]).T
+
+    def generate(self, N):
+        x1 = np.concatenate((np.array([-1]),
+                             np.random.uniform(-1, 1., size=int(N-2)),
+                             np.array([1])))
+        x2 = np.concatenate((np.array([-1]),
+                             np.random.uniform(-1, 1., size=int(N-2)),
+                             np.array([1])))
+        x3 = np.random.choice([0, 1], int(N), p=[0.5, 0.5])
+
+        x = np.stack((x1, x2, x3), axis=-1)
+        return x
+
+
+class RepidSimpleModel:
+    def __init__(self, a1=8, a2=-16):
+        self.a1 = a1
+        self.a2 = a2
+
+    def predict(self, x):
+        y = np.zeros_like(x[:,0])
+
+        cond = np.logical_and(x[:, 0] > 0, x[:, 2] == 1)
+        y[cond] += self.a1*x[cond, 1]
+
+        eps = np.random.normal(loc=0, scale=0.1, size=y.shape[0])
+        y += eps
+        return y
+
+    def jacobian(self, x):
+        y = np.zeros_like(x)
+
+        cond = np.logical_and(x[:, 0] > 0, x[:, 2] == 1)
+        y[cond, 1] = self.a1
+        return y
+
+def get_effect_from_ebm(ebm_model, ii):
+    explanation = ebm_model.explain_global()
[…]

### 2023-06-07 examples/synthetic_repid_third_degree.py
not important changes

@@ -1,12 +1,16 @@
 import sys, os
 import timeit
+
+import pandas as pd
+
 sys.path.append(os.path.dirname(os.getcwd()))
 import numpy as np
 import pythia
 import pythia.regions as regions
 import pythia.interaction as interaction
 import matplotlib.pyplot as plt
-# from nodegam.sklearn import NodeGAMClassifier, NodeGAMRegressor
+from nodegam.sklearn import NodeGAMClassifier, NodeGAMRegressor
+import pandas as pd
 from sklearn.model_selection import train_test_split
 from interpret.glassbox import ExplainableBoostingRegressor
 
@@ -46,7 +50,8 @@ class RepidSimpleModel:
         self.a1 = a1
 
     def predict(self, x):
-        y = np.zeros_like(x[:,0])
+        """f(x) = a1*x2 if x1 > 0 and x3 == 0 else 0"""
+        y = np.zeros_like(x[:, 0])
 
         cond = np.logical_and(x[:, 0] > 0, x[:, 2] == 0)
         y[cond] += self.a1*x[cond, 1]
@@ -56,6 +61,7 @@ class RepidSimpleModel:
         return y
 
     def jacobian(self, x):
+        """dfdx = [0, a1, 0] if x1 > 0 and x3 == 0 else [0, 0, 0]"""
         y = np.zeros_like(x)
 
         cond = np.logical_and(x[:, 0] > 0, x[:, 2] == 0)
@@ -84,7 +90,6 @@ X = dist.generate(N=1000)
 Y = model.predict(X)
 
 
-
 # Our method
 reg = pythia.regions.Regions(data=X, model=model.predict, model_jac=model.jacobian, cat_limit=25)
 reg.search_splits(nof_levels=2, nof_candidate_splits=20, criterion="rhale")
@@ -98,14 +103,35 @@ seed = 21
 X_train, X_test, y_train, y_test = train_test_split(X, Y, test_size=0.20, random_state=seed)
 X_train_new, X_test_new, y_train_new, y_test_new = train_test_split(new_X, Y, test_size=0.20, random_state=seed)
 
+# GAM to the original data
+gam = ExplainableBoostingRegressor(interactions=0)
+gam.fit(X_train, y_train)
+print("GAM:", gam.score(X_test, y_test))
 
 # fit a GAM to the transformed data
 gam_subspaces = ExplainableBoostingRegressor(interactions=0)
[…]

### 2023-06-07 pythia/binning_methods.py

@@ -49,14 +49,14 @@ def find_limits(
 
     Parameters
     ----------
-    data: numpy array
+    data: numpy array, shape (n_samples, n_features)
         Data matrix
-    data_effect: numpy array
-        Data effect matrix
+    data_effect: numpy array, shape (n_samples, n_features)
+        Jacobian matrix
     feature: int
-        Index of the feature
+        Index of the feature of interest
     axis_limits: numpy array
-        Axis limits
+        Axis limits, shape (n_features, 2)
     binning_method: str or instance of appropriate binning class
         Binning method to use
 
@@ -67,6 +67,7 @@ def find_limits(
     """
     if isinstance(binning_method, str):
         assert binning_method in ["dp", "greedy", "fixed"]
+    assert isinstance(axis_limits, np.ndarray)
 
     if isinstance(binning_method, Fixed):
         bin_est = be.Fixed(data, data_effect, feature=feature, axis_limits=axis_limits)

### 2023-06-07 pythia/rhale.py

@@ -45,8 +45,8 @@ class RHALE(FeatureEffectBase):
 
         super(RHALE, self).__init__(axis_limits)
 
-        # init as None, it will get gradients after compile
-        self.data_effect = None if data_effect is None else data_effect
+        # if data_effect is None, it will be computed after compile
+        self.data_effect = data_effect
 
     def compile(self):
         """Prepare everything for fitting, i.e., compute the gradients on data points.

### 2023-06-07 pythia/bin_estimation.py
made none_bin_valid check to come first

@@ -80,9 +80,15 @@ class BinBase:
         Returns:
             Boolean, True if it is impossible to create any binning, False otherwise
         """
+        # if there is only one unique value, then it is impossible to create any binning
+        cond_1 = len(np.unique(self.data[:, self.feature])) == 1
 
+        # if there are less than min_points, then it is impossible to create any binning
         dy_dxs = self.data_effect[:, self.feature]
-        return dy_dxs.size < self.method_args["min_points"]
+        cond_2 = dy_dxs.size < self.method_args["min_points"]
+
+        # if either is true, then it is impossible to create any binning
+        return cond_1 or cond_2
 
     def _only_one_bin_possible(self):
         """Check if the only possible binning is all points in one bin
@@ -178,15 +184,14 @@ class Greedy(BinBase):
             "cat_limit": cat_limit,
         }
 
-        # if categorical, then return the limits
-        if self._is_categorical(cat_limit):
-            return self.limits
 
         xs_min = self.xs_min
         xs_max = self.xs_max
 
         if self._none_valid_binning():
             self.limits = False
+        elif self._is_categorical(cat_limit):
+            return self.limits
         elif self._only_one_bin_possible():
             self.limits = np.array([self.xs_min, self.xs_max])
         else:
@@ -304,17 +309,17 @@ class DP(BinBase):
 
         self.method_args = {"max_nof_bins": max_nof_bins, "min_points": min_points, "discount": discount, "cat_limit": cat_limit}
 
-        # if is categorical, then only one bin is possible
-        if self._is_categorical(cat_limit):
-            return self.limits
-
         self.min_points = min_points
         big_M = self.big_M
         nof_limits = max_nof_bins + 1
         nof_bins = max_nof_bins
 
+
+        # if is categorical, then only one bin is possible
         if self._none_valid_binning():
             self.limits = False
+        if self._is_categorical(cat_limit):
+            return self.limits
         elif self._only_one_bin_possible():
             self.limits = np.array([self.xs_min, self.xs_max])
             self.dx_list = np.array([self.xs_max - self.xs_min])
@@ -362,6 +367,10 @@ class Fixed(BinBase):
 
     def find(self, nof_bins: int, min_points: typing.Union[None, int] = 10, cat_limit: int = 10):
         self.method_args = {"nof_bins": nof_bins, "min_points": min_points, "cat_limit": cat_limit}
+
+        if self._none_valid_binning():
+            self.limits = False
+
         if self._is_categorical(cat_limit):
             return self.limits
 

### 2023-06-10 report/report.tex
started report

--- 2023-05-24 report/report.tex
@@
 \usepackage{bm}
+\usepackage{bbm}
+\usepackage{graphicx}
+\usepackage{geometry}[margin=1in]
+\usepackage{subcaption}
+\usepackage{algorithm}
+\usepackage{algorithmic}
 
-\title{From global to regional effects: a comparison of the different approaches}
+\title{Regionally Additive Models: Explainable-by-design models minimizing feature interactions}
 
@@
 \newcommand{\Ycal}{\mathcal{Y}}
+\newcommand{\when}[1]{\mathbbm{1}_{#1}}
 
 \author{Vasilis Gkolemis}
+
 \begin{document}
 \maketitle
-\section{Introduction}
 
-\section{Background}
+\begin{abstract}
+Generalized Additive Models (GAMs) are a popular class of explainable-by-design models that are widely used in practice.
+GAMs are based on the assumption that the effect of each feature on the target is independent of the values of the
+other features, however, in cases where this assumption is violated they may lead to poor performance.
+To address this limitation we propose Regionally Additive Models (RAMs), a novel class of explainable-by-design models,
+that fits multiple GAMs to subregions of the feature space where interactions are minimized.
+Our approach consists of two steps: first, we fit a black-box model and we identify the subregions where the black-box model is nearly locally additive,
+i.e., where the effect of each feature on the target is independent of the values of the other features.
+Secondly, we train a GAM specifically for each identified subregion.
+
+We show that RAMs are more expressive than GAMs while they are still interpretable.
+
+\end{abstract}
+
+\section{Introduction and Related Work}
+
+% Paragraph for Motivating about Regionally Additive Models
+Generalized Additive Models (GAMs) are a widely recognized class of explainable-by-design models.
+Their popularity stems from their seamless interpretability; since they are a linear (additive) combination of univariate functions,
+$f(\xb) = c + \sum_{s=1}^D f_s(x_s)$, each individual univariate function can be readily visualized and comprehended in isolation.
+However, the expressiveness of GAMs is limited when it comes to learn interactions between features.
+To overcome this limitation, a family of models extends GAMs enabling pairwise interactions,
+i.e., $f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{c \neq s} f_{sc}(x_s, x_c)$.
+Pairwise interactions can also be visualized and understood in isolation, so these models also maintain their explainable-by-design nature.
+Unfortunately, this does not hold for any interaction that involves more than two features, thus, the expressiveness of GAMs is limited to capturing up to two-feature interactions.
+
+% Paragraph for Motivating about Regionally Additive Models
+To overcome this limitation, we propose Regionally Additive Models (RAMs), a novel class of explainable-by-design models,
+that fits multiple GAMs to subregions of the feature space where interactions are minimized.
+Our approach consists of a three-step pipeline.
+First, we fit a black-box model to capture all high-order interactions.
+Second, we identify the subregions where the black-box model is nearly locally additive.
+Finally, we train a GAM specifically for each identified subregion.
+
+\section{Motivation}
+
+Consider the black-box function
+\(f(\xb) = 8x_2\when{x_1 > 0}\when{x_3=0}\) with \(x_1, x_2 \sim \mathcal{U}(-1,1)\) and \(x_3 \sim Bernoulli(0,1)\).
+The black-box function $f$ involves a three-term interaction term which cannot be captured by a GAM (Figure~\ref{subfig:global_gam}).
+However, if splitting the input space in two subregions we observe that $f$ is locally additive
+in each one, i.e.,
+%
+\begin{equation}
+    f(\xb) = \begin{cases} 8x_2 & \text{if } x_1 > 0 \text{ and } x_3 = 0 \\ 0 & \text{otherwise} \end{cases}
+\end{equation}
+%
+and a Regionally Additive Model (RAM)
+
+\begin{equation}
+f^{\mathtt{RAM}}(\xb) = c + \sum_{s=1}^D f^{\mathtt{RAM}}_s(x_s) = c + \sum_{s=1}^D \sum_k f_{sk}(x_s) \when{\xc \in \mathcal{R}_{sk}}
+\end{equation}
+%
+can perfectly capture the black-box function.
+In this case, $f^{\mathtt{RAM}}_1(x_1)$, $f^{\mathtt{RAM}}_3(x_3)$ are zero and $f^{\mathtt{RAM}}_2(x_2)$ has two cases;
+$f_{21}(x_2) = 8x_2$ for $\mathcal{R}_{21} = \{x_1 > 0 \text{ and } x_3 = 0\}$ and $f_{22}(x_2) = 0$ for $\mathcal{R}_{22} = \{x_1 \leq 0 \text{ or } x_3 = 1\}$.
+Importantly, RAM's expressiveness does not come at the expense of interpretability.
+We can still visualize and comprehend each univariate function in isolation, exactly as we would do with a GAM,
+with the only difference being that we have to consider the subregions where each univariate function is active, as we observe in
+Figures~\ref{subfig:regional_gam_1} and~\ref{subfig:regional_gam_2}.
+
+The key idea and the main challenge of RAMs is to identify the subregions where the black-box function is locally
+closer to additive.
+Each split adds a new univariate function to the final model and enforces the user to interpret more univariate functions to understand the model behavior.
+Thus, the number of splits should be kept as small as possible.
+
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/global_GAM.pdf}
+        \caption{$f_2(x_2)$}
+        \label{subfig:global_gam}
+    \end{subfigure}
+    \begin{subfigure}{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/regional_gam_subreg_1.pdf}
+        \caption{$f_2(x_2) \when{x_1 > 0} \when{x_3=0}$}
+        \label{subfig:regional_gam_1}
+    \end{subfigure}
+    \begin{subfigure}{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/regional_gam_subreg_2.pdf}
+        \caption{$f_2(x_2) \when{x_1 \leq 0 \text{ or } x_3 = 1}$}
+        \label{subfig:regional_gam_2}
+    \end{subfigure}
+    \caption{Caption for the entire figure}
+    \label{fig:ram_example}
+\end{figure}
+
+\section{The Regionally Additive Models (RAMs) Framework}
+
+\begin{algorithm}
+\caption{Regionally Additive Models (RAMs)}
+\label{alg:ram}
+\begin{algorithmic}[1]
+\REQUIRE Dataset $\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n$, black-box model $f(\mathbf{x})$, regional effect method $\mathcal{R}$, RAM $f^{\mathtt{RAM}}(\mathbf{x})$
+\STATE Fit a black-box model $f(\mathbf{x})$ to $\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n$
+\STATE Identify the subregions $\mathcal{R}_{sk}$ using a regional effect method
+\STATE Define the RAM $f^{\mathtt{RAM}}(\mathbf{x}) = c + \sum_{s=1}^D f^{\mathtt{RAM}}_s(x_s) = c + \sum_{s=1}^D \sum_k f_{sk}(x_s) \when{\xc \in \mathcal{R}_{sk}}$
+\STATE Fit the univariate functions $f_{sk}(x_s)$ for each $s$ and $k$.
+\end{algorithmic}
+\end{algorithm}
+
+RAM is a framework for fitting explainable-by-design models that are expressive enough to capture any interaction.
+This is achieved using a pipeline that consists of a sequence of steps (Algorithm~\ref{alg:ram}).
+In the first step, we fit a black-box model $f$ to the data.
+The black-box model can be any model that is expressive enough to accurately fit the data.
+For example, we can use a random forest or a neural network.
+In the second step, we identify the subregions where the black-box model is nearly locally additive.
+This is achieved using a regional effect method $\mathcal{R}$.
+For example, we can use REPID (add citation) or any method presented by \citet{ribeiro2016should}.
+At section \ref{sec:regional_effect_methods} we present a regional effect method based on the ALE plot.
+This step defines the form of the final model as defined in equation \ref{eq:ram}.
+Finally, we fit a GAM to the identified subregions, using any GAM model.
+
+\subsection{Background}
 
@@
 
-\section{Feature Effect}
+\subsection{Identifying the Regions of Locally Additivity}
 
-The purpose of any feature effect (FE) method is to explain the
-`black-box` function $f: \mathbb{R}^D \rightarrow \mathbb{R}$
-using a Generalized Additive Model $f_{\mathtt{<method>}}(x) = c + f_1(x_1) + \cdots + f_D(x_D)$, as a global surrogate.
+\subsection{Fitting the GAMs}
 
-\subsection{Approaches}
+\section{Experiments}
 
-\begin{table}[htbp]
-    \centering
-    \caption{Table Caption}
-    \begin{tabular}{c|c|c}
-      \hline
-      \textbf{Name} & \textbf{Definition} \((f)\) & \textbf{Approximation} \((\hat{f})\)\\
-      \hline
-      \textbf{PDP} & $\mathbb{E}_{X_c}[f(x_s,X_c)]$ & $\frac{1}{N} \sum f(x_s,x_c^{(i)})$ \\
-      \textbf{dPDP} & $\mathbb{E}_{X_c}[ \frac{\partial f(x_s,X_c)}{\partial x_s}]$ & $\frac{1}{N} \sum \frac{\partial f(x_s,x_c^{(i)})}{\partial x_s}$ \\
-    \end{tabular}
-\end{table}
+\subsection{Synthetic Examples}
 
-    \section{Interaction Index}
-    \subsection{Approaches}
-    \section{Regional Effects}
-    \subsection{Approaches}
+\subsection{Real-World Datasets}
 
-    \section{Can we evaluate the approaches?}
-
-    \subsection{Idea 1}
-
-    We may split every $f: \mathbb{R}^D \rightarrow \mathbb{R}$ into
-    a model without interaction between $\xc$ and $x_s$,
-    i.e., $f_{ni}(\xb) = \fxs(x_s) +  \fxc(\xc)$,
-    and the interaction term $\kappa(\xc, x_s)$:
-
-    \[
-      f(\xb) = \underbrace{\fxs(x_s) +  \fxc(\xc)}_{f_{ni}(\xb)} + \kappa(\xc, x_s)
-    \]
-
-    A simple approach is defining \(f\) to be a Neural Network and \(f_{ni}\) a Neural Additive Model without interaction
-    between \(x_s\) and \(\xc\).
-Then \(\kappa(\xc, x_s) = f(\xb) - f_{ni}(\xb)\) and we quantify the importance of \(\kappa\) as \(\mathbb{E}_{X_c, X_s} \left [ |\kappa(X_c, X_s)| \right ] \approx \sqrt{\frac{1}{N} \sum_i \kappa^2(\xc, x_s)} \).
-
-    \subsection{Idea 2}
 \end{document}

### 2023-06-10 report/synthetic_repid.py

@@ -40,12 +40,11 @@ class RepidSimpleDist:
 
 
 class RepidSimpleModel:
-    def __init__(self, a1=8, a2=-16):
+    def __init__(self, a1=8):
         self.a1 = a1
-        self.a2 = a2
 
     def predict(self, x):
-        y = np.zeros_like(x[:,0])
+        y = np.zeros_like(x[:, 0])
 
         cond = np.logical_and(x[:, 0] > 0, x[:, 2] == 1)
         y[cond] += self.a1*x[cond, 1]

### 2023-06-12 report/report.tex
started algorithm

@@
 \usepackage{subcaption}
+
 \usepackage{algorithm}
+\usepackage{algpseudocode}
+
+\usepackage{natbib}
 \usepackage{algorithmic}
+\bibliographystyle{plainnat}
+% set path for bibliography
 
@@
 
+\newcommand{\dfdx}{\frac{\partial f}{\partial x_s}}
 \newcommand{\Rd}{\mathbb{R}^d}
@@
 \newcommand{\xc}{\mathbf{x_c}}
+\newcommand{\xcc}{\mathbf{x_{/s}}}
 \newcommand{\fxc}{f^{(\xc)}}
@@
 Their popularity stems from their seamless interpretability; since they are a linear (additive) combination of univariate functions,
-$f(\xb) = c + \sum_{s=1}^D f_s(x_s)$, each individual univariate function can be readily visualized and comprehended in isolation.
+\(f(\xb) = c + \sum_{s=1}^D f_s(x_s)\), each individual univariate function can be readily visualized and comprehended in isolation.
 However, the expressiveness of GAMs is limited when it comes to learn interactions between features.
 To overcome this limitation, a family of models extends GAMs enabling pairwise interactions,
-i.e., $f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{c \neq s} f_{sc}(x_s, x_c)$.
+i.e., \(f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{c \neq s} f_{sc}(x_s, x_c)\).
 Pairwise interactions can also be visualized and understood in isolation, so these models also maintain their explainable-by-design nature.
@@
 \section{Motivation}
+\label{sec:motivation}
 
-Consider the black-box function
-\(f(\xb) = 8x_2\when{x_1 > 0}\when{x_3=0}\) with \(x_1, x_2 \sim \mathcal{U}(-1,1)\) and \(x_3 \sim Bernoulli(0,1)\).
-The black-box function $f$ involves a three-term interaction term which cannot be captured by a GAM (Figure~\ref{subfig:global_gam}).
-However, if splitting the input space in two subregions we observe that $f$ is locally additive
-in each one, i.e.,
+Consider the black-box function \(f(\xb) = 8x_2\when{x_1 > 0}\when{x_3=0}\)
+with \(x_1, x_2 \sim \mathcal{U}(-1,1)\) and \(x_3 \sim Bernoulli(0,1)\).
+Although very simple, the existence of the interaction term $8x_2\when{x_1 > 0}\when{x_3=0}$ that involves
+three features $x_1, x_2, x_3$ makes it impossible to be learned by a GAM.
+As we see in Figure~\ref{subfig:global_gam}, a GAM misleadingly learns that $f(x_2) = 2x_2$,
+because in $\frac{1}{4}$ of the cases ($x_1 > 0 \text{ and } x_3 = 0$) the impact of $x_2$ to the output is $8x_2$,
+and in the rest $\frac{3}{4}$ of the cases the impact of $x_2$ to the output is $0$.
+However, if splitting the input space in two subregions we observe that \(f\) is additive in each one (regionally additive):
 %
 \begin{equation}
-    f(\xb) = \begin{cases} 8x_2 & \text{if } x_1 > 0 \text{ and } x_3 = 0 \\ 0 & \text{otherwise} \end{cases}
+    \label{eq:regionally_additive}
+    f(\xb) = \begin{cases} 8x_2 & \text{if } x_1 > 0 \text{ and } x_3 = 1 \\ 0 & \text{otherwise} \end{cases}
 \end{equation}
 %
-and a Regionally Additive Model (RAM)
+Therefore, if we knew the appropriate subregions for $x_2$,
+namely, \(\mathcal{R}_{21} = \{x_1 > 0 \text{ and } x_3 = 0\}\)
+and  \(\mathcal{R}_{22} = \{x_1 \leq 0 \text{ or } x_3 = 1\}\),
+we could fit the following model to the data
 
 \begin{equation}
-f^{\mathtt{RAM}}(\xb) = c + \sum_{s=1}^D f^{\mathtt{RAM}}_s(x_s) = c + \sum_{s=1}^D \sum_k f_{sk}(x_s) \when{\xc \in \mathcal{R}_{sk}}
+    \label{eq:regional_gam}
+    f^{\mathtt{RAM}}(\xb) = f_1(x_1) + f_{21}(x_2) \when{(x_1, x_3) \in \mathcal{R}_{21}} + f_{22}(x_2) \when{(x_1, x_3) \in \mathcal{R}_{22}} + f_3(x_3)
 \end{equation}
 %
-can perfectly capture the black-box function.
-In this case, $f^{\mathtt{RAM}}_1(x_1)$, $f^{\mathtt{RAM}}_3(x_3)$ are zero and $f^{\mathtt{RAM}}_2(x_2)$ has two cases;
-$f_{21}(x_2) = 8x_2$ for $\mathcal{R}_{21} = \{x_1 > 0 \text{ and } x_3 = 0\}$ and $f_{22}(x_2) = 0$ for $\mathcal{R}_{22} = \{x_1 \leq 0 \text{ or } x_3 = 1\}$.
-Importantly, RAM's expressiveness does not come at the expense of interpretability.
+Equation~\eqref{eq:regional_gam} represents a Regionally Additive Model (RAM). A RAM is nothing more than a GAM fitted on each subregion of the feature space.
+Importantly, RAM's enhanced expressiveness does not come at the expense of interpretability.
 We can still visualize and comprehend each univariate function in isolation, exactly as we would do with a GAM,
@@
 Figures~\ref{subfig:regional_gam_1} and~\ref{subfig:regional_gam_2}.
-
-The key idea and the main challenge of RAMs is to identify the subregions where the black-box function is locally
-closer to additive.
-Each split adds a new univariate function to the final model and enforces the user to interpret more univariate functions to understand the model behavior.
-Thus, the number of splits should be kept as small as possible.
+The key challenge of RAMs is to appropriately identify the subregions where the black-box function is (close to) regionally additive.
+For this purpose, as we will see in Section~\ref{subsec:regional_effect_methods}, we propose a novel algorithm that is based on the idea of
+regional effect plots.
 
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/global_GAM.pdf}
-        \caption{$f_2(x_2)$}
+        \includegraphics[width=\textwidth]{figures/global_GAM}
+        \caption{\(f_2(x_2)\)}
         \label{subfig:global_gam}
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/regional_gam_subreg_1.pdf}
-        \caption{$f_2(x_2) \when{x_1 > 0} \when{x_3=0}$}
+        \includegraphics[width=\textwidth]{figures/regional_gam_subreg_1}
+        \caption{\(f_2(x_2) \when{x_1 > 0 \text{ and } x_3 = 1}\)}
         \label{subfig:regional_gam_1}
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/regional_gam_subreg_2.pdf}
-        \caption{$f_2(x_2) \when{x_1 \leq 0 \text{ or } x_3 = 1}$}
+        \includegraphics[width=\textwidth]{figures/regional_gam_subreg_2}
+        \caption{\(f_2(x_2) \when{x_1 \leq 0 \text{ or } x_3 \neq 1}\)}
         \label{subfig:regional_gam_2}
@@
 
-\section{The Regionally Additive Models (RAMs) Framework}
+\section{The RAM framework}
 
@@
 \begin{algorithmic}[1]
-\REQUIRE Dataset $\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n$, black-box model $f(\mathbf{x})$, regional effect method $\mathcal{R}$, RAM $f^{\mathtt{RAM}}(\mathbf{x})$
-\STATE Fit a black-box model $f(\mathbf{x})$ to $\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n$
-\STATE Identify the subregions $\mathcal{R}_{sk}$ using a regional effect method
-\STATE Define the RAM $f^{\mathtt{RAM}}(\mathbf{x}) = c + \sum_{s=1}^D f^{\mathtt{RAM}}_s(x_s) = c + \sum_{s=1}^D \sum_k f_{sk}(x_s) \when{\xc \in \mathcal{R}_{sk}}$
-\STATE Fit the univariate functions $f_{sk}(x_s)$ for each $s$ and $k$.
+\Require Dataset \(\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n\), black-box model \(f(\mathbf{x})\), regional effect method \(\mathcal{R}\), RAM \(f^{\mathtt{RAM}}(\mathbf{x})\)
+\State Fit a black-box model \(f(\mathbf{x})\) to \(\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n\)
+\State Identify the subregions \(\mathcal{R}_{sk}\) using a regional effect method
+\State Define the RAM \(f^{\mathtt{RAM}}(\mathbf{x}) = c + \sum_{s=1}^D f^{\mathtt{RAM}}_s(x_s) = c + \sum_{s=1}^D \sum_k f_{sk}(x_s) \when{\xcc \in \mathcal{R}_{sk}}\)
+\State Fit the univariate functions \(f_{sk}(x_s)\) for each \(s\) and \(k\).
 \end{algorithmic}
@@
 This is achieved using a pipeline that consists of a sequence of steps (Algorithm~\ref{alg:ram}).
-In the first step, we fit a black-box model $f$ to the data.
+In the first step, we fit a black-box model \(f\) to the data.
 The black-box model can be any model that is expressive enough to accurately fit the data.
@@
 In the second step, we identify the subregions where the black-box model is nearly locally additive.
-This is achieved using a regional effect method $\mathcal{R}$.
-For example, we can use REPID (add citation) or any method presented by \citet{ribeiro2016should}.
+This is achieved using a regional effect method \(\mathcal{R}\).
+For example, we can use REPID (add citation) or any method presented by % \citet{ribeiro2016should}.
 At section \ref{sec:regional_effect_methods} we present a regional effect method based on the ALE plot.
@@
 We use index \(s \in \{1, \ldots, d\}\) for the feature of interest and \(c = \{1, \ldots, d\} - s\) for the rest.
-For convenience, we use \((x_s, \xc)\) to refer to \((x_1, \cdots , x_s, \cdots, x_D)\) and, equivalently, \((X_s, X_c)\) instead of \((X_1, \cdots , X_s, \cdots, X_D)\) when we refer to random variables.
+For convenience, we use \((x_s, \xcc)\) to refer to \((x_1, \cdots , x_s, \cdots, x_D)\) and, equivalently, \((X_s, X_c)\) instead of \((X_1, \cdots , X_s, \cdots, X_D)\) when we refer to random variables.
 The training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) is sampled
@@
 
-\subsection{Identifying the Regions of Locally Additivity}
+\subsection{Fit a black-box function}
+
+\subsubsection{Objective}
+
+In the first step of the pipelin, a black-box function \(f(\cdot)\) is fitted to the training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\).
+In principle, the black-box function can be any model that is expressive enough to accurately fit the data,
+meaning that it is able to learn the underlying mapping \(\mathbb{P}_{Y|X}\).
+
+\subsubsection{Proposed Approach}
+
+For being able to use the DALE approximation that we propose in the next step,
+the black-box function should be differentiable.
+Recent developments have shown that differentiable Deep Learning models specifically designed for tabular
+data~\citep{arik2021tabnet} can achieve state-of-the-art performance, make them a good candidate for this step.
+In the running example, we use a simple neural network with three hidden layers as the black-box function,
+which achieves a test \(\texttt{MSE} \approx 0.01\).
+The neural network is trained using the Adam optimizer (add citation) with a learning rate of \(0.01\).
+Based on the small test \(\texttt{MSE}\), we can assume that the neural network is able to capture any interactions
+between the features.
+
+\subsection{Regions that minimize feature interactions}
+\label{subsec:regional_effect_methods}
+
+In this step, we use regional effect methods to identify the regions where the black-box function is nearly locally additive.
+
+\subsubsection{Objective}
+
+% Describe the goal of the regional effect methods
+Regional effect methods yield for each individual feature \(s\), a set of \(T_s\) non-overlapping regions,
+denoted as \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) where \(\mathcal{R}_{st} \subseteq \Xcal_{/s}\).
+We denote the number of non-overlapping regions for feature \(s\) as \(T_s\), because for each feature \(s\),
+the number of regions \(T_s\) can be different.
+The primary objective is to identify regions in which the impact of the \(s\)-th feature on the output is
+\textit{relatively independent} of the values of the other features \(\xcc\).
+To better grasp this objective, if we decompose the impact of the \(s\)-th feature on the output $y$ into two terms:
+\(f_s(x_s, \xcc) = f_{s,eff}(x_s) + f_{s, int}(x_s, \xcc)\),
+where \(f_{s,eff}(\cdot)\) represents the independent effect
+and \(f_{s, int}(\cdot)\) represents the interaction effect,
+the objective is to identify regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) where the interaction effect is minimized.
+It is important to note that the union of all regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) covers the entire feature space \(\Xcal_{/s}\).
+Regionally Additive Models (RAM) formulation is:
+
+\begin{equation}
+\label{eq:ram_formulation}
+f^{\mathtt{RAM}}(\xb) = c + \sum_{s=1}^D \sum_{t=1}^{T_s} f_{st}(x_s) \when{\xcc \in \mathcal{R}_{st}}
+\end{equation}
+%
+In the above formulation, \(f_{st}(\cdot)\) is the component of the \(s\)-th feature which is active on the \(t\)-th region.
+RAM can be viewed as a GAM with \(T_s\) components per feature where each component is applied to a specific region \(\mathcal{R}_{st}\).
+To facilitate this formulation, we introduce a new feature space \(\Xcal^\mathtt{RAM}\) defined as:
+
+\begin{equation}
+\label{eq:ram_feature_space}
+\begin{aligned}
+\Xcal^{\mathtt{RAM}} &= \{x_{st} | s \in \{1, \ldots, D\}, t \in \{1, \ldots, T_s\}\} \\
+x_{sk} &= \begin{cases}
+x_s, & \text{if } \xcc \in \mathcal{R}_{sk} \\
+0, & \text{otherwise}
+\end{cases}
+\end{aligned}
+\end{equation}
+%
+Finally, we can simply formulate RAM as a typical GAM on the extended feature space \(\Xcal^{\mathtt{RAM}}\):
+
+\begin{equation}
+\label{eq:ram_formulation2}
+f^{\mathtt{RAM}}(\xb) = c + \sum_{s,t} f_{st}(x_{st})
+\end{equation}
+%
+In the running example, to minimize the effect of feature interactions, we have to split the effect of feature \(x_2\) into two regions,
+\(\mathcal{R}_{21} = \{x_1 > 0 \text{ and } x_3 = 1\}\) and \(\mathcal{R}_{22} = \{x_1 \leq 0 \text{ or } x_3 = 0\}\).
+The RAM formulation is \(f^{\mathtt{RAM}}(\xb) = f_1(x_1) + f_{21}(x_{21}) + f_{22}(x_{22}) + f_3(x_3)\).
+
+\subsubsection{Proposed Approach}
+
+To identify the regions of the input space where the impact of feature interactions is reduced,
+we have developed a regional effect method influenced by the research conducted by
+\citet{herbinger2023decomposing} and \citet{gkolemis2023dale}.
+\citet{herbinger2023decomposing} introduced a versatile framework for detecting such regions,
+where one of the proposed methods is the Accumulated Local Effects~\citep{apley2020visualizing}.
+We have adopted their approach with two notable modifications.
+First, instead of using the ALE plot, we employ the Differential ALE (DALE) method introduced by \citet{gkolemis2023dale},
+which provides considerable computational advantages when the underlying black-box function is differentiable.
+Second, we utilize variable-size bins, instead of the fixed-size ones in DALE, because the result in a more accurate approximation.
+
+Given a black-box function \(f(\cdot)\) and a dataset \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\),
+the DALE plot is defined as:
+
+\begin{equation}  \label{eq:DALE_accumulated_mean_est}
+  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \underbrace{\frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^{(i)} \in
+    \mathcal{S}_k} \dfdx(\mathbf{x}^i)}_{\hat{\mu}(z_{k-1}, z_k)})
+\end{equation}
+%
+where \(k_x\) the index of the bin such that
+\(z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\)
+is the set of the instances of the \(k\)-th bin, i.e.
+\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^{(i)}_s < z_{k} \} \).
+DALE computes the average effect (impact) of the feature \(x_s\) on the output of the black-box function \(f(\cdot)\).
+The average effect is computed by dividing the feature space into bins and computing the average effect of
+the feature \(x_s\) in each bin.
+In cases where there are interactions between the features, the instance-level effects inside each bin
+start to deviate from the average effect.
+We can measure such deviation using the standard deviation of the instance-level effects inside each bin:
+\begin{equation}
+  \label{eq:var_bin_approx}
+  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k| - 1}
+\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
+  \hat{\mu}(z_{k-1}, z_k) \right )^2
+\end{equation}
+%
+and afterwards, the level of interaction between the feature \(x_s\) and the rest of the features can be quantified
+as the aggregated bin-deviation \(\sqrt{\sum_{k=1}^{k_x} (z_k - z_{k-1})^2 \hat{\sigma}^2(z_{k-1}, z_k)}\).
+Therefore, in order to minimize the interaction we search for a set of regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\),
+that minimize the aggregated bin-deviation defined on the subregions.
+For this reason we define the bin-effect and bin-deviation on a subregion \(\mathcal{R}_{st}\) as:
+
+\begin{equation}
+    \label{eq:bin_effect_subregion}
+    \hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}|} \sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \dfdx(\mathbf{x}^i)
+\end{equation}
+
+\begin{equation}
+  \label{eq:bin_deviation_subregion}
+  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
+\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
+    \hat{\mu}(\mathcal{R}_{st}) \right )^2
+\end{equation}
+
+Therefore, the interaction of the feature \(x_s\) with the rest of the features within a subregion \(\mathcal{R}_{st}\)
+can be quantified as:
+
+\begin{equation}
+  \label{eq:SE}
+  \mathcal{L}(\mathcal{R}_{sk}) = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
+\end{equation}
+
+For minimizing Eq.~\eqref{eq:SE}, we set up a CART-like algorithm, as proposed by \citet{herbinger2023decomposing}.
+The algorithm is described in Algorithm~\ref{alg:subregion_identification}.
+
+\begin{algorithm}
+\caption{DALE-based Region Detection}
+\label{alg:subregion_identification}
+\begin{algorithmic}
+\Procedure{DALE\_Region\_Detection}{$\mathcal{D}$, $J$, $T$}
+    \For {\(s=1, \cdots, D\)}
+        \State $\mathtt{D\_before\_split = [D]}$
+        \For {\(t=1, \cdots, T\)}
+        \EndFor
+    \EndFor
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+\caption{DALE-based Region Detection}
+\label{alg:single_feature_subregion}
+\begin{algorithmic}
+\For{\( c = 1, \cdots, s-1, s+1, \cdots, D\)}
+    \State Lalala
+    \State $\mathtt{D\_before\_split = [D]}$
+    \State $\mathtt{L\_before\_split = [}\mathcal{L}(\mathcal{D}_{s1}) \mathtt{]}$
+    \State $\mathtt{L\_after\_split = []}$
+\EndFor
+\end{algorithmic}
+\end{algorithm}
 
 \subsection{Fitting the GAMs}
+
+\subsubsection{Objective}
+
+\subsubsection{Proposed Approach}
+
+\subsection{Discussion}
+
+Recently, a number of methods have been proposed to extend traditional GAMs and make them more expressive.
+The majority of the ideas follow one of the following research directions;
+The first one targets on representing the main components of a GAM $\{ f_i(x_i) \}$ with novel models.
+For example, \citet{agarwal2021neural} who used an end-to-end neural network for learning the main components.
+The second one focuses on extending the GAMs to model interactions between the features.
+For example, \citet{lou2013accurate} proposed Explainable Boosting Machines (EBMs) which are a generalized additive model with pairwise interaction terms.
+It is worth noting that the proposed method is orthogonal to the aforementioned research directions.
+As we will show in the experiments, the proposed method can be used in conjunction with any of the
+aforementioned methods to improve the accuracy of the resulting model, while maintaining the interpretability of the model.
 
@@
 
+\bibliography{report}
+
 \end{document}

### 2023-06-12 report/synthetic_repid.py

@@ -92,98 +92,105 @@ model = RepidSimpleModel()
 X = dist.generate(N=1000)
 Y = model.predict(X)
 
-# # check interactions
-# h_index = interaction.HIndex(data=X, model=model.predict, nof_instances=950)
-# # print(h_index.eval_pairwise(0, 1))
-# print(h_index.eval_one_vs_all(0))
-# h_index.plot(interaction_matrix=False, one_vs_all=True)
+# fit a neural network
+from tensorflow import keras
+
+nn_model = keras.Sequential([
+    keras.layers.Dense(100, input_shape=(3,), activation="relu"),
+    keras.layers.Dense(100, activation="relu"),
+    keras.layers.Dense(100, activation="relu"),
+    keras.layers.Dense(1)
+])
+
+optimizer = keras.optimizers.Adam(learning_rate=0.0001)
+nn_model.compile(optimizer=optimizer, loss="mse")
+nn_model.fit(X, Y, epochs=10000, batch_size=1000)
+
+xx = np.array([-.5, 2., 1])
+print("NN prediction: ", nn_model.predict(xx.reshape(1, -1)))
+print("Ground truth: ", model.predict(xx.reshape(1, -1)))
+
+# # find regions
+# reg = pythia.regions.Regions(data=X, model=model.predict, model_jac=model.jacobian, cat_limit=25)
+# reg.find_splits(nof_levels=2, nof_candidate_splits=20, method="rhale")
+# opt_splits = reg.choose_important_splits(0.2)
 #
-# # check interactions with REPID (dPDP based method)
-# repid_index = interaction.REPID(data=X, model=model.predict, model_jac=model.jacobian, nof_instances=950)
-# print(repid_index.eval_one_vs_all(0))
-# repid_index.plot()
-
-# find regions
-reg = pythia.regions.Regions(data=X, model=model.predict, model_jac=model.jacobian, cat_limit=25)
-reg.find_splits(nof_levels=2, nof_candidate_splits=20, method="rhale")
-opt_splits = reg.choose_important_splits(0.2)
-
-transf = pythia.regions.DataTransformer(splits=opt_splits)
-new_X = transf.transform(X)
-
-# split data
-seed = 21
[…]

### 2023-06-13 report/report.tex
continued algorithm

@@
 
-\usepackage{algorithm}
-\usepackage{algpseudocode}
+% for algorithms
+\usepackage[linesnumbered,ruled]{algorithm2e}
 
 \usepackage{natbib}
-\usepackage{algorithmic}
 \bibliographystyle{plainnat}
@@
 
-\begin{algorithm}
-\caption{Regionally Additive Models (RAMs)}
-\label{alg:ram}
-\begin{algorithmic}[1]
-\Require Dataset \(\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n\), black-box model \(f(\mathbf{x})\), regional effect method \(\mathcal{R}\), RAM \(f^{\mathtt{RAM}}(\mathbf{x})\)
-\State Fit a black-box model \(f(\mathbf{x})\) to \(\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n\)
-\State Identify the subregions \(\mathcal{R}_{sk}\) using a regional effect method
-\State Define the RAM \(f^{\mathtt{RAM}}(\mathbf{x}) = c + \sum_{s=1}^D f^{\mathtt{RAM}}_s(x_s) = c + \sum_{s=1}^D \sum_k f_{sk}(x_s) \when{\xcc \in \mathcal{R}_{sk}}\)
-\State Fit the univariate functions \(f_{sk}(x_s)\) for each \(s\) and \(k\).
-\end{algorithmic}
-\end{algorithm}
+%\begin{algorithm}
+%\caption{Regionally Additive Models (RAMs)}
+%\label{alg:ram}
+%\begin{algorithmic}[1]
+%\Require Dataset \(\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n\), black-box model \(f(\mathbf{x})\), regional effect method \(\mathcal{R}\), RAM \(f^{\mathtt{RAM}}(\mathbf{x})\)
+%\State Fit a black-box model \(f(\mathbf{x})\) to \(\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n\)
+%\State Identify the subregions \(\mathcal{R}_{sk}\) using a regional effect method
+%\State Define the RAM \(f^{\mathtt{RAM}}(\mathbf{x}) = c + \sum_{s=1}^D f^{\mathtt{RAM}}_s(x_s) = c + \sum_{s=1}^D \sum_k f_{sk}(x_s) \when{\xcc \in \mathcal{R}_{sk}}\)
+%\State Fit the univariate functions \(f_{sk}(x_s)\) for each \(s\) and \(k\).
+%\end{algorithmic}
+%\end{algorithm}
 
@@
 \label{alg:subregion_identification}
-\begin{algorithmic}
-\Procedure{DALE\_Region\_Detection}{$\mathcal{D}$, $J$, $T$}
-    \For {\(s=1, \cdots, D\)}
-        \State $\mathtt{D\_before\_split = [D]}$
-        \For {\(t=1, \cdots, T\)}
-        \EndFor
-    \EndFor
-\EndProcedure
-\end{algorithmic}
+\SetAlgoLined
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+\BlankLine
+\Input{X, J, T}
+\BlankLine
+\Output{optimal\_splits}
+\BlankLine
+optimal\_splits[s, c, t] $\gets$ None\;
+\For{$s = 1$ to $D$}{
+    \For{$c \in \{1, \dots, D\} \setminus \{s\}$}{
+        X\_list $\gets [X]$\;
+        J\_list $\gets [J]$\;
+        \For{$t = 1$ to $T$}{
+            best\_split $\gets$ BestSplit(X\_list, J\_list, s)\;
+            optimal\_splits[s, c, t] $\gets$ best\_split\;
+            X\_list $\gets$ best\_split["X\_list"]\;
+            J\_list $\gets$ best\_split["J\_list"]\;
+        }
+    }
+}
 \end{algorithm}
@@
 \begin{algorithm}
-\caption{DALE-based Region Detection}
+\caption{BestSplit}
 \label{alg:single_feature_subregion}
-\begin{algorithmic}
-\For{\( c = 1, \cdots, s-1, s+1, \cdots, D\)}
-    \State Lalala
-    \State $\mathtt{D\_before\_split = [D]}$
-    \State $\mathtt{L\_before\_split = [}\mathcal{L}(\mathcal{D}_{s1}) \mathtt{]}$
-    \State $\mathtt{L\_after\_split = []}$
-\EndFor
-\end{algorithmic}
+\SetAlgoLined
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+
+\Input{X\_list, J\_list, s, c}
+\Output{BestSplits}
+
+% Initialize variables
+\BlankLine
+\uIf{c is categorical}{
+    positions $\gets$ GetUniqueValues(X\_list, c)\;
+    L $\gets$ []\; % \Comment{List to store interaction values}
+    \For{p \textnormal{in} positions}{
+        X\_split, J\_split $\gets$ SplitDatasetCategorical(X\_list, J\_list, c, p)\;
+        L $\gets$ GetInteraction(X\_split, J\_split)\;
+    }
+    split\_position = argmin(L)
+}
+\Else{
+    positions $\gets$ GetPosition(X\_list)\;
+    L $\gets$ []\; % \Comment{List to store interaction values}
+    \For{p \textnormal{in} positions}{
+        X\_split, J\_split $\gets$ SplitDatasetNumerical(X\_list, J\_list, c, p)\;
+        L $\gets$ GetInteraction(X\_split, J\_split)\;
+    }
+    split\_position = argmin(L) % \Comment{Store the minimum interaction value for the categorical feature}
+}
+
+\Return{BestSplits}
+\BlankLine
+\end{algorithm}
+%
+%
+%
+%\begin{algorithm}
+%\caption{BestSplit}
+%\label{alg:single_feature_subregion}
+%\SetAlgoLined
+%\SetKwInOut{Input}{Input}
+%\SetKwInOut{Output}{Output}
+%
+%\Input{X\_list, J\_list, s, t}
+%\Output{Splits\_list}
+%
+%% Initialize variables
+%\BlankLine
+%\For{c \in \{1, \dots, D\}}{
+%    \If{c is categorical}{
+%        unique\_values $\gets$ GetUniqueValues(X\_list, c)\;
+%        L $\gets$ []\;
+%        \For{p in unique\_values}{
+%            X\_split, J\_split $\gets$ SplitDatasetCategorical(X\_list, J\_list, c, p)\;
+%            L $\gets$ GetInteraction(X\_split, J\_split)\;
+%        }
+%        L[c] = min(L)\;
+%    }
+%    \Else{
+%        positions $\gets$ GetPosition(X\_list)\;
+%        L $\gets$ []\;
+%        \For{p in postions}{
+%            X\_split, J\_split $\gets$ SplitDatasetNumerical(X\_list, J\_list, c, p)\;
+%            L $\gets$ GetInteraction(X\_split, J\_split)\;
+%        }
+%    }
+%}
+%\BlankLine
+%\end{algorithm}
+
+\begin{algorithm}
+\caption{SplitDatasetCategorical}
+\label{alg:split_dataset_categorical}
+\SetAlgoLined
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+\Input{X\_list, J\_list, c, val}
+\Output{Split datasets}
+X\_split $\gets$ []\;
+J\_split $\gets$ []\;
+\For{i = 1 to len(X\_list)}{
+    ind\_equal $\gets$ X[:, c] $=$ val\;
+    ind\_diff $\gets$ X[:, c] $\neq$ val\;
+    Append X[ind\_equal], X[ind\_diff] to X\_split\;
+    Append J[ind\_equal], J[ind\_diff] to J\_split\;
+}
+\Return X\_split, J\_split
+\end{algorithm}
+
+\begin{algorithm}
+\caption{SplitDatasetNumerical}
+\label{alg:split_dataset_numerical}
+\SetAlgoLined
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+\Input{X\_list, J\_list, c, val}
+\Output{Split datasets}
+X\_split $\gets$ []\;
+J\_split $\gets$ []\;
+\For{i = 1 to len(X\_list)}{
+    ind\_left $\gets$ X[:, c] $\leq$ val\;
+    ind\_right $\gets$ X[:, c] $>$ val\;
+    Append X[ind\_left], X[ind\_right] to X\_split\;
+    Append J[ind\_left], J[ind\_right] to J\_split\;
+}
+\Return X\_split, J\_split
+\end{algorithm}
+
+\begin{algorithm}
+\caption{GetInteraction}
+\label{alg:get_interaction}
+\SetAlgoLined
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+\Input{$X\_list$, $J\_list$, $min\_points$}
+\Output{weighted average of interaction levels}
+$N \gets$ total number of items in $X\_list$\;
+$W \gets []$\;
+$L \gets []$\;
+\For{$i$ in len($X\_list$)}{
+    $X \gets X\_list[i]$\;
+    $J \gets J\_list[i]$\;
+    \uIf{$|X| < min\_points$}{
+      Append $\infty$ to $L$\;
+    }
+    \Else{
+        Append $\mathcal{L}(X, J)$ to $L$\;
+    }
+    Append $\frac{|X|}{N}$ to $W$\;
+}
+\Return $\text{sum}(L \cdot W)$\;
 \end{algorithm}

### 2023-06-13 pythia/regions.py

@@ -81,7 +81,6 @@ class Regions:
         # update state
         self.regions_found = True
 
-
     def choose_important_splits(self, heter_thres=0.1, pcg=0.2):
         assert self.regions_found is True
         optimal_splits = {}

### 2023-06-13 report/report.tex
updated

--- 2023-06-12 report/report.tex
@@
 \section{The RAM framework}
-
-\begin{algorithm}
-\caption{Regionally Additive Models (RAMs)}
-\label{alg:ram}
-\begin{algorithmic}[1]
-\Require Dataset \(\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n\), black-box model \(f(\mathbf{x})\), regional effect method \(\mathcal{R}\), RAM \(f^{\mathtt{RAM}}(\mathbf{x})\)
-\State Fit a black-box model \(f(\mathbf{x})\) to \(\mathcal{D} = \{(\mathbf{x}_i, y_i)\}_{i=1}^n\)
-\State Identify the subregions \(\mathcal{R}_{sk}\) using a regional effect method
-\State Define the RAM \(f^{\mathtt{RAM}}(\mathbf{x}) = c + \sum_{s=1}^D f^{\mathtt{RAM}}_s(x_s) = c + \sum_{s=1}^D \sum_k f_{sk}(x_s) \when{\xcc \in \mathcal{R}_{sk}}\)
-\State Fit the univariate functions \(f_{sk}(x_s)\) for each \(s\) and \(k\).
-\end{algorithmic}
-\end{algorithm}
-
-RAM is a framework for fitting explainable-by-design models that are expressive enough to capture any interaction.
-This is achieved using a pipeline that consists of a sequence of steps (Algorithm~\ref{alg:ram}).
-In the first step, we fit a black-box model \(f\) to the data.
-The black-box model can be any model that is expressive enough to accurately fit the data.
-For example, we can use a random forest or a neural network.
-In the second step, we identify the subregions where the black-box model is nearly locally additive.
-This is achieved using a regional effect method \(\mathcal{R}\).
-For example, we can use REPID (add citation) or any method presented by % \citet{ribeiro2016should}.
-At section \ref{sec:regional_effect_methods} we present a regional effect method based on the ALE plot.
-This step defines the form of the final model as defined in equation \ref{eq:ram}.
-Finally, we fit a GAM to the identified subregions, using any GAM model.
 
@@
 %
-and afterwards, the level of interaction between the feature \(x_s\) and the rest of the features can be quantified
-as the aggregated bin-deviation \(\sqrt{\sum_{k=1}^{k_x} (z_k - z_{k-1})^2 \hat{\sigma}^2(z_{k-1}, z_k)}\).
+and the global interaction between the feature \(x_s\) and the rest of the features with the
+aggregated bin-deviation \({\left ( \sum_{k=1}^{k_x} (z_k - z_{k-1})^2 \hat{\sigma}^2(z_{k-1}, z_k) \right)}^{\frac{1}{2}} \).
 Therefore, in order to minimize the interaction we search for a set of regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\),
 that minimize the aggregated bin-deviation defined on the subregions.
-For this reason we define the bin-effect and bin-deviation on a subregion \(\mathcal{R}_{st}\) as:
+For this reason we define the bin-effect, the bin-deviation and the aggregated deviation on a subregion \(\mathcal{R}_{st}\) as:
 
@@
 \sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
-    \hat{\mu}(\mathcal{R}_{st}) \right )^2
+    \hat{\mu}(z_{k-1}, z_k) \right )^2
 \end{equation}
-
-Therefore, the interaction of the feature \(x_s\) with the rest of the features within a subregion \(\mathcal{R}_{st}\)
-can be quantified as:
 
@@
   \label{eq:SE}
-  \mathcal{L}(\mathcal{R}_{sk}) = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
+  \mathcal{L}_{\mathcal{R}_{sk}} = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
+\end{equation}
+
+The goal is to find the optimal set of subregions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) that minimize
+the aggregated deviation. For this reason, we define the following optimization problem:
+
+\begin{equation}
+  \label{eq:optimal_subregions}
+  \begin{aligned}
+    & \underset{\{\mathcal{R}_{st}\}_{t=1}^{T_s}}{\text{minimize}}
+    & & \sum_{t=1}^{T_s} \mathcal{L}_{\mathcal{R}_{st}} \\
+    & \text{subject to}
+    & & \bigcup_{t=1}^{T_s} \mathcal{R}_{st} = \mathcal{X} \\
+    & & & \mathcal{R}_{st} \cap \mathcal{R}_{s\tau} = \emptyset, \quad \forall t \neq \tau
+  \end{aligned}
 \end{equation}
@@
 \label{alg:subregion_identification}
-\begin{algorithmic}
-\Procedure{DALE\_Region\_Detection}{$\mathcal{D}$, $J$, $T$}
-    \For {\(s=1, \cdots, D\)}
-        \State $\mathtt{D\_before\_split = [D]}$
-        \For {\(t=1, \cdots, T\)}
-        \EndFor
-    \EndFor
-\EndProcedure
-\end{algorithmic}
+\SetAlgoLined
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+\BlankLine
+\Input{X, J, T}
+\BlankLine
+\Output{optimal\_splits}
+\BlankLine
+splits[s, c, t] $\gets$ None\;
+positions[s, c, t] $\gets$ None\;
+\For{$s = 1$ to $D$}{
+    \For{$c \in \{1, \dots, D\} \setminus \{s\}$}{
+        is\_cat = True/False \;
+        X\_list $\gets [X]$\;
+        J\_list $\gets [J]$\;
+        \For{$t = 1$ to $T$}{
+            L, position, X\_split, J\_split $\gets$ BestSplit(X\_list, J\_list, s, is\_cat)\;
+            splits[s, c, t] $\gets$ L\;
+            positions[s, c, t] $\gets$ position\;
+        }
+    }
+}
 \end{algorithm}
@@
 \begin{algorithm}
-\caption{DALE-based Region Detection}
+\caption{BestSplit}
 \label{alg:single_feature_subregion}
-\begin{algorithmic}
-\For{\( c = 1, \cdots, s-1, s+1, \cdots, D\)}
-    \State Lalala
-    \State $\mathtt{D\_before\_split = [D]}$
-    \State $\mathtt{L\_before\_split = [}\mathcal{L}(\mathcal{D}_{s1}) \mathtt{]}$
-    \State $\mathtt{L\_after\_split = []}$
-\EndFor
-\end{algorithmic}
+\SetAlgoLined
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+
+\Input{X\_list, J\_list, s, c, is\_cat}
+\Output{BestSplits}
+
+% Initialize variables
+\BlankLine
+positions $\gets$ GetPositions(X\_list, c, is\_cat)\;
+L $\gets$ []\; % \Comment{List to store interaction values}
+\For{p \textnormal{in} positions}{
+    X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c, p, is\_cat)\;
+    L $\gets$ GetInteraction(X\_split, J\_split)\;
+}
+split\_position = argmin(L)\;
+position $\gets$ positions[split\_position]\;
+X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c, position, is\_cat)\;
+\Return{L, position, X\_split, J\_split}
+\BlankLine
+\end{algorithm}
+
+\begin{algorithm}
+\caption{SplitDataset}
+\label{alg:split_dataset}
+\SetAlgoLined
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+\Input{X\_list, J\_list, c, val, is\_cat}
+\Output{X\_split, J\_split}
+X\_split $\gets$ []\;
+J\_split $\gets$ []\;
+\For{i = 1 to len(X\_list)}{
+    X= X\_list[i]\;
+    J= J\_list[i]\;
+    \uIf{is\_cat}{
+        ind\_1 $\gets$ X[:, c] $=$ val\;
+        ind\_2 $\gets$ X[:, c] $\neq$ val\;
+    }
+    \Else{
+        ind\_1 $\gets$ X[:, c] $\leq$ val\;
+        ind\_2 $\gets$ X[:, c] $>$ val\;
+    }
+    Append X[ind\_1], X[ind\_2] to X\_split\;
+    Append J[ind\_1], J[ind\_2] to J\_split\;
+}
+\Return X\_split, J\_split
+\end{algorithm}
+
+\begin{algorithm}
+\caption{GetInteraction}
+\label{alg:get_interaction}
+\SetAlgoLined
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+\Input{$X\_list$, $J\_list$, $min\_points$}
+\Output{weighted average of interaction levels}
+N $\gets$ total number of items in $X\_list$\;
+W $\gets$ []\;
+L $\gets$ []\;
+\For{$i$ in len($X\_list$)}{
+    X $\gets$ X\_list[i]\;
+    J $\gets$ J\_list[i]\;
+    \uIf{$|X|$ < min\_points}{
+      Append $\infty$ to L\;
+    }
+    \Else{
+        Append $\mathcal{L}(X, J)$ to $L$\;
+    }
+    Append $\frac{|X|}{N}$ to $W$\;
+}
+\Return $\text{sum}(L \cdot W)$\;
 \end{algorithm}
@@
 
+\section{Appendix}
+
+    \subsubsection{Algorithmic Details of Subregion Detection}
+
 \end{document}

### 2023-06-15 report/report.tex
update

@@
 
-\section{Introduction and Related Work}
+\section{Introduction}
 
 % Paragraph for Motivating about Regionally Additive Models
-Generalized Additive Models (GAMs) are a widely recognized class of explainable-by-design models.
+Generalized Additive Models (GAMs)~\citep{hastie1987generalized} are a popular class of explainable by design (x-by-design) models.
 Their popularity stems from their seamless interpretability; since they are a linear (additive) combination of univariate functions,
-\(f(\xb) = c + \sum_{s=1}^D f_s(x_s)\), each individual univariate function can be readily visualized and comprehended in isolation.
-However, the expressiveness of GAMs is limited when it comes to learn interactions between features.
-To overcome this limitation, a family of models extends GAMs enabling pairwise interactions,
+\(f(\xb) = c + \sum_{s=1}^D f_s(x_s)\), each individual univariate function (component) can be readily visualized and comprehended in isolation.
+However, GAM's main limitation is that they cannot express interactions between features.
+To mitigation this limitation, some approaches~\citep{lou2013accurate} extend them enabling pairwise interactions,
 i.e., \(f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{c \neq s} f_{sc}(x_s, x_c)\).
-Pairwise interactions can also be visualized and understood in isolation, so these models also maintain their explainable-by-design nature.
+Pairwise interactions can also be visualized and understood in isolation, so these models also maintain their x-by-design nature.
 Unfortunately, this does not hold for any interaction that involves more than two features, thus, the expressiveness of GAMs is limited to capturing up to two-feature interactions.
@@
 % Paragraph for Motivating about Regionally Additive Models
-To overcome this limitation, we propose Regionally Additive Models (RAMs), a novel class of explainable-by-design models,
+To overcome this limitation, we propose Regionally Additive Models (RAMs), a novel class of x-by-design models,
 that fits multiple GAMs to subregions of the feature space where interactions are minimized.
@@
 
-\section{Motivation}
+\section{Background and motivation}
 \label{sec:motivation}
@@
 with \(x_1, x_2 \sim \mathcal{U}(-1,1)\) and \(x_3 \sim Bernoulli(0,1)\).
-Although very simple, the existence of the interaction term $8x_2\when{x_1 > 0}\when{x_3=0}$ that involves
-three features $x_1, x_2, x_3$ makes it impossible to be learned by a GAM.
-As we see in Figure~\ref{subfig:global_gam}, a GAM misleadingly learns that $f(x_2) = 2x_2$,
+Although very simple, a GAM would fail to learn this mapping due to
+the existence of the three-features interaction term $8x_2\when{x_1 > 0}\when{x_3=0}$.
+As we see in Figure~\ref{subfig:global_gam}, a GAM misleadingly learns that $\hat{f}(\xb) \approx 2x_2$,
 because in $\frac{1}{4}$ of the cases ($x_1 > 0 \text{ and } x_3 = 0$) the impact of $x_2$ to the output is $8x_2$,
@@
 %
-Therefore, if we knew the appropriate subregions for $x_2$,
+Therefore, if we knew the appropriate subregions,
 namely, \(\mathcal{R}_{21} = \{x_1 > 0 \text{ and } x_3 = 0\}\)
 and  \(\mathcal{R}_{22} = \{x_1 \leq 0 \text{ or } x_3 = 1\}\),
-we could fit the following model to the data
+we could split the impact of $x_2$ appropriately and fit the following model to the data:
 
@@
 %
-Equation~\eqref{eq:regional_gam} represents a Regionally Additive Model (RAM). A RAM is nothing more than a GAM fitted on each subregion of the feature space.
+Equation~\eqref{eq:regional_gam} represents a Regionally Additive Model (RAM), which is simply a GAM fitted on each subregion of the feature space.
 Importantly, RAM's enhanced expressiveness does not come at the expense of interpretability.
-We can still visualize and comprehend each univariate function in isolation, exactly as we would do with a GAM,
-with the only difference being that we have to consider the subregions where each univariate function is active, as we observe in
-Figures~\ref{subfig:regional_gam_1} and~\ref{subfig:regional_gam_2}.
+As we observe in Figures~\ref{subfig:regional_gam_1} and~\ref{subfig:regional_gam_2}, we can still visualize and comprehend each univariate function in isolation, exactly as we would do with a GAM,
+with the only difference being that we have to consider the subregions where each univariate function is active,
 The key challenge of RAMs is to appropriately identify the subregions where the black-box function is (close to) regionally additive.
@@
 
-In the first step of the pipelin, a black-box function \(f(\cdot)\) is fitted to the training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\).
+In the first step of the pipeline, a black-box function \(f(\cdot)\) is fitted to the training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\).
 In principle, the black-box function can be any model that is expressive enough to accurately fit the data,
@@
 denoted as \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) where \(\mathcal{R}_{st} \subseteq \Xcal_{/s}\).
-We denote the number of non-overlapping regions for feature \(s\) as \(T_s\), because for each feature \(s\),
-the number of regions \(T_s\) can be different.
+Note that, first, we use a subscript, i.e. $T_s$, to denote that the number of non-overlapping regions can be different for each feature and, second,
+the regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) are disjoint and their union covers the entire feature space \(\Xcal_{/s}\).
 The primary objective is to identify regions in which the impact of the \(s\)-th feature on the output is
@@
 To better grasp this objective, if we decompose the impact of the \(s\)-th feature on the output $y$ into two terms:
-\(f_s(x_s, \xcc) = f_{s,eff}(x_s) + f_{s, int}(x_s, \xcc)\),
-where \(f_{s,eff}(\cdot)\) represents the independent effect
+\(f_s(x_s, \xcc) = f_{s,ind}(x_s) + f_{s, int}(x_s, \xcc)\),
+where \(f_{s,ind}(\cdot)\) represents the independent effect
 and \(f_{s, int}(\cdot)\) represents the interaction effect,
-the objective is to identify regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) where the interaction effect is minimized.
-It is important to note that the union of all regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) covers the entire feature space \(\Xcal_{/s}\).
+the objective is to identify regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) such that the interaction effect is minimized.
 Regionally Additive Models (RAM) formulation is:
@@
 RAM can be viewed as a GAM with \(T_s\) components per feature where each component is applied to a specific region \(\mathcal{R}_{st}\).
-To facilitate this formulation, we introduce a new feature space \(\Xcal^\mathtt{RAM}\) defined as:
+To facilitate this interpretation, we can define an enhanced feature space \(\Xcal^\mathtt{RAM}\) defined as:
 
@@
 %
-Finally, we can simply formulate RAM as a typical GAM on the extended feature space \(\Xcal^{\mathtt{RAM}}\):
+and then define RAM as a typical GAM on the extended feature space \(\Xcal^{\mathtt{RAM}}\):
 
@@
 
-Given a black-box function \(f(\cdot)\) and a dataset \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\),
-the DALE plot is defined as:
+\paragraph{DALE and Feature Interaction}
+
+DALE plot gets as input a black-box function \(f(\cdot)\) and a dataset \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\),
+and decomposes the \texttt{effect} (impact) of a specific feature $s$ on the output $y$.
+DALE's formulation is:
 
@@
 \( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^{(i)}_s < z_{k} \} \).
-DALE computes the average effect (impact) of the feature \(x_s\) on the output of the black-box function \(f(\cdot)\).
-The average effect is computed by dividing the feature space into bins and computing the average effect of
-the feature \(x_s\) in each bin.
-In cases where there are interactions between the features, the instance-level effects inside each bin
-start to deviate from the average effect.
-We can measure such deviation using the standard deviation of the instance-level effects inside each bin:
+DALE computes the average effect (impact) of the feature \(x_s\) on the output of the black-box function \(f(\cdot)\),
+first by dividing the feature space into bins and computing the average effect of the feature \(x_s\) in each bin
+(bin-effect) and then by aggregating the bin-level effects.
+In cases where there are strong interactions between the features, the instance-level effects inside each bin
+start to deviate from the average bin-effect.
+We can measure such deviation using the standard deviation of the instance-level effects inside each bin (bin-deviation):
+
 \begin{equation}
@@
 and the global interaction between the feature \(x_s\) and the rest of the features with the
-aggregated bin-deviation \({\left ( \sum_{k=1}^{k_x} (z_k - z_{k-1})^2 \hat{\sigma}^2(z_{k-1}, z_k) \right)}^{\frac{1}{2}} \).
+aggregated bin-deviation
+\begin{equation}
+  \label{eq:DALE_interaction}
+  \mathcal{H}_s = \sqrt{ \sum_{k=1}^{k_x} (z_k - z_{k-1})^2 \hat{\sigma}^2(z_{k-1}, z_k) }
+\end{equation}
+%
+Eq.~\eqref{eq:DALE_interaction} measures the interaction between the feature \(x_s\) and the rest of the features.
+It takes values in the range \([0, \infty)\) and it is zero when \(x_s\) does not interact with the rest of the features,
+i.e., when $f(\xb) = f_s(x_s) + f_{s}(x_s)$ and it is higher when the interaction is stronger.
+
+\paragraph{DALE for Regional Effect}
+
+If instead of applying DALE globally, we restrict it on a subregion of the feature space
+\(\mathcal{R}_{st} \subset \mathcal{X}\), we can measure the interaction between the feature \(x_s\) and the rest of the features
+inside the subregion \(\mathcal{R}_{st}\).
+In compuatational terms, this means that instead of using the whole dataset \(\mathcal{D}\) to compute the bin-effect and the bin-deviation,
+we use only the instances that belong to the subregion \(\mathcal{R}_{st}\).
+In this way, we can define $\hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k), \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)$ and $\mathcal{H}_{\mathcal{R}_{st}}$ exaclty as in
+Eq.~\eqref{eq:DALE_accumulated_mean_est}, Eq.~\eqref{eq:var_bin_approx} and Eq.~\eqref{eq:DALE_interaction} respectively,
+but using only the instances that belong to the subregion \(\mathcal{R}_{st}\), i.e. $\xb^i: x_s^i \in \mathcal{S}_k \land x_{s}^i \in \mathcal{R}_{st}$.
+Analytical formulas in the Appendix~\ref{sec:regional_dale}.
+
 Therefore, in order to minimize the interaction we search for a set of regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\),
-that minimize the aggregated bin-deviation defined on the subregions.
-For this reason we define the bin-effect, the bin-deviation and the aggregated deviation on a subregion \(\mathcal{R}_{st}\) as:
-
-\begin{equation}
-    \label{eq:bin_effect_subregion}
-    \hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}|} \sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \dfdx(\mathbf{x}^i)
-\end{equation}
-
-\begin{equation}
-  \label{eq:bin_deviation_subregion}
-  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
-\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
-    \hat{\mu}(z_{k-1}, z_k) \right )^2
-\end{equation}
-
-\begin{equation}
-  \label{eq:SE}
-  \mathcal{L}_{\mathcal{R}_{sk}} = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
-\end{equation}
-
-The goal is to find the optimal set of subregions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) that minimize
-the aggregated deviation. For this reason, we define the following optimization problem:
+that minimize:
 
@@
     & \underset{\{\mathcal{R}_{st}\}_{t=1}^{T_s}}{\text{minimize}}
-    & & \sum_{t=1}^{T_s} \mathcal{L}_{\mathcal{R}_{st}} \\
+    & & \mathcal{L} = \sum_{s=1}^S \sum_{t=1}^{T} \mathcal{H}_{\mathcal{R}_{st}} \\
     & \text{subject to}
-    & & \bigcup_{t=1}^{T_s} \mathcal{R}_{st} = \mathcal{X} \\
+    & & \bigcup_{t=1}^{T} \mathcal{R}_{st} = \mathcal{X} \\
     & & & \mathcal{R}_{st} \cap \mathcal{R}_{s\tau} = \emptyset, \quad \forall t \neq \tau
@@
 
-For minimizing Eq.~\eqref{eq:SE}, we set up a CART-like algorithm, as proposed by \citet{herbinger2023decomposing}.
-The algorithm is described in Algorithm~\ref{alg:subregion_identification}.
+For minimizing Eq.~\eqref{eq:SE}, we set up a CART-like algorithm, as proposed by \citep{herbinger2023decomposing}.
+The backbone of the algorithm is described in Algorithm~\ref{alg:subregion_identification}.
+The basic idea is for each feature $s \in \{1, \ldots, D\}$, to exhaustively search level-by-level, for
+all possible splits before choosing the one that minimizes the interaction.
+For the first level, we search
 
@@
 \BlankLine
-splits[s, c, t] $\gets$ None\;
-positions[s, c, t] $\gets$ None\;
+splits[s, t] $\gets$ None\;
+positions[s, t] $\gets$ None\;
+feat\_c[s, t] $\gets$ None\;
 \For{$s = 1$ to $D$}{
-    \For{$c \in \{1, \dots, D\} \setminus \{s\}$}{
-        is\_cat = True/False \;
     X\_list $\gets [X]$\;
@@
     \For{$t = 1$ to $T$}{
-            L, position, X\_split, J\_split $\gets$ BestSplit(X\_list, J\_list, s, is\_cat)\;
-            splits[s, c, t] $\gets$ L\;
-            positions[s, c, t] $\gets$ position\;
+            L, p, c, X\_list, J\_list $\gets$ BestSplit(X\_list, J\_list, s, is\_cat)\;
+            splits[s, t] $\gets$ L\;
+            position[s, t] $\gets$ p\;
+            feat\_c[s, t] $\gets$ c\;
         }
@@
 
+\subsection{Regional DALE formulation}
+\label{sec:regional_dale}
+
+For this reason we define the bin-effect, the bin-deviation and the aggregated deviation on a subregion \(\mathcal{R}_{st}\) as:
+
+\begin{equation}
+    \label{eq:bin_effect_subregion}
+    \hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}|} \sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \dfdx(\mathbf{x}^i)
+\end{equation}
+
+\begin{equation}
+  \label{eq:bin_deviation_subregion}
+  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
+\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
+    \hat{\mu}(z_{k-1}, z_k) \right )^2
+\end{equation}
+
+\begin{equation}
+  \label{eq:SE}
+  \mathcal{L}_{\mathcal{R}_{sk}} = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
+\end{equation}
+
     \subsubsection{Algorithmic Details of Subregion Detection}

### 2023-06-15 report/report.tex
completed the algorithmic part

@@
 For minimizing Eq.~\eqref{eq:SE}, we set up a CART-like algorithm, as proposed by \citep{herbinger2023decomposing}.
-The backbone of the algorithm is described in Algorithm~\ref{alg:subregion_identification}.
-The basic idea is for each feature $s \in \{1, \ldots, D\}$, to exhaustively search level-by-level, for
-all possible splits before choosing the one that minimizes the interaction.
-For the first level, we search
+The backbone of the algorithm is described in Algorithm~\ref{alg:subregion_detection};
+for each feature $s \in \{1, \ldots, D\}$, we exhaustively search level-by-level, for the optimal split.
+For example, suppose that the optimal first-level split for some feature $s$ is
+$x_c \pm p_1$ where $x_c$ is a numerical feature.
+The first-level split adds to the RAM to terms for feature $x_s$, $f_{s1}(x_s) \when{x_c < p_1}$ and $f_{s2}(x_s) \when{x_c \geq p_1}$,
+and splits the dataset into two subsets, $X_1 = \{ \xb^i : x_s^i \leq x_c + p_1 \}$ and $X_2 = \{ \xb^i : x_s^i > x_c + p_1 \}$.
+Afterwards, suppose that the optimal second-level split for feature $s$ is $x_c' = (\neq) p_2$ for some categorical feature $x_c'$.
+The second-level split adds to the RAM four terms for feature $x_s$, $f_{s1}(x_s) \when{x_c < p_1 \land x_c' = p_2}$,
+$f_{s2}(x_s) \when{x_c < p_1 \land x_c' \neq p_2}$, $f_{s3}(x_s) \when{x_c \geq p_1 \land x_c' = p_2}$ and $f_{s4}(x_s) \when{x_c \geq p_1 \land x_c' \neq p_2}$.
+
+The output of algorithm~\ref{alg:subregion_detection} is a set of $T$ splits per feature $s$.
+Since splitting does guarantee an important decrease in the loss function, we refine the splits by
+pruning all splits that do not decrease the loss function by more than a threshold $\epsilon$.
 
 \begin{algorithm}
-\caption{DALE-based Region Detection}
-\label{alg:subregion_identification}
+\caption{Detection of Subregions using DALE}
+\label{alg:subregion_detection}
 \SetAlgoLined
@@
 \BlankLine
-\Input{X, J, T}
+\Input{Data matrix $X$, Jacobian matrix $J$, Maximum Depth $T$}
 \BlankLine
-\Output{optimal\_splits}
+\Output{Optimal splits}
 \BlankLine
-splits[s, t] $\gets$ None\;
-positions[s, t] $\gets$ None\;
-feat\_c[s, t] $\gets$ None\;
+Initialize empty arrays for \texttt{loss}, \texttt{position}, and \texttt{feature\_c}
+\BlankLine
 \For{$s = 1$ to $D$}{
-    X\_list $\gets [X]$\;
-    J\_list $\gets [J]$\;
+X\_list, J\_list = [X], [J] \tcp*{Init lists}
 \For{$t = 1$ to $T$}{
+    \tcc{Find best split}
     L, p, c, X\_list, J\_list $\gets$ BestSplit(X\_list, J\_list, s, is\_cat)\;
-            splits[s, t] $\gets$ L\;
-            position[s, t] $\gets$ p\;
-            feat\_c[s, t] $\gets$ c\;
+    \tcc{Store best split}
+    loss[s, t], position[s, t], feature\_c[s, t] $\gets$ L, p, c
 }
 }
-}
+\Return{loss, position, feature\_c}
 \end{algorithm}
@@
 \BlankLine
+L[c, p] $\gets$ None\; % \Comment{List to store interaction values}
+\For{c $s = 1$ to $D$ \textnormal{if} c $\neq$ s}{
+    is\_cat $\gets$ IsCategorical(X\_list, c)\;
     positions $\gets$ GetPositions(X\_list, c, is\_cat)\;
-L $\gets$ []\; % \Comment{List to store interaction values}
+    L\_pos $\gets$ []\; % \Comment{List to store interaction values}
     \For{p \textnormal{in} positions}{
@@
     }
-split\_position = argmin(L)\;
-position $\gets$ positions[split\_position]\;
-X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c, position, is\_cat)\;
-\Return{L, position, X\_split, J\_split}
+}
+L\_min $\gets$ min(L)\;
+c\_min, p\_min = argmin(L)\;
+is\_cat $\gets$ IsCategorical(X\_list, c\_min)\;
+X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c\_min, p\_min, is\_cat)\;
+
+\Return{L\_min, p\_min c\_min, X\_split, J\_split}
 \BlankLine
 \end{algorithm}
+
+\paragraph{Computational Complexity}
+
+Algorithm~\ref{alg:subregion_detection} has a computational complexity of $\mathcal{O}(D \cdot T)$.
+and Algorithm~\ref{alg:single_feature_subregion} has a computational complexity of $\mathcal{O}(D-1 \cdot P \cdot N)$
+where $D$ is the number of feature, $N$ is the number of samples, $T$ is the maximum depth of the tree and $P$ is the number of query postitions.
+The computational complexity of the proposed method is therefore $\mathcal{O}(D \cdot (D-1) \cdot T \cdot P \cdot N)$.
+However, in practice, the number of query positions $P$ and the maximum depth $T$ are small numbers.
+Therefore, the computational complexity of the proposed method is $\mathcal{O}(D^2 \cdot N)$.
+
+\subsection{Fitting the GAMs}
+
+\subsubsection{Objective}
+
+\subsubsection{Proposed Approach}
+
+\subsection{Discussion}
+
+Recently, a number of methods have been proposed to extend traditional GAMs and make them more expressive.
+The majority of the ideas follow one of the following research directions;
+The first one targets on representing the main components of a GAM $\{ f_i(x_i) \}$ with novel models.
+For example, \citet{agarwal2021neural} who used an end-to-end neural network for learning the main components.
+The second one focuses on extending the GAMs to model interactions between the features.
+For example, \citet{lou2013accurate} proposed Explainable Boosting Machines (EBMs) which are a generalized additive model with pairwise interaction terms.
+It is worth noting that the proposed method is orthogonal to the aforementioned research directions.
+As we will show in the experiments, the proposed method can be used in conjunction with any of the
+aforementioned methods to improve the accuracy of the resulting model, while maintaining the interpretability of the model.
+
+\section{Experiments}
+
+\subsection{Synthetic Examples}
+
+\subsection{Real-World Datasets}
+
+\bibliography{report}
+
+\section{Appendix}
+
+\subsection{Regional DALE formulation}
+\label{sec:regional_dale}
+
+For this reason we define the bin-effect, the bin-deviation and the aggregated deviation on a subregion \(\mathcal{R}_{st}\) as:
+
+\begin{equation}
+    \label{eq:bin_effect_subregion}
+    \hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}|} \sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \dfdx(\mathbf{x}^i)
+\end{equation}
+
+\begin{equation}
+  \label{eq:bin_deviation_subregion}
+  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
+\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
+    \hat{\mu}(z_{k-1}, z_k) \right )^2
+\end{equation}
+
+\begin{equation}
+  \label{eq:SE}
+  \mathcal{L}_{\mathcal{R}_{sk}} = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
+\end{equation}
+
+\subsection{Algorithmic Details of Subregion Detection}
 
@@
 
-\subsection{Fitting the GAMs}
-
-\subsubsection{Objective}
-
-\subsubsection{Proposed Approach}
-
-\subsection{Discussion}
-
-Recently, a number of methods have been proposed to extend traditional GAMs and make them more expressive.
-The majority of the ideas follow one of the following research directions;
-The first one targets on representing the main components of a GAM $\{ f_i(x_i) \}$ with novel models.
-For example, \citet{agarwal2021neural} who used an end-to-end neural network for learning the main components.
-The second one focuses on extending the GAMs to model interactions between the features.
-For example, \citet{lou2013accurate} proposed Explainable Boosting Machines (EBMs) which are a generalized additive model with pairwise interaction terms.
-It is worth noting that the proposed method is orthogonal to the aforementioned research directions.
-As we will show in the experiments, the proposed method can be used in conjunction with any of the
-aforementioned methods to improve the accuracy of the resulting model, while maintaining the interpretability of the model.
-
-\section{Experiments}
-
-\subsection{Synthetic Examples}
-
-\subsection{Real-World Datasets}
-
-\bibliography{report}
-
-\section{Appendix}
-
-\subsection{Regional DALE formulation}
-\label{sec:regional_dale}
-
-For this reason we define the bin-effect, the bin-deviation and the aggregated deviation on a subregion \(\mathcal{R}_{st}\) as:
-
-\begin{equation}
-    \label{eq:bin_effect_subregion}
-    \hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}|} \sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \dfdx(\mathbf{x}^i)
-\end{equation}
-
-\begin{equation}
-  \label{eq:bin_deviation_subregion}
-  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
-\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
-    \hat{\mu}(z_{k-1}, z_k) \right )^2
-\end{equation}
-
-\begin{equation}
-  \label{eq:SE}
-  \mathcal{L}_{\mathcal{R}_{sk}} = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
-\end{equation}
-
-    \subsubsection{Algorithmic Details of Subregion Detection}
-
 \end{document}

### 2023-06-17 pythia/regions.py
continued the example

@@ -127,48 +127,63 @@ class DataTransformer:
         self.splits = splits
 
     def transform(self, X):
-        new_features = []
+        # feat_mapping <- to how many features each feature is mapped
+        feat_mapping = []
         for split in self.splits.values():
             if len(split) == 0:
-                new_features.append(1)
+                feat_mapping.append(1)
             else:
-                new_features.append(2**len(split))
+                feat_mapping.append(2**len(split))
 
+        # the enhaned data, without masking
         new_data = []
         for i in range(X.shape[1]):
-            new_data.append(np.repeat(X[:, i, np.newaxis], new_features[i], axis=-1))
+            new_data.append(np.repeat(X[:, i, np.newaxis], feat_mapping[i], axis=-1))
         new_data = np.concatenate(new_data, axis=-1)
 
         # create mask, based on splits
         mask = np.ones(new_data.shape)
+        new_columns = []
         for feat in range(X.shape[1]):
-            # position in new data
-            cur_pos = int(np.sum(new_features[:feat]))
+            # jj = j in the enhanced dataset
+            pos = int(np.sum(feat_mapping[:feat]))
 
-            if new_features[feat] == 1:
+            if feat_mapping[feat] == 1:
+                new_columns.append("x{}".format(feat))
                 continue
             else:
                 feat_splits = self.splits["feat_{}".format(feat)]
                 lst = [list(i) for i in itertools.product([0, 1], repeat=len(feat_splits))]
                 for ii, bin in enumerate(lst):
+                    new_name = "x{} | ".format(feat)
                     init_col_mask = np.ones(new_data.shape[0]) * True
                     for jj, b in enumerate(bin):
                         if b == 0:
                             if feat_splits[jj]["type"] == "cat":
                                 init_col_mask = np.logical_and(init_col_mask, X[:, feat_splits[jj]["feature"]] == feat_splits[jj]["position"])
+                                # add with two decimals
+                                new_name += "x{}={:.2f} & ".format(feat_splits[jj]["feature"], feat_splits[jj]["position"])
                             else:
                                 init_col_mask = np.logical_and(init_col_mask, X[:, feat_splits[jj]["feature"]] <= feat_splits[jj]["position"])
+                                new_name += "x{}<={:.2f} & ".format(feat_splits[jj]["feature"], feat_splits[jj]["position"])
                         else:
                             if feat_splits[jj]["type"] == "cat":
[…]

### 2023-06-17 report/real_bike_sharing.py
continued the example

@@ -11,6 +11,16 @@ from tensorflow import keras
 import matplotlib.pyplot as plt
 from interpret.glassbox import ExplainableBoostingRegressor
 import sklearn.metrics as metrics
+import importlib
+from sklearn.utils import check_random_state
+import random
+pythia = importlib.reload(pythia)
+
+# make everyting reproducible
+np.random.seed(42)
+tf.random.set_seed(42)
+check_random_state(42)
+random.seed(42)
 
 def preprocess(df):
     # shuffle
@@ -102,36 +112,7 @@ X_train, Y_train, X_test, Y_test = split(X_df, Y_df)
 
 cols = df.columns
 
-# # train model
-# lin_model = linear_model.LinearRegression()
-# lin_model.fit(X_train, Y_train)
-# # root mean squared error
-# print("RMSE: ", metrics.mean_squared_error(Y_train, lin_model.predict(X_train), squared=False)*y_std)
-# print("R2 score: ", lin_model.score(X_train, Y_train))
-#
-# # Same on the test set
-# print("RMSE: ", metrics.mean_squared_error(Y_test, lin_model.predict(X_test), squared=False)*y_std)
-# print("R2 score: ", lin_model.score(X_test, Y_test))
-#
-# def lin_model_jac(x):
-#     return np.ones_like(x) * lin_model.coef_
-#
-# # Explain
-# ale = pythia.ALE(data=X_train.to_numpy(), model=lin_model.predict)
-# binning_method = pythia.binning_methods.Fixed(nof_bins=30)
-# ale.fit(features="all")
-# ale.plot(feature=8)
-#
-# rhale = pythia.RHALE(data=X_train.to_numpy(), model=lin_model.predict, model_jac=lin_model_jac)
-# binning_method = pythia.binning_methods.DynamicProgramming(max_nof_bins=30, min_points_per_bin=10)
-# rhale.fit(features="all", binning_method=binning_method)
-# rhale.plot(feature=8)
-#
-# # pdp = pythia.PDP(data=X_train.to_numpy(), model=lin_model.predict)
-# # pdp.plot(feature=8)
-#
-#
-#
+# Train - Evaluate - Explain a neural network
 model = keras.Sequential([
     keras.layers.Dense(1024, activation="relu"),
     keras.layers.BatchNormalization(),
@@ -148,8 +129,9 @@ model = keras.Sequential([
     keras.layers.Dense(1)
 ])
 
[…]

### 2023-06-18 report/report.tex
continued the example
finished the example

@@
 \usepackage{subcaption}
+\usepackage{array}
 
@@
 \usepackage{natbib}
+\usepackage{multicol}
 \bibliographystyle{plainnat}
@@
 
+We evaluate the proposed approach on two typical tabular datasets, namely the Bike-Sharing Dataset~\citep{misc_bike_sharing_dataset_275}
+and the California Housing Dataset~\citep{pace1997sparse}.
+
+\paragraph{Bike-Sharing Dataset}
+
+The Bike-Sharing dataset contains the hourly bike rentals in Washington DC. over the period 2011 and 2012.
+The dataset contains 14 features and we select the 11 features that are relevant to the prediction task.
+Most of the features are measurements of the environmental conditions, like,
+month, hour, temperature, humidity, windspeed, while some others containt information about the
+day-type, e.g. if it is working-day.
+The target value \( Y_{\mathtt{count}}\) is the bike rentals per hour, which has mean value
+\(\mu_{\mathtt{count}} = 189\) and standard deviation \(\sigma_{\mathtt{count}} = 181\).
+
+% Training - Evaluation - Region Extraction
+As black-box model, we train a deep fully-connected Neural Network with 6 hidden layers,
+for \(20\) epochs using the Adam optimizer with learning rate 0.001, that
+leads a root mean squared error of \( 0.38 \cdot 181 \approx 69\) counts on the test set.
+
+We extract the subregions, setting the maximum depth of the tree to \(T=3\).
+The only feature that is split is $X_{\mathtt{hour}}$, which is split into $X_{\mathtt{hour}} | \when{X_{\mathtt{workingday}} \neq 1}$ and $X_{\mathtt{hour}} \when{X_{\mathtt{workingday} = 1}}$.
+As we observe in Figure (add citation), it is confirmed
+
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/bike_rentals_gam}
+        \caption{\(f(X_{\mathtt{hour}})\)}
+        \label{subfig:bike_rentals_gam}
+    \end{subfigure}
+    \begin{subfigure}{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/bike_rentals_ram_1}
+        \caption{\(f(X_{\mathtt{hour}}) \when{X_{\mathtt{workingday}} \neq 1}\)}
+        \label{subfig:bike_rentals_regional_gam_1}
+    \end{subfigure}
+    \begin{subfigure}{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/bike_rentals_ram_2}
+        \caption{\(f(X_{\mathtt{hour}}) \when{X_{\mathtt{workingday}} = 1}\)}
+        \label{subfig:bike_rentals_regional_gam_2}
+    \end{subfigure}
+    \caption{Caption for the entire figure}
+    \label{fig:bike_sharing}
+\end{figure}
+
+\paragraph{California Housing Dataset}
+The California Housing is a largely-studied dataset with approximately
+\(20000\) training instances, making it appropriate for robust
+approximation with large \(K\).  The dataset contains \(D=8\)
+numerical features with characteristics about the building blocks of
+California, e.g., latitude, longitude, population of the block or
+median age of houses in the block.  The target variable is the median
+value of the houses inside the block in dollars that ranges between
+\([15, 500] \cdot 10^3\), with a mean value of
+\(\mu_Y \approx 201 \cdot 10^3 \) and a standard deviation of
+\(\sigma_Y \approx 110 \cdot 10^3\). We exclude instances with missing
+or outlier values and we normalize all features to zero-mean and unit
+standard deviation. We split the dataset into \(N_{tr} = 15639\)
+training and \(N_{test} = 3910\) test examples (80/20 split) and we
+fit a Neural Network with 3 hidden layers of 256, 128 and 36 units
+respectively.  After 15 epochs using the Adam optimizer with learning
+rate \(\eta = 0.02\), the model achieves a MAE of \(37 \cdot 10^3\)
+dollars.
+
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/california_gam}
+        \caption{\(f(X_{\mathtt{longitude}})\)}
+        \label{subfig:california_gam}
+    \end{subfigure}
+    \begin{subfigure}{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/california_ram_1}
+        \caption{\(f(X_{\mathtt{longitude}}) \when{X_{\mathtt{latitude}} \leq 34.89}\)}
+        \label{subfig:california_ram_1}
+    \end{subfigure}
+    \begin{subfigure}{0.32\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/california_ram_2}
+        \caption{\(f(X_{\mathtt{longitude}}) \when{X_{\mathtt{latitude}} > 34.89}\)}
+        \label{subfig:california_ram_2}
+    \end{subfigure}
+    \caption{Caption for the entire figure}
+    \label{fig:california_housing}
+\end{figure}
+
+\begin{table}[htbp]
+  \centering
+  \caption{Result Comparison}
+  \label{tab:sample}
+  \begin{tabular}{l|c|cccc}
+      \hline
+      Type & \textbf{Black-box} & \multicolumn{4}{c}{\textbf{x-by-design}} \\
+      \hline
+      \hline
+      Interactions & \textbf{all orders} & \multicolumn{2}{c}{1\textsuperscript{st} order} & \multicolumn{2}{c}{2\textsuperscript{nd} order} \\
+      \hline
+      \hline
+      Model & \textbf{DNN} & \textbf{GAM} & \textbf{RAM} & \textbf{GAM}$^2$ & \textbf{RAM}$^2$ \\
+      \hline
+      Training & 0.135 & 0.374 & 0.274 & 0.181 & 0.176 \\
+      Test     & 0.254 & 0.549 & 0.430 & 0.298 & 0.278 \\
+      \hline
+      Training & 0.282 & 0.322 & 0.314 & 0.278 & 0.265 \\
+      Test     & 0.369 & 0.600 & 0.535 & 0.554 & 0.514 \\
+  \end{tabular}
+\end{table}
+
 \bibliography{report}

### 2023-06-18 report/california_housing.py

@@ -1,0 +1,197 @@
+import sys, os
+import timeit
+sys.path.append(os.path.dirname(os.getcwd()))
+import pythia
+import numpy as np
+import pythia.interaction as interaction
+import pythia.regions as regions
+import pandas as pd
+import tensorflow as tf
+from tensorflow import keras
+import matplotlib.pyplot as plt
+from interpret.glassbox import ExplainableBoostingRegressor
+import sklearn.metrics as metrics
+import importlib
+from sklearn.utils import check_random_state
+import random
+pythia = importlib.reload(pythia)
+import report_utils
+
+
+# make everything reproducible
+np.random.seed(4452)
+tf.random.set_seed(4542)
+check_random_state(4452)
+random.seed(4245)
+
+
+def plot_subregions(splits, feat, feat_ram, X_train, model, model_jac, gam, ram):
+    foc = splits["feat_{0}".format(feat)][0]["feature"]
+    position = splits["feat_{0}".format(feat)][0]["position"]
+    type = splits["feat_{0}".format(feat)][0]["type"]
+
+    # Regional Plot
+    if type == "cat":
+        rhale = pythia.RHALE(data=X_train.to_numpy(), model=model, model_jac=model_jac)
+        rhale_1 = pythia.RHALE(data=X_train[X_train.iloc[:, foc] == position].to_numpy(), model=model, model_jac=model_jac)
+        rhale_2 = pythia.RHALE(data=X_train[X_train.iloc[:, foc] != position].to_numpy(), model=model, model_jac=model_jac)
+    else:
+        rhale = pythia.RHALE(data=X_train.to_numpy(), model=model, model_jac=model_jac)
+        rhale_1 = pythia.RHALE(data=X_train[X_train.iloc[:, foc] <= position].to_numpy(), model=model, model_jac=model_jac)
+        rhale_2 = pythia.RHALE(data=X_train[X_train.iloc[:, foc] > position].to_numpy(), model=model, model_jac=model_jac)
+
+    rhale.fit(features=feat, binning_method=pythia.binning_methods.Greedy(), centering="zero_integral")
+    rhale.plot(feature=feat, uncertainty=True)
+    rhale_1.fit(features=feat, binning_method=pythia.binning_methods.Greedy(), centering="zero_integral")
+    rhale_1.plot(feature=feat, uncertainty=True)
[…]

### 2023-06-18 report/real_bike_sharing.py

@@ -49,53 +49,53 @@ def split(X_df, Y_df):
     return X_train, Y_train, X_test, Y_test
 
 
-def plot_subregions_rhale(feat, feature, type, position, X_train, model, model_jac):
+def plot_subregions(splits, feat, feat_ram, X_train, model, model_jac, gam, ram):
+    foc = splits["feat_{0}".format(feat)][0]["feature"]
+    position = splits["feat_{0}".format(feat)][0]["position"]
+    type = splits["feat_{0}".format(feat)][0]["type"]
+
     # Regional Plot
     if type == "cat":
         rhale = pythia.RHALE(data=X_train.to_numpy(), model=model, model_jac=model_jac)
-        rhale_1 = pythia.RHALE(data=X_train[X_train.iloc[:, feature] == position].to_numpy(), model=model, model_jac=model_jac)
-        rhale_2 = pythia.RHALE(data=X_train[X_train.iloc[:, feature] != position].to_numpy(), model=model, model_jac=model_jac)
+        rhale_1 = pythia.RHALE(data=X_train[X_train.iloc[:, foc] == position].to_numpy(), model=model, model_jac=model_jac)
+        rhale_2 = pythia.RHALE(data=X_train[X_train.iloc[:, foc] != position].to_numpy(), model=model, model_jac=model_jac)
     else:
         rhale = pythia.RHALE(data=X_train.to_numpy(), model=model, model_jac=model_jac)
-        rhale_1 = pythia.RHALE(data=X_train[X_train.iloc[:, feature] <= position].to_numpy(), model=model, model_jac=model_jac)
-        rhale_2 = pythia.RHALE(data=X_train[X_train.iloc[:, feature] > position].to_numpy(), model=model, model_jac=model_jac)
+        rhale_1 = pythia.RHALE(data=X_train[X_train.iloc[:, foc] <= position].to_numpy(), model=model, model_jac=model_jac)
+        rhale_2 = pythia.RHALE(data=X_train[X_train.iloc[:, foc] > position].to_numpy(), model=model, model_jac=model_jac)
 
-    def plot(rhale):
-        binning_method = pythia.binning_methods.Fixed(nof_bins=100)
-        rhale.fit(features=feat, binning_method=binning_method, centering="zero_integral")
[…]

### 2023-06-18 report/report_utils.py

@@ -1,0 +1,80 @@
+import pythia
+import matplotlib.pyplot as plt
+import pandas as pd
+import numpy as np
+
+def preprocess_bike(df):
+    # shuffle
+    df.sample(frac=1).reset_index(drop=True)
+
+    # Standarize X
+    X_df = df.drop(["cnt"], axis=1)
+    x_mean = X_df.mean()
+    x_std = X_df.std()
+    X_df = (X_df - X_df.mean()) / X_df.std()
+
+    # Standarize Y
+    Y_df = df["cnt"]
+    y_mean = Y_df.mean()
+    y_std = Y_df.std()
+    Y_df = (Y_df - Y_df.mean()) / Y_df.std()
+    return X_df, Y_df, x_mean, x_std, y_mean, y_std
+
+
+def preprocess_california(df):
+    # shuffle
+    df.sample(frac=1).reset_index(drop=True)
+
+    df = df.dropna()
+    X_df = df.iloc[:, :-2]
+    Y_df = df.iloc[:, -2]
+
+    # normalize
+    X_mean = X_df.mean()
+    X_std = X_df.std()
+    X_df = (X_df - X_df.mean()) / X_df.std()
+
+    Y_mean = Y_df.mean()
+    Y_std = Y_df.std()
+    Y_df = (Y_df - Y_df.mean()) / Y_df.std()
+
+    # remove points 3 std away from the mean
+    ind = (X_df.abs() > 3).sum(1) == 0
+    X_df = X_df.loc[ind, :]
+    Y_df = Y_df.loc[ind]
+
+    return X_df, Y_df, X_mean, X_std, Y_mean, Y_std
+
+
+def split(X_df, Y_df):
+    # data split
+    X_train = X_df[:int(0.8 * len(X_df))]
+    Y_train = Y_df[:int(0.8 * len(Y_df))]
+    X_test = X_df[int(0.8 * len(X_df)):]
+    Y_test = Y_df[int(0.8 * len(Y_df)):]
+    return X_train, Y_train, X_test, Y_test
+
+
+# def plot_subregions_rhale(feat, feature, type, position, X_train, model, model_jac):
+#     # Regional Plot
+#     if type == "cat":
+#         rhale = pythia.RHALE(data=X_train.to_numpy(), model=model, model_jac=model_jac)
+#         rhale_1 = pythia.RHALE(data=X_train[X_train.iloc[:, feature] == position].to_numpy(), model=model, model_jac=model_jac)
[…]

### 2023-06-19 report/report.tex
completed the Bike-sharing dataset

@@
 
-\subsection{Synthetic Examples}
-
-\subsection{Real-World Datasets}
-
 We evaluate the proposed approach on two typical tabular datasets, namely the Bike-Sharing Dataset~\citep{misc_bike_sharing_dataset_275}
@@
 
-The Bike-Sharing dataset contains the hourly bike rentals in Washington DC. over the period 2011 and 2012.
-The dataset contains 14 features and we select the 11 features that are relevant to the prediction task.
-Most of the features are measurements of the environmental conditions, like,
-month, hour, temperature, humidity, windspeed, while some others containt information about the
-day-type, e.g. if it is working-day.
+The Bike-Sharing dataset contains the hourly bike rentals in the state of Washington DC over the period 2011 and 2012.
+The dataset contains a total of 14 features, out of which 11 are selected as relevant for the purpose of prediction.
+The majority of these features involve measurements related to environmental conditions,
+such as $X_{\mathtt{month}}$, $X_{\mathtt{hour}}$, $X_{\mathtt{temperature}}$, $X_{\mathtt{humidity}}$ and $X_{\mathtt{windspeed}}$.
+Additionally, certain features provide information about the type of day, for example, whether it is a working day ($X_{\mathtt{workingday}}$) or not.
 The target value \( Y_{\mathtt{count}}\) is the bike rentals per hour, which has mean value
@@
 % Training - Evaluation - Region Extraction
-As black-box model, we train a deep fully-connected Neural Network with 6 hidden layers,
-for \(20\) epochs using the Adam optimizer with learning rate 0.001, that
-leads a root mean squared error of \( 0.38 \cdot 181 \approx 69\) counts on the test set.
+As a black-box model, we train for \(60\) epochs a fully-connected Neural Network with 6 hidden layers, using the Adam optimizer with a learning rate of $0.001$.
+The model attains a root mean squared error of \( 0.25 \cdot 181 \approx 45\) counts on the test set.
+Subsequently, we extract the subregions, searching for splits up to a maximum spliting depth of \(T=3\).
+Following the postprocessing step, we find that the only split that substantially reduces the level of interactions within the subregions is based on the feature
+$X_{\mathtt{hour}}$. This feature is divided into two subgroups: $X_{\mathtt{hour}} | \when{X_{\mathtt{workingday}} \neq 1}$ and $X_{\mathtt{hour}} \when{X_{\mathtt{workingday} = 1}}$.
 
-We extract the subregions, setting the maximum depth of the tree to \(T=3\).
-The only feature that is split is $X_{\mathtt{hour}}$, which is split into $X_{\mathtt{hour}} | \when{X_{\mathtt{workingday}} \neq 1}$ and $X_{\mathtt{hour}} \when{X_{\mathtt{workingday} = 1}}$.
-As we observe in Figure (add citation), it is confirmed
+Figure~\ref{fig:bike_sharing} clearly illustrates that the impact of the hour of the day on bike rentals varies
+significantly depending on whether it is a working day or a non-working day.
+Specifically, during working days, there is higher demand for bike rentals in the morning and afternoon hours,
+which aligns with the typical commuting times (\ref{subfig:bike_rentals_regional_gam_1}).
+On the other hand, during non-working days, bike rentals peak in the afternoon as individuals engage in
+leisure activities (\ref{subfig:bike_rentals_regional_gam_2}).
+The proposed RAM method effectively captures and detects this interaction by establishing two distinct subregions,
+each corresponding to working days and non-working days, respectively.
+Subsequently, the EBM that is fitted to each subregion, successfully learns these patterns,
+achieving a root mean squared error of approximately \( 0.43 \cdot 181 \approx 77\) counts on the test set.
+It is noteworthy that RAM not only preserves the interpretability of the model,
+but it also enhances the interpretation of the underlying modeling process.
+By identifying and highlighting the interaction between the hour of the day and the day type,
+RAM provides valuable insights into the relationship between these variables and their influence on bike rentals.
+In contrast, the GAM model~\ref{subfig:bike_rentals_gam} is not able to capture this interaction and
+achieves a root mean squared error of \( 0.55 \cdot 181 \approx 100\) counts on the test set.
 
@@
     \end{subfigure}
-    \caption{Caption for the entire figure}
+    \caption{Comparison of different models' predictions for bike rentals based on the hour of the day.
+    Subfigure (a) depicts the generalized additive model (GAM),
+        while subfigures (b) and (c) illustrate the RAM model's predictions for different day types: non-working days
+        \(f(X_{\mathtt{hour}}) \when{X_{\mathtt{workingday}} \neq 1}\) and
+        working days \(f(X_{\mathtt{hour}}) \when{X_{\mathtt{workingday}} = 1}\), respectively.
+        The RAM model successfully captures the interaction between the hour of the day and the day type,
+        leading to improved predictions and enhanced interpretability.}
     \label{fig:bike_sharing}

### 2023-06-19 report/report.tex
added abstract

@@
 \usepackage{array}
+\usepackage{multirow}
 
@@
 \begin{abstract}
-Generalized Additive Models (GAMs) are a popular class of explainable-by-design models that are widely used in practice.
-GAMs are based on the assumption that the effect of each feature on the target is independent of the values of the
-other features, however, in cases where this assumption is violated they may lead to poor performance.
-To address this limitation we propose Regionally Additive Models (RAMs), a novel class of explainable-by-design models,
-that fits multiple GAMs to subregions of the feature space where interactions are minimized.
-Our approach consists of two steps: first, we fit a black-box model and we identify the subregions where the black-box model is nearly locally additive,
-i.e., where the effect of each feature on the target is independent of the values of the other features.
-Secondly, we train a GAM specifically for each identified subregion.
-
-We show that RAMs are more expressive than GAMs while they are still interpretable.
-
+Generalized Additive Models (GAMs) are widely utilized explainable-by-design models in various applications.
+However, their assumption of independence among features can lead to suboptimal performance when violated.
+To overcome this limitation, we introduce Regionally Additive Models (RAMs), a novel class of explainable-by-design models.
+RAMs mitigate the issue by identifying subregions in the feature space where interactions are minimized and fitting multiple GAMs accordingly.
+The RAM framework consists of three steps.
+Firstly, we train a black-box model.
+Secondly, using Regional Effect Plots, we identify subregions where the black-box model exhibits near-local additivity.
+In these subregions, the effect of each feature on the target is independent of the values of other features.
+Lastly, we fit a GAM specifically for each identified subregion.
+We validate the effectiveness of RAMs through experiments on both a synthetic example and real-world datasets.
+The results confirm that RAMs offer improved expressiveness compared to GAMs while maintaining interpretability.
 \end{abstract}
@@
 
+\begin{table}[htbp]
+  \centering
+  \caption{Result Comparison}
+  \label{tab:sample}
+  \begin{tabular}{l|c|cccc}
+      \hline
+      & \textbf{Black-box} & \multicolumn{4}{c}{\textbf{x-by-design}} \\
+      \hline
+      \hline
+      & all orders & \multicolumn{2}{c}{1\textsuperscript{st} order} & \multicolumn{2}{c}{2\textsuperscript{nd} order} \\
+      \hline
+      \hline
+      & \textbf{DNN} & \textbf{GAM} & \textbf{RAM} & \textbf{GAM}$^2$ & \textbf{RAM}$^2$ \\
+      \hline
+      Bike Sharing      & 0.254 & 0.549 & 0.430 & 0.298 & 0.278 \\
+      \hline
+      California Housing      & 0.369 & 0.600 & 0.535 & 0.554 & 0.514 \\
+  \end{tabular}
+\end{table}
+
 We evaluate the proposed approach on two typical tabular datasets, namely the Bike-Sharing Dataset~\citep{misc_bike_sharing_dataset_275}
@@
 achieves a root mean squared error of \( 0.55 \cdot 181 \approx 100\) counts on the test set.
+Finally, in table~\ref{tab:sample}, we also observe that the RAM$^2$, i.e., RAM with second-order interactions,
+outperforms the equivalent GAM$^2$ model in terms of predictive performance.
 
@@
 \paragraph{California Housing Dataset}
-The California Housing is a largely-studied dataset with approximately
-\(20000\) training instances, making it appropriate for robust
-approximation with large \(K\).  The dataset contains \(D=8\)
-numerical features with characteristics about the building blocks of
-California, e.g., latitude, longitude, population of the block or
-median age of houses in the block.  The target variable is the median
-value of the houses inside the block in dollars that ranges between
-\([15, 500] \cdot 10^3\), with a mean value of
-\(\mu_Y \approx 201 \cdot 10^3 \) and a standard deviation of
-\(\sigma_Y \approx 110 \cdot 10^3\). We exclude instances with missing
-or outlier values and we normalize all features to zero-mean and unit
-standard deviation. We split the dataset into \(N_{tr} = 15639\)
-training and \(N_{test} = 3910\) test examples (80/20 split) and we
-fit a Neural Network with 3 hidden layers of 256, 128 and 36 units
-respectively.  After 15 epochs using the Adam optimizer with learning
-rate \(\eta = 0.02\), the model achieves a MAE of \(37 \cdot 10^3\)
-dollars.
+
+The California Housing dataset consists of approximately $20,000$ of housing blocks situated in California.
+Each housing block is described by eight numerical features, namely,
+$X_{\mathtt{latitude}}$, $X_{\mathtt{longitude}}$, $X_{\mathtt{median\_age}}$, $X_{\mathtt{total\_rooms}}$, $X_{\mathtt{total\_bedrooms}}$, $X_{\mathtt{population}}$, $X_{\mathtt{households}}$, and $X_{\mathtt{median\_income}}$.
+The target variable, $Y_{\mathtt{value}}$, is the median house value in dollars for each block.
+The target value ranges in the interval \([15, 500] \cdot 10^3\), with a mean value of
+\(\mu_Y \approx 201 \cdot 10^3 \) and a standard deviation of \(\sigma_Y \approx 110 \cdot 10^3\).
+
+As a black-box model, we train for $45$ epochs a fully-connected Neural Network with 6 hidden layers,
+using the Adam optimizer with a learning rate of $0.001$.
+The model achieves a root mean square error (RMSE) of about \(40\)K dollars on the test set.
+Subsequently, we perform subregion extraction by searching for splits up to a maximum depth of \(T=3\).
+After the postprocessing step, we discover that several splits significantly reduce the level of interactions,
+resulting in an expanded input space consisting of \(16\) features, as we show in table~\ref{tab:california_housing_subregions}.
+Out of them, we randomly select and illustrate in Figure~\ref{fig:california_housing} the effect of the feature $X_{\mathtt{longitude}}$.
+As we observe, for the house blocks located in the southern part of California ($X_{\mathtt{latitude}} \leq 34.9$),
+the house value decreases in an almost linear fashion as we move eastward ($X_{\mathtt{longitude}}$ increases).
+In contrast, for the house blocks located in the northern part of California ($X_{\mathtt{latitude}} > 34.9$),
+the house value performs a rapid (non-linear) decrease as we move eastward ($X_{\mathtt{longitude}}$ increases).
+We also observe that although the EBM fitted to each subregion captures the general trend,
+it does not align perfectly with the regional effect.
+As in the Bike-Sharing Example, the RMSE of the RAM model, i.e. \( 0.53 \cdot 110 \cdot 10^3 \approx 58.3 \cdot 10^3\) dollars on the test set,
+is lower than the one of the GAM model, i.e.\( 0.6 \cdot 110 \cdot 10^3 \approx 66 \cdot 10^3\) dollars.
+These results indicate that the RAM model provides superior predictions compared to the GAM model.
+The same conclusion holds is when comparing the RAM$^2$ and the GAM$^2$ models.
+\begin{table}[htbp]
+  \centering
+  \caption{California Housing: Subregions Detected by RAM}
+  \label{tab:california_housing_subregions}
+  \begin{tabular}{c|c|}
+      Feature & Subregions \\
+      \hline
+      \multirow{2}{*}{$X_{\mathtt{longitude}}$} & $X_{\mathtt{longitude}} \when{X_{\mathtt{latitude}} \leq 34.9}$ \\
+      & $X_{\mathtt{longitude}} \when{X_{\mathtt{latitude}} > 34.9}$ \\
+      \hline
+      \multirow{2}{*}{$X_{\mathtt{latitude}}$} & $X_{\mathtt{latitude}} \when{X_{\mathtt{longitude}} \leq -120.31}$ \\
+      & $X_{\mathtt{latitude}} \when{X_{\mathtt{longitude}} > -120.31}$ \\
+      \hline
+      \multirow{2}{*}{$X_{\mathtt{total\_rooms}}$} & $X_{\mathtt{total\_rooms}} \when{X_{\mathtt{total\_bedrooms}} \leq 449.37}$ \\
+        & $X_{\mathtt{total\_rooms}} \when{X_{\mathtt{total\_bedrooms}} > 449.37}$ \\
+      \hline
+      \multirow{4}{*}{$X_{\mathtt{total\_bedrooms}}$} & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} \leq 411} \when{X_{\mathtt{total\_bedrooms}} \leq 647}$ \\
+        & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} \leq 411} \when{X_{\mathtt{total\_bedrooms}} > 647}$ \\
+        & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} > 411} \when{X_{\mathtt{total\_bedrooms}} \leq 647}$ \\
+        & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} > 411} \when{X_{\mathtt{total\_bedrooms}} > 647}$ \\
+      \hline
+      \multirow{2}{*}{$X_{\mathtt{population}}$} & $X_{\mathtt{population}} \when{X_{\mathtt{households}} \leq 411.5}$ \\
+      & $X_{\mathtt{population}} \when{X_{\mathtt{households}} > 411.5}$ \\
+      \hline
+      \multirow{2}{*}{$X_{\mathtt{households}}$} & $X_{\mathtt{households}} \when{X_{\mathtt{total\_bedrooms}} \leq 630.57}$ \\
+        & $X_{\mathtt{households}} \when{X_{\mathtt{total\_bedrooms}} > 630.57}$ \\
+  \end{tabular}
+\end{table}
 
@@
 
-\begin{table}[htbp]
-  \centering
-  \caption{Result Comparison}
-  \label{tab:sample}
-  \begin{tabular}{l|c|cccc}
-      \hline
-      Type & \textbf{Black-box} & \multicolumn{4}{c}{\textbf{x-by-design}} \\
-      \hline
-      \hline
-      Interactions & \textbf{all orders} & \multicolumn{2}{c}{1\textsuperscript{st} order} & \multicolumn{2}{c}{2\textsuperscript{nd} order} \\
-      \hline
-      \hline
-      Model & \textbf{DNN} & \textbf{GAM} & \textbf{RAM} & \textbf{GAM}$^2$ & \textbf{RAM}$^2$ \\
-      \hline
-      Training & 0.135 & 0.374 & 0.274 & 0.181 & 0.176 \\
-      Test     & 0.254 & 0.549 & 0.430 & 0.298 & 0.278 \\
-      \hline
-      Training & 0.282 & 0.322 & 0.314 & 0.278 & 0.265 \\
-      Test     & 0.369 & 0.600 & 0.535 & 0.554 & 0.514 \\
-  \end{tabular}
-\end{table}
+\section{Conclusion and Future Work}
+
+In this paper we have introduced the Regional Additive Models (RAM) framework, a novel approach for learning accurate
+x-by-design models from data.
+RAMs operate by decomposing the data into subregions, where the relationship between the target variable and the
+features exhibits an approximately additive nature.
+Subsequently, Generalized Additive Models (GAMs) are fitted to each subregion and combined to create the final model.
+Our experiments on two standard regression datasets have shown promising results, indicating that RAMs can provide more accurate predictions compared to GAMs while maintaining the same level of interpretability.
+
+Nevertheless, there are still several unresolved questions that require attention and further experimentation.
+Firstly, it is essential to systematically evaluate the performance of RAMs on a larger set of datasets to ensure that the observed improvements are not specific to particular datasets.
+Secondly, we need to explore different approaches for each step of the RAM framework.
+For the initial step, we should experiment with various black-box models.
+Regarding the subregion detection step, we can explore alternative clustering algorithms.
+Finally, in the last step, we should investigate different types of GAM models to fit within each subregion.
+
+Another important area of investigation involves exploring the impact of second-order effects within the RAM framework.
+While our experimenation demonstrated that even with the current subregion detection, RAM$^2$s outperform GAM$^2$s,
+it may be the case, that for second-order models the optimal subregions are not necessarily those that maximize the additive effect of individual features,
+but rather those that maximize the additive effect of feature pairs.
 

### 2023-06-19 pythia/regions.py

@@ -57,10 +57,16 @@ class Regions:
 
         # method
         for feat in tqdm(range(self.dim)):
+            # if feature is categorical, skip
+            if self.feature_types[feat] == "cat":
+                self.splits["feat_{}".format(feat)] = []
+                continue
+
             foi, foc = feat, [i for i in range(self.dim) if i != feat]
 
             # get feature types
             foc_types = [self.feature_types[i] for i in foc]
+
 
             # get splits
             splits = find_splits(
@@ -85,6 +91,16 @@ class Regions:
         assert self.regions_found is True
         optimal_splits = {}
         for feat in range(self.dim):
+            # if split is empy, skip
+            if len(self.splits["feat_{}".format(feat)]) == 0:
+                optimal_splits["feat_{}".format(feat)] = {}
+                continue
+
+            # if initial heterogeneity is BIG_M, skip
+            if self.splits["feat_{}".format(feat)][0]["weighted_heter"] == BIG_M:
+                optimal_splits["feat_{}".format(feat)] = {}
+                continue
+
 
             # if initial heterogeneity is small right from the beginning, skip
             if self.splits["feat_{}".format(feat)][0]["weighted_heter"] < heter_thres:
@@ -313,7 +329,7 @@ def split_rhale(model: callable,
     weighted_heter = np.ones([len(foc), max(nof_splits, cat_limit)]) * big_M
 
     # limits of the feature of interest for each dataset in x_list
-    lims = np.array([pythia.helpers.axis_limits_from_data(xx)[:,foi] for xx in x_list])
+    # lims = np.array([pythia.helpers.axis_limits_from_data(xx)[:,foi] for xx in x_list])
 
     # list with len(foc) elements
     # each element is a list with the split positions for the corresponding feature of conditioning
@@ -377,6 +393,10 @@ def find_splits(nof_levels: int,
     x_jac_list = [model_jac(data)]
     splits = [{"heterogeneity": [heter_init], "weighted_heter": heter_init, "nof_instances": [len(data)], "split_i": -1, "split_j": -1, "foc": foc}]
     for lev in range(nof_levels):
+        # if any x in x_list is empty then stop
+        if any([len(x) == 0 for x in x_list]):
+            break
+
         # find split
         split_fn = split_pdp if criterion == "pdp" else split_rhale
         split = split_fn(model, model_jac, data, x_list, x_jac_list, foi, foc_types, foc, nof_splits, cat_limit, splits[-1]["heterogeneity"], min_points)

### 2023-06-20 report/report.tex
almost finished the method

@@
 \usepackage{multirow}
+\usepackage{hyperref}
 
@@
 
-\subsection{Background}
+The RAM framework consists of a three-step pipeline; (a) fit a black-box model (Section~\ref{subsec:fit_black_box}),
+(b) identify subregions with minimal interactions (Section~\ref{subsec:regional_effect_methods}) and
+(c) fit a GAM to each subregion (Section~\ref{subsec:fitting_gams}).
+Throughout this section, we will use the following notation.
+Let \(\Xcal \in \Rd\) be the \(d\)-dimensional feature space, \(\Ycal\) the target space and \(f(\cdot) : \Xcal \rightarrow \Ycal\) the black-box function.
+We use index \(s \in \{1, \ldots, d\}\) for the feature of interest and \(/s = \{1, \ldots, D\} - s\) for the rest.
+For convenience, we use \((x_s, \xcc)\) to refer to \((x_1, \cdots , x_s, \cdots, x_D)\) and, equivalently, \((X_s, X_{/s})\) instead of \((X_1, \cdots , X_s, \cdots, X_D)\) when we refer to random variables.
+The training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) is sampled
+i.i.d.\ from the distribution \(\mathbb{P}_{X,Y}\).
+%Finally, \(f^{\mathtt{<method>}}(x_s)\) denotes how \(\mathtt{<method>}\)
+%defines the feature effect and \(\hat{f}^{\mathtt{<method>}}(x_s)\)
+%how it estimates it from the training set.
 
-Let \(\Xcal \in \Rd\) be the \(d\)-dimensional feature space, \(\Ycal\) the target space and \(f(\cdot) : \Xcal \rightarrow \Ycal\) the black-box function.
-We use index \(s \in \{1, \ldots, d\}\) for the feature of interest and \(c = \{1, \ldots, d\} - s\) for the rest.
-For convenience, we use \((x_s, \xcc)\) to refer to \((x_1, \cdots , x_s, \cdots, x_D)\) and, equivalently, \((X_s, X_c)\) instead of \((X_1, \cdots , X_s, \cdots, X_D)\) when we refer to random variables.
-The training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) is sampled
-i.i.d.\ from the distribution \(\mathbb{P}_{X,Y}\).  Finally,
-\(f^{\mathtt{<method>}}(x_s)\) denotes how \(\mathtt{<method>}\)
-defines the feature effect and \(\hat{f}^{\mathtt{<method>}}(x_s)\)
-how it estimates it from the training set.
+\subsection{First step: Fit a black-box function}
+\label{subsec:fit_black_box}
 
-\subsection{Fit a black-box function}
+In the initial step of the pipeline, we fit a black-box function \(f(\cdot)\) to the training set
+\(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) to accurately learn the underlying mapping \(f(\cdot) : \Xcal \rightarrow \Ycal\).
+While any black-box function can theoretically be employed in this stage,
+for utilizing the DALE approximation, as we will show in the next step,
+it is necessary to select a differentiable function.
+Recent advancements have demonstrated that differentiable Deep Learning models,
+specifically designed for tabular data~\citep{arik2021tabnet}, are capable of achieving state-of-the-art performance,
+making them a suitable choice for this step.
 
-\subsubsection{Objective}
-
-In the first step of the pipeline, a black-box function \(f(\cdot)\) is fitted to the training set \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\).
-In principle, the black-box function can be any model that is expressive enough to accurately fit the data,
-meaning that it is able to learn the underlying mapping \(\mathbb{P}_{Y|X}\).
-
-\subsubsection{Proposed Approach}
-
-For being able to use the DALE approximation that we propose in the next step,
-the black-box function should be differentiable.
-Recent developments have shown that differentiable Deep Learning models specifically designed for tabular
-data~\citep{arik2021tabnet} can achieve state-of-the-art performance, make them a good candidate for this step.
-In the running example, we use a simple neural network with three hidden layers as the black-box function,
-which achieves a test \(\texttt{MSE} \approx 0.01\).
-The neural network is trained using the Adam optimizer (add citation) with a learning rate of \(0.01\).
-Based on the small test \(\texttt{MSE}\), we can assume that the neural network is able to capture any interactions
-between the features.
-
-\subsection{Regions that minimize feature interactions}
+\subsection{Second step: Find subregions}
 \label{subsec:regional_effect_methods}
 
-In this step, we use regional effect methods to identify the regions where the black-box function is nearly locally additive.
-
-\subsubsection{Objective}
-
+In this step, we use regional effect methods~\citep{herbinger2023decomposing, herbinger2022repid}
+to identify the regions where the black-box function is (close to) regionally additive.
 % Describe the goal of the regional effect methods
@@
 denoted as \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) where \(\mathcal{R}_{st} \subseteq \Xcal_{/s}\).
-Note that, first, we use a subscript, i.e. $T_s$, to denote that the number of non-overlapping regions can be different for each feature and, second,
-the regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) are disjoint and their union covers the entire feature space \(\Xcal_{/s}\).
+Note that, the number of non-overlapping regions can be different for each feature ($T_s$),
+the regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) are disjoint
+and their union covers the entire feature space \(\Xcal_{/s}\).
 The primary objective is to identify regions in which the impact of the \(s\)-th feature on the output is
@@
 the objective is to identify regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\) such that the interaction effect is minimized.
-Regionally Additive Models (RAM) formulation is:
+Regionally Additive Models (RAM) formulate the mapping \(\mathcal{X} \rightarrow \mathcal{Y}\) as:
 
@@
 \label{eq:ram_formulation}
-f^{\mathtt{RAM}}(\xb) = c + \sum_{s=1}^D \sum_{t=1}^{T_s} f_{st}(x_s) \when{\xcc \in \mathcal{R}_{st}}
+f^{\mathtt{RAM}}(\xb) = c + \sum_{s=1}^D \sum_{t=1}^{T_s} f_{st}(x_s) \when{\xcc \in \mathcal{R}_{st}}, \quad \xb \in \Xcal
 \end{equation}
@@
 \label{eq:ram_formulation2}
-f^{\mathtt{RAM}}(\xb) = c + \sum_{s,t} f_{st}(x_{st})
+f^{\mathtt{RAM}}(\xb) = c + \sum_{s,t} f_{st}(x_{st}) \quad \xb \in \Xcal^{\mathtt{RAM}}
 \end{equation}
 %
-In the running example, to minimize the effect of feature interactions, we have to split the effect of feature \(x_2\) into two regions,
+Equations~\ref{eq:ram_formulation} and~\ref{eq:ram_formulation2} are equivalent.
+To gain a better understanding of the latter formulation, consider the toy example described in Section~\ref{sec:motivation}.
+To minimize the impact of feature interactions, we need to divide feature \(x_2\) into two subregions,
 \(\mathcal{R}_{21} = \{x_1 > 0 \text{ and } x_3 = 1\}\) and \(\mathcal{R}_{22} = \{x_1 \leq 0 \text{ or } x_3 = 0\}\).
-The RAM formulation is \(f^{\mathtt{RAM}}(\xb) = f_1(x_1) + f_{21}(x_{21}) + f_{22}(x_{22}) + f_3(x_3)\).
+This division results in an augmented feature space \(\Xcal^{\mathtt{RAM}} = (x_1, x_{21}, x_{22}, x_3)\)
+and a RAM formulation of the form: \(f^{\mathtt{RAM}}(\xb) = f_1(x_1) + f_{21}(x_{21}) + f_{22}(x_{22}) + f_3(x_3)\).
 
@@
 
-\paragraph{DALE and Feature Interaction}
+\paragraph{DALE formulation}
 
-DALE plot gets as input a black-box function \(f(\cdot)\) and a dataset \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\),
-and decomposes the \texttt{effect} (impact) of a specific feature $s$ on the output $y$.
-DALE's formulation is:
-
+DALE gets as input the black-box function \(f(\cdot)\)
+and the dataset \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\),
+and returns the effect (impact) of a specific feature $s$ on the output $y$:
+%
 \begin{equation}  \label{eq:DALE_accumulated_mean_est}
@@
 %
-where \(k_x\) the index of the bin such that
+For more details on the DALE method, please refer to the original paper~\citep{gkolemis2023dale}.
+In the above equation, \(k_x\) is the index of the bin such that
 \(z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\)
@@
 \( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^{(i)}_s < z_{k} \} \).
-DALE computes the average effect (impact) of the feature \(x_s\) on the output of the black-box function \(f(\cdot)\),
-first by dividing the feature space into bins and computing the average effect of the feature \(x_s\) in each bin
-(bin-effect) and then by aggregating the bin-level effects.
-In cases where there are strong interactions between the features, the instance-level effects inside each bin
-start to deviate from the average bin-effect.
+In short, DALE computes the average effect (impact) of the feature \(x_s\) on the output,
+by, first, dividing the feature space into $K$ equally-sized bins, i.e., \(z_0, \ldots, z_K\)
+second, computing the average effect in each bin \(\hat{\mu}(z_{k-1}, z_k)\) (bin-effect) as the average of the instance-level effects inside the bin,
+and, finally, aggregating the bin-level effects.
+
+\paragraph{DALE for feature interactions}
+
+In cases where there are strong interactions between the features,
+the instance-level effects inside each bin deviate from the average bin-effect.
 We can measure such deviation using the standard deviation of the instance-level effects inside each bin (bin-deviation):
@@
 %
-and the global interaction between the feature \(x_s\) and the rest of the features with the
-aggregated bin-deviation
+and the interaction between the feature \(x_s\) and the rest of the features along the whole \(s\)-th dimension
+with the aggregated bin-deviation:
 \begin{equation}
@@
 Eq.~\eqref{eq:DALE_interaction} measures the interaction between the feature \(x_s\) and the rest of the features.
-It takes values in the range \([0, \infty)\) and it is zero when \(x_s\) does not interact with the rest of the features,
-i.e., when $f(\xb) = f_s(x_s) + f_{s}(x_s)$ and it is higher when the interaction is stronger.
+It takes values in the range \([0, \infty)\) with zero indicating that \(x_s\) does not interact with the rest of the features,
+i.e., the underlying black box function can be written as $f(\xb) = f_s(x_s) + f_{/s}(x_{/s})$.
+In all other cases, $\mathcal{H}_s$ is greater than zero and the higher the value, the stronger the interaction.
 
-\paragraph{DALE for Regional Effect}
+A final detail, is that in order to have a more robust estimation of the bin-effect and the bin-deviation,
+we use variable-size bins instead of the fixed-size ones in DALE.
+In particular, we start with a dense fixed-size grid of bins and we iteratively merge the neighboring bins with similar
+bin-effect and bin-deviation until all bins have at least $n_{\min}$ instances.
+In this way, we can have a more accurate approximation of the bin-effect and the bin-deviation.
 
-If instead of applying DALE globally, we restrict it on a subregion of the feature space
-\(\mathcal{R}_{st} \subset \mathcal{X}\), we can measure the interaction between the feature \(x_s\) and the rest of the features
-inside the subregion \(\mathcal{R}_{st}\).
-In compuatational terms, this means that instead of using the whole dataset \(\mathcal{D}\) to compute the bin-effect and the bin-deviation,
-we use only the instances that belong to the subregion \(\mathcal{R}_{st}\).
-In this way, we can define $\hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k), \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)$ and $\mathcal{H}_{\mathcal{R}_{st}}$ exaclty as in
-Eq.~\eqref{eq:DALE_accumulated_mean_est}, Eq.~\eqref{eq:var_bin_approx} and Eq.~\eqref{eq:DALE_interaction} respectively,
-but using only the instances that belong to the subregion \(\mathcal{R}_{st}\), i.e. $\xb^i: x_s^i \in \mathcal{S}_k \land x_{s}^i \in \mathcal{R}_{st}$.
-Analytical formulas in the Appendix~\ref{sec:regional_dale}.
+\paragraph{Subregions as an optimization problem}
 
-Therefore, in order to minimize the interaction we search for a set of regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\),
+In the same way that we can estimate the feature effect and the feature interactions for the $s$-th feature in the whole input space,
+using Eq.~\eqref{eq:DALE_accumulated_mean_est} and Eq.~\eqref{eq:DALE_interaction},
+we can also estimate the effect and the interactions in a subregion of the input space \(\mathcal{R}_{st} \subset \mathcal{X}\).
+We denote the equivalent regional qunatities as $f^{\mathtt{DALE}}_{\mathcal{R}_{st}}(x_s)$ and $\mathcal{H}_{\mathcal{R}_{st}}$.
+$f^{\mathtt{DALE}}_{\mathcal{R}_{st}}(x_s)$ and $\mathcal{H}_{\mathcal{R}_{st}}$ are defined exactly as in
+Eq.~\eqref{eq:DALE_accumulated_mean_est} and Eq.~\eqref{eq:DALE_interaction} respectively,
+with the only difference that instead of using the whole dataset \(\mathcal{D}\) to compute the regional bin-effect $\hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k)$
+and the regional bin-deviation $\hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)$,
+we use only the instances that belong to the subregion \(\mathcal{R}_{st}\), i.e. $\xb^i: x_s^i \in \mathcal{S}_k \land x_{s}^i \in \mathcal{R}_{st}$.
+Therefore, in order to minimize the interactions of a particular feature $s$ we search for a set of regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\),
 that minimize:
@@
     & \underset{\{\mathcal{R}_{st}\}_{t=1}^{T_s}}{\text{minimize}}
-    & & \mathcal{L} = \sum_{s=1}^S \sum_{t=1}^{T} \mathcal{H}_{\mathcal{R}_{st}} \\
+    & & \mathcal{L} = \sum_{t=1}^{T_s} \mathcal{H}_{\mathcal{R}_{st}} \\
     & \text{subject to}
@@
 \end{equation}
+%
 
-For minimizing Eq.~\eqref{eq:SE}, we set up a CART-like algorithm, as proposed by \citep{herbinger2023decomposing}.
-The backbone of the algorithm is described in Algorithm~\ref{alg:subregion_detection};
-for each feature $s \in \{1, \ldots, D\}$, we exhaustively search level-by-level, for the optimal split.
-For example, suppose that the optimal first-level split for some feature $s$ is
-$x_c \pm p_1$ where $x_c$ is a numerical feature.
-The first-level split adds to the RAM to terms for feature $x_s$, $f_{s1}(x_s) \when{x_c < p_1}$ and $f_{s2}(x_s) \when{x_c \geq p_1}$,
-and splits the dataset into two subsets, $X_1 = \{ \xb^i : x_s^i \leq x_c + p_1 \}$ and $X_2 = \{ \xb^i : x_s^i > x_c + p_1 \}$.
-Afterwards, suppose that the optimal second-level split for feature $s$ is $x_c' = (\neq) p_2$ for some categorical feature $x_c'$.
-The second-level split adds to the RAM four terms for feature $x_s$, $f_{s1}(x_s) \when{x_c < p_1 \land x_c' = p_2}$,
-$f_{s2}(x_s) \when{x_c < p_1 \land x_c' \neq p_2}$, $f_{s3}(x_s) \when{x_c \geq p_1 \land x_c' = p_2}$ and $f_{s4}(x_s) \when{x_c \geq p_1 \land x_c' \neq p_2}$.
+\paragraph{Proposed solution}
+
+For minimizing Eq.~\eqref{eq:SE}, we develop a tree-based algorithm, similar to the one proposed by~\citep{herbinger2023decomposing}.
+The backbone of the algorithm is described in Algorithm~\ref{alg:subregion_detection}.
+For each feature $s$, we search for the $T$ optimal splits.
+To better understand the algorithm, let's take the toy example of Section~\ref{sec:motivation}.
+The subregions are searched independently for each feature $s \in \{1, \ldots, D\}$.
+For feture $s=2$, we start by searching for the optimal split for the first level of the tree.
+The candidate features for defining the first level split are $x_1$ and $x_3$.
+The algorithm~\ref{alg:single_feature_subregion} is responsible to find the optimal first level split.
+First, it defines the candidate split points for each feature.
+Since $x_1$ is a continuous feature, the candidate split points are a linearly spaced grid of $p$ points in the range $[-1, 1]$,
+where $P$ is a hyperparameter of the algorithm, set to 10 in the experiments.
+Therefore, the candidate split points are $p = \{-1, -0.8, -0.6, \ldots, 0.8, 1\}$ and
+the corrsponding subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_1 \leq p \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_1 > p \}$.
+Since $x_3$ is categorical, the candidate split points are the unique values of $x_3$, i.e. $\{0, 1\}$
+and the corresponding subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 = 1 \}$.
+For each candidate split point, the algorithm computes the corresponding subregions and the corresponding $\mathcal{H}_{\mathcal{R}_{st}}$.
+and selects the split point that minimizes the weighted level of interactions, as defined by Algorithm~\ref{alg:get_interaction}.
+Suppose that the optimal first-level subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 \neq 0 \}$.
+The algorithm then proceeds to the second level of the tree.
+For the second level, the first split is fixed, i.e. the are two datasets split according to the first-level split and the algorithm searches for the optimal second-level split.
+The candidate features for defining the second level split are $x_3$.
 
 The output of algorithm~\ref{alg:subregion_detection} is a set of $T$ splits per feature $s$.
-Since splitting does guarantee an important decrease in the loss function, we refine the splits by
-pruning all splits that do not decrease the loss function by more than a threshold $\epsilon$.
+It is imporant to note, that since not all splits lead to an important decrease in the level of interactions,
+we post-process the splits by pruning all splits that do not decrease the loss function by more than a threshold $\epsilon$.
+The final set of splits is denoted as $\{\mathcal{R}_{st}\}_{t=1}^{T_s}$.
 
@@
 
+\begin{algorithm}
+\caption{GetInteraction}
+\label{alg:get_interaction}
+\SetAlgoLined
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+\Input{$X\_list$, $J\_list$, $min\_points$}
+\Output{weighted average of interaction levels}
+N $\gets$ total number of items in $X\_list$\;
+W $\gets$ []\;
+L $\gets$ []\;
+\For{$i$ in len($X\_list$)}{
+    X $\gets$ X\_list[i]\;
+    J $\gets$ J\_list[i]\;
+    \uIf{$|X|$ < min\_points}{
+      Append $\infty$ to L\;
+    }
+    \Else{
+        Append $\mathcal{L}(X, J)$ to $L$\;
+    }
+    Append $\frac{|X|}{N}$ to $W$\;
+}
+\Return $\text{sum}(L \cdot W)$\;
+\end{algorithm}
+
 \paragraph{Computational Complexity}
@@
 \subsection{Fitting the GAMs}
+\label{subsec:fitting_gams}
 
@@
 
-\bibliography{report}
-
 \section{Appendix}
@@
 
-\begin{algorithm}
-\caption{GetInteraction}
-\label{alg:get_interaction}
-\SetAlgoLined
-\SetKwInOut{Input}{Input}
-\SetKwInOut{Output}{Output}
-\Input{$X\_list$, $J\_list$, $min\_points$}
-\Output{weighted average of interaction levels}
-N $\gets$ total number of items in $X\_list$\;
-W $\gets$ []\;
-L $\gets$ []\;
-\For{$i$ in len($X\_list$)}{
-    X $\gets$ X\_list[i]\;
-    J $\gets$ J\_list[i]\;
-    \uIf{$|X|$ < min\_points}{
-      Append $\infty$ to L\;
-    }
-    \Else{
-        Append $\mathcal{L}(X, J)$ to $L$\;
-    }
-    Append $\frac{|X|}{N}$ to $W$\;
-}
-\Return $\text{sum}(L \cdot W)$\;
-\end{algorithm}
+\bibliography{report}
 

### 2023-06-21 report/report.tex
update

@@
 \begin{document}
+    \def\MakeUppercaseUnsupportedInPdfStrings{\scshape}
 \maketitle
@@
     \end{subfigure}
-    \caption{Caption for the entire figure}
+    \caption{The left image showcases the global GAM which erroneously learns an approximation of \(f(\xb) \approx 2x_2\).
+    In contrast, the middle and right images demonstrate the RAM's ability to identify two distinct subregions where \(f\) exhibits regional additivity.
+    By fitting a GAM to each subregion, the RAM accurately captures the true function $f$ while retaining interpretability.}
     \label{fig:ram_example}
@@
 making them a suitable choice for this step.
+% Driver, teacher, guide
 
@@
 
-For minimizing Eq.~\eqref{eq:SE}, we develop a tree-based algorithm, similar to the one proposed by~\citep{herbinger2023decomposing}.
-The backbone of the algorithm is described in Algorithm~\ref{alg:subregion_detection}.
-For each feature $s$, we search for the $T$ optimal splits.
-To better understand the algorithm, let's take the toy example of Section~\ref{sec:motivation}.
-The subregions are searched independently for each feature $s \in \{1, \ldots, D\}$.
-For feture $s=2$, we start by searching for the optimal split for the first level of the tree.
-The candidate features for defining the first level split are $x_1$ and $x_3$.
-The algorithm~\ref{alg:single_feature_subregion} is responsible to find the optimal first level split.
-First, it defines the candidate split points for each feature.
-Since $x_1$ is a continuous feature, the candidate split points are a linearly spaced grid of $p$ points in the range $[-1, 1]$,
+To minimize the objective of Eq.~\eqref{eq:SE}, we have developed a tree-based algorithm based on the approach proposed
+by~\citep{herbinger2023decomposing}.
+The core of the algorithm is outlined in Algorithm~\ref{alg:subregion_detection}.
+We search for subregions independently for each feature $s \in \{1, \ldots, D\}$.
+and we select the $T$ optimal splits, in a greedy manner.
+To better comprehend the algorithm, let's consider the illustrative example of Section~\ref{sec:motivation}.
+For feature $s=2$, the algorithm starts with the first level of the tree, with $/s = \{1, 3\}$ as candidate split-features.
+Algorithm~\ref{alg:single_feature_subregion} identifies the optimal first-level split, with the following procedure.
+Initially, it determines the candidate split points for each feature.
+
+Since $x_1$ is a continuous feature, are generated as a linearly spaced grid of $P$ points within the range of the feature, i.e. $[-1, 1]$,
 where $P$ is a hyperparameter of the algorithm, set to 10 in the experiments.
-Therefore, the candidate split points are $p = \{-1, -0.8, -0.6, \ldots, 0.8, 1\}$ and
-the corrsponding subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_1 \leq p \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_1 > p \}$.
-Since $x_3$ is categorical, the candidate split points are the unique values of $x_3$, i.e. $\{0, 1\}$
-and the corresponding subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 = 1 \}$.
-For each candidate split point, the algorithm computes the corresponding subregions and the corresponding $\mathcal{H}_{\mathcal{R}_{st}}$.
-and selects the split point that minimizes the weighted level of interactions, as defined by Algorithm~\ref{alg:get_interaction}.
-Suppose that the optimal first-level subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 \neq 0 \}$.
-The algorithm then proceeds to the second level of the tree.
-For the second level, the first split is fixed, i.e. the are two datasets split according to the first-level split and the algorithm searches for the optimal second-level split.
-The candidate features for defining the second level split are $x_3$.
+Consequently, the candidate split points $p \in \{-1, -0.8, -0.6, \ldots, 0.8, 1\}$ define the candidate subregions
+$\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_1 \leq p \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_1 > p \}$.
+As for $x_3$, being a categorical feature, the candidate split points are its unique values, i.e., $\{0, 1\}$,
+and the corresponding subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 \neq 0 \}$.
 
-The output of algorithm~\ref{alg:subregion_detection} is a set of $T$ splits per feature $s$.
-It is imporant to note, that since not all splits lead to an important decrease in the level of interactions,
-we post-process the splits by pruning all splits that do not decrease the loss function by more than a threshold $\epsilon$.
-The final set of splits is denoted as $\{\mathcal{R}_{st}\}_{t=1}^{T_s}$.
+For each candidate position, the algorithm splits the initial dataset into two subsets,
+$[\mathcal{D}_1, \mathcal{D}_2]$ one for each subregion and computes the level of interaction
+as the weighted sum of the interactions of the two subregions, as defined by Algorithm~\ref{alg:get_interaction}.
+After the iteration over all points and all candidate positions,
+the split point that minimizes the weighted level of interactions is selected.
+In the illustrative example, the optimal first-level subregions are
+$\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 \neq 0 \}$.
+The algorithm then proceeds to the second level of the tree, where the only candidate feature is $x_3$.
+In this step, the first split is considered fixed, i.e., the algorithm uses a list of two datasets,
+and searches in an analogous manner for the optimal second-level split that will be applied on top of the first split,
+leading to four subregions.
+The algorithm proceeds in a similar manner, until it reaches the maximum depth $T$.
+
+The output of algorithm~\ref{alg:subregion_detection} is a collection of $T$ splits for each feature $s$.
+It is important to note that not all splits lead to a significant decrease in the level of interactions.
+To address this, we post-process the splits by pruning any splits that do not result in a decrease of the
+weighted interaction greater than a threshold value $\epsilon$ (set to $20\%$ drop in the experiments).
+The pruned set of splits is denoted as $\{\mathcal{R}_{st}\}_{t=1}^{T_s}$.
 
@@
 
-Algorithm~\ref{alg:subregion_detection} has a computational complexity of $\mathcal{O}(D \cdot T)$.
-and Algorithm~\ref{alg:single_feature_subregion} has a computational complexity of $\mathcal{O}(D-1 \cdot P \cdot N)$
-where $D$ is the number of feature, $N$ is the number of samples, $T$ is the maximum depth of the tree and $P$ is the number of query postitions.
-The computational complexity of the proposed method is therefore $\mathcal{O}(D \cdot (D-1) \cdot T \cdot P \cdot N)$.
-However, in practice, the number of query positions $P$ and the maximum depth $T$ are small numbers.
-Therefore, the computational complexity of the proposed method is $\mathcal{O}(D^2 \cdot N)$.
+Algorithm~\ref{alg:subregion_detection} has a computational complexity of $\mathcal{O}(D \cdot T)$
+as it iterates over all features and levels of the tree.
+Algorithm~\ref{alg:single_feature_subregion} has a computational complexity of $\mathcal{O}(D \cdot P \cdot N)$
+as it iterates over all features, query positions, and performs indexing operations on the data (SplitDataset and GetInteraction).
+The use of DALE eliminates the need to compute the Jacobian matrix for each split,
+resulting in computational complexity independent of the model complexity.
+Considering the entire algorithm, the computational complexity is $\mathcal{O}(D \cdot (D-1) \cdot T \cdot P \cdot N)$.
+However, in practice, $P$ and $T$ are small numbers.
+Therefore, the computational complexity of the proposed method simplifies to $\mathcal{O}(D^2 \cdot N)$,
+making it suitable for large datasets, heavy models, and reasonably high-dimensional data.
+In practice, for a typical scenario with $N=10 \times 10^3$ samples and $D=20$ features, the proposed method takes less than $10$ seconds to run.
 
-\subsection{Fitting the GAMs}
+\subsection{Third step: Fit a GAM in each subregion}
 \label{subsec:fitting_gams}
 
-\subsubsection{Objective}
-
-\subsubsection{Proposed Approach}
-
-\subsection{Discussion}
-
-Recently, a number of methods have been proposed to extend traditional GAMs and make them more expressive.
-The majority of the ideas follow one of the following research directions;
-The first one targets on representing the main components of a GAM $\{ f_i(x_i) \}$ with novel models.
-For example, \citet{agarwal2021neural} who used an end-to-end neural network for learning the main components.
-The second one focuses on extending the GAMs to model interactions between the features.
-For example, \citet{lou2013accurate} proposed Explainable Boosting Machines (EBMs) which are a generalized additive model with pairwise interaction terms.
-It is worth noting that the proposed method is orthogonal to the aforementioned research directions.
-As we will show in the experiments, the proposed method can be used in conjunction with any of the
-aforementioned methods to improve the accuracy of the resulting model, while maintaining the interpretability of the model.
+Once the subregions are detected, any Generalized Additive Model (GAM) family can be fitted to the augmented input space $\mathcal{X}^{\mathtt{RAM}}$.
+Recently, several methods have been proposed to extend GAMs and enhance their expressiveness.
+These methods can be categorized into two main research directions.
+The first direction targets on representing the main components of a GAM $\{ f_i(x_i) \}$ using novel models.
+For example,~\citep{agarwal2021neural} introduced an approach that employs an end-to-end neural network to learn the main components.
+The second direction aims to extend GAMs to model feature interactions.
+Examples of such extensions include Explainable Boosting Machines (EBMs)~\citep{lou2013accurate} or
+Node-GAMs~\citep{chang2021node}.
+These models are generalized additive models that incorporate pairwise interaction terms.
+It is worth noticing, that the RAM framework and can be used on top of both these research directions
+to further enhance the expressiveness of the models while maintaining their interpretability.
+In our experiments, we use the Explainable Boosting Machines (EBMs).
 
@@
 
+We evaluate the proposed approach on two typical tabular datasets, namely the Bike-Sharing Dataset~\citep{misc_bike_sharing_dataset_275}
+and the California Housing Dataset~\citep{pace1997sparse}.
+
 \begin{table}[htbp]
   \centering
-  \caption{Result Comparison}
+  \caption{The table compares RMSE values of different models on two datasets: Bike-Sharing and California Housing.
+  The models include DNN, GAM, RAM, GAM², and RAM² (representing 2nd order interactions).
+  Lower values indicate better performance.
+  RAM consistently outperforms GAM and approaches DNN performance.}
   \label{tab:sample}
@@
 \end{table}
-
-We evaluate the proposed approach on two typical tabular datasets, namely the Bike-Sharing Dataset~\citep{misc_bike_sharing_dataset_275}
-and the California Housing Dataset~\citep{pace1997sparse}.
 
@@
 Each housing block is described by eight numerical features, namely,
-$X_{\mathtt{latitude}}$, $X_{\mathtt{longitude}}$, $X_{\mathtt{median\_age}}$, $X_{\mathtt{total\_rooms}}$, $X_{\mathtt{total\_bedrooms}}$, $X_{\mathtt{population}}$, $X_{\mathtt{households}}$, and $X_{\mathtt{median\_income}}$.
+$X_{\mathtt{lat}}$, $X_{\mathtt{long}}$, $X_{\mathtt{median\_age}}$, $X_{\mathtt{total\_rooms}}$, $X_{\mathtt{total\_bedrooms}}$, $X_{\mathtt{population}}$, $X_{\mathtt{households}}$, and $X_{\mathtt{median\_income}}$.
 The target variable, $Y_{\mathtt{value}}$, is the median house value in dollars for each block.
@@
 resulting in an expanded input space consisting of \(16\) features, as we show in table~\ref{tab:california_housing_subregions}.
-Out of them, we randomly select and illustrate in Figure~\ref{fig:california_housing} the effect of the feature $X_{\mathtt{longitude}}$.
-As we observe, for the house blocks located in the southern part of California ($X_{\mathtt{latitude}} \leq 34.9$),
-the house value decreases in an almost linear fashion as we move eastward ($X_{\mathtt{longitude}}$ increases).
-In contrast, for the house blocks located in the northern part of California ($X_{\mathtt{latitude}} > 34.9$),
-the house value performs a rapid (non-linear) decrease as we move eastward ($X_{\mathtt{longitude}}$ increases).
+Out of them, we randomly select and illustrate in Figure~\ref{fig:california_housing} the effect of the feature $X_{\mathtt{long}}$.
+As we observe, for the house blocks located in the southern part of California ($X_{\mathtt{lat}} \leq 34.9$),
+the house value decreases in an almost linear fashion as we move eastward ($X_{\mathtt{long}}$ increases).
+In contrast, for the house blocks located in the northern part of California ($X_{\mathtt{lat}} > 34.9$),
+the house value performs a rapid (non-linear) decrease as we move eastward ($X_{\mathtt{long}}$ increases).
 We also observe that although the EBM fitted to each subregion captures the general trend,
@@
   \label{tab:california_housing_subregions}
-  \begin{tabular}{c|c|}
+  \begin{tabular}{c|c}
       Feature & Subregions \\
       \hline
-      \multirow{2}{*}{$X_{\mathtt{longitude}}$} & $X_{\mathtt{longitude}} \when{X_{\mathtt{latitude}} \leq 34.9}$ \\
-      & $X_{\mathtt{longitude}} \when{X_{\mathtt{latitude}} > 34.9}$ \\
+      \multirow{2}{*}{$X_{\mathtt{long}}$} & $X_{\mathtt{long}} \when{X_{\mathtt{lat}} \leq 34.9}$ \\
+      & $X_{\mathtt{long}} \when{X_{\mathtt{lat}} > 34.9}$ \\
       \hline
-      \multirow{2}{*}{$X_{\mathtt{latitude}}$} & $X_{\mathtt{latitude}} \when{X_{\mathtt{longitude}} \leq -120.31}$ \\
-      & $X_{\mathtt{latitude}} \when{X_{\mathtt{longitude}} > -120.31}$ \\
+      \multirow{2}{*}{$X_{\mathtt{lat}}$} & $X_{\mathtt{lat}} \when{X_{\mathtt{long}} \leq -120.31}$ \\
+      & $X_{\mathtt{lat}} \when{X_{\mathtt{long}} > -120.31}$ \\
       \hline
@@
         \includegraphics[width=\textwidth]{figures/california_gam}
-        \caption{\(f(X_{\mathtt{longitude}})\)}
+        \caption{\(f(X_{\mathtt{long}})\)}
         \label{subfig:california_gam}
@@
         \includegraphics[width=\textwidth]{figures/california_ram_1}
-        \caption{\(f(X_{\mathtt{longitude}}) \when{X_{\mathtt{latitude}} \leq 34.89}\)}
+        \caption{\(f(X_{\mathtt{long}}) \when{X_{\mathtt{lat}} \leq 34.89}\)}
         \label{subfig:california_ram_1}
@@
         \includegraphics[width=\textwidth]{figures/california_ram_2}
-        \caption{\(f(X_{\mathtt{longitude}}) \when{X_{\mathtt{latitude}} > 34.89}\)}
+        \caption{\(f(X_{\mathtt{long}}) \when{X_{\mathtt{lat}} > 34.89}\)}
         \label{subfig:california_ram_2}
     \end{subfigure}
-    \caption{Caption for the entire figure}
+    \caption{Comparison of different predictions for housing prices in California based on the longitude.
+    Subfigure (a) showcases the generalized additive model (GAM),
+        while subfigures (b) and (c) demonstrate the RAM components for different latitude ranges:
+        \(f(X_{\mathtt{long}}) \when{X_{\mathtt{lat}} \leq 34.89}\) and
+        \(f(X_{\mathtt{long}}) \when{X_{\mathtt{lat}} > 34.89}\), respectively.
+        We observe, that although the EBM model is able to capture the overall trend in the data,
+        it also exhibits a large amount of variance.}
     \label{fig:california_housing}
@@
 
-\section{Appendix}
+%\section{Appendix}
+%
+%\subsection{Regional DALE formulation}
+%\label{sec:regional_dale}
+%
+%Below we provide the regional DALE formulation for the bin-effect, the bin-deviation and the aggregated deviation on a subregion \(\mathcal{R}_{st}\):
+%
+%\begin{equation}
+%    \label{eq:bin_effect_subregion}
+%    \hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}|} \sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \dfdx(\mathbf{x}^i)
+%\end{equation}
+%
+%\begin{equation}
+%  \label{eq:bin_deviation_subregion}
+%  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
+%\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
+%    \hat{\mu}(z_{k-1}, z_k) \right )^2
+%\end{equation}
+%
+%\begin{equation}
+%  \label{eq:SE}
+%  \mathcal{H}_{\mathcal{R}_{sk}} = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
+%\end{equation}
 
-\subsection{Regional DALE formulation}
-\label{sec:regional_dale}
-
-For this reason we define the bin-effect, the bin-deviation and the aggregated deviation on a subregion \(\mathcal{R}_{st}\) as:
-
-\begin{equation}
-    \label{eq:bin_effect_subregion}
-    \hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}|} \sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \dfdx(\mathbf{x}^i)
-\end{equation}
-
-\begin{equation}
-  \label{eq:bin_deviation_subregion}
-  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
-\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
-    \hat{\mu}(z_{k-1}, z_k) \right )^2
-\end{equation}
-
-\begin{equation}
-  \label{eq:SE}
-  \mathcal{L}_{\mathcal{R}_{sk}} = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
-\end{equation}
-
-\subsection{Algorithmic Details of Subregion Detection}
-
-\begin{algorithm}
-\caption{SplitDataset}
-\label{alg:split_dataset}
-\SetAlgoLined
-\SetKwInOut{Input}{Input}
-\SetKwInOut{Output}{Output}
-\Input{X\_list, J\_list, c, val, is\_cat}
-\Output{X\_split, J\_split}
-X\_split $\gets$ []\;
-J\_split $\gets$ []\;
-\For{i = 1 to len(X\_list)}{
-    X= X\_list[i]\;
-    J= J\_list[i]\;
-    \uIf{is\_cat}{
-        ind\_1 $\gets$ X[:, c] $=$ val\;
-        ind\_2 $\gets$ X[:, c] $\neq$ val\;
-    }
-    \Else{
-        ind\_1 $\gets$ X[:, c] $\leq$ val\;
-        ind\_2 $\gets$ X[:, c] $>$ val\;
-    }
-    Append X[ind\_1], X[ind\_2] to X\_split\;
-    Append J[ind\_1], J[ind\_2] to J\_split\;
-}
-\Return X\_split, J\_split
-\end{algorithm}
+%\subsection{Algorithmic Details of Subregion Detection}
+%
+%\begin{algorithm}
+%\caption{SplitDataset}
+%\label{alg:split_dataset}
+%\SetAlgoLined
+%\SetKwInOut{Input}{Input}
+%\SetKwInOut{Output}{Output}
+%\Input{X\_list, J\_list, c, val, is\_cat}
+%\Output{X\_split, J\_split}
+%X\_split $\gets$ []\;
+%J\_split $\gets$ []\;
+%\For{i = 1 to len(X\_list)}{
+%    X= X\_list[i]\;
+%    J= J\_list[i]\;
+%    \uIf{is\_cat}{
+%        ind\_1 $\gets$ X[:, c] $=$ val\;
+%        ind\_2 $\gets$ X[:, c] $\neq$ val\;
+%    }
+%    \Else{
+%        ind\_1 $\gets$ X[:, c] $\leq$ val\;
+%        ind\_2 $\gets$ X[:, c] $>$ val\;
+%    }
+%    Append X[ind\_1], X[ind\_2] to X\_split\;
+%    Append J[ind\_1], J[ind\_2] to J\_split\;
+%}
+%\Return X\_split, J\_split
+%\end{algorithm}
 

### 2023-06-21 report/real_bike_sharing_tfdf.py

@@ -1,0 +1,228 @@
+import sys, os
+import timeit
+sys.path.append(os.path.dirname(os.getcwd()))
+import pythia
+import numpy as np
+import pythia.interaction as interaction
+import pythia.regions as regions
+import pandas as pd
+import tensorflow as tf
+from tensorflow import keras
+import matplotlib.pyplot as plt
+import tensorflow_decision_forests as tfdf
+import tensorflow_decision_trees as tfdt
+from interpret.glassbox import ExplainableBoostingRegressor
+import sklearn.metrics as metrics
+import importlib
+from sklearn.utils import check_random_state
+import random
+pythia = importlib.reload(pythia)
+
+# make everyting reproducible
+np.random.seed(42)
+tf.random.set_seed(42)
+check_random_state(42)
+random.seed(42)
+
+def preprocess(df):
+    # shuffle
+    df.sample(frac=1).reset_index(drop=True)
+
+    # Standarize X
+    X_df = df.drop(["cnt"], axis=1)
+    x_mean = X_df.mean()
+    x_std = X_df.std()
+    X_df = (X_df - X_df.mean()) / X_df.std()
+
+    # Standarize Y
+    Y_df = df["cnt"]
+    y_mean = Y_df.mean()
+    y_std = Y_df.std()
+    Y_df = (Y_df - Y_df.mean()) / Y_df.std()
+    return X_df, Y_df, x_mean, x_std, y_mean, y_std
+
+
+def split(X_df, Y_df):
+    # data split
+    X_train = X_df[:int(0.8 * len(X_df))]
+    Y_train = Y_df[:int(0.8 * len(Y_df))]
+    X_test = X_df[int(0.8 * len(X_df)):]
+    Y_test = Y_df[int(0.8 * len(Y_df)):]
+    return X_train, Y_train, X_test, Y_test
+
+
+def plot_subregions(splits, feat, feat_ram, X_train, model, model_jac, gam, ram):
+    foc = splits["feat_{0}".format(feat)][0]["feature"]
+    position = splits["feat_{0}".format(feat)][0]["position"]
+    type = splits["feat_{0}".format(feat)][0]["type"]
+
+    # Regional Plot
+    if type == "cat":
+        rhale = pythia.RHALE(data=X_train.to_numpy(), model=model, model_jac=model_jac)
[…]

### 2023-06-21 report/report.tex
Algorithm 1 change

--- 2023-06-20 report/report.tex
@@
 \begin{algorithm}
+\caption{Regionally Additive Model (RAM) training}
+\label{alg:ram}
+\SetAlgoLined
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+\BlankLine
+\Input{A dataset $(X, y)$}
+\Output{A trained RAM model $f^{\mathtt{RAM}}$}
+  \BlankLine
+  Train a differentiable black box model $f$ using $(X, y)$\;
+  \tcc{Detect subregions for all features}
+  $\left\{\mathcal{R}_{st}|s\in\left\{1,\ldots,D\right\}, t\in\left\{1,\ldots,T_s\right\}\right\}$ = DetectSubregions($X$, $y$, $f$)\;
+  Create the extended feature space $\mathcal{X}^{\mathtt{RAM}}$ using all
+  $\mathcal{R}_{st}$, as in Eq. \eqref{eq:ram_feature_space} \;
+  Fit a GAM in $\mathcal{X}^{\mathtt{RAM}}$ \tcp*{i.e., train each $f_{st}$ using
+    only data in $\mathcal{R}_{st}$}
+  \Return{$f^{\mathtt{RAM}}(\xb) = c + \sum_{s,t} f_{st}(x_{st}), \quad \xb \in \Xcal^{\mathtt{RAM}}$}
+\end{algorithm}
+
+\begin{algorithm}
 \caption{Detection of Subregions using DALE}
@@
 \BlankLine
-\Input{Data matrix $X$, Jacobian matrix $J$, Maximum Depth $T$}
+\Input{Data matrix $X$, Black box model $f$, Maximum Depth $T$}
 \BlankLine
-\Output{Optimal splits}
+\Output{Subregions $\mathcal{R}_{sK}$ for all features $s$ and depth $K=2^T$}
 \BlankLine

### 2023-06-21 report/report.tex
Algorithms 1 and 2

@@
 \BlankLine
-\Input{A dataset $(X, y)$}
+\Input{A dataset $(X, y)$ and a maximum level $T$}
 \Output{A trained RAM model $f^{\mathtt{RAM}}$}
@@
   \tcc{Detect subregions for all features}
-  $\left\{\mathcal{R}_{st}|s\in\left\{1,\ldots,D\right\}, t\in\left\{1,\ldots,T_s\right\}\right\}$ = DetectSubregions($X$, $y$, $f$)\;
+  $\left\{\mathcal{R}_{st}|s\in\left\{1,\ldots,D\right\},
+  t\in\left\{1,\ldots,T_s\right\}\right\}$ = DetectSubregions($X$, $y$, $f$, $T$)\;
   Create the extended feature space $\mathcal{X}^{\mathtt{RAM}}$ using all
@@
 \BlankLine
-\Input{Data matrix $X$, Black box model $f$, Maximum Depth $T$}
+\Input{Dataset $(X, y)$, Black box model $f$, Maximum depth $T$}
 \BlankLine
-\Output{Subregions $\mathcal{R}_{sK}$ for all features $s$ and depth $K=2^T$}
+\Output{Subregions $\mathcal{R}_{sT_s}$ for all features $s$, where $T_s \leq 2^T$}
 \BlankLine
-Initialize empty arrays for \texttt{loss}, \texttt{position}, and \texttt{feature\_c}
-\BlankLine
+\tcc{Needed for DALE feature effect estimation}
+Compute the Jacobian w.r.t. features $\xb$,  $J = \nabla_{\xb}f(\xb)$ \;
 \For{$s = 1$ to $D$}{
-X\_list, J\_list = [X], [J] \tcp*{Init lists}
   \For{$t = 1$ to $T$}{
-    \tcc{Find best split}
-    L, p, c, X\_list, J\_list $\gets$ BestSplit(X\_list, J\_list, s, is\_cat)\;
-    \tcc{Store best split}
-    loss[s, t], position[s, t], feature\_c[s, t] $\gets$ L, p, c
+    \tcc{Find best split feature $c$ at point $p$, leading to loss
+      $\mathcal{H}_{st}$ using regions of previous level}
+    $\mathcal{H}_{st}, c, p$ $\gets$
+    BestSplit($\left\{\mathcal{R}_{sk}|k\in\left\{1,\ldots,T_{t-1}\right\}\right\}$)\;
+    \If{$\lVert\mathcal{H}_{st} - \mathcal{H}_{s(t-1)}\rVert < \epsilon$}{
+      break\;
     }
   }
-\Return{loss, position, feature\_c}
+}
+\Return{$\left\{\mathcal{R}_{st}|s\in\left\{1,\ldots,D\right\},
+  t\in\left\{1,\ldots,T_s\right\}\right\}$}
 \end{algorithm}

### 2023-06-22 report/report.tex
intro updated

@@
 % Paragraph for Motivating about Regionally Additive Models
-Generalized Additive Models (GAMs)~\citep{hastie1987generalized} are a popular class of explainable by design (x-by-design) models.
-Their popularity stems from their seamless interpretability; since they are a linear (additive) combination of univariate functions,
-\(f(\xb) = c + \sum_{s=1}^D f_s(x_s)\), each individual univariate function (component) can be readily visualized and comprehended in isolation.
-However, GAM's main limitation is that they cannot express interactions between features.
-To mitigation this limitation, some approaches~\citep{lou2013accurate} extend them enabling pairwise interactions,
-i.e., \(f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{c \neq s} f_{sc}(x_s, x_c)\).
-Pairwise interactions can also be visualized and understood in isolation, so these models also maintain their x-by-design nature.
-Unfortunately, this does not hold for any interaction that involves more than two features, thus, the expressiveness of GAMs is limited to capturing up to two-feature interactions.
+Generalized Additive Models (GAMs)~\citep{hastie1987generalized} are a popular class of explainable by design (x-by-design) models~\citep{rudin2019stop, ghassemi2021false}.
+Their popularity stems from their inherent interpretability.
+GAMs represent an aggregation of univariate functions, where the overall model can be expressed as \(f(\xb) = c + \sum_{s=1}^D f_s(x_s)\).
+Due to this structure, each individual univariate function (component) can be visualized and interpreted independently.
+Consequently, understanding the behavior of the overall model simply requires visualizing all components, each with a one-dimensional plot.
 
+However, GAMs encounter limitations when predicting outcomes that depend on multiple features simultaneously.
+In such cases, the unknown predictive function includes terms representing interactions between features.
+To mitigate this limitation, GA$^2$Ms~\citep{lou2013accurate} extended GAMs adding pairwise interactions in the
+formulation, i.e., \(f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{s_1 \neq s_2} f_{s_1 s_2}(x_{s_1}, x_{s_2})\).
+Interpreting GA$^2$Ms involves visualizing both the first-order ($1D$ plots) and second-order (($2D$ plots)) components.
+However, as the number of features increases, the number of pairwise interactions grows exponentially,
+making it impractical for users to remember and interpret a large number of two-dimensional plots.
+Therefore, these methods target on automatically selecting the most significant interaction terms.
+
+Both GAMs and GA$^2$Ms have limitations in modeling interactions beyond the second degree,
+and as far as we know, there is no existing work that focuses on x-by-design GAM-based models capable of
+representing higher-order interactions.
+The main challenge lies in the difficulty of visualizing more than three terms simultaneously,
+which would violate the x-by-design principle.
+To address this limitation, we propose a new class of x-by-design models called Regionally Additive Models (RAMs).
+
+RAMs tackle the challenge of visualizing general interaction terms involving more than two variables by focusing
+on a specific class of terms.
+These terms have a general structure of these terms is
+$f(x_{s_1} | \when{x_{c_1}}, \when{x_{c_2}}, \cdots)$ for first-degree interactions and
+$f(x_{s_1}, x_{s_2} | \when{x_{c_1}}, \when{x_{c_2}}, \cdots)$ for second-degree interactions.
+To better grasp the idea, consider a prediction task where the outcome depends, among others,
+on a combination of three features: $x_1 \in [20, 80]$ (age), $x_2 \in [0, 40]$ (years in work), and $x_3 \in \{True, False\}$ (married).
+Both GAM and GA$^2$M would fail to capture the effect of the three-feature interaction.
+However, the three-feature effect can be decomposed in two sets of second-degree conditional terms based on the marital status:
+$f_{1}(x_1, x_2 | x_3 = True)$
+and
+$f_{2}(x_1, x_2 | x_3 = False)$, which can be visualized and interpreted as using two-dimensional plots.
+In a different scenario, it may be better to it in four sets of first-degree conditional terms:
+$f_{1}(x_1 | x_2 < 10, x_3 = True)$,
+$f_{2}(x_1 | x_2 \geq 10, x_3 = True)$,
+$f_{3}(x_1 | x_2 < 10, x_3 = False)$, and
+$f_{4}(x_1 | x_2 \geq 10, x_3 = False)$, which can be visualized and interpreted as four one-dimensional plots.
 % Paragraph for Motivating about Regionally Additive Models
-To overcome this limitation, we propose Regionally Additive Models (RAMs), a novel class of x-by-design models,
-that fits multiple GAMs to subregions of the feature space where interactions are minimized.
-Our approach consists of a three-step pipeline.
-First, we fit a black-box model to capture all high-order interactions.
-Second, we identify the subregions where the black-box model is nearly locally additive.
-Finally, we train a GAM specifically for each identified subregion.
+
+The objective of RAMs is to automatically identify the most significant conditional terms and fit a GAM for each of them.
+The RAM framework involves three steps.
+First, a black-box model is fitted to capture all high-order interactions.
+Then, the subregions where the black-box model exhibits near-local additivity are identified using Regional Effect Plots (REPs).
+Finally, a GAM is trained specifically for each identified subregion.
 

### 2023-06-23 report/report.tex
finished the paper

--- 2023-06-21 report/report.tex
@@
 \usepackage{multicol}
+\usepackage{algpseudocode}
 \bibliographystyle{plainnat}
@@
 % Paragraph for Motivating about Regionally Additive Models
-Generalized Additive Models (GAMs)~\citep{hastie1987generalized} are a popular class of explainable by design (x-by-design) models.
-Their popularity stems from their seamless interpretability; since they are a linear (additive) combination of univariate functions,
-\(f(\xb) = c + \sum_{s=1}^D f_s(x_s)\), each individual univariate function (component) can be readily visualized and comprehended in isolation.
-However, GAM's main limitation is that they cannot express interactions between features.
-To mitigation this limitation, some approaches~\citep{lou2013accurate} extend them enabling pairwise interactions,
-i.e., \(f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{c \neq s} f_{sc}(x_s, x_c)\).
-Pairwise interactions can also be visualized and understood in isolation, so these models also maintain their x-by-design nature.
-Unfortunately, this does not hold for any interaction that involves more than two features, thus, the expressiveness of GAMs is limited to capturing up to two-feature interactions.
+Generalized Additive Models (GAMs)~\citep{hastie1987generalized} are a popular class of explainable by design
+(x-by-design) models~\citep{rudin2019stop, ghassemi2021false}.
+Their popularity stems from their inherent interpretability.
+GAMs represent an aggregation of univariate functions, where the overall model can be expressed as
+\(f(\xb) = c + \sum_{s=1}^D f_s(x_s)\).
+Due to this structure, each individual univariate function (component) can be visualized and interpreted independently.
+Consequently, understanding the behavior of the overall model simply requires visualizing all components, each with a one-dimensional plot.
 
-% Paragraph for Motivating about Regionally Additive Models
-To overcome this limitation, we propose Regionally Additive Models (RAMs), a novel class of x-by-design models,
-that fits multiple GAMs to subregions of the feature space where interactions are minimized.
-Our approach consists of a three-step pipeline.
-First, we fit a black-box model to capture all high-order interactions.
-Second, we identify the subregions where the black-box model is nearly locally additive.
-Finally, we train a GAM specifically for each identified subregion.
+However, GAMs have limited accuracy in cases where the outcome depends on multiple features simultaneously,
+i.e., when the unknown predictive function includes terms that combine multiple features.
+To mitigate this limitation, GA$^2$Ms~\citep{lou2013accurate} extend the traditional GAMs by adding pairwise interactions in their
+    formulation, i.e., \(f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{s_1 \neq s_2} f_{s_1 s_2}(x_{s_1}, x_{s_2})\).
+GA$^2$Ms are also x-by-design, because the user can visualize both the first-order ($1D$ plots) and
+second-order (($2D$ plots)) components.
+However, as the number of features increases, the number of second-order interactions grows exponentially,
+making it impractical for users to remember and interpret a large number of two-dimensional plots.
+Therefore, methods like GA$^2$Ms target on automatically selecting the most significant interaction terms.
 
-\section{Background and motivation}
+Both GAMs and GA$^2$Ms have limitations in modeling interactions of more than two features,
+and as far as we know, there is no existing work that focuses on x-by-design GAM-based models capable of
+representing higher-order interactions.
+The main reason for that is the difficulty of visualizing more than three terms simultaneously,
+which would violate the x-by-design principle.
+
+To address this limitation, we propose a new class of x-by-design models called Regionally Additive Models (RAMs).
+Since in the general case, it is infeasible to visualize terms with more than two variables
+RAMs focus on learning terms with the following conditional structure:
+$f(x_{s_1} | \when{x_{c_1}}, \when{x_{c_2}}, \cdots)$ for first-degree interactions and
+$f(x_{s_1}, x_{s_2} | \when{x_{c_1}}, \when{x_{c_2}}, \cdots)$ for second-degree interactions.
+The symbol $\when{x_{c_1}}$ denotes the condition that the feature $x_{c_1}$ takes a specific value or belongs to a specific range.
+
+To better grasp the idea, consider a prediction task where the outcome depends, among others,
+on a combination $f(x_1, x_2, x_3)$ of three features:
+$x_1 \in [20, 80]$ (age), $x_2 \in [0, 40]$ (years in work), and $x_3 \in \{True, False\}$ (married).
+Both GAM and GA$^2$M would fail to accurately learn this term of the underlying predictive function.
+However, the three-feature effect can be decomposed in two sets of second-degree conditional terms based
+on the marital status: $f_{1}(x_1, x_2 | x_3 = True)$ and $f_{2}(x_1, x_2 | x_3 = False)$.
+In this way, RAM can accurately represent $f$ through learning two second-degree conditional terms, one for each marital status.
+Furthermore, the two sets of terms can be visualized and interpreted as using two-dimensional plots.
+It is worth noting that the conditional terms can also include numerical features.
+For example, it could be more accurate to learn instead a set of four first-degree terms, conditioned on the marital status and the years in work:
+$f_{1}(x_1 | x_2 < 10, x_3 = True)$,
+$f_{2}(x_1 | x_2 \geq 10, x_3 = True)$,
+$f_{3}(x_1 | x_2 < 10, x_3 = False)$, and
+$f_{4}(x_1 | x_2 \geq 10, x_3 = False)$, which can be visualized and interpreted as four one-dimensional plots.
+
+To adhere to the x-by-design principle, RAMs should be able to automatically identify the
+most significant conditional terms.
+As the number of these terms increases, it becomes difficult for users to retain and interpret numerous plots
+associated with each feature or pair of features.
+Therefore, RAMs use Regional Effect Plots~\citep{herbinger2023decomposing} to identify a small set
+of conditional terms that have the greatest impact in minimizing feature interactions.
+The RAM framework consists of three key steps.
+First, a black-box model is fitted to capture all high-order interactions.
+Then, the subregions where the black-box model exhibits near-local additivity are identified using Regional Effect Plots.
+Finally, a GAM component is fit to each identified subregion.
+
+The main contributions of this paper are as follows:
+
+\begin{itemize}
+    \item We formulate a new class of x-by-design models called Regionally Additive Models (RAMs).
+    \item We propose a generic framework for learning RAMs and we propose a novel method for identifying the most significant conditional terms.
+    \item We demonstrate the effectiveness of RAMs in modeling high-order interactions on a synthetic toy example and two real-world datasets.
+\end{itemize}
+
+\section{Motivation}
 \label{sec:motivation}
@@
 with \(x_1, x_2 \sim \mathcal{U}(-1,1)\) and \(x_3 \sim Bernoulli(0,1)\).
-Although very simple, a GAM would fail to learn this mapping due to
-the existence of the three-features interaction term $8x_2\when{x_1 > 0}\when{x_3=0}$.
+Although very simple, GAM and GA$^2$M would fail to learn this mapping due to
+the the three-features interaction term.
 As we see in Figure~\ref{subfig:global_gam}, a GAM misleadingly learns that $\hat{f}(\xb) \approx 2x_2$,
@@
 
-\section{The RAM framework}
+\section{RAM formulation}
+\label{sec:ram_formulation}
 
-The RAM framework consists of a three-step pipeline; (a) fit a black-box model (Section~\ref{subsec:fit_black_box}),
-(b) identify subregions with minimal interactions (Section~\ref{subsec:regional_effect_methods}) and
-(c) fit a GAM to each subregion (Section~\ref{subsec:fitting_gams}).
-Throughout this section, we will use the following notation.
+\paragraph{Notation.}
 Let \(\Xcal \in \Rd\) be the \(d\)-dimensional feature space, \(\Ycal\) the target space and \(f(\cdot) : \Xcal \rightarrow \Ycal\) the black-box function.
@@
 
-\subsection{First step: Fit a black-box function}
-\label{subsec:fit_black_box}
+The RAM consists of a three-step pipeline; (a) fit a black-box model (Section~\ref{subsec:fit_black_box}),
+(b) identify subregions with minimal interactions (Section~\ref{subsec:regional_effect_methods}) and
+(c) fit a GAM component to each subregion (Section~\ref{subsec:fitting_gams}).
 
-In the initial step of the pipeline, we fit a black-box function \(f(\cdot)\) to the training set
-\(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) to accurately learn the underlying mapping \(f(\cdot) : \Xcal \rightarrow \Ycal\).
-While any black-box function can theoretically be employed in this stage,
-for utilizing the DALE approximation, as we will show in the next step,
-it is necessary to select a differentiable function.
-Recent advancements have demonstrated that differentiable Deep Learning models,
-specifically designed for tabular data~\citep{arik2021tabnet}, are capable of achieving state-of-the-art performance,
-making them a suitable choice for this step.
-% Driver, teacher, guide
-
-\subsection{Second step: Find subregions}
-\label{subsec:regional_effect_methods}
-
-In this step, we use regional effect methods~\citep{herbinger2023decomposing, herbinger2022repid}
+In step (b), we use regional effect methods~\citep{herbinger2023decomposing, herbinger2022repid}
 to identify the regions where the black-box function is (close to) regionally additive.
@@
 Equations~\ref{eq:ram_formulation} and~\ref{eq:ram_formulation2} are equivalent.
-To gain a better understanding of the latter formulation, consider the toy example described in Section~\ref{sec:motivation}.
+To better understand of the formulations, consider the toy example described in Section~\ref{sec:motivation}.
 To minimize the impact of feature interactions, we need to divide feature \(x_2\) into two subregions,
 \(\mathcal{R}_{21} = \{x_1 > 0 \text{ and } x_3 = 1\}\) and \(\mathcal{R}_{22} = \{x_1 \leq 0 \text{ or } x_3 = 0\}\).
-This division results in an augmented feature space \(\Xcal^{\mathtt{RAM}} = (x_1, x_{21}, x_{22}, x_3)\)
-and a RAM formulation of the form: \(f^{\mathtt{RAM}}(\xb) = f_1(x_1) + f_{21}(x_{21}) + f_{22}(x_{22}) + f_3(x_3)\).
+Using Eq.~\ref{eq:ram_formulation}, RAM formulation is:
+\(f^{\mathtt{RAM}}(\xb) = f_1(x_1) + f_{21}(x_2) \when{x_1 > 0 \text{ and } x_3 = 1} + f_{22}(x_2) \when{x_1 \leq 0 \text{ or } x_3 = 0} + f_3(x_3)\).
+Using Eq.~\ref{eq:ram_feature_space}, we should first define the augmented feature space
+\(\Xcal^{\mathtt{RAM}} = (x_1, x_{21}, x_{22}, x_3)\),
+where \(x_{21} = x_2 \when{x_1 > 0 \text{ and } x_3 = 1}\) and \(x_{22} = x_2 \when{x_1 \leq 0 \text{ or } x_3 = 0}\)
+and then RAM formulation is: \(f^{\mathtt{RAM}}(\xb) = f_1(x_1) + f_{21}(x_{21}) + f_{22}(x_{22}) + f_3(x_3)\).
 
-\subsubsection{Proposed Approach}
+\section{RAM framework}
+\label{sec:ram_framework}
+
+\subsection{First step: Fit a black-box function}
+\label{subsec:fit_black_box}
+
+In the initial step of the pipeline, we fit a black-box function \(f(\cdot)\) to the training set
+\(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) to accurately learn the underlying mapping \(f(\cdot) : \Xcal \rightarrow \Ycal\).
+While any black-box function can theoretically be employed in this stage,
+for utilizing the DALE approximation, as we will show in the next step,
+it is necessary to select a differentiable function.
+Recent advancements have demonstrated that differentiable Deep Learning models,
+specifically designed for tabular data~\citep{arik2021tabnet}, are capable of achieving state-of-the-art performance,
+making them a suitable choice for this step.
+% Driver, teacher, guide
+
+\subsection{Second step: Find subregions}
+\label{subsec:regional_effect_methods}
 
@@
 
-\paragraph{DALE formulation}
+\paragraph{DALE}
 
@@
 and the dataset \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\),
-and returns the effect (impact) of a specific feature $s$ on the output $y$:
+and returns the effect (impact) of the $s$-th feature $s$ on the output $y$:
 %
@@
 In cases where there are strong interactions between the features,
-the instance-level effects inside each bin deviate from the average bin-effect.
-We can measure such deviation using the standard deviation of the instance-level effects inside each bin (bin-deviation):
+the instance-level effects inside each bin deviate from the average bin-effect (bin-deviation).
+We can measure such deviation using the standard deviation of the instance-level effects inside each bin:
 
@@
 %
-and the interaction between the feature \(x_s\) and the rest of the features along the whole \(s\)-th dimension
+The bin-deviation is a measure of the interaction between the feature \(x_s\) and the rest of the features inside the \(k\)-th bin.
+Therefore, we can measure the global interaction between the feature \(x_s\) and the rest of the features along the whole \(s\)-th dimension
 with the aggregated bin-deviation:
@@
 %
-Eq.~\eqref{eq:DALE_interaction} measures the interaction between the feature \(x_s\) and the rest of the features.
-It takes values in the range \([0, \infty)\) with zero indicating that \(x_s\) does not interact with the rest of the features,
+Eq.~\eqref{eq:DALE_interaction} outputs values in the range \([0, \infty)\) with zero indicating that \(x_s\) does not interact with any other feature,
 i.e., the underlying black box function can be written as $f(\xb) = f_s(x_s) + f_{/s}(x_{/s})$.
@@
 In particular, we start with a dense fixed-size grid of bins and we iteratively merge the neighboring bins with similar
-bin-effect and bin-deviation until all bins have at least $n_{\min}$ instances.
+bin-effect and bin-deviation until all bins have at least a minimum number of instances.
 In this way, we can have a more accurate approximation of the bin-effect and the bin-deviation.
@@
 
-In the same way that we can estimate the feature effect and the feature interactions for the $s$-th feature in the whole input space,
-using Eq.~\eqref{eq:DALE_accumulated_mean_est} and Eq.~\eqref{eq:DALE_interaction},
+In the same way that we can estimate the feature effect (Eq.~\eqref{eq:DALE_accumulated_mean_est}) and
+the feature interactions (Eq.~\eqref{eq:DALE_interaction}) for the $s$-th feature in the whole input space,
 we can also estimate the effect and the interactions in a subregion of the input space \(\mathcal{R}_{st} \subset \mathcal{X}\).
-We denote the equivalent regional qunatities as $f^{\mathtt{DALE}}_{\mathcal{R}_{st}}(x_s)$ and $\mathcal{H}_{\mathcal{R}_{st}}$.
+We denote the equivalent regional qunatities as $\hat{f}^{\mathtt{DALE}}_{st}(x_s)$ and $\mathcal{H}_{st}$.
 $f^{\mathtt{DALE}}_{\mathcal{R}_{st}}(x_s)$ and $\mathcal{H}_{\mathcal{R}_{st}}$ are defined exactly as in
 Eq.~\eqref{eq:DALE_accumulated_mean_est} and Eq.~\eqref{eq:DALE_interaction} respectively,
-with the only difference that instead of using the whole dataset \(\mathcal{D}\) to compute the regional bin-effect $\hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k)$
-and the regional bin-deviation $\hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)$,
-we use only the instances that belong to the subregion \(\mathcal{R}_{st}\), i.e. $\xb^i: x_s^i \in \mathcal{S}_k \land x_{s}^i \in \mathcal{R}_{st}$.
+with the only difference that instead of using the whole dataset \(\mathcal{D}\), to compute the regional bin-effect $\hat{\mu}_{st}(z_{k-1}, z_k)$
+and the regional bin-deviation $\hat{\sigma}_{st}^2(z_{k-1}, z_k)$,
+we use $\mathcal{D}_{st}$ which includes only the instances that belong to the subregion \(\mathcal{R}_{st}\),
+i.e., $\mathcal{D}_{st} = \{\xb^i: x_s^i \in \mathcal{S}_k \land \xc^i \in \mathcal{R}_{st}\}$.
 Therefore, in order to minimize the interactions of a particular feature $s$ we search for a set of regions \(\{\mathcal{R}_{st}\}_{t=1}^{T_s}\),
-that minimize:
+that minimizes the following objective:
 
@@
     & \underset{\{\mathcal{R}_{st}\}_{t=1}^{T_s}}{\text{minimize}}
-    & & \mathcal{L} = \sum_{t=1}^{T_s} \mathcal{H}_{\mathcal{R}_{st}} \\
+    & & \mathcal{L}_s = \sum_{t=1}^{T_s} \frac{| \mathcal{D}_{st} |}{|\mathcal{D}|} \mathcal{H}_{st} \\
     & \text{subject to}
@@
 %
+In Eq.~\eqref{eq:optimal_subregions}, the objective function is the weighted sum of the regional interactions $\mathcal{H}_{st}$,
+where the weights are the number of instances in each subregion.
+In this way, we give more importance to the subregions that contain more instances.
+The first constraint ensures that the subregions cover the whole input space and the second constraint ensures that the subregions are disjoint.
 
@@
 
-To minimize the objective of Eq.~\eqref{eq:SE}, we have developed a tree-based algorithm based on the approach proposed
-by~\citep{herbinger2023decomposing}.
-The core of the algorithm is outlined in Algorithm~\ref{alg:subregion_detection}.
-We search for subregions independently for each feature $s \in \{1, \ldots, D\}$.
-and we select the $T$ optimal splits, in a greedy manner.
-To better comprehend the algorithm, let's consider the illustrative example of Section~\ref{sec:motivation}.
-For feature $s=2$, the algorithm starts with the first level of the tree, with $/s = \{1, 3\}$ as candidate split-features.
-Algorithm~\ref{alg:single_feature_subregion} identifies the optimal first-level split, with the following procedure.
-Initially, it determines the candidate split points for each feature.
+The core of the method is outlined in Algorithm~\ref{alg:ram}.
+First, we fit a differentiable black box model to the data (Step 1) and we copute the Jacobian matrix w.r.t. the input features (Step 2).
+Then we search for a set of subregions by minimizing the objective of Eq.~\eqref{eq:optimal_subregions}
+for each feature $s$ independently (Steps 3-4-5).
+Based on the optimal subregions, we define the extended feature space (Step 6) and we fit a GAM in the extended feature space (Step 7).
 
-Since $x_1$ is a continuous feature, are generated as a linearly spaced grid of $P$ points within the range of the feature, i.e. $[-1, 1]$,
-where $P$ is a hyperparameter of the algorithm, set to 10 in the experiments.
-Consequently, the candidate split points $p \in \{-1, -0.8, -0.6, \ldots, 0.8, 1\}$ define the candidate subregions
-$\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_1 \leq p \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_1 > p \}$.
+For solving Eq.~\eqref{eq:optimal_subregions}, we have developed a tree-based algorithm
+based on the approach proposed by~\citep{herbinger2023decomposing}, which we describe in detail in Algorithm~\ref{alg:subregion_detection}.
+To describe the algorithm, we define some additional notation: $\mathcal{R}_s^l$ is the set of optimal subregion of the $s$-th feature at level $l$ of the tree.
+Since at each level of the tree we divide the input space into two subregions, at level $l$ we have $2^l$ subregions,
+i.e., $\mathcal{R}_s^l = \{\mathcal{R}_{st}\}_{t=1}^{2^l}$.
+Equivalently, $\mathcal{L}_s^l$ is the optimal objective value of Eq.~\eqref{eq:optimal_subregions} at level $l$ of the tree.
+Although the algorithm can search for an abritrary number of subregions per feature,
+in order to preserve the smooth interpretation of the method,
+we limit the maximum depth of the tree to $L=3$ levels,
+which stands for a maximum of $T = 2^L = 8$ subregions per feature.
+In general, the user can control the trade-off between the interpretability and the accuracy of the method by changing the maximum depth of the tree.
+Note that with three splits, we already have an interaction term of
+four ($f(x_s | \when{x_{c_1}}, \when{x_{c_2}}, \when{x_{c_3}}$) or
+five ($f(x_{s_1}, x_{s_{2}} | \when{x_{c_1}}, \when{x_{c_2}}, \when{x_{c_3}}$) features.
+
+To describe how the algorithm finds the optimal splits at each level $l$, let's consider the illustrative example of Section~\ref{sec:motivation}.
+For feature $s=2$, the algorithm starts with $/s = \{1, 3\}$ as candidate split-features for the first level of the tree.
+For each candidate split-feature, the algorithm determines the candidate split positions.
+Since $x_1$ is a continuous feature, the candidate splits postions are a linearly spaced grid of $P$ points within the range of the feature, i.e. $[-1, 1]$,
+where $P$ is a hyperparameter of the algorithm, set to $10$ in the experiments.
+Therefore, the candidate positions are $p \in \{-1, -0.8, -0.6, \ldots, 0.8, 1\}$ each on defining two subregions,
+$\mathcal{R}_{21} = \{ (x_1, x_3) : x_1 \leq p \}$ and $\mathcal{R}_{22} = \{ (x_1, x_3) : x_1 > p \}$.
 As for $x_3$, being a categorical feature, the candidate split points are its unique values, i.e., $\{0, 1\}$,
-and the corresponding subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 \neq 0 \}$.
-
-For each candidate position, the algorithm splits the initial dataset into two subsets,
-$[\mathcal{D}_1, \mathcal{D}_2]$ one for each subregion and computes the level of interaction
-as the weighted sum of the interactions of the two subregions, as defined by Algorithm~\ref{alg:get_interaction}.
-After the iteration over all points and all candidate positions,
-the split point that minimizes the weighted level of interactions is selected.
-In the illustrative example, the optimal first-level subregions are
-$\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 \neq 0 \}$.
-The algorithm then proceeds to the second level of the tree, where the only candidate feature is $x_3$.
-In this step, the first split is considered fixed, i.e., the algorithm uses a list of two datasets,
-and searches in an analogous manner for the optimal second-level split that will be applied on top of the first split,
-leading to four subregions.
-The algorithm proceeds in a similar manner, until it reaches the maximum depth $T$.
-
-The output of algorithm~\ref{alg:subregion_detection} is a collection of $T$ splits for each feature $s$.
-It is important to note that not all splits lead to a significant decrease in the level of interactions.
-To address this, we post-process the splits by pruning any splits that do not result in a decrease of the
-weighted interaction greater than a threshold value $\epsilon$ (set to $20\%$ drop in the experiments).
-The pruned set of splits is denoted as $\{\mathcal{R}_{st}\}_{t=1}^{T_s}$.
+and the corresponding subregions are $\mathcal{R}_{21} = \{ (x_1, x_3) : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ (x_1, x_3) : x_3 \neq 0 \}$.
+Each candidate position, creates a corresponding dataset $[\mathcal{D}_{21}, \mathcal{D}_{22}]$,
+and the algorithm computes the weighted level of interactions $\mathcal{H}_{21}$ and $\mathcal{H}_{22}$ for each dataset.
+After iterating over all features and all candidate positions for each feature,
+it selects the split point that minimizes the weighted level of interactions.
+In the illustrative example, the optimal first-level split is based on $x_3$ and the optimal split point is $p = 0$.
+The algorithm next proceeds to the second level, where the only candidate feature is $x_3$.
+In this step, the first split is considered fixed so the optimal second split is applied to the subregions $\mathcal{R}_{21}$ and $\mathcal{R}_{22}$,
+creating four subregions in total.
+The algorithm continues in a similar manner, until it reaches the maximum depth $T$ or the drop in the weighted level of interactions is below a threshold $\epsilon$ (set to $20\%$ drop in the experiments).
 
@@
   Train a differentiable black box model $f$ using $(X, y)$\;
-  \tcc{Detect subregions for all features}
-  $\left\{\mathcal{R}_{st}|s\in\left\{1,\ldots,D\right\},
-  t\in\left\{1,\ldots,T_s\right\}\right\}$ = DetectSubregions($X$, $y$, $f$, $T$)\;
+  Compute the Jacobian w.r.t. features $\xb$,  $J = \nabla_{\xb}f(\xb)$ \;
+  \For{$s \in \{1, \ldots, D\}$}{
+    $\left\{\mathcal{R}_{st}\right\}_{t=1}^{T_s}$ = DetectSubregions($X$, $J$, $T$, $s$)\;
+  }
   Create the extended feature space $\mathcal{X}^{\mathtt{RAM}}$ using all
@@
 \begin{algorithm}
-\caption{Detection of Subregions using DALE}
+\caption{DetectSubregions}
 \label{alg:subregion_detection}
@@
 \BlankLine
-\Input{Dataset $(X, y)$, Black box model $f$, Maximum depth $T$}
+\Input{Dataset $X$, Gradients $J$, Maximum depth $T$, Feature $s$}
 \BlankLine
-\Output{Subregions $\mathcal{R}_{sT_s}$ for all features $s$, where $T_s \leq 2^T$}
+\Output{Subregions $\{\mathcal{R}_{st}\}_{t=1}^{T_s}$, where $T_s \leq 2^T$}
 \BlankLine
-\tcc{Needed for DALE feature effect estimation}
-Compute the Jacobian w.r.t. features $\xb$,  $J = \nabla_{\xb}f(\xb)$ \;
-\For{$s = 1$ to $D$}{
-  \For{$t = 1$ to $T$}{
-    \tcc{Find best split feature $c$ at point $p$, leading to loss
-      $\mathcal{H}_{st}$ using regions of previous level}
-    $\mathcal{H}_{st}, c, p$ $\gets$
-    BestSplit($\left\{\mathcal{R}_{sk}|k\in\left\{1,\ldots,T_{t-1}\right\}\right\}$)\;
-    \If{$\lVert\mathcal{H}_{st} - \mathcal{H}_{s(t-1)}\rVert < \epsilon$}{
+$\mathcal{H}_s^0$ \tcp*{Compute the level of interactions before any split}
+$T_s = 0$ \tcp*{Initialize the number of splits for feature $s$}
+\For{$l = 1$ to $L$}{
+    \If{H_{s}^{l-1} = 0}{
         break\;
     }
+    \tcc{Find best split feature $c_s^l$ at point $p_s^l$, leading to loss
+        $\mathcal{H}_s^l$ using regions of previous level}
+    Find $\mathcal{H}_{st}, c, p$ of the optimal split based on $\mathcal{R}_s^l$ \;
+    \If{$1 - \frac{\mathcal{H}_s^l}{\mathcal{H}_s^{l-1}} > \epsilon$}{
+        break\;
     }
+    $T_s = 2^l$ \tcp*{Update the number of splits for feature $s$}
 }
@@
 
-\begin{algorithm}
-\caption{BestSplit}
-\label{alg:single_feature_subregion}
-\SetAlgoLined
-\SetKwInOut{Input}{Input}
-\SetKwInOut{Output}{Output}
-
-\Input{X\_list, J\_list, s, c, is\_cat}
-\Output{BestSplits}
-
-% Initialize variables
-\BlankLine
-L[c, p] $\gets$ None\; % \Comment{List to store interaction values}
-\For{c $s = 1$ to $D$ \textnormal{if} c $\neq$ s}{
-    is\_cat $\gets$ IsCategorical(X\_list, c)\;
-    positions $\gets$ GetPositions(X\_list, c, is\_cat)\;
-    L\_pos $\gets$ []\; % \Comment{List to store interaction values}
-    \For{p \textnormal{in} positions}{
-        X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c, p, is\_cat)\;
-        L $\gets$ GetInteraction(X\_split, J\_split)\;
-    }
-}
-L\_min $\gets$ min(L)\;
-c\_min, p\_min = argmin(L)\;
-is\_cat $\gets$ IsCategorical(X\_list, c\_min)\;
-X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c\_min, p\_min, is\_cat)\;
-
-\Return{L\_min, p\_min c\_min, X\_split, J\_split}
-\BlankLine
-\end{algorithm}
-
-\begin{algorithm}
-\caption{GetInteraction}
-\label{alg:get_interaction}
-\SetAlgoLined
-\SetKwInOut{Input}{Input}
-\SetKwInOut{Output}{Output}
-\Input{$X\_list$, $J\_list$, $min\_points$}
-\Output{weighted average of interaction levels}
-N $\gets$ total number of items in $X\_list$\;
-W $\gets$ []\;
-L $\gets$ []\;
-\For{$i$ in len($X\_list$)}{
-    X $\gets$ X\_list[i]\;
-    J $\gets$ J\_list[i]\;
-    \uIf{$|X|$ < min\_points}{
-      Append $\infty$ to L\;
-    }
-    \Else{
-        Append $\mathcal{L}(X, J)$ to $L$\;
-    }
-    Append $\frac{|X|}{N}$ to $W$\;
-}
-\Return $\text{sum}(L \cdot W)$\;
-\end{algorithm}
+%\begin{algorithm}
+%\caption{BestSplit}
+%\label{alg:single_feature_subregion}
+%\SetAlgoLined
+%\SetKwInOut{Input}{Input}
+%\SetKwInOut{Output}{Output}
+%
+%\Input{X\_list, J\_list, s, c, is\_cat}
+%\Output{BestSplits}
+%
+%% Initialize variables
+%\BlankLine
+%L[c, p] $\gets$ None\; % \Comment{List to store interaction values}
+%\For{c $s = 1$ to $D$ \textnormal{if} c $\neq$ s}{
+%    is\_cat $\gets$ IsCategorical(X\_list, c)\;
+%    positions $\gets$ GetPositions(X\_list, c, is\_cat)\;
+%    L\_pos $\gets$ []\; % \Comment{List to store interaction values}
+%    \For{p \textnormal{in} positions}{
+%        X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c, p, is\_cat)\;
+%        L $\gets$ GetInteraction(X\_split, J\_split)\;
+%    }
+%}
+%L\_min $\gets$ min(L)\;
+%c\_min, p\_min = argmin(L)\;
+%is\_cat $\gets$ IsCategorical(X\_list, c\_min)\;
+%X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c\_min, p\_min, is\_cat)\;
+%
+%\Return{L\_min, p\_min c\_min, X\_split, J\_split}
+%\BlankLine
+%\end{algorithm}
+%
+%\begin{algorithm}
+%\caption{GetInteraction}
+%\label{alg:get_interaction}
+%\SetAlgoLined
+%\SetKwInOut{Input}{Input}
+%\SetKwInOut{Output}{Output}
+%\Input{$X\_list$, $J\_list$, $min\_points$}
+%\Output{weighted average of interaction levels}
+%N $\gets$ total number of items in $X\_list$\;
+%W $\gets$ []\;
+%L $\gets$ []\;
+%\For{$i$ in len($X\_list$)}{
+%    X $\gets$ X\_list[i]\;
+%    J $\gets$ J\_list[i]\;
+%    \uIf{$|X|$ < min\_points}{
+%      Append $\infty$ to L\;
+%    }
+%    \Else{
+%        Append $\mathcal{L}(X, J)$ to $L$\;
+%    }
+%    Append $\frac{|X|}{N}$ to $W$\;
+%}
+%\Return $\text{sum}(L \cdot W)$\;
+%\end{algorithm}
 
@@
 
-Algorithm~\ref{alg:subregion_detection} has a computational complexity of $\mathcal{O}(D \cdot T)$
-as it iterates over all features and levels of the tree.
-Algorithm~\ref{alg:single_feature_subregion} has a computational complexity of $\mathcal{O}(D \cdot P \cdot N)$
-as it iterates over all features, query positions, and performs indexing operations on the data (SplitDataset and GetInteraction).
-The use of DALE eliminates the need to compute the Jacobian matrix for each split,
-resulting in computational complexity independent of the model complexity.
-Considering the entire algorithm, the computational complexity is $\mathcal{O}(D \cdot (D-1) \cdot T \cdot P \cdot N)$.
+Algorithm~\ref{alg:subregion_detection} has a computational complexity of $\mathcal{O}(D-1 \cdot L \cdot N)$
+as it iterates over all features, query positions, and performs indexing operations on the data (splitting the dataset and computing the level of interactions).
+Algorithm~\ref{alg:subregion_detection} is applied to each feature $s$ independently, and so computational complexity of the entire algorithm is $\mathcal{O}(D \cdot (D-1) \cdot L \cdot N)$.
 However, in practice, $P$ and $T$ are small numbers.
@@
 making it suitable for large datasets, heavy models, and reasonably high-dimensional data.
-In practice, for a typical scenario with $N=10 \times 10^3$ samples and $D=20$ features, the proposed method takes less than $10$ seconds to run.
+The key point is that the use of DALE eliminates the need to compute the Jacobian matrix for each split,
+which is the most computationally expensive step.
+This is because the Jacobian matrix is computed only once for the entire dataset, and then it is used as a lookup table for computing the level of interactions for each split.
+This makes the proposed method applicable to heavy models.
 

### 2023-06-29 report/main_paper.tex
latest paper

--- 2023-06-23 report/report.tex
@@
-\documentclass[12pt]{article}
+% This is samplepaper.tex, a sample chapter demonstrating the
+% LLNCS macro package for Springer Computer Science proceedings;
+% Version 2.21 of 2022/01/12
+%
+\documentclass[runningheads]{llncs}
+%
+\usepackage[T1]{fontenc}
+% T1 fonts will be used to generate the final print and online PDFs,
+% so please use T1 fonts in your manuscript whenever possible.
+% Other font encondings may result in incorrect characters.
+%
+\usepackage{graphicx}
+% Used for displaying a sample figure. If possible, figure files should
+% be included in EPS format.
+%
+% If you use the hyperref package, please uncomment the following two lines
+% to display URLs in blue roman font according to Springer's eBook style:
+%\usepackage{color}
+%\renewcommand\UrlFont{\color{blue}\rmfamily}
+%
 \usepackage{amsfonts}
@@
 \usepackage{bbm}
-\usepackage{graphicx}
 \usepackage{geometry}[margin=1in]
@@
 \usepackage{natbib}
-\usepackage{multicol}
 \usepackage{algpseudocode}
@@
 % set path for bibliography
-
-\title{Regionally Additive Models: Explainable-by-design models minimizing feature interactions}
 
@@
 
-\author{Vasilis Gkolemis}
-
 \begin{document}
-    \def\MakeUppercaseUnsupportedInPdfStrings{\scshape}
-\maketitle
-
+%
+\title{Regionally Additive Models: Explainable-by-design models minimizing feature interactions}
+%
+%\titlerunning{Abbreviated paper title}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+%
+\author{
+%    First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
+%Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
+%Third Author\inst{3}\orcidID{2222--3333-4444-5555}
+}
+%
+\authorrunning{F. Author et al.}
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+%
+\institute{
+%    Princeton University, Princeton NJ 08544, USA \and
+%Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
+%\email{lncs@springer.com}\\
+%\url{http://www.springer.com/gp/computer-science/lncs} \and
+%ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
+%\email{\{abc,lncs\}@uni-heidelberg.de}
+}
+%
+\maketitle              % typeset the header of the contribution
+%
 \begin{abstract}
-Generalized Additive Models (GAMs) are widely utilized explainable-by-design models in various applications.
-However, their assumption of independence among features can lead to suboptimal performance when violated.
-To overcome this limitation, we introduce Regionally Additive Models (RAMs), a novel class of explainable-by-design models.
-RAMs mitigate the issue by identifying subregions in the feature space where interactions are minimized and fitting multiple GAMs accordingly.
+Generalized Additive Models (GAMs) are widely used explainable-by-design models in various applications.
+GAMs assume that the output can be represented as a sum of univariate functions, referred to as components.
+However, this assumption fails in ML problems where the output depends on multiple features simultaneously.
+In these cases, GAMs fail to capture the interaction terms of the underlying function, leading to subpar accuracy.
+To (partially) address this issue, we propose Regionally Additive Models (RAMs), a novel class of explainable-by-design models.
+RAMs identify subregions within the feature space where interactions are minimized.
+Within these regions, it is more accurate to express the output as a sum of univariate functions (components).
+Consequently, RAMs fit one component per subregion of each feature instead of one component per feature.
+This approach yields a more expressive model compared to GAMs while retaining interpretability.
 The RAM framework consists of three steps.
@@
 Secondly, using Regional Effect Plots, we identify subregions where the black-box model exhibits near-local additivity.
-In these subregions, the effect of each feature on the target is independent of the values of other features.
-Lastly, we fit a GAM specifically for each identified subregion.
-We validate the effectiveness of RAMs through experiments on both a synthetic example and real-world datasets.
+Lastly, we fit a GAM component for each identified subregion.
+We validate the effectiveness of RAMs through experiments on both synthetic and real-world datasets.
 The results confirm that RAMs offer improved expressiveness compared to GAMs while maintaining interpretability.
+\keywords{Explainable AI \and Generalized Additive models \and x-by-design}
 \end{abstract}
-
+%
+%
+%
 \section{Introduction}
+\label{sec:intro}
 
@@
 \For{$l = 1$ to $L$}{
-    \If{H_{s}^{l-1} = 0}{
+    \If{$H_{s}^{l-1} = 0$}{
         break\;
@@
 \end{algorithm}
-
-%\begin{algorithm}
-%\caption{BestSplit}
-%\label{alg:single_feature_subregion}
-%\SetAlgoLined
-%\SetKwInOut{Input}{Input}
-%\SetKwInOut{Output}{Output}
-%
-%\Input{X\_list, J\_list, s, c, is\_cat}
-%\Output{BestSplits}
-%
-%% Initialize variables
-%\BlankLine
-%L[c, p] $\gets$ None\; % \Comment{List to store interaction values}
-%\For{c $s = 1$ to $D$ \textnormal{if} c $\neq$ s}{
-%    is\_cat $\gets$ IsCategorical(X\_list, c)\;
-%    positions $\gets$ GetPositions(X\_list, c, is\_cat)\;
-%    L\_pos $\gets$ []\; % \Comment{List to store interaction values}
-%    \For{p \textnormal{in} positions}{
-%        X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c, p, is\_cat)\;
-%        L $\gets$ GetInteraction(X\_split, J\_split)\;
-%    }
-%}
-%L\_min $\gets$ min(L)\;
-%c\_min, p\_min = argmin(L)\;
-%is\_cat $\gets$ IsCategorical(X\_list, c\_min)\;
-%X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c\_min, p\_min, is\_cat)\;
-%
-%\Return{L\_min, p\_min c\_min, X\_split, J\_split}
-%\BlankLine
-%\end{algorithm}
-%
-%\begin{algorithm}
-%\caption{GetInteraction}
-%\label{alg:get_interaction}
-%\SetAlgoLined
-%\SetKwInOut{Input}{Input}
-%\SetKwInOut{Output}{Output}
-%\Input{$X\_list$, $J\_list$, $min\_points$}
-%\Output{weighted average of interaction levels}
-%N $\gets$ total number of items in $X\_list$\;
-%W $\gets$ []\;
-%L $\gets$ []\;
-%\For{$i$ in len($X\_list$)}{
-%    X $\gets$ X\_list[i]\;
-%    J $\gets$ J\_list[i]\;
-%    \uIf{$|X|$ < min\_points}{
-%      Append $\infty$ to L\;
-%    }
-%    \Else{
-%        Append $\mathcal{L}(X, J)$ to $L$\;
-%    }
-%    Append $\frac{|X|}{N}$ to $W$\;
-%}
-%\Return $\text{sum}(L \cdot W)$\;
-%\end{algorithm}
 
@@
 
-We evaluate the proposed approach on two typical tabular datasets, namely the Bike-Sharing Dataset~\citep{misc_bike_sharing_dataset_275}
-and the California Housing Dataset~\citep{pace1997sparse}.
+We evaluate the proposed approach on two typical tabular datasets:
+the Bike-Sharing Dataset~\citep{misc_bike_sharing_dataset_275} and the California Housing Dataset~\citep{pace1997sparse}.
 
@@
   \centering
-  \caption{The table compares RMSE values of different models on two datasets: Bike-Sharing and California Housing.
-  The models include DNN, GAM, RAM, GAM², and RAM² (representing 2nd order interactions).
+  \caption{The table compares the Mean Absolute Error (MAE) and the Root Mean Square Error (RMSE) of
+  DNN, GAM, RAM, GA$^2$M, and RA$^2$M (representing 2nd order interactions), on two datasets: Bike-Sharing and California Housing.
   Lower values indicate better performance.
@@
   \begin{tabular}{l|c|cccc}
-      \hline
       & \textbf{Black-box} & \multicolumn{4}{c}{\textbf{x-by-design}} \\
@@
       \hline
-      & \textbf{DNN} & \textbf{GAM} & \textbf{RAM} & \textbf{GAM}$^2$ & \textbf{RAM}$^2$ \\
+      & \textbf{DNN} & \textbf{GAM} & \textbf{RAM} & \textbf{GA}$^2$\textbf{M} & \textbf{RA}$^2$\textbf{M} \\
       \hline
-      Bike Sharing      & 0.254 & 0.549 & 0.430 & 0.298 & 0.278 \\
+      Bike Sharing (MAE)  & 0.254 & 0.549 & 0.430 & 0.298 & 0.278 \\
+      Bike Sharing (RMSE) & 0.389 & 0.734 & 0.563 & 0.438 & 0.412 \\
       \hline
-      California Housing      & 0.369 & 0.600 & 0.535 & 0.554 & 0.514 \\
+      California Housing (MAE)  & 0.373 & 0.600 & 0.553 & 0.554 & 0.533 \\
+      California Housing (RMSE) & 0.533 & 0.819 & 0.754 & 0.774 & 0.739 \\
   \end{tabular}
@@
 As a black-box model, we train for \(60\) epochs a fully-connected Neural Network with 6 hidden layers, using the Adam optimizer with a learning rate of $0.001$.
-The model attains a root mean squared error of \( 0.25 \cdot 181 \approx 45\) counts on the test set.
+The model attains a root mean squared error of \( 0.39 \cdot 181 \approx 70\) counts on the test set.
 Subsequently, we extract the subregions, searching for splits up to a maximum spliting depth of \(T=3\).
@@
 Specifically, during working days, there is higher demand for bike rentals in the morning and afternoon hours,
-which aligns with the typical commuting times (\ref{subfig:bike_rentals_regional_gam_1}).
+which aligns with the typical commuting times (Figure~\ref{subfig:bike_rentals_regional_gam_1}).
 On the other hand, during non-working days, bike rentals peak in the afternoon as individuals engage in
-leisure activities (\ref{subfig:bike_rentals_regional_gam_2}).
+leisure activities (Figure~\ref{subfig:bike_rentals_regional_gam_2}).
 The proposed RAM method effectively captures and detects this interaction by establishing two distinct subregions,
@@
 Subsequently, the EBM that is fitted to each subregion, successfully learns these patterns,
-achieving a root mean squared error of approximately \( 0.43 \cdot 181 \approx 77\) counts on the test set.
+achieving a root mean squared error of approximately \( 0.56 \cdot 181 \approx 101\) counts on the test set.
 It is noteworthy that RAM not only preserves the interpretability of the model,
@@
 In contrast, the GAM model~\ref{subfig:bike_rentals_gam} is not able to capture this interaction and
-achieves a root mean squared error of \( 0.55 \cdot 181 \approx 100\) counts on the test set.
-Finally, in table~\ref{tab:sample}, we also observe that the RAM$^2$, i.e., RAM with second-order interactions,
-outperforms the equivalent GAM$^2$ model in terms of predictive performance.
+achieves a root mean squared error of \( 0.73 \cdot 181 \approx 132\) counts on the test set.
+Finally, in table~\ref{tab:sample}, we also observe that the RA$^2$M, i.e., RAM with second-order interactions,
+outperforms the equivalent GA$^2$M model in terms of predictive performance.
+Specifically, the RA$^2$M model achieves a root mean squared error of \( 0.41 \cdot 181 \approx 74\) counts,
+while the GA$^2$M model of \( 0.44 \cdot 181 \approx 80\) counts on the test set.
+It is worth noticing that the RA$^2$M model's accuracy is comparable to the black-box model's accuracy.
 
@@
 using the Adam optimizer with a learning rate of $0.001$.
-The model achieves a root mean square error (RMSE) of about \(40\)K dollars on the test set.
+The model achieves a root mean square error (RMSE) of about \(58\)K dollars on the test set.
 Subsequently, we perform subregion extraction by searching for splits up to a maximum depth of \(T=3\).
@@
 it does not align perfectly with the regional effect.
-As in the Bike-Sharing Example, the RMSE of the RAM model, i.e. \( 0.53 \cdot 110 \cdot 10^3 \approx 58.3 \cdot 10^3\) dollars on the test set,
-is lower than the one of the GAM model, i.e.\( 0.6 \cdot 110 \cdot 10^3 \approx 66 \cdot 10^3\) dollars.
+As in the Bike-Sharing Example, the RMSE of the RAM model, i.e. \( 0.75 \cdot 110 \approx 82.5\)K dollars on the test set,
+is lower than the one of the GAM model, i.e.\( 0.82 \cdot 110 \approx 90\)K dollars.
 These results indicate that the RAM model provides superior predictions compared to the GAM model.
-The same conclusion holds is when comparing the RAM$^2$ and the GAM$^2$ models.
+The same conclusion holds is when comparing the RA$^2$M and the GA$^2$M models,
+where the former achieves a RMSE of \( 0.74 \cdot 110 \approx 81\)K dollars,
+while the latter of \( 0.77 \cdot 110 \approx 85\)K dollars.
+
 \begin{table}[htbp]
@@
 Another important area of investigation involves exploring the impact of second-order effects within the RAM framework.
-While our experimenation demonstrated that even with the current subregion detection, RAM$^2$s outperform GAM$^2$s,
+While our experimenation demonstrated that even with the current subregion detection, RA$^2$Ms outperform GA$^2$Ms,
 it may be the case, that for second-order models the optimal subregions are not necessarily those that maximize the additive effect of individual features,
@@
 
-%\section{Appendix}
-%
-%\subsection{Regional DALE formulation}
-%\label{sec:regional_dale}
-%
-%Below we provide the regional DALE formulation for the bin-effect, the bin-deviation and the aggregated deviation on a subregion \(\mathcal{R}_{st}\):
-%
-%\begin{equation}
-%    \label{eq:bin_effect_subregion}
-%    \hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}|} \sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \dfdx(\mathbf{x}^i)
-%\end{equation}
-%
-%\begin{equation}
-%  \label{eq:bin_deviation_subregion}
-%  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
-%\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
-%    \hat{\mu}(z_{k-1}, z_k) \right )^2
-%\end{equation}
-%
-%\begin{equation}
-%  \label{eq:SE}
-%  \mathcal{H}_{\mathcal{R}_{sk}} = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
-%\end{equation}
-
-%\subsection{Algorithmic Details of Subregion Detection}
-%
-%\begin{algorithm}
-%\caption{SplitDataset}
-%\label{alg:split_dataset}
-%\SetAlgoLined
-%\SetKwInOut{Input}{Input}
-%\SetKwInOut{Output}{Output}
-%\Input{X\_list, J\_list, c, val, is\_cat}
-%\Output{X\_split, J\_split}
-%X\_split $\gets$ []\;
-%J\_split $\gets$ []\;
-%\For{i = 1 to len(X\_list)}{
-%    X= X\_list[i]\;
-%    J= J\_list[i]\;
-%    \uIf{is\_cat}{
-%        ind\_1 $\gets$ X[:, c] $=$ val\;
-%        ind\_2 $\gets$ X[:, c] $\neq$ val\;
-%    }
-%    \Else{
-%        ind\_1 $\gets$ X[:, c] $\leq$ val\;
-%        ind\_2 $\gets$ X[:, c] $>$ val\;
-%    }
-%    Append X[ind\_1], X[ind\_2] to X\_split\;
-%    Append J[ind\_1], J[ind\_2] to J\_split\;
-%}
-%\Return X\_split, J\_split
-%\end{algorithm}
+\subsubsection{Acknowledgements} Please place your acknowledgments at
+the end of the paper, preceded by an unnumbered run-in heading (i.e.
+3rd-level heading).
 
